---
title: "NK project: Preliminary analysis"
author: "Sara Wernig-Zorc"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input: "/home/rstudio/mnt_data/"
  out: "/home/rstudio/mnt_out/"
  data: "/home/rstudio/mnt_data/"
  nb_tumors: "/home/rstudio/mnt_nb_tumors/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
---

Fetahu et al., 2023 scRNA-seq
Group I: non-tumor-infiltrated BM
Group II: MYCNA 
Group III: ATRXmutated
Group IV: Sporadic

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r R packages: load, include=FALSE, echo=FALSE}
suppressPackageStartupMessages(library("Seurat"))
suppressPackageStartupMessages(library("SingleCellExperiment"))
suppressPackageStartupMessages(library("readxl"))
suppressPackageStartupMessages(library("SCpubr"))
suppressPackageStartupMessages(library("scCustomize"))
suppressPackageStartupMessages(library("ggVennDiagram"))
suppressPackageStartupMessages(library("statmod"))
suppressPackageStartupMessages(library("miloR"))
suppressPackageStartupMessages(library("scran"))
suppressPackageStartupMessages(library("scater"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("patchwork"))
suppressPackageStartupMessages(library("RColorBrewer"))

source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
```

# __load integrated data__

```{r}
# input data - integrated data from Fetahu et al., 2023
scrna_data <- readRDS(paste0(params$out,
                             "/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"))
metadata <- readRDS(paste0(params$out,
                           "/scRNA-seq_stage_M/results/published/data/metadata.rds"))

rna <- as.Seurat(scrna_data, data=NULL)
rna <- RenameAssays(object = rna, originalexp = 'RNA')

# set cell type annotation to the original annotation
rna@active.ident <- setNames(pull(metadata[match(metadata$cell,colnames(rna)), "cellont_cluster"], "cellont_cluster"),
                             pull(metadata[match(metadata$cell,colnames(rna)),  "cell"], "cell"))

rna@meta.data$orig.ident <- rna@active.ident

# set assay to RNA 
DefaultAssay(rna) <- "RNA"
```

# __subset object to NK cells only__

```{r}
levels(Idents(rna))

# subset metadata
filter <- metadata$cellont_abbr == "NK"

# subset scRNA-seq data 
# to NK cells only

NK_cells <- "NK (4)" # change cluster number
fetahu.NK <- subset(rna, idents = NK_cells, invert = FALSE)
fetahu.NK

Idents(fetahu.NK) <- fetahu.NK@meta.data$sample
fetahu.NK <- RenameIdents(fetahu.NK, 
                    "2005_1702" = "ATRXmut_01", 
                    "2006_2684" = "sporadic_01",
                    "2014_0102" = "ctrl_01",
                    "2016_1853" = "ctrl_02",
                    "2016_2950" = "ctrl_03",
                    "2016_3924" = "ATRXmut_02",
                    "2016_4503" = "MYCNA_01",
                    "2018_1404" = "MYCNA_02",
                    "2018_1625" = "sporadic_02",
                    "2018_4252" = "ctrl_04",
                    "2018_6056" = "sporadic_03",
                    "2019_2495" = "sporadic_04",
                    "2019_5022" = "MYCNA_03",
                    "2019_5754" = "MYCNA_04",
                    "2020_1288" = "ctrl_05",
                    "2020_1667" = "sporadic_05")

fetahu.NK@meta.data$patients <- Idents(fetahu.NK)
```

# __Yang et al., 2019__

```{r}
yang2019_clust_mark_full <- read_excel("/home/rstudio/mnt_resources/NK_cells_gene_sets/Yang2019_supp_Table_1.xlsx")
yang2019_clust_mark <- yang2019_clust_mark_full$gene
length(yang2019_clust_mark)

yang2019_clust_filt <- filter(yang2019_clust_mark_full, 
                              avg_logFC > 0.2) 

yang2019_clust1 <- filter(yang2019_clust_filt, 
                          cluster == "1") # 149
yang2019_clust2 <- filter(yang2019_clust_filt, 
                          cluster == "2") # 21
yang2019_clust3 <- filter(yang2019_clust_filt, 
                          cluster == "3") # 65
yang2019_clust4 <- filter(yang2019_clust_filt, 
                          cluster == "4") # 6
yang2019_clust5 <- filter(yang2019_clust_filt, 
                          cluster == "5") # 143
yang2019_clust6 <- filter(yang2019_clust_filt, 
                          cluster == "6") # 79
yang2019_clust7 <- filter(yang2019_clust_filt, 
                          cluster == "7") # 65
```

```{r eval=FALSE, include=FALSE}
# loop through clusters 1-7
for (x in 1:7) {
  print(paste0("Processing cluster ", x))
  filter <- yang2019_clust_filt$cluster == x
  y <- yang2019_clust_filt[filter,]
  # write the gene list per cluster
  write.table(y$gene, sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE,
              file = paste0("/home/rstudio/mnt_resources/NK_cells_gene_sets/yang2019_cluster",x,".csv"))
}
```

```{r}
# normalize and scale count data
fetahu.NK <- fetahu.NK %>%
  NormalizeData() %>%
  ScaleData() %>%
  FindVariableFeatures(nfeatures = 8000)

# principal component analysis using 1088 genes from yang2019
fetahu.NK <- RunPCA(fetahu.NK, 
                    assay = "RNA", 
                    npcs = 20,
                    features = yang2019_clust_mark,
                    seed.use = 2805,
                    reduction.name = "pca_new")

# PCA plot
DimPlot(object = fetahu.NK,
        reduction = "pca_new",
        group.by = "group")

# find closest cells based on gene expression of the yang2019_clust_mark genes
fetahu.NK <- FindNeighbors(fetahu.NK, 
                           features = yang2019_clust_mark,
                           reduction = "pca_new", 
                           dims = 1:10,
                           k.param = 100)

# find clusters based on nearest neighbours
fetahu.NK <- FindClusters(fetahu.NK, 
                          resolution = 0.2,
                          random.seed = 28,
                          group.singletons = TRUE,
                          cluster.name = "yang2019_clusters")

# Dimension reductions with Uniform Manifold Approximation and Projection (UMAP)
fetahu.NK <- RunUMAP(fetahu.NK, 
                     reduction = "pca_new", 
                     dims = 1:10,
                     n.epochs = 500,
                     min.dist = 0.5,
                     seed.use = 28,
                     reduction.name = "umap_new")

# plot UMAP
DimPlot(fetahu.NK,
        group.by = "yang2019_clusters",
        reduction = "umap_new")

# Get number of cells per cluster and per group
table(fetahu.NK$yang2019_clusters, 
      fetahu.NK$group)

# set Seurat clusters as ident
Idents(fetahu.NK) <- fetahu.NK@meta.data$yang2019_clusters

# Heatmap of gene expression of yang2019_clust_mark genes
DoHeatmap(fetahu.NK, 
          features = yang2019_clust_filt$gene)

#saveRDS(fetahu.NK, file = paste0(params$out,"NK_project/Rds/fetahu2023_NK_reClust_yang2019.Rds"))

```

```{r Clusters per sample}
DimPlot(fetahu.NK,
        reduction = "umap_new",
        split.by = "group", 
        group.by = "yang2019_clusters",
        pt.size = 1)

fetahu.NK <- readRDS(file = paste0(params$out,
                                   "NK_project/Rds/fetahu2023_NK_reClust_yang2019.Rds"))

Idents(fetahu.NK) <- fetahu.NK@meta.data$patients

percent_NK_group <- table(fetahu.NK$yang2019_clusters,fetahu.NK$group)
percent_NK_group

percent_NK_sample <- table(fetahu.NK$yang2019_clusters, fetahu.NK$patients)
percent_NK_sample


# define colors
colors <- brewer.pal(6,"Pastel2") 

barplot(percent_NK_group,
        main = "Fetahu et al., 2023: reclustering with Yang2019 genes",
        col = colors, 
        border = "white",
        xlim=c(0, ncol(percent_NK_group) + 2),
        ylab = "Number of cells in the cluster",
        xlab = "Sample group",
        legend.text = c("0","1","2","3","4","5"),
        args.legend = list(x = ncol(percent_NK_group) + 2,
                           y = max(colSums(percent_NK_group)),
                           bty = "n"))

# Transform this data in %
data_percentage_group <- apply(percent_NK_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

data_percentage_sample <- apply(percent_NK_sample, 2, 
                         function(x){x*100/sum(x,na.rm=T)})
 

# Make a stacked barplot in %
barplot(data_percentage_group, 
        main = "Fetahu et al., 2023: reclustering with Yang2019 genes",
        col = colors, 
        border = "white",
        xlim=c(0, ncol(data_percentage_group) + 2),
        ylab = "Percentage of cells in the cluster",
        xlab = "Sample group",
        legend.text = c("0","1","2","3","4","5"),
        args.legend = list(x = ncol(data_percentage_group) + 2,
                           y = max(colSums(data_percentage_group)),
                           bty = "n"))
```

```{r fig.height=5, fig.width=15}
barplot(data_percentage_sample, 
        main = "Fetahu et al., 2023: reclustering with Yang2019 genes",
        col = colors, 
        border = "white",
        xlim = c(0, ncol(data_percentage_sample) + 4),
        ylab = "Percentage of cells in the cluster",
        xlab = "Sample groups",
        legend.text = c("0","1","2","3","4","5"),
        args.legend = list(x = ncol(data_percentage_sample) + 5,
                           y = max(colSums(data_percentage_sample)),
                           bty = "n"))
```


# ___NK sub-type enrichment___

```{r load gene lists}
yang2019_clust1 <- read.table("/home/rstudio/mnt_resources/NK_cells_gene_sets/yang2019_cluster1.csv")
nrow(yang2019_clust1)
yang2019_clust2 <- read.table("/home/rstudio/mnt_resources/NK_cells_gene_sets/yang2019_cluster2.csv")
nrow(yang2019_clust2)
yang2019_clust3 <- read.table("/home/rstudio/mnt_resources/NK_cells_gene_sets/yang2019_cluster3.csv")
nrow(yang2019_clust3)
yang2019_clust4 <- read.table("/home/rstudio/mnt_resources/NK_cells_gene_sets/yang2019_cluster4.csv")
nrow(yang2019_clust4)
yang2019_clust5 <- read.table("/home/rstudio/mnt_resources/NK_cells_gene_sets/yang2019_cluster5.csv")
nrow(yang2019_clust5)
yang2019_clust6 <- read.table("/home/rstudio/mnt_resources/NK_cells_gene_sets/yang2019_cluster6.csv")
nrow(yang2019_clust6)
yang2019_clust7 <- read.table("/home/rstudio/mnt_resources/NK_cells_gene_sets/yang2019_cluster7.csv")
nrow(yang2019_clust7)
```

```{r}
# extract counts
counts <- LayerData(fetahu.NK, assay="RNA", layer='data')
var_genes <- dimnames(counts)[[1]]

# remove genes not present in dataset
yang2019_clust1 <- intersect(yang2019_clust1$V1, var_genes)
length(yang2019_clust1)
yang2019_clust2 <- intersect(yang2019_clust2$V1, var_genes)
length(yang2019_clust2)
yang2019_clust3 <- intersect(yang2019_clust3$V1, var_genes)
length(yang2019_clust3)
yang2019_clust4 <- intersect(yang2019_clust4$V1, var_genes)
length(yang2019_clust4)
yang2019_clust5 <- intersect(yang2019_clust5$V1, var_genes)
length(yang2019_clust5)
yang2019_clust6 <- intersect(yang2019_clust6$V1, var_genes)
length(yang2019_clust6)
yang2019_clust7 <- intersect(yang2019_clust7$V1, var_genes)
length(yang2019_clust7)

yang2019_genes <- list(cluster1 = yang2019_clust1, 
                       cluster2 = yang2019_clust2,
                       cluster3 = yang2019_clust3,
                       cluster4 = yang2019_clust4,
                       cluster5 = yang2019_clust5,
                       cluster6 = yang2019_clust6,
                       cluster7 = yang2019_clust7)

#pdf(file = paste0(params$out,"NK_project/figures/yang2019_gene_list_overlap.pdf"),
#   height = 5, width = 10)
ggVennDiagram(yang2019_genes, set_size = 6, label_alpha = 0,label_size = 2,
  category.names = c("Cluster 1","Cluster 2","Cluster 3","Cluster 4",
                     "Cluster 5","Cluster 6","Cluster 7"),
  edge_size = 1, label_percent_digit = 2,label_color = "white") +
  ggplot2::scale_fill_gradient(low="#B59410",high = "darkblue") +
  labs(title = "\t \t \t Overlap of marker genes") + 
  theme(title = element_text(size = 20),legend.text = element_text(size = 10)) +
  scale_x_continuous(expand = expansion(mult = .2))
#dev.off()
```

```{r generate cluster module scores}
fetahu.NK <- AddModuleScore(fetahu.NK,
                         features = yang2019_clust1,
                         name = "Yang2019_cluster1_",
                         seed = 2805,
                         slot = "data")

fetahu.NK <- AddModuleScore(fetahu.NK,
                         features = yang2019_clust2,
                         name = "Yang2019_cluster2_",
                         seed = 2805)

fetahu.NK <- AddModuleScore(fetahu.NK,
                         features = yang2019_clust3,
                         name = "Yang2019_cluster3_",
                         seed = 2805)

fetahu.NK <- AddModuleScore(fetahu.NK,
                         features = yang2019_clust4,
                         name = "Yang2019_cluster4_",
                         seed = 2805)

fetahu.NK <- AddModuleScore(fetahu.NK,
                         features = yang2019_clust5,
                         name = "Yang2019_cluster5_",
                         seed = 2805)

fetahu.NK <- AddModuleScore(fetahu.NK,
                         features = yang2019_clust6,
                         name = "Yang2019_cluster6_",
                         seed = 2805)

fetahu.NK <- AddModuleScore(fetahu.NK,
                         features = yang2019_clust7,
                         name = "Yang2019_cluster7_",
                         seed = 2805)
```

```{r plot module scores, fig.height=6, fig.width=6}
#pdf(file = paste0(params$out,"NK_project/figures/fetahu2023_reClust_UMAP.pdf"), 
#    height = 6, width = 6)
do_DimPlot(sample = fetahu.NK,
           plot.title = "Fetahu et al., 2023",
           plot.subtitle = "Re-clustering using Yang et al., 2019 gene lists",
           reduction = "umap_new",
           colors.use = COLOR_CODE_CLUST_v3,
           group.by = "yang2019_clusters",
           shuffle = TRUE,
           repel = TRUE,
           font.size = 16,
           plot.axes = TRUE,
           pt.size = 1,
           legend.title = "Clusters",
           legend.position = "right")
#dev.off()

#pdf(file = paste0(params$out,"NK_project/figures/yang2019_moduleScore_cluster1.pdf"), 
#    height = 6, width = 6)
do_FeaturePlot(sample = fetahu.NK, 
               features = "Yang2019_cluster1_1",
               plot.subtitle = "Yang et al., 2019: Cluster 1 gene module score",
               plot.title = "Fetahu et al., 2023",
               order = TRUE,
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
#dev.off()

#pdf(file = paste0(params$out,"NK_project/figures/yang2019_moduleScore_cluster2.pdf"), 
#    height = 6, width = 6)
do_FeaturePlot(sample = fetahu.NK, 
               features = "Yang2019_cluster2_1",
               plot.subtitle = "Yang et al., 2019: Cluster 2 gene module score",
               plot.title = "Fetahu et al., 2023",
               order = TRUE,
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
#dev.off()

#pdf(file = paste0(params$out,"NK_project/figures/yang2019_moduleScore_cluster3.pdf"), 
#    height = 6, width = 6)
do_FeaturePlot(sample = fetahu.NK, 
               features = "Yang2019_cluster3_1",
               plot.subtitle = "Yang et al., 2019: Cluster 3 gene module score",
               plot.title = "Fetahu et al., 2023",
               order = TRUE,
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
#dev.off()

#pdf(file = paste0(params$out,"NK_project/figures/yang2019_moduleScore_cluster4.pdf"), 
#    height = 6, width = 6)
do_FeaturePlot(sample = fetahu.NK, 
               features = "Yang2019_cluster4_1",
               plot.subtitle = "Yang et al., 2019: Cluster 4 gene module score",
               plot.title = "Fetahu et al., 2023",
               order = TRUE,
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
#dev.off()

#pdf(file = paste0(params$out,"NK_project/figures/yang2019_moduleScore_cluster5.pdf"), 
#    height = 6, width = 6)
do_FeaturePlot(sample = fetahu.NK, 
               features = "Yang2019_cluster5_1",
               plot.subtitle = "Yang et al., 2019: Cluster 5 gene module score",
               plot.title = "Fetahu et al., 2023",
               order = TRUE,
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
#dev.off()

#pdf(file = paste0(params$out,"NK_project/figures/yang2019_moduleScore_cluster6.pdf"), 
#    height = 6, width = 6)
do_FeaturePlot(sample = fetahu.NK, 
               features = "Yang2019_cluster6_1",
               plot.subtitle = "Yang et al., 2019: Cluster 6 gene module score",
               plot.title = "Fetahu et al., 2023",
               order = TRUE,
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
#dev.off()


#pdf(file = paste0(params$out,"NK_project/figures/yang2019_moduleScore_cluster7.pdf"), 
#    height = 6, width = 6)
do_FeaturePlot(sample = fetahu.NK, 
               features = "Yang2019_cluster7_1",
               plot.subtitle = "Yang et al., 2019: Cluster 7 gene module score",
               plot.title = "Fetahu et al., 2023",
               order = TRUE,
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
#dev.off()
```

```{r fig.height=6, fig.width=6}
FeaturePlot(object = fetahu.NK, 
            features = "Yang2019_cluster1_1",
            reduction = "umap_new",
            cols = c("darkgray","blue","red"),
            pt.size = 1,
            order = FALSE)
FeaturePlot(object = fetahu.NK, 
            features = "Yang2019_cluster2_1",
            reduction = "umap_new",
            cols = c("darkgray","blue","red"),
            pt.size = 1,
            order = FALSE)
FeaturePlot(object = fetahu.NK, 
            features = "Yang2019_cluster3_1",
            reduction = "umap_new",
            cols = c("darkgray","blue","red"),
            pt.size = 1,
            order = FALSE)
FeaturePlot(object = fetahu.NK, 
            features = "Yang2019_cluster4_1",
            reduction = "umap_new",
            cols = c("darkgray","blue","red"),
            pt.size = 1,
            order = FALSE)
FeaturePlot(object = fetahu.NK, 
            features = "Yang2019_cluster5_1",
            reduction = "umap_new",
            cols = c("darkgray","blue","red"),
            pt.size = 1,
            order = FALSE)
FeaturePlot(object = fetahu.NK, 
            features = "Yang2019_cluster6_1",
            reduction = "umap_new",
            cols = c("darkgray","blue","red"),
            pt.size = 1,
            order = FALSE)
FeaturePlot(object = fetahu.NK, 
            features = "Yang2019_cluster7_1",
            reduction = "umap_new",
            cols = c("darkgray","blue","red"),
            pt.size = 1,
            order = FALSE)
```


```{r}
VlnPlot(fetahu.NK,features = "Yang2019_cluster1_1")
RidgePlot(fetahu.NK,features = "Yang2019_cluster1_1")
VlnPlot(fetahu.NK,features = "Yang2019_cluster2_1")
RidgePlot(fetahu.NK,features = "Yang2019_cluster2_1")
VlnPlot(fetahu.NK,features = "Yang2019_cluster3_1")
RidgePlot(fetahu.NK,features = "Yang2019_cluster3_1")
VlnPlot(fetahu.NK,features = "Yang2019_cluster4_1")
RidgePlot(fetahu.NK,features = "Yang2019_cluster4_1")
VlnPlot(fetahu.NK,features = "Yang2019_cluster5_1")
RidgePlot(fetahu.NK,features = "Yang2019_cluster5_1")
VlnPlot(fetahu.NK,features = "Yang2019_cluster6_1")
RidgePlot(fetahu.NK,features = "Yang2019_cluster6_1")
VlnPlot(fetahu.NK,features = "Yang2019_cluster7_1")
RidgePlot(fetahu.NK,features = "Yang2019_cluster7_1")
```

Intepretation: 
NB cluster 4 = Yang cluster 1 (CD56 bright NK)
*NB cluster 0 = Yang cluster 2 (Transitional NK)*
NB cluster 5 = Yang cluster 3 (Active NK)
NB cluster 1 = Yang cluster 4 (adaptive NK)
NB cluster 2 = Yang cluster 5 (Mature NK)
NA = Yang clusters 6 and 7 (Terminal and Inflamed NK, respectively)

# ___DA testing___

```{r}
fetahu.NK@meta.data$patients<- factor(fetahu.NK@meta.data$patients)
fetahu.NK@meta.data$yang2019_clusters<- factor(fetahu.NK@meta.data$yang2019_clusters)
```

```{r miloR - many groups}
# MiloR doesn't use the "original" clusters but rather creates neighborhoods
# based on similar logFC (note the fold change here is the number of cells
# per cell type)

#2. If comparing more than two sample / groups, use DA testing with contrasts

# For differential abundance analysis on graph neighborhoods we first construct a Milo object. 
# This extends the SingleCellExperiment class to store information about neighborhoods on the KNN graph.

rna_milo <- Milo(as.SingleCellExperiment(fetahu.NK))
rna_milo

reducedDim(rna_milo, "UMAP") <- reducedDim(as.SingleCellExperiment(fetahu.NK), 
                                           "UMAP_NEW")
plotUMAP(rna_milo, 
         colour_by="group") 
plotUMAP(rna_milo, 
        colour_by="patients")

# Define cell neighbourhoods
rna_milo <- buildGraph(rna_milo, 
                       k = 10, 
                       d = 30)

## Constructing kNN graph with k = 10
rna_milo <- makeNhoods(rna_milo, 
                       prop = 0.1, 
                       k = 10, 
                       d = 30, 
                       refined = TRUE, 
                       refinement_scheme = "graph") # make nhoods using graph-only as this is faster

# metadata
colData(rna_milo)$patients <- paste(colData(rna_milo)$group, colData(rna_milo)$patients, sep="_")

# Counting cells in neighbourhoods
rna_milo <- countCells(rna_milo, meta.data = data.frame(colData(rna_milo)), 
                       samples="patients") # make the nhood X sample counts matrix


# Defining experimental design
rna_design <- data.frame(colData(rna_milo))[,c("patients", "group")]
rna_design <- distinct(rna_design)
rownames(rna_design) <- rna_design$patients
rna_design

# Computing neighborhood connectivity
rna_milo <- calcNhoodDistance(rna_milo, 
                              d = 30, 
                              reduced.dim = "PCA")

# DA testing
da_results <- testNhoods(rna_milo, 
                         design = ~ 0 + group,
                         design.df = rna_design, 
                         reduced.dim="PCA")
head(da_results)

da_results %>%
  arrange(SpatialFDR) %>%
  head() 

rna_milo <- buildNhoodGraph(rna_milo)


# QUALITY CHECK OF THE DIFFERENTIAL ABUNDANCE TESTING

# Plot single-cell UMAP
umap_pl <- plotReducedDim(rna_milo, 
                          dimred = "UMAP", 
                          colour_by = "group", 
                          text_by = "yang2019_clusters", 
                          text_size = 3, point_size = 0.5,) +
  guides(fill = "none")

rna_milo <- buildNhoodGraph(rna_milo)

# Plot neighbourhood graph
nh_graph_pl <- plotNhoodGraphDA(rna_milo, 
                                da_results, 
                                layout="UMAP", 
                                alpha=0.05) 

umap_pl + nh_graph_pl +
  plot_layout(guides = "collect")

# Differential abundance testing with contrasts
rna_design <- data.frame(colData(rna_milo))[,c("patients", "group")]
rna_design <- distinct(rna_design)
rownames(rna_design) <- rna_design$patients

# Reorder rownames to match columns of nhoodCounts(milo)
rna_design <- rna_design[colnames(nhoodCounts(rna_milo)), , drop=FALSE]
table(rna_design$group)

# Comparisons
#A. ATRXmut vs.(MYCNA and sporadic)
model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupIII - (groupII + groupIV) / 2")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXmutvsAll <- testNhoods(rna_milo, design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXmutvsAll$SpatialFDR < 0.1)

# annotate hood with original cell type annotation
da_results_ATRXmutvsAll <- annotateNhoods(rna_milo, da_results_ATRXmutvsAll, 
                             coldata_col = "yang2019_clusters")

# remove hoods that contained mixed cell type populations
da_results_ATRXmutvsAll$yang2019_clusters <- 
  ifelse(da_results_ATRXmutvsAll$yang2019_clusters_fraction < 0.7, 
         "Mixed", da_results_ATRXmutvsAll$yang2019_clusters)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXmutvsAll %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

plotDAbeeswarm(da_results_ATRXmutvsAll, group.by = "yang2019_clusters")

#B. ATRXmut vs.(MYCNA and sporadic and control)

model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupIII - (groupII + groupIV + groupI) / 3")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXmutvsAll <- testNhoods(rna_milo, design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXmutvsAll$SpatialFDR < 0.1)

# annotate hood with original cell type annotation
da_results_ATRXmutvsAll <- annotateNhoods(rna_milo, da_results_ATRXmutvsAll, 
                             coldata_col = "yang2019_clusters")

# remove hoods that contained mixed cell type populations
da_results_ATRXmutvsAll$yang2019_clusters <- 
  ifelse(da_results_ATRXmutvsAll$yang2019_clusters_fraction < 0.7, 
         "Mixed", da_results_ATRXmutvsAll$yang2019_clusters)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXmutvsAll %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

plotDAbeeswarm(da_results_ATRXmutvsAll, group.by = "yang2019_clusters")
```


# ___SessionInfo___

```{r}
sessionInfo()
```

