---
title: "NK project: DEG analysis"
author: "Dr. Sara Wernig-Zorc"
date: "22.07.2024 - 26.07.2024"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/scRNA.NB.integration/metadata/"
---

PROJECT DESCRIPTION

Identify NK sub-populations in tumor-infiltrated bone marrow samples from
Neuroblastoma patients.

Here i am looking at the differences between adult and infant BM-derived
NK cells and between NB genetic subtypes


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)

source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_utils.R")
```

# ___Load data: Healthy BM-derived NK cells___

```{r}
crinier2021 <- readRDS(file = paste0(params$input_BM1))
```

```{r eval=FALSE, include=FALSE}
lee2023 <- readRDS(file = paste0(params$input_BM3))

DefaultAssay(lee2023) <- "RNA"
lee2023@assays$RNA@data <- lee2023@assays$RNA@counts

col_clusters <- read.csv2("/home/rstudio/mnt_data/BM/Lee/celltype_metacluster.csv", 
                          sep = ";")
col_clusters <- as.data.frame(col_clusters)

Idents(lee2023) <- lee2023$age
DefaultAssay(lee2023) <- "RNA"
hBm_age2 <- WhichCells(lee2023, idents = 2)

DimPlot(lee2023, 
        group.by = "age",
        cells.highlight = hBm_age2)

age2 <- "2"
hBM_age2 <- subset(lee2023, idents = age2)
hBM_age2

DimPlot(hBM_age2,group.by = "ct")

Idents(hBM_age2) <- lee2023$ct
cell_types_of_interest <- c("CD16+ NK",
                            "Tissue-resident NK cell",
                            "CD56+ NK",
                            "T/NK cell (cycling)")

hBM_age2_NK <- subset(hBM_age2, idents = cell_types_of_interest)
hBM_age2_NK # 771 cells
DimPlot(hBM_age2_NK,group.by = "ct")
```

# ___Load data: Diseased BM-MNCs___

```{r}
fetahu2023 <- readRDS(file = paste0(params$input_scRNA))
metadata <- readRDS("/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/metadata.rds")

fetahu2023 <- SeuratObject::as.Seurat(fetahu2023, data=NULL)
fetahu2023 <- RenameAssays(object = fetahu2023, originalexp = 'RNA')

# set cell type annotation to the original annotation
fetahu2023@active.ident <- setNames(pull(metadata[match(metadata$cell,colnames(fetahu2023)),"cellont_cluster"], "cellont_cluster"),pull(metadata[match(metadata$cell,colnames(fetahu2023)), "cell"], "cell"))

fetahu2023@meta.data$orig.ident <- fetahu2023@active.ident

# set assay to RNA 
DefaultAssay(fetahu2023) <- "RNA"
```

```{r}
# add sex and age information
# Merge the metadata with the Seurat object's metadata based on the sample_ID and sample columns
# Ensure column names match those in your data
full_metadata <- read_excel(paste0(params$metadata,
  "2024-07-29_Wernig-Zorc_Fetahu_complete_single_cell_cohort_EB_FINAL.xlsx"), 
            col_types = c("text", "numeric", "text", "text", "text", "text", 
                          "text", "date", "date", "text", "text", "text", 
                          "text", "text", "numeric", "text", "text", "text", 
                          "text", "text", "text", "text", "text", "text", 
                          "text", "text", "text", "text", "text", "text", 
                          "text", "text", "text", "text", "text", "text"))

metadata <- fetahu2023@meta.data
metadata$cell_ID <- row.names(metadata)

names(full_metadata)[names(full_metadata) == 'BM_ID'] <- 'sample'

# Perform the merge using left_join from dplyr, keeping the row order of df1
merged_metadata <- metadata %>%
  left_join(full_metadata, by = "sample") %>% 
  distinct(cell_ID, .keep_all = TRUE)

row.names(merged_metadata) <- merged_metadata$cell_ID 

# Update the Seurat object's metadata slot
fetahu2023@meta.data <- merged_metadata
fetahu2023@meta.data

saveRDS(fetahu2023, file = paste0(params$out,"/Rds/fetahu2023_withMetadata.Rds"))
```


## ___Subset Fetahu2023 to NK cells only___

```{r eval=FALSE, include=FALSE}
levels(Idents(fetahu2023))

# subset metadata
filter <- metadata$cellont_abbr == "NK"
NK.metadata <- metadata[filter,]

# subset scRNA-seq data 
# NK cells only

NK_cells <- "NK (4)" # change cluster number
fetahu2023_NK <- subset(fetahu2023, idents = NK_cells, invert = FALSE)

NK.metadata <- fetahu2023_NK@meta.data
NK.metadata$cell_ID <- row.names(NK.metadata)
predicted_labels$cell_ID <- row.names(predicted_labels)

merged_metadata <- NK.metadata %>%
  left_join(predicted_labels, by = "cell_ID", copy = TRUE) %>% 
  distinct(cell_ID, .keep_all = TRUE)

merged_metadata <- merged_metadata %>%
  rename(predicted_NK_type_de.n350 = labels)
  
row.names(merged_metadata) <- merged_metadata$cell_ID 
  
fetahu2023_NK@meta.data <- merged_metadata

fetahu2023_NK <- ScaleData(fetahu2023_NK)

#saveRDS(fetahu2023_NK, file = paste0(params$out,"/Rds/fetahu2023_NK_cluster.Rds"))
```

------------------------------ START HERE -------------------------------------

## ___Load data: NB BM-derived NK cells___

```{r}
predicted_labels <- readRDS(file = paste0(params$out,
                                          "/Rds/fetahu2023_predicted_NKclusters_from_Crinier2021_de.n=350.Rds"))
fetahu2023_NK <- readRDS(file = paste0(params$out,
                                       "/Rds/fetahu2023_NK_subclusters_from_Crinier2021_de.n=350.Rds"))
```


# ___NK types in tumor-infiltrated and non-infiltrated BM, vs. adult BM___

```{r}
percent_NK_group <- table(fetahu2023_NK$predicted_NK_type_de.n350_de.n350,
                          fetahu2023_NK$group)

percent_NK_group

# define colors
colors <- brewer.pal(4, "Pastel2") 

barplot(percent_NK_group,
        col = colors, 
        border = "white", 
        xlab = "NB genetic group")
```


# ___CD6 expression in NK sub-clusters___

```{r fig.height=5, fig.width=10}
fetahu2023_NK <- fetahu2023_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)

FeaturePlot(fetahu2023_NK,features = "CD6",
            split.by = "group")

VlnPlot(fetahu2023_NK,
         group.by = "predicted_NK_type_de.n350",
         split.by = "group",
         features = "CD6",
         log = TRUE,
         layer = "data")

VlnPlot(fetahu2023_NK,
         group.by = "predicted_NK_type_de.n350",
         features = "CD6",
         log = TRUE,
         layer = "data",
        pt.size = 0,cols = COLOR_CODE_v2)


VlnPlot(fetahu2023_NK,
         group.by = "group",
         features = "CD6",
         log = TRUE,
         layer = "data",
         pt.size = 0)

do_FeaturePlot(sample = fetahu2023_NK, 
               features = "CD6",
               plot.subtitle = "CD6 gene expression",
               plot.title = "Predicted NK sub-types in NB-BM samples",
               order = TRUE,
               reduction = "UMAP",
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               min.cutoff = 0,
               max.cutoff = 3,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
```

# ___NK type gene signatures in NB BM-NK cells___

```{r}
# extract counts
counts <- LayerData(fetahu2023_NK, assay="RNA", layer='data')
var_genes <- dimnames(counts)[[1]]

# remove genes not present in dataset
adaptive_NK_genes <- intersect(adaptive_NK_genes, var_genes)
conventional_NK_genes <- intersect(conventional_NK_genes, var_genes)
mature_NK_genes <- intersect(mature_NK_genes, var_genes)
NKP_genes <- intersect(NKP_genes, var_genes)

# log scale counts
fetahu2023_NK <- fetahu2023_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

## ___Calculate module scores___

```{r}
fetahu2023_NK <- AddModuleScore(object = fetahu2023_NK,
                                features = adaptive_NK_genes,
                                ctrl = 5,
                                name = "adaptive_NK_module",
                                seed = 2805)

fetahu2023_NK <- AddModuleScore(object = fetahu2023_NK,
                                features = conventional_NK_genes,
                                ctrl = 5,
                                name = "conventional_NK_module",
                                seed = 2805)

fetahu2023_NK <- AddModuleScore(object = fetahu2023_NK,
                                features = NKP_genes,
                                ctrl = 5,
                                name = "NKP_module",
                                seed = 2805)

fetahu2023_NK <- AddModuleScore(object = fetahu2023_NK,
                                features = mature_NK_genes,
                                ctrl = 5,
                                name = "mature_NK_module",
                                seed = 2805)
```

## ___Plot module scores___

```{r fig.height=12, fig.width=12}
do_FeaturePlot(sample = fetahu2023_NK, 
               features = "adaptive_NK_module1",
               plot.title = "Adaptive NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14, 
               max.cutoff = 2,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Gene expression",
               legend.length = 10,
               split.by = "group")

do_FeaturePlot(sample = fetahu2023_NK, 
               features = "adaptive_NK_module1",
               plot.title = "Adaptive NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               max.cutoff = 2,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Gene expression",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = fetahu2023_NK, 
               features = "conventional_NK_module1",
               plot.title = "Conventional NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Gene expression",
               legend.length = 10,
               split.by = "group")

do_FeaturePlot(sample = fetahu2023_NK, 
               features = "conventional_NK_module1",
               plot.title = "Conventional NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Gene expression",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = fetahu2023_NK, 
               features = "NKP_module1",
               plot.title = "progenitor NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Gene expression",
               legend.length = 10,
               split.by = "group")

do_FeaturePlot(sample = fetahu2023_NK, 
               features = "NKP_module1",
               plot.title = "progenitor NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Gene expression",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = fetahu2023_NK, 
               features = "mature_NK_module1",
               plot.title = "Mature NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Gene expression",
               legend.length = 10,
               split.by = "group")

do_FeaturePlot(sample = fetahu2023_NK, 
               features = "mature_NK_module1",
               plot.title = "Mature NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Gene expression",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")
```


```{r fig.height=7, fig.width=15}
VlnPlot(fetahu2023_NK,
        features = "adaptive_NK_module1", 
        group.by = "group", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_GROUP)

VlnPlot(fetahu2023_NK,
        features = "conventional_NK_module1", 
        group.by = "group", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_GROUP)

VlnPlot(fetahu2023_NK,
        features = "NKP_module1", 
        group.by = "group", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_GROUP)

VlnPlot(fetahu2023_NK,
        features = "mature_NK_module1", 
        group.by = "group", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_GROUP)

do_ViolinPlot(fetahu2023_NK,
              features = "mature_NK_module1", 
              group.by = "group", 
              colors.use = COLOR_CODE,
              font.size = 18,
              legend.title = "",
              legend.position = "none")

##########################

do_ViolinPlot(fetahu2023_NK,
              features = "adaptive_NK_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(fetahu2023_NK,
              features = "conventional_NK_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(fetahu2023_NK,
              features = "NKP_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(fetahu2023_NK,
              features = "mature_NK_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)


```


## ___healthy BM-NK data___

### ___toddlers___


```{r} 
# extract counts
counts <- LayerData(fetahu2023_NK, assay="RNA", layer='data')
var_genes <- dimnames(counts)[[1]]

# remove genes not present in dataset
adaptive_NK_genes <- intersect(adaptive_NK_genes, var_genes)
conventional_NK_genes <- intersect(conventional_NK_genes, var_genes)
mature_NK_genes <- intersect(mature_NK_genes, var_genes)
NKP_genes <- intersect(NKP_genes, var_genes)
```


```{r}
fetahu2023_NK <- AddModuleScore(object = fetahu2023_NK,
                                features = adaptive_NK_genes,
                                ctrl = 5,
                                name = "adaptive_NK_module",
                                seed = 2805)

fetahu2023_NK <- AddModuleScore(object = fetahu2023_NK,
                                features = conventional_NK_genes,
                                ctrl = 5,
                                name = "conventional_NK_module",
                                seed = 2805)

fetahu2023_NK <- AddModuleScore(object = fetahu2023_NK,
                                features = NKP_genes,
                                ctrl = 5,
                                name = "NKP_module",
                                seed = 2805)

fetahu2023_NK <- AddModuleScore(object = fetahu2023_NK,
                                features = mature_NK_genes,
                                ctrl = 5,
                                name = "mature_NK_module",
                                seed = 2805)
```


```{r}
do_ViolinPlot(fetahu2023_NK,
              features = "adaptive_NK_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(fetahu2023_NK,
              features = "conventional_NK_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(fetahu2023_NK,
              features = "NKP_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(fetahu2023_NK,
              features = "mature_NK_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)
```


### ___adults___

```{r}
# extract counts
counts <- LayerData(crinier2021, assay="RNA", layer='data')
var_genes <- dimnames(counts)[[1]]

# remove genes not present in dataset
adaptive_NK_genes <- intersect(adaptive_NK_genes, var_genes)
conventional_NK_genes <- intersect(conventional_NK_genes, var_genes)
mature_NK_genes <- intersect(mature_NK_genes, var_genes)
NKP_genes <- intersect(NKP_genes, var_genes)

# log scale counts
crinier2021 <- crinier2021 %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```


```{r}
crinier2021 <- AddModuleScore(object = crinier2021,
                                features = adaptive_NK_genes,
                                ctrl = 5,
                                name = "adaptive_NK_module",
                                seed = 2805)

crinier2021 <- AddModuleScore(object = crinier2021,
                                features = conventional_NK_genes,
                                ctrl = 5,
                                name = "conventional_NK_module",
                                seed = 2805)

crinier2021 <- AddModuleScore(object = crinier2021,
                                features = NKP_genes,
                                ctrl = 5,
                                name = "NKP_module",
                                seed = 2805)

crinier2021 <- AddModuleScore(object = crinier2021,
                                features = mature_NK_genes,
                                ctrl = 5,
                                name = "mature_NK_module",
                                seed = 2805)
```


```{r}
do_ViolinPlot(crinier2021,
              features = "adaptive_NK_module1", 
              group.by = "cluster2", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(crinier2021,
              features = "conventional_NK_module1", 
              group.by = "cluster2", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(crinier2021,
              features = "NKP_module1", 
              group.by = "cluster2", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(crinier2021,
              features = "mature_NK_module1", 
              group.by = "cluster2", 
              colors.use = COLOR_CODE_v2)
```



# ___DEG analysis: ATRXm cluster2 vs. other cluster2___

single cell RNA-seq DEG analysis

```{r}
# DE analysis for the whole NK cluster
DEG_NK <- DElegate::findDE(fetahu2023_NK,
                           group_column = "group",
                           replicate_column = "sample",
                           order_results = TRUE)

# results specific to ATRXm vs. rest
filter <- DEG_NK$group1 == "III" 
DEG_ATRXm <- DEG_NK[filter,]
head(DEG_ATRXm, 10)

# Change idents to cluster annotations
Idents(fetahu2023_NK) <- fetahu2023_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 2
cluster_of_interest <- "hNK_Bm2"

# subset scRNA-seq data to NK and NB cells 
fetahu2023_hNK_Bm2 <- subset(fetahu2023_NK, 
                             idents = cluster_of_interest, 
                             invert = FALSE)
fetahu2023_hNK_Bm2

# DE analysis for the hNK_Bm2 NK cluster
DEG_hNK_Bm2 <- DElegate::findDE(fetahu2023_hNK_Bm2,
                           group_column = "group",
                           order_results = TRUE)

# results specific to ATRXm vs. rest
filter <- DEG_hNK_Bm2$group1 == "III" 
DEG_hNK_Bm2_ATRXm <- DEG_hNK_Bm2[filter,]
head(DEG_hNK_Bm2_ATRXm,10)

DEG_hNK_Bm2_ATRXm <- subset(DEG_hNK_Bm2_ATRXm, DEG_hNK_Bm2_ATRXm$padj < 0.01)
head(DEG_hNK_Bm2_ATRXm, 10)

DEG_hNK_Bm2_ATRXm_top <- subset(DEG_hNK_Bm2_ATRXm, DEG_hNK_Bm2_ATRXm$log_fc > 1)
nrow(DEG_hNK_Bm2_ATRXm_top)
head(DEG_hNK_Bm2_ATRXm_top, 10)

write.table(DEG_hNK_Bm2_ATRXm_top, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_cluster_ATRXm_top.csv"))

# DE analysis for the hNK_Bm2 NK cluster
DEG_hNK_Bm2_ref <- DElegate::findDE(fetahu2023_hNK_Bm2,
                           group_column = "group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           compare = "III")

DEG_hNK_Bm2_ref <- subset(DEG_hNK_Bm2_ref, DEG_hNK_Bm2_ref$padj < 0.01)
head(DEG_hNK_Bm2_ref, 10)

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_ref, DEG_hNK_Bm2_ref$group1 == "I")
MYCNA_vs_ATRXm <- subset(DEG_hNK_Bm2_ref, DEG_hNK_Bm2_ref$group1 == "II")
sporadic_vs_ATRXm <- subset(DEG_hNK_Bm2_ref, DEG_hNK_Bm2_ref$group1 == "IV")


common_genes <- merge(sporadic_vs_ATRXm, MYCNA_vs_ATRXm,  by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, common_genes,  by = "feature")
head(common_genes,10)

head(common_genes_all,10)
```

### ___Vulcano plot___

```{r fig.height=7, fig.width=7}
common_genes_all
rownames(common_genes_all) <- common_genes_all$feature

library(EnhancedVolcano)

EnhancedVolcano(common_genes_all, 
                lab = rownames(common_genes_all),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXm vs. rest",
                pCutoff = 0.01,
                FCcutoff = 1,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(common_genes_all, 
                lab = rownames(common_genes_all),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs",
                subtitle = "ATRXm vs. rest",
                pCutoff = 0.01,
                FCcutoff = 1,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```


## ___Pathways enrichment analysis___

```{r}
# Function to map gene names to Entrez IDs
map_gene_to_entrez <- function(gene_names) {
  # Query the org.Hs.eg.db package for the Entrez IDs
  entrez_ids <- mapIds(org.Hs.eg.db,
                       keys = gene_names,
                       column = "ENTREZID",
                       keytype = "SYMBOL",
                       multiVals = "first")
  
  # Return the result as a named vector
  return(entrez_ids)
}
```

```{r pathway enrichment analysis, fig.height=10, fig.width=6}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- DEG_hNK_Bm2_ATRXm_top$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)
#print(entrez_ids)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable=TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

head(go_BP, 10)

barplot(go_BP, showCategory = 10)
dotplot(go_BP, showCategory = 20) + 
  ggtitle("NK cluster 2 in ATRXm: BP enrichment")

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable=TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

head(go_MF, 10)

barplot(go_MF, showCategory = 10)
dotplot(go_MF, showCategory = 20) + 
  ggtitle("NK cluster 2 in ATRXm: MF enrichment")

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable=TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

head(go_CC, 10)

barplot(go_CC, showCategory = 10)
dotplot(go_CC, showCategory = 20) + 
  ggtitle("NK cluster 2 in ATRXm: CC enrichment")


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.1)

head(kegg_enrichment, 10)

barplot(kegg_enrichment, showCategory = 10)
dotplot(kegg_enrichment, showCategory=20) + 
  ggtitle("NK cluster 2 in ATRXm: KEGG enrichment")
```


```{r eval=FALSE, include=FALSE}
# Wiki pathways
wiki.path <- enrichWP(gene = entrez_ids,
                      #universe = geneUniverse,
                      organism = "Homo sapiens",
                      pAdjustMethod = "fdr",
                      pvalueCutoff = 0.01
)

wiki.path <- DOSE::setReadable(wiki.path, 
                            org.Hs.eg.db, 
                            keyType = "ENTREZID")

head(wiki.path, 10)


barplot(wiki.path, showCategory = 10) + 
  ggtitle("NK cluster 2 in ATRXm: Wiki pathway enrichment")
dotplot(wiki.path, showCategory = 20) + 
  ggtitle("NK cluster 2 in ATRXm: Wiki pathway enrichment")
```


## ___Gene set enrichment analysis___

```{r eval=FALSE, include=FALSE}
# GSEAnalysis
genes <- DEG_hNK_Bm2_ATRXm$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)
print(entrez_ids)

ranked_genes <- data.frame("gene_name" = DEG_hNK_Bm2_ATRXm$feature,
                           "entrez_ids" = entrez_ids,
                            "log_fc" = DEG_hNK_Bm2_ATRXm$log_fc,
                            "pvalue" =  DEG_hNK_Bm2_ATRXm$pvalue,
                            "padj" = DEG_hNK_Bm2_ATRXm$padj)

ranked_genes$fcsign <- sign(DEG_hNK_Bm2_ATRXm$log_fc)
ranked_genes$logfdr <- -log10(DEG_hNK_Bm2_ATRXm$pvalue)
ranked_genes$sig <- ranked_genes$logfdr/ranked_genes$fcsign
ranked_genes <- na.omit(ranked_genes)
row.names(ranked_genes) <- ranked_genes$entrez_ids
head(ranked_genes, 10)

columns_to_remove <- c(1,3,4,5,6,7)
gene_list <- ranked_genes[,-columns_to_remove]

# Sort the gene list
gene_list <- gene_list %>% arrange(desc(sig),desc(entrez_ids))
gene_list

# Perform GSEA
gsea_result <- gseKEGG(geneList = list(gene_list$entrez_ids),
                       organism = "hsa", 
                       nPerm = 1000,
                       minGSSize = 10,
                       pvalueCutoff = 0.05,
                       verbose = FALSE)

# Print the GSEA results
print(gsea_result)

# Dot plot
dotplot(gsea_result, showCategory = 10)

# Enrichment map
enrichMap(gsea_result)
```

24.07.2024

# ___DEG analysis___

pseudo bulk RNA-seq DEG analysis

The aim here is to see if the low scores in the label transfer from adult BM
to toddler healthy and diseased BM, stem from the differences observed between
toddler and adult BM-NK cells or are these differences due to the tumor-infiltration
in the bone marrow and tumor-induced rewiring of the bone marrow immune cells.

inspo: https://hbctraining.github.io/scRNA-seq_online/lessons/pseudobulk_DESeq2_scrnaseq.html

## ___subset per group___

### ___subset control BM-NK cells___

```{r subset control BM-NK cells}
Idents(fetahu2023_NK) <- fetahu2023_NK$group
levels(Idents(fetahu2023_NK))

# subset scRNA-seq data 
# NK cells in control (toddler) BM only
CTRL_NK <- subset(fetahu2023_NK, 
                   idents = "I", 
                   invert = FALSE)

# sanity check
Idents(CTRL_NK) <- CTRL_NK$predicted_NK_type_de.n350
DimPlot(CTRL_NK, 
        group.by = "predicted_NK_type_de.n350", 
        cols = COLOR_CODE_v2, 
        raster = FALSE)

# number of cells per cluster
table(CTRL_NK$predicted_NK_type_de.n350)

#saveRDS(fetahu2023_NK, file = paste0(params$out,"Rds/fetahu2023_fetal_BM_NK.Rds"))
```

### ___subset ATRXm BM-NK cells___

```{r subset ATRXm BM-NK cells}
Idents(fetahu2023_NK) <- fetahu2023_NK$group
levels(Idents(fetahu2023_NK))

# subset scRNA-seq data 
# NK cells in control (toddler) BM only
ATRX_NK <- subset(fetahu2023_NK, 
                   idents = "III", 
                   invert = FALSE)

# sanity check
Idents(ATRX_NK) <- ATRX_NK$predicted_NK_type_de.n350
DimPlot(ATRX_NK, 
        group.by = "predicted_NK_type_de.n350", 
        cols = COLOR_CODE_v2, 
        raster = FALSE)

DimPlot(ATRX_NK, 
        group.by = "sample",
        cols = COLOR_CODE_PATIENTS_v6,
        raster = FALSE)

# number of cells per cluster
table(ATRX_NK$predicted_NK_type_de.n350)

saveRDS(ATRX_NK, file = paste0(params$out,"Rds/fetahu2023_ATRXmut_BM_NK.Rds"))
```

### ___subset MNA BM-NK cells___

```{r subset MNA BM-NK cells}
Idents(fetahu2023_NK) <- fetahu2023_NK$group
levels(Idents(fetahu2023_NK))

# subset scRNA-seq data 
# NK cells in control (toddler) BM only
MNA_NK <- subset(fetahu2023_NK, 
                   idents = "II", 
                   invert = FALSE)

# sanity check
Idents(MNA_NK) <- MNA_NK$predicted_NK_type_de.n350
DimPlot(MNA_NK, 
        group.by = "predicted_NK_type_de.n350", 
        cols = COLOR_CODE_v2, 
        raster = FALSE)

DimPlot(MNA_NK, 
        group.by = "sample",
        cols = COLOR_CODE_PATIENTS_v6,
        raster = FALSE)

# number of cells per cluster
table(MNA_NK$predicted_NK_type_de.n350)

saveRDS(MNA_NK, file = paste0(params$out,"Rds/fetahu2023_MNA_BM_NK.Rds"))
```

### ___subset sporadic BM-NK cells___

```{r subset sporadic BM-NK cells}
Idents(fetahu2023_NK) <- fetahu2023_NK$group
levels(Idents(fetahu2023_NK))

# subset scRNA-seq data 
# NK cells in control (toddler) BM only
sporadic_NK <- subset(fetahu2023_NK, 
                   idents = "IV", 
                   invert = FALSE)

# sanity check
Idents(sporadic_NK) <- sporadic_NK$predicted_NK_type_de.n350
DimPlot(sporadic_NK, 
        group.by = "predicted_NK_type_de.n350", 
        cols = COLOR_CODE_v2, 
        raster = FALSE)

DimPlot(sporadic_NK, 
        group.by = "sample",
        cols = COLOR_CODE_PATIENTS_v6,
        raster = FALSE)

# number of cells per cluster
table(sporadic_NK$predicted_NK_type_de.n350)

saveRDS(sporadic_NK, file = paste0(params$out,"Rds/fetahu2023_sporadic_BM_NK.Rds"))
```


## ___Pseudo bulk - aggregate counts___

```{r aggregate counts in ATRXmut BM, results='asis'}
ATRX_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXmut_BM_NK.Rds"))
ATRX_NK # 1823 cells in total

ATRX_NK@meta.data$orig.ident <- ATRX_NK@meta.data$sample 

# Create a named vector for the new values
new_names <- c("2005_1702" = "donor1",
               "2016_3924" = "donor2")

# Apply the re-naming
ATRX_NK@meta.data$orig.ident <- new_names[ATRX_NK@meta.data$orig.ident]

# convert to single cell experiment
ATRX_NK_sc <- as.SingleCellExperiment(ATRX_NK)

# explore sc object
#assays(ATRX_NK_sc)

# Dimensions of gene-by-cell count matrix
dim(counts(ATRX_NK_sc))

# Access dgCMatrix with gene counts
#head(counts(ATRX_NK_sc), 10)

# metadata
table(ATRX_NK_sc$sample)
table(ATRX_NK_sc$predicted_NK_type_de.n350)

head(colData(ATRX_NK_sc),10)

# aggregate counts
ATRX_NK_agg_counts <- aggregateBioVar(scExp = ATRX_NK_sc,
                                       subjectVar = "orig.ident", 
                                       cellVar = "predicted_NK_type_de.n350")

# Heatmap of log2 expression from aggregate gene-by-subject count matrix
ann_colors <- list(
    NK_cluster = "#419D78",
    Patients = COLOR_CODE_PATIENTS_v6)

scPHeatmap(
    sumExp = ATRX_NK_agg_counts$hNK_Bm2, 
    logSample = 1:100,
    subjectVar = "orig.ident",
    gtVar = "predicted_NK_type_de.n350",
    color = viridis::viridis(75), 
    annotation_colors = ann_colors,
    treeheight_row = 0, 
    treeheight_col = 0
)
```

```{r aggregate counts in toddler BM, results='asis'}
CTRL_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_fetal_BM_NK.Rds"))
#fetahu2023_NK$patients

CTRL_NK@meta.data$orig.ident <- CTRL_NK@meta.data$sample

# Create a named vector for the new values
new_names <- c("2014_0102" = "donor1", 
               "2020_1288" = "donor5", 
               "2018_4252" = "donor4", 
               "2016_1853" = "donor2",
               "2016_2950" = "donor3")

# Apply the renaming
CTRL_NK@meta.data$orig.ident <- new_names[CTRL_NK@meta.data$orig.ident]


# convert to single cell experiment
CTRL_NK_sc <- as.SingleCellExperiment(CTRL_NK)

# explore sc object
#assays(fetahu2023_NK_sc)

# Dimensions of gene-by-cell count matrix
dim(counts(CTRL_NK_sc))

# Access dgCMatrix with gene counts
#counts(fetahu2023_NK_sc)

# metadata
#table(CTRL_NK_sc$orig.ident)
#table(CTRL_NK_sc$predicted_NK_type_de.n350)

#head(colData(CTRL_NK_sc),10)

# aggregate counts
CTRL_NK_agg_counts <- aggregateBioVar(scExp = CTRL_NK_sc,
                                       subjectVar = "orig.ident", 
                                       cellVar = "predicted_NK_type_de.n350")

# Heatmap of log2 expression from aggregate gene-by-subject count matrix
ann_colors <- list(
    NK_cluster = "#419D78",
    Patients = COLOR_CODE_PATIENTS_v7)

scPHeatmap(
    sumExp = CTRL_NK_agg_counts$hNK_Bm2, 
    logSample = 1:100,
    subjectVar = "orig.ident",
    gtVar = "predicted_NK_type_de.n350",
    color = viridis::viridis(75), 
    annotation_colors = ann_colors,
    treeheight_row = 0, 
    treeheight_col = 0
)
```

```{r aggregate counts in adult BM, results='asis'}
crinier2021$orig.ident <- crinier2021$batch

# Create a named vector for the new values

new_names <- c(
  "Moelle 3" = "donor3",
  "Moelle 4" = "donor4",
  "Moelle 1" = "donor1",
  "Moelle 2" = "donor2",
  "Moelle 5" = "donor5",              
  "Moelle MoBxF10" = "donor6",
  "Moelle MoBx62151660270" = "donor7",
  "Moelle MoBxF20" = "donor8")

# Apply the renaming
crinier2021@meta.data$orig.ident <- new_names[crinier2021@meta.data$orig.ident]

# convert to single cell experiment
crinier2021_sc <- as.SingleCellExperiment(crinier2021)

# explore sc object
#assays(crinier2021_sc)

# Dimensions of gene-by-cell count matrix
dim(counts(crinier2021_sc))

# Access dgCMatrix with gene counts
#counts(crinier2021_sc)

# metadata
table(crinier2021_sc$batch)
table(crinier2021_sc$cluster2)

head(colData(crinier2021_sc),10)

# aggregate counts
adult_NK_agg_counts <- aggregateBioVar(scExp = crinier2021_sc,
                                       subjectVar = "orig.ident", 
                                       cellVar = "cluster2")

# Heatmap of log2 expression from aggregate gene-by-subject count matrix
ann_colors <- list(
    Sex = c(male = "darkblue", 
            female = "hotpink"),
    Donor = COLOR_CODE_DONORS)

scPHeatmap(
    sumExp = adult_NK_agg_counts$hNK_Bm2, 
    logSample = 1:100,
    subjectVar = "orig.ident",
    gtVar = "sex",
    color = viridis::viridis(75), 
    annotation_colors = ann_colors,
    treeheight_row = 0, 
    treeheight_col = 0
)
```


## ___DEG analysis: Adult vs. child BM-NK cells___

### ___DEG analysis: hNK_Bm1 cluster___

```{r DEG adult vs. toddler BM: hNK_Bm1, results='asis'}
df_toddler <- assay(CTRL_NK_agg_counts$hNK_Bm1, "counts")
df_adult <- assay(adult_NK_agg_counts$hNK_Bm1, "counts")

dim(df_toddler)
dim(df_adult)

# Check that the rownames match 
common_rownames <- intersect(rownames(df_adult), rownames(df_toddler))

# Subset the dataframes to the common genes
df_adult <- df_adult[common_rownames, , drop = FALSE]
df_toddler <- df_toddler[common_rownames, , drop = FALSE]

# Combine the dataframes by columns
merged_df <- cbind(df_adult, df_toddler)

# View the dimensions of the merged dataframe
dim(merged_df)

# merge metadata
metadata_toddler <- colData(CTRL_NK_agg_counts$hNK_Bm1)
metadata_adult <- colData(adult_NK_agg_counts$hNK_Bm1)

# match column names
colnames(metadata_adult) <- c("donor", "batch","cluster","predicted_NK_type_de.n350","sex","run") 
metadata_adult <- subset(metadata_adult,select = "donor")

metadata_toddler$donor <- metadata_toddler$orig.ident
metadata_toddler <- subset(metadata_toddler,select = "orig.ident")

colnames(metadata_toddler) <- "donor"


# merge metadata
merged_metadata <- rbind(metadata_adult,metadata_toddler)

# add comparison groups
merged_metadata$group <- c("adult","adult","adult","adult",
                           "adult","adult","adult","adult",
                           "toddler","toddler","toddler","toddler","toddler")


rownames(merged_metadata) <- paste(as.character(merged_metadata$group),
                                   as.character(merged_metadata$donor),
                                   sep="")
rownames(merged_metadata)
colnames(merged_df) <- c("adult4","adult1","adult2","adult5","adult6",
                         "adult7","adult3","adult8","toddler1","toddler5",
                         "toddler4","toddler2","toddler3")

# Drop unused factor levels
merged_metadata$donor <- factor(merged_metadata$donor)
merged_metadata$donor <- droplevels(merged_metadata$donor)
merged_metadata$donor 

merged_metadata$group <- factor(merged_metadata$group)
merged_metadata$group <- droplevels(merged_metadata$group)
merged_metadata$group

rownames(merged_metadata) <- NULL

# bulk RNA-seq DEG analysis
deseq <-
    DESeqDataSetFromMatrix(
        countData = merged_df,
        colData = merged_metadata,
        design = ~ group  + donor  
    )

# Transform counts for data visualization
rld <- rlog(deseq, blind=TRUE)

# Plot PCA
pca_data <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "group", returnData = TRUE) 
  
# Calculate percent variance
percentVar <- round(100 * attr(pca_data, "percentVar"))

# Custom PCA plot using ggplot2
ggplot(pca_data, aes(x = PC1, y = PC2, label = name, color = group)) +
  geom_text(vjust = -0.5, hjust = 0.5) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_color_manual(values=c("#161B33","#419D78"))  +
  labs(color = "Group", title = "DEGs Adult vs. toddler BM-NK cluster 1") +
  theme_light(base_size = 18) +  geom_point(size = 4)

# DEG analysis
deseq_analysis <- DESeq(deseq)

deseq_results_Adultvstoddler_c1 <-
    results(deseq_analysis, contrast = c("group", "adult", "toddler"))
head(deseq_results_Adultvstoddler_c1,50)
```

### ___DEG analysis: hNK_Bm2 cluster___

```{r DEG adult vs. toddler BM: hNK_Bm2, results='asis'}
df_toddler <- assay(CTRL_NK_agg_counts$hNK_Bm2, "counts")
df_adult <- assay(adult_NK_agg_counts$hNK_Bm2, "counts")

dim(df_toddler)
dim(df_adult)

# Check that the rownames match 
common_rownames <- intersect(rownames(df_adult), rownames(df_toddler))

# Subset the dataframes to the common genes
df_adult <- df_adult[common_rownames, , drop = FALSE]
df_toddler <- df_toddler[common_rownames, , drop = FALSE]

# Combine the dataframes by columns
merged_df <- cbind(df_adult, df_toddler)

# View the dimensions of the merged dataframe
dim(merged_df)

# merge metadata
metadata_toddler <- colData(CTRL_NK_agg_counts$hNK_Bm2)
metadata_adult <- colData(adult_NK_agg_counts$hNK_Bm2)

# match column names
colnames(metadata_adult) <- c("donor", "batch","cluster","predicted_NK_type_de.n350","sex","run") 
metadata_adult <- subset(metadata_adult,select = "donor")

metadata_toddler$donor <- metadata_toddler$orig.ident
metadata_toddler <- subset(metadata_toddler,select = "orig.ident")

colnames(metadata_toddler) <- "donor"


# merge metadata
merged_metadata <- rbind(metadata_adult,metadata_toddler)

# add comparison groups
merged_metadata$group <- c("adult","adult","adult","adult",
                           "adult","adult","adult","adult",
                           "toddler","toddler","toddler","toddler","toddler")


rownames(merged_metadata) <- paste(as.character(merged_metadata$group),
                                   as.character(merged_metadata$donor),
                                   sep="")
merged_metadata

colnames(merged_df) <- c("adult4","adult1","adult2","adult5","adult6",
                         "adult7","adult3","adult8","toddler1","toddler5",
                         "toddler4","toddler2","toddler3")

# Drop unused factor levels
merged_metadata$donor <- factor(merged_metadata$donor)
merged_metadata$donor <- droplevels(merged_metadata$donor)
merged_metadata$donor 

merged_metadata$group <- factor(merged_metadata$group)
merged_metadata$group <- droplevels(merged_metadata$group)
merged_metadata$group

rownames(merged_metadata) <- NULL

# bulk RNA-seq DEG analysis
deseq <-
    DESeqDataSetFromMatrix(
        countData = merged_df,
        colData = merged_metadata,
        design = ~ group  + donor  
    )

# Transform counts for data visualization
rld <- rlog(deseq, blind=TRUE)

# Plot PCA
pca_data <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "group", returnData = TRUE) 
  
# Calculate percent variance
percentVar <- round(100 * attr(pca_data, "percentVar"))

# Custom PCA plot using ggplot2
ggplot(pca_data, aes(x = PC1, y = PC2, label = name, color = group)) +
  geom_text(vjust = -0.5, hjust = 0.5) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_color_manual(values=c("#161B33","#419D78"))  +
  labs(color = "Group", title = "DEGs Adult vs. toddler BM-NK cluster 2") +
  theme_light(base_size = 18) +  geom_point(size = 4)


# DEG analysis
deseq_analysis <- DESeq(deseq)

deseq_results_Adultvstoddler_c2 <-
    results(deseq_analysis, contrast = c("group", "adult", "toddler"))
head(deseq_results_Adultvstoddler_c2,50)
```

### ___DEG analysis: hNK_Bm3 cluster___

```{r DEG adult vs. toddler BM: hNK_Bm3, results='asis'}
df_toddler <- assay(CTRL_NK_agg_counts$hNK_Bm3, "counts")
df_adult <- assay(adult_NK_agg_counts$hNK_Bm3, "counts")

dim(df_toddler)
dim(df_adult)

# Check that the rownames match 
common_rownames <- intersect(rownames(df_adult), rownames(df_toddler))

# Subset the dataframes to the common genes
df_adult <- df_adult[common_rownames, , drop = FALSE]
df_toddler <- df_toddler[common_rownames, , drop = FALSE]

# Combine the dataframes by columns
merged_df <- cbind(df_adult, df_toddler)

# View the dimensions of the merged dataframe
dim(merged_df)

# merge metadata
metadata_toddler <- colData(CTRL_NK_agg_counts$hNK_Bm3)
metadata_adult <- colData(adult_NK_agg_counts$hNK_Bm3)

# match column names
colnames(metadata_adult) <- c("donor", "batch","cluster","predicted_NK_type_de.n350","sex","run") 
metadata_adult <- subset(metadata_adult,select = "donor")
metadata_toddler$donor <- metadata_toddler$orig.ident
metadata_toddler <- subset(metadata_toddler,select = "orig.ident")

colnames(metadata_toddler) <- "donor"


# merge metadata
merged_metadata <- rbind(metadata_adult,metadata_toddler)

# add comparison groups
merged_metadata$group <- c("adult","adult","adult","adult",
                           "adult","adult","adult","adult",
                           "toddler","toddler","toddler","toddler","toddler")

rownames(merged_metadata) <- paste(as.character(merged_metadata$group),
                                   as.character(merged_metadata$donor),
                                   sep="")
merged_metadata
colnames(merged_df) <- c("adult4","adult1","adult2","adult5","adult6",
                         "adult7","adult3","adult8","toddler1","toddler5",
                         "toddler4","toddler2","toddler3")

# Drop unused factor levels
merged_metadata$donor <- factor(merged_metadata$donor)
merged_metadata$donor <- droplevels(merged_metadata$donor)
merged_metadata$donor 

merged_metadata$group <- factor(merged_metadata$group)
merged_metadata$group <- droplevels(merged_metadata$group)
merged_metadata$group

rownames(merged_metadata) <- NULL

# bulk RNA-seq DEG analysis
deseq <-
    DESeqDataSetFromMatrix(
        countData = merged_df,
        colData = merged_metadata,
        design = ~ group  + donor  
    )

# Transform counts for data visualization
rld <- rlog(deseq, blind=TRUE)

# Plot PCA
# Plot PCA
pca_data <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "group", returnData = TRUE) 
  
# Calculate percent variance
percentVar <- round(100 * attr(pca_data, "percentVar"))

# Custom PCA plot using ggplot2
ggplot(pca_data, aes(x = PC1, y = PC2, label = name, color = group)) +
  geom_text(vjust = -0.5, hjust = 0.5) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_color_manual(values=c("#161B33","#419D78"))  +
  labs(color = "Group", title = "DEGs Adult vs. toddler BM-NK cluster 3") +
  theme_light(base_size = 18) +  geom_point(size = 4)

# DEG analysis
deseq_analysis <- DESeq(deseq)

deseq_results_Adultvstoddler_c3 <-
    results(deseq_analysis, contrast = c("group", "adult", "toddler"))
head(deseq_results_Adultvstoddler_c3,50)
```

### ___DEG analysis: hNK_Bm4 cluster___

```{r DEG adult vs. toddler BM: hNK_Bm4, results='asis'}
df_toddler <- assay(CTRL_NK_agg_counts$hNK_Bm4, "counts")
df_adult <- assay(adult_NK_agg_counts$hNK_Bm4, "counts")

dim(df_toddler)
dim(df_adult)

# Check that the rownames match 
common_rownames <- intersect(rownames(df_adult), rownames(df_toddler))

# Subset the dataframes to the common genes
df_adult <- df_adult[common_rownames, , drop = FALSE]
df_toddler <- df_toddler[common_rownames, , drop = FALSE]

# Combine the dataframes by columns
merged_df <- cbind(df_adult, df_toddler)

# View the dimensions of the merged dataframe
dim(merged_df)

# merge metadata
metadata_toddler <- colData(CTRL_NK_agg_counts$hNK_Bm4)
metadata_adult <- colData(adult_NK_agg_counts$hNK_Bm4)

# match column names
colnames(metadata_adult) <- c("donor", "batch","cluster","predicted_NK_type_de.n350","sex","run") 
metadata_adult <- subset(metadata_adult,select = "donor")
metadata_toddler$donor <- metadata_toddler$orig.ident
metadata_toddler <- subset(metadata_toddler,select = "orig.ident")

colnames(metadata_toddler) <- "donor"


# merge metadata
merged_metadata <- rbind(metadata_adult,metadata_toddler)

# add comparison groups (not all adult BM samples have cluster 4)
merged_metadata$group <- c("adult","adult","adult", 
                           "toddler","toddler","toddler","toddler","toddler")


rownames(merged_metadata) <- paste(as.character(merged_metadata$group),
                                   as.character(merged_metadata$donor),
                                   sep="")
merged_metadata
colnames(merged_df) <- c("adult3","adult2","adult5",
                         "toddler1","toddler5",
                         "toddler4","toddler2","toddler3")

# Drop unused factor levels
merged_metadata$donor <- factor(merged_metadata$donor)
merged_metadata$donor <- droplevels(merged_metadata$donor)
merged_metadata$donor 

merged_metadata$group <- factor(merged_metadata$group)
merged_metadata$group <- droplevels(merged_metadata$group)
merged_metadata$group

rownames(merged_metadata) <- NULL

# bulk RNA-seq DEG analysis
deseq <-
    DESeqDataSetFromMatrix(
        countData = merged_df,
        colData = merged_metadata,
        design = ~ group  + donor  
    )

# Transform counts for data visualization
rld <- rlog(deseq, blind=TRUE)

# Plot PCA
# Plot PCA
pca_data <- DESeq2::plotPCA(rld, ntop = 1000, intgroup = "group", returnData = TRUE) 
  
# Calculate percent variance
percentVar <- round(100 * attr(pca_data, "percentVar"))

# Custom PCA plot using ggplot2
ggplot(pca_data, aes(x = PC1, y = PC2, label = name, color = group)) +
  geom_text(vjust = -0.5, hjust = 0.5) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_color_manual(values=c("#161B33","#419D78"))  +
  labs(color = "Group", title = "DEGs Adult vs. toddler BM-NK cluster 4") +
  theme_light(base_size = 18) +  geom_point(size = 4)

# DEG analysis
deseq_analysis <- DESeq(deseq)

deseq_results_Adultvstoddler_c4 <-
    results(deseq_analysis, contrast = c("group", "adult", "toddler"))
head(deseq_results_Adultvstoddler_c4,50)
```

## ___DEG analysis: tumor-infiltrated vs. non-infiltrated BM-NK___

### ___DEG analysis: hNK_Bm1 cluster___

Cluster 1 is only present in small amounts in one ATRX samples/patient, while
missing in the other sample/patient

### ___DEG analysis: hNK_Bm2 cluster___

```{r DEG NB vs. toddler BM: hNK_Bm2, results='asis'}
df_toddler <- assay(CTRL_NK_agg_counts$hNK_Bm2, "counts")
df_ATRX <- assay(ATRX_NK_agg_counts$hNK_Bm2, "counts")

dim(df_toddler)
dim(df_ATRX)

# Check that the rownames match 
common_rownames <- intersect(rownames(df_ATRX), rownames(df_toddler))

# Subset the dataframes to the common genes
df_ATRX <- df_ATRX[common_rownames, , drop = FALSE]
df_toddler <- df_toddler[common_rownames, , drop = FALSE]

# Combine the dataframes by columns
merged_df <- cbind(df_ATRX, df_toddler)

# View the dimensions of the merged dataframe
dim(merged_df)

# merge metadata
metadata_toddler <- colData(CTRL_NK_agg_counts$hNK_Bm2)
metadata_ATRX <- colData(ATRX_NK_agg_counts$hNK_Bm2)

# match column names
colnames(metadata_ATRX) <- c("donor", "sample","group","patients","predicted_NK_type_de.n350","ident") 
metadata_ATRX <- subset(metadata_ATRX,select = "donor")
colnames(metadata_toddler) <- c("orig.ident","sample", "group","donor","predicted_NK_type_de.n350","ident")
metadata_toddler <- subset(metadata_toddler,select = "orig.ident")

colnames(metadata_toddler) <- "donor"

# merge metadata
merged_metadata <- rbind(metadata_ATRX,metadata_toddler)

# add comparison groups
merged_metadata$group <- c("ATRX","ATRX",
                           "toddler","toddler","toddler","toddler","toddler")

# unique column names with group and donor info
merged_metadata$donor = gsub("ctrl_0","",merged_metadata$donor)
merged_metadata$donor = gsub("ATRXmut_0","",merged_metadata$donor)
kable(merged_metadata)

rownames(merged_metadata) <- paste(as.character(merged_metadata$group),
                                   as.character(merged_metadata$donor),
                                   sep="")
rownames(merged_metadata)
colnames(merged_df) <- c("ATRX2","ATRX1",
                         "toddler1","toddler5","toddler4", "toddler2","toddler3")

# Drop unused factor levels
merged_metadata$donor <- factor(merged_metadata$donor)
merged_metadata$donor <- droplevels(merged_metadata$donor)
merged_metadata$donor 

merged_metadata$group <- factor(merged_metadata$group)
merged_metadata$group <- droplevels(merged_metadata$group)
merged_metadata$group

rownames(merged_metadata) <- NULL

# bulk RNA-seq DEG analysis
deseq <-
    DESeqDataSetFromMatrix(
        countData = merged_df,
        colData = merged_metadata,
        design = ~ group + donor 
    )

# Transform counts for data visualization
rld <- rlog(deseq, blind=TRUE)

# Plot PCA
DESeq2::plotPCA(rld, ntop = 1000, intgroup = "group") + 
  scale_color_manual(values=c("#161B33","#419D78"))  +
  labs(color = "Group", title = "DEGs ATRXmut NB vs. toddler BM-NK cluster 2") +
  theme_light(base_size = 18) +  geom_point(size = 4)

DESeq2::plotPCA(rld, ntop = 500, intgroup = "sizeFactor") + 
  scale_color_continuous(type = "gradient")  +
  labs(color = "Cell number", title = "DEGs ATRXmut NB vs. toddler BM-NK cluster 2") +
  theme_light(base_size = 18) +  geom_point(size = 4)

# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- assay(rld)
rld_cor <- cor(rld_mat)

# Plot heatmap
pheatmap(rld_cor)

# DEG analysis
deseq_analysis <- DESeq(deseq)

plotDispEsts(deseq_analysis)

deseq_results_ATRXvstoddler_c2 <-
    results(deseq_analysis, contrast = c("group", "ATRX", "toddler"))

head(deseq_results_ATRXvstoddler_c2,50)

# Shrink the log2 fold changes to be more appropriate using the apeglm method - should cite [paper]() when using this method
# Generate results object
#resultsNames(deseq_results_ATRXvstoddler_c2)
deseq_results_ATRXvstoddler_c2 <- results(deseq_analysis, 
                                          name = "group_toddler_vs_ATRX",
                                          alpha = 0.05)

res <- lfcShrink(deseq_analysis, 
                 coef = "group_toddler_vs_ATRX",
                 res = deseq_results_ATRXvstoddler_c2,
                 type = "apeglm")

head(res, 50)
```

### ___DEG analysis: hNK_Bm3 cluster___

```{r DEG NB vs. toddler BM: hNK_Bm3, results='asis'}
df_toddler <- assay(CTRL_NK_agg_counts$hNK_Bm3, "counts")
df_ATRX <- assay(ATRX_NK_agg_counts$hNK_Bm3, "counts")

dim(df_toddler)
dim(df_ATRX)

# Check that the rownames match 
common_rownames <- intersect(rownames(df_ATRX), rownames(df_toddler))

# Subset the dataframes to the common genes
df_ATRX <- df_ATRX[common_rownames, , drop = FALSE]
df_toddler <- df_toddler[common_rownames, , drop = FALSE]

# Combine the dataframes by columns
merged_df <- cbind(df_ATRX, df_toddler)

# View the dimensions of the merged dataframe
dim(merged_df)

# merge metadata
metadata_toddler <- colData(CTRL_NK_agg_counts$hNK_Bm2)
metadata_ATRX <- colData(ATRX_NK_agg_counts$hNK_Bm2)

# match column names
colnames(metadata_ATRX) <- c("donor", "sample","group","patients","predicted_NK_type_de.n350","ident") 
metadata_ATRX <- subset(metadata_ATRX,select = "donor")
colnames(metadata_toddler) <- c("orig.ident","sample", "group","donor","predicted_NK_type_de.n350","ident")
metadata_toddler <- subset(metadata_toddler,select = "orig.ident")

colnames(metadata_toddler) <- "donor"

# merge metadata
merged_metadata <- rbind(metadata_ATRX,metadata_toddler)

# add comparison groups
merged_metadata$group <- c("ATRX","ATRX",
                           "toddler","toddler","toddler","toddler","toddler")

# unique column names with group and donor info
merged_metadata$donor = gsub("ctrl_0","",merged_metadata$donor)
merged_metadata$donor = gsub("ATRXmut_0","",merged_metadata$donor)
kable(merged_metadata)

rownames(merged_metadata) <- paste(as.character(merged_metadata$group),
                                   as.character(merged_metadata$donor),
                                   sep="")
rownames(merged_metadata)
colnames(merged_df) <- c("ATRX2","ATRX1",
                         "toddler1","toddler5","toddler4", "toddler2","toddler3")

# Drop unused factor levels
merged_metadata$donor <- factor(merged_metadata$donor)
merged_metadata$donor <- droplevels(merged_metadata$donor)
merged_metadata$donor 

merged_metadata$group <- factor(merged_metadata$group)
merged_metadata$group <- droplevels(merged_metadata$group)
merged_metadata$group

rownames(merged_metadata) <- NULL

# bulk RNA-seq DEG analysis
deseq <-
    DESeqDataSetFromMatrix(
        countData = merged_df,
        colData = merged_metadata,
        design = ~ group + donor 
    )

# Transform counts for data visualization
rld <- rlog(deseq, blind=TRUE)

# Plot PCA
DESeq2::plotPCA(rld, ntop = 1000, intgroup = "group") + 
  scale_color_manual(values=c("#161B33","#419D78"))  +
  labs(color = "Group", title = "DEGs ATRXmut NB vs. toddler BM-NK cluster 3") +
  theme_light(base_size = 18) +  geom_point(size = 4)

DESeq2::plotPCA(rld, ntop = 500, intgroup = "sizeFactor") + 
  scale_color_continuous(type = "gradient")  +
  labs(color = "Cell number", title = "DEGs ATRXmut NB vs. toddler BM-NK cluster 3") +
  theme_light(base_size = 18) +  geom_point(size = 4)

# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- assay(rld)
rld_cor <- cor(rld_mat)

# Plot heatmap
pheatmap(rld_cor)

# DEG analysis
deseq_analysis <- DESeq(deseq)

plotDispEsts(deseq_analysis)

deseq_results_ATRXvstoddler_c3 <-
    results(deseq_analysis, contrast = c("group", "ATRX", "toddler"))
head(deseq_results_ATRXvstoddler_c3,50)

# Shrink the log2 fold changes to be more appropriate using the apeglm method - should cite [paper]() when using this method
# Generate results object
#resultsNames(deseq_results_ATRXvstoddler_c3)
deseq_results_ATRXvstoddler_c3 <- results(deseq_analysis, 
                                          name = "group_toddler_vs_ATRX",
                                          alpha = 0.05)

res <- lfcShrink(deseq_analysis, 
                 coef = "group_toddler_vs_ATRX",
                 res = deseq_results_ATRXvstoddler_c3,
                 type = "apeglm")

head(res, 50)
```

### ___DEG analysis: hNK_Bm4 cluster___

```{r DEG NB vs. toddler BM: hNK_Bm4, results='asis'}
df_toddler <- assay(CTRL_NK_agg_counts$hNK_Bm4, "counts")
df_ATRX <- assay(ATRX_NK_agg_counts$hNK_Bm4, "counts")

dim(df_toddler)
dim(df_ATRX)

# Check that the rownames match 
common_rownames <- intersect(rownames(df_ATRX), rownames(df_toddler))

# Subset the dataframes to the common genes
df_ATRX <- df_ATRX[common_rownames, , drop = FALSE]
df_toddler <- df_toddler[common_rownames, , drop = FALSE]

# Combine the dataframes by columns
merged_df <- cbind(df_ATRX, df_toddler)

# View the dimensions of the merged dataframe
dim(merged_df)

# merge metadata
metadata_toddler <- colData(CTRL_NK_agg_counts$hNK_Bm2)
metadata_ATRX <- colData(ATRX_NK_agg_counts$hNK_Bm2)

# match column names
colnames(metadata_ATRX) <- c("donor", "sample","group","patients","predicted_NK_type_de.n350","ident") 
metadata_ATRX <- subset(metadata_ATRX,select = "donor")
colnames(metadata_toddler) <- c("orig.ident","sample", "group","donor","predicted_NK_type_de.n350","ident")
metadata_toddler <- subset(metadata_toddler,select = "orig.ident")

colnames(metadata_toddler) <- "donor"

# merge metadata
merged_metadata <- rbind(metadata_ATRX,metadata_toddler)

# add comparison groups
merged_metadata$group <- c("ATRX","ATRX",
                           "toddler","toddler","toddler","toddler","toddler")

# unique column names with group and donor info
merged_metadata$donor = gsub("ctrl_0","",merged_metadata$donor)
merged_metadata$donor = gsub("ATRXmut_0","",merged_metadata$donor)
kable(merged_metadata)

rownames(merged_metadata) <- paste(as.character(merged_metadata$group),
                                   as.character(merged_metadata$donor),
                                   sep="")
rownames(merged_metadata)
colnames(merged_df) <- c("ATRX2","ATRX1",
                         "toddler1","toddler5","toddler4", "toddler2","toddler3")

# Drop unused factor levels
merged_metadata$donor <- factor(merged_metadata$donor)
merged_metadata$donor <- droplevels(merged_metadata$donor)
merged_metadata$donor 

merged_metadata$group <- factor(merged_metadata$group)
merged_metadata$group <- droplevels(merged_metadata$group)
merged_metadata$group

rownames(merged_metadata) <- NULL

# bulk RNA-seq DEG analysis
deseq <-
    DESeqDataSetFromMatrix(
        countData = merged_df,
        colData = merged_metadata,
        design = ~ group + donor 
    )

# Transform counts for data visualization
rld <- rlog(deseq, blind=TRUE)

# Plot PCA
DESeq2::plotPCA(rld, ntop = 1000, intgroup = "group") + 
  scale_color_manual(values=c("#161B33","#419D78"))  +
  labs(color = "Group", title = "DEGs ATRXmut NB vs. toddler BM-NK cluster 4") +
  theme_light(base_size = 18) +  geom_point(size = 4)

DESeq2::plotPCA(rld, ntop = 500, intgroup = "sizeFactor") + 
  scale_color_continuous(type = "gradient")  +
  labs(color = "Cell number", title = "DEGs ATRXmut NB vs. toddler BM-NK cluster 4") +
  theme_light(base_size = 18) +  geom_point(size = 4)

# Extract the rlog matrix from the object and compute pairwise correlation values
rld_mat <- assay(rld)
rld_cor <- cor(rld_mat)

# Plot heatmap
pheatmap(rld_cor)

# DEG analysis
deseq_analysis <- DESeq(deseq)

plotDispEsts(deseq_analysis)

deseq_results_ATRXvstoddler_c4 <-
    results(deseq_analysis, contrast = c("group", "ATRX", "toddler"))
head(deseq_results_ATRXvstoddler_c4,50)

# Shrink the log2 fold changes to be more appropriate using the apeglm method - should cite [paper]() when using this method
# Generate results object
#resultsNames(deseq_results_ATRXvstoddler_c4)
deseq_results_ATRXvstoddler_c4 <- results(deseq_analysis, 
                                          name = "group_toddler_vs_ATRX",
                                          alpha = 0.05)

res <- lfcShrink(deseq_analysis, 
                 coef = "group_toddler_vs_ATRX",
                 res = deseq_results_ATRXvstoddler_c4,
                 type = "apeglm")

head(res, 50)
```

## ___Compare results: Adult vs. toddler BM-NK and Healthy vs. tumor-infiltrated BM-NK___

## ___Adult BM-NK vs. toddler BM-NK___

```{r compare results, results='asis'}
# Adult vs. toddler BM: Cluster 1
pos_val <- deseq_results_Adultvstoddler_c1$baseMean > 0
deseq_results_Adultvstoddler_posVal <- deseq_results_Adultvstoddler_c1[pos_val,]

deseq_results_Adultvstoddler_transf_c1 <- as.data.frame(deseq_results_Adultvstoddler_posVal) %>%
    bind_cols(feature = rownames(deseq_results_Adultvstoddler_posVal)) %>%
    mutate(log_padj = - log(.data$padj, base = 10))
head(deseq_results_Adultvstoddler_transf_c1,50)

deseq_results_Adultvstoddler_signif_c1 <- 
  deseq_results_Adultvstoddler_transf_c1 %>% 
  filter(padj < 0.1)

head(deseq_results_Adultvstoddler_signif_c1,50)

# Adult vs. toddler BM: Cluster 2
pos_val <- deseq_results_Adultvstoddler_c2$baseMean > 0
deseq_results_Adultvstoddler_posVal <- deseq_results_Adultvstoddler_c2[pos_val,]

deseq_results_Adultvstoddler_transf_c2 <- as.data.frame(deseq_results_Adultvstoddler_posVal) %>%
    bind_cols(feature = rownames(deseq_results_Adultvstoddler_posVal)) %>%
    mutate(log_padj = - log(.data$padj, base = 10))

deseq_results_Adultvstoddler_signif_c2 <- 
  deseq_results_Adultvstoddler_transf_c2 %>% 
  filter(padj < 0.1)

head(deseq_results_Adultvstoddler_signif_c2,50)

# Adult vs. toddler BM: Cluster 3
pos_val <- deseq_results_Adultvstoddler_c3$baseMean > 0
deseq_results_Adultvstoddler_posVal <- deseq_results_Adultvstoddler_c3[pos_val,]

deseq_results_Adultvstoddler_transf_c3 <- as.data.frame(deseq_results_Adultvstoddler_posVal) %>%
    bind_cols(feature = rownames(deseq_results_Adultvstoddler_posVal)) %>%
    mutate(log_padj = - log(.data$padj, base = 10))

deseq_results_Adultvstoddler_signif_c3 <- 
  deseq_results_Adultvstoddler_transf_c3 %>% 
  filter(padj < 0.1)

head(deseq_results_Adultvstoddler_signif_c3,50)

# Adult vs. toddler BM: Cluster 4
pos_val <- deseq_results_Adultvstoddler_c4$baseMean > 0
deseq_results_Adultvstoddler_posVal <- deseq_results_Adultvstoddler_c4[pos_val,]

deseq_results_Adultvstoddler_transf_c4 <- as.data.frame(deseq_results_Adultvstoddler_posVal) %>%
    bind_cols(feature = rownames(deseq_results_Adultvstoddler_posVal)) %>%
    mutate(log_padj = - log(.data$padj, base = 10))

deseq_results_Adultvstoddler_signif_c4 <- 
  deseq_results_Adultvstoddler_transf_c4 %>% 
  filter(padj < 0.1)

head(deseq_results_Adultvstoddler_signif_c4,50)
```

```{r}
# Vulcano plot: Adult vs toddler BM-NK cluster 1, DEGs
ggplot_Adultvstoddler_c1 <- ggplot(data = deseq_results_Adultvstoddler_transf_c1) +
    geom_point(aes(x = log2FoldChange, y = log_padj), na.rm = TRUE) +
    geom_point(
        data = filter(
            .data = deseq_results_Adultvstoddler_transf_c1,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.05
        ),
        aes(x = log2FoldChange, y = log_padj), color = "red"
    ) +
    geom_label(
        data = filter(
            .data = deseq_results_Adultvstoddler_transf_c1,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.01
        ),
        aes(x = log2FoldChange + 0.5, y = log_padj + 5, label = feature)
    ) +
    deseq_themes() + ylim(c(0,20))


# Vulcano plot: Adult vs toddler BM-NK cluster 2, DEGs
ggplot_Adultvstoddler_c2 <- ggplot(data = deseq_results_Adultvstoddler_transf_c2) +
    geom_point(aes(x = log2FoldChange, y = log_padj), na.rm = TRUE) +
    geom_point(
        data = filter(
            .data = deseq_results_Adultvstoddler_transf_c2,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.05
        ),
        aes(x = log2FoldChange, y = log_padj), color = "red"
    ) +
    geom_label(
        data = filter(
            .data = deseq_results_Adultvstoddler_transf_c2,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.01
        ),
        aes(x = log2FoldChange + 0.5, y = log_padj + 5, label = feature)
    ) +
    deseq_themes() + ylim(c(0,20))

# Vulcano plot: Adult vs toddler BM-NK cluster 3, DEGs
ggplot_Adultvstoddler_c3 <- ggplot(data = deseq_results_Adultvstoddler_transf_c3) +
    geom_point(aes(x = log2FoldChange, y = log_padj), na.rm = TRUE) +
    geom_point(
        data = filter(
            .data = deseq_results_Adultvstoddler_transf_c3,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.05
        ),
        aes(x = log2FoldChange, y = log_padj), color = "red"
    ) +
    geom_label(
        data = filter(
            .data = deseq_results_Adultvstoddler_transf_c3,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.01
        ),
        aes(x = log2FoldChange + 0.5, y = log_padj + 5, label = feature)
    ) +
    deseq_themes() + ylim(c(0,20))

# Vulcano plot: Adult vs toddler BM-NK cluster 4, DEGs
ggplot_Adultvstoddler_c4 <- ggplot(data = deseq_results_Adultvstoddler_transf_c4) +
    geom_point(aes(x = log2FoldChange, y = log_padj), na.rm = TRUE) +
    geom_point(
        data = filter(
            .data = deseq_results_Adultvstoddler_transf_c4,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.05
        ),
        aes(x = log2FoldChange, y = log_padj), color = "red"
    ) +
    geom_label(
        data = filter(
            .data = deseq_results_Adultvstoddler_transf_c4,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.01
        ),
        aes(x = log2FoldChange + 0.5, y = log_padj + 5, label = feature)
    ) +
    deseq_themes() + ylim(c(0,20))
```

```{r plots}
ggplot_Adultvstoddler_c1
ggplot_Adultvstoddler_c2
ggplot_Adultvstoddler_c3
ggplot_Adultvstoddler_c4
```

## ___ATRXmutated NB BM-NK vs. healthy toddler BM-NK___

```{r}
pos_val <- deseq_results_ATRXvstoddler_c2$baseMean > 0
deseq_results_ATRXvstoddler_c2_posVal <- deseq_results_ATRXvstoddler_c2[pos_val,]

deseq_results_ATRXvstoddler_c2_transf <- as.data.frame(deseq_results_ATRXvstoddler_c2_posVal) %>%
    bind_cols(feature = rownames(deseq_results_ATRXvstoddler_c2_posVal)) %>%
    mutate(log_padj = - log(.data$padj, base = 10))

deseq_results_ATRXvstoddler_c2_signif <- 
  deseq_results_ATRXvstoddler_c2_transf %>% 
  filter(padj < 0.1)

head(deseq_results_ATRXvstoddler_c2_signif,50)

# Vulcano plot: ATRXmut vs toddler BM-NK cluster 2, DEGs
ggplot_ATRXvstoddler_c2 <- ggplot(data = deseq_results_ATRXvstoddler_c2_transf) +
    geom_point(aes(x = log2FoldChange, y = log_padj), na.rm = TRUE) +
    geom_point(
        data = filter(
            .data = deseq_results_ATRXvstoddler_c2_transf,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.05
        ),
        aes(x = log2FoldChange, y = log_padj), color = "red"
    ) +
    geom_label(
        data = filter(
            .data = deseq_results_ATRXvstoddler_c2_transf,
            abs(.data$log2FoldChange) > 1, .data$padj < 0.01
        ),
        aes(x = log2FoldChange + 0.5, y = log_padj + 5, label = feature)
    ) +
    deseq_themes() + ylim(c(0,5))
```

```{r plots - ATRX}
ggplot_ATRXvstoddler_c2
```


# ___Session Info___

```{r}
sessionInfo()
```
