---
title: "NK project: Desensetized NK cells & NK-MO-DC regulation axis "
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
---

PROJECT DESCRIPTION

Identify NK sub-populations in tumor-infiltrated bone marrow samples from
Neuroblastoma patients.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_utils.R")
```

# ___Load data___

```{r}
fetahu2023 <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_withMetadata.Rds"))
predicted_labels <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_predicted_NKclusters_from_Crinier2021.Rds"))
ALL_NK <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_NK_cluster.Rds"))
ATRX_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXmut_BM_NK.Rds"))
toddler_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_fetal_BM_NK.Rds"))
```

#  ___Exploring prediction scores for cluster 2___

Some cells have a higher prediction score for cluster 3, but were predicted to
be part of cluster 2. Here i am exploring if this occurs also for clusters 1
and 4 or if this is exlusive to clusters 2 and 3. 

```{r}
# exploring prediction scores for cluster 2
Idents(ALL_NK) <- ALL_NK$predicted_NK_type
NKclust2 <- subset(ALL_NK, idents = "hNK_Bm2")
hist(NKclust2$scores.hNK_Bm2,plot = TRUE,breaks = 50)

# Extract the cell IDs where scores.hNK_Bm3 > scores.hNK_Bm2
cell_ids_might_be_clust3 <- WhichCells(NKclust2, expression = scores.hNK_Bm3 > scores.hNK_Bm2)

# Extract the cell IDs where scores.hNK_Bm1 > scores.hNK_Bm2
cell_ids_might_be_clust1 <- WhichCells(NKclust2, expression = scores.hNK_Bm1 > scores.hNK_Bm2)

# Extract the cell IDs where scores.hNK_Bm4 > scores.hNK_Bm2
cell_ids_might_be_clust4 <- WhichCells(NKclust2, expression = scores.hNK_Bm4 > scores.hNK_Bm2)

# Extract the cell IDs where scores.hNK_Bm2 > scores.hNK_Bm3
cell_ids_true_clust2 <- WhichCells(NKclust2, 
                                   expression =
                                     scores.hNK_Bm3 < scores.hNK_Bm2 &
                                     scores.hNK_Bm1 < scores.hNK_Bm2 &
                                     scores.hNK_Bm4  < scores.hNK_Bm2)


# Print or return the cell IDs
highlight <- c(cell_ids_might_be_clust3,cell_ids_might_be_clust1,cell_ids_might_be_clust4)

DimPlot(NKclust2, 
        cells.highlight = cell_ids_might_be_clust1,
                pt.size = 2,
        sizes.highlight = 2)
DimPlot(NKclust2, 
        cells.highlight = cell_ids_might_be_clust3,
                pt.size = 2,
        sizes.highlight = 2)
DimPlot(NKclust2, 
        cells.highlight = cell_ids_might_be_clust4,
        pt.size = 2,
        sizes.highlight = 2)
# all wrongly predicted cells 
DimPlot(NKclust2, 
        cells.highlight = highlight,
                pt.size = 2,
        sizes.highlight = 2)
DimPlot(NKclust2, 
        cells.highlight = cell_ids_true_clust2,
                pt.size = 2,
        sizes.highlight = 2)


# ATRX only
ATRX_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXmut_BM_NK.Rds"))

ATRX_clust2 <- subset(ATRX_NK, idents = "hNK_Bm2")

hist(predicted_labels$scores,plot = TRUE,breaks = 50)

```

# ___Desensitized/exhausted NK cells___

List of genes from NK Cell Exhaustion review paper from 2017 
DOI: 10.3389/fimmu.2017.00760

2B4 protein = CD244 gene
NKG2A protein = KLRC1 gene
NKG2D protein = KLRK1 gene
NRC proteins = NCOA1, NCOA2, NCOA3
CD16 protein = FCGR3A Gene gene
TIM3 protein = HAVCR2 gene


```{r}
exhausted_NK_genes <- list(up = c("PDCD1", "HAVCR2", "KLRC1"),
                           down = c("KLRK1","FCGR3A","NCOA1", "NCOA2", 
                                    "NCOA3","CD226","CD244","IFNG")) 

do_DotPlot(ALL_NK, 
           features = exhausted_NK_genes, 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(ALL_NK, 
           features = exhausted_NK_genes, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

```

# ___CCC analysis___

The degree to which NK cells are active against tumour cells is highly dependent 
on the activation state of the NK cells, which is regulated in large part by 
cytokines derived from dendritic cells (DCs) and monocytes

## ___NK cells: L-R expression per cluster per group___

```{r fig.height=5, fig.width=15}
# based on The cancer–natural killer cell immunity cycle review paper
# high or low expression of  these genes is associated with a good prognosis 
# in a pan-cancer analysis

NK_receptors <- c("ADORA2A","CD2","CD226", "CD244", "TIGIT","TMIGD2", 
                  "NCR1","NCR2","NCR3","NT5E", "ITGAL", "ITGB2","PDCD1",
                  "KLRC1","KLRC2", "KLRC3", "KLRC4", "KLRD1", "KLRF1",
                  "KLRG1", "KLRG2","KLRK1", "KIR2DL1", "KIR2DL3", "KIR2DL4", 
                  "KIR2DS4","KIR3DL1", "KIR3DL2", "KIR3DL3")

NK_secreted_cytokines <- c("CCL5","FLT3LG", "XCL1", "XCL2")

NK_prognosis_markers <- list(NK_receptors = c("ADORA2A","CD2","CD226", "CD244", 
                                              "TIGIT","TMIGD2", "NCR1","NCR2",
                                              "NCR3","NT5E", "ITGAL", "ITGB2",
                                              "PDCD1","KLRC1","KLRC2", "KLRC3", 
                                              "KLRC4", "KLRD1","KLRF1","KLRG1", 
                                              "KLRG2","KLRK1","KIR2DL1","KIR2DL3", 
                                              "KIR2DL4", "KIR3DL1", "KIR3DL2", 
                                              "KIR3DL3","KIR2DS4"),
                             NK_secreted_cytokines = c("CCL5","FLT3LG", "XCL1", "XCL2"))

do_DotPlot(ALL_NK, 
           features = NK_prognosis_markers, 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(ALL_NK, 
           features = NK_prognosis_markers, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

```


## ___NK - MO - DC axis___

```{r}
cell_types_of_interest_v2 <- c("NK (4)",
                               "pDC (12)",
                               "M (1)","M (2)", "M (10)","M (15)")

fetahu2023.NK.DC.M <- subset(fetahu2023,
                             idents = cell_types_of_interest_v2)

do_DotPlot(fetahu2023.NK.DC.M, 
           features = NK_activating_cytokines, 
           group.by = "orig.ident", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           plot.title = "NK-cell activating cytokines",
           plot.subtitle = "Monocytes",
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_ViolinPlot(fetahu2023.NK.DC.M, 
           features = c("IL2", "IL15", "IL12A", "IL12B","IL18"), 
           group.by = "orig.ident", 
           split.by = "group",
           plot_boxplot = FALSE,
           slot = "data")

saveRDS(fetahu2023.NK.DC.M, file = paste0(params$out,"Rds/fetahu2023_NK_DC_MO.Rds"))
```

### ___Immunosupresive Monocytes (M10)___

```{r}
cell_types_of_interest_v3 <- c("M (10)")

fetahu2023.M10 <- subset(fetahu2023,
                         idents = cell_types_of_interest_v3)

do_DotPlot(fetahu2023.M10, 
           features = NK_activating_cytokines, 
           plot.title = "NK-cell activating cytokines",
           plot.subtitle = "Immunosupresive Monocytes (M10)",
           group.by = "group", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )
```

### ___NK cells___

Chemokines secreated by NK cells and associated with recruiting type 1 
conventional dendritic cells

```{r}
do_DotPlot(ALL_NK, 
           features = c("CCL5","XCL1","FLT3LG"), 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(ALL_NK, 
           features = c("CCL5","XCL1","FLT3LG"), 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )
```

-------------------------------------------------------------------------------

# ___Session Info___

```{r}
sessionInfo()
```
