---
title: "NK project: Cluster markers"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
---

PROJECT DESCRIPTION

Identify NK sub-populations in tumor-infiltrated bone marrow samples from
Neuroblastoma patients.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(SingleR))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(stats))
suppressPackageStartupMessages(library(viridis))
suppressPackageStartupMessages(library(SCpubr))
suppressPackageStartupMessages(library(scCustomize))
suppressPackageStartupMessages(library(DElegate))
suppressPackageStartupMessages(library(clusterProfiler))
suppressPackageStartupMessages(library(org.Hs.eg.db))
suppressPackageStartupMessages(library(enrichplot))
suppressPackageStartupMessages(library(AnnotationDbi))
suppressPackageStartupMessages(library(pathview))
suppressPackageStartupMessages(library(aggregateBioVar))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(SingleCellExperiment))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(ggtext))

source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
```

# ___Load data___


```{r}
crinier2021 <- readRDS(file = paste0(params$input_BM1))
lee2023 <- readRDS(file = paste0(params$input_BM3))
```



```{r}
fetahu2023 <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_withMetadata.Rds"))
predicted_labels <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_predicted_NKclusters_from_Crinier2021.Rds"))
fetahu2023_NK <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_NK_cluster.Rds"))

ATRX_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXmut_BM_NK.Rds"))
fetal_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_fetal_BM_NK.Rds"))
```


#  ___Exploring prediction scores for cluster 2___

```{r}
# exploring prediction scores for cluster 2
Idents(fetahu2023_NK) <- fetahu2023_NK$predicted_NK_type
NKclust2 <- subset(fetahu2023_NK, idents = "hNK_Bm2")
hist(NKclust2$scores.hNK_Bm2,plot = TRUE,breaks = 50)

# Extract the cell IDs where scores.hNK_Bm3 > scores.hNK_Bm2
cell_ids_might_be_clust3 <- WhichCells(NKclust2, expression = scores.hNK_Bm3 > scores.hNK_Bm2)

# Extract the cell IDs where scores.hNK_Bm1 > scores.hNK_Bm2
cell_ids_might_be_clust1 <- WhichCells(NKclust2, expression = scores.hNK_Bm1 > scores.hNK_Bm2)

# Extract the cell IDs where scores.hNK_Bm4 > scores.hNK_Bm2
cell_ids_might_be_clust4 <- WhichCells(NKclust2, expression = scores.hNK_Bm4 > scores.hNK_Bm2)

# Extract the cell IDs where scores.hNK_Bm2 > scores.hNK_Bm3
cell_ids_true_clust2 <- WhichCells(NKclust2, 
                                   expression =
                                     scores.hNK_Bm3 < scores.hNK_Bm2 &
                                     scores.hNK_Bm1 < scores.hNK_Bm2 &
                                     scores.hNK_Bm4  < scores.hNK_Bm2)


# Print or return the cell IDs
highlight <- c(cell_ids_might_be_clust3,cell_ids_might_be_clust1,cell_ids_might_be_clust4)

DimPlot(NKclust2, 
        cells.highlight = cell_ids_might_be_clust1,
                pt.size = 2,
        sizes.highlight = 2)
DimPlot(NKclust2, 
        cells.highlight = cell_ids_might_be_clust3,
                pt.size = 2,
        sizes.highlight = 2)
DimPlot(NKclust2, 
        cells.highlight = cell_ids_might_be_clust4,
        pt.size = 2,
        sizes.highlight = 2)
# all wrongly predicted cells 
DimPlot(NKclust2, 
        cells.highlight = highlight,
                pt.size = 2,
        sizes.highlight = 2)
DimPlot(NKclust2, 
        cells.highlight = cell_ids_true_clust2,
                pt.size = 2,
        sizes.highlight = 2)




# ATRX only
ATRX_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXmut_BM_NK.Rds"))

ATRX_clust2 <- subset(ATRX_NK, idents = "hNK_Bm2")
hist(ATRX_clust2$scores.hNK_Bm4, plot = TRUE,breaks = 50)

hist(predicted_labels$scores,plot = TRUE,breaks = 50)

```


# ___Cluster markers in NB dataset on predicted NK sub-clusters___

```{r cluster markers}
# "healthy" fetal/toddler BM-NK cells
fetal_NK <- fetal_NK %>% ScaleData()


library(DElegate)

de_results <- DElegate::findDE(object = fetal_NK,
                               group_column = "predicted_NK_type")
marker_results <- DElegate::FindAllMarkers2(object = fetal_NK,
                                            group_column = "predicted_NK_type")

marker_results_top5 <- dplyr::filter(marker_results, feature_rank < 6)
marker_results_top5

marker_results %>%
    group_by(group1) %>%
    slice_max(n = 10, order_by = log_fc)

marker_results %>%
    group_by(group1) %>%
    top_n(n = 10, wt = log_fc) -> top10

DEG_tumor <- DoHeatmap(fetal_NK, 
                       features = top10$feature) + NoLegend()
DEG_tumor
```

# ___Exploration of age and sex dependence in the data___

```{r}
fetahu2023.metadata <- read.csv("/home/rstudio/mnt_nb_tumors/fetahu2023.scRNA/metadata/fetahu2023_scRNA_seq_metadata.csv", 
                                sep = "\t", header = TRUE)

metadata <- fetahu2023@meta.data
metadata$cell_ID <- row.names(metadata)


# filter by unique sample ID
library(dplyr) 
unique.metadata <- fetahu2023.metadata %>% 
  distinct(sample_ID,.keep_all = TRUE)

# plot age distribution per comparison group
ggplot(unique.metadata, aes(x = group, 
               y = age_at_Dx_years, 
               fill = group)) +
  geom_boxplot()

# input table
input <- table(unique.metadata$group, unique.metadata$sex)
input <- as.data.frame(input)
colnames(input) <- c("group","sex","value")


# plot sex distribution in the whole dataset
ggplot(input, aes(x=sex,y=value,fill=sex)) +
  geom_bar(stat = "identity") + 
  theme_minimal() + 
  scale_fill_manual(values=c("hotpink",
                             "darkblue")) +
  theme(text = element_text(size = 20, colour = "black"))

# plot sex distribution per comparison group
ggplot(input, aes(x=group,y=value,fill=sex)) +
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title="Number of male/female donors per group") + 
  theme_minimal() + 
  scale_fill_manual(values=c("hotpink",
                             "darkblue")) +
  theme(text = element_text(size = 20, colour = "black"))

```


```{r fig.height=7, fig.width=7}
do_FeaturePlot(fetahu2023_NK, 
               plot.title = "EOMES",
               plot.subtitle = "gene expression in NK cells",
               features = "EOMES",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 18
               )

do_FeaturePlot(fetahu2023_NK, 
               plot.title = "ID2",
               plot.subtitle = "gene expression in NK cells",
               features = "ID2",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 18
               )

do_FeaturePlot(fetahu2023_NK, 
               plot.title = "TBX21",
               plot.subtitle = "gene expression in NK cells",
               features = "TBX21",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 18
               )

do_FeaturePlot(fetahu2023_NK, 
               plot.title = "NFIL3",
               plot.subtitle = "gene expression in NK cells",
               features = "NFIL3",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 18
               )

do_FeaturePlot(fetahu2023_NK, 
               plot.title = "PLZF (ZBTB16)",
               plot.subtitle = "gene expression in NK cells",
               features = "ZBTB16",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 18
               )

```
```{r fig.height=5, fig.width=10}
NK_cytotoxic_markers <- list("canonical NK markers" =  c("EOMES","ID2","TBX21"),
                             "indirect killing" =  c("PRF1","GZMB"),
                             "direct killing" = c("TNFSF10","FASLG"),
                             "recruitment of immune cells" = c("IFNG","TNF"),
                              "NK activating cytokines" = c("IL2", "IL15", "IL12A", "IL12B","IL18"))

do_DotPlot(fetahu2023_NK, 
           features = NK_cytotoxic_markers, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(fetahu2023_NK, 
           features = NK_cytotoxic_markers, 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

# NK cell killing occurs either via secretion of 
# perforin-1 (PRF1 Gene) and granzyme B (GZMB gene) or 
# direct cell-to-cell contact via death receptors TRAIL (TNFRSF10B gene) and  
# FAS receptor (TNFSF6 gene). FAS ligand (FASLG gene); TRAIL ligand (TNFSF10 gene) 
# or mediate cell killing by recruiting other immune cells via cytokine and
# chemokine secretion, and induce cell death via Interferon-γ (IFNG gene) or  
# TNF secretion
```


# ___Desensitized/exhausted NK cells___

```{r}
# list of genes from NK Cell Exhaustion review paper from 2017
# DOI: 10.3389/fimmu.2017.00760
# 2B4 = CD244 gene
# NKG2A  = KLRC1 gene
# NKG2D = KLRK1 gene
# NRC proteins = NCOA1, NCOA2, NCOA3
# CD16 = FCGR3A Gene gene
# TIM3 = HAVCR2 gene

exhausted_NK_genes <- list(up = c("PDCD1", "HAVCR2", "KLRC1"),
                           down = c("KLRK1","FCGR3A","NCOA1", "NCOA2", 
                                    "NCOA3","CD226","CD244","IFNG")) 

do_DotPlot(fetahu2023_NK, 
           features = exhausted_NK_genes, 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(fetahu2023_NK, 
           features = exhausted_NK_genes, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

```


# ___CCC analysis___


The degree to which NK cells are active against tumour cells is highly dependent 
on the activation state of the NK cells, which is regulated in large part by 
cytokines derived from dendritic cells (DCs) and monocytes

## ___NK cells: L-R expression per cluster per group___

```{r fig.height=5, fig.width=15}
# based on The cancer–natural killer cell immunity cycle review paper
# high or low expression of  these genes is associated with a good prognosis 
# in a pan-cancer analysis

NK_receptors <- c("ADORA2A","CD2","CD226", "CD244", "TIGIT","TMIGD2", 
                  "NCR1","NCR2","NCR3","NT5E", "ITGAL", "ITGB2","PDCD1",
                  "KLRC1","KLRC2", "KLRC3", "KLRC4", "KLRD1", "KLRF1",
                  "KLRG1", "KLRG2","KLRK1", "KIR2DL1", "KIR2DL3", "KIR2DL4", 
                  "KIR2DS4","KIR3DL1", "KIR3DL2", "KIR3DL3")

NK_secreted_cytokines <- c("CCL5","FLT3LG", "XCL1", "XCL2")

NK_prognosis_markers <- list(NK_receptors = c("ADORA2A","CD2","CD226", "CD244", 
                                              "TIGIT","TMIGD2", "NCR1","NCR2",
                                              "NCR3","NT5E", "ITGAL", "ITGB2",
                                              "PDCD1","KLRC1","KLRC2", "KLRC3", 
                                              "KLRC4", "KLRD1","KLRF1","KLRG1", 
                                              "KLRG2","KLRK1","KIR2DL1","KIR2DL3", 
                                              "KIR2DL4", "KIR3DL1", "KIR3DL2", 
                                              "KIR3DL3","KIR2DS4"),
                             NK_secreted_cytokines = c("CCL5","FLT3LG", "XCL1", "XCL2"))

do_DotPlot(fetahu2023_NK, 
           features = NK_prognosis_markers, 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(fetahu2023_NK, 
           features = NK_prognosis_markers, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

```


## ___NK - M - DC axis___

```{r}
cell_types_of_interest_v2 <- c("NK (4)",
                               "pDC (12)",
                               "M (1)","M (2)", "M (10)","M (15)")

fetahu2023.NK.DC.M <- subset(fetahu2023,
                             idents = cell_types_of_interest_v2)

cell_types_of_interest_v3 <- c("M (10)")
fetahu2023.M10 <- subset(fetahu2023,
                         idents = cell_types_of_interest_v3)

do_DotPlot(fetahu2023.NK.DC.M, 
           features = NK_cytotoxic_markers, 
           group.by = "orig.ident", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(fetahu2023.M10, 
           features = NK_cytotoxic_markers, 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_ViolinPlot(fetahu2023.NK.DC.M, 
           features = c("IL2", "IL15", "IL12A", "IL12B","IL18"), 
           group.by = "orig.ident", 
           split.by = "group.y",
           plot_boxplot = FALSE,
           slot = "data")

# Chemokines secreated by NK cells and associated with 
# recruiting type 1 conventional dendritic cells

do_DotPlot(fetahu2023_NK, 
           features = c("CCL5","XCL1","FLT3LG"), 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(fetahu2023_NK, 
           features = c("CCL5","XCL1","FLT3LG"), 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 18,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

```

-------------------------------------------------------------------------------

# ___Session Info___

```{r}
sessionInfo()
```
