---
title: "NK project: Functional markers"
author: "Dr. Sara Wernig-Zorc"
date: "27.08.2024 - 30.08.2024"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/scRNA.NB.integration/metadata/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_utils.R")
```

# ___Input_data___

```{r}
ALL_NK <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_NK_cluster.Rds"))
CTRL_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_fetal_BM_NK.Rds"))
MNA_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_MNA_BM_NK.Rds"))
ATRX_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXmut_BM_NK.Rds"))
sporadic_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_sporadic_BM_NK.Rds"))
```

```{r}
ALL_NK <- ALL_NK %>% ScaleData() %>% NormalizeData()
CTRL_NK <- CTRL_NK %>% ScaleData() %>% NormalizeData()
MNA_NK <- MNA_NK %>% ScaleData() %>% NormalizeData()
ATRX_NK <- ATRX_NK %>% ScaleData() %>% NormalizeData()
sporadic_NK <- sporadic_NK %>% ScaleData() %>% NormalizeData()
```


# ___NK Cluster markers in the NB dataset___

Using predicted NK sub-clusters

```{r cluster markers}
de_results <- DElegate::findDE(object = ALL_NK,
                               group_column = "predicted_NK_type",
                               replicate_column = "patient")

marker_results <- DElegate::FindAllMarkers2(object = ALL_NK,
                                            group_column = "predicted_NK_type",
                                            replicate_column = "patient")

marker_results %>%
    group_by(group1) %>%
    top_n(n = 10, wt = log_fc) -> top10

# TOP 10 markers per cluster 
DEG_tumor <- DoHeatmap(ALL_NK, group.by = "predicted_NK_type",
                       features = top10$feature) + NoLegend()
DEG_tumor
```

# ___Exploration of age and sex dependence in the data___

```{r}
full_metadata <- read_excel(paste0(params$metadata,
  "2024-07-29_Wernig-Zorc_Fetahu_complete_single_cell_cohort_EB_FINAL.xlsx"), 
            col_types = c("text", "numeric", "text", "text", "text", "text", 
                          "text", "date", "date", "text", "text", "text", 
                          "text", "text", "numeric", "text", "text", "text", 
                          "text", "text", "text", "text", "text", "text", 
                          "text", "text", "text", "text", "text", "text", 
                          "text", "text", "text", "text", "text", "text"))

# filter by fetahu2023 dataset samples
unique.metadata <- full_metadata %>% filter(cohort == "fetahu2023")

# filter by unique sample ID
unique.metadata <- unique.metadata %>% 
  distinct(BM_ID,.keep_all = TRUE)

unique.metadata <- as.data.frame(unique.metadata)

# plot age distribution per comparison group
ggplot(unique.metadata, aes(x = genetic_subtype, 
               y = age_at_Dx_months, 
               fill = genetic_subtype)) +
  geom_boxplot()

# input table
input <- table(unique.metadata$genetic_subtype, 
               unique.metadata$sex,
               unique.metadata$stage)
input <- as.data.frame(input)
colnames(input) <- c("group","sex","stage","value")


# plot sex distribution in the whole dataset
ggplot(input, aes(x=sex,y=value,fill=sex)) +
  geom_bar(stat = "identity") + 
  theme_minimal() + 
  scale_fill_manual(values=c("hotpink",
                             "darkblue")) +
  theme(text = element_text(size = 20, colour = "black"))

# plot sex distribution per comparison group
ggplot(input, aes(x=group,y=value,fill=sex)) +
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title="Number of male/female donors per group") + 
  theme_minimal() + 
  scale_fill_manual(values=c("hotpink",
                             "darkblue")) +
  theme(text = element_text(size = 20, colour = "black"))

```

# ___NK functional markers___

NK cell killing occurs either via secretion of perforin-1 (PRF1 Gene) and 
granzyme B (GZMB gene) or direct cell-to-cell contact via death receptors 
TRAIL (TNFRSF10B gene) and FAS receptor (TNFSF6 gene). FAS ligand (FASLG gene);
TRAIL ligand (TNFSF10 gene) or mediate cell killing by recruiting other immune 
cells via cytokine and chemokine secretion, and induce cell death via 
Interferon-γ (IFNG gene) or TNF secretion

## ___Canonical NK markers___

### ___ALL___

#### ___Featureplot___

```{r fig.height=7, fig.width=7}
do_FeaturePlot(ALL_NK, 
               plot.title = "EOMES",
               plot.subtitle = "gene expression in NK cells",
               features = "EOMES",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 16
               )
do_FeaturePlot(ALL_NK, 
               plot.title = "ID2",
               plot.subtitle = "gene expression in NK cells",
               features = "ID2",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 16
               )
do_FeaturePlot(ALL_NK, 
               plot.title = "TBX21",
               plot.subtitle = "gene expression in NK cells",
               features = "TBX21",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 16
               )
do_FeaturePlot(ALL_NK, 
               plot.title = "NFIL3",
               plot.subtitle = "gene expression in NK cells",
               features = "NFIL3",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 16
               )
do_FeaturePlot(ALL_NK, 
               plot.title = "PLZF (ZBTB16)",
               plot.subtitle = "gene expression in NK cells",
               features = "ZBTB16",
               reduction = "UMAP",
               plot.axes = TRUE, 
               pt.size = 0.5,
               slot = "data",
               assay = "RNA",
               font.size = 16
               )
```

#### ___Dotplot___

```{r}
do_DotPlot(ALL_NK, 
           features = NK_functional_markers, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "ALL"
           )

do_DotPlot(ALL_NK, 
           features = NK_functional_markers, 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "ALL"
           )
```

### ___ATRX___

```{r}
do_DotPlot(ATRX_NK, 
           features = NK_functional_markers, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "ATRX"
           )
```

### ___MNA___

```{r}
do_DotPlot(MNA_NK, 
           features = NK_functional_markers, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "MNA"
           )
```

### ___Sporadic___

```{r}
do_DotPlot(sporadic_NK, 
           features = NK_functional_markers, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "Sporadic"
           )
```

### ___CTRL___

```{r}
do_DotPlot(CTRL_NK, 
           features = NK_functional_markers, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "CTRL"
           )
```

## ___NK TFs___

### ___ALL___

```{r fig.height=5, fig.width=8}
do_DotPlot(ALL_NK, 
           features = TFs, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "ALL",
           legend.title = "Scaled expression"
           )

do_DotPlot(ALL_NK, 
           features = TFs, 
           group.by = "group.y",
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "ALL",
           legend.title = "Scaled expression"
           )
```

### ___ATRX___

```{r fig.height=5, fig.width=8}
do_DotPlot(ATRX_NK,
           features = TFs, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "ATRX",
           legend.title = "Scaled expression"
           )
```

### ___MNA___

```{r fig.height=5, fig.width=8}
do_DotPlot(MNA_NK,
           features = TFs, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "MNA",
           legend.title = "Scaled expression"
           )
```

### ___Sporadic___

```{r fig.height=5, fig.width=8}
do_DotPlot(sporadic_NK,
           features = TFs, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "sporadic",
           legend.title = "Scaled expression"
           )
```

### ___CTRL___

```{r fig.height=5, fig.width=8}
do_DotPlot(CTRL_NK,
           features = TFs, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "CTRL",
           legend.title = "Scaled expression"
           )
```

## ___NK cytokines___

### ___ALL___

```{r fig.height=4, fig.width=6}
do_DotPlot(ALL_NK, 
           features = cytokines, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "ALL cytokines",
           legend.title = "Scaled expression",
           cluster = TRUE
           )

do_DotPlot(ALL_NK, 
           features = cytokines, 
           group.by = "group.y",
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "ALL cytokines",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___ATRX___

```{r}
do_DotPlot(ATRX_NK, 
           features = cytokines, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "Cytokines",
           plot.subtitle = "ATRX",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___MNA___

```{r}
do_DotPlot(MNA_NK, 
           features = cytokines, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "Cytokines",
           plot.subtitle = "MNA",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___Sporadic___

```{r}
do_DotPlot(sporadic_NK, 
           features = cytokines, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "Cytokines",
           plot.subtitle = "Sporadic",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___CTRL___

```{r}
do_DotPlot(CTRL_NK, 
           features = cytokines, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "Cytokines",
           plot.subtitle = "CTRL",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

## ___NK maturation genes___

```{r}
do_DotPlot(ALL_NK, 
           features = NK_maturation_genes, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "NK maturation genes",
           plot.subtitle = "ALL",
           legend.title = "Scaled expression",
           cluster = TRUE
           )

do_DotPlot(ALL_NK, 
           features = NK_maturation_genes, 
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "NK maturation genes",
           plot.subtitle = "ALL",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___ATRX___

```{r}
do_DotPlot(ATRX_NK, 
           features = NK_maturation_genes, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "NK maturation genes",
           plot.subtitle = "ATRX",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___Sporadic___

```{r}
do_DotPlot(sporadic_NK, 
           features = NK_maturation_genes, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "NK maturation genes",
           plot.subtitle = "sporadic",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___MNA___

```{r}
do_DotPlot(MNA_NK, 
           features = NK_maturation_genes, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "NK maturation genes",
           plot.subtitle = "MNA",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___CTRL___

```{r}
do_DotPlot(CTRL_NK, 
           features = NK_maturation_genes, 
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "NK maturation genes",
           plot.subtitle = "CTRL",
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

## ___NK subtype genes___

hNK_Bm2 subset was characterized by higher levels of CD122, CD27, and CD69, 
whereas hNK_Bm3 cells had higher levels of CD127.

Bone marrow NK cells revealed that the expression of CD160 and CD52 (CAMPATH-1) 
on the surface of CD56bright cells was mutually exclusive, further discriminating 
between hNK_Bm2 and hNK_Bm3, respectively.

IL2RB gene = CD122 protein
NCAM1 gene = CD56 protein
IL7R gene = CD127 protein
FCGR3A gene = CD16 protein

### ___Dotplot___

#### ___ALL___

```{r fig.height=5, fig.width=5}
do_DotPlot(ALL_NK,
           features = NK_functional_markers$NK_subtype_markers,
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           legend.title = "Scaled expression",
           cluster = TRUE
           )

do_DotPlot(ALL_NK,
           features = NK_functional_markers$NK_subtype_markers,
           group.by = "group.y", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

#### ___ATRX___

```{r fig.height=5, fig.width=5}
do_DotPlot(ATRX_NK,
           features = NK_functional_markers$NK_subtype_markers,
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

#### ___MNA___

```{r fig.height=5, fig.width=5}
do_DotPlot(MNA_NK,
           features = NK_functional_markers$NK_subtype_markers,
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

#### ___Sporadic___

```{r fig.height=5, fig.width=5}
do_DotPlot(sporadic_NK,
           features = NK_functional_markers$NK_subtype_markers,
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

#### ___CTRL___

```{r fig.height=5, fig.width=5}
do_DotPlot(CTRL_NK,
           features = NK_functional_markers$NK_subtype_markers,
           group.by = "predicted_NK_type", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           legend.title = "Scaled expression",
           cluster = TRUE
           )
```

### ___Violinplot___

#### ___ALL___

```{r fig.height=12, fig.width=10}
VlnPlot(ALL_NK,
        features = NK_functional_markers$NK_subtype_markers,
        log = FALSE,
        layer = "data",
        slot = "data")
```
#### ___ATRX___

```{r fig.height=12, fig.width=10}
VlnPlot(ATRX_NK,
        features = NK_functional_markers$NK_subtype_markers,
        group.by = "predicted_NK_type",
        log = FALSE,
        layer = "data",
        slot = "data")
```

#### ___MNA___

```{r fig.height=12, fig.width=10}
VlnPlot(MNA_NK,
        features = NK_functional_markers$NK_subtype_markers,
        group.by = "predicted_NK_type",
        log = FALSE,
        layer = "data",
        slot = "data")
```

#### ___CTRL___

```{r fig.height=12, fig.width=10}
VlnPlot(CTRL_NK,
        features = NK_functional_markers$NK_subtype_markers,
        group.by = "predicted_NK_type",
        log = FALSE,
        layer = "data",
        slot = "data")
```

#### ___Sporadic___

```{r fig.height=12, fig.width=10}
VlnPlot(sporadic_NK,
        features = NK_functional_markers$NK_subtype_markers,
        group.by = "predicted_NK_type",
        log = FALSE,
        layer = "data",
        slot = "data")
```

# ___Session Info___

```{r}
sessionInfo()
```
