---
title: "NK project: Trajectory analysis"
author: "Dr. Sara Wernig-Zorc"
date: "27.9.2024"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/scRNA.NB.integration/metadata/"
---

DESCRIPTION

The idea here is to see the the differentiation trajectory of NK cells and if 
<br>
the trajectory differs between healthy and tumor

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_utils.R")
```

 ___Input_data___

```{r}
adult_BM_NK <- readRDS(file = params$input_BM1)

scea <- as.SingleCellExperiment(adult_BM_NK)
colLabels(scea) <- adult_BM_NK$cluster2
```

```{r}
ALL_NK <-  readRDS(file = paste0(params$out,"Rds/fetahu2023_NK_subclusters_from_Crinier2021_de.n=350.Rds"))

sce <- as.SingleCellExperiment(ALL_NK)
colLabels(sce) <- ALL_NK$predicted_NK_type_de.n350
```

```{r}
Idents(ALL_NK) <- ALL_NK$group.y

ATRX_NK <- subset(ALL_NK, idents = "ATRXdel") 
ATRX <- as.SingleCellExperiment(ATRX_NK)
colLabels(ATRX) <- ATRX_NK$predicted_NK_type_de.n350

MNA_NK <- subset(ALL_NK, idents = "MNA") 
MNA <- as.SingleCellExperiment(MNA_NK)
colLabels(MNA) <- MNA_NK$predicted_NK_type_de.n350

sporadic_NK <- subset(ALL_NK, idents = "sporadic") 
sporadic <- as.SingleCellExperiment(sporadic_NK)
colLabels(sporadic) <- sporadic_NK$predicted_NK_type_de.n350

control_NK <- subset(ALL_NK, idents = "control") 
control <- as.SingleCellExperiment(control_NK)
colLabels(control) <- control_NK$predicted_NK_type_de.n350

Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350
```

# ___ALL NK___

## ___Slingshot___

```{r PCA}
# Extract PCA embeddings from Seurat object
pca_embeddings <- Embeddings(ALL_NK, "PCA")

# Extract cluster identities
Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350
cluster_labels <- Idents(ALL_NK)

# Perform trajectory inference 
sds <- slingshot(sce,
                 clusterLabels = cluster_labels, 
                 reducedDim = "PCA")

# plot the PCA from Seurat
DimPlot(ALL_NK,reduction = "PCA",dims = 1:2,pt.size = 1)
 
# Plot Slingshot results
plot(pca_embeddings[, 1], pca_embeddings[, 2], col = cluster_labels, pch = 10,
     xlab = "PC1", ylab = "PC2", main = "PCA with Slingshot Trajectories")

# Overlay the Slingshot trajectories
lines(SlingshotDataSet(sds), lwd = 4, col = 'black')
```

```{r UMAP}
# Extract PCA embeddings from Seurat object
pca_embeddings <- Embeddings(ALL_NK, "UMAP")

# Extract cluster identities
Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350
cluster_labels <- Idents(ALL_NK)

# Perform trajectory inference 
sds <- slingshot(pca_embeddings, 
                 clusterLabels = cluster_labels, 
                 reducedDim = "UMAP")

# plot the PCA from Seurat
DimPlot(ALL_NK,reduction = "UMAP",dims = 1:2,pt.size = 1)
 
# Plot Slingshot results
plot(pca_embeddings[, 1], pca_embeddings[, 2], col = cluster_labels, pch = 10,
     xlab = "PC2", ylab = "PC3", main = "UMAP with Slingshot Trajectories")

# Overlay the Slingshot trajectories
lines(SlingshotDataSet(sds), lwd = 4, col = 'black')

```

## ___TSCAN___

TSCAN uses the clustering to summarize the data into a smaller set of discrete units, 
computes cluster centroids by averaging the coordinates of its member cells, 
and then forms the minimum spanning tree (MST) across those centroids. 
The MST is simply an undirected acyclic graph that passes through each centroid 
exactly once and is thus the most parsimonious structure that captures the 
transitions between clusters. 

```{r}
library(TSCAN)
library(scater)

# 1.Cluster-based minimum spanning tree
by.cluster <- aggregateAcrossCells(sce, ids = sce$predicted_NK_type_de.n350)
centroids <- reducedDim(by.cluster, "PCA")

# Set clusters=NULL as we have already aggregated above.
mst <- createClusterMST(centroids, clusters=NULL)
mst

line.data <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="TSNE")

plotTSNE(sce, colour_by="predicted_NK_type_de.n350") + 
    geom_line(data=line.data, mapping=aes(x=tSNE_1, y=tSNE_2, group=edge))

#We obtain a pseudotime ordering by projecting the cells onto the MST with mapCellsToEdges(). More specifically, we move each cell onto the closest edge of the MST; the pseudotime is then calculated as the distance along the MST to this new position from a “root node” with orderCells(). For our purposes, we will arbitrarily pick one of the endpoint nodes as the root, though a more careful choice based on the biological annotation of each node may yield more relevant orderings (e.g., picking a node corresponding to a more pluripotent state).

map.tscan <- mapCellsToEdges(sce, mst=mst, use.dimred="PCA")
tscan.pseudo <- orderCells(map.tscan, mst)
head(tscan.pseudo)

common.pseudo <- averagePseudotime(tscan.pseudo) 
plotTSNE(sce, colour_by=I(common.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=line.data, mapping=aes(x=tSNE_1, y=tSNE_2, group=edge))
```

```{r eval=FALSE, include=FALSE}
#same as above but in one command
pseudo.all <- quickPseudotime(sce, use.dimred="PCA")
```

Alternative approach: In this mode, the MST focuses on the connectivity between clusters, which can be different from the shortest distance between centroids

```{r}
pseudo.mnn <- quickPseudotime(sce, use.dimred="PCA", dist.method=NULL)
mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

plotTSNE(sce, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))

plotUMAP(sce, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, mapping=aes(x=UMAP_1, y=UMAP_2,
                                                          group=edge))

plotPCA(sce, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, mapping=aes(x=PC_1, y=PC_2,
                                                          group=edge))
```



# ___Adolescent controls [healthy]___

```{r}
pseudo.mnn <- quickPseudotime(control, use.dimred="PCA", dist.method=NULL)
mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

plotTSNE(control, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))

plotUMAP(control, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, mapping=aes(x=UMAP_1, y=UMAP_2,
                                                          group=edge))

plotPCA(control, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, mapping=aes(x=PC_1, y=PC_2,
                                                          group=edge))
```

# ___ATRXm___

```{r}
pseudo.mnn <- quickPseudotime(ATRX, use.dimred="PCA", dist.method=NULL)
mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

plotTSNE(ATRX, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))

plotUMAP(ATRX, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, mapping=aes(x=UMAP_1, y=UMAP_2,
                                                          group=edge))

plotPCA(ATRX, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, mapping=aes(x=PC_1, y=PC_2,
                                                          group=edge))
```

# ___MNA___

```{r}
pseudo.mnn <- quickPseudotime(MNA, use.dimred="PCA", dist.method=NULL)
mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

plotTSNE(MNA, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))

plotUMAP(MNA, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, mapping=aes(x=UMAP_1, y=UMAP_2,
                                                          group=edge))

plotPCA(MNA, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, mapping=aes(x=PC_1, y=PC_2,
                                                          group=edge))
```

# ___Sporadic___

```{r}
pseudo.mnn <- quickPseudotime(sporadic, use.dimred="PCA", dist.method=NULL)
mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

plotTSNE(sporadic, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))

plotUMAP(sporadic, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, mapping=aes(x=UMAP_1, y=UMAP_2,
                                                          group=edge))

plotPCA(sporadic, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, mapping=aes(x=PC_1, y=PC_2,
                                                          group=edge))
```



# ___Adult controls [healthy]___

Crinier at al., 2021 adult BM-derived-NK cells from 8 healthy donors

```{r}
library(TSCAN)
library(scater)

by.cluster <- aggregateAcrossCells(scea, ids = scea$cluster2)
centroids <- reducedDim(by.cluster, "PCA")

mst <- createClusterMST(centroids, clusters=NULL)
mst

line.data <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="TSNE")

plotTSNE(scea, colour_by="cluster2") + 
    geom_line(data=line.data, mapping=aes(x=tSNE_1, y=tSNE_2, group=edge))

map.tscan <- mapCellsToEdges(scea, mst=mst, use.dimred="PCA")
tscan.pseudo <- orderCells(map.tscan, mst)
head(tscan.pseudo)

common.pseudo <- averagePseudotime(tscan.pseudo) 
plotTSNE(scea, colour_by=I(common.pseudo), 
        text_by="cluster2", text_colour="red") +
    geom_line(data=line.data, mapping=aes(x=tSNE_1, y=tSNE_2, group=edge))
```

# ___sessionInfo___

```{r}
sessionInfo()
```
