---
title: "NK project: Data integration"
author: "Dr. Sara Wernig-Zorc"
date: "06.09.2024 - "
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/scRNA.NB.integration/metadata/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```

```{r load project specific R packages and color code}
# install.packages(c("BPCells", "presto", "glmGamPoi"))
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_utils.R")
```

```{r eval=FALSE, include=FALSE}
remotes::install_github(repo = "satijalab/seurat-object", ref = "v5.0.2")
remotes::install_github(repo = "satijalab/seurat", ref = "v5.1.0")

# Install from source code
#path_to_file <- "/home/rstudio/mnt_tools/Seurat_5.1.0.tar.g"
#install.packages(path_to_file, repos = NULL, type="source")

devtools::install_github("bnprks/BPCells/r")
devtools::install_github("saeyslab/nichenetr")
devtools::install_github("saeyslab/multinichenetr")

BiocManager::install('multtest')
install.packages('metap')
```

# ___Input_data___

```{r}
ALL_NK <-  readRDS(file = paste0(params$out,"Rds/fetahu2023_NK_subclusters_from_Crinier2021_de.n=350.Rds"))
crinier2021 <- readRDS(file = paste0(params$input_BM1))

crinier2021$cohort <- "adult"
ALL_NK$cohort <- "child"

# sample/batch information with the same variable name
ALL_NK$batch <- ALL_NK$patients 

# convert a v4 assay to a v5 assay
ALL_NK[["RNA"]] <- as(object = ALL_NK[["RNA"]], 
                      Class = "Assay5")

crinier2021[["RNA"]] <- as(object = crinier2021[["RNA"]], 
                           Class = "Assay5")
```

# ___NEW clustering___

```{r}
NK.combined <- merge(x = ALL_NK, 
                     y = crinier2021, 
                     add.cell.ids = c("child", "adult"), 
                     project = "BM_NK_cells")
NK.combined

adult <- is.na(NK.combined$group.x)
NK.combined$group.y[adult] <- "V"

adult <- is.na(NK.combined$group.y)
NK.combined$group.y[adult] <- "adult_control"

# split into layers
# NK.combined[["RNA"]] <- split(NK.combined[["RNA"]], f = as.factor(NK.combined$cohort))
```

```{r}
# run standard analysis on non-integrated, merged data
NK.combined <- NormalizeData(NK.combined)
NK.combined <- FindVariableFeatures(NK.combined)
NK.combined <- ScaleData(NK.combined)
NK.combined <- RunPCA(NK.combined)
```

```{r fig.height=5, fig.width=15}
NK.combined <- FindNeighbors(NK.combined, 
                             dims = 1:30, 
                             reduction = "pca")
NK.combined <- FindClusters(NK.combined, 
                            resolution = 0.2, 
                            cluster.name = "unintegrated_clusters")
NK.combined <- RunUMAP(NK.combined, 
                       dims = 1:30, 
                       reduction = "pca", 
                       reduction.name = "umap.unintegrated")
DimPlot(NK.combined, reduction = "umap.unintegrated", 
        group.by = c("cohort", "seurat_clusters"))
```

```{r}
# Data integration
NK.combined <- IntegrateLayers(object = NK.combined, 
                        method = CCAIntegration, 
                        orig.reduction = "pca", 
                        new.reduction = "integrated.cca",
                        verbose = FALSE)

# re-join layers after integration
NK.combined[["RNA"]] <- JoinLayers(NK.combined[["RNA"]])

NK.combined <- FindNeighbors(NK.combined, 
                             reduction = "integrated.cca", 
                             dims = 1:30)
NK.combined <- FindClusters(NK.combined, 
                            resolution = 0.1,
                            cluster.name = "integrated_clusters")
# tested 0.2 and 2
```

```{r}
NK.combined <- RunUMAP(NK.combined, 
                dims = 1:10, 
                reduction = "integrated.cca",
                reduction.name = "umap.integrated")
# Visualization
DimPlot(NK.combined, 
        reduction = "umap.integrated", 
        group.by = c("cohort", "integrated_clusters"))

DimPlot(NK.combined, 
        reduction = "umap.integrated", 
        split.by = "cohort")

DimPlot(NK.combined, 
        reduction = "umap.integrated", 
        group.by = "predicted_NK_type_de.n350",
        split.by = "cohort")

DimPlot(NK.combined, 
        reduction = "umap.integrated", 
        group.by = "cluster2",
        split.by = "cohort")
```

```{r fig.height=12, fig.width=7}
plot <- DimPlot(NK.combined, 
        reduction = "umap.integrated", 
        group.by = c("integrated_clusters", 
                     "cluster2",
                     "predicted_NK_type_de.n350"),
        split.by = "cohort")

plot

png(paste0(params$out,"figures/UMAP_BM-NK_adult+child_integrated_4clust_v2.png"),
    width = 700, height = 1100)
plot
dev.off()
```

```{r NK Cell subset compositon}
cluster_anno <- c("0","1","2","3")

filter <- is.na(NK.combined$group.x)
NK.combined$group.x[filter] <- "V"

filter <- is.na(NK.combined$group.y)
NK.combined$group.x[filter] <- "adult"

factor(NK.combined$group.x)
factor(NK.combined$group.y)

DimPlot(NK.combined,
        group.by = "integrated_clusters",
        split.by = "group.y")

percent_NK_group <- table(NK.combined$integrated_clusters,
                          NK.combined$group.x)
percent_NK_group


percent_NK_sample <- table(NK.combined$integrated_clusters,
                          NK.combined$batch)

barplot(percent_NK_group,
        col = COLOR_CODE_CLUST_v2, 
        border = "white", 
        xlab = "sample groups",
        legend.text = cluster_anno)

# Transform this data in %
data_percentage <- apply(percent_NK_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})
 
data_percentage_sample <- apply(percent_NK_sample, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

# Make a stacked barplot in %
barplot(data_percentage, 
        col = COLOR_CODE_CLUST_v2, 
        border = "white", 
        xlab = "NB genetic group",
        legend.text = cluster_anno)

barplot(data_percentage_sample, 
        col = COLOR_CODE_CLUST_v2, 
        border = "white", 
        xlab = "NB genetic group",
        legend.text = cluster_anno)

table_NK_per_sample <- as.data.frame(percent_NK_sample)
table_NK_per_sample$group <- gsub("_[0-9][0-9]", "", table_NK_per_sample$Var2)
table_NK_per_sample$group <- gsub("*[0-9]*", "", table_NK_per_sample$group)
table_NK_per_sample$group <- gsub("Moelle ", "adult", table_NK_per_sample$group)
table_NK_per_sample$group <- gsub("MoBx*", "", table_NK_per_sample$group)
table_NK_per_sample$group <- gsub("F", "", table_NK_per_sample$group)
table_NK_per_sample

# CLUSTER 1
hNK_Bm1 <- filter(table_NK_per_sample, Var1 == "0")

# Perform ANOVA testing
hNK_Bm1_anova <- aov(Freq ~ group, data = hNK_Bm1)
summary(hNK_Bm1_anova)

# Extract p-value
p_value <- summary(hNK_Bm1_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm1, 
                                      group %in% c("ATRXmut", "sporadic")))$p.value
p_value2 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm1, 
                                      group %in% c("ATRXmut", "ctrl")))$p.value
p_value3 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm1, 
                                      group %in% c("ATRXmut", "MYCNA")))$p.value
p_value4 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm1, 
                                      group %in% c("ATRXmut", "adult")))$p.value

# Create boxplot and add p-value annotation
boxplot1 <- hNK_Bm1 %>%
  ggplot( aes(x = group, y = Freq, fill = group)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha = 0.6) +
    theme(legend.position="none",
          axis.text = element_text(size = 18),
          axis.title.y = element_text(size = 18),
          plot.title = element_text(size = 18)) +
    ggtitle("NK cluster per NB genetic group: cluster 0") +
    xlab("") + ylab("Number of cells per group") +
  geom_signif(comparisons = list(c("ATRXmut", "ctrl"),
                                 c("ATRXmut", "MYCNA"),
                                 c("ATRXmut", "sporadic"),
                                 c("ATRXmut","adult")),
              map_signif_level = FALSE, 
              y_position = c(max(hNK_Bm1$Freq) + 50, 
                             max(hNK_Bm1$Freq) + 65, 
                             max(hNK_Bm1$Freq) + 80))


# CLUSTER 2
hNK_Bm2 <- filter(table_NK_per_sample, Var1 == "1")

# Perform ANOVA
hNK_Bm2_anova <- aov(Freq ~ group, data = hNK_Bm2)
summary(hNK_Bm2_anova)

# Extract p-value
p_value <- summary(hNK_Bm2_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm2, 
                                      group %in% c("ATRXmut", "sporadic")))$p.value
p_value2 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm2, 
                                      group %in% c("ATRXmut", "ctrl")))$p.value
p_value3 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm2, 
                                      group %in% c("ATRXmut", "MYCNA")))$p.value
p_value4 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm2, 
                                      group %in% c("ATRXmut", "adult")))$p.value


# plot boxplot
boxplot2 <- hNK_Bm2 %>%
  ggplot( aes(x = group, y = Freq, fill = group)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha = 0.6) +
    theme(legend.position="none",
          axis.text = element_text(size = 18),
          axis.title.y = element_text(size = 18),
          plot.title = element_text(size = 18)) +
    ggtitle("NK cluster per NB genetic group: hNK_Bm2") +
    xlab("") + ylab("Number of cells per group") +
  geom_signif(comparisons = list(c("ATRXmut", "ctrl"),
                                 c("ATRXmut", "MYCNA"),
                                 c("ATRXmut", "sporadic"),
                                 c("ATRXmut","adult")),
              map_signif_level = FALSE, 
              y_position = c(max(hNK_Bm2$Freq) + 50, 
                             max(hNK_Bm2$Freq) + 80, 
                             max(hNK_Bm2$Freq) + 110)) 


# CLUSTER 3
hNK_Bm3 <- filter(table_NK_per_sample, Var1 == "2")

# Perform ANOVA
hNK_Bm3_anova <- aov(Freq ~ group, data = hNK_Bm3)
summary(hNK_Bm3_anova)

# Extract p-value
p_value <- summary(hNK_Bm3_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm3, 
                                      group %in% c("ATRXmut", "sporadic")))$p.value
p_value2 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm3, 
                                      group %in% c("ATRXmut", "ctrl")))$p.value
p_value3 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm3, 
                                      group %in% c("ATRXmut", "MYCNA")))$p.value
p_value4 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm3, 
                                      group %in% c("ATRXmut", "adult")))$p.value


# plot boxplot
boxplot3 <- hNK_Bm3 %>%
  ggplot( aes(x = group , y = Freq, fill = group)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha = 0.6) +
    theme(legend.position="none",
          axis.text = element_text(size = 18),
          axis.title.y = element_text(size = 18),
          plot.title = element_text(size = 18)) +
    ggtitle("NK cluster per NB genetic group: hNK_Bm3") +
    xlab("") + ylab("Number of cells per group")  +
  geom_signif(comparisons = list(c("ATRXmut", "ctrl"),
                                 c("ATRXmut", "MYCNA"),
                                 c("ATRXmut", "sporadic"),
                                 c("ATRXmut","adult")),
              map_signif_level = FALSE, 
              y_position = c(max(hNK_Bm3$Freq) + 50, 
                             max(hNK_Bm3$Freq) + 80, 
                             max(hNK_Bm3$Freq) + 110))

# CLUSTER 4
hNK_Bm4 <- filter(table_NK_per_sample, Var1 == "3")

# Perform ANOVA
hNK_Bm4_anova <- aov(Freq ~ group, data = hNK_Bm4)
summary(hNK_Bm4_anova)

# Extract p-value
p_value <- summary(hNK_Bm4_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm4, 
                                      group %in% c("ATRXmut", "sporadic")))$p.value
p_value2 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm4, 
                                      group %in% c("ATRXmut", "ctrl")))$p.value
p_value3 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm4, 
                                      group %in% c("ATRXmut", "MYCNA")))$p.value
p_value4 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm4, 
                                      group %in% c("ATRXmut", "adult")))$p.value

# plot boxplot
boxplot4 <- hNK_Bm4 %>%
  ggplot( aes(x = group , y = Freq, fill = group)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha = 0.6) +
    theme(legend.position="none",
          axis.text = element_text(size = 18),
          axis.title.y = element_text(size = 18),
          plot.title = element_text(size = 18)) +
    ggtitle("NK cluster per NB genetic group: hNK_Bm4") +
    xlab("") + ylab("Number of cells per group")  +
  geom_signif(comparisons = list(c("ATRXmut", "ctrl"),
                                 c("ATRXmut", "MYCNA"),
                                 c("ATRXmut", "sporadic"),
                                 c("ATRXmut","adult")),
              map_signif_level = FALSE,
              y_position = c(max(hNK_Bm4$Freq) + 5, 
                             max(hNK_Bm4$Freq) + 8, 
                             max(hNK_Bm4$Freq) + 11)) 

# adding ANOVA test p-values
boxplot1
boxplot2
boxplot3
boxplot4
```

# ___DA testing___

```{r}
NK.combined@meta.data$batch <- factor(NK.combined@meta.data$batch)
NK.combined@meta.data$integrated_clusters <- factor(NK.combined@meta.data$integrated_clusters)
```

If comparing more than two sample / groups, use DA testing with contrasts
MiloR doesn't use the "original" clusters but rather creates neighborhoods
based on similar logFC (note the fold change here is the number of cells
per cell type)

```{r miloR}
NK.combined <- readRDS(file = paste0(params$out,"Rds/BM-NK_adult+child_integrated_4clust_v2.Rds"))

library("statmod")
library("miloR")
library("SingleCellExperiment")
library("scater")
library("scran")
library("dplyr")
library("patchwork")

#2. If comparing more than two sample / groups, use DA testing with contrasts

# For differential abundance analysis on graph neighborhoods we first construct a Milo object. 
# This extends the SingleCellExperiment class to store information about neighbourhoods on the KNN graph.

rna_milo <- Milo(as.SingleCellExperiment(NK.combined))
rna_milo

reducedDim(rna_milo, "UMAP") <- reducedDim(as.SingleCellExperiment(NK.combined), 
                                           "UMAP.INTEGRATED")
plotUMAP(rna_milo, 
         colour_by="group.y") + 
  plotUMAP(rna_milo, 
           colour_by="cohort")

# Define cell neighbourhoods
rna_milo <- buildGraph(rna_milo, 
                       k = 10, 
                       d = 30)

## Constructing kNN graph with k = 10
rna_milo <- makeNhoods(rna_milo, 
                       prop = 0.1, 
                       k = 10, 
                       d = 30, 
                       refined = TRUE, 
                       refinement_scheme = "graph") 
# make nhoods using graph-only as this is faster

# metadata
colData(rna_milo)$patients <- paste(colData(rna_milo)$group.x, 
                                    colData(rna_milo)$cohort, sep="_")

# Counting cells in neighbourhoods
rna_milo <- countCells(rna_milo, meta.data = data.frame(colData(rna_milo)), 
                       samples = "batch") # make the nhood X sample counts matrix


# Defining experimental design
rna_design <- data.frame(colData(rna_milo))[,c("batch", "group.x")]
rna_design <- distinct(rna_design)
rownames(rna_design) <- rna_design$batch
rna_design

# Computing neighborhood connectivity
rna_milo <- calcNhoodDistance(rna_milo, 
                              d = 30, 
                              reduced.dim = "PCA")

# DA testing
da_results <- testNhoods(rna_milo, 
                         design = ~ 0 + group.x,
                         design.df = rna_design, 
                         reduced.dim = "PCA")
head(da_results)

da_results %>%
  arrange(SpatialFDR) %>%
  head() 

rna_milo <- buildNhoodGraph(rna_milo)


# QUALITY CHECK OF THE DIFFERENTIAL ABUNDANCE TESTING
# Plot single-cell UMAP
umap_pl <- plotReducedDim(rna_milo, 
                          dimred = "UMAP", 
                          colour_by = "group.x", 
                          text_by = "integrated_clusters", 
                          text_size = 3, point_size = 0.5,) +
  guides(fill = "none")

rna_milo <- buildNhoodGraph(rna_milo)

# Plot neighbourhood graph
nh_graph_pl <- plotNhoodGraphDA(rna_milo, 
                                da_results, 
                                layout = "UMAP", 
                                alpha = 0.05) 

umap_pl + nh_graph_pl +
  plot_layout(guides = "collect")

# Differential abundance testing with contrasts
rna_design <- data.frame(colData(rna_milo))[,c("batch", "group.x")]
rna_design <- distinct(rna_design)
rownames(rna_design) <- rna_design$batch

# Reorder rownames to match columns of nhoodCounts(milo)
rna_design <- rna_design[colnames(nhoodCounts(rna_milo)), , drop=FALSE]
table(rna_design$group.x)
```

## ___ATRX mutated vs. wild-type___

ATRXwt = NB MYCN amplified and sporadic samples

```{r eval=FALSE, include=FALSE}
# A. ATRXmut vs. ATRXwt ( MYCNA and sporadic )

model <- model.matrix(~ 0 + group.x, data = rna_design)
ave.contrast <- c("group.xIII - (group.xII + group.xIV) / 2")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXmutvsAll <- testNhoods(rna_milo, design = ~ 0 + group.x, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXmutvsAll$SpatialFDR < 0.1)

# annotate hood with original cell type annotation
da_results_ATRXmutvsAll <- annotateNhoods(rna_milo, 
                                          da_results_ATRXmutvsAll, 
                                          coldata_col = "integrated_clusters")

# remove hoods that contained mixed cell type populations
da_results_ATRXmutvsAll$integrated_clusters <- 
  ifelse(da_results_ATRXmutvsAll$integrated_clusters_fraction < 0.7, 
         "Mixed", da_results_ATRXmutvsAll$integrated_clusters)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXmutvsAll %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

plotDAbeeswarm(da_results_ATRXmutvsAll, 
               group.by = "integrated_clusters")
```

## ___ATRXm vs. healthy child___

```{r eval=FALSE, include=FALSE}
# B. ATRXmut vs. healthy child (control)

model <- model.matrix(~ 0 + group.x, data = rna_design)
ave.contrast <- c("(group.xIII - group.xI) / 3")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXmutvsAll <- testNhoods(rna_milo, design = ~ 0 + group.x, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXmutvsCTRL$SpatialFDR < 0.1)

# annotate hood with original cell type annotation
da_results_ATRXmutvsCTRL <- annotateNhoods(rna_milo, da_results_ATRXmutvsCTRL, 
                             coldata_col = "integrated_clusters")

# remove hoods that contained mixed cell type populations
da_results_ATRXmutvsCTRL$integrated_clusters <- 
  ifelse(da_results_ATRXmutvsCTRL$integrated_clusters_fraction < 0.7, 
         "Mixed", da_results_ATRXmutvsCTRL$integrated_clusters)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXmutvsCTRL %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

plotDAbeeswarm(da_results_ATRXmutvsCTRL, group.by = "integrated_clusters")
```

## ___heathy child vs. adult___

```{r eval=FALSE, include=FALSE}
# C. Child vs. Adult (both healthy)

model <- model.matrix(~ 0 + group.x, data = rna_design)
ave.contrast <- c("(group.xI - group.xV) / 3")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ChildvsAdult <- testNhoods(rna_milo, design = ~ 0 + group.x, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ChildvsAdult$SpatialFDR < 0.1)

# annotate hood with original cell type annotation
da_results_ChildvsAdult <- annotateNhoods(rna_milo, da_results_ChildvsAdult, 
                             coldata_col = "integrated_clusters")

# remove hoods that contained mixed cell type populations
da_results_ChildvsAdult$integrated_clusters <- 
  ifelse(da_results_ChildvsAdult$integrated_clusters_fraction < 0.7, 
         "Mixed", da_results_ChildvsAdult$integrated_clusters)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ChildvsAdult %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

plotDAbeeswarm(da_results_ChildvsAdult, group.by = "integrated_clusters")
```


# ___Identify conserved cell type markers___

## ___Adult vs. Child___

```{r}
Idents(NK.combined) <- "integrated_clusters"
cluster_markers_0 <- FindConservedMarkers(NK.combined, 
                                        ident.1 = "0",
                                        grouping.var = "cohort", 
                                        verbose = FALSE)
head(cluster_markers_0)

cluster_markers_1 <- FindConservedMarkers(NK.combined, 
                                        ident.1 = "1",
                                        grouping.var = "cohort", 
                                        verbose = FALSE)
head(cluster_markers_1)

cluster_markers_2 <- FindConservedMarkers(NK.combined, 
                                        ident.1 = "2",
                                        grouping.var = "cohort", 
                                        verbose = FALSE)
head(cluster_markers_2)

cluster_markers_3 <- FindConservedMarkers(NK.combined, 
                                        ident.1 = "3",
                                        grouping.var = "cohort", 
                                        verbose = FALSE)
head(cluster_markers_3)
```

## ___healthy vs. tumor___

```{r}
Idents(NK.combined) <- "integrated_clusters"
cluster_markers_0_perGroup <- FindConservedMarkers(NK.combined, 
                                        ident.1 = "0",
                                        grouping.var = "group.x", 
                                        verbose = FALSE)
head(cluster_markers_0_perGroup)

cluster_markers_1_perGroup <- FindConservedMarkers(NK.combined, 
                                        ident.1 = "1",
                                        grouping.var = "group.x", 
                                        verbose = FALSE)
head(cluster_markers_1_perGroup)

cluster_markers_2_perGroup <- FindConservedMarkers(NK.combined, 
                                        ident.1 = "2",
                                        grouping.var = "group.x", 
                                        verbose = FALSE)
head(cluster_markers_2_perGroup)

cluster_markers_3_perGroup <- FindConservedMarkers(NK.combined, 
                                        ident.1 = "3",
                                        grouping.var = "group.x", 
                                        verbose = FALSE)
head(cluster_markers_3_perGroup)
```

Here i am aggregating all reads into one pseudobulk expression profile per group

```{r}
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

aggregate_NK_combined <- AggregateExpression(NK.combined, 
                                             group.by = "group.x", 
                                             return.seurat = TRUE)

genes.to.label = c("CD6", "CCL4")

p1 <- CellScatter(aggregate_NK_combined,
                  cell1 = "I", 
                  cell2 = "III", 
                  highlight = genes.to.label)
p2 <- LabelPoints(plot = p1, 
                  points = genes.to.label, 
                  repel = TRUE)

p3 <- CellScatter(aggregate_NK_combined, 
                  cell1 = "I", 
                  cell2 = "V", 
                  highlight = genes.to.label)
p4 <- LabelPoints(plot = p3, 
                  points = genes.to.label, 
                  repel = TRUE)

p5 <- CellScatter(aggregate_NK_combined, 
                  cell1 = "III", 
                  cell2 = "II", 
                  highlight = genes.to.label)
p6 <- LabelPoints(plot = p5, 
                  points = genes.to.label, 
                  repel = TRUE)

p7 <- CellScatter(aggregate_NK_combined, 
                  cell1 = "III", 
                  cell2 = "IV", 
                  highlight = genes.to.label)
p8 <- LabelPoints(plot = p7, 
                  points = genes.to.label, 
                  repel = TRUE)

p2 + p4 + p6 + p8
```

```{r}
saveRDS(NK.combined, 
        file = paste0(params$out,"Rds/BM-NK_adult+child_integrated_4clust_v2.Rds"))
```

# ___END___

```{r}
sessionInfo()
```


