---
title: "NK project: DEG analysis"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/scRNA.NB.integration/metadata/"
---

PROJECT DESCRIPTION

Identify NK sub-populations in tumor-infiltrated bone marrow samples from
Neuroblastoma patients.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)

source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")

library(EnhancedVolcano)
```

# ___Load data___

## ___NB BM-derived NK cells___

```{r}
predicted_labels <- readRDS(paste0(params$out,
                                   "/Rds/fetahu2023_predicted_NKclusters_from_Crinier2021_de.n=350.Rds"))
ALL_NK <- readRDS(paste0(params$out,
                         "/Rds/fetahu2023_NK_subclusters_from_Crinier2021_de.n=350.Rds"))

ALL_NK <- ALL_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000) %>% ScaleData()

# Reorder the levels of the group-by column based
ALL_NK$group.y <- factor(ALL_NK$group.y,
                         levels = c("control",
                                    "ATRXdel",
                                    "MNA",
                                    "sporadic"))

# Reorder the levels of the NK subset annotation column
ALL_NK$predicted_NK_type_de.n350 <- factor(ALL_NK$predicted_NK_type_de.n350,
                         levels = c("hNK_Bm1",
                                    "hNK_Bm2",
                                    "hNK_Bm3",
                                    "hNK_Bm4"))
```

```{r eval=FALSE, include=FALSE}
Idents(ALL_NK) <- ALL_NK@meta.data$group.y

# identify cell IDs associated with each cluster
M <- WhichCells(ALL_NK, ident = "MNA")
S <- WhichCells(ALL_NK, ident = "sporadic")
C <- WhichCells(ALL_NK, ident = "control")
A <- WhichCells(ALL_NK, ident = "ATRXdel")

ALL_NK@active.ident <- ALL_NK$orig.ident

# re-label the NK cells to the corresponding predicted sub-clusters 
Idents(object = ALL_NK, cells = M) <- "ATRXwt"
Idents(object = ALL_NK, cells = S) <- "ATRXwt"
Idents(object = ALL_NK, cells = A) <- "ATRXmut"
Idents(object = ALL_NK, cells = C) <- "CTRL"

# Add active identities to metadata column comparison_group
ALL_NK$comparison_group <- ALL_NK@active.ident
```

# ___CD6 expression in NK sub-clusters___

```{r}
do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD6",
           plot.title = "RNA expression",
           plot.subtitle = "CD6",
           scale = TRUE,
           legend.length = 5,
           group.by = "group.y",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD6",
           plot.title = "RNA expression",
           plot.subtitle = "CD6",
           scale = TRUE,
           legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")
```

```{r fig.height=5, fig.width=10}
FeaturePlot(ALL_NK,features = "CD6",
            split.by = "group.y",cols = c("gray","purple4"))

VlnPlot(ALL_NK,
         group.by = "group.y",
         split.by = "predicted_NK_type_de.n350",
         features = "CD6",
         cols = COLOR_CODE_v2,
         log = TRUE,
         layer = "data",
        pt.size = 1)

VlnPlot(ALL_NK,
         group.by = "group.y",
         split.by = "predicted_NK_type_de.n350",
         features = "CD6",
         cols = COLOR_CODE_v2,
         log = TRUE,
         layer = "data",
        pt.size = 0)

VlnPlot(ALL_NK,
         group.by = "predicted_NK_type_de.n350",
        split.by = "group.y",
         features = "CD6",
         log = TRUE,
         layer = "data",
        pt.size = 1,
        cols = COLOR_CODE_GROUP)

VlnPlot(ALL_NK,
         group.by = "predicted_NK_type_de.n350",
        split.by = "group.y",
         features = "CD6",
         log = TRUE,
         layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP)

VlnPlot(ALL_NK,
         group.by = "group.y",
         features = "CD6",
         log = TRUE,
        cols = COLOR_CODE_GROUP,
         layer = "data",
         pt.size = 0)

do_FeaturePlot(sample = ALL_NK, 
               features = "CD6",
               plot.subtitle = "CD6 gene expression",
               plot.title = "Predicted NK sub-types in NB-BM samples",
               order = TRUE,
               reduction = "UMAP",
               pt.size = 0.5,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               min.cutoff = 0,
               max.cutoff = 3,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
```

# ___NK type gene signatures in NB BM-NK cells___

Gene lists as defined in Crinier at al., 2021

```{r}
# extract counts
counts <- LayerData(ALL_NK, assay="RNA", layer='data')
var_genes <- dimnames(counts)[[1]]

# remove genes not present in dataset
adaptive_NK_genes <- intersect(adaptive_NK_genes, var_genes)
print("Number of genes in the adaptive_NK_genes gene list: ")
length(adaptive_NK_genes)

conventional_NK_genes <- intersect(conventional_NK_genes, var_genes)
print("Number of genes in the conventional_NK_genes gene list: ")
length(conventional_NK_genes)

mature_NK_genes <- intersect(mature_NK_genes, var_genes)
print("Number of genes in the mature_NK_genes gene list: ")
length(mature_NK_genes)

NKP_genes <- intersect(NKP_genes, var_genes)
print("Number of genes in the NKP_genes gene list: ")
length(NKP_genes)

# log scale counts
ALL_NK <- ALL_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

## ___Calculate module scores___

```{r}
# remove previously calculated module scores
metadata <- ALL_NK@meta.data
cols_to_remove <- grep("module", colnames(metadata), value = TRUE)
ALL_NK@meta.data <- metadata[, !colnames(metadata) %in% cols_to_remove]

head(ALL_NK@meta.data)
```

```{r}
ALL_NK <- AddModuleScore(object = ALL_NK,
                         ctrl = 200,
                         features = adaptive_NK_genes,
                         name = "adaptive_NK_module",
                         seed = 2805)

ALL_NK <- AddModuleScore(object = ALL_NK,
                         ctrl = 200,
                         features = conventional_NK_genes,
                         name = "conventional_NK_module",
                         seed = 2805)

ALL_NK <- AddModuleScore(object = ALL_NK,
                         ctrl = 200,
                         features = NKP_genes,
                         name = "NKP_module",
                         seed = 2805)

ALL_NK <- AddModuleScore(object = ALL_NK,
                         ctrl = 200, 
                         features = mature_NK_genes,
                         name = "mature_NK_module",
                         seed = 2805)


NK_annotations <- c("mature_NK_module1",
                    "adaptive_NK_module1",
                    "conventional_NK_module1",
                    "NKP_module1")
```

## ___Plot module scores___

```{r fig.height=12, fig.width=12}
do_FeaturePlot(sample = ALL_NK, 
               features = "adaptive_NK_module1",
               plot.title = "Adaptive NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14, 
               max.cutoff = 2,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "group.y")

do_FeaturePlot(sample = ALL_NK, 
               features = "adaptive_NK_module1",
               plot.title = "Adaptive NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               max.cutoff = 2,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = ALL_NK, 
               features = "conventional_NK_module1",
               plot.title = "Conventional NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "group.y")

do_FeaturePlot(sample = ALL_NK, 
               features = "conventional_NK_module1",
               plot.title = "Conventional NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = ALL_NK, 
               features = "NKP_module1",
               plot.title = "progenitor NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "group.y")

do_FeaturePlot(sample = ALL_NK, 
               features = "NKP_module1",
               plot.title = "progenitor NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = ALL_NK, 
               features = "mature_NK_module1",
               plot.title = "Mature NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "group.y")

do_FeaturePlot(sample = ALL_NK, 
               features = "mature_NK_module1",
               plot.title = "Mature NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")
```

```{r fig.height=6.1, fig.width=7}
do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "mature_NK_module1",
           plot.title = "Mature NK cells",
           plot.subtitle = "Gene module scores",
           scale = TRUE,legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 22,
           dot.scale = 30,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Mature NK score")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "adaptive_NK_module1",
           plot.title = "adaptive NK",
           plot.subtitle = "Gene module scores",
           scale = TRUE,legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 22,
           dot.scale = 30,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Adaptive NK score")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "conventional_NK_module1",
           plot.title = "Conventional NK",
           plot.subtitle = "Gene module scores",
           scale = TRUE,legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 22,
           dot.scale = 30,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Conventional NK score")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "NKP_module1",
           plot.title = "NK Progenitors",
           plot.subtitle = "Gene module scores",
           scale = TRUE,legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 22,
           dot.scale = 30,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "NK progenitor score")
```

```{r}
VlnPlot(ALL_NK,
        features = "adaptive_NK_module1", 
        group.by = "group.y", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_v2)

VlnPlot(ALL_NK,
        features = "conventional_NK_module1", 
        group.by = "group.y", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_v2)

VlnPlot(ALL_NK,
        features = "NKP_module1", 
        group.by = "group.y", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_v2)

VlnPlot(ALL_NK,
        features = "mature_NK_module1", 
        group.by = "group.y", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_v2)
```

```{r fig.height=7, fig.width=15}
do_ViolinPlot(ALL_NK,
              features = "adaptive_NK_module1", 
              group.by = "predicted_NK_type_de.n350", 
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(ALL_NK,
              features = "conventional_NK_module1", 
              group.by = "predicted_NK_type_de.n350",
              legend.title = "NK clusters",
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(ALL_NK,
              features = "NKP_module1", 
              group.by = "predicted_NK_type_de.n350",
              legend.title = "NK clusters",
              colors.use = COLOR_CODE_v2)

do_ViolinPlot(ALL_NK,
              features = "mature_NK_module1", 
              group.by = "predicted_NK_type_de.n350",
              legend.title = "NK clusters",
              colors.use = COLOR_CODE_v2,
              font.size = 18,
              legend.position = "none")

do_ViolinPlot(ALL_NK,
              features = "mature_NK_module1", 
              group.by = "predicted_NK_type_de.n350",
              legend.title = "NK clusters",
              colors.use = COLOR_CODE_v2,
              font.size = 18,
              plot_boxplot = FALSE,
              legend.position = "none")

do_ViolinPlot(ALL_NK,
              features = "mature_NK_module1", 
              group.by = "group.y", 
              colors.use = COLOR_CODE_GROUP,
              font.size = 18,
              legend.title = "NK clusters",
              legend.position = "none")

do_ViolinPlot(ALL_NK,
              features = "mature_NK_module1", 
              group.by = "group.y", 
              colors.use = COLOR_CODE_GROUP,
              font.size = 18,
              legend.title = "NK clusters",
              legend.position = "none",
              plot_boxplot = FALSE)
```

# ___DEG analysis: ATRX mutated vs. ATRX wild-type___

___single cell RNA-seq DEG analysis___

Replicates set to biological replicates with the sample slot in the metadata

```{r}
# Access the counts matrix
counts_matrix <- GetAssayData(ALL_NK, slot = "counts")

# Filter genes with at least two reads in at least one sample
geneUniverse <- rownames(counts_matrix)[Matrix::rowSums(counts_matrix) > 1]
length(geneUniverse)
```



## ___Cluster 2___

```{r}
# Change idents to cluster annotations
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 2
cluster_of_interest <- "hNK_Bm2"

# subset scRNA-seq data to NK and NB cells 
hNK_Bm2 <- subset(ALL_NK, 
                  idents = cluster_of_interest, 
                  invert = FALSE)
hNK_Bm2 #2586 cells
```

___4-way comparison, always using ATRXm samples as the reference group [group2]___

```{r}
# DE analysis for the hNK_Bm2 NK cluster
DEG_hNK_Bm2_DEG <- DElegate::findDE(hNK_Bm2,
                           group_column = "group.y",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXdel")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_DEG, 
                        DEG_hNK_Bm2_DEG$group1 == "control")
MYCNA_vs_ATRXm <- subset(DEG_hNK_Bm2_DEG, 
                         DEG_hNK_Bm2_DEG$group1 == "MNA")
sporadic_vs_ATRXm <- subset(DEG_hNK_Bm2_DEG, 
                            DEG_hNK_Bm2_DEG$group1 == "sporadic")

MYCNA_vs_ATRXm_signif <- subset(MYCNA_vs_ATRXm, 
                                MYCNA_vs_ATRXm$padj < 0.1)
sporadic_vs_ATRXm_signif <- subset(sporadic_vs_ATRXm, 
                                   sporadic_vs_ATRXm$padj < 0.1)

common_genes <- merge(MYCNA_vs_ATRXm_signif, 
                      sporadic_vs_ATRXm_signif, 
                      by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, 
                          common_genes,  
                          by = "feature")

head(common_genes,10)
head(common_genes_all,10)

write.table(MYCNA_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_ATRXm-vs-MNA_2024-09-13.csv"))
write.table(sporadic_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_ATRXm-vs-Sporadic_2024-09-13.csv"))
```

___ATRX wild-type [MNA + sporadic] vs. ATRX mutated___

```{r}
# DE analysis for the hNK_Bm2 NK cluster
DEG_hNK_Bm2_ref <- DElegate::findDE(hNK_Bm2,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXmut")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_ref, 
                        DEG_hNK_Bm2_ref$group1 == "CTRL")
ATRXwt_vs_ATRXm <- subset(DEG_hNK_Bm2_ref, 
                          DEG_hNK_Bm2_ref$group1 == "ATRXwt")

common_genes <- merge(ctrl_vs_ATRXm, ATRXwt_vs_ATRXm,  by = "feature")
head(common_genes,10)

ctrl_vs_ATRXm_signif <- subset(ctrl_vs_ATRXm, ctrl_vs_ATRXm$padj < 0.10)
ATRXwt_vs_ATRXm_signif <- subset(ATRXwt_vs_ATRXm, ATRXwt_vs_ATRXm$padj < 0.10)

ctrl_vs_ATRXm_signif
ATRXwt_vs_ATRXm_signif

write.table(ATRXwt_vs_ATRXm, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_ATRX_mut-vs-wt_2024-09-13.csv"))

```

### ___Vulcano plot___

___1. ctrl vs. ATRXm___

```{r fig.height=7, fig.width=7}
ctrl_vs_ATRXm
rownames(ctrl_vs_ATRXm) <- ctrl_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___2. MNA vs. ATRXmut___

```{r fig.height=7, fig.width=7}
MYCNA_vs_ATRXm
rownames(MYCNA_vs_ATRXm) <- MYCNA_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "MNA vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "MNA vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___3. Sporadic vs. ATRXmut___

```{r fig.height=7, fig.width=7}
sporadic_vs_ATRXm
rownames(sporadic_vs_ATRXm) <- sporadic_vs_ATRXm$feature

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___4. ATRXwt vs. ATRXmut___

```{r fig.height=7, fig.width=7}
ATRXwt_vs_ATRXm
rownames(ATRXwt_vs_ATRXm) <- ATRXwt_vs_ATRXm$feature

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

### ___Pathways enrichment analysis___

```{r}

ATRXwt_vs_ATRXm_signif <- ATRXwt_vs_ATRXm_signif %>%
  arrange(log_fc)

sporadic_vs_ATRXm_signif <- sporadic_vs_ATRXm_signif %>%
  arrange(log_fc)

MYCNA_vs_ATRXm_signif  <- MYCNA_vs_ATRXm_signif %>%
  arrange(log_fc)

ATRXwt_vs_ATRXm_up <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc < 0) # up-regulated in ATRXmutated samples 
ATRXwt_vs_ATRXm_down <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc > 0) #down-regulated in ATRXmutated samples

MYCNA_vs_ATRXm_up <- MYCNA_vs_ATRXm_signif %>% subset(log_fc < 0)
MYCNA_vs_ATRXm_down <- MYCNA_vs_ATRXm_signif %>% subset(log_fc > 0)

sporadic_vs_ATRXm_up <- sporadic_vs_ATRXm_signif %>% subset(log_fc < 0)
sporadic_vs_ATRXm_down <- sporadic_vs_ATRXm_signif %>% subset(log_fc > 0)
```

___MNA vs. ATRXmut ___

```{r pathway enrichment MNA vs. ATRXm - cluster 2}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- MYCNA_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
               pAdjustMethod = "fdr",
               qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
                  ont = "MF",
                  OrgDb ="org.Hs.eg.db",
                  universe = names(geneUniverse),
                  readable = TRUE,
                  pAdjustMethod = "fdr",
                  qvalueCutoff = 0.10)

go_MF


if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              qvalueCutoff = 0.10, 
                              minGSSize = 2,
                              maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

___Sporadic vs. ATRXmut ___

```{r pathway enrichment Sporadic vs. ATRXm - cluster 2}

# the background genes here should be all expressed (and detected) in all NK cells
genes <- sporadic_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}


# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

___ATRXwt vs. ATRXmut___

```{r pathway enrichment analysis - Cluster 2}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ATRXwt_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}



# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

## ___Cluster 3___

```{r}
# Change idents to cluster annotations
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 3
cluster_of_interest <- "hNK_Bm3"

# subset scRNA-seq data to NK and NB cells 
hNK_Bm3 <- subset(ALL_NK, 
                  idents = cluster_of_interest, 
                  invert = FALSE)
hNK_Bm3 # 3732 cells
```

___4-way comparison, always using ATRXm samples as the reference group [group2]___

```{r}
# DE analysis for the hNK_Bm3 NK cluster
DEG_hNK_Bm3_DEG <- DElegate::findDE(hNK_Bm3,
                           group_column = "group.y",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXdel")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm3_DEG, 
                        DEG_hNK_Bm3_DEG$group1 == "control")
MYCNA_vs_ATRXm <- subset(DEG_hNK_Bm3_DEG, 
                         DEG_hNK_Bm3_DEG$group1 == "MNA")
sporadic_vs_ATRXm <- subset(DEG_hNK_Bm3_DEG, 
                            DEG_hNK_Bm3_DEG$group1 == "sporadic")

MYCNA_vs_ATRXm_signif <- subset(MYCNA_vs_ATRXm, 
                                MYCNA_vs_ATRXm$padj < 0.1)
sporadic_vs_ATRXm_signif <- subset(sporadic_vs_ATRXm, 
                                   sporadic_vs_ATRXm$padj < 0.1)

common_genes <- merge(MYCNA_vs_ATRXm_signif, 
                      sporadic_vs_ATRXm_signif, 
                      by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, 
                          common_genes,  
                          by = "feature")

head(common_genes,10)
head(common_genes_all,10)

write.table(MYCNA_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_ATRXm-vs-MNA_2024-09-13.csv"))
write.table(sporadic_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_ATRXm-vs-Sporadic_2024-09-13.csv"))
```

___ATRX wild-type [MNA + sporadic] vs. ATRX mutated___

```{r}
# DE analysis for the hNK_Bm3 NK cluster
DEG_hNK_Bm3_ref <- DElegate::findDE(hNK_Bm3,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXmut")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm3_ref, 
                        DEG_hNK_Bm3_ref$group1 == "CTRL")
ATRXwt_vs_ATRXm <- subset(DEG_hNK_Bm3_ref, 
                          DEG_hNK_Bm3_ref$group1 == "ATRXwt")

common_genes <- merge(ctrl_vs_ATRXm, ATRXwt_vs_ATRXm,  by = "feature")
head(common_genes,10)

ctrl_vs_ATRXm_signif <- subset(ctrl_vs_ATRXm, ctrl_vs_ATRXm$padj < 0.10)
ATRXwt_vs_ATRXm_signif <- subset(ATRXwt_vs_ATRXm, ATRXwt_vs_ATRXm$padj < 0.10)

ctrl_vs_ATRXm_signif
ATRXwt_vs_ATRXm_signif

write.table(ATRXwt_vs_ATRXm, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_ATRX_mut-vs-wt_2024-09-13.csv"))

```

### ___Vulcano plot___

___1. ctrl vs. ATRXm___

```{r vulcano plot - ctrl vs. ATRXm: cluster 3,fig.height=7, fig.width=7}
ctrl_vs_ATRXm
rownames(ctrl_vs_ATRXm) <- ctrl_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___2. MNA vs. ATRXmut___

```{r vulcano plot - MNA vs. ATRXmut: cluster 3,fig.height=7, fig.width=7}
MYCNA_vs_ATRXm
rownames(MYCNA_vs_ATRXm) <- MYCNA_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "MNA vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "MNA vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___3. Sporadic vs. ATRXmut___

```{r vulcano plot - Sporadic vs. ATRXmut: cluster 3, fig.height=7, fig.width=7}
sporadic_vs_ATRXm
rownames(sporadic_vs_ATRXm) <- sporadic_vs_ATRXm$feature

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___4. ATRXwt vs. ATRXmut___

```{r vulcano plot - ATRXwt vs. ATRXmut: Cluster 3, fig.height=7, fig.width=7}
ATRXwt_vs_ATRXm
rownames(ATRXwt_vs_ATRXm) <- ATRXwt_vs_ATRXm$feature

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

### ___Pathways enrichment analysis___

```{r}

ATRXwt_vs_ATRXm_signif <- ATRXwt_vs_ATRXm_signif %>%
  arrange(log_fc)

sporadic_vs_ATRXm_signif <- sporadic_vs_ATRXm_signif %>%
  arrange(log_fc)

MYCNA_vs_ATRXm_signif  <- MYCNA_vs_ATRXm_signif %>%
  arrange(log_fc)

ATRXwt_vs_ATRXm_up <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc < 0) # up-regulated in ATRXmutated samples 
ATRXwt_vs_ATRXm_down <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc > 0) #down-regulated in ATRXmutated samples

MYCNA_vs_ATRXm_up <- MYCNA_vs_ATRXm_signif %>% subset(log_fc < 0)
MYCNA_vs_ATRXm_down <- MYCNA_vs_ATRXm_signif %>% subset(log_fc > 0)

sporadic_vs_ATRXm_up <- sporadic_vs_ATRXm_signif %>% subset(log_fc < 0)
sporadic_vs_ATRXm_down <- sporadic_vs_ATRXm_signif %>% subset(log_fc > 0)
```

As there are no significant UP-regulated genes in comparison ATRXwt vs. ATRXmut,
I performed pathway enrichment on separated MNA and sporadic samples (vs. ATRXmut)

___MNA vs. ATRXmut ___

```{r pathway enrichment MNA vs. ATRXm - cluster 3}

# the background genes here should be all expressed (and detected) in all NK cells
genes <- MYCNA_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

___Sporadic vs. ATRXmut ___

```{r pathway enrichment Sporadic vs. ATRXm - cluster 3}

# the background genes here should be all expressed (and detected) in all NK cells
genes <- sporadic_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

## ___Cluster 2 + 3___

```{r}
# Change idents to cluster annotations
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 3
cluster_of_interest <- c("hNK_Bm2","hNK_Bm3")

# subset scRNA-seq data to NK and NB cells 
hNK_Bm2_and_Bm3 <- subset(ALL_NK, 
                  idents = cluster_of_interest, 
                  invert = FALSE)
hNK_Bm2_and_Bm3 # 6318 cells

Idents(hNK_Bm2_and_Bm3) <- hNK_Bm2_and_Bm3@meta.data$orig.ident 
```

___4-way comparison, always using ATRXm samples as the reference group [group2]___

```{r}
# DE analysis for the hNK_Bm3 NK cluster
DEG_hNK_Bm2_and_Bm3_DEG <- DElegate::findDE(hNK_Bm2_and_Bm3,
                           group_column = "group.y",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXdel")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_DEG, 
                        DEG_hNK_Bm2_and_Bm3_DEG$group1 == "control")
MYCNA_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_DEG, 
                         DEG_hNK_Bm2_and_Bm3_DEG$group1 == "MNA")
sporadic_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_DEG, 
                            DEG_hNK_Bm2_and_Bm3_DEG$group1 == "sporadic")

MYCNA_vs_ATRXm_signif <- subset(MYCNA_vs_ATRXm, 
                                MYCNA_vs_ATRXm$padj < 0.1)
sporadic_vs_ATRXm_signif <- subset(sporadic_vs_ATRXm, 
                                   sporadic_vs_ATRXm$padj < 0.1)

common_genes <- merge(MYCNA_vs_ATRXm_signif, 
                      sporadic_vs_ATRXm_signif, 
                      by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, 
                          common_genes,  
                          by = "feature")

head(common_genes,10)
head(common_genes_all,10)

write.table(MYCNA_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_NK_clust2+3_ATRXm-vs-MNA_2024-09-13.csv"))

write.table(sporadic_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_NK_clust2+3_ATRXm-vs-Sporadic_2024-09-13.csv"))
```

```{r}
hNK_Bm2_and_Bm3 <- hNK_Bm2_and_Bm3 %>% ScaleData()
DoHeatmap(object = hNK_Bm2_and_Bm3, 
          group.by = "group.y",
          features = common_genes$feature)

counts <- LayerData(ALL_NK, assay="RNA", layer='data')
```

___ATRX wild-type [MNA + sporadic] vs. ATRX mutated___

```{r}
Idents(hNK_Bm2_and_Bm3) <- hNK_Bm2_and_Bm3@meta.data$group.y

# DE analysis for the ALL_NK NK cluster
DEG_hNK_Bm2_and_Bm3_ref <- DElegate::findDE(hNK_Bm2_and_Bm3,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXmut")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_ref, 
                        DEG_hNK_Bm2_and_Bm3_ref$group1 == "CTRL")
ATRXwt_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_ref, 
                          DEG_hNK_Bm2_and_Bm3_ref$group1 == "ATRXwt")

common_genes <- merge(ctrl_vs_ATRXm, ATRXwt_vs_ATRXm,  by = "feature")
head(common_genes,10)

ctrl_vs_ATRXm_signif <- subset(ctrl_vs_ATRXm, ctrl_vs_ATRXm$padj < 0.10)
ATRXwt_vs_ATRXm_signif <- subset(ATRXwt_vs_ATRXm, ATRXwt_vs_ATRXm$padj < 0.10)

ctrl_vs_ATRXm_signif
ATRXwt_vs_ATRXm_signif

write.table(ATRXwt_vs_ATRXm, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, 
                          "tables/DEG_NK_clust2+3_ATRX_mut-vs-wt_2024-09-13.csv"))

```

### ___Vulcano plot___

___1. ctrl vs. ATRXm___

```{r vulcano plot - ctrl vs. ATRXm: cluster 2+3,fig.height=7, fig.width=7}
ctrl_vs_ATRXm
rownames(ctrl_vs_ATRXm) <- ctrl_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___2. MNA vs. ATRXmut___

```{r vulcano plot - MNA vs. ATRXmut: cluster 2+3, fig.height=7, fig.width=7}
MYCNA_vs_ATRXm
rownames(MYCNA_vs_ATRXm) <- MYCNA_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "MNA vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                subtitle = "MNA vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___3. Sporadic vs. ATRXmut___

```{r vulcano plot - Sporadic vs. ATRXmut: cluster 2+3, fig.height=7, fig.width=7}
sporadic_vs_ATRXm
rownames(sporadic_vs_ATRXm) <- sporadic_vs_ATRXm$feature

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___4. ATRXwt vs. ATRXmut___

```{r vulcano plot - ATRXwt vs. ATRXmut: cluster 2+3, fig.height=7, fig.width=7}
ATRXwt_vs_ATRXm
rownames(ATRXwt_vs_ATRXm) <- ATRXwt_vs_ATRXm$feature

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

### ___Pathways enrichment analysis___

```{r}

ATRXwt_vs_ATRXm_signif <- ATRXwt_vs_ATRXm_signif %>%
  arrange(log_fc)

sporadic_vs_ATRXm_signif <- sporadic_vs_ATRXm_signif %>%
  arrange(log_fc)

MYCNA_vs_ATRXm_signif  <- MYCNA_vs_ATRXm_signif %>%
  arrange(log_fc)

ATRXwt_vs_ATRXm_up <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc < 0) # up-regulated in ATRXmutated samples 
ATRXwt_vs_ATRXm_down <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc > 0) #down-regulated in ATRXmutated samples

MYCNA_vs_ATRXm_up <- MYCNA_vs_ATRXm_signif %>% subset(log_fc < 0)
MYCNA_vs_ATRXm_down <- MYCNA_vs_ATRXm_signif %>% subset(log_fc > 0)

sporadic_vs_ATRXm_up <- sporadic_vs_ATRXm_signif %>% subset(log_fc < 0)
sporadic_vs_ATRXm_down <- sporadic_vs_ATRXm_signif %>% subset(log_fc > 0)
```

___ATRX wild-type vs. mutated___

```{r pathway enrichment ATRXwt vs. ATRXm - cluster 2+3}

# the background genes here should be all expressed (and detected) in all NK cells
genes <- ATRXwt_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

___MNA vs. ATRXmut ___

```{r pathway enrichment MNA vs. ATRXm - clusters 2+3}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- MYCNA_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

head(go_BP, 10)

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

___Sporadic vs. ATRXmut ___

```{r pathway enrichment Sporadic vs. ATRXm - clusters 2+3}

# the background genes here should be all expressed (and detected) in all NK cells
genes <- sporadic_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

## ___All clusters___

___4-way comparison, always using ATRXm samples as the reference group [group2]___

```{r}
# DE analysis for the ALL_NK NK cluster
DEG_ALL_NK_DEG <- DElegate::findDE(ALL_NK,
                           group_column = "group.y",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXdel")

ctrl_vs_ATRXm <- subset(DEG_ALL_NK_DEG, 
                        DEG_ALL_NK_DEG$group1 == "control")
MYCNA_vs_ATRXm <- subset(DEG_ALL_NK_DEG, 
                         DEG_ALL_NK_DEG$group1 == "MNA")
sporadic_vs_ATRXm <- subset(DEG_ALL_NK_DEG, 
                            DEG_ALL_NK_DEG$group1 == "sporadic")

MYCNA_vs_ATRXm_signif <- subset(MYCNA_vs_ATRXm, 
                                MYCNA_vs_ATRXm$padj < 0.1)
sporadic_vs_ATRXm_signif <- subset(sporadic_vs_ATRXm, 
                                   sporadic_vs_ATRXm$padj < 0.1)

common_genes <- merge(MYCNA_vs_ATRXm_signif, 
                      sporadic_vs_ATRXm_signif, 
                      by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, 
                          common_genes,  
                          by = "feature")

head(common_genes,10)
head(common_genes_all,10)

write.table(MYCNA_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_ALL-NK_ATRXm-vs-MNA_2024-09-13.csv"))
write.table(sporadic_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_ALL-NK_ATRXm-vs-Sporadic_2024-09-13.csv"))
```

___ATRX wild-type [MNA + sporadic] vs. ATRX mutated___

```{r}
# DE analysis for the ALL_NK NK cluster
DEG_ALL_NK_ref <- DElegate::findDE(ALL_NK,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXmut")

ctrl_vs_ATRXm <- subset(DEG_ALL_NK_ref, 
                        DEG_ALL_NK_ref$group1 == "CTRL")
ATRXwt_vs_ATRXm <- subset(DEG_ALL_NK_ref, 
                          DEG_ALL_NK_ref$group1 == "ATRXwt")

common_genes <- merge(ctrl_vs_ATRXm, ATRXwt_vs_ATRXm,  by = "feature")
head(common_genes,10)

ctrl_vs_ATRXm_signif <- subset(ctrl_vs_ATRXm, ctrl_vs_ATRXm$padj < 0.10)
ATRXwt_vs_ATRXm_signif <- subset(ATRXwt_vs_ATRXm, ATRXwt_vs_ATRXm$padj < 0.10)

ctrl_vs_ATRXm_signif
ATRXwt_vs_ATRXm_signif

write.table(ATRXwt_vs_ATRXm, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_ALL-NK_ATRX_mut-vs-wt_2024-09-13.csv"))

```

### ___Vulcano plot___

___1. ctrl vs ATRXm___

```{r fig.height=7, fig.width=7}
ctrl_vs_ATRXm
rownames(ctrl_vs_ATRXm) <- ctrl_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___2. MNA vs. ATRXmut___

```{r fig.height=7, fig.width=7}
MYCNA_vs_ATRXm
rownames(MYCNA_vs_ATRXm) <- MYCNA_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "MNA vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                subtitle = "MNA vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___3. Sporadic vs. ATRXmut___

```{r fig.height=7, fig.width=7}
sporadic_vs_ATRXm
rownames(sporadic_vs_ATRXm) <- sporadic_vs_ATRXm$feature

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

___4. ATRXwt vs. ATRXmut___

```{r fig.height=7, fig.width=7}
ATRXwt_vs_ATRXm
rownames(ATRXwt_vs_ATRXm) <- ATRXwt_vs_ATRXm$feature

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                selectLab = c('CD6','ALCAM','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

### ___Pathways enrichment analysis___

```{r}
ATRXwt_vs_ATRXm_signif <- ATRXwt_vs_ATRXm_signif %>%
  arrange(log_fc)

sporadic_vs_ATRXm_signif <- sporadic_vs_ATRXm_signif %>%
  arrange(log_fc)

MYCNA_vs_ATRXm_signif  <- MYCNA_vs_ATRXm_signif %>%
  arrange(log_fc)

ATRXwt_vs_ATRXm_up <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc < 0) # up-regulated in ATRXmutated samples 
ATRXwt_vs_ATRXm_down <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc > 0) #down-regulated in ATRXmutated samples

MYCNA_vs_ATRXm_up <- MYCNA_vs_ATRXm_signif %>% subset(log_fc < 0)
MYCNA_vs_ATRXm_down <- MYCNA_vs_ATRXm_signif %>% subset(log_fc > 0)

sporadic_vs_ATRXm_up <- sporadic_vs_ATRXm_signif %>% subset(log_fc < 0)
sporadic_vs_ATRXm_down <- sporadic_vs_ATRXm_signif %>% subset(log_fc > 0)
```

As there are no significant UP-regulated genes in comparison ATRXwt vs. ATRXmut,
i performed pathway enrichment on separated MNA and sporadic samples (vs. ATRXmut)

___1. MNA vs. ATRXmut ___

```{r pathway enrichment - MNA vs. ATRXm: all clusters}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- MYCNA_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
                              minGSSize = 2,
                              maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

___2. Sporadic vs. ATRXmut ___

```{r pathway enrichment - Sporadic vs. ATRXm: all clusters}

# the background genes here should be all expressed (and detected) in all NK cells
genes <- sporadic_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
                              minGSSize = 2,
                              maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.05)) {
  # If there is at least one term with qvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

# ___The End___

```{r}
sessionInfo()
```

