---
title: "NK project: New clinical annotation"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged                 
params:
  input_BM1: "/home/rstudio/mnt_data/CancReg/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/CancReg/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/CancReg/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/CancReg/scRNA-seq_stage_M/results/published/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/mnt_out/CancReg/scRNA-seq_stage_M/results/metadata/"
---

DESCRIPTION

Evi re-analysed the SNP data for these patients and discovered that one of the
sporadic patients actually carries an ATRX deletion.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")
```

# ___Add new clinical information to the single cell object___

```{r}
fetahu2023 <- readRDS(file = paste0(params$input_scRNA))
metadata <- readRDS("/home/rstudio/mnt_out/CancReg/scRNA-seq_stage_M/results/published/metadata.rds")

fetahu2023 <- SeuratObject::as.Seurat(fetahu2023, data=NULL)
fetahu2023 <- RenameAssays(object = fetahu2023, originalexp = 'RNA')

# set cell type annotation to the original annotation
fetahu2023@active.ident <- setNames(pull(metadata[match(metadata$cell,colnames(fetahu2023)),"cellont_cluster"], "cellont_cluster"),pull(metadata[match(metadata$cell,colnames(fetahu2023)), "cell"], "cell"))

fetahu2023@meta.data$orig.ident <- fetahu2023@active.ident # cell type annotation

# set assay to RNA 
DefaultAssay(fetahu2023) <- "RNA"

fetahu2023@meta.data

columns_to_remove <- c("group")

# Drop the unwanted columns while preserving row names (cell IDs)
old_metadata <- fetahu2023@meta.data 
new_metadata <- old_metadata %>%
  dplyr::select(-all_of(columns_to_remove))
fetahu2023@meta.data <- new_metadata

fetahu2023@meta.data 
```

```{r}
# add the new clinical annotations and additional information 
full_metadata <- read.csv(file = paste0(params$metadata, 
  "fetahu2023_scRNA_seq_metadata.csv"), sep ="\t")

metadata <- fetahu2023@meta.data
metadata$cell_ID <- row.names(metadata)

# Perform the merge using left_join from dplyr, keeping the row order of df1
merged_metadata <- metadata %>%
  left_join(full_metadata, by = "sample") %>% 
  distinct(cell_ID, .keep_all = TRUE)

nrow(metadata)
nrow(merged_metadata)

row.names(merged_metadata) <- merged_metadata$cell_ID 

# Update the Seurat object's metadata slot
fetahu2023@meta.data <- merged_metadata
fetahu2023@meta.data
```

```{r eval=FALSE, include=FALSE}
colnames(fetahu2023@meta.data)
 
columns_to_remove <- c("sample_ID","File.name", "cell_ID", "sample")

# Drop the unwanted columns while preserving row names (cell IDs)
old_metadata <- fetahu2023@meta.data 
new_metadata <- old_metadata %>%
  dplyr::select(-all_of(columns_to_remove))
fetahu2023@meta.data <- new_metadata

fetahu2023@meta.data 
```

```{r}
# does log normalization make sense?
# was it already done for this object?
# does visualization change if we log normalize?

fetahu2023 <- fetahu2023 %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

```{r}
saveRDS(fetahu2023, file = paste0(params$out,"/Rds/fetahu2023_withNewMetadata.Rds"))
```

Subset to NK cells + add predicted NK subset labels

```{r}
predicted_labels <- readRDS(file = paste0(params$out,                                 "/Rds/2024_analysis/fetahu2023_predicted_NKclusters_from_Crinier2021_de.n=350.Rds"))

predicted_labels$cell_ID <- rownames(predicted_labels)

# subset scRNA-seq data to NK cells only
NK_cells <- "NK (4)"
ALL_NK <- subset(fetahu2023, 
                 idents = NK_cells, 
                 invert = FALSE)

# sanity check
nrow(predicted_labels) == nrow(ALL_NK@meta.data)

old_NK_metadata <- ALL_NK@meta.data

old_NK_metadata$cell_ID <- rownames(old_NK_metadata)

new_NK_metadata <- old_NK_metadata %>%
  dplyr::left_join(as.data.frame(predicted_labels), by = "cell_ID")

new_NK_metadata$predicted_NK_type_de.n350 <- new_NK_metadata$labels

row.names(new_NK_metadata) <- new_NK_metadata$cell_ID 

columns_to_remove <- c("cell_ID", "scores.hNK_Bm1", "scores.hNK_Bm2", 
                       "scores.hNK_Bm3", "scores.hNK_Bm4", "labels",
                       "delta.next","pruned.labels")
new_NK_metadata <- new_NK_metadata %>%
  dplyr::select(-all_of(columns_to_remove))

ALL_NK@meta.data <- new_NK_metadata
ALL_NK@meta.data

# log scale counts
ALL_NK <- ALL_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

```{r}
Idents(ALL_NK) <- ALL_NK@meta.data$group

# identify cell IDs associated with each cluster 
M <- WhichCells(ALL_NK, ident = "MYCNamp")        # MYCN amplified
S <- WhichCells(ALL_NK, ident = "ATRXwtMYCNwt") # sporadic
C <- WhichCells(ALL_NK, ident = "control")   # control / control (GNB or GNM)
A <- WhichCells(ALL_NK, ident = "ATRXdel")       # ATRX deletion or mutations

ALL_NK@active.ident <- ALL_NK$orig.ident

# Add ATRX wt vs. mutated annotation
Idents(object = ALL_NK, cells = M) <- "ATRXwt"
Idents(object = ALL_NK, cells = S) <- "ATRXwt"
Idents(object = ALL_NK, cells = A) <- "ATRXdel"
Idents(object = ALL_NK, cells = C) <- "CTRL"

# Add active identities to metadata column comparison_group
ALL_NK$comparison_group <- ALL_NK@active.ident
```

```{r}
saveRDS(ALL_NK, file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```


#############################################################################


# ___Load data___

```{r}
ALL_NK <- readRDS(file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```

```{r}
# sanity check
table(ALL_NK$sample, ALL_NK$group)
```

# ___RNA expression___

## ___CD6 expression [on NK sub-clusters]___

### ___Dot plot___

```{r fig.height=7, fig.width=7}
do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD6",
           plot.title = "RNA expression",
           plot.subtitle = "CD6",
           scale = TRUE,
           legend.length = 5,
           group.by = "group",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")

CD6_dotplot_per_NK_cluster_per_group <- do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD6",
           plot.title = "RNA expression",
           plot.subtitle = "CD6",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 20,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")
```

```{r fig.height=12, fig.width=7}
CD6_dotplot_per_NK_cluster_per_group

pdf(file = paste0(params$out,"figures/CD6_dotplot_per_NK_cluster_per_group.pdf")
    ,height = 12, width = 7)
CD6_dotplot_per_NK_cluster_per_group
dev.off()
```


### ___Feature plot___

```{r fig.height=5, fig.width=10}
do_FeaturePlot(sample = ALL_NK, 
               features = "CD6",
               plot.subtitle = "CD6 gene expression",
               plot.title = "Predicted NK sub-types in NB-BM samples",
               order = TRUE,
               reduction = "UMAP",
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               min.cutoff = 0,
               max.cutoff = 3,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
```

### ___Violin plot with stats___

```{r add stats to violin plots: CD6 expression}
gene_of_interest <- "CD6"

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$group

# Specify groups comparisons
ATRXdel_ctrl <- list("ATRXdel", "control")
ATRXdel_ATRXwt <- list("ATRXdel", c("MYCNamp","ATRXwtMYCNwt"))
ATRXdel_other <- list("ATRXdel", c("control","MYCNamp","ATRXwtMYCNwt"))

my_comparisons <- list(ATRXdel_ctrl,
                       ATRXdel_ATRXwt,
                       ATRXdel_other)

# I'M NOT USING THESE COMPARISONS BELOW!
# These were only used for testing purposes

# Create violin plot with adjusted p-values for all comparisons [to ATRXdel as ref]

# 1. PER GROUP

group_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "group",
             log = TRUE,
             cols = COLOR_CODE_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","03_DEG_", 
                      gene_of_interest, "_violinPlot_NB_groups",".png")
ggsave(group_save, plot = group_signif, width = 5, height = 4,dpi = 600)

# save the plot as PDF
pdf(paste0(params$out,"figures/","03_DEG_", 
                      gene_of_interest, "_violinPlot_NB_groups",".pdf"),
     width = 5, height = 4)
group_signif
dev.off()

# 2. PER SAMPLE/PATIENT

rename <- as.data.frame(PATIENT_GROUPS_v2)

# Rename the sample names in the Seurat object's metadata
ALL_NK@meta.data$sample <- plyr::mapvalues(ALL_NK@meta.data$sample, 
                                                   from = rownames(rename), 
                                                   to = rename$PATIENT_GROUPS_v2)

ALL_NK@meta.data$sample <- factor(ALL_NK@meta.data$sample)

patient_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "rados_et_al_2025.x",
             log = TRUE,
             cols = COLOR_CODE_PATIENTS_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
patient_signif

# Save the plot as a PNG file
patient_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample" ,".png")
ggsave(patient_save, plot = patient_signif, width = 8, height = 4,dpi = 600)

# save the plot as PDF
pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample" ,".pdf"), 
    width = 8, height = 4)
patient_signif
dev.off()

# COMPARISON BETWEEN NK SUB-TYPES, BETWEEN NB GENETIC GROUPS [ON NK CELLS!]

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350

# Check the range of the gene expression for y label positioning
gene_range <- range(FetchData(ALL_NK, vars = gene_of_interest)$CD6)
gene_max <- gene_range[2]
y_positions <- c(gene_max + 1)

# Extract the cluster identities from the Seurat object
NK_cluster_list <- unique(Idents(ALL_NK))

# plot violin plots with significance levels for each NK cluster
for (NK_cluster in NK_cluster_list){
  # Generate violin plot for each cluster
  signif_plot <- VlnPlot(ALL_NK,
        group.by = "group",
        idents = NK_cluster,
        features = gene_of_interest,
        fill.by = "group",
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") +
      ggtitle(paste("RNA expression for ", gene_of_interest, 
                    "in cluster: ", NK_cluster)) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "") + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black")) 
  
  # print the violin plot on the screen
  print(signif_plot)
  
  # Save the plot as a PNG file
  file_name <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_cluster_", NK_cluster, 
                      "_NEWCLINICALGROUPS" ,".png")
  ggsave(file_name, plot = signif_plot, width = 5, height = 4,dpi = 600)
  # save PDF
  pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_cluster_",NK_cluster ,".pdf"),
      width = 5, height = 4)
  print(signif_plot)
  dev.off()
  
}
```


## ___CD8A expression [on NK sub-clusters]___

### ___Dot plot___

```{r fig.height=17, fig.width=7}
pdf(paste0(params$out,"figures/dotplot_CD8_NB_groups.pdf"),
    height = 7,width = 7)
do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD8A",
           plot.title = "RNA expression",
           plot.subtitle = "CD8A",
           scale = TRUE,
           legend.length = 5,
           group.by = "group",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")
dev.off()

pdf(paste0(params$out,"figures/dotplot_CD8_per_NK_cluster_per_group.pdf"),
    height = 17,width = 7)
do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD8A",
           plot.title = "RNA expression",
           plot.subtitle = "CD8A",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")
dev.off()
```

### ___Feature plot___

```{r fig.height=5, fig.width=10}
do_FeaturePlot(sample = ALL_NK, 
               features = "CD8A",
               plot.subtitle = "CD8A gene expression",
               plot.title = "Predicted NK sub-types in NB-BM samples",
               order = TRUE,
               reduction = "UMAP",
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               min.cutoff = 0,
               max.cutoff = 3,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
```

### ___Violin plot with stats___

```{r add stats to violin plots: CD8A expression}
gene_of_interest <- "CD8A"

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$group

# Specify groups comparisons
ATRXdel_ctrl <- list("ATRXdel", "control")
ATRXdel_ATRXwt <- list("ATRXdel", c("MYCNamp","ATRXwtMYCNwt"))
ATRXdel_other <- list("ATRXdel", c("control","MYCNamp","ATRXwtMYCNwt"))

my_comparisons <- list(ATRXdel_ctrl,
                       ATRXdel_ATRXwt,
                       ATRXdel_other)

# I'M NOT USING THESE COMPARISONS BELOW!
# These were only used for testing purposes

# Create violin plot with adjusted p-values for all comparisons [to ATRXdel as ref]

# 1. PER GROUP

group_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "group",
             log = TRUE,
             cols = COLOR_CODE_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_NB_groups", "_NEWCLINICALGROUPS" ,".png")
ggsave(group_save, plot = group_signif, width = 5, height = 4,dpi = 600)

pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_NB_groups" ,".pdf"),
    width = 5,height = 4)
group_signif
dev.off()


# 2. PER SAMPLE/PATIENT

rename <- as.data.frame(PATIENT_GROUPS_v2)

# Rename the sample names in the Seurat object's metadata
ALL_NK@meta.data$sample <- plyr::mapvalues(ALL_NK@meta.data$sample, 
                                                   from = rownames(rename), 
                                                   to = rename$PATIENT_GROUPS_v2)

ALL_NK@meta.data$sample <- factor(ALL_NK@meta.data$sample)

patient_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "rados_et_al_2025.x",
             log = TRUE,
             cols = COLOR_CODE_PATIENTS_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
patient_signif

# Save the plot as a PNG file
patient_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".png")
ggsave(patient_save, plot = patient_signif, width = 8, height = 4,dpi = 600)

pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".pdf"),
    width = 8,height = 4)
patient_signif
dev.off()

# COMPARISON BETWEEN NK SUB-TYPES, BETWEEN NB GENETIC GROUPS [ON NK CELLS!]

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350

# Check the range of the gene expression for y label positioning
gene_range <- range(FetchData(ALL_NK, vars = gene_of_interest)$CD8A)
gene_max <- gene_range[2]
y_positions <- c(gene_max + 1)

# Extract the cluster identities from the Seurat object
NK_cluster_list <- unique(Idents(ALL_NK))

# plot violin plots with significance levels for each NK cluster
for (NK_cluster in NK_cluster_list){
  # Generate violin plot for each cluster
  signif_plot <- VlnPlot(ALL_NK,
        group.by = "group",
        idents = NK_cluster,
        features = gene_of_interest,
        fill.by = "group",
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") +
      ggtitle(paste("RNA expression for ", gene_of_interest, 
                    "in cluster: ", NK_cluster)) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "") + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black")) 
  
  # print the violin plot on the screen
  print(signif_plot)
  
  # Save the plot as a PNG file
  file_name <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_cluster_", NK_cluster, 
                      "_NEWCLINICALGROUPS" ,".png")
  ggsave(file_name, plot = signif_plot, width = 5, height = 4,dpi = 600)
  # save PDF
  pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_cluster_",NK_cluster ,".pdf"),
      width = 5,height = 4)
  print(signif_plot)
  dev.off()
}
```


## ___ALCAM expression [on NB tumor cells]___

```{r}
fetahu2023 <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_withNewMetadata.Rds"))
fetahu2023 <- fetahu2023 %>% ScaleData()
```

```{r}
gene_of_interest <- "ALCAM"

# 1. PER GROUP

group_signif <- VlnPlot(fetahu2023, 
                        idents = "NB (8)",
                        features = gene_of_interest, 
                        group.by = "group",
                        log = TRUE,
                        cols = COLOR_CODE_GROUP_v3,
                        layer = "data",
                        pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_NB_groups", "_NEWCLINICALGROUPS" ,".png")
ggsave(group_save, plot = group_signif, width = 5, height = 4,dpi = 600)

pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_NB_groups" ,".pdf"), 
    width = 5, height = 4)
group_signif
dev.off()

# 2. PER SAMPLE

sample_signif <- VlnPlot(fetahu2023, 
                        idents = "NB (8)",
                        features = gene_of_interest, 
                        group.by = "rados_et_al_2025.x",
                        log = TRUE,
                        cols = COLOR_CODE_PATIENTS_GROUP,
                        layer = "data",
                        pt.size = 0) +
  theme_classic() +
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
sample_signif

# Save the plot as a PNG file
sample_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".png")
ggsave(sample_save, plot = sample_signif, width = 5, height = 4,dpi = 600)

pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample" ,".pdf"), 
    width = 5, height = 4)
sample_signif
dev.off()
```

## ___ATRX expression [on NB tumor cells]___

```{r}
gene_of_interest <- "ATRX"

# 1. PER GROUP

group_signif <- VlnPlot(fetahu2023, 
                        idents = "NB (8)",
                        features = "rna_ATRX", 
                        group.by = "group",
                        log = TRUE,
                        cols = COLOR_CODE_GROUP_v3,
                        layer = "data",
                        pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_NB_groups", "_NEWCLINICALGROUPS" ,".png")
ggsave(group_save, plot = group_signif, width = 5, height = 4,dpi = 600)

pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_NB_groups", ".pdf"), 
    width = 5, height = 4)
group_signif
dev.off()


# 2. PER SAMPLE

sample_signif <- VlnPlot(fetahu2023, 
                        idents = "NB (8)",
                        features = "rna_ATRX", 
                        group.by = "rados_et_al_2025.x",
                        log = TRUE,
                        cols = COLOR_CODE_PATIENTS_GROUP,
                        layer = "data",
                        pt.size = 0) +
  theme_classic() +
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
sample_signif

pdf(paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", ".pdf"),
    width = 5, height = 4)
sample_signif
dev.off()

# Save the plot as a PNG file
sample_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".png")
ggsave(sample_save, plot = sample_signif, width = 5, height = 4,dpi = 600)


```



## ___IL15 expression [on Monocytes]___

```{r}
gene_of_interest <- "IL15"
clusters_of_interest <- c("M (1)","M (2)","M (10)","M (15)")

for (cluster in clusters_of_interest){
  plot <- VlnPlot(fetahu2023, 
        idents =  cluster, 
        features = "IL15",   
        group.by = "group", 
        log = TRUE,
        cols = COLOR_CODE_GROUP_v4,
        layer = "data",
        pt.size = 0)  +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXdel") + 
      ggtitle(paste(gene_of_interest, " RNA expression", "in cluster: ", cluster)) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))
  print(plot)
}
```


# ___Session Info___

```{r}
sessionInfo()
```

# ___The End___
