---
title: "NK project: New clinical annotation"
author: "Dr. Sara Wernig-Zorc"
date: "04.10.2024"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged                 
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/metadata/"
---

DESCRIPTION

Evi re-analysed the SNP data for these patients and discovered that one of the
sporadic patients actually carries an ATRX deletion.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_utils.R")

library(EnhancedVolcano)
```

# ___Add new clinical information to the single cell object___

```{r eval=FALSE, include=FALSE}
fetahu2023 <- readRDS(file = paste0(params$input_scRNA))
metadata <- readRDS("/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/metadata.rds")

fetahu2023 <- SeuratObject::as.Seurat(fetahu2023, data=NULL)
fetahu2023 <- RenameAssays(object = fetahu2023, originalexp = 'RNA')

# set cell type annotation to the original annotation
fetahu2023@active.ident <- setNames(pull(metadata[match(metadata$cell,colnames(fetahu2023)),"cellont_cluster"], "cellont_cluster"),pull(metadata[match(metadata$cell,colnames(fetahu2023)), "cell"], "cell"))

fetahu2023@meta.data$orig.ident <- fetahu2023@active.ident

# set assay to RNA 
DefaultAssay(fetahu2023) <- "RNA"

# add sex and age information
# Merge the metadata with the Seurat object's metadata based on the sample_ID and sample columns
# Ensure column names match those in your data
full_metadata <- read.csv(file = paste0(params$metadata, 
  "2024-10-04_Fetahu_complete_single_cell_cohort_EB_FINAL_Rimport.csv"), 
  sep =";",colClasses = c("character","character","character","character","character",
                          "character","character","numeric","character",
                          "character","character","character","character",
                          "numeric","character","character","character",
                          "character","character","character","character",
                          "character","character","character","character",
                          "character","character","character","character",
                          "character","character","character","character"))

full_metadata[, 1] <- as.Date(full_metadata[, 1], format = "%d/%m/%Y")
full_metadata[, 2] <- as.Date(full_metadata[, 2], format = "%d/%m/%Y")

metadata <- fetahu2023@meta.data
metadata$cell_ID <- row.names(metadata)

names(full_metadata)[names(full_metadata) == 'BM_ID'] <- 'sample'

full_metadata$sample <- gsub("-", "_", full_metadata$sample)

# Perform the merge using left_join from dplyr, keeping the row order of df1
merged_metadata <- metadata %>%
  left_join(full_metadata, by = "sample") %>% 
  distinct(cell_ID, .keep_all = TRUE)

row.names(merged_metadata) <- merged_metadata$cell_ID 

# Update the Seurat object's metadata slot
fetahu2023@meta.data <- merged_metadata
fetahu2023@meta.data

saveRDS(fetahu2023, file = paste0(params$out,"/Rds/fetahu2023_withNewMetadata.Rds"))
```

```{r eval=FALSE, include=FALSE}
predicted_labels <- readRDS(file = paste0(params$out,                                 "/Rds/fetahu2023_predicted_NKclusters_from_Crinier2021_de.n=350.Rds"))

levels(Idents(fetahu2023))

# subset scRNA-seq data 
# NK cells only

NK_cells <- "NK (4)" # change cluster number
ALL_NK <- subset(fetahu2023, idents = NK_cells, invert = FALSE)

# subset metadata
filter <- metadata$cellont_abbr == "NK"
NK.metadata <- metadata[filter,]
NK.metadata <- ALL_NK@meta.data
NK.metadata$cell_ID <- row.names(NK.metadata)
predicted_labels$cell_ID <- row.names(predicted_labels)

merged_metadata <- NK.metadata %>%
  left_join(predicted_labels, by = "cell_ID", copy = TRUE) %>% 
  distinct(cell_ID, .keep_all = TRUE)

merged_metadata <- merged_metadata %>%
  rename(predicted_NK_type_de.n350 = labels)
  
row.names(merged_metadata) <- merged_metadata$cell_ID 
  
ALL_NK@meta.data <- merged_metadata

# log scale counts
ALL_NK <- ALL_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

```{r eval=FALSE, include=FALSE}
ALL_NK@meta.data$genetic_subtype  <- factor(ALL_NK@meta.data$genetic_subtype)
Idents(ALL_NK) <- ALL_NK@meta.data$genetic_subtype 

Idents(ALL_NK) <- plyr::mapvalues(Idents(ALL_NK), 
                                      from = c("ATRXm",
                                              "localized",
                                              "MNA",
                                              "other"),
                                      to = c("ATRXm", 
                                             "control",
                                             "MYCNamp", 
                                             "ATRXwtMYCNwt"))


ALL_NK@meta.data$genetic_subtype  <- Idents(ALL_NK) 

# identify cell IDs associated with each cluster 
M <- WhichCells(ALL_NK, ident = "MYCNamp")        # MYCN amplified
S <- WhichCells(ALL_NK, ident = "ATRXwtMYCNwt") # sporadic
C <- WhichCells(ALL_NK, ident = "control")   # control / control (GNB or GNM)
A <- WhichCells(ALL_NK, ident = "ATRXm")       # ATRX deletion or mutations

ALL_NK@active.ident <- ALL_NK$orig.ident

# Add ATRX wt vs. mutated annotation
Idents(object = ALL_NK, cells = M) <- "ATRXwt"
Idents(object = ALL_NK, cells = S) <- "ATRXwt"
Idents(object = ALL_NK, cells = A) <- "ATRXmut"
Idents(object = ALL_NK, cells = C) <- "CTRL"

# Add active identities to metadata column comparison_group
ALL_NK$comparison_group <- ALL_NK@active.ident

saveRDS(ALL_NK, file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```


# ___Load data___

```{r}
ALL_NK <- readRDS(file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```

```{r}
# sanity check
table(ALL_NK$sample, ALL_NK$genetic_subtype)
```


# ___RNA expression___

## ___CD6 expression [on NK sub-clusters]___

### ___Dot plot___

```{r}
do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD6",
           plot.title = "RNA expression",
           plot.subtitle = "CD6",
           scale = TRUE,
           legend.length = 5,
           group.by = "genetic_subtype",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD6",
           plot.title = "RNA expression",
           plot.subtitle = "CD6",
           scale = TRUE,
           legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")
```

### ___Feature plot___

```{r fig.height=5, fig.width=10}
do_FeaturePlot(sample = ALL_NK, 
               features = "CD6",
               plot.subtitle = "CD6 gene expression",
               plot.title = "Predicted NK sub-types in NB-BM samples",
               order = TRUE,
               reduction = "UMAP",
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               min.cutoff = 0,
               max.cutoff = 3,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
```

### ___Violin plot with stats___

```{r add stats to violin plots: CD6 expression}
gene_of_interest <- "CD6"

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$genetic_subtype

# Specify groups comparisons
ATRXm_ctrl <- list("ATRXm", "control")
ATRXm_ATRXwt <- list("ATRXm", c("MYCNamp","ATRXwtMYCNwt"))
ATRXm_other <- list("ATRXm", c("control","MYCNamp","ATRXwtMYCNwt"))

my_comparisons <- list(ATRXm_ctrl,
                       ATRXm_ATRXwt,
                       ATRXm_other)

# I'M NOT USING THESE COMPARISONS BELOW!
# These were only used for testing purposes

# Create violin plot with adjusted p-values for all comparisons [to ATRXm as ref]

# 1. PER GROUP

group_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "genetic_subtype",
             log = TRUE,
             cols = COLOR_CODE_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_genetic_groups", "_NEWCLINICALGROUPS" ,".png")
ggsave(group_save, plot = group_signif, width = 5, height = 4,dpi = 600)

# 2. PER SAMPLE/PATIENT

rename <- as.data.frame(PATIENT_GROUPS_v2)

# Rename the sample names in the Seurat object's metadata
ALL_NK@meta.data$sample <- plyr::mapvalues(ALL_NK@meta.data$sample, 
                                                   from = rownames(rename), 
                                                   to = rename$PATIENT_GROUPS_v2)

ALL_NK@meta.data$sample <- factor(ALL_NK@meta.data$sample)

patient_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "sample",
             log = TRUE,
             cols = COLOR_CODE_PATIENTS_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
patient_signif

# Save the plot as a PNG file
patient_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".png")
ggsave(patient_save, plot = patient_signif, width = 8, height = 4,dpi = 600)



# COMPARISON BETWEEN NK SUB-TYPES, BETWEEN NB GENETIC GROUPS [ON NK CELLS!]

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350

# Check the range of the gene expression for y label positioning
gene_range <- range(FetchData(ALL_NK, vars = gene_of_interest)$CD6)
gene_max <- gene_range[2]
y_positions <- c(gene_max + 1)

# Extract the cluster identities from the Seurat object
NK_cluster_list <- unique(Idents(ALL_NK))

# plot violin plots with significance levels for each NK cluster
for (NK_cluster in NK_cluster_list){
  # Generate violin plot for each cluster
  signif_plot <- VlnPlot(ALL_NK,
        group.by = "genetic_subtype",
        idents = NK_cluster,
        features = gene_of_interest,
        fill.by = "genetic_subtype",
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") +
      ggtitle(paste("RNA expression for ", gene_of_interest, 
                    "in cluster: ", NK_cluster)) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "") + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black")) 
  
  # print the violin plot on the screen
  print(signif_plot)
  
  # Save the plot as a PNG file
  file_name <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_cluster_", NK_cluster, 
                      "_NEWCLINICALGROUPS" ,".png")
  ggsave(file_name, plot = signif_plot, width = 5, height = 4,dpi = 600)
}
```


## ___CD8A expression [on NK sub-clusters]___

### ___Dot plot___

```{r}
do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD8A",
           plot.title = "RNA expression",
           plot.subtitle = "CD8A",
           scale = TRUE,
           legend.length = 5,
           group.by = "genetic_subtype",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "CD8A",
           plot.title = "RNA expression",
           plot.subtitle = "CD8A",
           scale = TRUE,
           legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 25,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "scaled RNA expression")
```

### ___Feature plot___

```{r fig.height=5, fig.width=10}
do_FeaturePlot(sample = ALL_NK, 
               features = "CD8A",
               plot.subtitle = "CD8A gene expression",
               plot.title = "Predicted NK sub-types in NB-BM samples",
               order = TRUE,
               reduction = "UMAP",
               pt.size = 1,
               use_viridis = TRUE,
               viridis.direction = -1,
               viridis.palette = "A",
               font.size = 16,
               label.size = 16,
               min.cutoff = 0,
               max.cutoff = 3,
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Score",
               legend.length = 10,
               legend.position = "right")
```

### ___Violin plot with stats___

```{r add stats to violin plots: CD8A expression}
gene_of_interest <- "CD8A"

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$genetic_subtype

# Specify groups comparisons
ATRXm_ctrl <- list("ATRXm", "control")
ATRXm_ATRXwt <- list("ATRXm", c("MYCNamp","ATRXwtMYCNwt"))
ATRXm_other <- list("ATRXm", c("control","MYCNamp","ATRXwtMYCNwt"))

my_comparisons <- list(ATRXm_ctrl,
                       ATRXm_ATRXwt,
                       ATRXm_other)

# I'M NOT USING THESE COMPARISONS BELOW!
# These were only used for testing purposes

# Create violin plot with adjusted p-values for all comparisons [to ATRXm as ref]

# 1. PER GROUP

group_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "genetic_subtype",
             log = TRUE,
             cols = COLOR_CODE_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_genetic_groups", "_NEWCLINICALGROUPS" ,".png")
ggsave(group_save, plot = group_signif, width = 5, height = 4,dpi = 600)

# 2. PER SAMPLE/PATIENT

rename <- as.data.frame(PATIENT_GROUPS_v2)

# Rename the sample names in the Seurat object's metadata
ALL_NK@meta.data$sample <- plyr::mapvalues(ALL_NK@meta.data$sample, 
                                                   from = rownames(rename), 
                                                   to = rename$PATIENT_GROUPS_v2)

ALL_NK@meta.data$sample <- factor(ALL_NK@meta.data$sample)

patient_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "sample",
             log = TRUE,
             cols = COLOR_CODE_PATIENTS_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
patient_signif

# Save the plot as a PNG file
patient_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".png")
ggsave(patient_save, plot = patient_signif, width = 8, height = 4,dpi = 600)



# COMPARISON BETWEEN NK SUB-TYPES, BETWEEN NB GENETIC GROUPS [ON NK CELLS!]

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350

# Check the range of the gene expression for y label positioning
gene_range <- range(FetchData(ALL_NK, vars = gene_of_interest)$CD8A)
gene_max <- gene_range[2]
y_positions <- c(gene_max + 1)

# Extract the cluster identities from the Seurat object
NK_cluster_list <- unique(Idents(ALL_NK))

# plot violin plots with significance levels for each NK cluster
for (NK_cluster in NK_cluster_list){
  # Generate violin plot for each cluster
  signif_plot <- VlnPlot(ALL_NK,
        group.by = "genetic_subtype",
        idents = NK_cluster,
        features = gene_of_interest,
        fill.by = "genetic_subtype",
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") +
      ggtitle(paste("RNA expression for ", gene_of_interest, 
                    "in cluster: ", NK_cluster)) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "") + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black")) 
  
  # print the violin plot on the screen
  print(signif_plot)
  
  # Save the plot as a PNG file
  file_name <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_cluster_", NK_cluster, 
                      "_NEWCLINICALGROUPS" ,".png")
  ggsave(file_name, plot = signif_plot, width = 5, height = 4,dpi = 600)
}
```


## ___ALCAM expression [on NB tumor cells]___

```{r}
fetahu2023 <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_withNewMetadata.Rds"))
fetahu2023 <- fetahu2023 %>% ScaleData()
```

```{r}
gene_of_interest <- "ALCAM"

# 1. PER GROUP

group_signif <- VlnPlot(fetahu2023, 
                        idents = "NB (8)",
                        features = gene_of_interest, 
                        group.by = "genetic_subtype",
                        log = TRUE,
                        cols = COLOR_CODE_GROUP_v3,
                        layer = "data",
                        pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_genetic_groups", "_NEWCLINICALGROUPS" ,".png")
ggsave(group_save, plot = group_signif, width = 5, height = 4,dpi = 600)


# 2. PER SAMPLE

sample_signif <- VlnPlot(fetahu2023, 
                        idents = "NB (8)",
                        features = gene_of_interest, 
                        group.by = "sample",
                        log = TRUE,
                        cols = COLOR_CODE_PATIENTS_GROUP_v2,
                        layer = "data",
                        pt.size = 0) +
  theme_classic() +
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
sample_signif

# Save the plot as a PNG file
sample_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".png")
ggsave(sample_save, plot = sample_signif, width = 5, height = 4,dpi = 600)
```

## ___ATRX expression [on NB tumor cells]___

```{r}
gene_of_interest <- "ATRX"

# 1. PER GROUP

group_signif <- VlnPlot(fetahu2023, 
                        idents = "NB (8)",
                        features = "rna_ATRX", 
                        group.by = "genetic_subtype",
                        log = TRUE,
                        cols = COLOR_CODE_GROUP_v3,
                        layer = "data",
                        pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_genetic_groups", "_NEWCLINICALGROUPS" ,".png")
ggsave(group_save, plot = group_signif, width = 5, height = 4,dpi = 600)

# 2. PER SAMPLE

sample_signif <- VlnPlot(fetahu2023, 
                        idents = "NB (8)",
                        features = "rna_ATRX", 
                        group.by = "sample",
                        log = TRUE,
                        cols = COLOR_CODE_PATIENTS_GROUP_v2,
                        layer = "data",
                        pt.size = 0) +
  theme_classic() +
      ggtitle(paste(gene_of_interest, " RNA expression")) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
sample_signif

# Save the plot as a PNG file
sample_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".png")
ggsave(sample_save, plot = sample_signif, width = 5, height = 4,dpi = 600)
```



## ___IL15 expression [on Monocytes]___

```{r}
gene_of_interest <- "IL15"
clusters_of_interest <- c("M (1)","M (2)","M (10)","M (15)")

for (cluster in clusters_of_interest){
  plot <- VlnPlot(fetahu2023, 
        idents =  cluster, 
        features = "IL15",   
        group.by = "genetic_subtype", 
        log = TRUE,
        cols = COLOR_CODE_GROUP_v4,
        layer = "data",
        pt.size = 0)  +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " RNA expression", "in cluster: ", cluster)) + 
  labs(y = paste(gene_of_interest, " log scaled RNA expression"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))
  print(plot)
}
```


# ___DEG analysis: ATRX mutated vs. ATRX wild-type___

___single cell RNA-seq DEG analysis___

<br>
Replicates set to biological replicates with the sample slot in the metadata
<br>
<br>
ATRXm, n = 3
<br>
MYCNamp, n = 4
<br>
ATRXwtMYCNwt (MYCNwt, ATRXwt), n = 4
<br>
<br>
4-way comparison, always using ATRXm samples as the reference group [group2]
<br>
ctrl vs. ATRXm
<br>
MYCNamp vs. ATRXmut
<br>
Sporadic vs. ATRXmut
<br>
ATRX wild-type [MYCNamp + sporadic] vs. ATRX mutated
<br>

```{r}
# Access the counts matrix
counts_matrix <- GetAssayData(ALL_NK, slot = "counts")

# Filter genes with at least 11 reads in the dataset
geneUniverse <- rownames(counts_matrix)[Matrix::rowSums(counts_matrix) > 10]
length(geneUniverse)
```

## ___Cluster 2___

```{r}
# Change idents to cluster annotations
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 2
cluster_of_interest <- "hNK_Bm2"

# subset scRNA-seq data to NK and NB cells 
hNK_Bm2 <- subset(ALL_NK, 
                  idents = cluster_of_interest, 
                  invert = FALSE)
hNK_Bm2 #2586 cells
```

```{r}
# DE analysis for the hNK_Bm2 NK cluster
DEG_hNK_Bm2_DEG <- DElegate::findDE(hNK_Bm2,
                           group_column = "genetic_subtype",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXm")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_DEG, 
                        DEG_hNK_Bm2_DEG$group1 == "control")
MYCNA_vs_ATRXm <- subset(DEG_hNK_Bm2_DEG, 
                         DEG_hNK_Bm2_DEG$group1 == "MYCNamp")
sporadic_vs_ATRXm <- subset(DEG_hNK_Bm2_DEG, 
                            DEG_hNK_Bm2_DEG$group1 == "ATRXwtMYCNwt")

MYCNA_vs_ATRXm_signif <- subset(MYCNA_vs_ATRXm, 
                                MYCNA_vs_ATRXm$padj < 0.10)
sporadic_vs_ATRXm_signif <- subset(sporadic_vs_ATRXm, 
                                   sporadic_vs_ATRXm$padj < 0.10)

common_genes <- merge(MYCNA_vs_ATRXm_signif, 
                      sporadic_vs_ATRXm_signif, 
                      by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, 
                          common_genes,  
                          by = "feature")

head(common_genes,10)
head(common_genes_all,10)

write.table(MYCNA_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_ATRXm-vs-MYCNamp_2024-10-04.csv"))
write.table(sporadic_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_ATRXm-vs-Sporadic_2024-10-04.csv"))
```


```{r}
# DE analysis for the hNK_Bm2 NK cluster
DEG_hNK_Bm2_ref <- DElegate::findDE(hNK_Bm2,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXmut")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_ref, 
                        DEG_hNK_Bm2_ref$group1 == "CTRL")
ATRXwt_vs_ATRXm <- subset(DEG_hNK_Bm2_ref, 
                          DEG_hNK_Bm2_ref$group1 == "ATRXwt")

common_genes <- merge(ctrl_vs_ATRXm, ATRXwt_vs_ATRXm,  by = "feature")
head(common_genes,10)

ctrl_vs_ATRXm_signif <- subset(ctrl_vs_ATRXm, ctrl_vs_ATRXm$padj < 0.10)
ATRXwt_vs_ATRXm_signif <- subset(ATRXwt_vs_ATRXm, ATRXwt_vs_ATRXm$padj < 0.10)

ctrl_vs_ATRXm_signif
ATRXwt_vs_ATRXm_signif

write.table(ATRXwt_vs_ATRXm, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_ATRX_mut-vs-wt_2024-10-04.csv"))

```

### ___Vulcano plot___

#### ___1. ctrl vs. ATRXm___

```{r fig.height=7, fig.width=7}
ctrl_vs_ATRXm
rownames(ctrl_vs_ATRXm) <- ctrl_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___2. MYCNamp vs. ATRXmut___

```{r fig.height=7, fig.width=7}
MYCNA_vs_ATRXm
rownames(MYCNA_vs_ATRXm) <- MYCNA_vs_ATRXm$feature

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G','PFN1'),
                subtitle = "MYCNamp vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "MYCNamp vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___3. Sporadic vs. ATRXmut___

```{r fig.height=7, fig.width=7}
sporadic_vs_ATRXm
rownames(sporadic_vs_ATRXm) <- sporadic_vs_ATRXm$feature

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___4. ATRXwt vs. ATRXmut___

```{r fig.height=7, fig.width=7}
ATRXwt_vs_ATRXm
rownames(ATRXwt_vs_ATRXm) <- ATRXwt_vs_ATRXm$feature

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

##### ___DEG Heatmap___

```{r}
hNK_Bm2 <- hNK_Bm2 %>% ScaleData()
DoHeatmap(object = hNK_Bm2, 
           group.by = "genetic_subtype",
           features = ATRXwt_vs_ATRXm_signif$feature)
```


### ___Pathways enrichment analysis___

```{r}
ATRXwt_vs_ATRXm_signif <- ATRXwt_vs_ATRXm_signif %>%
  arrange(log_fc)

sporadic_vs_ATRXm_signif <- sporadic_vs_ATRXm_signif %>%
  arrange(log_fc)

MYCNA_vs_ATRXm_signif  <- MYCNA_vs_ATRXm_signif %>%
  arrange(log_fc)

ATRXwt_vs_ATRXm_up <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc < 0) # up-regulated in ATRXmutated samples !!
ATRXwt_vs_ATRXm_down <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc > 0) # down-regulated in ATRXmutated samples !!

MYCNA_vs_ATRXm_up <- MYCNA_vs_ATRXm_signif %>% subset(log_fc < 0)
MYCNA_vs_ATRXm_down <- MYCNA_vs_ATRXm_signif %>% subset(log_fc > 0)

sporadic_vs_ATRXm_up <- sporadic_vs_ATRXm_signif %>% subset(log_fc < 0)
sporadic_vs_ATRXm_down <- sporadic_vs_ATRXm_signif %>% subset(log_fc > 0)
```

#### ___MYCNamp vs. ATRXmut ___

```{r pathway enrichment MYCNamp vs. ATRXm - cluster 2, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- MYCNA_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
               pAdjustMethod = "fdr",
               qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
                  ont = "MF",
                  OrgDb ="org.Hs.eg.db",
                  universe = geneUniverse,
                  readable = TRUE,
                  pAdjustMethod = "fdr",
                  qvalueCutoff = 0.10)

go_MF


if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              qvalueCutoff = 0.10, 
                              minGSSize = 2,
                              maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

#### ___Sporadic vs. ATRXmut ___

```{r pathway enrichment Sporadic vs. ATRXm - cluster 2, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- sporadic_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}


# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

#### ___ATRXwt vs. ATRXmut___

```{r pathway enrichment analysis - Cluster 2, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ATRXwt_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}



# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

## ___Cluster 3___

```{r}
# Change idents to cluster annotations
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 3
cluster_of_interest <- "hNK_Bm3"

# subset scRNA-seq data to NK and NB cells 
hNK_Bm3 <- subset(ALL_NK, 
                  idents = cluster_of_interest, 
                  invert = FALSE)
hNK_Bm3 # 3732 cells
```

```{r}
# DE analysis for the hNK_Bm3 NK cluster
DEG_hNK_Bm3_DEG <- DElegate::findDE(hNK_Bm3,
                           group_column = "genetic_subtype",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXm")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm3_DEG, 
                        DEG_hNK_Bm3_DEG$group1 == "control")
MYCNA_vs_ATRXm <- subset(DEG_hNK_Bm3_DEG, 
                         DEG_hNK_Bm3_DEG$group1 == "MYCNamp")
sporadic_vs_ATRXm <- subset(DEG_hNK_Bm3_DEG, 
                            DEG_hNK_Bm3_DEG$group1 == "ATRXwtMYCNwt")

MYCNA_vs_ATRXm_signif <- subset(MYCNA_vs_ATRXm, 
                                MYCNA_vs_ATRXm$padj < 0.1)
sporadic_vs_ATRXm_signif <- subset(sporadic_vs_ATRXm, 
                                   sporadic_vs_ATRXm$padj < 0.1)

common_genes <- merge(MYCNA_vs_ATRXm_signif, 
                      sporadic_vs_ATRXm_signif, 
                      by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, 
                          common_genes,  
                          by = "feature")

head(common_genes,10)
head(common_genes_all,10)

write.table(MYCNA_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_ATRXm-vs-MYCNamp_2024-10-04.csv"))
write.table(sporadic_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_ATRXm-vs-Sporadic_2024-10-04.csv"))
```

```{r}
# DE analysis for the hNK_Bm3 NK cluster
DEG_hNK_Bm3_ref <- DElegate::findDE(hNK_Bm3,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXmut")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm3_ref, 
                        DEG_hNK_Bm3_ref$group1 == "CTRL")
ATRXwt_vs_ATRXm <- subset(DEG_hNK_Bm3_ref, 
                          DEG_hNK_Bm3_ref$group1 == "ATRXwt")

common_genes <- merge(ctrl_vs_ATRXm, ATRXwt_vs_ATRXm,  by = "feature")
head(common_genes,10)

ctrl_vs_ATRXm_signif <- subset(ctrl_vs_ATRXm, ctrl_vs_ATRXm$padj < 0.10)
ATRXwt_vs_ATRXm_signif <- subset(ATRXwt_vs_ATRXm, ATRXwt_vs_ATRXm$padj < 0.10)

ctrl_vs_ATRXm_signif
ATRXwt_vs_ATRXm_signif

write.table(ATRXwt_vs_ATRXm, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_ATRX_mut-vs-wt_2024-10-04.csv"))

```

### ___Vulcano plot___

#### ___1. ctrl vs. ATRXm___

```{r vulcano plot - ctrl vs. ATRXm: cluster 3,fig.height=7, fig.width=7}
ctrl_vs_ATRXm
rownames(ctrl_vs_ATRXm) <- ctrl_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___2. MYCNamp vs. ATRXmut___

```{r vulcano plot - MYCNamp vs. ATRXmut: cluster 3,fig.height=7, fig.width=7}
MYCNA_vs_ATRXm
rownames(MYCNA_vs_ATRXm) <- MYCNA_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "MYCNamp vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "MYCNamp vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___3. Sporadic vs. ATRXmut___

```{r vulcano plot - Sporadic vs. ATRXmut: cluster 3, fig.height=7, fig.width=7}
sporadic_vs_ATRXm
rownames(sporadic_vs_ATRXm) <- sporadic_vs_ATRXm$feature

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___4. ATRXwt vs. ATRXmut___

```{r vulcano plot - ATRXwt vs. ATRXmut: Cluster 3, fig.height=7, fig.width=7}
ATRXwt_vs_ATRXm
rownames(ATRXwt_vs_ATRXm) <- ATRXwt_vs_ATRXm$feature

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

### ___Pathways enrichment analysis___

```{r}
ATRXwt_vs_ATRXm_signif <- ATRXwt_vs_ATRXm_signif %>%
  arrange(log_fc)

sporadic_vs_ATRXm_signif <- sporadic_vs_ATRXm_signif %>%
  arrange(log_fc)

MYCNA_vs_ATRXm_signif  <- MYCNA_vs_ATRXm_signif %>%
  arrange(log_fc)

ATRXwt_vs_ATRXm_up <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc < 0) # up-regulated in ATRXmutated samples 
ATRXwt_vs_ATRXm_down <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc > 0) #down-regulated in ATRXmutated samples

MYCNA_vs_ATRXm_up <- MYCNA_vs_ATRXm_signif %>% subset(log_fc < 0)
MYCNA_vs_ATRXm_down <- MYCNA_vs_ATRXm_signif %>% subset(log_fc > 0)

sporadic_vs_ATRXm_up <- sporadic_vs_ATRXm_signif %>% subset(log_fc < 0)
sporadic_vs_ATRXm_down <- sporadic_vs_ATRXm_signif %>% subset(log_fc > 0)
```

#### ___MYCNamp vs. ATRXmut ___

```{r pathway enrichment MYCNamp vs. ATRXm - cluster 3, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- MYCNA_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

#### ___Sporadic vs. ATRXmut ___

```{r pathway enrichment Sporadic vs. ATRXm - cluster 3, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- sporadic_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

## ___Cluster 2 + 3___

```{r}
# Change idents to cluster annotations
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 3
cluster_of_interest <- c("hNK_Bm2","hNK_Bm3")

# subset scRNA-seq data to NK and NB cells 
hNK_Bm2_and_Bm3 <- subset(ALL_NK, 
                  idents = cluster_of_interest, 
                  invert = FALSE)
hNK_Bm2_and_Bm3 # 6318 cells

Idents(hNK_Bm2_and_Bm3) <- hNK_Bm2_and_Bm3@meta.data$orig.ident 
```

```{r}
# DE analysis for the hNK_Bm3 NK cluster
DEG_hNK_Bm2_and_Bm3_DEG <- DElegate::findDE(hNK_Bm2_and_Bm3,
                           group_column = "genetic_subtype",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXm")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_DEG, 
                        DEG_hNK_Bm2_and_Bm3_DEG$group1 == "control")
MYCNA_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_DEG, 
                         DEG_hNK_Bm2_and_Bm3_DEG$group1 == "MYCNamp")
sporadic_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_DEG, 
                            DEG_hNK_Bm2_and_Bm3_DEG$group1 == "ATRXwtMYCNwt")

MYCNA_vs_ATRXm_signif <- subset(MYCNA_vs_ATRXm, 
                                MYCNA_vs_ATRXm$padj < 0.1)
sporadic_vs_ATRXm_signif <- subset(sporadic_vs_ATRXm, 
                                   sporadic_vs_ATRXm$padj < 0.1)

common_genes <- merge(MYCNA_vs_ATRXm_signif, 
                      sporadic_vs_ATRXm_signif, 
                      by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, 
                          common_genes,  
                          by = "feature")

head(common_genes,10)
head(common_genes_all,10)

write.table(MYCNA_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_NK_clust2+3_ATRXm-vs-MYCNamp_2024-10-04.csv"))

write.table(sporadic_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_NK_clust2+3_ATRXm-vs-Sporadic_2024-10-04.csv"))
```

```{r}
Idents(hNK_Bm2_and_Bm3) <- hNK_Bm2_and_Bm3@meta.data$genetic_subtype

# DE analysis for the ALL_NK NK cluster
DEG_hNK_Bm2_and_Bm3_ref <- DElegate::findDE(hNK_Bm2_and_Bm3,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXmut")

ctrl_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_ref, 
                        DEG_hNK_Bm2_and_Bm3_ref$group1 == "CTRL")
ATRXwt_vs_ATRXm <- subset(DEG_hNK_Bm2_and_Bm3_ref, 
                          DEG_hNK_Bm2_and_Bm3_ref$group1 == "ATRXwt")

common_genes <- merge(ctrl_vs_ATRXm, ATRXwt_vs_ATRXm,  by = "feature")
head(common_genes,10)

ctrl_vs_ATRXm_signif <- subset(ctrl_vs_ATRXm, ctrl_vs_ATRXm$padj < 0.10)
ATRXwt_vs_ATRXm_signif <- subset(ATRXwt_vs_ATRXm, ATRXwt_vs_ATRXm$padj < 0.10)

ctrl_vs_ATRXm_signif
ATRXwt_vs_ATRXm_signif

write.table(ATRXwt_vs_ATRXm, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, 
                          "tables/DEG_NK_clust2+3_ATRX_mut-vs-wt_2024-10-04.csv"))

```

### ___Vulcano plot___

#### ___1. ctrl vs. ATRXm___

```{r vulcano plot - ctrl vs. ATRXm: cluster 2+3,fig.height=7, fig.width=7}
ctrl_vs_ATRXm
rownames(ctrl_vs_ATRXm) <- ctrl_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___2. MYCNamp vs. ATRXmut___

```{r vulcano plot - MYCNamp vs. ATRXmut: cluster 2+3, fig.height=7, fig.width=7}
MYCNA_vs_ATRXm
rownames(MYCNA_vs_ATRXm) <- MYCNA_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "MYCNamp vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                subtitle = "MYCNamp vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___3. Sporadic vs. ATRXmut___

```{r vulcano plot - Sporadic vs. ATRXmut: cluster 2+3, fig.height=7, fig.width=7}
sporadic_vs_ATRXm
rownames(sporadic_vs_ATRXm) <- sporadic_vs_ATRXm$feature

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___4. ATRXwt vs. ATRXmut___

```{r vulcano plot - ATRXwt vs. ATRXmut: cluster 2+3, fig.height=7, fig.width=7}
ATRXwt_vs_ATRXm
rownames(ATRXwt_vs_ATRXm) <- ATRXwt_vs_ATRXm$feature

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2+3",
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

### ___Pathways enrichment analysis___

```{r}
ATRXwt_vs_ATRXm_signif <- ATRXwt_vs_ATRXm_signif %>%
  arrange(log_fc)

sporadic_vs_ATRXm_signif <- sporadic_vs_ATRXm_signif %>%
  arrange(log_fc)

MYCNA_vs_ATRXm_signif  <- MYCNA_vs_ATRXm_signif %>%
  arrange(log_fc)

ATRXwt_vs_ATRXm_up <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc < 0) # up-regulated in ATRXmutated samples 
ATRXwt_vs_ATRXm_down <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc > 0) #down-regulated in ATRXmutated samples

MYCNA_vs_ATRXm_up <- MYCNA_vs_ATRXm_signif %>% subset(log_fc < 0)
MYCNA_vs_ATRXm_down <- MYCNA_vs_ATRXm_signif %>% subset(log_fc > 0)

sporadic_vs_ATRXm_up <- sporadic_vs_ATRXm_signif %>% subset(log_fc < 0)
sporadic_vs_ATRXm_down <- sporadic_vs_ATRXm_signif %>% subset(log_fc > 0)
```

#### ___ATRX wild-type vs. mutated___

```{r pathway enrichment ATRXwt vs. ATRXm - cluster 2+3, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ATRXwt_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

#### ___MYCNamp vs. ATRXmut ___

```{r pathway enrichment MYCNamp vs. ATRXm - clusters 2+3, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- MYCNA_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

head(go_BP, 10)

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

#### ___Sporadic vs. ATRXmut ___

```{r pathway enrichment Sporadic vs. ATRXm - clusters 2+3, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- sporadic_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+Bm_hNK3 in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

## ___All clusters___

```{r}
# DE analysis for the ALL_NK NK cluster
DEG_ALL_NK_DEG <- DElegate::findDE(ALL_NK,
                           group_column = "genetic_subtype",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXm")

ctrl_vs_ATRXm <- subset(DEG_ALL_NK_DEG, 
                        DEG_ALL_NK_DEG$group1 == "control")
MYCNA_vs_ATRXm <- subset(DEG_ALL_NK_DEG, 
                         DEG_ALL_NK_DEG$group1 == "MYCNamp")
sporadic_vs_ATRXm <- subset(DEG_ALL_NK_DEG, 
                            DEG_ALL_NK_DEG$group1 == "ATRXwtMYCNwt")

MYCNA_vs_ATRXm_signif <- subset(MYCNA_vs_ATRXm, 
                                MYCNA_vs_ATRXm$padj < 0.1)
sporadic_vs_ATRXm_signif <- subset(sporadic_vs_ATRXm, 
                                   sporadic_vs_ATRXm$padj < 0.1)

common_genes <- merge(MYCNA_vs_ATRXm_signif, 
                      sporadic_vs_ATRXm_signif, 
                      by = "feature")
common_genes_all <- merge(ctrl_vs_ATRXm, 
                          common_genes,  
                          by = "feature")

head(common_genes,10)
head(common_genes_all,10)

write.table(MYCNA_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_ALL-NK_ATRXm-vs-MYCNamp_2024-10-04.csv"))
write.table(sporadic_vs_ATRXm_signif, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_ALL-NK_ATRXm-vs-Sporadic_2024-10-04.csv"))
```

```{r}
# DE analysis for the ALL_NK NK cluster
DEG_ALL_NK_ref <- DElegate::findDE(ALL_NK,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "sample",
                           compare = "ATRXmut")

ctrl_vs_ATRXm <- subset(DEG_ALL_NK_ref, 
                        DEG_ALL_NK_ref$group1 == "CTRL")
ATRXwt_vs_ATRXm <- subset(DEG_ALL_NK_ref, 
                          DEG_ALL_NK_ref$group1 == "ATRXwt")

common_genes <- merge(ctrl_vs_ATRXm, ATRXwt_vs_ATRXm,  by = "feature")
head(common_genes,10)

ctrl_vs_ATRXm_signif <- subset(ctrl_vs_ATRXm, ctrl_vs_ATRXm$padj < 0.10)
ATRXwt_vs_ATRXm_signif <- subset(ATRXwt_vs_ATRXm, ATRXwt_vs_ATRXm$padj < 0.10)

ctrl_vs_ATRXm_signif
ATRXwt_vs_ATRXm_signif

write.table(ATRXwt_vs_ATRXm, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_ALL-NK_ATRX_mut-vs-wt_2024-10-04.csv"))

```

### ___Vulcano plot___

#### ___1. ctrl vs ATRXm___

```{r fig.height=7, fig.width=7}
ctrl_vs_ATRXm
rownames(ctrl_vs_ATRXm) <- ctrl_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ctrl_vs_ATRXm, 
                lab = rownames(ctrl_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                subtitle = "CTRL vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___2. MYCNamp vs. ATRXmut___

```{r fig.height=7, fig.width=7}
MYCNA_vs_ATRXm
rownames(MYCNA_vs_ATRXm) <- MYCNA_vs_ATRXm$feature

library(EnhancedVolcano)

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "MYCNamp vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(MYCNA_vs_ATRXm, 
                lab = rownames(MYCNA_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                subtitle = "MYCNamp vs. ATRXm",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___3. Sporadic vs. ATRXmut___

```{r fig.height=7, fig.width=7}
sporadic_vs_ATRXm
rownames(sporadic_vs_ATRXm) <- sporadic_vs_ATRXm$feature

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(sporadic_vs_ATRXm, 
                lab = rownames(sporadic_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                subtitle = "Sporadic vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

#### ___4. ATRXwt vs. ATRXmut___

```{r fig.height=7, fig.width=7}
ATRXwt_vs_ATRXm
rownames(ATRXwt_vs_ATRXm) <- ATRXwt_vs_ATRXm$feature

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                selectLab = c('CD6','TGFB1','IL32','KLRF1','FCER1G',
                              'PFN1'),
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6
                )

EnhancedVolcano(ATRXwt_vs_ATRXm, 
                lab = rownames(ATRXwt_vs_ATRXm),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in ALL NK cells",
                subtitle = "ATRXwt vs. ATRXmut",
                pCutoff = 0.1,
                FCcutoff = 0.5,
                pointSize = 3,
                labSize = 6,
                drawConnectors = TRUE,
                widthConnectors = 0.75
                )
```

### ___Pathways enrichment analysis___

```{r}
ATRXwt_vs_ATRXm_signif <- ATRXwt_vs_ATRXm_signif %>%
  arrange(log_fc)

sporadic_vs_ATRXm_signif <- sporadic_vs_ATRXm_signif %>%
  arrange(log_fc)

MYCNA_vs_ATRXm_signif  <- MYCNA_vs_ATRXm_signif %>%
  arrange(log_fc)

ATRXwt_vs_ATRXm_up <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc < 0) # up-regulated in ATRXmutated samples 
ATRXwt_vs_ATRXm_down <- ATRXwt_vs_ATRXm_signif %>% subset(log_fc > 0) #down-regulated in ATRXmutated samples

MYCNA_vs_ATRXm_up <- MYCNA_vs_ATRXm_signif %>% subset(log_fc < 0)
MYCNA_vs_ATRXm_down <- MYCNA_vs_ATRXm_signif %>% subset(log_fc > 0)

sporadic_vs_ATRXm_up <- sporadic_vs_ATRXm_signif %>% subset(log_fc < 0)
sporadic_vs_ATRXm_down <- sporadic_vs_ATRXm_signif %>% subset(log_fc > 0)
```

#### ___1. MYCNamp vs. ATRXmut ___

```{r pathway enrichment - MYCNamp vs. ATRXm: all clusters, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- MYCNA_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
                              minGSSize = 2,
                              maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

#### ___2. Sporadic vs. ATRXmut ___

```{r pathway enrichment - Sporadic vs. ATRXm: all clusters, fig.height=10,fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- sporadic_vs_ATRXm_up$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)


# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.10, 
               minGSSize = 2,
               maxGSSize = 50000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
                              minGSSize = 2,
                              maxGSSize = 50000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(as.data.frame(kegg_enrichment)$pvalue < 0.05)) {
  # If there is at least one term with pvalue < 0.05, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in NK cells in ATRXm: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}
```

## ___Cluster annotation___

### ___Calculate module scores___

Gene lists as defined in Crinier at al., 2021

```{r}
# remove previously calculated module scores
metadata <- ALL_NK@meta.data
cols_to_remove <- grep("module", colnames(metadata), value = TRUE)
ALL_NK@meta.data <- metadata[, !colnames(metadata) %in% cols_to_remove]

head(ALL_NK@meta.data)
```

```{r}
# extract counts
counts <- LayerData(ALL_NK, assay="RNA", layer='data')
var_genes <- dimnames(counts)[[1]]

# remove genes not present in dataset
adaptive_NK_genes <- intersect(adaptive_NK_genes, var_genes)
print("Number of genes in the adaptive_NK_genes gene list: ")
length(adaptive_NK_genes)

conventional_NK_genes <- intersect(conventional_NK_genes, var_genes)
print("Number of genes in the conventional_NK_genes gene list: ")
length(conventional_NK_genes)

mature_NK_genes <- intersect(mature_NK_genes, var_genes)
print("Number of genes in the mature_NK_genes gene list: ")
length(mature_NK_genes)

NKP_genes <- intersect(NKP_genes, var_genes)
print("Number of genes in the NKP_genes gene list: ")
length(NKP_genes)

# log scale counts
ALL_NK <- ALL_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

```{r}
ALL_NK <- AddModuleScore(object = ALL_NK,
                         ctrl = 50,
                         features = adaptive_NK_genes,
                         name = "adaptive_NK_module",
                         seed = 2805)

ALL_NK <- AddModuleScore(object = ALL_NK,
                         ctrl = 50,
                         features = conventional_NK_genes,
                         name = "conventional_NK_module",
                         seed = 2805)

ALL_NK <- AddModuleScore(object = ALL_NK,
                         ctrl = 50,
                         features = NKP_genes,
                         name = "NKP_module",
                         seed = 2805)

ALL_NK <- AddModuleScore(object = ALL_NK,
                         ctrl = 50, 
                         features = mature_NK_genes,
                         name = "mature_NK_module",
                         seed = 2805)
```

### ___Plot module scores___

#### ___Feature plot___

```{r fig.height=12, fig.width=12}
do_FeaturePlot(sample = ALL_NK, 
               features = "adaptive_NK_module1",
               plot.title = "Adaptive NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14, 
               max.cutoff = 0.5,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "genetic_subtype")

do_FeaturePlot(sample = ALL_NK, 
               features = "adaptive_NK_module1",
               plot.title = "Adaptive NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               max.cutoff = 0.5,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = ALL_NK, 
               features = "conventional_NK_module1",
               plot.title = "Conventional NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               max.cutoff = 0.5,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "genetic_subtype")

do_FeaturePlot(sample = ALL_NK, 
               features = "conventional_NK_module1",
               plot.title = "Conventional NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               max.cutoff = 0.5,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = ALL_NK, 
               features = "NKP_module1",
               plot.title = "progenitor NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               max.cutoff = 0.5,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "genetic_subtype")

do_FeaturePlot(sample = ALL_NK, 
               features = "NKP_module1",
               plot.title = "progenitor NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               max.cutoff = 0.5,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")

do_FeaturePlot(sample = ALL_NK, 
               features = "mature_NK_module1",
               plot.title = "Mature NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "genetic_subtype")

do_FeaturePlot(sample = ALL_NK, 
               features = "mature_NK_module1",
               plot.title = "Mature NK cells",
               plot.subtitle = "Gene module scores",
               pt.size = 1,
               reduction = "UMAP",
               viridis.direction = -1,
               viridis.palette = "D",
               font.size = 14,
               label.size = 14,
               slot = "data",
               legend.title.face = "bold",
               axis.text.face = "bold",
               plot.title.face = "bold",
               plot.axes = TRUE,
               legend.title = "Module score",
               legend.length = 10,
               split.by = "predicted_NK_type_de.n350")
```

#### ___Dot plot___

```{r fig.height=6.1, fig.width=7}
do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "mature_NK_module1",
           plot.title = "Mature NK cells",
           plot.subtitle = "Gene module scores",
           scale = TRUE,legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 22,
           dot.scale = 30,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Mature NK score")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "adaptive_NK_module1",
           plot.title = "adaptive NK",
           plot.subtitle = "Gene module scores",
           scale = TRUE,legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 22,
           dot.scale = 30,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Adaptive NK score")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "conventional_NK_module1",
           plot.title = "Conventional NK",
           plot.subtitle = "Gene module scores",
           scale = TRUE,legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 22,
           dot.scale = 30,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Conventional NK score")

do_DotPlot(sample = ALL_NK, 
           axis.text.x.angle = 0,
           features = "NKP_module1",
           plot.title = "NK Progenitors",
           plot.subtitle = "Gene module scores",
           scale = TRUE,legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 22,
           dot.scale = 30,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "NK progenitor score")
```

#### ___Violin plot___

##### ___Adaptive NK gene signature___

```{r}
VlnPlot(ALL_NK,
        features = "adaptive_NK_module1", 
        group.by = "genetic_subtype", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_v2, 
        y.max = 0.01)
```

##### ___Conventional NK gene signature___

```{r}
VlnPlot(ALL_NK,
        features = "conventional_NK_module1", 
        group.by = "genetic_subtype", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_v2, 
        y.max = 0.01)
```

##### ___NK progenitors gene signature___

```{r}
VlnPlot(ALL_NK,
        features = "NKP_module1", 
        group.by = "genetic_subtype", 
        split.by = "predicted_NK_type_de.n350",
        cols = COLOR_CODE_v2, 
        y.max = 0.01)
```

##### ___NK maturity gene signature [with stats]___

```{r add stats to violin plots: NK maturity score}
gene_of_interest <- "mature_NK_module1"

# 1. PER GROUP

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$genetic_subtype

# Create violin plot with adjusted p-values for all comparisons [to ATRXm as ref]
group_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "genetic_subtype",
             log = FALSE,
             cols = COLOR_CODE_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " module score")) + 
  labs(y = paste(gene_of_interest, " module score"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
group_signif

# Save the plot as a PNG file
group_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_genetic_groups", "_NEWCLINICALGROUPS" ,".png")
ggsave(group_save, plot = group_signif, width = 6, height = 4,dpi = 600)


# 2. PER SAMPLE/PATIENT

patient_signif <- VlnPlot(ALL_NK, 
             features = gene_of_interest, 
             group.by = "sample",
             log = TRUE,
             cols = COLOR_CODE_PATIENTS_GROUP,
             layer = "data",
             pt.size = 0) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") + 
      ggtitle(paste(gene_of_interest, " module scores")) + 
  labs(y = paste(gene_of_interest, " module scores"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))

# Show the plot
patient_signif

# Save the plot as a PNG file
patient_save <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_sample", "_NEWCLINICALGROUPS" ,".png")
ggsave(patient_save, plot = patient_signif, width = 8, height = 4,dpi = 600)


# COMPARISON BETWEEN NK SUB-TYPES, BETWEEN NB GENETIC GROUPS [ON NK CELLS!]

# Change identities in the Seurat object to genetic groups
Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350

# Check the range of the gene expression for y label positioning
gene_range <- range(FetchData(ALL_NK, vars = gene_of_interest)$CD6)
gene_max <- gene_range[2]
y_positions <- c(gene_max + 1)

# Extract the cluster identities from the Seurat object
NK_cluster_list <- unique(Idents(ALL_NK))

# plot violin plots with significance levels for each NK cluster
for (NK_cluster in NK_cluster_list){
  # Generate violin plot for each cluster
  signif_plot <- VlnPlot(ALL_NK,
        group.by = "genetic_subtype",
        idents = NK_cluster,
        features = gene_of_interest,
        fill.by = "genetic_subtype",
        log = FALSE, # FOR MODULE SCORE SHOULD BE SET TO FALSE
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP) +
  theme_classic() +
  stat_compare_means(method = "wilcox.test", 
                     label = "p.signif",
                     ref.group = "ATRXm") +
      ggtitle(paste("Module scores for ", gene_of_interest, 
                    "in cluster: ", NK_cluster)) + 
  labs(y = paste(gene_of_interest, " module score"),
       x = "")  + 
    theme(legend.position = "none",  # Remove the legend
          axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"))
  
  # print the violin plot on the screen
  print(signif_plot)
  
  # Save the plot as a PNG file
  file_name <- paste0(params$out,"figures/","fetahu2023_violinPlot_", 
                      gene_of_interest, "_cluster_", NK_cluster, 
                      "_NEWCLINICALGROUPS" ,".png")
  ggsave(file_name, plot = signif_plot, width = 6, height = 4,dpi = 600)
}
```



# ___Session Info___

```{r}
sessionInfo()
```

# ___The End___
