---
title: "NK project: Desensetized NK cells & NK-MO-DC regulation axis "
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
---

PROJECT DESCRIPTION

Identify NK sub-populations in tumor-infiltrated bone marrow samples from
Neuroblastoma patients.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")
```

# ___Load data___

```{r}
fetahu2023 <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_withNewMetadata.Rds"))
predicted_labels <- readRDS(file = paste0(params$out,"/Rds/",params$input_scRNA))
ALL_NK <- readRDS(file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```

```{r eval=FALSE, include=FALSE}
group <- "genetic_subtype" 
NK_subtype <- "predicted_NK_type_de.n350"  

# Create a new metadata column that concatenates values from both columns
ALL_NK$group_per_NK_subtype <- paste(ALL_NK@meta.data[[group]], 
                                      ALL_NK@meta.data[[NK_subtype]], 
                                      sep = "_")

# Check if the new column is added
ALL_NK@meta.data$group_per_NK_subtype <- as.factor(ALL_NK@meta.data$group_per_NK_subtype)
table(ALL_NK@meta.data$group_per_NK_subtype)
```

```{r eval=FALSE, include=FALSE}
saveRDS(ALL_NK, file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```


# ___Desensitized/exhausted NK cells___

```{r}
ALL_NK <- readRDS(file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```


List of genes from NK Cell Exhaustion review paper from 2017 
DOI: 10.3389/fimmu.2017.00760

2B4 protein = CD244 gene
NKG2A protein = KLRC1 gene
NKG2D protein = KLRK1 gene
NRC proteins = NCOA1, NCOA2, NCOA3
CD16 protein = FCGR3A Gene gene
TIM3 protein = HAVCR2 gene


```{r fig.height=8.36, fig.width=8.20}
exhausted_NK_genes <- list(up = c("PDCD1", "HAVCR2", "KLRC1"),
                           down = c("KLRK1","FCGR3A","NCOA1", "NCOA2", 
                                    "NCOA3","CD226","CD244","IFNG")) 

do_DotPlot(ALL_NK, 
           features = exhausted_NK_genes, 
           group.by = "group_per_NK_subtype", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_DotPlot(ALL_NK, 
           features = exhausted_NK_genes, 
           group.by = "NK_subtype_per_group", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

```


```{r}
# Consensus NK cell exhaustion genes from 
# https://onlinelibrary.wiley.com/doi/epdf/10.1111/imm.13464

exhausted_NK_genes <- list("Increased in NK cell exhaustion" = c("TIGIT", "LAG3", "HAVCR2","PDCD1","KLRC1","CD69",
                                  "KIR2DL1", "KIR2DL2", "KIR2DL3", "KIR2DL5A",
                                  "KIR2DL5B", "KIR3DL1", "KIR3DL2", "KIR3DL3"),
                           "Decreased in NK cell exhaustion" = c("KLRK1","CD226","LAMP1",
                                    "IFNG","TNF","PRF1",
                                    "FCGR3A",
                                    "GZMA","GZMB", "GZMH", "GZMK","GZMM")) 

exhust_NK_plot_v1 <- do_DotPlot(ALL_NK, 
           features = exhausted_NK_genes, 
           group.by = "group_per_NK_subtype", 
           legend.title = "Scaled RNA expression",
           scale = TRUE,
           dot.scale = 10,
           dot_border = NA,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

exhust_NK_plot_v1

pdf(file = paste0(params$out,"figures/NK_cell_exhustion_dotplot_perNKgroup.pdf"),
    width = 9.82,  height = 7.63)
exhust_NK_plot_v1
dev.off()

exhust_NK_plot_v2 <- do_DotPlot(ALL_NK, 
           features = exhausted_NK_genes, 
           group.by = "NK_subtype_per_group",
           legend.title = "Scaled RNA expression",
           scale = TRUE,
           dot.scale = 10,
           dot_border = NA,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1)

exhust_NK_plot_v2

pdf(file = paste0(params$out,"figures/NK_cell_exhustion_dotplot_perNBgroup.pdf"),
    width = 9.82,  height = 7.63)
exhust_NK_plot_v2
dev.off()

exhust_NK_plot_v3 <- do_DotPlot(ALL_NK, 
           features = exhausted_NK_genes, 
           group.by = "group",
           legend.title = "Scaled RNA expression",
           scale = TRUE,
           dot_border = NA,
           dot.scale = 10,
           font.size = 16,
           dot_border = NA,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1)

exhust_NK_plot_v3

pdf(file = paste0(params$out,"figures/NK_cell_exhustion_dotplot_perAllNK.pdf"),
    width = 8.38,  height = 3.99)
exhust_NK_plot_v3
dev.off()

exhust_NK_plot_v4 <- do_DotPlot(ALL_NK, 
           features = exhausted_NK_genes, 
           group.by = "group",
           legend.title = "Normalized RNA expression",
           scale = FALSE,
           dot_border = NA,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1)

exhust_NK_plot_v4

pdf(file = paste0(params$out,"figures/NK_cell_exhustion_dotplot_perAllNK_noScaling.pdf"),
    width = 8.38,  height = 3.99)
exhust_NK_plot_v4
dev.off()
```
```{r}
# Module score for NK cell exhaustion (up-regulated genes)
NK_cell_exhaustion <- c("TIGIT", "LAG3", "HAVCR2","PDCD1","KLRC1","CD69")

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_cell_exhaustion,
                         name = "NK_cell_exhaustion_score",
                         seed = "2805",
                         slot = "data",
                         nbin = 10)

moduleScore_plot <- do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 90,
           features = "NK_cell_exhaustion_score1",
           plot.title = "NK cell exhaustion score",
           plot.subtitle = "Upregulated genes",
           scale = TRUE,
           legend.length = 10,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled module score")

moduleScore_plot

pdf(file = paste0(params$out,"figures/NK_cell_exhustion_moduleScore.pdf"),
    width = 7,  height =10)
moduleScore_plot
dev.off()
```

```{r}
# Module score for NK cell exhaustion (up-regulated genes)
NK_cell_exhaustion_down <- c("KLRK1","CD226","LAMP1","IFNG","TNF","PRF1", "FCGR3A",
                        "GZMA","GZMB", "GZMH", "GZMK","GZMM")

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_cell_exhaustion_down,
                         name = "NK_cell_exhaustion_down",
                         seed = "2805",
                         slot = "data",
                         nbin = 10)

moduleScore_plot_down <- do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 90,
           features = "NK_cell_exhaustion_down1",
           plot.title = "NK cell exhaustion score",
           plot.subtitle = "Downregulated genes",
           scale = TRUE,
           legend.length = 10,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled module score")

moduleScore_plot_down

pdf(file = paste0(params$out,"figures/NK_cell_exhustion_moduleScore_downGenes.pdf"),
    width = 7,  height =10)
moduleScore_plot_down
dev.off()
```


# ___CCC analysis___

The degree to which NK cells are active against tumour cells is highly dependent 
on the activation state of the NK cells, which is regulated in large part by 
cytokines derived from dendritic cells (DCs) and monocytes

## ___NK cells: L-R expression per cluster per group___

```{r fig.height=8.36, fig.width=15}
# based on The cancer–natural killer cell immunity cycle review paper
# high or low expression of  these genes is associated with a good prognosis 
# in a pan-cancer analysis

NK_receptors <- c("ADORA2A","CD2","CD226", "CD244", "TIGIT","TMIGD2", 
                  "NCR1","NCR2","NCR3","NT5E", "ITGAL", "ITGB2","PDCD1",
                  "KLRC1","KLRC2", "KLRC3", "KLRC4", "KLRD1", "KLRF1",
                  "KLRG1", "KLRG2","KLRK1", "KIR2DL1", "KIR2DL3", "KIR2DL4", 
                  "KIR2DS4","KIR3DL1", "KIR3DL2", "KIR3DL3")

NK_secreted_cytokines <- c("CCL5","FLT3LG", "XCL1", "XCL2")

NK_prognosis_markers <- list("NK receptors" = c("ADORA2A","CD2","CD226", "CD244", 
                                              "TIGIT","TMIGD2", "NCR1","NCR2",
                                              "NCR3","NT5E", "ITGAL", "ITGB2",
                                              "PDCD1","KLRC1","KLRC2", "KLRC3", 
                                              "KLRC4", "KLRD1","KLRF1","KLRG1", 
                                              "KLRG2","KLRK1","KIR2DL1","KIR2DL3", 
                                              "KIR2DL4", "KIR3DL1", "KIR3DL2", 
                                              "KIR3DL3","KIR2DS4"),
                             "NK secreted 
                             cytokines" = c("CCL5","FLT3LG", "XCL1", "XCL2"))

do_DotPlot(ALL_NK, 
           features = NK_prognosis_markers, 
           group.by = "group_per_NK_subtype", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )
```


## ___NK - MO - DC axis___

```{r}
cell_types_of_interest_v2 <- c("NK (4)",
                               "pDC (12)",
                               "M (1)","M (2)", "M (10)","M (15)")

fetahu2023.NK.DC.M <- subset(fetahu2023,
                             idents = cell_types_of_interest_v2)

do_DotPlot(fetahu2023.NK.DC.M, 
           features = NK_activating_cytokines, 
           group.by = "orig.ident", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           plot.title = "NK-cell activating cytokines",
           plot.subtitle = "Monocytes",
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )

do_ViolinPlot(fetahu2023.NK.DC.M, 
           features = c("IL2", "IL15", "IL12A", "IL12B","IL18"), 
           group.by = "orig.ident", 
           split.by = "genetic_subtype",
           plot_boxplot = FALSE,
           slot = "data")
```

```{r eval=FALSE, include=FALSE}
saveRDS(fetahu2023.NK.DC.M, 
        file = paste0(params$out,"Rds/fetahu2023_NK_DC_MO.Rds"))
```


### ___Immunosupresive Monocytes (M10)___

```{r}
cell_types_of_interest_v3 <- c("M (10)")

fetahu2023.M10 <- subset(fetahu2023,
                         idents = cell_types_of_interest_v3)

do_DotPlot(fetahu2023.M10, 
           features = NK_activating_cytokines, 
           plot.title = "NK-cell activating cytokines",
           plot.subtitle = "Immunosupresive Monocytes (M10)",
           group.by = "genetic_subtype", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )
```

### ___NK cells___

Chemokines secreated by NK cells and associated with recruiting type 1 
conventional dendritic cells

```{r fig.height=8.34, fig.width=5}
do_DotPlot(ALL_NK, 
           features = c("CCL5","XCL1","FLT3LG"), 
           group.by = "group_per_NK_subtype", 
           scale = TRUE,
           legend.length = 10,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1
           )
```

-------------------------------------------------------------------------------

# ___Session Info___

```{r}
sessionInfo()
```
