---
title: "NK project: Functional markers"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/scRNA.scATAC/metadata/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_setup.R")
source("/home/rstudio/scRNA.scATAC/R/00_NK_paper_utils.R")
```

# ___Load data I___

```{r}
fetahu2023 <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_withNewMetadata.Rds"))
ALL_NK <- readRDS(file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```


## ___some processing___

```{r eval=FALSE, include=FALSE}
Idents(ALL_NK) <- ALL_NK$group
levels(Idents(ALL_NK))

# subset scRNA-seq data 
## ATRX
ATRX_NK <- subset(ALL_NK, 
                   idents = "ATRXdel", 
                   invert = FALSE)

## MNA
MNA_NK <- subset(ALL_NK, 
                   idents = "MYCNamp", 
                   invert = FALSE)

## ATRXwtMYCNwt
sporadic_NK <- subset(ALL_NK, 
                   idents = "ATRXwtMYCNwt", 
                   invert = FALSE)

## control
CTRL_NK <- subset(ALL_NK, 
                   idents = "control", 
                   invert = FALSE)
```

```{r eval=FALSE, include=FALSE}
ALL_NK <- ALL_NK %>% NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000) %>% ScaleData()
CTRL_NK <- CTRL_NK %>% NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000) %>% ScaleData()
MNA_NK <- MNA_NK %>% NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000) %>% ScaleData()
ATRX_NK <- ATRX_NK %>% NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000) %>% ScaleData()
sporadic_NK <- sporadic_NK %>% NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000) %>% ScaleData()
```

```{r eval=FALSE, include=FALSE}
saveRDS(ATRX_NK, 
        file = paste0(params$out,"/Rds/fetahu2023_ATRXdel_BM_NK.Rds"))
saveRDS(ATRX_NK_sub, 
        file = paste0(params$out,"/Rds/fetahu2023_ATRXdel-sub_BM_NK.Rds"))
saveRDS(sporadic_NK, 
        file = paste0(params$out,"/Rds/fetahu2023_sporadic_BM_NK.Rds"))
saveRDS(CTRL_NK, 
        file = paste0(params$out,"/Rds/fetahu2023_fetal_BM_NK.Rds"))
saveRDS(MNA_NK, 
        file = paste0(params$out,"/Rds/fetahu2023_MNA_BM_NK.Rds"))
```

# ___Load data II___

```{r}
CTRL_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_fetal_BM_NK.Rds"))
MNA_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_MNA_BM_NK.Rds"))
ATRX_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXdelut_BM_NK.Rds"))
sporadic_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_sporadic_BM_NK.Rds"))
```

```{r eval=FALSE, include=FALSE}
Idents(ATRX_NK) <- ATRX_NK$predicted_NK_type_de.n350
table(ATRX_NK$predicted_NK_type_de.n350)

clusters_of_interest <- c("hNK_Bm2","hNK_Bm3")

# subset to hNK_Bm2 & hNK_Bm3 only !
ATRX_NK_sub <- subset(ATRX_NK, 
                   idents = clusters_of_interest, 
                   invert = FALSE)

ATRX_NK_sub <- ATRX_NK_sub %>% NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000) %>% ScaleData()
```

```{r eval=FALSE, include=FALSE}
saveRDS(ATRX_NK_sub, 
        file = paste0(params$out,"/Rds/fetahu2023_ATRXdelut_BM_NK_subset_hNK_Bm2_3.Rds"))
```


```{r}
ATRX_NK_sub <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_ATRXdelut_BM_NK_subset_hNK_Bm2_3.Rds"))
```


# ___NK Cluster markers in the NB dataset___

```{r cluster markers: CTRL}
de_results <- DElegate::findDE(object = CTRL_NK,
                               group_column = "predicted_NK_type_de.n350",
                               replicate_column = "rados_et_al_2025")

marker_results <- DElegate::FindAllMarkers2(object = CTRL_NK,
                                            group_column = "predicted_NK_type_de.n350",
                                            replicate_column = "rados_et_al_2025")

marker_results %>%
    group_by(group1) %>%
    top_n(n = 10, wt = log_fc) -> top10

signif_DEGs <- top10 %>% filter(padj < 0.05)

# TOP 10 markers per cluster 
DEG_tumor <- DoHeatmap(CTRL_NK, 
                       group.by = "predicted_NK_type_de.n350",
                       features = signif_DEGs$feature) + NoLegend()
DEG_tumor

do_DotPlot(CTRL_NK,
           features = signif_DEGs$feature, 
           group.by = "predicted_NK_type_de.n350", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "CTRL"
           )

signif_DEGs_hNK_Bm1 <- top10 %>% filter(group1 == "hNK_Bm1")
signif_DEGs_hNK_Bm2 <- top10 %>% filter(group1 == "hNK_Bm2")
signif_DEGs_hNK_Bm3 <- top10 %>% filter(group1 == "hNK_Bm3")
signif_DEGs_hNK_Bm4 <- top10 %>% filter(group1 == "hNK_Bm4")
```

hNK_Bm1: SPON2+, CX3CR1+, CXCR2+ NK cells = Cytotoxic, inflammatory, tissue-resident NK cells	- typically found in peripheral tissues or inflamed sites - Function: High cytotoxicity, tissue adaptation
hNK_Bm2: TOX2+, CXCR6+, XCL1+ NK cells = Activated, effector, and exhausted-like NK cells - typically found in tumor microenvironment, chronic infections - Function: Effector function, but signs of exhaustion
hNK_Bm3: SELL+, CCR7+, IFIT3+ NK cells = Circulating, naïve/memory-like, and antiviral NK cells - typically found in blood, lymphoid tissues - Function: Naïve or memory-like, migratory, antiviral response
hNK_Bm4: B3GAT1+, GPR27+, SLC1A7+ NK cells =	Metabolically adapted, tissue-resident, suppressive NK cells - typically found in specialized tissues (brain, tumor niches) - Function:	Long-lived, stress-adapted, immune regulation

Now let's look at ATRXdel samples and how do those differ (we have mainly hNK_Bm2
and hNK_Bm3 NK cells here so i will focus on those):

```{r cluster markers: ATRX}
de_results <- DElegate::findDE(object = ATRX_NK,
                               group_column = "predicted_NK_type_de.n350",
                               replicate_column = "rados_et_al_2025")

marker_results <- DElegate::FindAllMarkers2(object = ATRX_NK,
                                            group_column = "predicted_NK_type_de.n350",
                                            replicate_column = "rados_et_al_2025")

marker_results %>%
    group_by(group1) %>%
    top_n(n = 10, wt = log_fc) -> top10

signif_DEGs <- top10 %>% filter(padj < 0.05)

# TOP 10 markers per cluster 
DEG_tumor <- DoHeatmap(ATRX_NK, group.by = "predicted_NK_type_de.n350",
                       features = signif_DEGs$feature) + NoLegend()
DEG_tumor

do_DotPlot(ATRX_NK,
           features = signif_DEGs$feature, 
           group.by = "predicted_NK_type_de.n350", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "ATRXdel"
           )
# should we remove cluster 1 and 4 when plotting dotplots of ATRXdel samples?
# since we have only few cells mapping to clusters 1 and 4, they tend to
# give a wrong impression

signif_DEGs_hNK_Bm2 <- signif_DEGs %>% filter(group1  == "hNK_Bm2")
signif_DEGs_hNK_Bm3 <- signif_DEGs %>% filter(group1  == "hNK_Bm3")

print(signif_DEGs_hNK_Bm2$feature)
print(signif_DEGs_hNK_Bm3$feature)

```

In Rebuffet et al., 2024 (Eric Vivier lab), they defined these genes as markers of
NK cytotoxicity: GZMB and PRF1, and markers of NK cell maturity: CD160, CD247, 
ADGRG1, NKG7, FCER1G, LAIR2, SPON2, CLIC3 and CHST2 

If we look at the genes that are differentialy expressed in NK cluster 2 and 3,
compared to all other clusters:

hNK_Bm3: FGFBP2+, GZMB+, FCGR3A+ NK cells = tumor-infiltrating cytotoxic NK cells;
are highly cytotoxic NK cells, likely involved in direct tumor killing. 
Their strong ADCC function suggests they are effector cells involved in rapid immune responses.

hNK_Bm2: ID3+, LTB+, CD28+ NK cells = resident, adaptive-like NK cells involved in immune surveillance. They appear more memory-like, tissue-adapted, and immune-regulatory. 
Their expression of ID3, CD28, and TNFSF8 suggests they could be long-lived, 
adaptive-like NK cells that respond to persistent stimuli, such as chronic inflammation 
or tumor microenvironment. They might be quiescent, metabolically flexible NK population 
capable of responding to chronic immune challenges rather than acting as immediate cytotoxic effectors, as hNK_Bm3 NK cluster.

```{r cluster markers: ATRXsub}
de_results <- DElegate::findDE(object = ATRX_NK_sub,
                               group_column = "predicted_NK_type_de.n350",
                               replicate_column = "rados_et_al_2025")

marker_results <- DElegate::FindAllMarkers2(object = ATRX_NK_sub,
                                            group_column = "predicted_NK_type_de.n350",
                                            replicate_column = "rados_et_al_2025")

marker_results %>%
    group_by(group1) %>%
    top_n(n = 10, wt = log_fc) -> top10

signif_DEGs <- top10 %>% filter(padj < 0.05)

# TOP 10 markers per cluster 
DEG_tumor <- DoHeatmap(ATRX_NK_sub, group.by = "predicted_NK_type_de.n350",
                       features = signif_DEGs$feature) + NoLegend()
DEG_tumor

do_DotPlot(ATRX_NK_sub,
           features = signif_DEGs$feature, 
           group.by = "predicted_NK_type_de.n350", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "ATRXdel"
           )
# should we remove cluster 1 and 4 when plotting dotplots of ATRXdel samples?
# since we have only few cells mapping to clusters 1 and 4, they tend to
# give a wrong impression

signif_DEGs_hNK_Bm2 <- signif_DEGs %>% filter(group1  == "hNK_Bm2")
signif_DEGs_hNK_Bm3 <- signif_DEGs %>% filter(group1  == "hNK_Bm3")

print(signif_DEGs_hNK_Bm2$feature)
print(signif_DEGs_hNK_Bm3$feature)

```

# ___Exploration of age and sex dependence in the data___

```{r}
full_metadata <- read.csv(paste0(params$metadata,
  "2024-10-04_Fetahu_complete_single_cell_cohort_EB_FINAL_Rimport.csv"), 
  sep = ";")


# filter by unique sample ID
unique.metadata <- full_metadata %>% 
  distinct(BM_ID,.keep_all = TRUE)

unique.metadata <- as.data.frame(unique.metadata)

# plot age distribution per comparison group
ggplot(unique.metadata, aes(x = genetic_subtype, 
               y = age_at_Dx_months, 
               fill = genetic_subtype)) +
  geom_boxplot()

# input table
input <- table(unique.metadata$genetic_subtype, 
               unique.metadata$sex,
               unique.metadata$stage)
input <- as.data.frame(input)
colnames(input) <- c("group","sex","stage","value")


# plot sex distribution in the whole dataset
ggplot(input, aes(x=sex,y=value,fill=sex)) +
  geom_bar(stat = "identity") + 
  theme_minimal() + 
  scale_fill_manual(values=c("hotpink",
                             "darkblue")) +
  theme(text = element_text(size = 20, colour = "black"))

# plot sex distribution per comparison group
ggplot(input, aes(x=group,y=value,fill=sex)) +
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title="Number of male/female donors per group") + 
  theme_minimal() + 
  scale_fill_manual(values=c("hotpink",
                             "darkblue")) +
  theme(text = element_text(size = 20, colour = "black"))

```

Since in ATRXdel samples, only 30 cells map to the NK clusters 1 or 4, 
i removed those from downstream analysis (below)

# ___NK functional markers___

NK cell killing occurs either via secretion of perforin-1 (PRF1 Gene) and 
granzyme B (GZMB gene) or direct cell-to-cell contact via death receptors 
TRAIL (TNFRSF10B gene) and FAS receptor (TNFSF6 gene). FAS ligand (FASLG gene);
TRAIL ligand (TNFSF10 gene) or mediate cell killing by recruiting other immune 
cells via cytokine and chemokine secretion, and induce cell death via 
Interferon-γ (IFNG gene) or TNF secretion

## ___Canonical NK markers___

### ___ALL___

#### ___Dotplot___

```{r fig.height=7.9, fig.width=11.38}
do_DotPlot(ALL_NK, 
           features = NK_functional_markers, 
           group.by = "group_per_NK_subtype", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "ALL"
           )
```

### ___ATRX clusters 2&3 only___

```{r}
do_DotPlot(ATRX_NK_sub, 
           features = NK_functional_markers, 
           group.by = "predicted_NK_type_de.n350", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "ATRX"
           )
```

## ___NK TFs___

### ___ALL___

```{r fig.height=7.9, fig.width=11.38}
do_DotPlot(ALL_NK, 
           features = TFs, 
           group.by = "group_per_NK_subtype",
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "ALL",
           legend.title = "Scaled expression"
           )
```

### ___ATRX clusters 2&3 only___

```{r fig.height=5, fig.width=8}
do_DotPlot(ATRX_NK_sub,
           features = TFs, 
           group.by = "predicted_NK_type_de.n350", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1,
           plot.title = "ATRX",
           legend.title = "Scaled expression"
           )
```

## ___NK cytokines___

### ___ALL___

```{r fig.height=7.9, fig.width=11.38}
do_DotPlot(ALL_NK, 
           features = cytokines, 
           group.by = "group_per_NK_subtype",
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "ALL cytokines",
           legend.title = "Scaled expression",
           cluster = FALSE
           )
```

### ___ATRX clusters 2&3 only___

```{r}
do_DotPlot(ATRX_NK_sub, 
           features = cytokines, 
           group.by = "predicted_NK_type_de.n350", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "Cytokines",
           plot.subtitle = "ATRX",
           legend.title = "Scaled expression",
           cluster = FALSE
           )
```

## ___NK maturation genes___

### ___ALL___

```{r fig.height=7.9, fig.width=11.38}
do_DotPlot(ALL_NK, 
           features = NK_maturation_genes, 
           group.by = "group_per_NK_subtype", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "NK maturation genes",
           plot.subtitle = "ALL",
           legend.title = "Scaled expression",
           cluster = FALSE
           )
```

### ___ATRX clusters 2&3 only___

```{r}
do_DotPlot(ATRX_NK_sub, 
           features = NK_maturation_genes, 
           group.by = "predicted_NK_type_de.n350", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           plot.title = "NK maturation genes",
           plot.subtitle = "ATRX",
           legend.title = "Scaled expression",
           cluster = FALSE
           )
```

## ___NK subtype genes___

hNK_Bm2 subset was characterized by higher levels of CD122, CD27, and CD69, 
whereas hNK_Bm3 cells had higher levels of CD127.

Bone marrow NK cells revealed that the expression of CD160 and CD52 (CAMPATH-1) 
on the surface of CD56bright cells was mutually exclusive, further discriminating 
between hNK_Bm2 and hNK_Bm3, respectively.

IL2RB gene = CD122 protein
NCAM1 gene = CD56 protein
IL7R gene = CD127 protein
FCGR3A gene = CD16 protein

### ___Dotplot___

#### ___ALL___

```{r fig.height=7.9, fig.width=11.38}
do_DotPlot(ALL_NK,
           features = NK_functional_markers$NK_subtype_markers,
           group.by = "group_per_NK_subtype", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           legend.title = "Scaled expression",
           cluster = FALSE
           )
```

### ___ATRX clusters 2&3 only___

```{r}
do_DotPlot(ATRX_NK_sub,
           features = NK_functional_markers$NK_subtype_markers,
           group.by = "predicted_NK_type_de.n350", 
           scale = TRUE,
           dot.scale = 10,
           font.size = 16,
           number.breaks = 4,
           use_viridis = TRUE,
           viridis.palette = "A",
           viridis.direction = -1, 
           legend.title = "Scaled expression",
           cluster = FALSE
           )
```

### ___Violinplot___

#### ___ALL___

```{r fig.height=12, fig.width=10}
VlnPlot(ALL_NK,
        features = NK_functional_markers$NK_subtype_markers,
        log = FALSE,
        layer = "data",
        slot = "data")
```

#### ___ATRX clusters 2&3 only___

```{r fig.height=12, fig.width=10}
VlnPlot(ATRX_NK_sub,
        features = NK_functional_markers$NK_subtype_markers,
        group.by = "predicted_NK_type_de.n350",
        log = FALSE,
        layer = "data",
        slot = "data")
```

#### ___MNA___

```{r fig.height=12, fig.width=10}
VlnPlot(MNA_NK,
        features = NK_functional_markers$NK_subtype_markers,
        group.by = "predicted_NK_type_de.n350",
        log = FALSE,
        layer = "data",
        slot = "data")
```

#### ___CTRL___

```{r fig.height=12, fig.width=10}
VlnPlot(CTRL_NK,
        features = NK_functional_markers$NK_subtype_markers,
        group.by = "predicted_NK_type_de.n350",
        log = FALSE,
        layer = "data",
        slot = "data")
```

#### ___Sporadic___

```{r fig.height=12, fig.width=10}
VlnPlot(sporadic_NK,
        features = NK_functional_markers$NK_subtype_markers,
        group.by = "predicted_NK_type_de.n350",
        log = FALSE,
        layer = "data",
        slot = "data")
```

# ___Session Info___

```{r}
sessionInfo()
```
