---
title: "NK project: Paper figures"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/CancReg/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/CancReg/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/CancReg/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/NK_project/Rds/fetahu2023_withNewMetadata.Rds"
  input_NK: "/home/rstudio/mnt_out/NK_project/Rds/ALL_NK_cluster_withNewMetadata.Rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/NK-NB_study/metadata/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")
```

# ___Input_data___

```{r}
fetahu2023 <- readRDS(file = params$input_scRNA)
ALL_NK <-  readRDS(file = params$input_NK)
```

```{r}
ATRX_NK <- readRDS(file = paste0(params$out,"/Rds/fetahu2023_ATRXmut_BM_NK_subset_hNK_Bm2_3.Rds"))
```

```{r eval=FALSE, include=FALSE}
fetahu2023@meta.data$case_control <- NA
fetahu2023@meta.data <- fetahu2023@meta.data %>% 
  mutate(case_control = ifelse(fetahu2023$histology == "NB","NB", "CTRL"))

fetahu2023@meta.data$case_control  <- factor(fetahu2023@meta.data$case_control)
fetahu2023@meta.data$case_control  <- droplevels(fetahu2023@meta.data$case_control )

# Change to the new group identitites
Idents(fetahu2023) <- fetahu2023@meta.data$case_control


#crinier2021 <- readRDS(file = paste0(params$input_BM1))
#crinier2021[["RNA"]] <- as(object = crinier2021[["RNA"]], Class = "Assay5")

#crinier2021$cohort <- "adults"
#ALL_NK$cohort <- "children"

# convert a v4 assay to a v5 assay
#ALL_NK[["RNA"]] <- as(object = ALL_NK[["RNA"]],  Class = "Assay5")

ALL_NK@meta.data$case_control <- NA
ALL_NK@meta.data <- ALL_NK@meta.data %>% 
  mutate(case_control = ifelse(ALL_NK$histology.x == "NB","NB", "CTRL"))

ALL_NK@meta.data$case_control  <- factor(ALL_NK@meta.data$case_control)
ALL_NK@meta.data$case_control  <- droplevels(ALL_NK@meta.data$case_control )

# Change to the new group identitites
Idents(ALL_NK) <- ALL_NK@meta.data$case_control


group <- "group" 
NK_subtype <- "predicted_NK_type_de.n350"  

# Create a new metadata column that concatenates values from both columns
ALL_NK$NK_subtype_per_group <- paste(ALL_NK@meta.data[[NK_subtype]], 
                                     ALL_NK@meta.data[[group]],
                                      sep = "_")

# Create a new metadata column that concatenates values from both columns
ALL_NK$group_per_NK_subtype <- paste(ALL_NK@meta.data[[group]],
                                     ALL_NK@meta.data[[NK_subtype]], 
                                      sep = "_")

# Check if the new column is added
ALL_NK@meta.data$NK_subtype_per_group <- as.factor(ALL_NK@meta.data$NK_subtype_per_group)
ALL_NK@meta.data$group_per_NK_subtype <- as.factor(ALL_NK@meta.data$group_per_NK_subtype)
table(ALL_NK@meta.data$NK_subtype_per_group)
table(ALL_NK@meta.data$group_per_NK_subtype)
```

```{r eval=FALSE, include=FALSE}
## add NK class prediction!!!
metadata <- read.csv(file = "/home/rstudio/mnt_data/nb_tumors/fetahu2023.scRNA/metadata/fetahu2023_scRNA_seq_metadata.csv",
                     sep = "\t")
metadata_unique <- metadata %>% distinct(sample, .keep_all = TRUE)

table(metadata_unique$group)

# ATRX NK cells
old_metadata <- ATRX_NK@meta.data
new_metadata <- old_metadata %>%
  rownames_to_column("cell_id") %>%    # Temporarily store cell names as a column
  left_join(metadata_unique, by = "sample") %>%  # Merge by sample
  column_to_rownames("cell_id")        # Restore cell IDs as row names

ATRX_NK@meta.data <- new_metadata

# ALL NK cells
old_metadata <- ALL_NK@meta.data
new_metadata <- old_metadata %>%
  rownames_to_column("cell_id") %>%    # Temporarily store cell names as a column
  left_join(metadata_unique, by = "sample") %>%  # Merge by sample
  column_to_rownames("cell_id")        # Restore cell IDs as row names

ALL_NK@meta.data <- new_metadata

# whole object
old_metadata <- fetahu2023@meta.data
new_metadata <- old_metadata %>%
  rownames_to_column("cell_id") %>%    # Temporarily store cell names as a column
  left_join(metadata_unique, by = "sample") %>%  # Merge by sample
  column_to_rownames("cell_id")        # Restore cell IDs as row names

fetahu2023@meta.data <- new_metadata


# Convert sample names to a factor with the defined order
ATRX_NK@meta.data$rados_et_al_2025  <- 
  factor(ATRX_NK@meta.data$rados_et_al_2025, 
         levels = PATIENT_ORDER)
levels(ATRX_NK@meta.data$rados_et_al_2025)

ALL_NK@meta.data$rados_et_al_2025  <- 
  factor(ALL_NK@meta.data$rados_et_al_2025, 
         levels = PATIENT_ORDER)
levels(ALL_NK@meta.data$rados_et_al_2025)

fetahu2023@meta.data$rados_et_al_2025.y  <- 
  factor(fetahu2023@meta.data$rados_et_al_2025.y, 
         levels = PATIENT_ORDER)
levels(fetahu2023@meta.data$rados_et_al_2025.y)

# Convert group names to a factor with the defined order
ALL_NK@meta.data$group  <- 
  factor(ALL_NK@meta.data$group, 
         levels = GROUP_ORDER)
levels(ALL_NK@meta.data$group)

fetahu2023@meta.data$group  <- 
  factor(fetahu2023@meta.data$group, 
         levels = GROUP_ORDER)
levels(fetahu2023@meta.data$group)

ALL_NK@meta.data$NK_subtype_per_group  <- 
  factor(ALL_NK@meta.data$NK_subtype_per_group, 
         levels = GROUP_CLUSTER_ORDER)
levels(ALL_NK@meta.data$NK_subtype_per_group)

ALL_NK@meta.data$group_per_NK_subtype  <- 
  factor(ALL_NK@meta.data$group_per_NK_subtype, 
         levels = CLUSTER_GROUP_ORDER)
levels(ALL_NK@meta.data$group_per_NK_subtype)


# 
columns_to_remove <- c("sex.x", "age_at_Dx_months.x", "diagnosis", "stage.x",
                       "MYCN", "ATRX", "group", "patient_ID",
                       "OMICS_patient_ID", "OMICS_data_ID", "Clinical_trial_ID", 
                       "TU_DNA_available", "TU_piece_available",
                       "BM_DNA_available", "germline_DNA_available",
                       "germline_ID", "IF", "File.name")

# Drop the unwanted columns while preserving row names (cell IDs)
old_metadata <- ALL_NK@meta.data 
new_metadata <- old_metadata %>%
  select(-all_of(columns_to_remove))
ALL_NK@meta.data <- new_metadata


# save Rds objects
saveRDS(ATRX_NK, 
        file = paste0(params$out,"Rds/fetahu2023_ATRXdel_BM_NK.Rds"))
saveRDS(ALL_NK, 
        file = paste0(params$out,"Rds/ALL_NK_cluster_withNewMetadata.Rds"))
saveRDS(fetahu2023, 
        file = paste0(params$out,"Rds/fetahu2023_withNewMetadata.Rds"))
```

```{r eval=FALSE, include=FALSE}
#remotes::install_github("10xGenomics/loupeR")
#loupeR::setup()

library("loupeR")

create_loupe_from_seurat(
    ATRX_NK,
    output_dir = paste0(params$out,"Rds"),
    output_name = "fetahu2023_ATRX_NK",
    dedup_clusters = FALSE,
    feature_ids = NULL,
    executable_path = NULL,
    force = FALSE
    )

create_loupe_from_seurat(
    ALL_NK,
    output_dir = paste0(params$out,"Rds"),
    output_name = "fetahu2023_ALL_NK",
    dedup_clusters = FALSE,
    feature_ids = NULL,
    executable_path = NULL,
    force = FALSE
    )

for_MaRa <- fetahu2023
for_MaRa@meta.data <- for_MaRa@meta.data[, c("orig.ident", "sample", "histology.x",
                                             "at_diagnosis.x", "at_relapse.x",
                                             "sex","rados_et_al_2025.x","group")]

create_loupe_from_seurat(
    for_MaRa,
    output_dir = paste0(params$out,"Rds"),
    output_name = "fetahu2023",
    dedup_clusters = FALSE,
    feature_ids = NULL,
    executable_path = NULL,
    force = FALSE
    )
```

```{r eval=FALSE, include=FALSE}
Idents(fetahu2023) <- fetahu2023$orig.ident
NB_cells <- subset(fetahu2023, idents = "NB (8)")

# save Robject
saveRDS(NB_cells, file = paste0(params$out,"Rds/fetahu2023_NB_cells.Rds"))
```


```{r}
NB_cells <- readRDSfile = paste0(params$out,"Rds/fetahu2023_NB_cells.Rds")()
```

# ___1. INTRO/LABEL FIGURE___

```{r}
library(reshape2)
cell_prop_per_group <- table(fetahu2023$orig.ident,fetahu2023$group)
cell_prop_per_group <- cell_prop_per_group[-20,] # remove NB tumor cluster
cell_prop_per_group_percent <- apply(cell_prop_per_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

cell_prop_per_group_percent_trans <- melt(cell_prop_per_group_percent) 
data_wide <- t(cell_prop_per_group_percent_trans)
```

```{r}
cell_prop_per_group_percent_trans$Var2 <- factor(cell_prop_per_group_percent_trans$Var2)
cell_prop_per_group_percent_trans <- cell_prop_per_group_percent_trans %>% arrange(Var2)


barplot <-  ggplot(cell_prop_per_group_percent_trans, aes(x = Var2, 
                                                          y = value, 
                                                          fill = Var1)) + 
  geom_bar(stat = "identity", 
           position = "stack") +
  theme_void() + 
  labs(x = "Clusters", y = "Percentage of cells assigned to each cluster") +  
  ggtitle("") +
  theme_minimal()  + 
  scale_fill_manual(values = COLOR_CODE_UMAP_v2)

barplot
```

## ___FIgure 1, A____

```{r fig.height=5, fig.width=5}
figure_1_A <- SCpubr::do_DimPlot(fetahu2023,
                                 border.color = NA,
                                 plot_cell_borders = FALSE,
                                 reduction = "UMAP",
                                 colors.use =  COLOR_CODE_UMAP,
                                 group.by = "orig.ident",
                                 repel = TRUE,
                                 font.size = 18,
                                 label.size  = 18,
                                 plot.axes = FALSE,
                                 label = FALSE,
                                 legend.title = "Clusters",
                                 plot.title = "Fetahu et al., 2023",
                                 legend.position = "none")

figure_1_A

pdf(paste0(params$out,"figures/01_INTRO_figure_UMAP_whole_dataset.pdf"),
    height = 5, width = 5)
figure_1_A
dev.off()

png(paste0(params$out,"figures/01_INTRO_figure_UMAP_whole_dataset.png"),
    height = 500, width = 500) 
figure_1_A
dev.off()
```

```{r fig.height=5, fig.width=5}
figure_S1 <- SCpubr::do_DimPlot(fetahu2023,
                                 border.color = NA,
                                 plot_cell_borders = FALSE,
                                 reduction = "UMAP",
                                 colors.use =  COLOR_CODE_UMAP_v3,
                                 group.by = "orig.ident",
                                 repel = TRUE,
                                 font.size = 18,
                                 label.size  = 18,
                                 plot.axes = FALSE,
                                 label = FALSE,
                                 legend.title = "Clusters",
                                 plot.title = "Fetahu et al., 2023",
                                 legend.position = "none")

figure_S1

pdf(paste0(params$out,"figures/01_INTRO_figure_UMAP_whole_dataset_metaClusters.pdf"),
    height = 5, width = 5)
figure_S1
dev.off()

png(paste0(params$out,"figures/01_INTRO_figure_UMAP_whole_dataset_metaClusters.png"),
    height = 500, width = 500) 
figure_S1
dev.off()
```


```{r fig.height=5, fig.width=5}
figure_1_B <-SCpubr::do_DimPlot(sample = ALL_NK,
           plot.title = "Fetahu et al., 2023",
           plot.subtitle = "NK cluster",
           reduction = "UMAP",
           group.by =  "case_control",
           colors.use = c("NB" = "#b4d1ee",
                          "CTRL" = "#6c77c1"),
           shuffle = TRUE,
           repel = TRUE,
           plot_density_contour = FALSE,
           label.size = 16,
           font.size = 14,
           plot.axes = TRUE,
           pt.size = 1,
           legend.title = "Groups",
           legend.position = "right",
           raster.dpi = 300,
           border.color = NA,
           plot_cell_borders = FALSE)

figure_1_B

pdf(paste0(params$out,"figures/01_INTRO_figure_UMAP_NK_cluster.pdf"),
    height = 5, width = 5)
figure_1_B
dev.off()

png(paste0(params$out,"figures/01_INTRO_figure_UMAP_NK_cluster.png"),
    height = 500, width = 500)
figure_1_B
dev.off()
```

```{r fig.height=10, fig.width=10}
figure_1_C <- do_DimPlot(sample = ALL_NK,
           plot.title = "Fetahu et al., 2023",
           plot.subtitle = "NK cluster",
           reduction = "UMAP",
           colors.use = COLOR_CODE_v2,
           border.color = NA,
           group.by = "predicted_NK_type_de.n350",
           split.by = "group",
           shuffle = TRUE,
           repel = TRUE,
           plot_density_contour = FALSE,
           label.size = 20,
           font.size = 18,
           plot.axes = TRUE,
           pt.size = 1,
           legend.title = "Clusters",
           legend.position = "right",
           raster.dpi = 300,
           plot_cell_borders = FALSE)

figure_1_C

pdf(paste0(params$out,"figures/01_INTRO_figure_UMAP_NK_cluster_per_group.pdf"),
    height = 10, width = 12)
figure_1_C
dev.off()

png(paste0(params$out,"figures/01_INTRO_figure_UMAP_NK_cluster_per_group.png"),
    height = 1000, width = 1200)
figure_1_C
dev.off()
```

## ___Figure S1, B___

```{r}
# Step 1: Ensure Var2 is correctly converted into a factor with the desired order
percent_NK_samples_trans$Var2 <- factor(percent_NK_samples_trans$Var2, levels = PATIENT_ORDER)

# Arrange the data frame by the ordered Var2 column
percent_NK_samples_trans <- percent_NK_samples_trans %>% 
  arrange(Var2,value)

barplot_all_samples <- ggplot(percent_NK_samples_trans, aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each sub-cluster", 
       fill = "NK clusters") +  
  ggtitle("") +
  theme_minimal(base_size = 22) +
  scale_fill_manual(values = COLOR_CODE_v2)

barplot_all_samples

pdf(paste0(params$out,"figures/01_SUPP_barplot_NK_subtypes_for_all_samples.pdf"), 
    height = 10, width = 10)
barplot_all_samples
dev.off()
```


## ___Stats for ALL clusters___

```{r fig.height=10, fig.width=10}
percent_cell_type_group <- table(TME_cells$cell_type,
                          TME_cells$group)

# Transform this data in %
data_percentage <- apply(percent_cell_type_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

percent_cell_type_group_trans <- melt(data_percentage) 

barplots_cell_types <- ggplot(percent_cell_type_group_trans, 
                             aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of cells assigned to each cluster", 
       fill = "Clusters") +  
  ggtitle("") +
  theme_minimal(base_size = 22) + 
  theme(axis.text.x = element_text(angle = 45,

                                   hjust = 1)) +
  scale_fill_manual(values = COLOR_CODE_FETAHU2023)

barplots_cell_types

pdf(paste0(params$out,"figures/barplot_ALL_cell_types_per_NB_group.pdf"),
    height = 10, width = 10)
barplots_cell_types
dev.off()


```

## ___Stats for NK clusters___

```{r}
table_NK_per_sample <- as.data.frame(data_percentage_sample)
table_NK_per_sample$group <- gsub("_[0-9][0-9]", "", table_NK_per_sample$Var2)
table_NK_per_sample
```

### ___hNK_Bm1___

```{r}
hNK_Bm1 <- filter(as.data.frame(table_NK_per_sample), Var1 == "hNK_Bm1")

# Perform ANOVA testing
hNK_Bm1_anova <- aov(Freq ~ group, data = hNK_Bm1)
summary(hNK_Bm1_anova)

# Extract p-value
p_value <- summary(hNK_Bm1_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm1, 
                                      group %in% c("ATRXdel", "ATRXwtMYCNwt")))$p.value
p_value2 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm1, 
                                      group %in% c("ATRXdel", "ctrl")))$p.value
p_value3 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm1, 
                                      group %in% c("ATRXdel", "MYCNamp")))$p.value

# Create boxplot and add p-value annotation
boxplot1 <- hNK_Bm1 %>%
  ggplot( aes(x = group , y = Freq, fill = group)) +
    geom_boxplot() +
    theme_minimal(legend.position="none",
          axis.text = element_text(size = 18),
          axis.title.y = element_text(size = 18),
          plot.title = element_text(size = 18)) +
    ggtitle("NK cluster per NB genetic group: hNK_Bm1") +
    xlab("") + ylab("Number of cells per group") +
  geom_signif(comparisons = list(c("ATRXdel", "ctrl"),
                                 c("ATRXdel", "MYCNamp"),
                                 c("ATRXdel", "ATRXwtMYCNwt")),
              map_signif_level = FALSE, 
              y_position = c(max(hNK_Bm1$Freq) + 50, 
                             max(hNK_Bm1$Freq) + 65, 
                             max(hNK_Bm1$Freq) + 80)) +
  scale_fill_manual(values = COLOR_CODE_GROUP_v1)

```

```{r}
boxplot1

pdf(paste0(params$out,"Rds/barplot_hNK_Bm1_wStats.pdf"))
boxplot1
dev.off()
```

### ___hNK_Bm2___

```{r}
hNK_Bm2 <- filter(as.data.frame(table_NK_per_sample), Var1 == "hNK_Bm2")

# Perform ANOVA
hNK_Bm2_anova <- aov(Freq ~ group, data = hNK_Bm2)
summary(hNK_Bm2_anova)

# Extract p-value
p_value <- summary(hNK_Bm2_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm2, 
                                      group %in% c("ATRXdel", "ATRXwtMYCNwt")))$p.value
p_value2 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm2, 
                                      group %in% c("ATRXdel", "ctrl")))$p.value
p_value3 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm2, 
                                      group %in% c("ATRXdel", "MYCNamp")))$p.value

# plot boxplot
boxplot2 <- hNK_Bm2 %>%
  ggplot( aes(x = group , y = Freq, fill = group)) +
    geom_boxplot() +
    theme_minimal(legend.position="none",
          axis.text = element_text(size = 18),
          axis.title.y = element_text(size = 18),
          plot.title = element_text(size = 18)) +
    ggtitle("NK cluster per NB genetic group: hNK_Bm2") +
    xlab("") + ylab("Number of cells per group") +
  geom_signif(comparisons = list(c("ATRXdel", "ctrl"),
                                 c("ATRXdel", "MYCNamp"),
                                 c("ATRXdel", "ATRXwtMYCNwt")),
              map_signif_level = FALSE, 
              y_position = c(max(hNK_Bm2$Freq) + 50, 
                             max(hNK_Bm2$Freq) + 80, 
                             max(hNK_Bm2$Freq) + 110))  +
  scale_fill_manual(values = COLOR_CODE_GROUP_v1)
```

```{r}
boxplot2

pdf(paste0(params$out,"Rds/barplot_hNK_Bm2_wStats.pdf"))
boxplot3
dev.off()
```

### ___hNK_Bm3___

```{r}
hNK_Bm3 <- filter(as.data.frame(table_NK_per_sample), Var1 == "hNK_Bm3")

# Perform ANOVA
hNK_Bm3_anova <- aov(Freq ~ group, data = hNK_Bm3)
summary(hNK_Bm3_anova)

# Extract p-value
p_value <- summary(hNK_Bm3_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm3, 
                                      group %in% c("ATRXdel", "ATRXwtMYCNwt")))$p.value
p_value2 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm3, 
                                      group %in% c("ATRXdel", "ctrl")))$p.value
p_value3 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm3, 
                                      group %in% c("ATRXdel", "MYCNamp")))$p.value

# plot boxplot
boxplot3 <- hNK_Bm3 %>%
  ggplot( aes(x = group , y = Freq, fill = group)) +
    geom_boxplot() +
    theme_minimal(legend.position="none",
          axis.text = element_text(size = 18),
          axis.title.y = element_text(size = 18),
          plot.title = element_text(size = 18)) +
    ggtitle("NK cluster per NB genetic group: hNK_Bm3") +
    xlab("") + ylab("Number of cells per group")  +
  geom_signif(comparisons = list(c("ATRXdel", "ctrl"),
                                 c("ATRXdel", "MYCNamp"),
                                 c("ATRXdel", "ATRXwtMYCNwt")),
              map_signif_level = FALSE, 
              y_position = c(max(hNK_Bm3$Freq) + 50, 
                             max(hNK_Bm3$Freq) + 80, 
                             max(hNK_Bm3$Freq) + 110)) +
  scale_fill_manual(values = COLOR_CODE_GROUP_v1)
```

```{r}
boxplot3

pdf(paste0(params$out,"Rds/barplot_hNK_Bm3_wStats.pdf"))
boxplot3
dev.off()
```

### ___hNK_Bm4___

```{r}
hNK_Bm4 <- filter(as.data.frame(table_NK_per_sample), Var1 == "hNK_Bm4")

# Perform ANOVA
hNK_Bm4_anova <- aov(Freq ~ group, data = hNK_Bm4)
summary(hNK_Bm4_anova)

# Extract p-value
p_value <- summary(hNK_Bm4_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm4, 
                                      group %in% c("ATRXdel", "ATRXwtMYCNwt")))$p.value
p_value2 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm4, 
                                      group %in% c("ATRXdel", "ctrl")))$p.value
p_value3 <- wilcox.test(Freq ~ group, 
                        data = subset(hNK_Bm4, 
                                      group %in% c("ATRXdel", "MYCNamp")))$p.value

# plot boxplot
boxplot4 <- hNK_Bm4 %>%
  ggplot( aes(x = group , y = Freq, fill = group)) +
    geom_boxplot() +
    theme_minimal(legend.position="none",
          axis.text = element_text(size = 18),
          axis.title.y = element_text(size = 18),
          plot.title = element_text(size = 18)) +
    ggtitle("NK cluster per NB genetic group: hNK_Bm4") +
    xlab("") + ylab("Number of cells per group")  +
  geom_signif(comparisons = list(c("ATRXdel", "ctrl"),
                                 c("ATRXdel", "MYCNamp"),
                                 c("ATRXdel", "ATRXwtMYCNwt")),
              map_signif_level = FALSE,
              y_position = c(max(hNK_Bm4$Freq) + 5, 
                             max(hNK_Bm4$Freq) + 8, 
                             max(hNK_Bm4$Freq) + 11))  +
  scale_fill_manual(values = COLOR_CODE_GROUP_v1)
```

```{r}
boxplot4

pdf(paste0(params$out,"Rds/barplot_hNK_Bm4_wStats.pdf"))
boxplot4
dev.off()
```



## ___DA testing___

```{r}
ALL_NK@meta.data$rados_et_al_2025.x <- factor(ALL_NK@meta.data$rados_et_al_2025.x)
ALL_NK@meta.data$predicted_NK_type_de.n350 <- factor(ALL_NK@meta.data$predicted_NK_type_de.n350)
```

If comparing more than two sample / groups, use DA testing with contrasts
MiloR doesn't use the "original" clusters but rather creates neighborhoods
based on similar logFC (note the fold change here is the number of cells
per cell type)

```{r miloR prepare data}
library("statmod")
library("miloR")
library("SingleCellExperiment")
library("scater")
library("scran")
library("dplyr")
library("patchwork")

rna_milo <- Milo(as.SingleCellExperiment(ALL_NK))
rna_milo

reducedDim(rna_milo, "UMAP") <- reducedDim(as.SingleCellExperiment(ALL_NK), 
                                           "UMAP")
plotUMAP(rna_milo, 
         colour_by="group") + 
  plotUMAP(rna_milo, 
           colour_by="rados_et_al_2025.x")

# Define cell neighbourhoods
rna_milo <- buildGraph(rna_milo, 
                       k = 100,
                       d = 30)

## Constructing kNN graph with k = 10
rna_milo <- makeNhoods(rna_milo, 
                       prop = 0.1, 
                       k = 100,
                       d = 30, 
                       refined = TRUE, 
                       refinement_scheme = "graph") # make nhoods using graph-only as this is faster

# metadata
colData(rna_milo)$patients <- paste(colData(rna_milo)$group, 
                                    colData(rna_milo)$rados_et_al_2025.x, sep="_")

# Counting cells in neighbourhoods
rna_milo <- countCells(rna_milo, meta.data = data.frame(colData(rna_milo)), 
                       samples="patients") # make the nhood X sample counts matrix


# Defining experimental design
rna_design <- data.frame(colData(rna_milo))[,c("patients", "group")]
rna_design <- distinct(rna_design)
rownames(rna_design) <- rna_design$patients
rna_design

# Computing neighborhood connectivity
rna_milo <- calcNhoodDistance(rna_milo, 
                              d = 30, 
                              reduced.dim = "PCA")

# DA testing
da_results <- testNhoods(rna_milo, 
                         design = ~ 0 + group,
                         design.df = rna_design, 
                         reduced.dim="PCA")
head(da_results)

da_results %>%
  arrange(SpatialFDR) %>%
  head() 

rna_milo <- buildNhoodGraph(rna_milo)


# QUALITY CHECK OF THE DIFFERENTIAL ABUNDANCE TESTING

# Plot single-cell UMAP
umap_pl <- plotReducedDim(rna_milo, 
                          dimred = "UMAP", 
                          colour_by = "group", 
                          text_by = "predicted_NK_type_de.n350", 
                          text_size = 3, point_size = 0.5,) +
  guides(fill = "none")

rna_milo <- buildNhoodGraph(rna_milo)

# Plot neighbourhood graph
nh_graph_pl <- plotNhoodGraphDA(rna_milo, 
                                da_results, 
                                layout="UMAP", 
                                alpha=0.05) 


umap_pl + nh_graph_pl +
  plot_layout(guides = "collect")

# Differential abundance testing with contrasts
rna_design <- data.frame(colData(rna_milo))[,c("patients", "group")]
rna_design <- distinct(rna_design)
rownames(rna_design) <- rna_design$patients

# Reorder rownames to match columns of nhoodCounts(milo)
rna_design <- rna_design[colnames(nhoodCounts(rna_milo)), , drop=FALSE]
table(rna_design$group)

saveRDS(rna_milo, file = paste0(params$out,"Rds/miloR_NK_cells.Rds"))
```

```{r}
####### Comparisons #######
levels(ALL_NK$group)

COLOR_CODE_miloR <- c( "hNK_Bm1" ="#161B33",
                       "hNK_Bm1" = "#419D78",
                       "hNK_Bm3" = "#AF1B3F",
                       "hNK_Bm4" = "#E88873",
                       "Mixed" = "darkgray")
```

## ___A. ATRXdel vs. ATRXwt (MYCNamp and ATRXwtMYCNwt)___

```{r}
model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupATRXdel - ((groupMYCNamp + groupATRXwtMYCNwt) / 2)")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXdel_vs_ATRXwt <- testNhoods(rna_milo, 
                                      design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXdel_vs_ATRXwt$SpatialFDR < 0.1)
table(da_results_ATRXdel_vs_ATRXwt$SpatialFDR < 0.05)


# annotate hood with original cell type annotation
da_results_ATRXdel_vs_ATRXwt <- annotateNhoods(rna_milo, 
                                          da_results_ATRXdel_vs_ATRXwt, 
                                          coldata_col = "predicted_NK_type_de.n350")

# remove hoods that contained mixed cell type populations
da_results_ATRXdel_vs_ATRXwt$predicted_NK_type_de.n350 <- 
  ifelse(da_results_ATRXdel_vs_ATRXwt$predicted_NK_type_de.n350_fraction < 0.7, 
         "Mixed", da_results_ATRXdel_vs_ATRXwt$predicted_NK_type_de.n350)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXdel_vs_ATRXwt %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

da_results_ATRXdel_vs_ATRXwt$predicted_NK_type_de.n350 <- as.factor(da_results_ATRXdel_vs_ATRXwt$predicted_NK_type_de.n350)

DA_plot_ATRXdel_vs_ATRXwt <- plotDAbeeswarm(da_results_ATRXdel_vs_ATRXwt, 
               group.by = "predicted_NK_type_de.n350", 
               alpha = 0.1) # cut-off for spatial FDR

DA_plot_ATRXdel_vs_ATRXwt

pdf(paste0(params$out,"figures/DA_plot_NK_clusters_ATRXdel_vs_ATRXwt.pdf"),
    width =10, height = 6)
DA_plot_ATRXdel_vs_ATRXwt
dev.off()
```

## ___B. ATRXdel vs. ALL (MYCNamp + ATRXwtMYCNwt + control)___

```{r}
model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupATRXdel - ((groupMYCNamp + groupATRXwtMYCNwt + groupcontrol) / 3)")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXdelvsAll <- testNhoods(rna_milo, design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXdelvsAll$SpatialFDR < 0.1)

# annotate hood with original cell type annotation
da_results_ATRXdelvsAll <- annotateNhoods(rna_milo, da_results_ATRXdelvsAll, 
                             coldata_col = "predicted_NK_type_de.n350")

# remove hoods that contained mixed cell type populations
da_results_ATRXdelvsAll$predicted_NK_type_de.n350 <- 
  ifelse(da_results_ATRXdelvsAll$predicted_NK_type_de.n350_fraction < 0.7, 
         "Mixed", da_results_ATRXdelvsAll$predicted_NK_type_de.n350)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXdelvsAll %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results


DA_plot_ATRXdelvsAll <- plotDAbeeswarm(da_results_ATRXdelvsAll, 
               group.by = "predicted_NK_type_de.n350", 
               alpha = 0.1) # cut-off for spatial FDR

DA_plot_ATRXdelvsAll

pdf(paste0(params$out,"figures/DA_plot_NK_clusters_ATRXdelvsAll.pdf"),
    width =10, height = 6)
DA_plot_ATRXdelvsAll
dev.off()
```

## ___C. CTRL vs. NB (MYCNamp and ATRXwtMYCNwt and ATRXdel)___

```{r}
model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupcontrol - ((groupMYCNampmp + groupATRXwtMYCNwt + groupATRXdel) / 3)")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXdelvsAll <- testNhoods(rna_milo, design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXdelvsAll$SpatialFDR < 0.1)

# annotate hood with original cell type annotation
da_results_ATRXdelvsAll <- annotateNhoods(rna_milo, da_results_ATRXdelvsAll, 
                             coldata_col = "predicted_NK_type_de.n350")

# remove hoods that contained mixed cell type populations
da_results_ATRXdelvsAll$predicted_NK_type_de.n350 <- 
  ifelse(da_results_ATRXdelvsAll$predicted_NK_type_de.n350_fraction < 0.7, 
         "Mixed", da_results_ATRXdelvsAll$predicted_NK_type_de.n350)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXdelvsAll %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

plotDAbeeswarm(da_results_ATRXdelvsAll, 
               group.by = "predicted_NK_type_de.n350", 
               alpha=0.1) # alfa is the significance value for spatial FDR
```

# ___3. DEG FIGURE___

```{r fig.height=5, fig.width=10}
genes.to.plot <- c("CD6","IFNG","TNF")

Figure_3_C <- SCpubr::do_FeaturePlot(sample = ALL_NK,
                       reduction = "UMAP",
                       features = genes.to.plot,
                       order = TRUE,
                       label = FALSE,
                       font.size = 16,
                       pt.size = 1,
                       plot.title = "RNA expression",
                       ncol = 3,
                       number.breaks = 10,
                       individual.subtitles = c("\t \t \t \t CD6", 
                                                "\t \t \t \t IFNG", 
                                                "\t \t \t \t TNF"),
                       legend.length = 10,
                       border.color = NA,
                       legend.title = "Scaled RNA expression",
                       plot.axes = TRUE,
                       slot = "scale.data",
                       min.cutoff = c(0,0,0),
                       max.cutoff = c(3,3,3)
                       )

Figure_3_C


```

## ___per CASE/CONTROL group___

```{r}
library(reshape2)
percent_NK_group <- table(ALL_NK$predicted_NK_type_de.n350,
                          ALL_NK$case_control)

# Transform this data in %
data_percentage <- apply(percent_NK_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})


percent_NK_group_trans <- melt(data_percentage) 

barplots_NK_case_control <- ggplot(percent_NK_group_trans, 
                                   aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each cluster", 
       fill = "NK clusters") +  
  ggtitle("") +
  theme_minimal(base_size = 18) +
  scale_fill_manual(values = COLOR_CODE_v2)

barplots_NK_case_control

pdf(paste0(params$out,"figures/barplot_NK_subtypes_per_case_control.pdf"))
barplots_NK_case_control
dev.off()
```

## ___per NB group___

```{r}
percent_NK_group <- table(ALL_NK$predicted_NK_type_de.n350,
                          ALL_NK$group)

# Transform this data in %
data_percentage <- apply(percent_NK_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

percent_NK_group_trans <- melt(data_percentage) 

barplots_NK_groups <- ggplot(percent_NK_group_trans, 
                             aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each cluster", 
       fill = "NK clusters") +  
  ggtitle("") +
  theme_minimal(base_size = 18) + 
  theme(axis.text.x = element_text(angle = 45,

                                   hjust = 1)) +
  scale_fill_manual(values = COLOR_CODE_v2)

barplots_NK_groups

pdf(paste0(params$out,"figures/barplot_NK_subtypes_per_NB_group.pdf"))
barplots_NK_groups
dev.off()
```

## ___per patient___

```{r}
Idents(ALL_NK) <- ALL_NK@meta.data$rados_et_al_2025.x

data_sample <- table(ALL_NK$predicted_NK_type_de.n350,
                          ALL_NK$rados_et_al_2025.x)

# Transform this data in %
data_percentage <- apply(data_sample, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

# all samples
percent_NK_samples_trans <- melt(data_percentage) 

# Step 1: Reshape the data into wide format for clustering
data_wide <- t(data_percentage)
```

## ___Samples clustered based on NK cell subtype composition___

```{r}
# Compute the Euclidean distance matrix
d <- dist(data_wide, method = "euclidean")

# Perform hierarchical clustering using complete linkage
hc <- hclust(d, method = "complete")

# Step 3: Reorder the samples based on clustering
ordered_samples <- rownames(data_wide)[hc$order]

# Arrange the data frame by the ordered Var2 column
percent_NK_samples_trans$Var2 <- factor(percent_NK_samples_trans$Var2, 
                                        levels = ordered_samples)

percent_NK_samples_trans <- percent_NK_samples_trans %>% arrange(Var2)

# add dendogram
library(ggdendro)

# Convert hc to dendrogram object
dendrogram_data <- dendro_data(hc)

# Create Dendrogram Plot
dendrogram_plot <- ggplot() + 
  geom_segment(data = segment(dendrogram_data), 
               aes(x = x, y = y, xend = xend, yend = yend)) +
  theme_minimal() +
  labs(y = "Distance", x = NULL)


# Step 4: Plot the clustered bar plot
barplot <-  ggplot(percent_NK_samples_trans, aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each sub-cluster", 
       fill = "NK clusters") +
  theme_minimal(base_size = 18) +
  scale_fill_manual(values = COLOR_CODE_v2)

# Combine Plots: Stack Dendrogram on Top of the Bar Plot
final_plot <- dendrogram_plot / barplot
final_plot
```



# ___5. SECRETION FIGURE___

## ___Figure 5, C___

### ___NK cells____

#### ___Group per NK subtype___

```{r fig.height=10, fig.width=25}
## ___ALL___
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = secreted_factors,
           plot.title = "Secreted NK factors",
           plot.subtitle = "All groups",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

```{r fig.height=10, fig.width=14}
## ___ALL___
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = secreted_factors_high,
           plot.title = "Secreted NK factors",
           plot.subtitle = "All groups",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

```{r  fig.height=10, fig.width=18}
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = secreted_factors_low,
           plot.title = "Secreted NK factors",
           plot.subtitle = "All groups",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

#### ___NK subtype per group___

```{r fig.height=10, fig.width=25}
## ___ALL___
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = secreted_factors,
           plot.title = "Secreted NK factors",
           plot.subtitle = "All groups",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

```{r fig.height=10, fig.width=14}
## ___ALL___
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = secreted_factors_high,
           plot.title = "Secreted NK factors",
           plot.subtitle = "All groups",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

```{r  fig.height=10, fig.width=18}
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = secreted_factors_low,
           plot.title = "Secreted NK factors",
           plot.subtitle = "All groups",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

#### ___ATRX clusters 2&3 only___

```{r fig.height=4, fig.width=20}
do_DotPlot(sample = ATRX_NK,
           axis.text.x.angle = 45,
           features = secreted_factors,
           plot.title = "Secreted NK factors",
           plot.subtitle = "ATRXdel",
           scale = FALSE,
           legend.length = 5,
           number.breaks = 7,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

### ___NB tumor cells____

```{r fig.height=4, fig.width=20}
do_DotPlot(sample = NB_cells,
           axis.text.x.angle = 45,
           features = secreted_factors,
           plot.title = "Secreted NB factors",
           plot.subtitle = "Tumor cells per group",
           scale = TRUE,
           legend.length = 5,
           group.by = "group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

```{r fig.height=4, fig.width=24}
do_DotPlot(sample = NB_cells,
           axis.text.x.angle = 45,
           features = secreted_factors_MinusMIF,
           plot.title = "Secreted NB factors",
           plot.subtitle = "Tumor cells per group",
           scale = TRUE,
           legend.length = 5,
           group.by = "group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```
### ___Only secreted factors in the coculture experiments___

```{r fig.height=10, fig.width=12}

# NK cells
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = coculture_secreted_factors,
           plot.title = "Secreted NK factors",
           plot.subtitle = "NK clusters per group",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```


```{r fig.height=5, fig.width=12}
# NK cells for ATRXdel samples only
do_DotPlot(sample = ATRX_NK,
           axis.text.x.angle = 45,
           features = coculture_secreted_factors,
           plot.title = "Secreted NK factors",
           plot.subtitle = "ATRXdel samples",
           scale = FALSE,
           legend.length = 5,
           group.by = "predicted_NK_type_de.n350",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```


```{r fig.height=4, fig.width=12}
# NB cells
do_DotPlot(sample = NB_cells,
           axis.text.x.angle = 45,
           features = coculture_secreted_factors,
           plot.title = "Secreted NB factors",
           plot.subtitle = "NB tumor cells per group",
           scale = TRUE,
           legend.length = 5,
           group.by = "group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```



# ___THE END___

```{r}
sessionInfo()
```

