---
title: "NK project: Trajectory analysis"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/CancReg/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/CancReg/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_scRNA: "/home/rstudio/mnt_out/NK_project/Rds/fetahu2023_withNewMetadata.Rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/scRNA.NB.integration/metadata/"
---

DESCRIPTION
<br>
The idea here is to see the the differentiation trajectory of healthy BM-derived
<br>
NK cells in adolescent children vs. adolescents with NB, with a focus on ATRX deleted NBs.
<br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)

print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")
```

```{r eval=FALSE, include=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
remotes::install_github("huayc09/SeuratExtend")
```

```{r include=FALSE}
library(reticulate)
library(SeuratExtend)
library(tradeSeq)
```

```{r}
# Check if conda is installed
CONDA_PATH = "/opt/conda/bin/conda"
Sys.setenv(RETICULATE_CONDA = CONDA_PATH)
reticulate::conda_version()

# Use seuratextend env for downstream processinfg
CONDA_ENV = "seuratextend"
reticulate::use_condaenv(CONDA_ENV, required = TRUE)
```

# ___Input_data___

```{r}
adult_BM_NK <- readRDS(file = params$input_BM1)
ALL_NK <-  readRDS(file = paste0(params$out,"Rds/ALL_NK_cluster_withNewMetadata.Rds"))
CTRL_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_fetal_BM_NK.Rds"))
ATRX_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXdel_BM_NK.Rds"))
ATRX_NK_sub <- readRDS(file = paste0(params$out,"Rds/fetahu2023_ATRXdel-sub_BM_NK.Rds"))
MNA_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_MNA_BM_NK.Rds"))
sporadic_NK <- readRDS(file = paste0(params$out,"Rds/fetahu2023_sporadic_BM_NK.Rds"))
```

# ___Diffusion map, pseudotime and temporal gene expression___

## ___Whole Fetahu2023 dataset - BM-derived NK cells___

### ___Palantir___

```{r}
# pre-processing
ALL_NK <- FindVariableFeatures(ALL_NK)
#ALL_NK <- RunPCA(ALL_NK)

# Difussion map
ALL_NK <- Palantir.RunDM(ALL_NK,reduction = "PCA")

# check if DM was created
ALL_NK@reductions

# View the first 2 ms dimensions:
pdf(paste0(params$out,"figures/", "ALL_NK","_Palantir.pdf"))
DimPlot2(ALL_NK, 
         reduction = "ms", 
         group.by = "predicted_NK_type_de.n350", 
         label = TRUE)
dev.off()

# Determining Cell Fates and Calculating Pseudotime

p <- DimPlot(ALL_NK, 
             reduction = "ms", 
             group.by = "predicted_NK_type_de.n350")
cells <- CellSelector(p)

ALL_NK <- Palantir.Pseudotime(ALL_NK, 
                               start_cell = cells)

ps <- ALL_NK@misc$Palantir$Pseudotime
head(ps)

# Visualize cell fate on UMAP
colnames(ps)[3:4] <- c("fate1", "fate2","fate3","fate4","fate5")
ALL_NK@meta.data[,colnames(ps)] <- ps

pdf(paste0(params$out,"figures/", "ALL_NK","_DM.pdf"))
DimPlot2(ALL_NK, 
         features = colnames(ps), 
         reduction = "ms", 
         cols = list(Entropy = "D"), 
         theme = NoAxes())
dev.off()
```

<br>
CD27⁺ITGAM⁻ → Immature NK cells  --> blue --> #3 (hNK_BM3)
<br>
CD27⁺ITGAM⁺ → Intermediate   --> red --> #1 (hNK_BM2)
<br>
CD27⁻ITGAM⁺ → Fully mature NK cells  --> green --> #2 (hNK_BM4 and hNK_BM1)
<br>
[with hNK_BM4 being slightly less mature than hNK_BM1]
<br>

### ___Slingshot___

```{r}
run_sligshot <- function(current_object, object_name, params, CONDA_ENV) {
  cat("Processing group: ", object_name, "\n") 

  set.seed(2805)
  current_object <- RunSlingshot(current_object, 
                    group.by = "predicted_NK_type_de.n350", 
                    start.clus = "hNK_Bm2")
  sling <- current_object@misc$slingshot$PCA$SlingPseudotime
  current_object@meta.data[,colnames(sling)] <- as.data.frame(sling)
  slig_data <- current_object@misc$slingshot$PCA$SlingshotDataSet
  sce <- as.SingleCellExperiment(current_object)
  #sce <- fitGAM(counts = as.matrix(GetAssayData(current_object, slot = "counts")),
  #            sds = slig_data, 
  #            nknots = 4) 
  pdf(paste0(params$out,"figures/", object_name,"_Slingshot.pdf"))
  print(DimPlot2(current_object, features = colnames(sling), 
         cols = "C", 
         theme = NoAxes()))
  dev.off()
  
  genes <-  c("ID2", "EOMES", "TBX21", "TOX", "GATA3", "ZNF683", "NCR1", "KLRG1", 
          "CD27", "ITGAM", "KLRC1", "KIR2DL1", "KIR2DL3", "KIR3DL1", "KLRD1", 
          "IL2RB", "IFNGR1","IL15", "IL15RA", "IL18R1", "IL18RAP", "IL12RB2", 
          "CX3CR1", "CXCR4", "GZMB", "PRF1", "IFNG", "TNF", "LAMP1", "SIRT1", 
           "EZH2")
  
  xlim <- c(0, 120)
  ylim <- c(0, 3) 
  
  tempExpression <- GeneTrendCurve.Slingshot(current_object, 
                         features = genes,
                         ncol = 4,
                         conda_env = CONDA_ENV)
  
  tempExpression_final <- tempExpression + coord_cartesian(xlim = xlim, 
                                                           ylim = ylim)
  print(tempExpression_final)
  
  pdf(paste0(params$out,"figures/", object_name,"_temporal_geneExpress.pdf"),
      height = 20, width = 10)
  print(tempExpression_final)
  dev.off()
  
  # developmental NK genes 
  genes <- c("ID2" , "RUNX3" , 
           "NFIL3", "ETS1",  
           "GATA2","EOMES", 
           "TBX21",  "ZEB2")
  
  xlim <- c(0, 120) 
  ylim <- c(0, 2) 
  tempExpression <- GeneTrendCurve.Slingshot(current_object, 
                         features = genes,
                         ncol = 2,
                         conda_env = CONDA_ENV)
  tempExpression_final <- tempExpression + coord_cartesian(xlim = xlim, 
                                                           ylim = ylim)
  print(tempExpression_final)
  
  pdf(paste0(params$out,"figures/", object_name,"_tempExp_NK_development.pdf"),
    height = 10, width = 5)
  print(tempExpression_final)
  dev.off()
  
  genes <- c("CAPG" , "CXCR3" , 
           "GPR183", "IL7R",  
           "NCF1", "SCML1", 
           "YBX3",  "COTL1",
           "DUSP4","GZMK",
           "LTB","PPP1R14B","XCL1")
  
  xlim <- c(0, 120) 
  ylim <- c(0, 2) 
  tempExpression <- GeneTrendCurve.Slingshot(current_object, 
                         features = genes,
                         ncol = 4,
                         conda_env = CONDA_ENV)
  
  tempExpression_final <- tempExpression + coord_cartesian(xlim = xlim, 
                                                           ylim = ylim)
  print(tempExpression_final)

  pdf(paste0(params$out,"figures/", object_name,"_tempExp_NK_development_v2.pdf"),
    height = 10, width = 10)
  print(tempExpression_final)
  dev.off()
}
```

```{r}
groups <- c(ALL_NK = ALL_NK,
            CTRL_NK = CTRL_NK,
            ATRX_NK_sub = ATRX_NK_sub,
            MNA_NK = MNA_NK,
            sporadic_NK = sporadic_NK)

output <- list() 

for (object_name in names(groups)) {
    current_object <- groups[[object_name]]

    processed_object <- run_sligshot(
        current_object = current_object,
        object_name = object_name,
        params = params,
        CONDA_ENV = CONDA_ENV)

    output[[object_name]] <- processed_object
}
```

# ___Palantir analysis___
## ___Healthy adolescent BM-derived NK cells___

```{r}
# pre-processing
CTRL_NK <- FindVariableFeatures(CTRL_NK)
CTRL_NK <- RunPCA(CTRL_NK)

# Difussion map
CTRL_NK <- Palantir.RunDM(CTRL_NK)

# check if DM was created
CTRL_NK@reductions

# View the first 2 ms dimensions:
pdf(paste0(params$out,"figures/", "CTRL_NK","_Palantir.pdf"))
DimPlot2(CTRL_NK, 
         reduction = "ms", 
         group.by = "predicted_NK_type_de.n350", 
         label = TRUE)
dev.off()

# Determining Cell Fates and Calculating Pseudotime

p <- DimPlot(CTRL_NK, 
             reduction = "ms", 
             group.by = "predicted_NK_type_de.n350")
cells <- CellSelector(p)

CTRL_NK <- Palantir.Pseudotime(CTRL_NK, 
                               start_cell = cells)

ps <- CTRL_NK@misc$Palantir$Pseudotime
head(ps)

# Visualize cell fate on UMAP
colnames(ps)[3:4] <- c("fate1", "fate2","fate3","fate4","fate5")
CTRL_NK@meta.data[,colnames(ps)] <- ps

pdf(paste0(params$out,"figures/", "CTRL_NK","_DM.pdf"))
DimPlot2(CTRL_NK, 
         features = colnames(ps), 
         reduction = "ms", 
         cols = list(Entropy = "D"), 
         theme = NoAxes())
dev.off()
```

<br>
CD27⁺ITGAM⁻ → Immature NK cells  →  blue →  #3 (hNK_BM3)
<br>
CD27⁺ITGAM⁺ → Intermediate →  red →  #1 (hNK_BM2)
<br>
CD27⁻ITGAM⁺ → Fully mature NK cells →  green →  #2 (hNK_BM4 and hNK_BM1)
<br>
[with hNK_BM4 being slightly less mature than hNK_BM1]
<br>


## ___ATRXdel NB patient BM-derived NK cells___

```{r}
# pre-processing
ATRX_NK <- FindVariableFeatures(ATRX_NK)
ATRX_NK <- RunPCA(ATRX_NK)

# Difussion map
ATRX_NK <- Palantir.RunDM(ATRX_NK)

# check if DM was created
ATRX_NK@reductions

# View the first 2 ms dimensions:
pdf(paste0(params$out,"figures/", "ATRX_NK","_Palantir.pdf"))
DimPlot2(ATRX_NK, 
         reduction = "ms", 
         group.by = "predicted_NK_type_de.n350", 
         label = TRUE)
dev.off()

# Determining Cell Fates and Calculating Pseudotime

p <- DimPlot(ATRX_NK, 
             reduction = "ms", 
             group.by = "predicted_NK_type_de.n350")
cells <- CellSelector(p)

ATRX_NK <- Palantir.Pseudotime(ATRX_NK, 
                               start_cell = cells)

ps <- ATRX_NK@misc$Palantir$Pseudotime
head(ps)

# Visualize cell fate on UMAP
colnames(ps)[3:4] <- c("fate1", "fate2","fate3","fate4","fate5")
ATRX_NK@meta.data[,colnames(ps)] <- ps

pdf(paste0(params$out,"figures/", "ATRX_NK","_DM.pdf"))
DimPlot2(ATRX_NK, 
         features = colnames(ps), 
         reduction = "ms", 
         cols = list(Entropy = "D"), 
         theme = NoAxes())
dev.off()
```

CD27⁻ITGAM⁺ markers indicate → Fully mature NK cells →  
but actually at the same time they resemble immature NK cells


## ___MNA NB patient BM-derived NK cells___

```{r}
# pre-processing
MNA_NK <- FindVariableFeatures(MNA_NK)
MNA_NK <- RunPCA(MNA_NK)

# Difussion map
MNA_NK <- Palantir.RunDM(MNA_NK)

# check if DM was created
MNA_NK@reductions

# View the first 2 ms dimensions:
pdf(paste0(params$out,"figures/", "MNA_NK","_Palantir.pdf"))
DimPlot2(MNA_NK, 
         reduction = "ms", 
         group.by = "predicted_NK_type_de.n350", 
         label = TRUE)
dev.off()

# Determining Cell Fates and Calculating Pseudotime

p <- DimPlot(MNA_NK, 
             reduction = "ms", 
             group.by = "predicted_NK_type_de.n350")
cells <- CellSelector(p)

MNA_NK <- Palantir.Pseudotime(MNA_NK, 
                               start_cell = cells)

ps <- MNA_NK@misc$Palantir$Pseudotime
head(ps)

# Visualize cell fate on UMAP
colnames(ps)[3] <- "fate1"
MNA_NK@meta.data[,colnames(ps)] <- ps

pdf(paste0(params$out,"figures/", "MNA_NK","_DM.pdf"))
DimPlot2(MNA_NK, 
         features = colnames(ps), 
         reduction = "ms", 
         cols = list(Entropy = "D"), 
         theme = NoAxes())
dev.off()
```


## ___sporadic NB patient BM-derived NK cells___

```{r}
# pre-processing
sporadic_NK <- FindVariableFeatures(sporadic_NK)
sporadic_NK <- RunPCA(sporadic_NK)

# Difussion map
sporadic_NK <- Palantir.RunDM(sporadic_NK)

# check if DM was created
sporadic_NK@reductions

# View the first 2 ms dimensions:
pdf(paste0(params$out,"figures/", "sporadic_NK","_Palantir.pdf"))
DimPlot2(sporadic_NK, 
         reduction = "ms", 
         group.by = "predicted_NK_type_de.n350", 
         label = TRUE)
dev.off()

# Determining Cell Fates and Calculating Pseudotime

p <- DimPlot(sporadic_NK, 
             reduction = "ms", 
             group.by = "predicted_NK_type_de.n350")
cells <- CellSelector(p)

sporadic_NK <- Palantir.Pseudotime(sporadic_NK, 
                               start_cell = cells)

ps <- sporadic_NK@misc$Palantir$Pseudotime
head(ps)

# Visualize cell fate on UMAP
colnames(ps)[3:4] <- c("fate1", "fate2","fate3","fate4","fate5")
sporadic_NK@meta.data[,colnames(ps)] <- ps

pdf(paste0(params$out,"figures/", "sporadic_NK","_DM.pdf"))
DimPlot2(sporadic_NK, 
         features = colnames(ps), 
         reduction = "ms", 
         cols = list(Entropy = "D"), 
         theme = NoAxes())
dev.off()
```


## ___Healthy ADULT BM-derived NK cells___

```{r}
# pre-processing
adult_BM_NK <- FindVariableFeatures(adult_BM_NK)
adult_BM_NK <- RunPCA(adult_BM_NK)

# Difussion map
adult_BM_NK <- Palantir.RunDM(adult_BM_NK)

# check if DM was created
adult_BM_NK@reductions

# View the first 2 ms dimensions:
pdf(paste0(params$out,"figures/", "adult_NK","_Palantir.pdf"))
DimPlot2(adult_BM_NK, 
         reduction = "ms", 
         group.by = "cluster2", 
         label = TRUE)
dev.off()

# Determining Cell Fates and Calculating Pseudotime

p <- DimPlot(adult_BM_NK, 
             reduction = "ms", 
             group.by = "cluster2")
cells <- CellSelector(p)

adult_BM_NK <- Palantir.Pseudotime(adult_BM_NK,
                              start_cell = cells)

ps <- adult_BM_NK@misc$Palantir$Pseudotime
head(ps)

# Visualize cell fate on UMAP
colnames(ps)[3:4] <- c("fate1", "fate2","fate3","fate4","fate5")
adult_BM_NK@meta.data[,colnames(ps)] <- ps

pdf(paste0(params$out,"figures/", "adult_NK","_DM.pdf"))
DimPlot2(adult_BM_NK, 
         features = colnames(ps), 
         reduction = "ms", 
         cols = list(Entropy = "D"), 
         theme = NoAxes())
dev.off()
```


# ___TSCAN Trajectory analysis___

TSCAN uses the clustering to summarize the data into a smaller set of discrete units, 
computes cluster centroids by averaging the coordinates of its member cells, 
and then forms the minimum spanning tree (MST) across those centroids. 
The MST is simply an undirected acyclic graph that passes through each centroid 
exactly once and is thus the most parsimonious structure that captures the 
transitions between clusters. 

```{r}
library(TSCAN)
library(scater)
```

## ___Whole Fetahu2023 dataset - BM-derived NK cells___

```{r eval=FALSE, include=FALSE}
sce <- as.SingleCellExperiment(ALL_NK)
colLabels(sce) <- sce$predicted_NK_type_de.n350

geneFilter <- apply(assays(sce)$counts,1,function(x){
    sum(x >= 5) >= 20
})
sce <- sce[geneFilter,]
sce

# 1.Cluster-based minimum spanning tree
by.cluster <- aggregateAcrossCells(sce, ids = sce$predicted_NK_type_de.n350)
centroids <- reducedDim(by.cluster, "UMAP")

# Set clusters=NULL as we have already aggregated above.
mst <- createClusterMST(centroids, clusters=NULL)
mst


line.data <- reportEdges(by.cluster, 
                         mst=mst, 
                         clusters=NULL, 
                         use.dimred="UMAP")

plotUMAP(sce, colour_by="predicted_NK_type_de.n350") + 
    geom_line(data=line.data, mapping=aes(x=UMAP_1, y=UMAP_2, group=edge))

#We obtain a pseudotime ordering by projecting the cells onto the MST with mapCellsToEdges(). More specifically, we move each cell onto the closest edge of the MST; the pseudotime is then calculated as the distance along the MST to this new position from a “root node” with orderCells(). For our purposes, we will arbitrarily pick one of the endpoint nodes as the root, though a more careful choice based on the biological annotation of each node may yield more relevant orderings (e.g., picking a node corresponding to a more pluripotent state).

map.UMAP <- mapCellsToEdges(sce, mst=mst, use.dimred="UMAP")
UMAP.pseudo <- orderCells(map.UMAP, mst, start = "hNK_Bm2")
head(UMAP.pseudo)

common.pseudo <- averagePseudotime(UMAP.pseudo) 
plotUMAP(sce, colour_by=I(common.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="black") +
    geom_line(data=line.data, linewidth = 2, colour = "#b4d1ee",
              mapping=aes(x=UMAP_1, y=UMAP_2, group=edge)) + 
  theme(text = element_text(size = 14,face = "bold")) + 
  scale_color_gradientn(
        colours = c("#4B0082", "darkgreen", "#CDA646"),
        values = scales::rescale(c(0, 1.5, 3)),  # center green around 1.5, covering 0–3
        limits = c(0, 3),
        oob = scales::squish
    )
```

```{r}
sce <- as.SingleCellExperiment(ALL_NK)
colLabels(sce) <- sce$predicted_NK_type_de.n350

#same as above but in one command
pseudo.all <- quickPseudotime(sce, 
                              use.dimred = "UMAP",
                              start = "hNK_Bm2")

all <- averagePseudotime(pseudo.all$ordering)
```

```{r}
pdf(paste0(params$out,"figures/", "ALL_NK","_TSCAN","_TSNE.pdf"))
plotTSNE(sce, colour_by=I(all), 
        text_by="predicted_NK_type_de.n350", text_colour="#b4d1ee") +
    geom_line(data=pseudo.all$connected$TSNE, linewidth = 1, colour = "black",
              mapping=aes(x=tSNE_1, y=tSNE_2, group=edge)) + 
  theme(text = element_text(size = 14,face = "bold")) + 
  scale_color_gradientn(
        colours = c("#4B0082", "darkgreen", "#CDA646"),
        values = scales::rescale(c(0, 1.5, 3)),  # center green around 1.5, covering 0–3
        limits = c(0, 3),
        oob = scales::squish
    )
dev.off()

pdf(paste0(params$out,"figures/", "ALL_NK","_TSCAN","_UMAP.pdf"))
plotUMAP(sce, colour_by=I(all), 
         text_by="predicted_NK_type_de.n350", text_colour="#b4d1ee") +
    geom_line(data=pseudo.all$connected$UMAP, 
              mapping=aes(x=UMAP_1, y=UMAP_2, group=edge)) + 
  theme(text = element_text(size = 14)) + 
  scale_color_gradientn(
        colours = c("#4B0082", "darkgreen", "#CDA646"),
        values = scales::rescale(c(0, 1.5, 3)),  # center green around 1.5, covering 0–3
        limits = c(0, 3),
        oob = scales::squish
    )
dev.off()

pdf(paste0(params$out,"figures/", "ALL_NK","_TSCAN","_PCA.pdf"))
plotPCA(sce, colour_by=I(all), 
        text_by="predicted_NK_type_de.n350", text_colour="#b4d1ee") +
    geom_line(data=pseudo.all$connected$PCA, 
              mapping=aes(x=pca_1, y=pca_2, group=edge)) + 
  theme(text = element_text(size = 14)) + 
  scale_color_gradientn(
        colours = c("#4B0082", "darkgreen", "#CDA646"),
        values = scales::rescale(c(0, 1.5, 3)),  # center green around 1.5, covering 0–3
        limits = c(0, 3),
        oob = scales::squish
    )
dev.off()
```

## __ATRXdel cells on top of the ALL_NK plot__

```{r fig.height=6, fig.width=7}
tsne_coords <- as.data.frame(reducedDim(sce, "TSNE"))
colnames(tsne_coords) <- c("TSNE1", "TSNE2")

cell_metadata <- as.data.frame(colData(sce))
cell_metadata$average_pseudotime <- all
plot_data <- bind_cols(tsne_coords, cell_metadata)

subset_data <- plot_data %>%
  filter(group == "ATRXdel")

trajectory_lines_data <- as.data.frame(pseudo.all$connected$TSNE)
colnames(trajectory_lines_data) <- c("edge", "TSNE1", "TSNE2")

background <- ggplot(plot_data, 
                     aes(x = TSNE1, y = TSNE2)) +
  geom_point(color = "darkgray", 
             size = 1.5, 
             alpha = 0.5)

p <- background + 
  geom_point(data = subset_data, 
             aes(color = average_pseudotime), 
             size = 1.5) +
  geom_line(data = trajectory_lines_data,
            aes(x = TSNE1, y = TSNE2, group = edge),
            linewidth = 1,
            colour = "black") +
  labs(title = "ATRXdel",
       x = "TSNE 1",
       y = "TSNE 2",
       color = "Pseudotime") + 
  scale_color_gradientn(
       colours = c("#4B0082", "darkgreen", "#CDA646"),
       values = scales::rescale(c(0, 1.5, 3)),
       limits = c(0, 3),
       oob = scales::squish) + 
  theme(text = element_text(size = 14, face = "bold"))

print(p)

pdf(paste0(params$out,"figures/TSCAN_trajectory_analysis_tsne_plot_ATRXdel.pdf"))
p
dev.off()
```

## __MYCNamp cells on top of the ALL_NK plot__

```{r fig.height=6, fig.width=7}
tsne_coords <- as.data.frame(reducedDim(sce, "TSNE"))
colnames(tsne_coords) <- c("TSNE1", "TSNE2")

cell_metadata <- as.data.frame(colData(sce))
cell_metadata$average_pseudotime <- all
plot_data <- bind_cols(tsne_coords, cell_metadata)

subset_data <- plot_data %>%
  filter(group == "MYCNamp")

trajectory_lines_data <- as.data.frame(pseudo.all$connected$TSNE)
colnames(trajectory_lines_data) <- c("edge", "TSNE1", "TSNE2")

background <- ggplot(plot_data, 
                     aes(x = TSNE1, y = TSNE2)) +
  geom_point(color = "darkgray", 
             size = 1.5, 
             alpha = 0.5)

p <- background + 
  geom_point(data = subset_data, 
             aes(color = average_pseudotime), 
             size = 1.5) +
  geom_line(data = trajectory_lines_data,
            aes(x = TSNE1, y = TSNE2, group = edge),
            linewidth = 1,
            colour = "black") +
  labs(title = "MYCNamp",
       x = "TSNE 1",
       y = "TSNE 2",
       color = "Pseudotime") + 
  scale_color_gradientn(
       colours = c("#4B0082", "darkgreen", "#CDA646"),
       values = scales::rescale(c(0, 1.5, 3)),
       limits = c(0, 3),
       oob = scales::squish) + 
  theme(text = element_text(size = 14, face = "bold"))

print(p)

pdf(paste0(params$out,"figures/TSCAN_trajectory_analysis_tsne_plot_MYCNamp.pdf"))
p
dev.off()
```

## __ATRXwtMYCNwt cells on top of the ALL_NK plot__

```{r fig.height=6, fig.width=7}
tsne_coords <- as.data.frame(reducedDim(sce, "TSNE"))
colnames(tsne_coords) <- c("TSNE1", "TSNE2")

cell_metadata <- as.data.frame(colData(sce))
cell_metadata$average_pseudotime <- all
plot_data <- bind_cols(tsne_coords, cell_metadata)

subset_data <- plot_data %>%
  filter(group == "ATRXwtMYCNwt")

trajectory_lines_data <- as.data.frame(pseudo.all$connected$TSNE)
colnames(trajectory_lines_data) <- c("edge", "TSNE1", "TSNE2")

background <- ggplot(plot_data, 
                     aes(x = TSNE1, y = TSNE2)) +
  geom_point(color = "darkgray", 
             size = 1.5, 
             alpha = 0.5)

p <- background + 
  geom_point(data = subset_data, 
             aes(color = average_pseudotime), 
             size = 1.5) +
  geom_line(data = trajectory_lines_data,
            aes(x = TSNE1, y = TSNE2, group = edge),
            linewidth = 1,
            colour = "black") +
  labs(title = "ATRXwtMYCNwt",
       x = "TSNE 1",
       y = "TSNE 2",
       color = "Pseudotime") + 
  scale_color_gradientn(
       colours = c("#4B0082", "darkgreen", "#CDA646"),
       values = scales::rescale(c(0, 1.5, 3)),
       limits = c(0, 3),
       oob = scales::squish) + 
  theme(text = element_text(size = 14, face = "bold"))

print(p)

pdf(paste0(params$out,"figures/TSCAN_trajectory_analysis_tsne_plot_ATRXwtMYCNwt.pdf"))
p
dev.off()
```


## __Adolescent healthy control cells on top of the ALL_NK plot__

```{r fig.height=6, fig.width=7}
tsne_coords <- as.data.frame(reducedDim(sce, "TSNE"))
colnames(tsne_coords) <- c("TSNE1", "TSNE2")

cell_metadata <- as.data.frame(colData(sce))
cell_metadata$average_pseudotime <- all
plot_data <- bind_cols(tsne_coords, cell_metadata)

subset_data <- plot_data %>%
  filter(group == "control")

trajectory_lines_data <- as.data.frame(pseudo.all$connected$TSNE)
colnames(trajectory_lines_data) <- c("edge", "TSNE1", "TSNE2")

background <- ggplot(plot_data, 
                     aes(x = TSNE1, y = TSNE2)) +
  geom_point(color = "darkgray", 
             size = 1.5, 
             alpha = 0.5)

p <- background + 
  geom_point(data = subset_data, 
             aes(color = average_pseudotime), 
             size = 1.5) +
  geom_line(data = trajectory_lines_data,
            aes(x = TSNE1, y = TSNE2, group = edge),
            linewidth = 1,
            colour = "black") +
  labs(title = "control",
       x = "TSNE 1",
       y = "TSNE 2",
       color = "Pseudotime") + 
  scale_color_gradientn(
       colours = c("#4B0082", "darkgreen", "#CDA646"),
       values = scales::rescale(c(0, 1.5, 3)),
       limits = c(0, 3),
       oob = scales::squish) + 
  theme(text = element_text(size = 14, face = "bold"))

print(p)

pdf(paste0(params$out,"figures/TSCAN_trajectory_analysis_tsne_plot_control.pdf"))
p
dev.off()
```

## ___each group separately___

### ___Healthy adolescent BM-derived NK cells___

```{r}
control <- as.SingleCellExperiment(CTRL_NK)
colLabels(control) <- control$predicted_NK_type_de.n350

pseudo.mnn <- TSCAN::quickPseudotime(control, 
                              use.dimred="UMAP", 
                              start = "hNK_Bm2")

mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

pdf(paste0(params$out,"figures/", "control_NK","_TSCAN","_TSNE.pdf"))
plotTSNE(control, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "control_NK","_TSCAN","_UMAP.pdf"))
plotUMAP(control, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, 
              mapping=aes(x=UMAP_1, y=UMAP_2, group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "control_NK","_TSCAN","_PCA.pdf"))
plotPCA(control, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, 
              mapping=aes(x=pca_1, y=pca_2, group=edge))
dev.off()
```

### ___ATRXdel NB patient BM-derived NK cells___

```{r}
ATRX <- as.SingleCellExperiment(ATRX_NK)
colLabels(ATRX) <- ATRX$predicted_NK_type_de.n350

pseudo.mnn <- quickPseudotime(ATRX, 
                              use.dimred="UMAP", 
                              start = "hNK_Bm2")
mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

pdf(paste0(params$out,"figures/", "ATRX_NK","_TSCAN","_TSNE.pdf"))
plotTSNE(ATRX, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350",, text_colour="#b4d1ee") +
    geom_line(data=pseudo.all$connected$TSNE, linewidth = 1, colour = "black",
              mapping=aes(x=tSNE_1, y=tSNE_2, group=edge)) + 
  theme(text = element_text(size = 14,face = "bold")) + 
  scale_color_gradientn(
        colours = c("#4B0082", "darkgreen", "#CDA646"),
        values = scales::rescale(c(0, 1.5, 3)),  # center green around 1.5, covering 0–3
        limits = c(0, 3),
        oob = scales::squish
    )
dev.off()

pdf(paste0(params$out,"figures/", "ATRX_NK","_TSCAN","_UMAP.pdf"))
plotUMAP(ATRX, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, 
              mapping=aes(x=UMAP_1, y=UMAP_2, group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "ATRX_NK","_TSCAN","_PCA.pdf"))
plotPCA(ATRX, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, 
              mapping=aes(x=pca_1, y=pca_2, group=edge))
dev.off()
```

### ___MNA NB patient BM-derived NK cells___

```{r}
MNA <- as.SingleCellExperiment(MNA_NK)
colLabels(MNA) <- MNA$predicted_NK_type_de.n350

pseudo.mnn <- quickPseudotime(MNA, 
                              use.dimred="UMAP", 
                              start = "hNK_Bm2")

mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

pdf(paste0(params$out,"figures/", "MNA_NK","_TSCAN","_TSNE.pdf"))
plotTSNE(MNA, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "MNA_NK","_TSCAN","_UMAP.pdf"))
plotUMAP(MNA, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, 
              mapping=aes(x=UMAP_1, y=UMAP_2, group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "MNA_NK","_TSCAN","_PCA.pdf"))
plotPCA(MNA, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, 
              mapping=aes(x=pca_1, y=pca_2, group=edge))
dev.off()
```

### ___sporadic NB patient BM-derived NK cells___

```{r}
sporadic <- as.SingleCellExperiment(sporadic_NK)
colLabels(sporadic) <- sporadic$predicted_NK_type_de.n350

pseudo.mnn <- quickPseudotime(sporadic, 
                              use.dimred="UMAP", 
                              start = "hNK_Bm2")

mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

pdf(paste0(params$out,"figures/", "sporadic_NK","_TSCAN","_TSNE.pdf"))
plotTSNE(sporadic, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "sporadic_NK","_TSCAN","_UMAP.pdf"))
plotUMAP(sporadic, colour_by=I(mnn.pseudo), 
         text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, 
              mapping=aes(x=UMAP_1, y=UMAP_2, group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "sporadic_NK","_TSCAN","_PCA.pdf"))
plotPCA(sporadic, colour_by=I(mnn.pseudo), 
        text_by="predicted_NK_type_de.n350", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, 
              mapping=aes(x=pca_1, y=pca_2, group=edge))
dev.off()
```

### ___Healthy ADULT BM-derived NK cells___

Crinier at al., 2021 adult BM-derived-NK cells from 8 healthy donors

#### ___hNK_Bm2 cluster as START___

```{r}
adult_sce <- as.SingleCellExperiment(adult_BM_NK)
colLabels(adult_sce) <- adult_sce$cluster2

pseudo.mnn <- quickPseudotime(adult_sce, 
                              use.dimred="UMAP", 
                              start = "hNK_Bm2")

mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)

pdf(paste0(params$out,"figures/", "adult_NK","_TSCAN_hNK_Bm2","_TSNE.pdf"))
plotTSNE(adult_sce, colour_by=I(mnn.pseudo), 
        text_by="cluster2", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "adult_NK","_TSCAN_hNK_Bm2","_UMAP.pdf"))
plotUMAP(adult_sce, colour_by=I(mnn.pseudo), 
         text_by="cluster2", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, 
              mapping=aes(x=UMAP_1, y=UMAP_2, group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "adult_NK","_TSCAN_hNK_Bm2","_PCA.pdf"))
plotPCA(adult_sce, colour_by=I(mnn.pseudo), 
        text_by="cluster2", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, 
              mapping=aes(x=PC_1, y=PC_2, group=edge))
dev.off()
```

#### ___hNK_Bm3 cluster as START___

```{r}
adult_sce <- as.SingleCellExperiment(adult_BM_NK)
colLabels(adult_sce) <- adult_sce$cluster2

pseudo.mnn <- quickPseudotime(adult_sce, 
                              use.dimred="UMAP", 
                              start = "hNK_Bm3")

mnn.pseudo <- averagePseudotime(pseudo.mnn$ordering)


pdf(paste0(params$out,"figures/", "adult_NK","_TSCAN_hNK_Bm3","_TSNE.pdf"))
plotTSNE(adult_sce, colour_by=I(mnn.pseudo), 
        text_by="cluster2", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$TSNE, mapping=aes(x=tSNE_1, y=tSNE_2, 
                                                          group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "adult_NK","_TSCAN_hNK_Bm3","_UMAP.pdf"))
plotUMAP(adult_sce, colour_by=I(mnn.pseudo), 
         text_by="cluster2", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$UMAP, 
              mapping=aes(x=UMAP_1, y=UMAP_2, group=edge))
dev.off()

pdf(paste0(params$out,"figures/", "adult_NK","_TSCAN_hNK_Bm3","_PCA.pdf"))
plotPCA(adult_sce, colour_by=I(mnn.pseudo), 
        text_by="cluster2", text_colour="red") +
    geom_line(data=pseudo.mnn$connected$PCA, 
              mapping=aes(x=PC_1, y=PC_2, group=edge))
dev.off()
```


# ___ENV___

```{r}
sessionInfo()
```
