---
title: "NK project: DA analysis for TME cell types"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/CancReg/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/CancReg/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/CancReg/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/NK_project/Rds/fetahu2023_withNewMetadata.Rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/mnt_out/CancReg/scRNA-seq_stage_M/results/metadata/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```

```{r load project specific R packages and color code}
source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")
```

```{r}
library(reshape2)
```

Supplemental Figure 1 of the manuscript Rados et al., 2025

Since clinical groups have been misannotated, we wanted to show DA of TME cell
type with the new clinical annotations. 

We show that cluster 6 of T-cells, NK cells and cluster 1 of Monocytes, as well
as cluster 3 of B-cells are present in higher abundance in ATRXdel NB bone marrow
samples. 

```{r}
fetahu2023 <- readRDS(params$input_scRNA)
```

```{r}
fetahu2023$celltype <- fetahu2023$orig.ident
table(fetahu2023$celltype, fetahu2023$group)
```

# ___DA testing___

If comparing more than two sample / groups, use DA testing with contrasts
MiloR doesn't use the "original" clusters but rather creates neighborhoods
based on similar logFC (note the fold change here is the number of cells
per cell type)

```{r}
Idents(fetahu2023) <- fetahu2023$orig.ident
# remove TU cells for DA analysis
TME_cells <- subset(fetahu2023, 
       idents = "NB (8)", 
       invert = TRUE)

TME_cells$celltype <- TME_cells$orig.ident

TME_cells@meta.data$group <- 
  factor(TME_cells@meta.data$group, 
         levels = GROUP_ORDER)
levels(TME_cells@meta.data$group)

TME_cells@meta.data$rados_et_al_2025 <- 
  factor(TME_cells@meta.data$rados_et_al_2025, 
         levels = PATIENT_ORDER)
levels(TME_cells@meta.data$rados_et_al_2025)
```


```{r}
# Create meta-clusters
TME_cells$celltype_meta <- as.character(Idents(TME_cells))

# Annotate all T-cell subsets as 'T-cells'
TME_cells$celltype_meta[grepl("^T \\(", TME_cells$celltype_meta)] <- "T-cells"
TME_cells$celltype_meta[grepl("^B \\(", TME_cells$celltype_meta)] <- "B-cells"
TME_cells$celltype_meta[grepl("^NK \\(", TME_cells$celltype_meta)] <- "NK-cells"
TME_cells$celltype_meta[grepl("^M \\(", TME_cells$celltype_meta)] <- "Myeloid cells"
TME_cells$celltype_meta[grepl("^pDC \\(", TME_cells$celltype_meta)] <- "Dendritic cells"
TME_cells$celltype_meta[grepl("^SC \\(", TME_cells$celltype_meta)] <- "Progenitor cells"
TME_cells$celltype_meta[grepl("^E \\(", TME_cells$celltype_meta)] <- "Erythroid cells"
TME_cells$celltype_meta[grepl("^other \\(", TME_cells$celltype_meta)] <- "Unknown"

# If needed, convert back to factor with proper ordering
TME_cells$celltype_meta <- factor(TME_cells$celltype_meta, 
                                      levels = unique(TME_cells$celltype_meta))

Idents(TME_cells) <- TME_cells$celltype_meta
```


```{r fig.height=10, fig.width=10}
percent_celltype_group <- table(TME_cells$celltype,
                          TME_cells$group)

# Transform this data in %
data_percentage <- apply(percent_celltype_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

percent_celltype_group_trans <- melt(data_percentage) 

barplots_celltypes <- ggplot(percent_celltype_group_trans, 
                             aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of cells assigned to each cluster", 
       fill = "Clusters") +  
  ggtitle("") +
  theme_minimal(base_size = 22) + 
  theme(axis.text.x = element_text(angle = 45,

                                   hjust = 1)) +
  scale_fill_manual(values = COLOR_CODE_FETAHU2023_v2)

barplots_celltypes

pdf(paste0(params$out,"figures/barplot_ALL_celltypes_per_NB_group.pdf"),
    height = 10, width = 10)
barplots_celltypes
dev.off()


```


```{r fig.height=10, fig.width=10}
percent_celltype_group <- table(TME_cells$celltype_meta,
                          TME_cells$group)

# Transform this data in %
data_percentage <- apply(percent_celltype_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

percent_celltype_group_trans <- melt(data_percentage) 

barplots_celltypes <- ggplot(percent_celltype_group_trans, 
                             aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of cells assigned to each cluster", 
       fill = "Clusters") +  
  ggtitle("") +
  theme_minimal(base_size = 22) + 
  theme(axis.text.x = element_text(angle = 45,

                                   hjust = 1)) +
  scale_fill_manual(values = COLOR_CODE_FETAHU2023_META)

barplots_celltypes

pdf(paste0(params$out,"figures/barplot_ALL_meta_celltypes_per_NB_group.pdf"),
    height = 10, width = 10)
barplots_celltypes
dev.off()


```


```{r}
TME_cells@meta.data$rados_et_al_2025<- factor(TME_cells@meta.data$rados_et_al_2025)
TME_cells@meta.data$celltype <- factor(TME_cells@meta.data$celltype)
TME_per_group <- table(TME_cells$celltype, TME_cells$group)
TME_per_group

write.table(TME_per_group, file = paste0(params$out,"tables/TME_per_group.csv"))
```



```{r}
TME_cells@meta.data$rados_et_al_2025 <- factor(TME_cells@meta.data$rados_et_al_2025)
TME_cells@meta.data$celltype_meta <- factor(TME_cells@meta.data$celltype_meta)
TME_meta_per_group <- table(TME_cells$celltype_meta, TME_cells$group)

TME_meta_per_group

write.table(TME_meta_per_group, 
            file = paste0(params$out,"tables/TME_meta_per_group.csv"))
```

# ___metacluster box plots with stats___

## ___Stats for celltype clusters___

```{r}
table_celltype_per_sample <- table(TME_cells$celltype_meta,TME_cells$rados_et_al_2025)
write.table(TME_meta_per_group, 
            file = paste0(params$out,"tables/TME_meta_per_sample.csv"))

table_celltype_per_sample <- as.data.frame(table_celltype_per_sample)

table_celltype_per_sample$group <- gsub("*[0-9]", "", table_celltype_per_sample$Var2)
table_celltype_per_sample

library(dplyr)

table_celltype_per_sample_percent <- table_celltype_per_sample %>%
  group_by(Var2, Var1) %>%           
  summarise(total = sum(Freq, na.rm = TRUE)) %>%  
  mutate(percentage = total * 100 / sum(total)) %>% 
  ungroup()

table_celltype_per_sample_percent <- as.data.frame(table_celltype_per_sample_percent)
table_celltype_per_sample_percent$group <- gsub("*[0-9]", "", table_celltype_per_sample_percent$Var2)
table_celltype_per_sample_percent

data_wide <- t(table_celltype_per_sample_percent)
```


```{r}
# TME clusters per sample

library(ggplot2)
library(ggpubr)
library(dplyr)

TME_cells_per_meta_cluster <- ggplot(table_celltype_per_sample_percent, 
       aes(x = Var2, y = percentage, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of cells per cluster", 
       fill = "Clusters") +  
  ggtitle("") +
  theme_minimal(base_size = 22) + 
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  scale_fill_manual(values = COLOR_CODE_FETAHU2023_META_v2)


pdf(paste0(params$out,"figures/barplot_ALL_meta_celltypes_per_sample.pdf"),
    height = 5, width = 8)
TME_cells_per_meta_cluster
dev.off()
```


### ___NK cells___

```{r}
NK_table <- filter(as.data.frame(table_celltype_per_sample_percent), 
                   Var1 == "NK-cells")

# Perform ANOVA testing
NK_table_anova <- aov(percentage ~ group, data = NK_table)
summary(NK_table_anova)

# Extract p-value
p_value <- summary(NK_table_anova)[[1]][["Pr(>F)"]][1]

# perform wilcox testing
p_value1 <- wilcox.test(percentage ~ group, 
                        data = subset(NK_table, 
                                      group %in% c("A", "S")))$p.value
p_value2 <- wilcox.test(percentage ~ group, 
                        data = subset(NK_table, 
                                      group %in% c("A", "C")))$p.value
p_value3 <- wilcox.test(percentage ~ group, 
                        data = subset(NK_table, 
                                      group %in% c("A", "M")))$p.value

# Adjust for multiple testing using FDR (Benjamini-Hochberg)
pvals <- c(p_value1, p_value2, p_value3)
names(pvals) <- c("ATRXdel_vs_ATRXwtMYCNwt", "ATRXdel_vs_control", "ATRXdel_vs_MYCNamp")
pvals_adj <- p.adjust(pvals, method = "fdr")
pvals_adj

# Ensure group order
NK_table$group <- factor(NK_table$group, levels = c("C", "A", "M", "S"))

# Create boxplot and add p-value annotation
boxplot1 <- NK_table %>%
  ggplot( aes(x = group , y =  percentage, fill = group)) +
    geom_boxplot() +
    ggtitle("NK cells per group") +
    xlab("") + ylab("Percentage of NK cells per group") +
  geom_signif(comparisons = list(c("C", "A"),
                                 c("A", "M"),
                                 c("A", "S")),
              map_signif_level = FALSE, 
              y_position = c(max(NK_table$percentage) + 0, 
                             max(NK_table$percentage) + 2, 
                             max(NK_table$percentage) + 4)) +
  scale_fill_manual(values = COLOR_CODE_GROUP_v4) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    strip.background = element_blank(),
    plot.background = element_blank(),
    axis.line = element_line(color = "black")
  )
```

```{r}
# FINAL PLOT

library(ggplot2)
library(ggpubr)
library(dplyr)

NK_table <- NK_table %>%
  mutate(group_label = recode(group,
                              "C" = "control",
                              "A" = "ATRXdel",
                              "M" = "MYCNamp",
                              "S" = "ATRXwtMYCNwt"),
         group_label = factor(group_label,
                              levels = c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt")))

comparisons <- list(c("control", "ATRXdel"),
                    c("control", "MYCNamp"),
                    c("control", "ATRXwtMYCNwt"))


boxplot1 <- NK_table %>%
  ggplot(aes(x = group_label, y = percentage, fill = group_label)) +
  geom_boxplot(outlier.shape = NA, color = NA, width = 0.5) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, color = "black", fatten = 0.9) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2, color = "black") +
  geom_point(position = position_dodge(width = 0.5), color = "black", size = 2, alpha = 0.7) +
  stat_compare_means(comparisons = comparisons,
                     method = "wilcox.test",
                     p.adjust.method = "BH",
                     label = "p.format", # "p.signif"
                     tip.length = 0.01,
                     size = 3.5) +
  scale_fill_manual(values = COLOR_CODE_GROUP) +
  ggtitle("NK cells per group") +
  xlab("") +
  ylab("Percentage of NK cells per group") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    strip.background = element_blank(),
    plot.background = element_blank(),
    axis.line = element_line(color = "black")
  )
```


```{r fig.height=4, fig.width=5}
boxplot1

pdf(paste0(params$out,"figures/barplot_metaclusters_NKcells_wStats_p-value_FINAL.pdf"),
    height = 4, width = 5)
boxplot1
dev.off()
```

```{r fig.height=7, fig.width=7}
table_celltype_per_sample_percent <- as.data.frame(table_celltype_per_sample_percent)
# Ensure group order
table_celltype_per_sample_percent$group <- factor(table_celltype_per_sample_percent$group, 
                                                   levels = c("C", "A", "M", "S"))

my_comparisons <- list(c("A", "C"), c("A", "M"), c("A", "S"))

# Create facetted boxplot with p-values
boxplot_all <- ggplot(table_celltype_per_sample_percent, 
                      aes(x = group, y = percentage, fill = group)) +
  geom_boxplot() +
  facet_wrap(~ Var1, scales = "free_y") +
  stat_compare_means(comparisons = my_comparisons, 
                     label = "p.signif", 
                     method = "wilcox.test",
                     tip.length = 0.01,
                     size = 3) +
  scale_fill_manual(values = COLOR_CODE_GROUP_v4) +
  theme(strip.text = element_text(face = "bold")) +
  labs(title = "Metacluster frequencies per group",
       y = "Percentage of cells",
       x = "") +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    strip.background = element_blank(),
    plot.background = element_blank(),
    axis.line = element_line(color = "black")
  )

boxplot_all
```

```{r}
pdf(paste0(params$out,"figures/barplot_metaclusters_wStats.pdf"),
    height = 8, width = 8)
boxplot_all
dev.off()
```


# ___DA analysis___

```{r miloR prepare data}
library("statmod")
library("miloR")
library("SingleCellExperiment")
library("scater")
library("scran")
library("dplyr")
library("patchwork")

rna_milo <- Milo(as.SingleCellExperiment(TME_cells))
rna_milo

reducedDim(rna_milo, "UMAP") <- reducedDim(as.SingleCellExperiment(TME_cells), 
                                           "UMAP")
plotUMAP(rna_milo, 
         colour_by="group") + 
  plotUMAP(rna_milo, 
           colour_by="rados_et_al_2025.x")

# Define cell neighbourhoods
rna_milo <- buildGraph(rna_milo, 
                       k = 100)

## Constructing kNN graph with k = 100
rna_milo <- makeNhoods(rna_milo,
                       prop = 0.1, 
                       k = 100, # 100 nearest neighbors for clustering
                       d = 50, 
                       refined = TRUE, 
                       refinement_scheme = "graph") # make nhoods using graph-only as this is faster

# metadata
colData(rna_milo)$patients <- paste(colData(rna_milo)$group, 
                                    colData(rna_milo)$rados_et_al_2025.x, sep="_")

# Counting cells in neighborhoods
rna_milo <- countCells(rna_milo, meta.data = data.frame(colData(rna_milo)), 
                       samples = "patients") # make the nhood X sample counts matrix


# Defining experimental design
rna_design <- data.frame(colData(rna_milo))[,c("patients", "group")]
rna_design <- distinct(rna_design)
rownames(rna_design) <- rna_design$patients
rna_design

# Computing neighborhood connectivity
rna_milo <- calcNhoodDistance(rna_milo, 
                              d = 50, 
                              reduced.dim = "PCA")

# DA testing
da_results <- testNhoods(rna_milo, 
                         design = ~ 0 + group,
                         design.df = rna_design, 
                         reduced.dim = "PCA")
head(da_results)

da_results %>%
  arrange(SpatialFDR) %>%
  head() 

rna_milo <- buildNhoodGraph(rna_milo)


# QUALITY CHECK OF THE DIFFERENTIAL ABUNDANCE TESTING
# Plot single-cell UMAP
umap_pl <- plotReducedDim(rna_milo, 
                          dimred = "UMAP", 
                          colour_by = "group", 
                          text_by = "celltype", 
                          text_size = 3, point_size = 0.5,) +
  guides(fill = "none")

rna_milo <- buildNhoodGraph(rna_milo)

# Plot neighbourhood graph
nh_graph_pl <- plotNhoodGraphDA(rna_milo, 
                                da_results, 
                                layout = "UMAP", 
                                alpha = 0.05) 

umap_pl + nh_graph_pl +
  plot_layout(guides = "collect")

# Differential abundance testing with contrasts
rna_design <- data.frame(colData(rna_milo))[,c("patients", "group")]
rna_design <- distinct(rna_design)
rownames(rna_design) <- rna_design$patients

# Reorder rownames to match columns of nhoodCounts(milo)
rna_design <- rna_design[colnames(nhoodCounts(rna_milo)), , drop=FALSE]
table(rna_design$group)

saveRDS(rna_milo, file = paste0(params$out,"Rds/miloR_TME_cells.Rds"))
```

## ___A. ATRXdel vs. ATRXwt [MYCNamp + ATRXwtMYCNwt]___

```{r fig.height=6, fig.width=10}
model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupATRXdel - ((groupMYCNamp + groupATRXwtMYCNwt) / 2)")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXdelvsWT <- testNhoods(rna_milo, 
                                      design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXdelvsWT$SpatialFDR < 0.1)
table(da_results_ATRXdelvsWT$SpatialFDR < 0.05)
table(da_results_ATRXdelvsWT$SpatialFDR < 0.01)

# annotate hood with original cell type annotation
da_results_ATRXdelvsWT <- annotateNhoods(rna_milo, 
                                          da_results_ATRXdelvsWT, 
                                          coldata_col = "celltype")

# remove hoods that contained mixed cell type populations
da_results_ATRXdelvsWT$celltype <- 
  ifelse(da_results_ATRXdelvsWT$celltype_fraction < 0.7, 
         "Mixed", da_results_ATRXdelvsWT$celltype)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXdelvsWT %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

da_results_ATRXdelvsWT$celltype <- as.factor(da_results_ATRXdelvsWT$celltype)

DA_plot_ATRTXdel_vs_ATRXwt <- plotDAbeeswarm(da_results_ATRXdelvsWT, 
               group.by = "celltype", 
               alpha = 0.1) # cut-off for spatial FDR

DA_plot_ATRTXdel_vs_ATRXwt

pdf(paste0(params$out,"figures/DA_plot_all_clusters_ATRTXdel_vs_ATRXwt.pdf"),
    width = 10, height = 6)
DA_plot_ATRTXdel_vs_ATRXwt
dev.off()
```

```{r fig.height=6, fig.width=10}
model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupATRXdel - ((groupMYCNamp + groupATRXwtMYCNwt) / 2)")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXdelvsWT <- testNhoods(rna_milo, 
                                      design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXdelvsWT$SpatialFDR < 0.1)
table(da_results_ATRXdelvsWT$SpatialFDR < 0.05)
table(da_results_ATRXdelvsWT$SpatialFDR < 0.01)

# annotate hood with original cell type annotation
da_results_ATRXdelvsWT <- annotateNhoods(rna_milo, 
                                          da_results_ATRXdelvsWT, 
                                          coldata_col = "celltype_meta")

# remove hoods that contained mixed cell type populations
da_results_ATRXdelvsWT$celltype_anno <- 
  ifelse(da_results_ATRXdelvsWT$celltype_meta_fraction < 0.7, 
         "Mixed", da_results_ATRXdelvsWT$celltype_meta)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXdelvsWT %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

da_results_ATRXdelvsWT$celltype_anno <- as.factor(da_results_ATRXdelvsWT$celltype_anno)

DA_plot_ATRTXdel_vs_ATRXwt <- plotDAbeeswarm(da_results_ATRXdelvsWT, 
               group.by = "celltype_anno", 
               alpha = 0.1) # cut-off for spatial FDR

DA_plot_ATRTXdel_vs_ATRXwt

pdf(paste0(params$out,"figures/DA_plot_all_meta-clusters_ATRTXdel_vs_ATRXwt.pdf"),
    width = 10, height = 6)
DA_plot_ATRTXdel_vs_ATRXwt
dev.off()
```


## ___B. ATRXdel vs. ALL (MYCNamp and ATRXwtMYCNwt and controls)___

```{r fig.height=7, fig.width=11}
model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupATRXdel - ((groupMYCNamp + groupATRXwtMYCNwt + groupcontrol) / 3)")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXdelvsALL <- testNhoods(rna_milo, 
                                      design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXdelvsALL$SpatialFDR < 0.1)
table(da_results_ATRXdelvsALL$SpatialFDR < 0.05)
table(da_results_ATRXdelvsALL$SpatialFDR < 0.01)

# annotate hood with original cell type annotation
da_results_ATRXdelvsALL <- annotateNhoods(rna_milo, 
                                          da_results_ATRXdelvsALL, 
                                          coldata_col = "celltype")

# remove hoods that contained mixed cell type populations
da_results_ATRXdelvsALL$celltype <- 
  ifelse(da_results_ATRXdelvsALL$celltype_fraction < 0.7, 
         "Mixed", da_results_ATRXdelvsALL$celltype)

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXdelvsALL %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

da_results_ATRXdelvsALL$celltype <- as.factor(da_results_ATRXdelvsALL$celltype)


DA_plot_ATRXdelvsALL <- plotDAbeeswarm(da_results_ATRXdelvsALL, 
               group.by = "celltype", 
               alpha = 0.01) # cut-off for spatial FDR

DA_plot_ATRXdelvsALL

pdf(paste0(params$out,"figures/DA_plot_all_clusters_ATRTXdel_vs_ALL.pdf"),
    width = 10, height = 6)
DA_plot_ATRXdelvsALL
dev.off()
```


```{r fig.height=6, fig.width=10}
model <- model.matrix(~ 0 + group, data = rna_design)
ave.contrast <- c("groupATRXdel - ((groupMYCNamp + groupATRXwtMYCNwt + groupcontrol) / 3)")
mod.constrast <- makeContrasts(contrasts = ave.contrast, levels = model)
mod.constrast

da_results_ATRXdelvsALL <- testNhoods(rna_milo, 
                                      design = ~ 0 + group, 
                         design.df = rna_design, 
                         model.contrasts = ave.contrast, 
                         fdr.weighting = "graph-overlap")

table(da_results_ATRXdelvsALL$SpatialFDR < 0.1)
table(da_results_ATRXdelvsALL$SpatialFDR < 0.05)
table(da_results_ATRXdelvsALL$SpatialFDR < 0.01)

# annotate hood with original cell type annotation
da_results_ATRXdelvsALL <- annotateNhoods(rna_milo, 
                                          da_results_ATRXdelvsALL, 
                                          coldata_col = "celltype_meta")

# remove hoods that contained mixed cell type populations
da_results_ATRXdelvsALL$celltype_anno <- 
  ifelse(da_results_ATRXdelvsALL$celltype_meta_fraction < 0.7, 
         "Mixed", da_results_ATRXdelvsALL$celltype_meta)

da_results_ATRXdelvsALL$celltype_anno <- as.factor(da_results_ATRXdelvsALL$celltype_anno)

DA_plot_ATRXdelvsALL <- plotDAbeeswarm(da_results_ATRXdelvsALL, 
               group.by = "celltype_anno", 
               alpha = 0.01) # cut-off for spatial FDR

DA_plot_ATRXdelvsALL

pdf(paste0(params$out,"figures/DA_plot_all_meta-clusters_ATRTXdel_vs_ALL.pdf"),
    width = 10, height = 6)
DA_plot_ATRXdelvsALL
dev.off()
```

```{r}

# filter only significant populations (less than 10% false discovery rate and 
# fold change bigger than 0.5)
results <- da_results_ATRXdelvsALL %>% 
  arrange(FDR) %>% filter(FDR < 0.1 & abs(logFC) >= 0.5)
results

results_table <- data.frame(da_results_ATRXdelvsALL$celltype_anno,
                            da_results_ATRXdelvsALL$logFC)

# Ensure your data frame has proper column names
colnames(results_table) <- c("celltype_anno", "logFC")

# Plot
ggplot(results_table, aes(x = logFC, fill = celltype_anno)) +
  geom_density(alpha = 0.5, color = NA) +  # No outline to avoid clutter
  scale_fill_manual(values = COLOR_CODE_FETAHU2023_META) +
  theme_minimal() +
  labs(
    title = "DA of cell types",
    x = "logFC",
    y = "Density"
  ) +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(hjust = 0.5)
  )

```

# ___dotplots___

```{r fig.height=10, fig.width=10}

genes_of_interest <- c("CD6","CD8A","FCGR3A","ID2","ZEB2","STAT5A","STAT4","NCR1")

do_DotPlot(ALL_NK, 
           group.by = "NK_subtype_per_group",
           features = genes_of_interest,
           scale = TRUE,
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

do_DotPlot(ALL_NK, 
           group.by = "group_per_NK_subtype",
           features = genes_of_interest,
           scale = TRUE,
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

```{r fig.height=10, fig.width=10}

genes_of_interest <- c("IL18","IL2", "IL15")

do_DotPlot(ALL_NK, 
           group.by = "NK_subtype_per_group",
           features = genes_of_interest,
           scale = TRUE,
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

do_DotPlot(ALL_NK, 
           group.by = "group_per_NK_subtype",
           features = genes_of_interest,
           scale = TRUE,
           use_viridis = TRUE,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

```{r}
sessionInfo()
```

