---
title: "CCC analysis: NB-NK axis"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/CancReg/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/CancReg/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_BM3: "/home/rstudio/mnt_data/CancReg/BM/Lee/disco_bone_marrow_v2.0.rds"
  input_scRNA: "/home/rstudio/mnt_out/NK_project/Rds/fetahu2023_withNewMetadata.Rds"
  input_NK: "/home/rstudio/mnt_out/NK_project/Rds/ALL_NK_cluster_withNewMetadata.Rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/NK-NB_study/metadata/"
  nicheNetDB: "/home/rstudio/mnt_resources/multiNicheNetDB/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```

```{r eval=FALSE, include=FALSE}
if(!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools") 
}

devtools::install_github("saeyslab/nichenetr")
devtools::install_github("saeyslab/multinichenetr")

install.packages("igraph")
```

```{r load project specific R packages and color code}
library(nichenetr)
library(multinichenetr)
library(igraph)

source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")
```

# ___Input_data___

```{r}
fetahu2023 <- readRDS(file = params$input_scRNA)
ALL_NK <-  readRDS(file = params$input_NK)
```

```{r define comparison groups and clusters of interest}
Idents(fetahu2023) <- fetahu2023$orig.ident
fetahu2023$cell_type <- fetahu2023$orig.ident
table(fetahu2023$cell_type, fetahu2023$group)
```

```{r Add NK subclusters}
# add NK sub-cluster information to the Seurat object
# set cell type annotation of the tumor cell object to the new seurat clusters
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# identify cell IDs associated with each cluster
cluster1 <- WhichCells(ALL_NK, ident = "hNK_Bm1")
cluster2 <- WhichCells(ALL_NK, ident = "hNK_Bm2")
cluster3 <- WhichCells(ALL_NK, ident = "hNK_Bm3")
cluster4 <- WhichCells(ALL_NK, ident = "hNK_Bm4")

fetahu2023@active.ident <- fetahu2023$cell_type 

# re-label the NK cells to the corresponding predicted sub-clusters 
Idents(object = fetahu2023, cells = cluster1) <- "NK (hNK_Bm1)"
Idents(object = fetahu2023, cells = cluster2) <- "NK (hNK_Bm2)"
Idents(object = fetahu2023, cells = cluster3) <- "NK (hNK_Bm3)"
Idents(object = fetahu2023, cells = cluster4) <- "NK (hNK_Bm4)"

# Add actATRXwtMYCNwte identities to metadata column new_clusters
fetahu2023$cell_type_detailed <- fetahu2023@active.ident

table(fetahu2023$cell_type_detailed)
```

# ___CCC analysis w/multinichenet: NB - NK axis___

```{r LR database}
organism = "human"

# 0. prepare db
lr_network_all = 
    readRDS(paste0(params$nicheNetDB,"lr_network_human_allInfo_30112033.rds")) %>% 
    mutate(
      ligand = convert_alias_to_symbols(ligand, organism = organism), 
      receptor = convert_alias_to_symbols(receptor, organism = organism))
  lr_network_all = lr_network_all  %>% 
    mutate(ligand = make.names(ligand), receptor = make.names(receptor)) 
  lr_network = lr_network_all %>% 
    distinct(ligand, receptor)
  
ligand_target_matrix = readRDS(paste0(params$nicheNetDB,
                                      "ligand_target_human_matrix_nsga2r_final.rds"))
  colnames(ligand_target_matrix) = colnames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  rownames(ligand_target_matrix) = rownames(ligand_target_matrix) %>% 
    convert_alias_to_symbols(organism = organism) %>% make.names()
  lr_network = lr_network %>% filter(ligand %in% colnames(ligand_target_matrix))
  ligand_target_matrix = ligand_target_matrix[, lr_network$ligand %>% unique()]
```

```{r get_lfc_celltype function}
# rewrite the "get_lfc_celltype" function, as the select function from dplyr R package is causing issues (i explicitly write the name of the package in the function, otherwise i get an error that select cannot be applied to this object)
get_lfc_celltype <- function(celltype_oi, seurat_obj, 
                              condition_colname, condition_oi, 
                              condition_reference, celltype_col = "celltype", ...) {
  seuratObj_sender = subset(seurat_obj, idents = celltype_oi)
  seuratObj_sender = SetIdent(seuratObj_sender, 
                              value = seuratObj_sender[[condition_colname, 
                                                        drop = TRUE]])
  DE_table_sender = FindMarkers(object = seuratObj_sender, 
    ident.1 = condition_oi, ident.2 = condition_reference, random.seed = 28,
    min.cells.group = 50, logfc.threshold = logfc.threshold, min.pct = min.pct) %>% rownames_to_column("gene")
  DE_table_sender = DE_table_sender %>% as_tibble() %>% 
    dplyr::select(-p_val) %>% dplyr::select(gene, avg_log2FC)
  return(DE_table_sender)
}
```

```{r visualize_network function}
visualize_network <- function (network, colors) {
  requireNamespace("dplyr")
  requireNamespace("ggplot2")
  requireNamespace("ggraph")
  nodes = network$nodes %>% data.frame() %>% magrittr::set_rownames(network$nodes$node)
  colors_regulation = NULL
  colors_regulation["up"] = "#50C878"
  colors_regulation["down"] = "#C70039"
  network_igraph = igraph::graph_from_data_frame(d = network$links %>% 
    dplyr::filter(type == "Ligand-Target"), vertices = nodes, 
    directed = T)
  network_tidygraph = tidygraph::as_tbl_graph(network_igraph)
  set.seed(2805)
  plot = ggraph::ggraph(network_tidygraph, layout = "dh") + 
    ggraph::geom_edge_fan(aes(color = direction_regulation), 
      edge_width = 1, arrow = arrow(length = unit(3, "mm")), 
      end_cap = ggraph::circle(5.5, "mm"), start_cap = ggraph::circle(3, 
        "mm")) + ggraph::geom_node_label(aes(label = gene, 
    fill = celltype), fontface = "bold", size = 3.5, nudge_x = 0, 
    nudge_y = 0, color = "whitesmoke") + ggraph::theme_graph(foreground = "black", 
    fg_text_colour = "white", base_family = "Helvetica") + 
    facet_grid(. ~ group) + ggraph::scale_edge_color_manual(values = colors_regulation) + 
    scale_fill_manual(values = colors)
  return(list(plot = plot, network_igraph = network_igraph, 
    network_tidygraph = network_tidygraph))
}
```

## ___ATRXdel vs. ATRXwt [MYCNamp and ATRXwtMYCNwt]____

```{r MultiNicheNet analysis: ATRXdel vs ATRXwt}
fetahu2023@meta.data$ATRX_groups <- NA
fetahu2023@meta.data <- fetahu2023@meta.data %>% 
  mutate(ATRX_groups = ifelse(group %in% c("control"), "control",
                       ifelse(group %in% c("MYCNamp", "ATRXwtMYCNwt"), "ATRXwt", 
                              "ATRXdel")))

fetahu2023@meta.data$ATRX_groups <- factor(fetahu2023@meta.data$ATRX_groups)
fetahu2023@meta.data$ATRX_groups <- droplevels(fetahu2023@meta.data$ATRX_groups)

# sanity check
table(fetahu2023@meta.data$ATRX_groups)
table(fetahu2023@meta.data$group)

# Get the cell IDs present in the Seurat object (cell barcodes in counts)
cells_present <- colnames(fetahu2023[['RNA']]$counts)

# Filter the metadata to keep only the rows corresponding to cells in the current Seurat object
filtered_metadata <- fetahu2023@meta.data %>%
  rownames_to_column(var = "cell_id") %>%    # Convert rownames to a column
  filter(cell_id %in% cells_present) %>%     # Filter out cells not present in the object
  column_to_rownames(var = "cell_id")        # Convert cell_id back to rownames

# Update the Seurat object with the filtered metadata
fetahu2023@meta.data <- filtered_metadata

# 1. Change to the new group identitites
Idents(fetahu2023) <- fetahu2023@meta.data$ATRX_groups
  
# 2. convert to single cell experiment
sce = Seurat::as.SingleCellExperiment(fetahu2023, assay = "RNA")

# 3. Define which columns in the metadata contain the neccessary information
SummarizedExperiment::colData(sce)$rados_et_al_2025.x = SummarizedExperiment::colData(sce)$rados_et_al_2025.x %>% make.names()
SummarizedExperiment::colData(sce)$ATRX_groups = SummarizedExperiment::colData(sce)$ATRX_groups %>% make.names()
SummarizedExperiment::colData(sce)$cell_type_detailed = SummarizedExperiment::colData(sce)$cell_type_detailed %>% make.names()

sample_id = "rados_et_al_2025.x"
group_id = "ATRX_groups" 
celltype_id = "cell_type_detailed"

# 4. Set covariates and batches
covariates = NA
batches = NA 
 
# 5. Set comparison groups
contrasts_oi = c("'ATRXdel-ATRXwt','ATRXwt-ATRXdel'")

# Note the format to indicate the contrasts! This formatting should be adhered to very strictly, and white spaces are not allowed! Check ?get_DE_info for explanation about how to define this well. The most important points are that: *each contrast is surrounded by single quotation marks *contrasts are separated by a comma without any white space *all contrasts together are surrounded by double quotation marks.

contrast_tbl = tibble(contrast =  c("ATRXdel-ATRXwt","ATRXwt-ATRXdel"), 
                      group = c("ATRXdel","ATRXwt"))

# 6. Define the sender and reciever cell types of interest
# In our case NK and NB clusters
senders_oi = c("NB..8.","NK..hNK_Bm3.","NK..hNK_Bm1.","NK..hNK_Bm2.","NK..hNK_Bm4.")
receivers_oi = c("NB..8.","NK..hNK_Bm3.","NK..hNK_Bm1.","NK..hNK_Bm2.","NK..hNK_Bm4.")
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)]

# 6. Define which conditions to use for the analysis
conditions_keep = c("ATRXdel","ATRXwt") # in this case we keep all samples
sce = sce[, SummarizedExperiment::colData(sce)[,group_id] %in% 
            conditions_keep]

# 7. MultiNicheNet analysis
## parameters
min_cells = 10
min_sample_prop = 0.50 # genes should be expressed in at least 50% of the samples in the group
fraction_cutoff = 0.05 # genes should show non-zero expression values in at least 5% of cells in a sample
empirical_pval = FALSE
logFC_threshold = 0.50     
p_val_threshold = 0.05
p_val_adj = FALSE 
top_n_target = 250
verbose = TRUE
n.cores = 8
scenario = "regular" 
ligand_activity_down = FALSE 
save = TRUE

# core analysis
multinichenet_output = multi_nichenet_analysis(
  sce = sce, 
  celltype_id = celltype_id, 
  sample_id = sample_id, 
  group_id = group_id, 
  batches = batches, 
  covariates = covariates, 
  lr_network = lr_network, 
  ligand_target_matrix = ligand_target_matrix, 
  contrasts_oi = contrasts_oi, 
  contrast_tbl = contrast_tbl, 
  senders_oi = senders_oi, 
  receivers_oi = receivers_oi,
  min_cells = min_cells, 
  fraction_cutoff = fraction_cutoff, 
  min_sample_prop = min_sample_prop,
  scenario = scenario, 
  ligand_activity_down = ligand_activity_down,
  logFC_threshold = logFC_threshold, 
  p_val_threshold = p_val_threshold, 
  p_val_adj = p_val_adj, 
  empirical_pval = empirical_pval, 
  top_n_target = top_n_target, 
  n.cores = n.cores, 
  verbose = TRUE
  )

# "NK..hNK_Bm4." removed from analysis because it is not present in all samples
# "NK..hNK_Bm1." & "NK..hNK_Bm2." removed from analysis because there are 
# not enough cells present in all comparison groups! (Mainly due to ATRXdel)

if(save == TRUE){
  saveRDS(multinichenet_output, paste0(params$out,"Rds/multiNicheNet_NB-NK-axis_ATRXdel-vs-ATRXwt.Rds"))
}
```

```{r MultiNicheNet figures: ATRXdel vs ATRXwt, fig.height=15, fig.width=15}
multinichenet_output <- readRDS(paste0(params$out,"Rds/multiNicheNet_NB-NK-axis_ATRXdel-vs-ATRXwt.Rds"))

# Visualization of differential cell-cell interactions
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 100, 
  rank_per_group = TRUE
  )

prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

# save results
output_ATRXdelvsATRXwt <- as.data.frame(prioritized_tbl_oi)
write.table(output_ATRXdelvsATRXwt, 
            file = paste0(params$out,"tables/multiNicheNet_LR_TOP100_NB-NK-axis_ATRXdel-vs-ATRXwt.csv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), 
                          prioritized_tbl_oi$receiver %>% unique()) %>% sort()

prioritized_tbl_oi_filtered <- prioritized_tbl_oi %>%
  filter(sender != receiver) # removing CCC within the same group

table(prioritized_tbl_oi_filtered$group)

colors_sender = NULL
colors_receiver = NULL

sender_oi <- c("NK..hNK.Bm1.","NK..hNK.Bm2.","NK..hNK.Bm3.","NK..hNK.Bm4.",
               "NB..8.")

receiver_oi <- c("NK..hNK.Bm1.","NK..hNK.Bm2.","NK..hNK.Bm3.","NK..hNK.Bm4.",
               "NB..8.")

# COLOR_CODE
colors_sender["NK..hNK_Bm1."]  = "#6376AE"
colors_sender["NK..hNK_Bm2."]  = "#679F8F"
colors_sender["NK..hNK_Bm3."]  = "#BF4E4E"
colors_sender["NK..hNK.Bm4."]  = "#CDA646"
colors_sender["NB..8."]        = "#520c61"

colors_receiver["NK..hNK_Bm1."]  = "#6376AE"
colors_receiver["NK..hNK_Bm2."]  = "#679F8F"
colors_receiver["NK..hNK_Bm3."]  = "#BF4E4E"
colors_receiver["NK..hNK_Bm4."]  = "#CDA646"
colors_receiver["NB..8."]        = "#520c61"


circos_list = make_circos_group_comparison(prioritized_tbl_oi, 
                                           colors_sender, 
                                          colors_receiver)

pdf(paste0(params$out,"figures/MultiNicheNet_circos_plot_ATRXdel-vs-ATRXwt_top50.pdf"),
    height = 8, width = 8)
circos_list$ATRXdel
circos_list$ATRXwt
dev.off()


####################### ATRXwt patient group #################################
group_oi = "ATRXwt" 

# R-L expression
prioritized_tbl_oi_M_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 300, 
  rank_per_group = TRUE,
  groups_oi = group_oi)

filter_group <- prioritized_tbl_oi_filtered$group == "ATRXwt"
prioritized_ATRXwt_filtered <- prioritized_tbl_oi_filtered[filter_group,]

prioritized_ATRXwt_filtered <- prioritized_ATRXwt_filtered %>%
  filter(sender != receiver) # removing CCC within the same group

prioritized_ATRXwt_filtered <- head(prioritized_ATRXwt_filtered, n = 51)

plot_oi_ATRXwt = make_sample_lr_prod_activity_plots(
  multinichenet_output$prioritization_tables, 
  prioritized_ATRXwt_filtered)

plot_oi_ATRXwt 
# Samples that were left out of the DE analysis are indicated with a smaller dot 
# (this helps to indicate the samples that did not contribute to the calculation 
# of the logFC, and thus not contributed to the final prioritization)

# save plots
pdf(paste0(params$out,"figures/multiNicheNet_ATRXwt_top50.pdf"),
    height = 12, width = 14,pointsize = 1)
plot_oi_ATRXwt
dev.off()

png(paste0(params$out,"figures/multiNicheNet_ATRXwt_top50.png"),
   height = 1700, width = 2000, units = "px", pointsize = 1, res = 150)
plot_oi_ATRXwt 
dev.off()

tiff(paste0(params$out,"figures/multiNicheNet_ATRXwt_top50.tiff"),
    height = 1700, width = 2000, units = "px", res = 150, pointsize = 1)
plot_oi_ATRXwt 
dev.off() 

############################## ATRXdel patient group #############################
group_oi = "ATRXdel"

prioritized_tbl_oi_M_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 300,  
  rank_per_group = TRUE,
  groups_oi = group_oi,
  receivers_oi = clusters_of_interest,
  senders_oi = clusters_of_interest)

prioritized_tbl_oi_M_50 <- prioritized_tbl_oi_M_50 %>%
  filter(sender != receiver) # removing CCC within the same group
plot_oi_ATRXdel = make_sample_lr_prod_activity_plots(
  multinichenet_output$prioritization_tables,  prioritized_tbl_oi_M_50)

filter_group <- prioritized_tbl_oi_filtered$group == "ATRXdel"
prioritized_ATRXdel_filtered <- prioritized_tbl_oi_filtered[filter_group,]

prioritized_ATRXdel_filtered <- prioritized_ATRXdel_filtered %>%
  filter(sender != receiver) # removing CCC within the same group

prioritized_ATRXdel_filtered <- head(prioritized_ATRXdel_filtered, n = 50)

plot_oi_ATRXdel = make_sample_lr_prod_activity_plots(
  multinichenet_output$prioritization_tables, 
  prioritized_ATRXdel_filtered)

plot_oi_ATRXdel

# save plots
pdf(paste0(params$out,"figures/multiNicheNet_ATRXdel_top50.pdf"),
    height = 12,width = 14,pointsize = 1)
plot_oi_ATRXdel
dev.off()

png(paste0(params$out,"figures/multiNicheNet_ATRXdel_top50.png"),
   height = 1700, width = 2000, units = "px", pointsize = 1, res = 150)
plot_oi_ATRXdel 
dev.off()

tiff(paste0(params$out,"figures/multiNicheNet_ATRXdel_top50.tiff"),
    height = 1700, width = 2000, units = "px", res = 150, pointsize = 1)
plot_oi_ATRXdel 
dev.off() 


# Filtering of target genes based on LR-target expression correlation
top_n_target = 100

contrast_tbl = tibble(contrast =  c("ATRXdel-ATRXwt","ATRXdel-ATRXwt"), 
                      group = c("ATRXdel","ATRXwt"))

lr_target_prior_cor_filtered = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl$group %>% unique() %>% 
  lapply(function(group_oi){
    lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
      inner_join(
        multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>%
          distinct(ligand, target, direction_regulation, contrast)
        ) %>% 
      inner_join(contrast_tbl) %>% filter(group == group_oi)
    
    lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% 
      filter(direction_regulation == "up") %>% 
      filter( (rank_of_target < top_n_target) & (pearson > 0.5))
    
    lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% 
      filter(direction_regulation == "down") %>% 
      filter( (rank_of_target < top_n_target) & (pearson < -0.5))
    lr_target_prior_cor_filtered = bind_rows(
      lr_target_prior_cor_filtered_up, 
      lr_target_prior_cor_filtered_down
      )
}) %>% bind_rows()

# Network plot
lr_target_df = lr_target_prior_cor_filtered %>% 
  distinct(group, sender, receiver, ligand, receptor, id, target, direction_regulation) 

# increase the top CCC
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 100, 
  rank_per_group = TRUE 
  )

network = infer_intercellular_regulatory_network(lr_target_df, prioritized_tbl_oi_all)
network$links %>% head()

network_graph = visualize_network(network, colors_sender)

final_network <- network_graph$plot
final_network

pdf(paste0(params$out,"figures/multiNicheNet_networkGraph_NB-NK-axis_ATRXdel-vs-ATRXwt.pdf"),
    height = 8, width = 12)
final_network
dev.off()

tiff(paste0(params$out,"figures/multiNicheNet_networkGraph_NB-NK-axis_ATRXdel-vs-ATRXwt.tiff"),
     width = 265.9449, height = 265.9449,units = "in", res = 100)
final_network
dev.off()

output_ATRXdelvsATRXwt <- as.data.frame(prioritized_tbl_oi)
write.table(output_ATRXdelvsATRXwt, 
            file = paste0(params$out,"tables/multiNicheNet_output_NB-NK-axis_ATRXdel-vs-ATRXwt.csv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

## ___ATRXdel vs. ATRXwtMYCNwt____

MYCNamp samples excluded

```{r MultiNicheNet analysis: ARTX deleted vs. ATRXwtMYCNwt}
# 1. Change to the new group identitites
Idents(fetahu2023) <- fetahu2023@meta.data$group
  
# 2. Convert to single cell object
sce = Seurat::as.SingleCellExperiment(fetahu2023, assay = "RNA")

# 3. Define which columns in the metadata contain the neccessary information
SummarizedExperiment::colData(sce)$rados_et_al_2025.x = SummarizedExperiment::colData(sce)$rados_et_al_2025.x %>% make.names()
SummarizedExperiment::colData(sce)$group = SummarizedExperiment::colData(sce)$group %>% make.names()
SummarizedExperiment::colData(sce)$cell_type_detailed = SummarizedExperiment::colData(sce)$cell_type_detailed %>% make.names()

sample_id = "rados_et_al_2025.x"
group_id = "group" # change to ATRX_groups
celltype_id = "cell_type_detailed"

# 4. Set covariates and batches
covariates = NA
batches = NA 
 
# 5. Set comparison groups
contrasts_oi = c("'ATRXdel-ATRXwtMYCNwt','ATRXwtMYCNwt-ATRXdel'")

# Note the format to indicate the contrasts! This formatting should be adhered to very strictly, and white spaces are not allowed! Check ?get_DE_info for explanation about how to define this well. The most important points are that: *each contrast is surrounded by single quotation marks *contrasts are separated by a comma without any white space *all contrasts together are surrounded by double quotation marks.

contrast_tbl = tibble(contrast =  c("ATRXdel-ATRXwtMYCNwt","ATRXwtMYCNwt-ATRXdel"), 
                      group = c("ATRXdel","ATRXwtMYCNwt"))

# 6. Define the sender and receiver cell types of interest
# In our case NB and NK clusters
senders_oi = c("NB..8.","NK..hNK_Bm3.","NK..hNK_Bm1.","NK..hNK_Bm2.","NK..hNK_Bm4.")
receivers_oi = c("NB..8.","NK..hNK_Bm3.","NK..hNK_Bm1.","NK..hNK_Bm2.","NK..hNK_Bm4.")
sce = sce[, SummarizedExperiment::colData(sce)[,celltype_id] %in% 
            c(senders_oi, receivers_oi)]

# 6. Define which conditions to use for the analysis
conditions_keep = c("ATRXdel","ATRXwtMYCNwt") 
sce = sce[, SummarizedExperiment::colData(sce)[,group_id] %in% 
            conditions_keep]

# 7. MultiNicheNet analysis

# parameters
min_cells = 10
min_sample_prop = 0.50 # genes should be expressed in at least 50% of the samples in the group
fraction_cutoff = 0.05 # genes should show non-zero expression values in at least 5% of cells in a sample
empirical_pval = FALSE
logFC_threshold = 0.50      # set based on the histogram distribution
p_val_threshold = 0.05
p_val_adj = FALSE 
top_n_target = 250
verbose = TRUE
n.cores = 1
scenario = "regular" 
ligand_activity_down = TRUE 
save = TRUE

# core analysis
multinichenet_output = multi_nichenet_analysis(
  sce = sce, 
  celltype_id = celltype_id, sample_id = sample_id, group_id = group_id, 
  batches = batches, covariates = covariates, 
  lr_network = lr_network, ligand_target_matrix = ligand_target_matrix, 
  contrasts_oi = contrasts_oi, contrast_tbl = contrast_tbl, 
  senders_oi = senders_oi, receivers_oi = receivers_oi,
  min_cells = min_cells, 
  fraction_cutoff = fraction_cutoff, 
  min_sample_prop = min_sample_prop,
  scenario = scenario, 
  ligand_activity_down = ligand_activity_down,
  logFC_threshold = logFC_threshold, 
  p_val_threshold = p_val_threshold, 
  p_val_adj = p_val_adj, 
  empirical_pval = empirical_pval, 
  top_n_target = top_n_target, 
  n.cores = n.cores, 
  verbose = TRUE
  )

if(save == TRUE){
  saveRDS(multinichenet_output, paste0(params$out,"Rds/multiNicheNet_NB-NK-axis_ATRXdel-vs-ATRXwtMYCNwt.Rds"))
}
```

```{r MultiNicheNet figures: ATRXdel vs ATRXwtMYCNwt, fig.height=15, fig.width=15}
multinichenet_output <- readRDS(paste0(params$out,"Rds/multiNicheNet_NB-NK-axis_ATRXdel-vs-ATRXwtMYCNwt.Rds"))

# Visualization of differential cell-cell interactions
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 100, 
  rank_per_group = TRUE
  )

prioritized_tbl_oi = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  filter(id %in% prioritized_tbl_oi_all$id) %>%
  distinct(id, sender, receiver, ligand, receptor, group) %>% 
  left_join(prioritized_tbl_oi_all)
prioritized_tbl_oi$prioritization_score[is.na(prioritized_tbl_oi$prioritization_score)] = 0

# save results
output_ATRXdelvsATRXwt <- as.data.frame(prioritized_tbl_oi)
write.table(output_ATRXdelvsATRXwt, 
            file = paste0(params$out,"tables/multiNicheNet_LR_TOP100_ATRXdel-vs-ATRXwtMYCNwt.csv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

senders_receivers = union(prioritized_tbl_oi$sender %>% unique(), 
                          prioritized_tbl_oi$receiver %>% unique()) %>% sort()

prioritized_tbl_oi_filtered <- prioritized_tbl_oi %>%
  filter(sender != receiver) # removing CCC within the same group

table(prioritized_tbl_oi_filtered$group)

colors_sender = NULL
colors_receiver = NULL

# COlOR code
colors_sender["NK..hNK.Bm1."]  = "#6376AE"
colors_sender["NK..hNK.Bm2."]  = "#679F8F"
colors_sender["NK..hNK.Bm3."]  = "#BF4E4E"
colors_sender["NK..hNK.Bm4."]  = "#CDA646"
colors_sender["NB..8."]        = "#520c61"

colors_receiver["NK..hNK.Bm1."]  = "#6376AE"
colors_receiver["NK..hNK.Bm2."]  = "#679F8F"
colors_receiver["NK..hNK.Bm3."]  = "#BF4E4E"
colors_receiver["NK..hNK.Bm4."]  = "#CDA646"
colors_receiver["NB..8."]        = "#520c61"


circos_list = make_circos_group_comparison(prioritized_tbl_oi, 
                                           colors_sender, 
                                           colors_receiver)

pdf(paste0(params$out,"figures/MultiNicheNet_circos_plot_ATRXdel-vs-ATRXwtMYCNwt_top50.pdf"),
    height = 8, width = 8)
circos_list
dev.off()

####################### ATRXwtMYCNwt patient group #################################

group_oi = "ATRXwtMYCNwt" 

# R-L expression
prioritized_tbl_oi_M_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 100, 
  rank_per_group = TRUE,
  groups_oi = group_oi)

filter_group <- prioritized_tbl_oi_filtered$group == "ATRXwtMYCNwt"
prioritized_ATRXwt_filtered <- prioritized_tbl_oi_filtered[filter_group,]

prioritized_ATRXwt_filtered <- prioritized_ATRXwt_filtered %>%
  filter(sender != receiver) # removing CCC within the same group

prioritized_ATRXwt_filtered <- head(prioritized_ATRXwt_filtered, n = 50)

plot_oi_ATRXwt = make_sample_lr_prod_activity_plots(
  multinichenet_output$prioritization_tables, 
  prioritized_ATRXwt_filtered)

plot_oi_ATRXwt 
# Samples that were left out of the DE analysis are indicated with a smaller dot 
# (this helps to indicate the samples that did not contribute to the calculation 
# of the logFC, and thus not contributed to the final prioritization)

# save plots
pdf(paste0(params$out,"figures/multiNicheNet_ATRXwtMYCNwt_top50.pdf"),
    height = 12, width = 14,pointsize = 1)
plot_oi_ATRXwt
dev.off()

png(paste0(params$out,"figures/multiNicheNet_ATRXwtMYCNwt_top50.png"),
   height = 1700, width = 2000, units = "px", pointsize = 1, res = 150)
plot_oi_ATRXwt 
dev.off()

############################## ATRXdel patient group #############################

group_oi = "ATRXdel"

prioritized_tbl_oi_M_50 = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 100,  
  rank_per_group = TRUE,
  groups_oi = group_oi,
  receivers_oi = clusters_of_interest,
  senders_oi = clusters_of_interest)

#prioritized_tbl_oi_M_50 <- prioritized_tbl_oi_M_50 %>%
#  filter(sender != receiver) # removing CCC within the same group

#plot_oi_ATRXdel = make_sample_lr_prod_activity_plots(
#  multinichenet_output$prioritization_tables, 
#  prioritized_tbl_oi_M_50)

filter_group <- prioritized_tbl_oi_filtered$group == "ATRXdel"
prioritized_ATRXdel_filtered <- prioritized_tbl_oi_filtered[filter_group,]

prioritized_ATRXdel_filtered <- prioritized_ATRXdel_filtered %>%
  filter(sender != receiver) # removing CCC within the same group

prioritized_ATRXdel_filtered <- head(prioritized_ATRXdel_filtered, n = 50)

plot_oi_ATRXdel = make_sample_lr_prod_activity_plots(
  multinichenet_output$prioritization_tables, 
  prioritized_ATRXdel_filtered)

plot_oi_ATRXdel

# save plots
pdf(paste0(params$out,"figures/multiNicheNet_ATRXdel_top50_v2.pdf"),
    height = 12,width = 14,pointsize = 1)
plot_oi_ATRXdel
dev.off()

png(paste0(params$out,"figures/multiNicheNet_ATRXdel_top50_v2.png"),
   height = 1700, width = 2000, units = "px", pointsize = 1, res = 150)
plot_oi_ATRXdel 
dev.off()

# Filtering of target genes based on LR-target expression correlation
top_n_target = 300

contrast_tbl = tibble(contrast =  c("ATRXdel-ATRXwtMYCNwt","ATRXwtMYCNwt-ATRXdel"), 
                      group = c("ATRXdel","ATRXwtMYCNwt"))

lr_target_prior_cor_filtered = 
  multinichenet_output$prioritization_tables$group_prioritization_tbl$group %>% unique() %>% 
  lapply(function(group_oi){
    lr_target_prior_cor_filtered = multinichenet_output$lr_target_prior_cor %>%
      inner_join(
        multinichenet_output$ligand_activities_targets_DEgenes$ligand_activities %>%
          distinct(ligand, target, direction_regulation, contrast)
        ) %>% 
      inner_join(contrast_tbl) %>% filter(group == group_oi)
    
    lr_target_prior_cor_filtered_up = lr_target_prior_cor_filtered %>% 
      filter(direction_regulation == "up") %>% 
      filter( (rank_of_target < top_n_target) & (pearson > 0.5))
    
    lr_target_prior_cor_filtered_down = lr_target_prior_cor_filtered %>% 
      filter(direction_regulation == "down") %>% 
      filter( (rank_of_target < top_n_target) & (pearson < -0.5))
    lr_target_prior_cor_filtered = bind_rows(
      lr_target_prior_cor_filtered_up, 
      lr_target_prior_cor_filtered_down
      )
}) %>% bind_rows()

# Network plot
lr_target_df = lr_target_prior_cor_filtered %>% 
  distinct(group, sender, receiver, ligand, receptor, id, target, direction_regulation) 

# increase the top CCC
prioritized_tbl_oi_all = get_top_n_lr_pairs(
  multinichenet_output$prioritization_tables, 
  top_n = 150, 
  rank_per_group = TRUE 
  )

network = infer_intercellular_regulatory_network(lr_target_df, prioritized_tbl_oi_all)
network$links %>% head()

network_graph = visualize_network(network, colors_sender)

final_network <- network_graph$plot
final_network

pdf(paste0(params$out,"figures/multiNicheNet_networkGraph_ATRXdel-vs-ATRXwtMYCNwt.pdf"),
    height = 8, width = 12)
final_network
dev.off()

tiff(paste0(params$out,"figures/multiNicheNet_networkGraph_ATRXdel-vs-ATRXwtMYCNwt.tiff"),
     width = 265.9449, height = 265.9449,units = "in", res = 100)
final_network
dev.off()

output_ATRXdelvsATRXwtMYCNwt <- as.data.frame(prioritized_tbl_oi)
write.table(output_ATRXdelvsATRXwtMYCNwt, 
            file = paste0(params$out,"tables/multiNicheNet_output_ATRXdel-vs-ATRXwtMYCNwt.csv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
```

# ___CCC analysis w/cellchat: NB - NK axis___

To infer the cell state-specific communications, CellChat v2 identifies 
over-expressed ligands or receptors in one cell group and then identifies 
over-expressed ligand-receptor interactions if either ligand or receptor are 
over-expressed.

```{r}
library(CellChat)
```

```{r}
# define which clusters to compare
clusters_of_interest <- c("NK (hNK_Bm1)",
                          "NK (hNK_Bm2)",
                          "NK (hNK_Bm3)",
                          "NK (hNK_Bm4)",
                          "NB (8)") 

# subset object to clusters of interest
NK_and_NB <- subset(fetahu2023, idents = clusters_of_interest)
```

```{r}
Idents(NK_and_NB) <- NK_and_NB$group

NK_and_NB$cell_type_detailed <- droplevels(NK_and_NB$cell_type_detailed)

ATRXdel <- subset(NK_and_NB, idents = "ATRXdel")
ATRXwt <- subset(NK_and_NB, idents = c("ATRXwtMYCNwt","MYCNamp"))

# separate MYCNamp and ATRXwtMYCNwt sampples
MYCNamp <- subset(NK_and_NB, idents = "MYCNamp")
ATRXwtMYCNwt <- subset(NK_and_NB, idents = "ATRXwtMYCNwt")
```

```{r ccc function - advanced}
ccc.analysis.comprehensive <- function(seurat) {
  n = n + 1
  print(seurat)
  print("Loop number: ", n)
  
  Idents(seurat) <- seurat@meta.data$cell_type_detailed
  # extract read count matrix
  mat <- GetAssayData(seurat, assay = "RNA", slot = "data")
  # create cell chat object
  cellchat <- createCellChat(mat, 
                             meta = seurat@meta.data,
                             group.by = "cell_type_detailed",
                             )
  
  
  # Prepare DB
  cellchat@DB <- CellChatDB.human
  # subset the expression data of signaling genes for saving computation cost
  cellchat <- subsetData(cellchat) # This step is necessary even if using the whole database (it removes genes which are not expressed in the data)
  cellchat <- identifyOverExpressedGenes(cellchat)
  cellchat <- identifyOverExpressedInteractions(cellchat)
  cellchat <- projectData(cellchat, PPI.human)
  # Compute the communication probability and infer cellular communication network
  # project gene expression data onto PPI (Use `raw.use = FALSE` in the function 
  # computeCommunProb() in order to use the projected data)
  cellchat <- computeCommunProb(cellchat, raw.use = FALSE, population.size = TRUE)
  # population.size = TRUE <- cell numbers per cell type/cluster 
  # used for the probability calculation
  # only consider R-L pairs which are expressed by minumum 50 cells
  cellchat <- filterCommunication(cellchat, min.cells = 50)
  df.net <- subsetCommunication(cellchat)
  df.net.pathways <- subsetCommunication(cellchat, slot.name = "netP")
  # slot.name = "netP" for inferred communications at the level of signaling pathways
  print(df.net.pathways)
  # Infer the cell-cell communication at a signaling pathway level
  cellchat <- computeCommunProbPathway(cellchat)
  # Calculate the aggregated cell-cell communication network
  cellchat <- aggregateNet(cellchat)
  groupSize <- as.numeric(table(cellchat@idents))
  
  circo_plot <- netVisual_circle(cellchat@net$weight, 
                                 vertex.weight = groupSize, 
                                 weight.scale = T, 
                                 label.edge = F, 
                                 title.name = "CC interaction strength",
                                 color.use = COLOR_CODE_CCC)
  print(circo_plot)
  
  pdf(paste0(params$out,"figures/Figure7-B_final_",n,".pdf"),
    height = 5, width = 8)
  circo_plot
  dev.off()
  
  # Bubble plot: all the significant interactions (Ligand-Receptor pairs) from 
  # one cell groups (defined by 'sources.use') to other cell groups (defined by 'targets.use')
  netVisual_bubble(cellchat, sources.use = "NB (8)",  
                   remove.isolate = FALSE)
  netVisual_chord_gene(cellchat, sources.use = "NB (8)", 
                       slot.name = "netP", 
                       legend.pos.x = 10)
  # Compute the network centrality scores
  cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP") 
  # the slot 'netP' means the inferred intercellular communication network of signaling pathways
  # Visualize the computed centrality scores using heatmap, allowing ready identification of major signaling roles of cell groups
  #netAnalysis_signalingRole_network(cellchat, signaling = pathway, 
  #                                  width = 8, height = 2.5, font.size = 12)
  # Visualize dominant senders (sources) and receivers (targets) in a 2D space
  gg1 <- netAnalysis_signalingRole_scatter(cellchat)
  # Signaling role analysis on the cell-cell communication networks of interest
  #gg2 <- netAnalysis_signalingRole_scatter(cellchat, signaling = pathway)
  # Signaling role analysis on the cell-cell communication network from user's input
  print(gg1)
  #gg2
  # Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
  ht1 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing",color.use = COLOR_CODE_CCC)
  ht2 <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming",color.use = COLOR_CODE_CCC)
  print(ht1 + ht2)
  # run selectK to infer the number of patterns.
  # selectK(cellchat, pattern = "outgoing") # define the number of patterns/cut-off
  nPatterns = 7 # based on the output of the previous command
  cellchat <- identifyCommunicationPatterns(cellchat, 
                                            pattern = "outgoing", 
                                            k = nPatterns)
  # river plot
  netAnalysis_river(cellchat, pattern = "outgoing",color.use = COLOR_CODE_CCC)
  # dot plot
  netAnalysis_dot(cellchat, pattern = "outgoing",color.use = COLOR_CODE_CCC)
  # the same for incoming communication patterns
  #selectK(cellchat, pattern = "incoming") # define the number of patterns/cut-off
  nPatterns = 7
  cellchat <- identifyCommunicationPatterns(cellchat, 
                                            pattern = "incoming", 
                                            k = nPatterns)
  # river plot
  netAnalysis_river(cellchat, pattern = "incoming",color.use = COLOR_CODE_CCC)
  # dot plot
  netAnalysis_dot(cellchat, pattern = "incoming",color.use = COLOR_CODE_CCC)
  #saveRDS(cellchat, file = "/home/out/MapMetRds/CCC_analysis_Fetahu2023.Rds"))
  results_network <- subsetCommunication(cellchat, slot.name = "net")
  results_pathways <- subsetCommunication(cellchat, slot.name = "netP")
  
  df_net_all <-
  cellchat %>% 
  subsetCommunication() %>% 
  as_tibble() %>% 
  group_by(interaction_name_2) %>% 
  mutate(prob.norm = prob / max(prob)) %>% # Normalize the probabilities 0-1
  ungroup() %>% 
  mutate(
    source = factor(source, clusters_of_interest),
    target = factor(target, clusters_of_interest),
    )
  
  print(df_net_all)
}
```

```{r Run comprehensive CCC analysis}
n = 0

sample.list <- list(ATRXdel, ATRXwt)
setNames(sample.list, c("ATRXdel", "ATRXwt"))

CCC_analysis_full <- lapply(sample.list, ccc.analysis.comprehensive)

saveRDS(CCC_analysis_full, file = paste0(params$out,"Rds/CCC_analysis_full.Rds"))
```

```{r set a cutoff II}
# set a cutoff
hist(CCC_analysis_full[[1]]$prob.norm)
hist(CCC_analysis_full[[2]]$prob.norm)
```

```{r Save CCC comprehensive results}
CCC_analysis_full[[1]]  # ATRXdel
write.csv(CCC_analysis_full[[1]], file = "/home/out/MapMet/tables/CellChat_ATRXdel_CCC_advanced.csv",
          quote = FALSE,sep = "\t",row.names = FALSE)
CCC_analysis_full[[2]]  # ATRXwt
write.csv(CCC_analysis_full[[1]], file = "/home/out/MapMet/tables/CellChat_ATRXwt_CCC_advanced.csv",
          quote = FALSE,sep = "\t",row.names = FALSE)
```

```{r ccc function for the publication}
ccc.publication <- function(seurat) {
  count <- count + 1
  print(seurat)
  print(paste0("Loop number: ", count))
  
  Idents(seurat) <- seurat@meta.data$cell_type_detailed
  # extract read count matrix
  mat <- GetAssayData(seurat, assay = "RNA", slot = "data")
  # create cell chat object
  cellchat <- createCellChat(mat, 
                             meta = seurat@meta.data,
                             group.by = "cell_type_detailed",
                             )

  cellchat@DB <- CellChatDB.human
  cellchat <- subsetData(cellchat) 
  cellchat <- identifyOverExpressedGenes(cellchat)
  cellchat <- identifyOverExpressedInteractions(cellchat)
  cellchat <- computeCommunProb(cellchat, 
                                raw.use = TRUE, 
                                population.size = TRUE)
  cellchat <- filterCommunication(cellchat, min.cells = 50)
  df.net <- subsetCommunication(cellchat)
  df.net.pathways <- subsetCommunication(cellchat, slot.name = "netP")
  print(df.net.pathways)
  cellchat <- computeCommunProbPathway(cellchat)
  cellchat <- aggregateNet(cellchat)
  groupSize <- as.numeric(table(cellchat@idents))
  circo_plot <- netVisual_circle(cellchat@net$weight, 
                                 vertex.weight = groupSize, 
                                 weight.scale = T, 
                                 label.edge = F,
                                 title.name = "CC interaction strength",
                                 color.use = COLOR_CODE_CCC)
  # colors are set based on the position of the cluster within the data structure!
  pdf(paste0(params$out,"figures/CellChat_circos_plot_",count,".pdf"))
  circo_plot
  dev.off()
  
  print(circo_plot)
}
```

```{r final circos plots}
count = 0

sample.list <- list(ATRXdel, ATRXwt, MYCNamp, ATRXwtMYCNwt)
setNames(sample.list, c("ATRXdel", "ATRXwt", "MYCNamp","ATRXwtMYCNwt"))

CCC_analysis_full <- lapply(sample.list, ccc.publication)

saveRDS(CCC_analysis_full, paste0(params$out,"Rds/cellchat_circos_plots_ATRXdel_vs_ATRXwt.Rds"))
```

```{r}
pdf(paste0(params$out,"figures/CellChat_circos_plot_","ATRXdel",".pdf"))
CCC_analysis_full[[1]]
dev.off()

pdf(paste0(params$out,"figures/CellChat_circos_plot_","ATRXwt",".pdf"))
CCC_analysis_full[[2]]
dev.off()
```


```{r final circos plots}
count = 0

sample.list <- list(MYCNamp, ATRXwtMYCNwt)
setNames(sample.list, c("MYCNamp","ATRXwtMYCNwt"))

CCC_analysis_full <- lapply(sample.list, ccc.publication)

saveRDS(CCC_analysis_full, paste0(params$out,"Rds/cellchat_circos_plots_MYCNamp_ATRXwtMYCNwt.Rds"))
```

```{r}
pdf(paste0(params$out,"figures/CellChat_circos_plot_","MYCNamp",".pdf"))
CCC_analysis_full[[1]]
dev.off()

pdf(paste0(params$out,"figures/CellChat_circos_plot_","ATRXwtMYCNwt",".pdf"))
CCC_analysis_full[[2]]
dev.off()
```


# ___SessionInfo___

```{r}
sessionInfo()
```



