---
title: "NK project: DEG analysis"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/CancReg/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_BM2: "/home/rstudio/mnt_data/CancReg/BM/yang2019/Rds/Human_BM_NK_seurat_v4.Rds"
  input_scRNA: "/home/rstudio/mnt_out/NK_project/Rds/fetahu2023_withNewMetadata.Rds"
  input_NK: "/home/rstudio/mnt_out/NK_project/Rds/ALL_NK_cluster_withNewMetadata.Rds"
  input_NB: "/home/rstudio/mnt_out/NK_project/Rds/ALL_NB_cluster_withNewMetadata.Rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
  metadata: "/home/rstudio/NK_project/metadata/"
---

PROJECT DESCRIPTION

Identify NK sub-populations in tumor-infiltrated bone marrow rados_et_al_2025 from
Neuroblastoma patients.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r echo=FALSE}
source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")
```

# ___Load data___

```{r eval=FALSE, include=FALSE}
install.packages('pheatmap')
install.packages("DOSE")
remotes::install_github("GuangchuangYu/enrichplot")
install.packages("ggupset")
```

```{r}
suppressPackageStartupMessages(library('pheatmap'))
suppressPackageStartupMessages(library("DOSE"))
suppressPackageStartupMessages(library("enrichplot"))
suppressPackageStartupMessages(library("ggupset"))
suppressPackageStartupMessages(library("EnhancedVolcano"))
suppressPackageStartupMessages(library("ggVennDiagram"))
library("enrichplot")
```

```{r}
ALL_NK <-  readRDS(file = params$input_NK)
```

```{r eval=FALSE, include=FALSE}
# Only run one time
fetahu2023 <-  readRDS(file = params$input_scRNA)
Idents(fetahu2023) <- fetahu2023$orig.ident
ALL_NB <- subset(fetahu2023, idents = "NB (8)") # 4925 cells

ALL_NB$group <- factor(ALL_NB$group)
Idents(ALL_NB) <- ALL_NB@meta.data$group

# identify cell IDs associated with each cluster
M <- WhichCells(ALL_NB, ident = "MYCNamp")
S <- WhichCells(ALL_NB, ident = "ATRXwtMYCNwt")
A <- WhichCells(ALL_NB, ident = "ATRXdel")

# re-label the NK cells to the corresponding predicted sub-clusters 
Idents(object = ALL_NB, cells = M) <- "ATRXwt"
Idents(object = ALL_NB, cells = S) <- "ATRXwt"
Idents(object = ALL_NB, cells = A) <- "ATRXdel"

# Add active identities to metadata column comparison_group
ALL_NB$ATRXdel_vs_ATRXwt <- ALL_NB@active.ident

saveRDS(ALL_NB, file = paste0(params$out, "Rds/ALL_NB_cluster_withNewMetadata.Rds"))
```

```{r}
ALL_NB <-  readRDS(file = params$input_NB)
```


# ___NK cells___

```{r}
ALL_NK$group <- factor(ALL_NK$group)

Idents(ALL_NK) <- ALL_NK@meta.data$group

# identify cell IDs associated with each cluster
M <- WhichCells(ALL_NK, ident = "MYCNamp")
S <- WhichCells(ALL_NK, ident = "ATRXwtMYCNwt")
C <- WhichCells(ALL_NK, ident = "control")
A <- WhichCells(ALL_NK, ident = "ATRXdel")

# re-label the NK cells to the corresponding predicted sub-clusters 
Idents(object = ALL_NK, cells = M) <- "ATRXwt"
Idents(object = ALL_NK, cells = S) <- "ATRXwt"
Idents(object = ALL_NK, cells = A) <- "ATRXdel"
Idents(object = ALL_NK, cells = C) <- "ATRXwt"

# Add active identities to metadata column comparison_group
ALL_NK$ATRXdel_vs_rest <- ALL_NK@active.ident
```

## ___DEG analysis___

Replicates set to biological replicates with the rados_et_al_2025 slot in the metadata
<br>
ATRXdel, n = 3
<br>
MYCNamp, n = 4
<br>
ATRXwtMYCNwt, n = 4


```{r}
# Access the counts matrix
counts_matrix <- GetAssayData(ALL_NK, slot = "counts")

# set a meaningful cutoff for expressed genes 
hist(Matrix::rowSums(counts_matrix),breaks = 100000000,xlim = (c(0,100)))
abline(v=2)

hist(log(Matrix::rowSums(counts_matrix)),breaks = 1000)

# Filter genes with at least 11 reads over all samples (rados_et_al_2025)
# these will be used as background genes for pathways enrichment analysis
geneUniverse <- rownames(counts_matrix)[Matrix::rowSums(counts_matrix) > 20]
length(geneUniverse) # 20270 expressed genes

# justify chosen cut-off of expressed genes
hist(log(Matrix::rowSums(counts_matrix)),breaks = 500) 
abline(v=0.7)
abline(v=3)

# remap to ENTREZ IDs
geneUniverse_entrezIDs <- map_gene_to_entrez(geneUniverse)
```

### ___Cluster 2___

```{r}
# Change idents to cluster annotations
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 2
cluster_of_interest <- "hNK_Bm2"

# subset scRNA-seq data to NK and NB cells 
hNK_Bm2 <- subset(ALL_NK, 
                  idents = cluster_of_interest, 
                  invert = FALSE)
hNK_Bm2 # 2586 cells
```



#### ___1. Control vs. ATRXdel___

```{r}
# DE analysis for the hNK_Bm2 NK cluster
DEG_hNK_Bm2_DEG <- DElegate::findDE(hNK_Bm2,
                           group_column = "group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "rados_et_al_2025",
                           compare = "control")

control_vs_ATRXdel <- subset(DEG_hNK_Bm2_DEG, 
                        DEG_hNK_Bm2_DEG$group1 == "ATRXdel")

control_vs_MYCNamp <- subset(DEG_hNK_Bm2_DEG, 
                        DEG_hNK_Bm2_DEG$group1 == "MYCNamp")

control_vs_ATRXwtMYCNwt <- subset(DEG_hNK_Bm2_DEG, 
                        DEG_hNK_Bm2_DEG$group1 == "ATRXwtMYCNwt")

control_vs_ATRXdel_signif <- subset(control_vs_ATRXdel, 
                                control_vs_ATRXdel$padj < 0.1)
```

```{r}
write.table(control_vs_ATRXdel, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_control_vs_ATRXdel_2025-04-24.csv"))
```


#### ___2. ATRX wild-type [MYCNamp + ATRXwtMYCNwt] vs. ATRXdel___

```{r}
# DE analysis for the hNK_Bm2 NK cluster
ATRXwt_vs_ATRXdel <- DElegate::findDE(hNK_Bm2,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = FALSE,
                           replicate_column = "rados_et_al_2025",
                           compare = "ATRXdel")

ATRXwt_vs_ATRXdel_signif <- subset(ATRXwt_vs_ATRXdel, 
                                   ATRXwt_vs_ATRXdel$padj < 0.10)

ATRXwt_vs_ATRXdel_signif
```

```{r}
write.table(ATRXwt_vs_ATRXdel, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_ATRXdel-vs-ATRXwt_2025-04-24.csv"))
```


#### ___3. ALL vs. ATRXdel___

```{r}
ALL_vs_ATRXdel <- DElegate::findDE(hNK_Bm2,
                           group_column = "group",
                           order_results = TRUE,
                           compare_is_ref = FALSE,
                           replicate_column = "rados_et_al_2025",
                           compare = "ATRXdel")

ALL_vs_ATRXdel_signif <- subset(ALL_vs_ATRXdel, 
                                ALL_vs_ATRXdel$padj < 0.10)

ALL_vs_ATRXdel_signif
```

```{r}
write.table(ALL_vs_ATRXdel, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm2_ATRXdel-vs-ALL_2025-04-24.csv"))
```


### ___Vulcano plot___

#### ___1. control vs. ATRXdel___

```{r fig.height=7, fig.width=7}
control_vs_ATRXdel
rownames(control_vs_ATRXdel) <- control_vs_ATRXdel$feature

library(EnhancedVolcano)

pdf(paste0(params$out,"figures/DEGs_NK_cluster2_control_vs_ATRXdel_vulcano_plot_wGenes.pdf"),
    height = 5,width = 7)
EnhancedVolcano(control_vs_ATRXdel, 
                lab = rownames(control_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','TGFB1', 'F2R', 'PTPRJ'),
                subtitle = "control vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                ylim = c(0,4),
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()

pdf(paste0(params$out,"figures/DEGs_NK_cluster2_control_vs_ATRXdel_vulcano_plot.pdf"),
    height = 5,width = 7)
EnhancedVolcano(control_vs_ATRXdel, 
                lab = rownames(control_vs_ATRXdel),
                selectLab = NULL, # does this remove all gene names?
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "control vs. ATRXdel",
                pCutoff = 0.05,
                ylim = c(0,4),
                FCcutoff = 1,
                pointSize = 2,
                labSize = 6,
                col = c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()
```

#### ___2. ATRXwt vs. ATRXdel___

```{r fig.height=7, fig.width=7}
ATRXwt_vs_ATRXdel
rownames(ATRXwt_vs_ATRXdel) <- ATRXwt_vs_ATRXdel$feature

pdf(paste0(params$out,"figures/DEGs_NK_cluster2_ATRXwt_vs_ATRXdel_vulcano_plot_wGenes.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ATRXwt_vs_ATRXdel, 
                lab = rownames(ATRXwt_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','TGFB1', 'F2R', 'PTPRJ'),
                subtitle = "ATRXwt vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 2,
                ylim = c(0,4),
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()

pdf(paste0(params$out,"figures/DEGs_NK_cluster2_ATRXwt_vs_ATRXdel_vulcano_plot.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ATRXwt_vs_ATRXdel, 
                lab = rownames(ATRXwt_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "ATRXwt vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                ylim = c(0,4),
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()
```

#### ___3. ALL vs. ATRXdel___

```{r fig.height=7, fig.width=7}
rownames(ALL_vs_ATRXdel) <- ALL_vs_ATRXdel$feature

pdf(paste0(params$out,"figures/DEGs_NK_cluster2_ALL_vs_ATRXdel_vulcano_plot_wGenes.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ALL_vs_ATRXdel, 
                lab = rownames(ALL_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                selectLab = c('CD6','TGFB1', 'F2R', 'PTPRJ'),
                subtitle = "(ATRXwt + control) vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                ylim = c(0,6),
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()


pdf(paste0(params$out,"figures/DEGs_NK_cluster2_ALL_vs_ATRXdel_vulcano_plot.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ALL_vs_ATRXdel, 
                lab = rownames(ALL_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 2",
                subtitle = "(ATRXwt + control) vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                ylim = c(0,6),
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()
```

### ___Pathways enrichment analysis___

```{r}
control_vs_ATRXdel_signif <- control_vs_ATRXdel_signif %>%
  arrange(log_fc)

control_vs_ATRXdel_NK2_up <- control_vs_ATRXdel_signif %>% subset(log_fc > 1) %>% subset(padj < 0.05) 
control_vs_ATRXdel_NK2_down <- control_vs_ATRXdel_signif %>% subset(log_fc < -1) %>% subset(padj < 0.05) 

ATRXwt_vs_ATRXdel_signif <- ATRXwt_vs_ATRXdel %>%
  arrange(log_fc)

ATRXwt_vs_ATRXdel_NK2_up <- ATRXwt_vs_ATRXdel %>%  subset(log_fc > 1) %>% subset(padj < 0.05) 
ATRXwt_vs_ATRXdel_NK2_down <- ATRXwt_vs_ATRXdel %>%  subset(log_fc < -1) %>% subset(padj < 0.05) 

ALL_vs_ATRXdel_signif <- ALL_vs_ATRXdel_signif %>%
  arrange(log_fc)

ALL_vs_ATRXdel_NK2_up <- ALL_vs_ATRXdel_signif %>% subset(log_fc > 1) %>% subset(padj < 0.05) # up-regulated in ATRXdeleted rados_et_al_2025s 
ALL_vs_ATRXdel_NK2_down <- ALL_vs_ATRXdel_signif %>%  subset(log_fc < -1)  %>% subset(padj < 0.05) #down-regulated in ATRXdeleted rados_et_al_2025s
```

```{r}
hallmark_genes <- msigdbr(species = "human",category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)
```


#### ___control vs. ATRXdel___

```{r fig.height=5, fig.width=9}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- control_vs_ATRXdel_NK2_up$feature
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05,
              qvalueCutoff = 0.1,
               minGSSize = 2,
               maxGSSize = 5000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
              qvalueCutoff = 0.1,
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05,
              qvalueCutoff = 0.1,
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              pAdjustMethod = "fdr",
                              #universe = geneUniverse,
                              )

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.05,
                             qvalueCutoff = 0.10,
                             pAdjustMethod = "fdr",
                             ##universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)

# TREE PLOT
# calculate pairwise similarities of the enriched terms using Jaccard’s similarity index
enrichres2 <- pairwise_termsim(kegg_enrichment) 
treeplot_NK2_ctrl <- treeplot(enrichres2,
  showCategory = 15,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK2_ctrl

```

```{r}
# publication plot
pdf(paste0(params$out,"figures/DEG_hBm-NK2_ATRXdel_vs_CTRL_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK2_ctrl
dev.off()
```

___Gene Set Enrichment Analysis (GSEA)___

```{r}
genes <- control_vs_ATRXdel$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

control_vs_ATRXdel$entrez_ids <- entrez_ids
genes <- control_vs_ATRXdel$log_fc
names(genes) <- control_vs_ATRXdel$entrez_ids
genes = sort(genes, decreasing = TRUE)
genes <- genes[!is.na(names(genes))]
length(genes)

# filter by expressed in NK cells
overlapping_genes <- intersect(entrez_ids, geneUniverse_entrezIDs)
length(overlapping_genes) #12119

# Filter the logFC vector based on the gene list
filtered_genes <- genes[names(genes) %in% overlapping_genes]
length(filtered_genes)

# H gene set
hallmark_genes <- msigdbr(species = "human",category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes)

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

gseaplot(hallmark_gene_gsea, 
         by = "all", 
         title = hallmark_gene_gsea$Description[1], 
         geneSetID = 1)

# C5 gene set
hallmark_genes <- msigdbr(species = "human",category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C7 gene set (immune cells)
hallmark_genes <- msigdbr(species = "human",category = "C7", 
                          subcategory = "IMMUNESIGDB") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes,
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C2:CP gene set (KEGG and REACTOME)
hallmark_genes <- msigdbr(species = "human",category = "C2") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr",pvalueCutoff = 0.05,)

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# Cell adhesion PART I
focal_adhesion_set <- hallmark_genes[hallmark_genes$gs_name == "KEGG_FOCAL_ADHESION", ]
focal_adhesion_genes_list <- list("KEGG_FOCAL_ADHESION" = unique(focal_adhesion_set$entrez_gene))

hallmark_gene_gsea <- fgsea::fgsea(pathways = focal_adhesion_genes_list, 
                                   stats = genes, 
                                   nperm = 1000 )

print(hallmark_gene_gsea)

fgsea::plotEnrichment(focal_adhesion_genes_list[["KEGG_FOCAL_ADHESION"]], genes) +
  ggtitle("FGSEA Running Score Plot: KEGG_FOCAL_ADHESION")

hallmark_gene_gsea[hallmark_gene_gsea$pathway == "KEGG_FOCAL_ADHESION", 
                   c("NES", "pval", "padj")]

# Cell adhesion PART II
focal_adhesion_set <- hallmark_genes[hallmark_genes$gs_name == 
                                       "KEGG_CELL_ADHESION_MOLECULES_CAMS", ]
focal_adhesion_genes_list <- list("KEGG_CELL_ADHESION_MOLECULES_CAMS" = unique(focal_adhesion_set$entrez_gene))

hallmark_gene_gsea <- fgsea::fgsea(pathways = focal_adhesion_genes_list, 
                                   stats = genes, 
                                   nperm = 1000)

print(hallmark_gene_gsea)

fgsea::plotEnrichment(focal_adhesion_genes_list[["KEGG_CELL_ADHESION_MOLECULES_CAMS"]], genes) +
  ggtitle("FGSEA Running Score Plot: KEGG_CELL_ADHESION_MOLECULES_CAMS")

hallmark_gene_gsea[hallmark_gene_gsea$pathway == "KEGG_CELL_ADHESION_MOLECULES_CAMS", 
                   c("NES", "pval", "padj")]
```

#### ___ATRXwt vs. ATRXdel___

```{r}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ATRXwt_vs_ATRXdel_NK2_up$feature
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.1)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.20, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.20, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              #universe = geneUniverse,
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.1,
                             qvalueCutoff = 0.20,
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)

# TREE PLOT
# calculate pairwise similarities of the enriched terms using Jaccard’s similarity index
enrichres2 <- pairwise_termsim(kegg_enrichment) 

treeplot_NK2_wt <- treeplot(enrichres2,
  showCategory = 10,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1)

treeplot_NK2_wt
```


```{r}
pdf(paste0(params$out,"figures/DEG_hBm-NK2_ATRXdel_vs_ATRXwt_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK2_wt
dev.off()
```


___Gene Set Enrichment Analysis (GSEA)___

```{r}
genes <- ATRXwt_vs_ATRXdel$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

control_vs_ATRXdel$entrez_ids <- entrez_ids
genes <- control_vs_ATRXdel$log_fc
names(genes) <- control_vs_ATRXdel$entrez_ids
genes = sort(genes, decreasing = TRUE)
genes <- genes[!is.na(names(genes))]
length(genes)

# filter by expressed in NK cells
overlapping_genes <- intersect(entrez_ids, geneUniverse_entrezIDs)
length(overlapping_genes) #12119

# Filter the logFC vector based on the gene list
filtered_genes <- genes[names(genes) %in% overlapping_genes]
length(filtered_genes)

# H gene set
hallmark_genes <- msigdbr(species = "human",category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes)

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

gseaplot(hallmark_gene_gsea, 
         by = "all", 
         title = hallmark_gene_gsea$Description[1], 
         geneSetID = 1)

# C5 gene set
hallmark_genes <- msigdbr(species = "human",category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C7 gene set (immune cells)
hallmark_genes <- msigdbr(species = "human",category = "C7", 
                          subcategory = "IMMUNESIGDB") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes,
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C2:CP gene set (KEGG and REACTOME)
hallmark_genes <- msigdbr(species = "human",category = "C2") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)


# Cell adhesion PART I
focal_adhesion_set <- hallmark_genes[hallmark_genes$gs_name == "KEGG_FOCAL_ADHESION", ]
focal_adhesion_genes_list <- list("KEGG_FOCAL_ADHESION" = unique(focal_adhesion_set$entrez_gene))

hallmark_gene_gsea <- fgsea::fgsea(pathways = focal_adhesion_genes_list, 
                                   stats = genes, 
                                   nperm = 1000 )

print(hallmark_gene_gsea)

fgsea::plotEnrichment(focal_adhesion_genes_list[["KEGG_FOCAL_ADHESION"]], genes) +
  ggtitle("FGSEA Running Score Plot: KEGG_FOCAL_ADHESION")

hallmark_gene_gsea[hallmark_gene_gsea$pathway == "KEGG_FOCAL_ADHESION", 
                   c("NES", "pval", "padj")]

# Cell adhesion PART II
focal_adhesion_set <- hallmark_genes[hallmark_genes$gs_name == 
                                       "KEGG_CELL_ADHESION_MOLECULES_CAMS", ]
focal_adhesion_genes_list <- list("KEGG_CELL_ADHESION_MOLECULES_CAMS" = unique(focal_adhesion_set$entrez_gene))

hallmark_gene_gsea <- fgsea::fgsea(pathways = focal_adhesion_genes_list, 
                                   stats = genes, 
                                   nperm = 1000)

print(hallmark_gene_gsea)

fgsea::plotEnrichment(focal_adhesion_genes_list[["KEGG_CELL_ADHESION_MOLECULES_CAMS"]], genes) +
  ggtitle("FGSEA Running Score Plot: KEGG_CELL_ADHESION_MOLECULES_CAMS")

hallmark_gene_gsea[hallmark_gene_gsea$pathway == "KEGG_CELL_ADHESION_MOLECULES_CAMS", 
                   c("NES", "pval", "padj")]
```


#### ___ALL vs. ATRXdel___

```{r fig.height=5, fig.width=10}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ALL_vs_ATRXdel_NK2_up$feature
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.1)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.20, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.20, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.10, 
                              #universe = geneUniverse,
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 10) +
    ggtitle("up- in Bm_hNK2 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.1,
                             qvalueCutoff = 0.20,
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)

# TREE PLOT
# calculate pairwise similarities of the enriched terms using Jaccard’s similarity index
enrichres2 <- pairwise_termsim(kegg_enrichment) 

treeplot_NK2_all <- treeplot(enrichres2,
  showCategory = 5,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.4, 
  nCluster = 3, 
  nWords = 2)

treeplot_NK2_all


pdf(paste0(params$out,"figures/DEG_hBm-NK2_ATRXdel_vs_all_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK2_all
dev.off()
```

```{r}
genes <- ALL_vs_ATRXdel$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

control_vs_ATRXdel$entrez_ids <- entrez_ids
genes <- control_vs_ATRXdel$log_fc
names(genes) <- control_vs_ATRXdel$entrez_ids
genes = sort(genes, decreasing = TRUE)
genes <- genes[!is.na(names(genes))]
length(genes)

# filter by expressed in NK cells
overlapping_genes <- intersect(entrez_ids, geneUniverse_entrezIDs)
length(overlapping_genes) #12119

# Filter the logFC vector based on the gene list
filtered_genes <- genes[names(genes) %in% overlapping_genes]
length(filtered_genes)

# H gene set
hallmark_genes <- msigdbr(species = "human",category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes)

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

gseaplot(hallmark_gene_gsea, 
         by = "all", 
         title = hallmark_gene_gsea$Description[1], 
         geneSetID = 1)

# C5 gene set
hallmark_genes <- msigdbr(species = "human",category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C7 gene set (immune cells)
hallmark_genes <- msigdbr(species = "human",category = "C7", 
                          subcategory = "IMMUNESIGDB") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes,
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C2:CP gene set (KEGG and REACTOME)
hallmark_genes <- msigdbr(species = "human",category = "C2") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# Cell adhesion PART I
focal_adhesion_set <- hallmark_genes[hallmark_genes$gs_name == "KEGG_FOCAL_ADHESION", ]
focal_adhesion_genes_list <- list("KEGG_FOCAL_ADHESION" = unique(focal_adhesion_set$entrez_gene))

hallmark_gene_gsea <- fgsea::fgsea(pathways = focal_adhesion_genes_list, 
                                   stats = genes, 
                                   nperm = 1000 )

print(hallmark_gene_gsea)

fgsea::plotEnrichment(focal_adhesion_genes_list[["KEGG_FOCAL_ADHESION"]], genes) +
  ggtitle("FGSEA Running Score Plot: KEGG_FOCAL_ADHESION")

hallmark_gene_gsea[hallmark_gene_gsea$pathway == "KEGG_FOCAL_ADHESION", 
                   c("NES", "pval", "padj")]

# Cell adhesion PART II
focal_adhesion_set <- hallmark_genes[hallmark_genes$gs_name == 
                                       "KEGG_CELL_ADHESION_MOLECULES_CAMS", ]
focal_adhesion_genes_list <- list("KEGG_CELL_ADHESION_MOLECULES_CAMS" = unique(focal_adhesion_set$entrez_gene))

hallmark_gene_gsea <- fgsea::fgsea(pathways = focal_adhesion_genes_list, 
                                   stats = genes, 
                                   nperm = 1000)

print(hallmark_gene_gsea)

fgsea::plotEnrichment(focal_adhesion_genes_list[["KEGG_CELL_ADHESION_MOLECULES_CAMS"]], genes) +
  ggtitle("FGSEA Running Score Plot: KEGG_CELL_ADHESION_MOLECULES_CAMS")

hallmark_gene_gsea[hallmark_gene_gsea$pathway == "KEGG_CELL_ADHESION_MOLECULES_CAMS", 
                   c("NES", "pval", "padj")]
```


### ___Cluster 3___

```{r}
# Change idents to cluster annotations
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

# subset to cluster 3
cluster_of_interest <- "hNK_Bm3"

# subset scRNA-seq data to NK and NB cells 
hNK_Bm3 <- subset(ALL_NK, 
                  idents = cluster_of_interest, 
                  invert = FALSE)
hNK_Bm3 # 3732 cells
```

#### ___1. Control vs. ATRXdel___

```{r}
# DE analysis for the hNK_Bm3 NK cluster
DEG_hNK_Bm3_DEG <- DElegate::findDE(hNK_Bm3,
                           group_column = "group",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "rados_et_al_2025",
                           compare = "control")

control_vs_ATRXdel <- subset(DEG_hNK_Bm3_DEG, 
                        DEG_hNK_Bm3_DEG$group1 == "ATRXdel")

control_vs_MYCNamp <- subset(DEG_hNK_Bm3_DEG, 
                        DEG_hNK_Bm3_DEG$group1 == "MYCNamp")

control_vs_ATRXwtMYCNwt <- subset(DEG_hNK_Bm3_DEG, 
                        DEG_hNK_Bm3_DEG$group1 == "ATRXwtMYCNwt")

control_vs_ATRXdel_signif <- subset(control_vs_ATRXdel, 
                                control_vs_ATRXdel$padj < 0.1)
```
```{r}
write.table(control_vs_ATRXdel, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_control_vs_ATRXdel_2025-04-24.csv"))
```



#### ___2. ATRX wild-type [MYCNamp + ATRXwtMYCNwt] vs. ATRXdel___

```{r}
# DE analysis for the hNK_Bm3 NK cluster
ATRXwt_vs_ATRXdel <- DElegate::findDE(hNK_Bm3,
                           group_column = "comparison_group",
                           order_results = TRUE,
                           compare_is_ref = FALSE,
                           replicate_column = "rados_et_al_2025",
                           compare = "ATRXdel")

ATRXwt_vs_ATRXdel_signif <- subset(ATRXwt_vs_ATRXdel, ATRXwt_vs_ATRXdel$padj < 0.10)

ATRXwt_vs_ATRXdel_signif
```
```{r}
write.table(ATRXwt_vs_ATRXdel, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_ATRXdel-vs-ATRXwt_2025-04-24.csv"))
```


#### ___3. ALL vs. ATRXdel___

```{r}
ALL_vs_ATRXdel <- DElegate::findDE(hNK_Bm3,
                           group_column = "group",
                           order_results = TRUE,
                           compare_is_ref = FALSE,
                           replicate_column = "rados_et_al_2025",
                           compare = "ATRXdel")


ALL_vs_ATRXdel_signif <- subset(ALL_vs_ATRXdel, ALL_vs_ATRXdel$padj < 0.10)

ALL_vs_ATRXdel_signif
```
```{r}
write.table(ALL_vs_ATRXdel, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNK-Bm3_ATRXdel-vs-ALL_2025-04-24.csv"))
```

### ___Vulcano plot___

#### ___1. control vs. ATRXdel___

```{r fig.height=7, fig.width=7}
control_vs_ATRXdel
rownames(control_vs_ATRXdel) <- control_vs_ATRXdel$feature

library(EnhancedVolcano)

pdf(paste0(params$out,"figures/DEGs_NK_cluster3_control_vs_ATRXdel_vulcano_plot_wGenes.pdf"),
    height = 5,width = 7)
EnhancedVolcano(control_vs_ATRXdel, 
                lab = rownames(control_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','TGFB1', 'F2R', 'PTPRJ'),
                subtitle = "control vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                ylim = c(0,3),
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()

pdf(paste0(params$out,"figures/DEGs_NK_cluster3_control_vs_ATRXdel_vulcano_plot.pdf"),
    height = 5,width = 7)
EnhancedVolcano(control_vs_ATRXdel, 
                lab = rownames(control_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "control vs. ATRXdel",
                pCutoff = 0.05,
                ylim = c(0,3),
                FCcutoff = 1,
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()

```

#### ___2. ATRXwt vs. ATRXdel___

```{r fig.height=7, fig.width=7}
ATRXwt_vs_ATRXdel
rownames(ATRXwt_vs_ATRXdel) <- ATRXwt_vs_ATRXdel$feature

pdf(paste0(params$out,"figures/DEGs_NK_cluster3_ATRXwt_vs_ATRXdel_vulcano_plot_wGenes.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ATRXwt_vs_ATRXdel, 
                lab = rownames(ATRXwt_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','TGFB1', 'F2R', 'PTPRJ'),
                subtitle = "ATRXwt vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                ylim = c(0,4),
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()

pdf(paste0(params$out,"figures/DEGs_NK_cluster3_ATRXwt_vs_ATRXdel_vulcano_plot.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ATRXwt_vs_ATRXdel, 
                lab = rownames(ATRXwt_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "ATRXwt vs. ATRXdel",
                pCutoff = 0.05,
                ylim = c(0,4),
                FCcutoff = 1,
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()

```

#### ___3. ALL vs. ATRXdel___

```{r fig.height=7, fig.width=7}
rownames(ALL_vs_ATRXdel) <- ALL_vs_ATRXdel$feature

pdf(paste0(params$out,"figures/DEGs_NK_cluster3_ALL_vs_ATRXdel_vulcano_plot_wGenes.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ALL_vs_ATRXdel, 
                lab = rownames(ALL_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                selectLab = c('CD6','TGFB1', 'F2R', 'PTPRJ'),
                subtitle = "(ATRXwt + control) vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                pointSize = 2,
                ylim = c(0,6),
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()

pdf(paste0(params$out,"figures/DEGs_NK_cluster3_ALL_vs_ATRXdel_vulcano_plot.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ALL_vs_ATRXdel, 
                lab = rownames(ALL_vs_ATRXdel),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in cluster 3",
                subtitle = "(ATRXwt + control) vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                ylim = c(0,6),
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()

```

### ___Pathways enrichment analysis___

```{r}
control_vs_ATRXdel_signif <- control_vs_ATRXdel_signif %>%
  arrange(log_fc)

control_vs_ATRXdel_NK3_up <- control_vs_ATRXdel_signif %>% 
  subset(log_fc > 1) %>% 
  subset(padj < 0.05) 
control_vs_ATRXdel_NK3_down <- control_vs_ATRXdel_signif %>% 
  subset(log_fc < -1) %>% 
  subset(padj < 0.05) 

ATRXwt_vs_ATRXdel_signif <- ATRXwt_vs_ATRXdel %>%
  arrange(log_fc)

ATRXwt_vs_ATRXdel_NK3_up <- ATRXwt_vs_ATRXdel %>%  
  subset(log_fc > 1) %>% 
  subset(padj < 0.05) 
ATRXwt_vs_ATRXdel_NK3_down <- ATRXwt_vs_ATRXdel %>% 
  subset(log_fc < -1) %>% 
  subset(padj < 0.05) 

ALL_vs_ATRXdel_signif <- ALL_vs_ATRXdel_signif %>%
  arrange(log_fc)

ALL_vs_ATRXdel_NK3_up <- ALL_vs_ATRXdel_signif %>% 
  subset(log_fc > 1) %>% 
  subset(padj < 0.05) # up-regulated in ATRXdeleted rados_et_al_2025s 

ALL_vs_ATRXdel_NK3_down <- ALL_vs_ATRXdel_signif %>%  
  subset(log_fc < -1)  %>% 
  subset(padj < 0.05) #down-regulated in ATRXdeleted rados_et_al_2025s
```

#### ___control vs. ATRXdel___

```{r}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- control_vs_ATRXdel_NK3_up$feature
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05,
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              #universe = geneUniverse,
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.05,
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched HALLMARK terms")
} else if (any(hallmark_gene_enrich@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(hallmark_gene_enrich, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: HALLMARK enrichment"))
} else {
  print("0 significantly enriched HALLMARK terms")
}

```

```{r}

pdf(paste0(params$out,"figures/DEG_hBm-NK3_ATRXdel_vs_CTRL_hallmark.pdf"),
    height = 4, width = 9)
dotplot(hallmark_gene_enrich, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: HALLMARK genes")
dev.off()

enrichres2 <- pairwise_termsim(kegg_enrichment) 
treeplot_NK3_ctrl <- treeplot(enrichres2,
  showCategory = 15,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK3_ctrl

pdf(paste0(params$out,"figures/DEG_hBm-NK3_ATRXdel_vs_CTRL_KEGG_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK3_ctrl
dev.off()

enrichres2 <- pairwise_termsim(go_BP) 
treeplot_NK3_ctrl <- treeplot(enrichres2,
  showCategory = 15,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK3_ctrl

pdf(paste0(params$out,"figures/DEG_hBm-NK3_ATRXdel_vs_CTRL_GO-BP_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK3_ctrl
dev.off()

```


```{r fgsea}
genes <- control_vs_ATRXdel$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

control_vs_ATRXdel$entrez_ids <- entrez_ids
genes <- control_vs_ATRXdel$log_fc
names(genes) <- control_vs_ATRXdel$entrez_ids
genes = sort(genes, decreasing = TRUE)
genes <- genes[!is.na(names(genes))]
length(genes)

# filter by expressed in NK cells
overlapping_genes <- intersect(entrez_ids, geneUniverse_entrezIDs)
length(overlapping_genes) #12119

# Filter the logFC vector based on the gene list
filtered_genes <- genes[names(genes) %in% overlapping_genes]
length(filtered_genes)

# H gene set
hallmark_genes <- msigdbr(species = "human",category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes)

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

gseaplot(hallmark_gene_gsea, 
         by = "all", 
         title = hallmark_gene_gsea$Description[1], 
         geneSetID = 1)

# C5 gene set
hallmark_genes <- msigdbr(species = "human",category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C7 gene set (immune cells)
hallmark_genes <- msigdbr(species = "human",category = "C7", 
                          subcategory = "IMMUNESIGDB") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes,
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C2:CP gene set (KEGG and REACTOME)
hallmark_genes <- msigdbr(species = "human",category = "C2",
                          subcategory = "CP") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)
```


#### ___ATRXwt vs. ATRXdel___

```{r}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ATRXwt_vs_ATRXdel_NK3_up$feature
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              #universe = geneUniverse,
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.05,
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)

enrichres2 <- pairwise_termsim(kegg_enrichment) 
treeplot_NK3_wt <- treeplot(enrichres2,
  showCategory = 15,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK3_wt

```

```{r fgsea}
genes <- ATRXwt_vs_ATRXdel$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

control_vs_ATRXdel$entrez_ids <- entrez_ids
genes <- control_vs_ATRXdel$log_fc
names(genes) <- control_vs_ATRXdel$entrez_ids
genes = sort(genes, decreasing = TRUE)
genes <- genes[!is.na(names(genes))]
length(genes)

# filter by expressed in NK cells
overlapping_genes <- intersect(entrez_ids, geneUniverse_entrezIDs)
length(overlapping_genes) #12119

# Filter the logFC vector based on the gene list
filtered_genes <- genes[names(genes) %in% overlapping_genes]
length(filtered_genes)

# H gene set
hallmark_genes <- msigdbr(species = "human",category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes)

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

gseaplot(hallmark_gene_gsea, 
         by = "all", 
         title = hallmark_gene_gsea$Description[1], 
         geneSetID = 1)

# C5 gene set
hallmark_genes <- msigdbr(species = "human",category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C7 gene set (immune cells)
hallmark_genes <- msigdbr(species = "human",category = "C7", 
                          subcategory = "IMMUNESIGDB") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes,
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C2:CP gene set (KEGG and REACTOME)
hallmark_genes <- msigdbr(species = "human",category = "C2",
                          subcategory = "CP") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)
```

#### ___ALL vs. ATRXdel___

```{r}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ALL_vs_ATRXdel_NK3_up$feature
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,  
                              #universe = geneUniverse,
                              minGSSize = 2,
                              maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK3 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.05, 
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)

enrichres2 <- pairwise_termsim(kegg_enrichment) 
treeplot_NK3_all <- treeplot(enrichres2,
  showCategory = 15,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK3_all
```

```{r fgsea}
genes <- ALL_vs_ATRXdel$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

control_vs_ATRXdel$entrez_ids <- entrez_ids
genes <- control_vs_ATRXdel$log_fc
names(genes) <- control_vs_ATRXdel$entrez_ids
genes = sort(genes, decreasing = TRUE)
genes <- genes[!is.na(names(genes))]
length(genes)

# filter by expressed in NK cells
overlapping_genes <- intersect(entrez_ids, geneUniverse_entrezIDs)
length(overlapping_genes) #12119

# Filter the logFC vector based on the gene list
filtered_genes <- genes[names(genes) %in% overlapping_genes]
length(filtered_genes)

# H gene set
hallmark_genes <- msigdbr(species = "human",category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes)

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

gseaplot(hallmark_gene_gsea, 
         by = "all", 
         title = hallmark_gene_gsea$Description[1], 
         geneSetID = 1)

# C5 gene set
hallmark_genes <- msigdbr(species = "human",category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C7 gene set (immune cells)
hallmark_genes <- msigdbr(species = "human",category = "C7", 
                          subcategory = "IMMUNESIGDB") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes,
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C2:CP gene set (KEGG and REACTOME)
hallmark_genes <- msigdbr(species = "human",category = "C2",
                          subcategory = "CP") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)
```
#### ___Cluster 2 + 3___

```{r}
# all genes
control_vs_ATRXdel <- c(control_vs_ATRXdel_NK2_up$feature, control_vs_ATRXdel_NK3_up$feature)
control_vs_ATRXdel <- unique(control_vs_ATRXdel)
length(control_vs_ATRXdel)
ATRXwt_vs_ATRXdel <- c(ATRXwt_vs_ATRXdel_NK2_up$feature, ATRXwt_vs_ATRXdel_NK3_up$feature)
ATRXwt_vs_ATRXdel <- unique(ATRXwt_vs_ATRXdel)
length(ATRXwt_vs_ATRXdel)
ALL_vs_ATRXdel <- c(ALL_vs_ATRXdel_NK2_up$feature, ALL_vs_ATRXdel_NK3_up$feature)
ALL_vs_ATRXdel <- unique(ALL_vs_ATRXdel)
length(ALL_vs_ATRXdel)

# common genes
```

###### ___pathway enrichment____

___control vs. ATRXdel___

```{r}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- control_vs_ATRXdel
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05,
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.05,
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched HALLMARK terms")
} else if (any(hallmark_gene_enrich@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(hallmark_gene_enrich, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: HALLMARK enrichment"))
} else {
  print("0 significantly enriched HALLMARK terms")
}


pdf(paste0(params$out,"figures/DEG_hBm-NK2+hBm-NK3_ATRXdel_vs_CTRL_hallmark.pdf"),
    height = 4, width = 9)
dotplot(hallmark_gene_enrich, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: HALLMARK genes")
dev.off()

# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              #universe = geneUniverse,
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

```

```{r}
enrichres2 <- pairwise_termsim(go_BP) 
treeplot_NK2_NK3_ctrl <- treeplot(enrichres2,
  showCategory = 10,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK2_NK3_ctrl

pdf(paste0(params$out,"figures/DEG_hBm-NK2+hBm-NK3_ATRXdel_vs_CTRL_GO-BP_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK2_NK3_ctrl
dev.off()

enrichres2 <- pairwise_termsim(kegg_enrichment) 
treeplot_NK2_NK3_ctrl <- treeplot(enrichres2,
  showCategory = 10,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK2_NK3_ctrl

pdf(paste0(params$out,"figures/DEG_hBm-NK2+hBm-NK3_ATRXdel_vs_CTRL_KEGG_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK3_ctrl
dev.off()


```

___ATRXwt vs. ATRXdel___

```{r}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ATRXwt_vs_ATRXdel
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,
                              #universe = geneUniverse,
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.05,
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)
```

```{r}
enrichres2 <- pairwise_termsim(go_BP) 
treeplot_NK2_NK3_wt <- treeplot(enrichres2,
  showCategory = 10,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK2_NK3_wt

pdf(paste0(params$out,"figures/DEG_hBm-NK2+hBm-NK3_ATRXdel_vs_ATRXwt_GO-BP_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK2_NK3_wt 
dev.off()

enrichres2 <- pairwise_termsim(kegg_enrichment) 
treeplot_NK2_NK3_wt  <- treeplot(enrichres2,
  showCategory = 10,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK2_NK3_wt

pdf(paste0(params$out,"figures/DEG_hBm-NK2+hBm-NK3_ATRXdel_vs_ATRXwt_KEGG_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK2_NK3_wt
dev.off()
```

___ALL vs. ATRXdel___

```{r}
# the background genes here should be all expressed (and detected) in all NK cells
genes <- ALL_vs_ATRXdel
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.05, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.05,  
                              #universe = geneUniverse,
                              minGSSize = 2,
                              maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("up- in Bm_hNK2+3 in ATRXdel: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.05, 
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)
```

```{r}
enrichres2 <- pairwise_termsim(go_BP) 
treeplot_NK2_NK3_all <- treeplot(enrichres2,
  showCategory = 10,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK2_NK3_all

pdf(paste0(params$out,"figures/DEG_hBm-NK2+hBm-NK3_ATRXdel_vs_all_GO-BP_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK2_NK3_all 
dev.off()

enrichres2 <- pairwise_termsim(kegg_enrichment) 
treeplot_NK2_NK3_all  <- treeplot(enrichres2,
  showCategory = 10,
  color = "p.adjust",
  offset = 10, 
  offset_tiplab = 0.5, 
  nCluster = 5, 
  nWords = 1,
  split = "subcategory")

treeplot_NK2_NK3_all

pdf(paste0(params$out,"figures/DEG_hBm-NK2+hBm-NK3_ATRXdel_vs_all_KEGG_treeplot.pdf"),
    height = 4, width = 9)
treeplot_NK2_NK3_all
dev.off()
```

## ___Justify CD6 selection___

### ___Cluster 2 DEGs___

```{r}
cluster2_gene_lists <- list(vs_control = control_vs_ATRXdel_NK2_up$feature,
                       vs_ATRXwt = ATRXwt_vs_ATRXdel_NK2_up$feature,
                       vs_all = ALL_vs_ATRXdel_NK2_up$feature)

ggVennDiagram(cluster2_gene_lists, set_size = 6, label_alpha = 0,label_size = 6,
  category.names = c("vs_control","vs_ATRXwt","vs_all"),
  edge_size = 1, label_percent_digit = 1,label_color = "white") +
  ggplot2::scale_fill_gradient(low="#B59410",high = "darkblue") +
  labs(title = "Overlap") + 
  theme(title = element_text(size = 16),legend.text = element_text(size = 14)) +
  scale_x_continuous(expand = expansion(mult = .2))

# Find common genes in all three lists
common_genes <- Reduce(intersect, cluster2_gene_lists)

# Print result
print(common_genes)
```

### ___Cluster 3 DEGs___

```{r}
cluster3_gene_lists <- list(vs_control = control_vs_ATRXdel_NK3_up$feature,
                       vs_ATRXwt = ATRXwt_vs_ATRXdel_NK3_up$feature,
                       vs_all = ALL_vs_ATRXdel_NK3_up$feature)

ggVennDiagram(cluster3_gene_lists, set_size = 6, label_alpha = 0,label_size = 6,
  category.names = c("vs_control","vs_ATRXwt","vs_all"),
  edge_size = 1, label_percent_digit = 1,label_color = "white") +
  ggplot2::scale_fill_gradient(low="#B59410",high = "darkblue") +
  labs(title = "Overlap") + 
  theme(title = element_text(size = 16),legend.text = element_text(size = 14)) +
  scale_x_continuous(expand = expansion(mult = .2))

# Find common genes in all three lists
common_genes <- Reduce(intersect, cluster3_gene_lists)

# Print result
print(common_genes)
```

### ___Cluster 2 and cluster 3 DEGs___

#### ___Up-regulated genes___

```{r}
cluster2_and_3_gene_lists <- list(NK2 = ALL_vs_ATRXdel_NK2_up$feature,
                            NK3 = ALL_vs_ATRXdel_NK3_up$feature)

ggVennDiagram(cluster2_and_3_gene_lists, set_size = 6, label_alpha = 0,label_size = 6,
  category.names = c("NK2","NK3"),
  edge_size = 1, label_percent_digit = 1,label_color = "white") +
  ggplot2::scale_fill_gradient(low="#B59410",high = "darkblue") +
  labs(title = "Overlap") + 
  theme(title = element_text(size = 16),legend.text = element_text(size = 14)) +
  scale_x_continuous(expand = expansion(mult = .2))

common_genes <- Reduce(intersect,cluster2_and_3_gene_lists)
common_genes
```

Up-regulated genes associated with cell adhesion:
ABLIM1, BTBD11, AC058791.1, CD6, ARHGEF5

```{r}
# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(common_genes)

# enrichment BP
go_BP <- enrichGO(gene = entrez_ids, 
               ont = "BP",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              pvalueCutoff = 0.1,
               minGSSize = 2,
               maxGSSize = 5000)

go_BP

if (is.null(go_BP) == TRUE) {
  print("no significantly enriched BP terms")
} else if (any(go_BP@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_BP, showCategory = 20) +
    ggtitle("common genes: BP enrichment"))
} else {
  print("0 significantly enriched BP terms")
}

# enrichment MF
go_MF <- enrichGO(gene = entrez_ids, 
               ont = "MF",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.20, 
               minGSSize = 2,
               maxGSSize = 5000)

go_MF

if (is.null(go_MF) == TRUE) {
  print("no significantly enriched MF terms")
} else if (any(go_MF@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_MF, showCategory = 20) +
    ggtitle("common genes: MF enrichment"))
} else {
  print("0 significantly enriched MF terms")
}

# enrichment CC
go_CC <- enrichGO(gene = entrez_ids, 
               ont = "CC",
               OrgDb ="org.Hs.eg.db",
               #universe = geneUniverse,
               readable = TRUE,
              pAdjustMethod = "fdr",
              qvalueCutoff = 0.20, 
               minGSSize = 2,
               maxGSSize = 5000)

go_CC

if (is.null(go_CC) == TRUE) {
  print("no significantly enriched CC terms")
} else if (any(go_CC@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(go_CC, showCategory = 20) +
    ggtitle("common genes: CC enrichment"))
} else {
  print("0 significantly enriched CC terms")
}


# enrichment KEGG
kegg_enrichment <- enrichKEGG(gene = entrez_ids,
                              organism = "hsa", 
                              keyType = "kegg",
                              pvalueCutoff = 0.1,
                              qvalueCutoff = 0.20, 
                              #universe = geneUniverse,
               minGSSize = 2,
               maxGSSize = 5000)

kegg_enrichment

if (is.null(kegg_enrichment) == TRUE) {
  print("no significantly enriched KEGG terms")
} else if (any(kegg_enrichment@result$qvalue < 0.1)) {
  # If there is at least one term with qvalue < 0.1, show the dotplot
  print(dotplot(kegg_enrichment, showCategory = 20) +
    ggtitle("common genes: KEGG enrichment"))
} else {
  print("0 significantly enriched KEGG terms")
}

# HALLMARK GENE SET
hallmark_gene_enrich <- enricher(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             qvalueCutoff = 0.20,
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_enrich)
```
 
```{r eval=FALSE, include=FALSE}
hallmark_gene_gsea <- GSEA(entrez_ids, 
                             TERM2GENE = hallmark_genes,
                             pvalueCutoff = 0.1,
                             qvalueCutoff = 0.2,
                             pAdjustMethod = "fdr",
                             #universe = geneUniverse,
                             minGSSize = 2,
                             maxGSSize = 5000)

head(hallmark_gene_gsea)
```

#### ___Down-regulated genes___

```{r}
cluster2_and_3_gene_lists <- list(NK2 = ALL_vs_ATRXdel_NK2_down$feature,
                            NK3 = ALL_vs_ATRXdel_NK3_down$feature)

ggVennDiagram(cluster2_and_3_gene_lists, set_size = 6, label_alpha = 0,label_size = 6,
  category.names = c("NK2","NK3"),
  edge_size = 1, label_percent_digit = 1,label_color = "white") +
  ggplot2::scale_fill_gradient(low="#B59410",high = "darkblue") +
  labs(title = "Overlap") + 
  theme(title = element_text(size = 16),legend.text = element_text(size = 14)) +
  scale_x_continuous(expand = expansion(mult = .2))

common_genes <- Reduce(intersect,cluster2_and_3_gene_lists)
common_genes
```

# ___NB cells___

```{r}
# Access the counts matrix
counts_matrix <- GetAssayData(ALL_NB, slot = "counts")

# set a meaningful cutoff for expressed genes 
hist(Matrix::rowSums(counts_matrix),breaks = 1000000,xlim = (c(0,100)))
abline(v=2)

hist(log(Matrix::rowSums(counts_matrix)),breaks = 500)

# Filter genes with at least 11 reads over all samples (rados_et_al_2025)
# these will be used as background genes for pathways enrichment analysis
geneUniverse <- rownames(counts_matrix)[Matrix::rowSums(counts_matrix) > 20]
length(geneUniverse) # 16480 expressed genes

# justify chosen cut-off of expressed genes
hist(log(Matrix::rowSums(counts_matrix)),breaks = 500) 
abline(v=0.7)
abline(v=3)

# remap to ENTREZ IDs
geneUniverse_entrezIDs <- map_gene_to_entrez(geneUniverse)
```

## ___ATRX wild-type [MYCNamp + ATRXwtMYCNwt] vs. ATRXdel___

```{r}
ALL_NB_DEG <- DElegate::findDE(ALL_NB,
                           group_column = "ATRXdel_vs_ATRXwt",
                           order_results = TRUE,
                           compare_is_ref = TRUE,
                           replicate_column = "rados_et_al_2025",
                           compare = "ATRXwt")

ALL_NB_DEG_signif <- subset(ALL_NB_DEG, 
                            ALL_NB_DEG$padj < 0.10)

ALL_NB_DEG_signif

write.table(ALL_NB_DEG, 
            sep = "\t",
            row.names = FALSE,
            quote = FALSE,
            file = paste0(params$out, "tables/DEG_NB_ATRXdel-vs-ATRXwt_2025-04-30.csv"))
```

```{r}
rownames(ALL_NB_DEG) <- ALL_NB_DEG$feature

library(EnhancedVolcano)

pdf(paste0(params$out,"figures/DEGs_NB_ATRXwt_vs_ATRXdel_vulcano_plot.pdf"),
    height = 5,width = 7)
EnhancedVolcano(ALL_NB_DEG, 
                lab = rownames(ALL_NB_DEG),
                x = 'log_fc',
                y = 'padj',
                title = "DEGs in NB tumor cells",
                #selectLab = c('CD6','TGFB1', 'F2R', 'PTPRJ'),
                subtitle = "ATRXwt vs. ATRXdel",
                pCutoff = 0.05,
                FCcutoff = 1,
                ylim = c(0,4),
                pointSize = 2,
                labSize = 6,
                col=c('darkgray', '#28428D', '#28428D', '#61E786'),
                boxedLabels = TRUE,
                colAlpha = 4/5,
                legendPosition = 'right',
                legendLabSize = 14,
                legendIconSize = 4.0,
                drawConnectors = TRUE,
                widthConnectors = 1.0,
                colConnectors = 'black'
                )
dev.off()
```

___Gene Set Enrichment Analysis (GSEA)___

```{r}
genes <- ALL_NB_DEG$feature

# remap to ENTREZ IDs
entrez_ids <- map_gene_to_entrez(genes)

control_vs_ATRXdel$entrez_ids <- entrez_ids
genes <- control_vs_ATRXdel$log_fc
names(genes) <- control_vs_ATRXdel$entrez_ids
genes = sort(genes, decreasing = TRUE)
genes <- genes[!is.na(names(genes))]
length(genes)

# filter by expressed in NK cells
overlapping_genes <- intersect(entrez_ids, geneUniverse_entrezIDs)
length(overlapping_genes) #12119

# Filter the logFC vector based on the gene list
filtered_genes <- genes[names(genes) %in% overlapping_genes]
length(filtered_genes)

# Hallmark gene set
hallmark_genes <- msigdbr(species = "human",category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes)

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

gseaplot(hallmark_gene_gsea, 
         by = "all", 
         title = hallmark_gene_gsea$Description[1], 
         geneSetID = 1)

# C5 Onotology gene set
hallmark_genes <- msigdbr(species = "human",category = "C5") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C2:CP gene set (KEGG and REACTOME)
hallmark_genes <- msigdbr(species = "human",category = "C2",
                          subcategory = "CP") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)

# C6 oncogenic signature gene sets
hallmark_genes <- msigdbr(species = "human",category = "C6") %>% 
  dplyr::select(gs_name, entrez_gene)

hallmark_gene_gsea <- GSEA(filtered_genes, 
                             TERM2GENE = hallmark_genes, 
                           pAdjustMethod = "fdr")

head(hallmark_gene_gsea)
dotplot(hallmark_gene_gsea, showCategory = 20)
```


# ___The End___

```{r}
sessionInfo()
```

