---
title: "NK project: Paper figures"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_BM1: "/home/rstudio/mnt_data/CancReg/BM/crinier2021/Rds/scBM_healthy_seurat_v4.Rds"
  input_fetahu2023: "/home/rstudio/mnt_out/CancReg/scRNA-seq_stage_M/results/published/data/rna_decontaminated.rds"
  input_scRNA: "/home/rstudio/mnt_out/NK_project/Rds/fetahu2023_withNewMetadata.Rds"
  input_NK: "/home/rstudio/mnt_out/NK_project/Rds/ALL_NK_cluster_withNewMetadata.Rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
  metadata: "/home/rstudio/mnt_out/CancReg/scRNA-seq_stage_M/results/metadata/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```

```{r load project specific R packages and color code, echo = FALSE}
source("/home/rstudio/NK-NB_study/code/00_NK_paper_setup.R")
source("/home/rstudio/NK-NB_study/code/00_NK_paper_utils.R")
```

```{r}
library("zen4R")
```


# ___Add new clinical information to the single cell object___

```{r}
fetahu2023 <- readRDS(file = paste0(params$input_scRNA))
metadata <- readRDS("/home/rstudio/mnt_out/CancReg/scRNA-seq_stage_M/results/published/data/metadata.rds")

# set cell type annotation to the original annotation
fetahu2023@active.ident <- setNames(pull(metadata[match(metadata$cell,colnames(fetahu2023)),"cellont_cluster"], "cellont_cluster"),pull(metadata[match(metadata$cell,colnames(fetahu2023)), "cell"], "cell"))

fetahu2023@meta.data$orig.ident <- fetahu2023@active.ident # cell type annotation

# set assay to RNA 
DefaultAssay(fetahu2023) <- "RNA"

fetahu2023@meta.data
```

```{r}
# add the new clinical annotations and additional information 
full_metadata <- read.csv(file = paste0(params$metadata, 
  "fetahu2023_scRNA_seq_metadata.csv"), sep ="\t")

metadata <- fetahu2023@meta.data
metadata$cell_ID <- row.names(metadata)

# Perform the merge using left_join from dplyr, keeping the row order of df1
merged_metadata <- metadata %>%
  left_join(full_metadata, by = "sample") %>% 
  distinct(cell_ID, .keep_all = TRUE)

nrow(metadata)
nrow(merged_metadata)

row.names(merged_metadata) <- merged_metadata$cell_ID 

# Update the Seurat object's metadata slot
fetahu2023@meta.data <- merged_metadata
fetahu2023@meta.data
```

```{r}
colnames(fetahu2023@meta.data)
 
columns_to_remove <- c("sample_ID","File.name", "stage.x")

# Drop the unwanted columns while preserving row names (cell IDs)
old_metadata <- fetahu2023@meta.data 
new_metadata <- old_metadata %>%
  dplyr::select(-all_of(columns_to_remove))
fetahu2023@meta.data <- new_metadata

fetahu2023@meta.data 
```

```{r}
# does log normalization make sense?
# was it already done for this object?
# does visualization change if we log normalize?

fetahu2023 <- fetahu2023 %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

```{r}
saveRDS(fetahu2023, file = paste0(params$out,"/Rds/fetahu2023_withNewMetadata.Rds"))
```

```{r}
predicted_labels <- readRDS(file = paste0(params$out,                                 "/Rds/2024_analysis/fetahu2023_predicted_NKclusters_from_Crinier2021_de.n=350.Rds"))

predicted_labels$cell_ID <- rownames(predicted_labels)

# subset scRNA-seq data to NK cells only
NK_cells <- "NK (4)"
ALL_NK <- subset(fetahu2023, 
                 idents = NK_cells, 
                 invert = FALSE)

old_NK_metadata <- ALL_NK@meta.data

old_NK_metadata$cell_ID <- rownames(old_NK_metadata)

new_NK_metadata <- old_NK_metadata %>%
  dplyr::left_join(as.data.frame(predicted_labels), by = "cell_ID")

new_NK_metadata <- new_NK_metadata %>%
  rename("labels" = "predicted_NK_type_de.n350")
  
row.names(new_NK_metadata) <- new_NK_metadata$cell_ID 
  
ALL_NK@meta.data <- merged_metadata
ALL_NK@meta.data

# log scale counts
ALL_NK <- ALL_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

```{r}
Idents(ALL_NK) <- ALL_NK@meta.data$group

# identify cell IDs associated with each cluster 
M <- WhichCells(ALL_NK, ident = "MYCNamp")        # MYCN amplified
S <- WhichCells(ALL_NK, ident = "ATRXwtMYCNwt") # sporadic
C <- WhichCells(ALL_NK, ident = "control")   # control / control (GNB or GNM)
A <- WhichCells(ALL_NK, ident = "ATRXdel")       # ATRX deletion or mutations

ALL_NK@active.ident <- ALL_NK$orig.ident

# Add ATRX wt vs. mutated annotation
Idents(object = ALL_NK, cells = M) <- "ATRXwt"
Idents(object = ALL_NK, cells = S) <- "ATRXwt"
Idents(object = ALL_NK, cells = A) <- "ATRXdel"
Idents(object = ALL_NK, cells = C) <- "CTRL"

# Add active identities to metadata column comparison_group
ALL_NK$comparison_group <- ALL_NK@active.ident

saveRDS(ALL_NK, file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```

# ___Add sub-group categories___

```{r eval=FALSE, include=FALSE}
group <- "group" 
NK_subtype <- "predicted_NK_type_de.n350"  

# Create a new metadata column that concatenates values from both columns
ALL_NK$group_per_NK_subtype <- paste(ALL_NK@meta.data[[group]], 
                                      ALL_NK@meta.data[[NK_subtype]], 
                                      sep = "_")

# Check if the new column is added
ALL_NK@meta.data$group_per_NK_subtype <- as.factor(ALL_NK@meta.data$group_per_NK_subtype)
table(ALL_NK@meta.data$group_per_NK_subtype)
```

```{r eval=FALSE, include=FALSE}
group <- "group" 
NK_subtype <- "predicted_NK_type_de.n350"  

# Create a new metadata column that concatenates values from both columns
ALL_NK$NK_subtype_per_group <- paste(ALL_NK@meta.data[[NK_subtype]], 
                                     ALL_NK@meta.data[[group]], 
                                      sep = "_")

# Check if the new column is added
ALL_NK@meta.data$NK_subtype_per_group <- as.factor(ALL_NK@meta.data$NK_subtype_per_group)
table(ALL_NK@meta.data$NK_subtype_per_group)
```

```{r eval=FALSE, include=FALSE}
saveRDS(ALL_NK, file = paste0(params$out,"/Rds/ALL_NK_cluster_withNewMetadata.Rds"))
```


# ___Input_data___

```{r}
fetahu2023 <- readRDS(file = params$input_scRNA)
ALL_NK <-  readRDS(file = params$input_NK)
# ALL_NK <-  readRDS(file = paste0(params$out,"Rds/2024_analysis/ALL_NK_cluster.Rds"))
```

# ___Figure 1: Differential abundance and label transfer analysis___

```{r}

```

# ___Figure 2: NK annotation, maturation and trajectory analysis___

## ___Mei et al., 2024___

```{r}
NK_Mei_et_al_markers
Mei_et_al_CD56dim_genes <- list(CD56dim_1 = NK_Mei_et_al_markers$CD56dim_1, 
                                 CD56dim_2 = NK_Mei_et_al_markers$CD56dim_2)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = Mei_et_al_CD56dim_genes,
                         name = "Mei_et_al_CD56dim_genes",
                         seed = "2805",
                         slot = "data",
                         nbin = 10)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_Mei_et_al_markers$CD56dim_1,
                         name = "CD56dim1_module_",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_Mei_et_al_markers$CD56dim_2,
                         name = "CD56dim2_module_",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)
```

```{r fig.height=12, fig.width=8}
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = "Mei_et_al_CD56dim_genes1",
           plot.title = "NK CD56dim genes",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = "Mei_et_al_CD56dim_genes1",
           plot.title = "NK CD56dim genes",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15, 
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

pdf(paste0(params$out,"figures/dotplot_Mei_et_al_CD56dim_NK_subtypes_per_group.pdf"),
    height = 12, width = 8)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = "Mei_et_al_CD56dim_genes1",
           plot.title = "NK CD56dim genes",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
dev.off()

pdf(paste0(params$out,"figures/dotplot_Mei_et_al_CD56dim_group_per_NK_subtypes.pdf"),
    height = 12, width = 8)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = "Mei_et_al_CD56dim_genes1",
           plot.title = "NK CD56dim genes",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
dev.off()
```

```{r fig.height=12, fig.width=8}
CD56dim_module_list <- c("CD56dim1_module_1", "CD56dim2_module_4")

do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = CD56dim_module_list,
           plot.title = "NK CD56dim genes",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Module score")

do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = CD56dim_module_list,
           plot.title = "NK CD56dim genes",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Module score")
```

```{r fig.height=10, fig.width=10}
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = Mei_et_al_CD56dim_genes,
           plot.title = "NK CD56dim genes",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = Mei_et_al_CD56dim_genes,
           plot.title = "NK CD56dim genes",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
```

```{r fig.height=12, fig.width=22}
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_markers,
           plot.title = "",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_markers,
           plot.title = "",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

pdf(paste0(params$out,"figures/dotplot_Mei_et_al_genes_NK_subtypes_per_group.pdf"),
    height = 12, width = 22)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_markers,
           plot.title = "",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
dev.off()

pdf(paste0(params$out,"figures/dotplot_Mei_et_al_genes_NK_group_per_subtype.pdf"),
    height = 12, width = 22)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_markers,
           plot.title = "",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
dev.off()
```

```{r fig.height=12, fig.width=18}
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_markers_short,
           plot.title = "",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_markers_short,
           plot.title = "",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")

pdf(paste0(params$out,"figures/dotplot_Mei_et_al_short_genes_NK_subtypes_per_group.pdf"),
    height = 12, width = 22)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_markers_short,
           plot.title = "",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
dev.off()

pdf(paste0(params$out,"figures/dotplot_Mei_et_al_short_genes_NK_group_per_subtype.pdf"),
    height = 12, width = 22)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_markers_short,
           plot.title = "",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Scaled RNA expression")
dev.off()
```


```{r}
ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_Mei_et_al_markers$CD56bright,
                         name = "CD56bright",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_Mei_et_al_markers$CD56dim_1,
                         name = "CD56dim_1",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_Mei_et_al_markers$CD56dim_2,
                         name = "CD56dim_2",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_Mei_et_al_markers$Immature_NK,
                         name = "Immature_NK",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_Mei_et_al_markers$NKT_cells,
                         name = "NKT_cells",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = NK_Mei_et_al_markers$Active_NK_cells,
                         name = "Active_NK_cells",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)
```

```{r fig.height=12, fig.width=10}
NK_Mei_et_al_modules <- c("CD56bright1",
                          "CD56dim_11",
                          "CD56dim_21",
                          "Immature_NK",
                          "NKT_cells1",
                          "Active_NK_cells1")

pdf(paste0(params$out,"figures/dotplot_Mei_et_al_moduleScores_NK_group_per_subtype.pdf"),
    height = 12, width = 10)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_modules,
           plot.title = "Module scores",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Module score")
dev.off()

pdf(paste0(params$out,"figures/dotplot_Mei_et_al_moduleScores_NK_subtype_per_group.pdf"),
    height = 12, width = 10)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = NK_Mei_et_al_modules,
           plot.title = "Module scores",
           plot.subtitle = "Mei et al., 2024",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Module score")
dev.off()
```

## ___Crinier et al., 2021___

```{r}
# extract counts
counts_matrix <- LayerData(ALL_NK, 
                           assay = "RNA", 
                           layer = 'data')
var_genes <- dimnames(counts_matrix)[[1]]

# remove genes not present in the dataset
adaptive_NK <- intersect(crinier_et_al_genes$adaptive_NK, var_genes)
conventional_NK <- intersect(crinier_et_al_genes$conventional_NK, var_genes)
mature_NK <- intersect(crinier_et_al_genes$mature_NK, var_genes)
progenitor_NK <- intersect(crinier_et_al_genes$progenitor_NK, var_genes)
```


```{r}
ALL_NK <- AddModuleScore(ALL_NK, 
                         features = adaptive_NK,
                         name = "adaptive_NK",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = conventional_NK,
                         name = "conventional_NK",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = mature_NK,
                         name = "mature_NK",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)

ALL_NK <- AddModuleScore(ALL_NK, 
                         features = progenitor_NK,
                         name = "progenitor_NK",
                         seed = "2805",
                         slot = "data",
                         nbin = 5)
```


```{r}
crinier_et_al_module_scores <- c("mature_NK1",
                                 "adaptive_NK1",
                                 "conventional_NK1",
                                 "progenitor_NK1")

pdf(paste0(params$out,"figures/dotplot_crinier_et_al_genes_NK_subtypes_per_group.pdf"),
    height = 10, width = 8)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = crinier_et_al_module_scores,
           plot.title = "NK subset module scores",
           plot.subtitle = "Crinier et al., 2021",
           scale = TRUE,
           legend.length = 5,
           group.by = "NK_subtype_per_group",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Module scores")
dev.off()

pdf(paste0(params$out,"figures/dotplot_crinier_et_al_genes_group_per_NK_subtype.pdf"),
    height = 10, width = 8)
do_DotPlot(sample = ALL_NK,
           axis.text.x.angle = 45,
           features = crinier_et_al_module_scores,
           plot.title = "NK subset module scores",
           plot.subtitle = "Crinier et al., 2021",
           scale = TRUE,
           legend.length = 5,
           group.by = "group_per_NK_subtype",
           use_viridis = TRUE,
           number.breaks = 7,
           viridis.direction = -1,
           viridis.palette = "magma",
           font.size = 18,
           dot.scale = 15,
           dot_border = FALSE, 
           legend.position = "right",
           legend.title = "Module scores")
dev.off()
```

# ___Figure 3: DGE and CCC analysis___

```{r}

```

# ___Figure 4: ALCAM, CD6 gene expression___

```{r}
# subset scRNA-seq data to NK cells only
NK_NB_cells <- c("NK (4)", "NB (8)")
ALL_NK_NB <- subset(fetahu2023, 
                 idents = NK_NB_cells, 
                 invert = FALSE)

ALL_NK_NB <- ALL_NK_NB %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

```{r}
predicted_labels <- readRDS(file = paste0(params$out,                                 "/Rds/2024_analysis/fetahu2023_predicted_NKclusters_from_Crinier2021_de.n=350.Rds"))

predicted_labels$cell_ID <- rownames(predicted_labels)
NK_subset_labels <- data.frame(cell_ID = predicted_labels$cell_ID, 
                               NK_subsets = predicted_labels$labels)

old_metadata <- ALL_NK_NB@meta.data
old_metadata$cell_ID <- rownames(old_metadata)

new_metadata <- old_metadata %>%
  dplyr::left_join(as.data.frame(NK_subset_labels), by = "cell_ID")

#new_metadata <- new_metadata %>% rename("NK_subsets" = "predicted_NK_type_de.n350")
  
row.names(new_metadata) <- new_metadata$cell_ID 
  
ALL_NK_NB@meta.data <- new_metadata
ALL_NK_NB@meta.data

ALL_NK_NB@meta.data$cell_type_detailed <- NA

# Replace NA values in 'NK_subsets' with "NB"
ALL_NK_NB@meta.data <- ALL_NK_NB@meta.data %>%
  mutate(cell_type_detailed = ifelse(is.na(NK_subsets), "NB", NK_subsets))

head(ALL_NK_NB@meta.data$cell_type_detailed)
```


```{r fig.height=4, fig.width=9}
genes <- c("CD6","ALCAM")

combined_RNA_plot <- VlnPlot_scCustom(ALL_NK_NB, 
                 features = genes, 
                 group.by = "group",
                 split.by = "cell_type_detailed",
                 colors_use = COLOR_CODE_CCC_2,
                 pt.size = 0) 

combined_RNA_plot

pdf(paste0(params$out,"figures/vlnPlot_CD6_ALCAM_NK_NB.pdf"),
    height = 4, width = 9)
combined_RNA_plot
dev.off()

combined_RNA_plot_v2 <- VlnPlot_scCustom(ALL_NK_NB, 
                 features = genes, 
                 group.by = "cell_type_detailed",
                 split.by = "group",
                 colors_use = COLOR_CODE_GROUP,
                 pt.size = 0) 

combined_RNA_plot_v2

pdf(paste0(params$out,"figures/vlnPlot_CD6_ALCAM_NK_NB_v2.pdf"),
    height = 4, width = 9)
combined_RNA_plot_v2
dev.off()
```

```{r}
stacked_vln <- Stacked_VlnPlot(ALL_NK_NB, 
                features = genes, 
                plot_legend = TRUE,
                group.by = "group",
                split.by = "cell_type_detailed",
                colors_use = COLOR_CODE_CCC_2,
                x_lab_rotate = 45) 

stacked_vln

pdf(paste0(params$out,"figures/stacked_vlnPlot_CD6_ALCAM.pdf"),
    height = 4, width = 6)
stacked_vln
dev.off()

stacked_vln_v2 <- Stacked_VlnPlot(ALL_NK_NB, 
                features = genes, 
                plot_legend = TRUE,
                group.by = "cell_type_detailed",
                split.by = "group",
                colors_use = COLOR_CODE_GROUP,
                x_lab_rotate = 45) 

stacked_vln_v2

pdf(paste0(params$out,"figures/stacked_vlnPlot_CD6_ALCAM_v2.pdf"),
    height = 4, width = 6)
stacked_vln_v2
dev.off()
```


```{r}
combined_dotplot <- do_DotPlot(ALL_NK_NB, 
                 features = genes, 
                 group.by = "cell_type_detailed",
                 flip = TRUE,
                 dot_border = FALSE,
                 grid.type = "blank",
                 axis.text.x.angle = 45,
                 plot.title = "RNA expression",
                 scale = TRUE,
                 legend.length = 5,
                 use_viridis = TRUE,
                 number.breaks = 7,
                 viridis.direction = -1,
                 viridis.palette = "magma",
                 font.size = 18,
                 dot.scale = 15,
                 legend.title = "Log scaled RNA expression",
                 legend.position = "right"
                 )
```

```{r}
# subset scRNA-seq data to NK cells only
NK_cells <- c("NK (4)")
ALL_NK <- subset(fetahu2023, 
                 idents = NK_cells, 
                 invert = FALSE)

ALL_NK <- ALL_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```

```{r}
NB_cells <- c("NB (8)")
ALL_NB <- subset(ALL_NK_NB, 
                 idents = NB_cells, 
                 invert = FALSE)

ALL_NB <- ALL_NB %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)
```


```{r}
NK_dotplot <- do_DotPlot(ALL_NK, 
                 features = genes, 
                 group.by = "group",
                 flip = TRUE,
                 dot_border = FALSE,
                 grid.type = "blank",
                 axis.text.x.angle = 45,
                 plot.title = "NK cells",
                 scale = TRUE,
                 legend.length = 10,
                 use_viridis = TRUE,
                 number.breaks = 7,
                 viridis.direction = -1,
                 viridis.palette = "magma",
                 font.size = 18,
                 dot.scale = 15,
                 legend.title = "Scaled RNA expression"
                 )

NB_dotplot <- do_DotPlot(ALL_NB, 
                 features = genes, 
                 group.by = "group",
                 flip = TRUE,
                 dot_border = FALSE,
                 grid.type = "blank",
                 axis.text.x.angle = 45,
                 plot.title = "NB cells",
                 scale = TRUE,
                 legend.length = 5,
                 use_viridis = TRUE,
                 number.breaks = 7,
                 viridis.direction = -1,
                 viridis.palette = "magma",
                 font.size = 18,
                 dot.scale = 15,
                 legend.position = "none" 
                 )

                 
NK_dotplot + NB_dotplot

pdf(paste0(params$out,"figures/CD6_ALCAM_expression_dotplot.pdf"))
NK_dotplot + NB_dotplot
dev.off()
```
```{r}
NB_vln <- VlnPlot_scCustom(ALL_NB, 
                 features = genes, 
                 group.by = "group",
                 colors_use = COLOR_CODE_GROUP,
                 pt.size = 0) 

NK_vln <- VlnPlot_scCustom(ALL_NK, 
                 features = genes, 
                 group.by = "group",
                 colors_use = COLOR_CODE_GROUP,
                 pt.size = 0) 
NK_vln + NB_vln

```




## ___Bulk RNA-seq: ALCAM expression___

```{r}
count_data <- read.csv("~/mnt_data/nb_tumors/tumorbiologylab.bulk.RNA/nb-full_omics/full_omics_expression_data_normalized_counts_liblayout_corrected.csv")
names(count_data)[1] <- "Gene_ID"

metadata_table <- read_excel("~/mnt_data/nb_tumors/tumorbiologylab.bulk.RNA/metadata/metadata_table_wStage.xlsx")
metadata_table <- as.data.frame(metadata_table)
metadata_table$classification<- as.factor(metadata_table$classification)
```

```{r}
counts_t <- as.data.frame(melt(count_data))
names(counts_t)[3] <- "omics_id"

df <- merge(metadata_table, counts_t, by = "omics_id")
df$classification <- as.factor(df$classification)

df_filtered <- df[df$gene_name == "ALCAM", ]
ALCAM_counts <- df_filtered[, c("classification","value")]
colnames(ALCAM_counts) <- c("Group", "RNA_expression")

# sample order
SAMPLE_ORDER <- c("control", # non-tumor-infiltrated-BM
                 "ATRXdel", # sorted tumor cells
                 "MYCNamp", # sorted tumor cells
                 "ATRXwtMYCNwt", # sorted tumor cells
                 "TMEcells" # TME cells from tumor-infiltrated-BM
                 )

ALCAM_counts$Group <- factor(ALCAM_counts$Group, levels = SAMPLE_ORDER)

ALCAM_counts_clean <- ALCAM_counts %>%
  filter(!grepl("TMEcells", Group))
```


```{r}
library(ggpubr)

bulkRNA_ALCAM_plot <- ggviolin(ALCAM_counts_clean, 
                               x = "Group", 
                               y = "RNA_expression",
                               fill = "Group") + 
  labs(title = "ALCAM RNA expression", 
       y = "Normalized expression") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right") +
  scale_fill_manual(values = COLOR_CODE_GROUP) + 
  stat_compare_means(comparisons = list(c("ATRXdel", "MYCNamp"),
                                        c("ATRXdel", "ATRXwtMYCNwt")),
                     method = "wilcox.test", label = "p.format",
                     p.adjust.method = "BH")

bulkRNA_ALCAM_plot

pdf(paste0(params$out,"figures/bulkRNA_ALCAM_expression.pdf"),
    height = 5, width = 5)
bulkRNA_ALCAM_plot
dev.off()
```


# ___Figure 5: Secretome analysis___

```{r}

```

# ___Session Info___

```{r}
sessionInfo()
```

# ___The End___
