---
title: "NK project: snATAC-seq analysis"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  out: "/home/rstudio/mnt_out/"
  anno: "/home/rstudio/mnt_resources/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```


```{r load project specific R packages and color code}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
```

```{r}
readRDS(NK_ATAC, file = paste0(params$out, "/Rds/fetahu2023_scATAC_NK_processed.Rds"))
```

"scOpen"      # latent low-rank embedding from scOpen
"umap_scopen" # a UMAP embedding already computed from scOpen
"lsi"         # LSI embedding from Signac
"umap"        # another UMAP embedding (likely from LSI)

```{r}
reduction.use <- "scOpen"    
dims.use <- 2:30             

NK_ATAC <- RunUMAP(
  NK_ATAC,
  reduction = reduction.use,
  dims = dims.use
)

NK_ATAC <- FindNeighbors(object = NK_ATAC, 
                         k.param = 20,
                         reduction = reduction.use, 
                         dims = dims.use, 
                         return.neighbor = TRUE)

NK_ATAC <- FindClusters(object = NK_ATAC, 
                        verbose = FALSE, 
                        resolution = 0.2,
                        algorithm = 3)

DimPlot(object = NK_ATAC, label = TRUE) + NoLegend()
```

```{r testing functions}
run_neighbors <- function(obj, k, reduction, dims) {
  obj <- FindNeighbors(
    obj,
    reduction = reduction,
    dims = dims,
    k.param = k,
    return.neighbor = TRUE
  )
  return(obj)
}

run_clustering_resolutions <- function(obj, resolutions, algorithm = 3) {
  for (res in resolutions) {
    obj <- FindClusters(
      obj,
      resolution = res,
      algorithm = algorithm,
      verbose = FALSE
    )
  }
  return(obj)
}

compute_connectivity <- function(obj, graph.name) {
  g <- obj@graphs[[graph.name]]
  mean(Matrix::rowSums(g > 0))
}

compute_trustworthiness <- function(obj, reduction, dims, k) {
  emb <- Embeddings(obj, reduction)[, dims]
  # (Insert trustworthiness function from previous message)
  trustworthiness(emb, k = k)
}
```

```{r}
k.values    <- c(15, 20, 30, 40, 50)
res.values  <- seq(0.1, 2.0, by = 0.1)

results.list <- list()

for (k in k.values) {
  message("Running k = ", k)

  obj_k <- run_neighbors(
    obj = NK_ATAC,
    k = k,
    reduction = reduction.use,
    dims = dims.use
  )

  obj_k <- run_clustering_resolutions(
    obj = obj_k,
    resolutions = res.values
  )

  results.list[[paste0("k_", k)]] <- obj_k
}
```

```{r}
library(clustree)

obj_k20 <- results.list[["k_20"]]
clustree(obj_k20)
```

```{r}
NK_ATAC <- FindNeighbors(object = NK_ATAC, 
                         k.param = 20,
                         reduction = reduction.use, 
                         dims = dims.use, 
                         return.neighbor = TRUE)

NK_ATAC <- FindClusters(object = NK_ATAC, 
                        verbose = FALSE, 
                        resolution = 0.3,
                        algorithm = 3)

DimPlot(object = NK_ATAC, label = TRUE) + NoLegend()
```


```{r}
#saveRDS(NK_ATAC, file = paste0(params$out, "/Rds/fetahu2023_scATAC_NK_processed_v2025-12-09.Rds"))
```

# ___NK subset composition barplots___

```{r}
library(reshape2)

Idents(NK_ATAC) <- NK_ATAC$seurat_clusters

NK_ATAC$group <- as.factor(NK_ATAC$group)
NK_ATAC$seurat_clusters <- as.factor(NK_ATAC$seurat_clusters)
percent_group <- table(NK_ATAC$seurat_clusters, NK_ATAC$group)



# Transform this data in %
data_percentage <- apply(percent_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

percent_group_trans <- reshape2::melt(data_percentage) 

percent_group_trans$Var1 <- factor(percent_group_trans$Var1)

COLOR_CODE_BAR_PLOT <- c("0" = "#0A369D",
                         "1" = "#8AB17D",
                         "2" = "#E05263",
                         "3" = "#F4A261")

ggplot(percent_group_trans, aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each sub-cluster", 
       fill = "NK clusters") +  
  ggtitle("Clustered Stacked Bar Plot of NK Groups") +
  theme(axis.text.x = element_text(angle = 0, 
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",),                        
        axis.title.y = element_text(size = 16,
                                    color = "black",angle = 90),                 
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                                    color = "black")) +  
  scale_fill_manual(values = COLOR_CODE_BAR_PLOT) 
```


```{r}
NK_ATAC <- readRDS(file = paste0(params$out, "/Rds/fetahu2023_scATAC_NK_processed_v2025-12-09.Rds"))
```

```{r}
# extrapolating RNA activity based on chromatin accessibility of the corresponding promoter regions
gene.activities <- GeneActivity(NK_ATAC)

# add the gene activity matrix to the Seurat object as a new assay and normalize it
NK_ATAC[['RNA']] <- CreateAssayObject(counts = gene.activities)
NK_ATAC <- NormalizeData(
  object = NK_ATAC,
  assay = 'RNA',
  normalization.method = 'LogNormalize',
  scale.factor = median(NK_ATAC$nCount_RNA)
)
```

```{r}
DefaultAssay(NK_ATAC) <- 'RNA'

FeaturePlot(
  object = NK_ATAC,
  features = c("ZEB2","KLF2"),
  pt.size = 0.3,
  max.cutoff = 'q25',
  ncol = 2)
```

```{r fig.height=20, fig.width=20}
FeaturePlot(
  object = NK_ATAC,
  features = TFs,
  pt.size = 0.3,
  max.cutoff = 'q75',
  ncol = 4)
```

# ___Differentially accessible peaks between NK sub-types___

## ___Add promoter, enhancer regions___

https://screen.encodeproject.org/

Promoter-like (PLS)
Proximal enhancer-like (pELS)
Distal enhancer-like (dELS)
CTCF-only
DNase-H3K4me3


```{r}
library(AnnotationHub)
library(EnsDb.Hsapiens.v86)
library(rtracklayer)
library(data.table)

bed_file <- "/home/rstudio/mnt_resources/regulatory_regions/GRCh38-cCREs.bed"
# Read all columns as character to avoid scan() guessing numeric
bed_df <- fread(
  bed_file,
  header = FALSE,
  sep = "\t",
  colClasses = "character"
) 


colnames(bed_df)[1:3] <- c("chr", "start", "end")

bed_gr <- makeGRangesFromDataFrame(
  bed_df,
  seqnames.field      = "chr",
  start.field         = "start",
  end.field           = "end",
  keep.extra.columns  = TRUE
)

bed_gr

enhancer_terms <- c("dELS", "pELS", "DNase-H3K4me3")

bed_gr$region_type[bed_gr$V6 %in% enhancer_terms] <- "Enhancer"
bed_gr$region_type[bed_gr$V6 == "PLS"] <- "Promoter"

Annotation(NK_ATAC) <- bed_gr
Annotation(NK_ATAC)
```

## ___Compare groups___

if we compare every peak, we get none significant due to multiple testing correction
(> 300k tests)

```{r}
Idents(NK_ATAC) <- NK_ATAC$group
DefaultAssay(NK_ATAC) <- "peaks"
table(NK_ATAC$seurat_clusters, NK_ATAC$group)

find_DA_within_cluster <- function(obj, cluster_id, group1, group2,
                                   latent.vars = "nCount_peaks") {
  # subset to one NK cluster
  obj_sub <- subset(obj, subset = seurat_clusters == cluster_id)
  Idents(obj_sub) <- "group"
  
  da <- FindMarkers(
    obj_sub,
    ident.1      = group1,
    ident.2      = group2,
    test.use     = "LR",
    only.pos     = FALSE,
    min.pct      = 0.1,
    logfc.threshold = 0.5
  )
  
  da$peak <- rownames(da)
  da$NK_cluster <- cluster_id
  da$group1 <- group1
  da$group2 <- group2
  da
}

groups <- levels(NK_ATAC$group)
clusters <- levels(NK_ATAC$seurat_clusters)
g1 <- "ATRXmut"
g2 <- "control"
da_list <- list()

for (cl in clusters) {
    da <- find_DA_within_cluster(NK_ATAC, cl, g1,g2)
    da_list[[paste0("cl", cl, "_", g1, "_vs_", g2)]] <- da
}

# Example: DA peaks for NK cluster 0, group A vs B
da_list[["cl3_ATRXmut_vs_control"]] %>% dplyr::filter(p_val_adj < 0.05) 
```

```{r}
DefaultAssay(NK_ATAC) <- "peaks"


NK_ATAC <- FindTopFeatures(NK_ATAC, assay = "peaks", min.cutoff = "q90")

top_var_regions <- VariableFeatures(NK_ATAC)
all_peaks <- granges(NK_ATAC)  # all peaks
```

## ___Find motifs in DARs___





# ___Pseudobulk analysis___

```{r}
NK_ATAC$bulk_id <- paste0( NK_ATAC$group,"_", NK_ATAC$seurat_clusters )
NK_ATAC$bulk_id <- factor(NK_ATAC$bulk_id)

# Aggregate pseudobulk counts 
pb <- AggregateExpression(
  NK_ATAC,
  assays = "peaks",
  group.by = "bulk_id",
  return.seurat = FALSE
)
pb_counts <- pb$peaks

bulk_meta <- data.frame(bulk_id = colnames(pb_counts))

bulk_meta$NK_clusters <- c("0","1","2","3",
                           "0","1","2","3",
                           "0","1","2","3",
                           "0","1","2","3")
bulk_meta$group <- c("ATRXdel","ATRXdel","ATRXdel","ATRXdel",
                     "ATRXwtMYCNwt","ATRXwtMYCNwt","ATRXwtMYCNwt","ATRXwtMYCNwt",
                     "control","control","control","control",
                     "MYCNamp","MYCNamp","MYCNamp","MYCNamp")

bulk_meta$NK_clusters <- factor(bulk_meta$NK_clusters)
bulk_meta$group      <- factor(bulk_meta$group, ordered = FALSE,
                               levels = c("control","ATRXdel","MYCNamp","ATRXwtMYCNwt"))
bulk_meta$group <- relevel(bulk_meta$group, ref = "ATRXdel")
bulk_meta

dds <- DESeqDataSetFromMatrix(
  countData = pb_counts,
  colData   = bulk_meta,
  design    = ~ group
)


library(edgeR)

keep <- filterByExpr(dds, group = dds$group)
dds <- dds[keep,]

dds <- calcNormFactors(dds)

design <- model.matrix(~ group, data = dds$samples)
dds <- estimateDisp(dds, design)

fit <- glmFit(dds, design)
lrt <- glmLRT(fit, coef = 2) 

DARs <- as.data.frame(topTags(lrt, n = Inf))
head(DARs)

DARs_signif <- DARs %>% filter(FDR < 0.05)
granges()
```

```{r}
library(stringr)

dar_tbl <- DARs_signif %>%
tibble::rownames_to_column("region") %>%
tidyr::separate(region, into = c("chr", "start", "end"), sep = "-", convert = TRUE)
dar_gr <- GRanges(
seqnames = dar_tbl$chr,
ranges   = IRanges(start = dar_tbl$start, end = dar_tbl$end),
mcols    = dar_tbl %>% select(-chr, -start, -end)
)

table(bed_gr$region_type)
hits <- findOverlaps(dar_gr, bed_gr, ignore.strand = TRUE)
hits

# annotate
annotated <- data.frame(
  #dar_id     = dar_gr[queryHits(hits)],
  seqnames   = as.character(seqnames(dar_gr)[queryHits(hits)]),
  dar_start  = start(dar_gr)[queryHits(hits)],
  dar_end    = end(dar_gr)[queryHits(hits)],
  logFC      = dar_gr$mcols.logFC[queryHits(hits)],
  FDR        = dar_gr$mcols.FDR[queryHits(hits)],
  
  RE_type    = bed_gr$region_type[subjectHits(hits)],
  RE_id      = bed_gr$V5[subjectHits(hits)],
  RE_feature = bed_gr$V6[subjectHits(hits)]
)

annotated

table(annotated$RE_feature)

# --------------------------------------- #
gr <- GRanges(
    seqnames = annotated$seqnames,
    ranges   = IRanges(start = annotated$dar_start,
                       end   = annotated$dar_end),
    mcols    = annotated
  )

gr

hits <- findOverlaps(gr, Annotation(NK_ATAC), ignore.strand = TRUE)
hits

if (length(hits) > 0) {
    gene_df <- data.frame(
      index       = queryHits(hits),
      gene_seq    = as.character(seqnames(Annotation(NK_ATAC))[subjectHits(hits)]),
      gene_start  = start(Annotation(NK_ATAC))[subjectHits(hits)],
      gene_end    = end(Annotation(NK_ATAC))[subjectHits(hits)],
      mcols(Annotation(NK_ATAC))[subjectHits(hits), , drop = FALSE] %>% as.data.frame()
    )

    # Many-to-one mapping: join back using row index
    re_out <- annotated %>%
      mutate(index = row_number()) %>%
      left_join(gene_df, by = "index") %>%
      select(-index)

  } else {

    # If no gene hits, append NA columns for gene annotations
    gene_meta_cols <- colnames(mcols(Annotation(NK_ATAC)))

    re_out <- annotated %>%
      mutate(
        gene_seq   = NA,
        gene_start = NA,
        gene_end   = NA
      ) %>%
      bind_cols(
        as.data.frame(matrix(
          NA,
          nrow = nrow(annotated),
          ncol = length(gene_meta_cols),
          dimnames = list(NULL, gene_meta_cols)
        ))
      )
  }

table(re_out$gene_name)
```

```{r}
gene_of_interest <- "ZNF215"

gene_of_interest <- TFs[2]

CoveragePlot(peaks = TRUE,
             group.by = "bulk_id",
             window = 2000,
             object = NK_ATAC,
             region = gene_of_interest,
             extend.upstream = 10000,
             extend.downstream = 5000
)

CoveragePlot(peaks = TRUE,
             group.by = "group",
             window = 2000,
             object = NK_ATAC,
             region = gene_of_interest,
             extend.upstream = 10000,
             extend.downstream = 5000
)
```

# ___TF footprinting___

```{r}
library(motifmatchr)
library(JASPAR2024)
library(TFBSTools)
library(BSgenome.Hsapiens.UCSC.hg38)
library(RSQLite)

JASPAR2024 <- JASPAR2024()
con24 <- RSQLite::dbConnect(RSQLite::SQLite(), db(JASPAR2024))

# extract position frequency matrices for the motifs
pwm <- getMatrixSet(
  x = con24, # why does JASPAR2024 not work?
  opts = list(species = 9606, all_versions = TRUE)
)
pwm

# add motif information
NK_ATAC <- AddMotifs(NK_ATAC, genome = BSgenome.Hsapiens.UCSC.hg38, pfm = pwm)

# gather the footprinting information for set of motifs
NK_ATAC <- Footprint(
  object = NK_ATAC,
  motif.name = "TBX21",
  genome = BSgenome.Hsapiens.UCSC.hg38
)

# plot the TF footprints
PlotFootprint(NK_ATAC, 
              group.by = "bulk_id",
              features = "TBX21")
```


## ___ZNF215 motif___

```{r}
ZNF215_motif <- "/home/rstudio/mnt_resources/TFs/MA2547.1_ZNF215.jaspar"

# Read PFM(s) 
pfm_list <- readJASPARMatrix(ZNF215_motif, matrixClass = "PFM")
pfm_znf215 <- pfm_list[[1]]

# create PFMatrix from matrix
ZNF215_pwm <- PFMatrix(
  ID = "ZNF215",
  name = "ZNF215",
  matrixClass = "PFM",
  strand = "+",
  bg = c(A = 0.25, C = 0.25, G = 0.25, T = 0.25),
  profileMatrix = as.matrix(pfm_znf215)
)

pwm_znf215 <- toPWM(pfm_znf215)

pfm_list <- PFMatrixList(pfm_znf215)
names(pfm_list) <- pfm_znf215@name  

NK_ATAC <- AddMotifs(
  object = NK_ATAC,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  pfm    = pfm_list
)

NK_ATAC <- Footprint(
  object    = NK_ATAC,
  genome    = BSgenome.Hsapiens.UCSC.hg38,
  motif.name = "ZNF215",    
  assay     = "peaks"     
)

PlotFootprint(NK_ATAC, 
              group.by = "bulk_id",
              features = "ZNF215",
              assay     = "peaks")

PlotFootprint(NK_ATAC, 
              group.by = "group",
              features = "ZNF215",
              assay     = "peaks",)
```

# ___Label transfer from scRNA-seq data___

two options: Original reference data crinier2021 OR fetahu2023 RNA data w/Label transfer from crinier2021 

```{r}
NK_ATAC <- readRDS(file = paste0(params$out, "/Rds/fetahu2023_scATAC_NK_processed_v2025-12-09.Rds"))
```

```{r}
frag_file <- paste0(params$out, "/Rds/feathu2023_ATAC_NKcells_fragments.tsv.gz")
fr_obj <- CreateFragmentObject(path = frag_file, cells = colnames(NK_ATAC))
NK_ATAC[["peaks"]]@fragments <- list(fr_obj)

# fix fragment file
frag <- Fragments(NK_ATAC)[[1]]@path
file.exists(frag)
file.exists(paste0(frag, ".tbi"))
```

```{r}
# Compute gene activity from peaks
gene.activities <- GeneActivity(NK_ATAC)

# Add as a new assay
NK_ATAC[["GeneActivity"]] <- CreateAssayObject(counts = gene.activities)
DefaultAssay(NK_ATAC) <- "GeneActivity"

# Standard normalization on gene activity
NK_ATAC <- NormalizeData(NK_ATAC)
NK_ATAC <- FindVariableFeatures(NK_ATAC, nfeatures = 10000)
NK_ATAC <- ScaleData(NK_ATAC)
```

```{r}
NK_RNA <- readRDS(paste0(params$out, "/Rds/ALL_NK_cluster_withNewMetadata.Rds"))

NK_RNA <- NormalizeData(NK_RNA)
NK_RNA <- FindVariableFeatures(NK_RNA, nfeatures = 5000)
NK_RNA <- ScaleData(NK_RNA)

DefaultAssay(NK_RNA) <- "RNA"
DefaultAssay(NK_ATAC)  <- "GeneActivity"  # for anchoring

# Intersect features
common.features <- intersect(
  VariableFeatures(NK_RNA),
  rownames(NK_ATAC[["GeneActivity"]])
)
length(common.features)
```

## ___1. Seurat/Signac CCA integration and label transfer___

```{r}
transfer.anchors <- FindTransferAnchors(
  reference        = NK_RNA,
  query            = NK_ATAC,
  features         = common.features,
  reference.assay  = "RNA",
  query.assay      = "GeneActivity",
  reduction        = "cca",     # CCA between RNA and gene activity
  dims             = 2:30
)

# Use LSI as weight reduction (chromatin space) for the ATAC object
DefaultAssay(NK_ATAC) <- "peaks"  # ensure LSI is linked to this assay

predictions <- TransferData(
  anchorset        = transfer.anchors,
  refdata          = NK_RNA$predicted_NK_type_de.n350,  
  weight.reduction = NK_ATAC[["lsi"]],
  dims             = 2:30
)

# Add predicted labels and scores
NK_ATAC <- AddMetaData(NK_ATAC, metadata = predictions)

# Inspect metadata columns
head(NK_ATAC@meta.data)
```

```{r}
DimPlot(NK_ATAC,group.by = "predicted.id")

table(NK_ATAC$predicted.id,NK_ATAC$group)
table(NK_ATAC$predicted.id,NK_ATAC$seurat_clusters)
```

```{r}
Idents(NK_ATAC) <- "predicted.id"

VlnPlot(NK_ATAC, features = "prediction.score.max", pt.size = 0)

# Filter out low-confidence assignments, if desired
high_conf_cells <- WhichCells(NK_ATAC, expression = prediction.score.max > 0.6)
atac_nk_high_conf <- subset(NK_ATAC, cells = high_conf_cells)

DimPlot(NK_ATAC, reduction = "umap_scopen", group.by = "predicted.id")
```

```{r}
library(reshape2)

Idents(NK_ATAC) <- NK_ATAC$predicted.id

NK_ATAC$group <- as.factor(NK_ATAC$group)
NK_ATAC$predicted.id <- as.factor(NK_ATAC$predicted.id)
percent_group <- table(NK_ATAC$predicted.id, NK_ATAC$group)


# Transform this data in %
data_percentage <- apply(percent_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

percent_group_trans <- reshape2::melt(data_percentage) 

percent_group_trans$Var1 <- factor(percent_group_trans$Var1)

ggplot(percent_group_trans, aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each sub-cluster", 
       fill = "NK clusters") +  
  ggtitle("Clustered Stacked Bar Plot of NK Groups") +
  theme(axis.text.x = element_text(angle = 0, 
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",),                        
        axis.title.y = element_text(size = 16,
                                    color = "black",angle = 90),                 
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                                    color = "black")) +  
  scale_fill_manual(values = COLOR_CODE_v2) 
```


```{r}
Idents(NK_ATAC) <- NK_ATAC$predicted.id
DefaultAssay(NK_ATAC) <- "peaks"
table(NK_ATAC$predicted.id, NK_ATAC$group)

find_DA_within_cluster <- function(obj, cluster_id, group1, group2,
                                   latent.vars = "nCount_peaks") {
  # subset to one NK cluster
  obj_sub <- subset(obj, subset = predicted.id == cluster_id)
  Idents(obj_sub) <- "group"
  
  da <- FindMarkers(
    obj_sub,
    ident.1      = group1,
    ident.2      = group2,
    test.use     = "LR",
    only.pos     = FALSE,
    min.pct      = 0.1,
    logfc.threshold = 0.5
  )
  
  da$peak <- rownames(da)
  da$NK_cluster <- cluster_id
  da$group1 <- group1
  da$group2 <- group2
  da
}

groups <- levels(NK_ATAC$group)
clusters <- levels(NK_ATAC$predicted.id)
g1 <- "ATRXmut"
g2 <- "control"
da_list <- list()

for (cl in clusters) {
    da <- find_DA_within_cluster(NK_ATAC, cl, g1,g2)
    da_list[[paste0("cl", cl, "_", g1, "_vs_", g2)]] <- da
}

#  DA peaks for NK cluster 3, group A vs B
DARs_hNK_Bm3 <- da_list[["clhNK_Bm3_ATRXmut_vs_control"]] %>% dplyr::filter(p_val_adj < 0.05) 
```

```{r}
library(stringr)

dar_tbl <- DARs_hNK_Bm3 %>%
tibble::rownames_to_column("region") %>%
tidyr::separate(region, into = c("chr", "start", "end"), sep = "-", convert = TRUE)

dar_gr <- GRanges(
seqnames = dar_tbl$chr,
ranges   = IRanges(start = dar_tbl$start, end = dar_tbl$end),
mcols    = dar_tbl %>% select(-chr, -start, -end)
)

table(bed_gr$region_type)
hits <- findOverlaps(dar_gr, bed_gr, ignore.strand = TRUE)
hits

# annotate
annotated <- data.frame(
  #dar_id     = dar_gr[queryHits(hits)],
  seqnames   = as.character(seqnames(dar_gr)[queryHits(hits)]),
  dar_start  = start(dar_gr)[queryHits(hits)],
  dar_end    = end(dar_gr)[queryHits(hits)],
  logFC      = dar_gr$mcols.avg_log2FC[queryHits(hits)],
  p.adj        = dar_gr$mcols.p_val_adj[queryHits(hits)],
  
  RE_type    = bed_gr$region_type[subjectHits(hits)],
  RE_id      = bed_gr$V5[subjectHits(hits)],
  RE_feature = bed_gr$V6[subjectHits(hits)]
)

annotated

table(annotated$RE_feature)

# --------------------------------------- #
gr <- GRanges(
    seqnames = annotated$seqnames,
    ranges   = IRanges(start = annotated$dar_start,
                       end   = annotated$dar_end),
    mcols    = annotated
  )

gr

hits <- findOverlaps(gr, Annotation(NK_ATAC), ignore.strand = TRUE)
hits

if (length(hits) > 0) {
    gene_df <- data.frame(
      index       = queryHits(hits),
      gene_seq    = as.character(seqnames(Annotation(NK_ATAC))[subjectHits(hits)]),
      gene_start  = start(Annotation(NK_ATAC))[subjectHits(hits)],
      gene_end    = end(Annotation(NK_ATAC))[subjectHits(hits)],
      mcols(Annotation(NK_ATAC))[subjectHits(hits), , drop = FALSE] %>% as.data.frame()
    )

    # Many-to-one mapping: join back using row index
    re_out <- annotated %>%
      mutate(index = row_number()) %>%
      left_join(gene_df, by = "index") %>%
      select(-index)

  } else {

    # If no gene hits, append NA columns for gene annotations
    gene_meta_cols <- colnames(mcols(Annotation(NK_ATAC)))

    re_out <- annotated %>%
      mutate(
        gene_seq   = NA,
        gene_start = NA,
        gene_end   = NA
      ) %>%
      bind_cols(
        as.data.frame(matrix(
          NA,
          nrow = nrow(annotated),
          ncol = length(gene_meta_cols),
          dimnames = list(NULL, gene_meta_cols)
        ))
      )
  }

table(re_out$gene_name)
```

```{r}
NK_ATAC$predicted_NK_subtypes <- paste0(NK_ATAC$group,"_", NK_ATAC$predicted.id)
NK_ATAC$predicted_NK_subtypes <- factor(NK_ATAC$predicted_NK_subtypes)

candidate_genes <- c("ZNF471","DOCK1")

for (gene in candidate_genes){
  CoveragePlot(peaks = TRUE,
             group.by = "predicted_NK_subtypes",
             window = 2000,
             object = NK_ATAC,
             region = gene,
             extend.upstream = 10000,
             extend.downstream = 5000)
  
  CoveragePlot(peaks = TRUE,
             group.by = "group",
             window = 2000,
             object = NK_ATAC,
             region = gene,
             extend.upstream = 10000,
             extend.downstream = 5000)
}

```

## ___2. SingleR label transfer___

```{r}

# Compute gene activity from peaks
gene.activities <- GeneActivity(NK_ATAC)

# Add as a new assay
NK_ATAC[["GeneActivity"]] <- CreateAssayObject(counts = gene.activities)
DefaultAssay(NK_ATAC) <- "GeneActivity"

# Standard normalization on gene activity
NK_ATAC <- NormalizeData(NK_ATAC)

NK_ATAC <- 
  NK_ATAC %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)

NK_ATAC <- FindVariableFeatures(NK_ATAC, nfeatures = 10000)
NK_ATAC <- ScaleData(NK_ATAC)

Assay(NK_ATAC) <- "GeneActivity"

# RNA data
NK_RNA <- 
  NK_RNA %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)


# RNA as reference
NK_clusters <- NK_RNA$predicted_NK_type_de.n350
original_clusters <- NK_ATAC$seurat_clusters

NK_ATAC_mat <- GetAssayData(NK_ATAC, 
                        assay = "RNA", 
                        slot = "data")

NK_RNA_mat <- GetAssayData(NK_RNA, 
                            assay = "RNA", 
                            slot = "data")
```

(optional) remove NK cluster 4

```{r (optional) remove NK cluster 4}
# remove cluster hNK_Bm4 due to low abundance
Idents(NK_RNA) <- NK_RNA$predicted_NK_type_de.n350

NK_RNA_core <- subset(NK_RNA, idents = "hNK_Bm4", invert = TRUE)
NK_RNA_core$predicted_NK_type_de.n350 <- factor(NK_RNA_core$predicted_NK_type_de.n350)

# RNA data
NK_RNA_core <- 
  NK_RNA_core %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)

NK_RNA_mat <- GetAssayData(NK_RNA_core, 
                            assay = "RNA", 
                            slot = "data")

NK_clusters <- NK_RNA_core$predicted_NK_type_de.n350 
```

#### ___Sensitivity analysis for de.n parameter___

```{r}
# Store de.n values [Number of DE genes] from 5 to 500 with a 10 step increment
de_n_values <- seq(5, 500, by=10) 

# store results
results <- data.frame(de_n = integer(), cluster = character(), num_cells = integer())

# Loop through de.n values
for (de_n in de_n_values) {
 
  # Run SingleR with the current de.n parameter
  singleR_result <- SingleR(test = NK_ATAC_mat, 
                            ref = NK_RNA_mat, 
                            labels = NK_clusters, 
                            de.n = de_n,
                            num.threads = 30)
  
  # Get the number of cells per cluster
  cluster_counts <- table(singleR_result$labels)
  
  # Add the results to the data frame
  results <- rbind(results, data.frame(de_n = de_n, 
                                       cluster = names(cluster_counts), 
                                       num_cells = as.integer(cluster_counts)))
}

saveRDS(singleR_result, file = paste0(params$out,                       "Rds/Label_transfer_sensitivity_analysis_20260112_without-cluster-NK4.Rds"))

write.table(results, file = paste0(params$out,                                "tables/label_transfer_sensitivity_analysis_20260112_without-cluster-NK4.csv"),
            sep = "\t", quote = FALSE,row.names = FALSE)
```

```{r}
# Plot the results using ggplot2
sensitivity_analysis <- ggplot(results, aes(x = de_n, y = num_cells, color = cluster)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  labs(title = "Number of cells per cluster vs. de.n parameter",
       x = "de.n parameter",
       y = "Number of cells per cluster") +
  theme_minimal() +  scale_color_manual(values = COLOR_CODE_v2)

sensitivity_analysis

# save as PDF
pdf(paste0(params$out,"figures/label_transfer_sensitivity_analysis_results_20260112_without-cluster-NK4.pdf"),
    height = 3, width = 5)
sensitivity_analysis
dev.off()
```

```{r}
# Load required libraries
library(reshape2)

#  range for de.n parameter
de_n_values <- seq(5, 1000, by=10)  

# Initialize a list to store results
results_list <- list()

# Loop through de.n values
for (de_n in de_n_values) {
  
  # Run SingleR with the current de.n parameter
  singleR_result <- SingleR(test = NK_ATAC_mat, 
                            ref = NK_RNA_mat, 
                            labels = NK_clusters, 
                            de.n = de_n,
                            num.threads = 30)
  
  # Store the results in a list, using de.n as the name
  results_list[[as.character(de_n)]] <- singleR_result$labels
}

# Convert the results list to a data frame
results_df <- do.call(cbind, results_list)
colnames(results_df) <- de_n_values
results_df <- as.data.frame(results_df)
results_df$de_n <- de_n_values #
results_df$Cell_ID <- rownames(results_df)

# Melt the data frame for plotting
results_melted <- reshape2::melt(results_df, 
                       id.vars = "Cell_ID", 
                       variable.name = "de_n", 
                       value.name = "Predicted_Type")

results_melted

# Make sure de.n is treated as a numeric variable
results_melted$de_n <- as.numeric(as.character(results_melted$de_n))

# Plot the heatmap
params_sensitivity_heatmap_plot <- ggplot(results_melted, aes(x = de_n, 
                                                              y = Cell_ID, 
                                                              fill = Predicted_Type)) +
  geom_tile() +
  scale_fill_manual(values = COLOR_CODE_v2) +
  labs(title = "Heatmap of predicted NK sub-types across different de.n values",
       x = "de.n parameter",
       y = "Cell ID",
       fill = "Predicted NK Type") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),  # Optional: Hide y-axis text if there are too many cells
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())


write.table(results_df, file = paste0(params$out,"tables/label_transfer_sensitivity_analysis_per_cell_ID_20260112_without-cluster-NK4.csv"),
            sep = "\t")
```

```{r}
params_sensitivity_heatmap_plot

pdf(paste0(params$out,"figures/label_transfer_heatmap_plot_20260112_without-cluster-NK4.pdf"))
params_sensitivity_heatmap_plot
dev.off()
```

```{r subsampled}
set.seed(123)
num_cells <- nrow(results_df)

# If the number of cells is greater than 500, sub-sample
if (num_cells > 500) {
  # Randomly sample 500 cell IDs
  sampled_cell_ids <- sample(results_df$Cell_ID, 500, replace = FALSE)
  
  # Subset results_df to include only the sampled cells
  results_df_subsampled <- results_df[results_df$Cell_ID %in% sampled_cell_ids, ]
} else {
  # If there are fewer than 500 cells, use the entire dataset
  results_df_subsampled <- results_df
  warning("The number of cells is less than or equal to 500; using the entire dataset.")
}

# Display the first few rows of the subsampled data frame
head(results_df_subsampled)

# Melt the data frame for plotting
df_subsampled_melted <- reshape2::melt(results_df_subsampled, 
                       id.vars = "Cell_ID", 
                       variable.name = "de_n", 
                       value.name = "Predicted_Type")

df_subsampled_melted

# Make sure de.n is treated as a numeric variable
df_subsampled_melted$de_n <- as.numeric(as.character(df_subsampled_melted$de_n))

# Plot the heatmap
params_sensitivity_plot <- ggplot(df_subsampled_melted, 
                                  aes(x = de_n, y = Cell_ID, fill = Predicted_Type)) +
  geom_tile() +
  scale_fill_manual(values = COLOR_CODE_v2) +
  labs(title = "Heatmap of predicted NK sub-types across different de.n values",
       x = "de.n parameter",
       y = "Cell ID",
       fill = "Predicted NK Type") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),  # Optional: Hide y-axis text if there are too many cells
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())
```

```{r}
params_sensitivity_plot

pdf(params$out,"figures/label_transfer_sensitivity_analysis_v2_20260112_without-cluster-NK4.pdf")
params_sensitivity_plot
dev.off()
```

### ___Label transfer___

```{r}
predicted_labels <- SingleR(test = NK_ATAC_mat, 
                            ref = NK_RNA_mat, 
                            labels = NK_clusters, 
                            de.n = 250, # based on above sensitivity analysis
                            num.threads = 30) # GPU server
```

### ___QC of label transfer___

```{r QC of predicted labels: NK_RNA}
# filtering step
predicted_label <- predicted_labels$labels

final_labels <- predicted_labels$pruned.labels
final_labels[is.na(final_labels)] <- "unassigned"

# Compute an empirical delta threshold from non-pruned cells
ok <- predicted_labels$pruned.labels %in% unique(predicted_labels$pruned.labels[!is.na(predicted_labels$pruned.labels)])
delta_ok <- predicted_labels$delta[ok]

# Example: conservative threshold = 20th percentile of delta among pruned-kept cells
delta_thr <- unname(quantile(delta_ok, probs = 0.20, na.rm = TRUE))

# Apply: require delta >= threshold
low_delta <- predicted_labels$delta < delta_thr | is.na(predicted_labels$delta)
final_labels[low_delta] <- "unassigned"
final_labels <- factor(final_labels)

##########

# Start with SingleR pruning
final <- predicted_labels$pruned.labels
final[is.na(final)] <- "unassigned"
table(final)

# restrict threshold estimation to cells that passed pruning
keep0 <- predicted_labels$pruned.labels[!is.na(predicted_labels$pruned.labels)]
delta0 <- predicted_labels$delta.next[!is.na(predicted_labels$pruned.labels)]

# percentile-based threshold (increase p for stricter filtering)
p <- 0.50
delta_thr <- unname(quantile(delta0, probs = p, na.rm = TRUE))

# Apply threshold
final[final != "unassigned" & predicted_labels$delta.next < delta_thr] <- "unassigned"
table(final)

##########

plotScoreHeatmap(predicted_labels, 
                 normalize = TRUE,
                 show.labels = TRUE,
                 annotation_col = as.data.frame(NK_ATAC@meta.data[,"sample",
                                                         drop = FALSE]))

plotDeltaDistribution(predicted_labels, 
                      ncol = 2)

# defult delta treshold
to.remove <- is.na(predicted_labels$pruned.labels)
table(Label=predicted_labels$labels, Removed = to.remove)
# manual delta treshold
to.remove <- pruneScores(predicted_labels, min.diff.med = 0.01)
table(Label=predicted_labels$labels, Removed = to.remove)

plotScoreHeatmap(predicted_labels[to.remove,], 
                 normalize = TRUE,
                 show.labels = TRUE,
                 annotation_col = as.data.frame(NK_ATAC@meta.data[,"sample",
                                                         drop = FALSE]))

to.remove <- pruneScores(predicted_labels, min.diff.med = 0.001)
table(Label=predicted_labels$labels, Removed = to.remove)

plotScoreHeatmap(predicted_labels[to.remove,], 
                 normalize = TRUE,
                 show.labels = TRUE,
                 annotation_col = as.data.frame(NK_ATAC@meta.data[,"sample",
                                                         drop = FALSE]))
```


```{r}
# save plot
pdf(paste0(params$out,"figures/RNAtoATAC_label_transfer_QC_without-cluster-NK4.pdf"),
    height = 5,width = 7)
plotScoreHeatmap(predicted_labels, 
                 normalize = TRUE,
                 show.labels = TRUE,
                 annotation_col = as.data.frame(NK_ATAC@meta.data[,"sample",
                                                         drop = FALSE]))
dev.off()
```


```{r label transfer results: NK_RNA}
NK_ATAC$predicted_NK_type <- predicted_label

cluster_anno <- c("hNK_Bm1","hNK_Bm2","hNK_Bm3","hNK_Bm4")

DimPlot(NK_ATAC,
        group.by = "predicted_NK_type",
        cols = COLOR_CODE_GROUP)

DimPlot(NK_ATAC,
        group.by = "predicted_NK_type",
        split.by = "group",
        cols = COLOR_CODE_GROUP)

percent_NK_group <- table(NK_ATAC$predicted_NK_type,
                          NK_ATAC$group)
percent_NK_group

percent_NK_sample <- table(NK_ATAC$predicted_NK_type,
                          NK_ATAC$sample)

barplot(percent_NK_group,
        col = COLOR_CODE_GROUP, 
        border = "white", 
        xlab = "Group",
        legend.text = cluster_anno)
        
# Transform this data in %
data_percentage <- apply(percent_NK_group, 2, 
                         function(x){x*100/sum(x,na.rm=T)})
 
data_percentage_sample <- apply(percent_NK_sample, 2, 
                         function(x){x*100/sum(x,na.rm=T)})

# Make a stacked barplot in %
barplot(data_percentage, 
        col = COLOR_CODE_v2, 
        border = "white", 
        xlab = "Group",
        legend.text = cluster_anno)

barplot(data_percentage_sample, 
        col = COLOR_CODE_v2, 
        border = "white", 
        xlab = "Group",
        legend.text = cluster_anno)

# with ggplot2

percent_NK_group_trans <- reshape2::melt(data_percentage) 

ggplot(percent_NK_group_trans, aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each sub-cluster", 
       fill = "NK clusters") +  
  ggtitle("Clustered Stacked Bar Plot of NK Groups") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",),                        
        axis.title.y = element_text(size = 16,
                                    angle = 90,
                                    color = "black",),                 
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                                    color = "black")) +
  scale_fill_manual(values = COLOR_CODE_v2)


# all samples
percent_NK_samples_trans <- reshape2::melt(data_percentage_sample) 
data_wide <- t(data_percentage_sample)
d <- dist(data_wide, method = "euclidean")
hc <- hclust(d, method = "complete")
ordered_samples <- rownames(data_wide)[hc$order]
percent_NK_samples_trans <- percent_NK_samples_trans %>%
  arrange(Var2)

print(percent_NK_samples_trans)

ggplot(percent_NK_samples_trans, aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each sub-cluster", 
       fill = "NK clusters") +  
  ggtitle("Clustered Stacked Bar Plot of NK Groups") +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",),                        
        axis.title.y = element_text(size = 16,
                                    angle = 90,
                                    color = "black",),                 
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                                    color = "black")) +
  scale_fill_manual(values = COLOR_CODE_v2)


# version 2
ordered_samples <- sort(ordered_samples)
percent_NK_samples_trans$Var2 <- factor(percent_NK_samples_trans$Var2, levels = ordered_samples)
percent_NK_samples_trans <- percent_NK_samples_trans %>% 
  arrange(Var2,value)

ggplot(percent_NK_samples_trans, aes(x = Var2, y = value, fill = Var1)) + 
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells assigned to each sub-cluster", 
       fill = "NK clusters") +  
  ggtitle("Stacked Bar Plot of NK Groups") +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",),                        
        axis.title.y = element_text(size = 16,
                                    angle = 90,
                                    color = "black",),                 
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                                    color = "black")) +
  scale_fill_manual(values = COLOR_CODE_v2)
```

### ___Save predicted labels___

```{r eval=FALSE, include=FALSE}
saveRDS(predicted_labels, 
        file = paste0(params$out,
                      "/Rds/NK_ATAC_predicted_NKclusters_without-cluster-NK4_RNAtoATAC_de.n=250.Rds"))

saveRDS(NK_ATAC, 
       file = paste0(params$out, "/Rds/NK_ATAC_wMatchedRNAcells_without-cluster-NK4_v2026-01_de.n=250.Rds"))
```


## ___3. ArchR label transfer___

```{r}
# devtools::install_github("GreenleafLab/ArchR", ref="master", repos = BiocManager::repositories())

library("ArchR")
set.seed(2805)

# archR params setup
Sys.setenv(PATH = paste("/opt/conda/envs/atacseq/bin", 
                        Sys.getenv("PATH"), sep = .Platform$path.sep))

addArchRThreads(threads = 24) 
addArchRGenome("hg38")
macs2 <- "/opt/conda/envs/atacseq/bin/macs2"

#frag_file <- paste0(params$out,"/Rds/feathu2023_ATAC_NKcells_fragments.tsv.gz")
names(frag_file) <- "NK_ATAC"

ArrowFiles <- createArrowFiles(
  inputFiles = frag_file, # loaded in the code above
  sampleNames = names(frag_file),
  minTSS = 4, # adjust
  minFrags = 1000, # adjust
  addTileMat = TRUE,
  addGeneScoreMat = TRUE
)
```

Per-cell quality control information that was calculated during Arrow file creation:
TSSEnrichment - the per-cell TSS enrichment score
ReadsInTSS - the number of reads that fall within TSS regions (default is 100 bp around TSS)
ReadsInPromoter - the number of reads that fall in promoter regions (default is -2000 to +100 from the TSS)
PromoterRatio - the ratio of reads in promoters to reads outside of promoters
ReadsInBlacklist - the number of reads that fall in blacklist regions
BlacklistRatio - the ratio of reads in blacklist regions to reads outside of blacklist regions
NucleosomeRatio - similar to above but for reads mapping to nucleosome-sized reads, defined as: (nDiFrags + nMultiFrags) / nMonoFrags)
nFrags - the number of fragments recovered per cell
nMonoFrags - the number of fragments that have length less than 2 * nucLength (nucLength is the length of DNA that wraps around a nucleosome (147 bp by default))
nDiFrags - the number of fragments that have length less than 3 * nucLength but greater than 2 * nucLength
nMultiFrags - the number of fragments that have length >= 3*nucLength
PassQC - equal to 1 if the cell passed QC filters or 0 if not

```{r}
# create archR project file
proj <- ArchRProject(
  ArrowFiles       = ArrowFiles,
  outputDirectory  = paste0(params$out,"/ArchR"),
  copyArrows       = TRUE
)

# TSS enrichment plot
df <- getCellColData(proj, select = c("log10(nFrags)", "TSSEnrichment"))

p <- ggPoint(
    x = df[,1], 
    y = df[,2], 
    colorDensity = TRUE,
    continuousSet = "sambaNight",
    xlabel = "Log10 Unique Fragments",
    ylabel = "TSS Enrichment",
    xlim = c(log10(500), quantile(df[,1], probs = 0.99)),
    ylim = c(0, quantile(df[,2], probs = 0.99))
) + geom_hline(yintercept = 4, lty = "dashed") + geom_vline(xintercept = 3, lty = "dashed")

p

plotPDF(p, name = "NK_ATAC_archR_TSS-vs-frag_QC.pdf", ArchRProj = proj, addDOC = FALSE)

# fragment size distribution plot
p1 <- plotFragmentSizes(
  ArchRProj = proj,
  sample    = "NK_ATAC",
  plotAs    = "line"
)

p2 <- plotTSSEnrichment(ArchRProj = proj)

plotPDF(p1,p2, name = "NK_ATAC_archR_FragSizes+TSSProfile.pdf", 
        ArchRProj = proj, addDOC = FALSE, width = 5, height = 5)

```
 
```{r}
doubScores <- addDoubletScores(
    input = ArrowFiles,
    k = 10, # adjust! 
    # how many cells near a "pseudo-doublet" the algorithm should count
    knnMethod = "UMAP", # can change to  knnMethod = "LSI" and force = TRUE 
    LSIMethod = 1
)
```


ArchR reports the R2 value for the UMAP projection for each Arrow file. If these R2 values are much lower (i.e. less than 0.9), this often indicates that the cells within the Arrow file have very little heterogeneity. This makes the accuracy of doublet calling worse because the majority of doublets would be “homotypic” - or a single droplet with two very similar cells. 

In these cases, ArchR will skip doublet prediction because it is not accurate. Alternatively, you can try setting knnMethod = "LSI" and force = TRUE to perform projection in the LSI subspace; however, you should manually assess the results and make sure this is performing as you would expect.

Adding doublet scores will create plots in the “QualityControl” directory. There are 3 plots associated with each of your samples in this folder:

Doublet Enrichments - These represent the enrichment of simulated doublets nearby each single cell compared to the expected if we assume a uniform distribution.

Doublet Scores - These represent the significance (-log10(binomial adjusted p-value)) of simulated doublets nearby each single cell compared to the expected if we assume a uniform distribution. We have found this value to be less consistent than the doublet enrichments and therefore use doublet enrichments for doublet identification.

Doublet Density - This represents the density of the simulated doublet projections. This allows you to visualize where the synthetic doublets were located after projection into your 2-dimensional embedding.

### ___QC filtering and Doublet removal___

```{r}
cd <- getCellColData(proj, select = c("nFrags", "TSSEnrichment"))
keep <- rownames(cd)[
  cd$nFrags >= 3000 &
  cd$nFrags <= 25000 &
  cd$TSSEnrichment >= 10
]

proj_filt <- subsetArchRProject(
  ArchRProj = proj,
  cells = keep,
  outputDirectory = "ArchR_QCfiltered",
  dropCells = TRUE,
  force = TRUE
)

table(proj_filt$DoubletEnrichment) 
summary(proj_filt$DoubletScore)
```

```{r}
# Confirm available per-cell QC fields
colnames(getCellColData(proj_filt))

# Pull the relevant columns
cd <- as.data.frame(getCellColData(
  proj_filt,
  select = c("nFrags", "TSSEnrichment", "DoubletScore", "DoubletEnrichment", "BlacklistRatio")
))

summary(cd$DoubletScore)
summary(cd$DoubletEnrichment)
quantile(cd$DoubletEnrichment, probs = c(0.5, 0.9, 0.95, 0.99))

ggplot(cd, aes(x = log10(nFrags), y = DoubletEnrichment)) +
  geom_point(size = 0.6, alpha = 0.6) +
  theme_classic() +
  xlab("log10(nFrags)") + ylab("DoubletEnrichment")

ggplot(cd, aes(x = DoubletScore, y = DoubletEnrichment)) +
  geom_point(size = 0.6, alpha = 0.6) +
  theme_classic()
```

```{r}
proj_auto <- filterDoublets(proj)
c(before = nCells(proj_filt), after = nCells(proj_auto), removed = nCells(proj_filt) - nCells(proj_auto))
```

ATAC data pre-processing with archR

```{r}
# Standard ATAC preprocessing
proj_filt <- addIterativeLSI(
  ArchRProj = proj_filt,
  useMatrix = "TileMatrix",
  name = "IterativeLSI",
  iterations = 2,
  clusterParams = list(
    resolution = c(0.2),
    sampleCells = 10000,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30
)

proj_filt <- addClusters(
  input = proj_filt,
  reducedDims = "IterativeLSI",
  method = "Seurat",
  name = "Clusters",
  resolution = 0.6
)

proj_filt <- addUMAP(
  ArchRProj = proj_filt,
  reducedDims = "IterativeLSI",
  name = "UMAP",
  nNeighbors = 30,
  minDist = 0.5,
  metric = "cosine"
)

# Gene activity-like matrix for integration
proj_filt <- addGeneScoreMatrix(proj_filt,force = TRUE)
```

### ___Label transfer___

```{r}
# LOAD SAVED FILE
proj <- loadArchRProject(path  = "mnt_out/ArchR_filt/")
proj_filt <- readRDS( "mnt_out/ArchR_filt/Save-ArchR-Project.rds")
```


```{r}
library(SingleCellExperiment)
library(Seurat)

sceRNA <- as.SingleCellExperiment(NK_RNA)

proj_filt <- addGeneIntegrationMatrix(
  ArchRProj = proj_filt,
  useMatrix = "GeneScoreMatrix",
  matrixName = "GeneIntegrationMatrix_v3",
  reducedDims = "IterativeLSI",
  seRNA = sceRNA,
  addToArrow = TRUE,   # add to arrow file
  groupRNA = "predicted_NK_type_de.n350", 
  nameCell = "predicted_cell", # predicted cell ID
  nameGroup = "predicted_NK_subtype", # predicted label
  nameScore = "prediction_score", # prediction score - higher score, higher confidence
  dimsToUse = 1:30,
  force = TRUE,
  groupATAC = "group"
)
```

### ___Label transfer QC___

```{r}
p1 <- plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "cellColData",
  name = "predicted_NK_subtype",
  embedding = "UMAP",
  baseSize = 14,
  size = 2,
  plotAs = "points"
)

p2 <- plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "cellColData",
  name = "prediction_score",
  embedding = "UMAP",
  baseSize = 14,
  size = 2,
  plotAs = "points"
)


p1
p2
```

```{r}
table(proj_filt$predicted_NK_subtype)

proj_filt$predicted_NK_subtype <- as.character(proj_filt$predicted_NK_subtype)
proj_filt$predicted_NK_subtype <- as.factor(proj_filt$predicted_NK_subtype)

#validGroups <- setdiff(names(table(proj_filt$predicted_NK_subtype)), "hNK_Bm4")

getFeatures(
  ArchRProj = proj_filt,
  useMatrix = "GeneScoreMatrix",
  select = NULL,
  ignoreCase = TRUE
)

plotEmbedding(ArchRProj = proj_filt, 
              colorBy = "cellColData", 
              name = "Clusters", 
              embedding = "UMAP",
              baseSize = 14,
              size = 2,
              plotAs = "points")

plotEmbedding(ArchRProj = proj_filt, 
              colorBy = "cellColData", 
              name = "group", 
              embedding = "UMAP",  
              baseSize = 14,
              size = 2,
              plotAs = "points")
```

```{r}
# BARPLOT
predict_NK_per_clust <- table(proj_filt$Clusters,proj_filt$predicted_NK_subtype)

# Convert to proportions (per cluster/group)
predict_NK_per_clust_prop <- prop.table(predict_NK_per_clust, margin = 1) * 100 

input <- reshape2::melt(predict_NK_per_clust_prop)

# Create the plot
ggplot(input, aes(x = Var1, y = value, fill = Var2)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells", 
       fill = "NK subsets") +
  ggtitle("") +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black"),                        
        axis.title.y = element_text(size = 16,
                                    color = "black", angle = 90),                 
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                             color = "black")) +  
  scale_fill_manual(values = COLOR_CODE_v2)
```

  C1  =  hNK_Bm2_hNK_Bm3 [transitioning]
  C2  =  hNK_Bm3
  C3  =  hNK_Bm3
  C4  =  hNK_Bm3
  C5  =  hNK_Bm1     
  C6  =  hNK_Bm3
  
```{r}
# BARPLOT
predict_NK_per_group <- table(proj_filt$group,proj_filt$Clusters)

# Convert to proportions (per cluster/group)
predict_NK_per_group_prop <- prop.table(predict_NK_per_group, margin = 1) * 100 

input <- reshape2::melt(predict_NK_per_group_prop)

# Create the plot
ggplot(input, aes(x = Var1, y = value, fill = Var2)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_void() + 
  labs(x = "", y = "Percentage of NK cells", 
       fill = "NK subsets") +
  ggtitle("") +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0, 
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black"),                        
        axis.title.y = element_text(size = 16,
                                    color = "black", angle = 90),                 
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                             color = "black")) 
```


```{r}
head(names(proj_filt@cellColData))

proj_filt <- addImputeWeights(proj_filt)

genes_of_interest <- c("ID2","RUNX3","ETS1",
                       "EOMES","TBX21","ZEB2"
                       )

p <- plotEmbedding(
    ArchRProj = proj_filt, 
    colorBy = "GeneScoreMatrix", 
    name = genes_of_interest, 
    embedding = "UMAP",
    baseSize = 14,
    size = 2,
    plotAs = "points",
    imputeWeights = getImputeWeights(proj_filt)
)

pdf("/home/rstudio/mnt_out/figures/snATAC_NKcells_UMAP_TF_geneActivity.pdf")
p
dev.off()


# Add Tile Matrix
proj_filt <- addTileMatrix(proj_filt, 
                           force = TRUE, 
                           tileSize = 1000)

proj_filt <- getFeatures(proj_filt, 
            select = genes_of_interest, 
            useMatrix = "TileMatrix")


a <- plotEmbedding(
    ArchRProj = proj_filt, 
    colorBy = "TileMatrix", 
    name = genes_of_interest, 
    embedding = "UMAP",
    baseSize = 14,
    size = 2,
    plotAs = "points",
    imputeWeights = getImputeWeights(proj_filt)
)

a
```
chr     start      end        name    strand
chr2    8678845    8684461    ID2     +
chr1    24899511   24965121   RUNX3   -
chr11   128328656  128457501  ETS1    +
chr3    27715949   27722711   EOMES   -
chr17   47733236   47746126   TBX21   +
chr2    144364364  144521057  ZEB2    -


### ___Add metadata___

```{r}
# ArchR cell IDs
archr_cells <- proj_filt$cellNames

# Seurat cell IDs
seurat_cells <- colnames(NK_ATAC)


# Normalize to base barcode sequence:
# - remove ArchR "prefix#"
# - remove trailing "-<digits>" (e.g., -1, -7)
norm_bc <- function(x) {
  x <- sub("^.*#", "", x)       # drop prefix up to '#'
  x <- sub("-\\d+$", "", x)     # drop trailing -<number>
  x
}

archr_bc  <- norm_bc(archr_cells)
seurat_bc <- norm_bc(seurat_cells)

# Sanity: overlap rate
overlap <- length(intersect(archr_bc, seurat_bc))
cat("Overlap:", overlap, "of", length(archr_bc), "ArchR cells\n")

# Create Seurat lookup table for metadata you want to transfer
lut <- data.frame(
  bc = seurat_bc,
  sample = NK_ATAC$sample,
  group  = NK_ATAC$group,
  stringsAsFactors = FALSE
)

# Critical: ensure Seurat base barcodes are unique, otherwise mapping is ambiguous
dup <- duplicated(lut$bc)
if (any(dup)) {
  # show how many duplicates (often indicates barcode reuse across multiple samples)
  cat("WARNING: duplicated Seurat base barcodes =", sum(dup), "\n")
  print(head(lut[lut$bc %in% lut$bc[dup], ], 10))
  stop("Mapping is ambiguous because Seurat base barcodes are not unique. See notes below for resolution.")
}

# Map Seurat metadata onto ArchR cells
m <- match(archr_bc, lut$bc)

# How many ArchR cells did not map?
cat("Unmapped ArchR cells:", sum(is.na(m)), "\n")

proj_filt$sample <- lut$sample[m]
proj_filt$group  <- lut$group[m]

# QC check
table(proj_filt$sample, useNA = "ifany")
table(proj_filt$group,  useNA = "ifany")
```

```{r}
tab <- table(proj_filt$group, proj_filt$predicted_NK_subtype)
tab
tab_prop <- prop.table(tab, margin = 1) 

input <- reshape2::melt(tab_prop)

barplot(
  tab_prop,
  beside = FALSE,
  las = 2,
  xlab = "Fraction of NK cells",
  ylab = "Group",
  legend.text = TRUE,
  args.legend = list(x = "topright", bty = "n", cex = 0.9)
)

COLOR_CODE_GROUP <- c("control"   = "#7EA3AC",
                      "ATRXmut"   = "#4f4e92",
                      "MYCNamp"       = "#A7563C",
                      "ATRXwtMYCNwt"  = "#FAA21B")

ggplot(input, aes(x = Var1, y = value, fill = Var2)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_void() +
  labs(x = "", y = "Percentage of NK cells",
       fill = "NK subsets") +  
  ggtitle("snATAC-seq of BM metastasis samples") +
  theme(axis.text.x = element_text(angle = 45,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",
                                    angle = 45),
        axis.title.y = element_text(size = 16,
                                    color = "black", 
                                    angle = 90),
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                             color = "black")) +
  scale_fill_manual(values = COLOR_CODE_v2)   


```

```{r}
# INSTALL MACS2 with PIP not CONDA ?!

# Make pseudo-bulk coverages (often by cell state or sample)
proj_filt <- addGroupCoverages(
  ArchRProj = proj_filt,
  groupBy = "group"   # or "predicted_NK_subtype"  for NK subsets
)

#addArchRThreads(threads = 1) # set to run MACS2 on a single core ?!
#options(mc.cores = 1)

# Call peaks on those pseudo-bulks
proj_filt <- addReproduciblePeakSet(
  ArchRProj = proj_filt,
  peakMethod = "Tiles", # using tiles instead of MACS2
  method = "p",
  groupBy = "group",  # same as above
  #pathToMacs2 = macs2 
)
```



```{r}
celltypes <- levels(proj_filt$predicted_NK_subtype)

# differentially accesible regions
marker_list <- lapply(celltypes, function(ct) {
  cells_ct <- proj_filt$cellNames[proj_filt$predicted_NK_subtype == ct]
  proj_ct <- subsetArchRProject(
    ArchRProj = proj_filt,
    cells = cells_ct,
    outputDirectory = paste0("ArchR_", ct),
    dropCells = TRUE
  )

  getMarkerFeatures(
    ArchRProj = proj_ct,
    useMatrix = "TileMatrix",            # or "GeneScoreMatrix" for RNA or "TileMatrix" if MACS2  isnot available
    groupBy = "group",                   # your condition (e.g., control vs ATRXmut)
   # bias = c("TSSEnrichment", "log10(nFrags)")
  )
})

names(marker_list) <- celltypes

# differential gene scores
seMarker <- getMarkerFeatures(
  ArchRProj = proj_ct,
  useMatrix = "GeneScoreMatrix", 
  groupBy = "group",
  bias = c("TSSEnrichment", "log10(nFrags)")
)
```


### ___save Rds___

```{r}
# SAVE
saveArchRProject(ArchRProj = proj_filt, 
                 outputDirectory = paste0(params$out,"/ArchR_filt/"), 
                 load = FALSE)


saveRDS(proj_filt, file = paste0(params$out,"/ArchR_filt/ATAC_analysis_20260115.Rds"))
```

```{r}
save.image("~/archR_analysis_20260115.RData")
```


# ___The End___

```{r}
sessionInfo()
```

