---
title: "NK project: snATAC-seq label transfer"
author: "Dr. Sara Wernig-Zorc"
date: "2026-01-26 - 2026-02-04"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  out: "/home/rstudio/mnt_out/"
  anno: "/home/rstudio/mnt_resources/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```

# ___Project specific setup___

```{r load project specific R packages and color code, echo = FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
library("ArchR")
set.seed(2805)

conda_path <- "/opt/conda/envs/macs2" # conda env to use
macs2_path <- "/opt/conda/envs/macs2/bin/macs2" # path to MACS2 tool
system2(macs2_path, args = "--version")

library(reticulate)
use_condaenv(conda_path, required = TRUE)

library("BSgenome.Hsapiens.UCSC.hg38")
```


### ___Load ATAC data___

```{r}
# LOAD SAVED FILE
#proj_filt <- loadArchRProject(path  = "mnt_out/ArchR_filt/")
proj_filt <- readRDS( "mnt_out/ArchR_filt/Save-ArchR-Project.rds")
```

### ___Load RNA data___

```{r}
NK_RNA <- readRDS(paste0(params$out, "/Rds/ALL_NK_cluster_withNewMetadata.Rds"))

NK_RNA <- NormalizeData(NK_RNA)
NK_RNA <- FindVariableFeatures(NK_RNA, nfeatures = 5000)
NK_RNA <- ScaleData(NK_RNA)

DefaultAssay(NK_RNA) <- "RNA"
```

I will test if the label transfer scores improve if i combine NK2+NK3 clusters,
my hypothesis is that the scores are low for those because the two clusters are
very similar to each other, thus combining them would make the algorithm more
confident in the NK subset assignment,

```{r}
# 1. combine NK2+3 clusters
NK_RNA$meta_clusters <- NK_RNA$predicted_NK_type_de.n350
NK_RNA$meta_clusters <- case_when(
  NK_RNA$predicted_NK_type_de.n350 %in% c("hNK_Bm2", "hNK_Bm3") ~ "immuno",
  NK_RNA$predicted_NK_type_de.n350 == "hNK_Bm1" ~ "cyto"
)

NK_RNA$meta_clusters <- factor(NK_RNA$meta_clusters) 

NK_RNA <- 
  NK_RNA %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)

Idents(NK_RNA) <- NK_RNA$meta_clusters
```

# ___NK_ATAC Seurat object: Differential peaks between all NK cells___

```{r one gene, eval=FALSE, include=FALSE}
library(ggpubr)

DefaultAssay(NK_ATAC) <- "GeneActivity"
gene_of_interest <- "PRDM1"

VlnPlot(NK_ATAC, 
        features = gene_of_interest, 
        group.by = "group",
        pt.size = 0.1) +
  stat_compare_means(method = "wilcox.test") +
  ggtitle("PRDM1 Gene Activity in NK cells") +
  theme(plot.title = element_text(hjust = 0.5))

DefaultAssay(NK_ATAC) <- "peaks"

CoveragePlot(peaks = TRUE,
             group.by = "group",
             window = 2000,
             object = NK_ATAC,
             region = gene_of_interest,
             extend.upstream = 10000,
             extend.downstream = 5000
)

# Get gene coordinates
anno <- Annotation(NK_ATAC)
gene_coords <- anno[anno$gene_name == gene_of_interest]

# Extend the region (matching your coverage plot)
region <- GRanges(
  seqnames = seqnames(gene_coords),
  ranges = IRanges(
    start = start(gene_coords) - 10000,
    end = end(gene_coords) + 5000
  )
)

# Find peaks in this region
peaks_in_region <- findOverlaps(StringToGRanges(rownames(NK_ATAC)), region)
peak_names <- rownames(NK_ATAC)[queryHits(peaks_in_region)]

# View peaks found
peak_names

# Extract counts for peaks in region
peak_counts <- GetAssayData(NK_ATAC, 
                            assay = "peaks", 
                            slot = "counts")[peak_names, ]

# Sum counts per cell
total_accessibility <- colSums(peak_counts)

# Add to metadata
NK_ATAC$region_accessibility <- total_accessibility

# Compare between groups
VlnPlot(NK_ATAC,
        features = "region_accessibility",
        group.by = "group",
        pt.size = 0,log = TRUE) +
  stat_compare_means(method = "wilcox.test")
```

# ___ArchR analysis___

## ___ATAC data pre-processing with archR___

```{r}
proj_filt <- addIterativeLSI(
  ArchRProj = proj_filt,
  useMatrix = "TileMatrix",
  name = "IterativeLSI_20260203",
  iterations = 2,
  clusterParams = list(
    resolution = c(0.1,0.2,0.3,0.4,0.5), 
    n.start = 10),
  varFeatures = 25000,
  dimsToUse = 1:30)
# resolution: re-selects variable features based on those clusters to refine the embedding

# Test multiple resolutions
for (res in c(0.2, 0.4, 0.6, 0.8, 1.0, 1.2)) {
  proj_filt <- addClusters(
    input = proj_filt,
    reducedDims = "IterativeLSI_20260203",
    method = "Seurat",
    name = paste0("Clusters_res", res),
    resolution = res,
    force = TRUE
  )
}

proj_filt <- addUMAP(
  ArchRProj = proj_filt,
  reducedDims = "IterativeLSI_20260203",
  name = "UMAP",
  nNeighbors = 30,
  minDist = 0.5,
  metric = "cosine", 
  force = TRUE
)

# Gene activity-like matrix for integration
proj_filt <- addGeneScoreMatrix(proj_filt,force = TRUE)
```

## ___clustering QC [Choosing best resolution parameter & number of clusters]___

```{r}
# QC Metrics 
# 1. Cluster Stability / Silhouette Score
# Are cells confidently assigned or sitting on boundaries?
# Rough stability check - compare nearby resolutions
library(pheatmap)
confusionMatrix <- confusionMatrix(
  getCellColData(proj_filt)$Clusters_res0.4,
  getCellColData(proj_filt)$Clusters_res1
)
pheatmap(confusionMatrix)

# 2. Cluster Size Distribution
# Watch for very small clusters (could be noise) or one dominant cluster (under-clustering).
# Check cluster sizes

for (res in c(0.2, 0.4, 0.6, 0.8, 1.0, 1.2)) {
  cluster_name <- paste0("Clusters_res", res)
  cluster_data <- getCellColData(proj_filt)[[cluster_name]]
  
  cat("\n--- Resolution:", res, "---\n")
  print(table(cluster_data))
}

# Or visualize
UMAP_final_clust <- plotEmbedding(proj_filt, 
              colorBy = "cellColData", 
              name = "Clusters_res1",
              baseSize = 14,
              size = 2,
              plotAs = "points")

UMAP_final_clust

# 3. Technical vs Biological Signal
# Clusters shouldn't correlate strongly with technical artifacts.
# Check for batch effects
plotEmbedding(proj_filt, colorBy = "cellColData", name = "group")

# QC metrics per cluster
plotGroups(
  ArchRProj = proj_filt,
  groupBy = "Clusters_res1",
  colorBy = "cellColData",
  name = "TSSEnrichment"  # or nFrags, DoubletScore, etc.
)

plotGroups(
  ArchRProj = proj_filt,
  groupBy = "Clusters_res1",
  colorBy = "cellColData",
  name = "nFrags"  # or nFrags, DoubletScore, etc.
)


# Red flags:
# Clusters driven by sample/batch rather than biology
# Clusters with very low TSS enrichment (poor quality cells)
# Clusters with high doublet scores

# 4. Marker Gene/Peak Accessibility
# Do clusters have biologically meaningful differences?
# Find marker genes
markersGS <- getMarkerFeatures(
  ArchRProj = proj_filt,
  useMatrix = "GeneScoreMatrix",
  groupBy = "Clusters_res1",
  bias = c("TSSEnrichment", "log10(nFrags)")
)

# Heatmap of top markers
heatmapGS <- plotMarkerHeatmap(
  seMarker = markersGS,
  cutOff = "FDR <= 0.05 & Log2FC >= 0.5",
  transpose = TRUE, 
  #returnMatrix = TRUE, 
  clusterCols = TRUE,
  scaleRows = TRUE,
  plotLog2FC = TRUE,
  log2Norm = TRUE,
  nLabel = 5,
  pal = paletteContinuous(set = "greyMagma", n = 265, reverse = FALSE)
)

heatmapGS

pdf(paste0(params$out,"/figures/snATAC_clusterMarkers_basedOnATACdata.pdf"),
           height = 6,width = 7)
heatmapGS
dev.off()

# 5. Known Marker Genes
# Check if expected cell types appear in distinct clusters.
# Plot known markers

plotEmbedding(proj_filt, 
              colorBy = "cellColData", 
              name = "Clusters_res1",
              )

df <- getCellColData(proj_filt, select = c("log10(nFrags)", "TSSEnrichment"))
df

p <- ggPoint(
    x = df[,1], 
    y = df[,2], 
    colorDensity = TRUE,
    continuousSet = "sambaNight",
    xlabel = "Log10 Unique Fragments",
    ylabel = "TSS Enrichment",
    xlim = c(log10(500), quantile(df[,1], probs = 0.99)),
    ylim = c(0, quantile(df[,2], probs = 0.99))) + 
  geom_hline(yintercept = 4, lty = "dashed") + 
  geom_vline(xintercept = 3, lty = "dashed")

p

getCellColData(proj_filt)
```

## ___Asssign cluster labels___

```{r}
# Get cluster and group information
cellData <- getCellColData(proj_filt)
clusters <- cellData$Clusters_res1
groups <- cellData$predicted_NK_subtype  

# Create a table of cluster vs group proportions
cluster_group_table <- table(clusters, groups)
cluster_props <- prop.table(cluster_group_table, margin = 1)  # row proportions

# Function to assign majority label with 60% threshold
assign_majority_label <- function(props, threshold = 0.6) {
  max_prop <- max(props)
  if (max_prop > threshold) {
    return(names(props)[which.max(props)])
  } else {
    return("Mixed")
  }
}

# Apply to each cluster
majority_labels <- apply(cluster_props, 1, assign_majority_label, threshold = 0.6)

# View the assignments
print(data.frame(
  Cluster = names(majority_labels),
  Majority_Label = majority_labels,
  Max_Proportion = apply(cluster_props, 1, max)
))

# Create a mapping from cluster to majority label
cluster_to_label <- setNames(majority_labels, names(majority_labels))

# Add to ArchR project
proj_filt$majority_vote_label <- cluster_to_label[as.character(proj_filt$Clusters_res1)]

# Verify
table(proj_filt$Clusters_res1, proj_filt$majority_vote_label)

# Plot UMAP with new labels
p <- plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "cellColData",
  name = "majority_vote_label",
  embedding = "UMAP"
)
p
```


## ___TF occupancy___

```{r}
proj_filt <- addImputeWeights(proj_filt,
                              reducedDims = "IterativeLSI_20260203")

ATAC_features <- getFeatures(proj_filt)

# Multiple TFs at once
genes_of_interest <- c("ETS-1", "PU.1", "Ikaros", "TOX", "E4BP4", "ID2", "GATA-3", 
                       "IRF-2",   "MEF", "MITF" , "TBX21", "Eomes", "PRDM1")

genes_for_plotting <- intersect(ATAC_features, genes_of_interest)

inferred_gene_activity <- plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "GeneScoreMatrix",
  name = genes_for_plotting,
  embedding = "UMAP",
  quantCut = c(0.01, 0.99),
  imputeWeights = getImputeWeights(proj_filt), 
  baseSize = 14,
  size = 2,
  plotAs = "points"
)

inferred_gene_activity

combined_RNA_ATAC_gene_activity <- plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "GeneIntegrationMatrix_v3",
  name = genes_for_plotting,
  embedding = "UMAP",
  quantCut = c(0.01, 0.99),
  imputeWeights = getImputeWeights(proj_filt), 
  baseSize = 14,
  size = 2,
  plotAs = "points"
)

combined_RNA_ATAC_gene_activity

pdf(paste0(params$out,"figures/ATAC-NK_TF_gene_activity_wGeneScoreMatrix.pdf"),
    height = 5,width = 7)
inferred_gene_activity
dev.off()

pdf(paste0(params$out,"figures/ATAC-NK_TF_gene_activity_wGeneIntegrationMatrix.pdf"),
    height = 5,width = 7)
combined_RNA_ATAC_gene_activity
dev.off()
```


## ___archR label transfer with 2 NK clusters, cytotoxic and immuno___

```{r}
library(SingleCellExperiment)
library(Seurat)

sceRNA <- as.SingleCellExperiment(NK_RNA)

proj_filt <- addGeneIntegrationMatrix(
  ArchRProj = proj_filt,
  useMatrix = "GeneScoreMatrix",
  matrixName = "GeneIntegrationMatrix_v3",
  reducedDims = "IterativeLSI_20260203",
  seRNA = sceRNA,
  addToArrow = TRUE,   # add to arrow file
  groupRNA = "meta_clusters",  # label transfer with 2 NK clusters, cytotoxic and immuno
  nameCell = "predicted_cell", # predicted cell ID
  nameGroup = "predicted_NK_group", # predicted label {using here two NK clusters}
  nameScore = "prediction_score", # prediction score - higher score, higher confidence
  dimsToUse = 1:30,
  force = TRUE,
  groupATAC = "group"
)
```


# ---------2026.02.04---------

```{r eval=FALSE, include=FALSE}
proj_filt <- readRDS( "mnt_out/ArchR_v20260203/Save-ArchR-Project.rds")
```

```{r}
# Maturation markers
maturation_markers <- c("SELL", "TCF7", "LEF1",      # less mature / CD56bright / immunomodulatory NK
                        "FCGR3A", "CX3CR1", "ZEB2")  # more mature / CD56dim / cytotoxic NK 

# Cytotoxic vs immunomodulatory
functional_markers <- c("PRF1", "GZMB", "GZMK", "IFNG", "TNF", "XCL1")

# Compare matrices
p_gs <- plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "GeneScoreMatrix",
  name = "PRF1",
  baseSize = 14,
  size = 2,
  plotAs = "points",
  imputeWeights = getImputeWeights(proj_filt)
) + ggtitle("PRF1 - Accessibility\n (may look similar across subsets)")

p_gi <- plotEmbedding(ArchRProj = proj_filt,
  colorBy = "GeneIntegrationMatrix_v3",
  name = "PRF1",
  baseSize = 14,
  size = 2,
  plotAs = "points",
) + ggtitle("PRF1 - Expression\n (should distinguish CD56dim)")

p_gs + p_gi

# TCF7/LEF1 - should be high in less mature
# ZEB2 - should be high in more mature (represses TCF7)

# GeneScoreMatrix problem:
# TCF7 locus may be accessible in all NK (poised)
# Only GeneIntegrationMatrix shows actual TCF7 expression gradient

markers_maturation <- c("TCF7", "LEF1", "ZEB2", "CX3CR1", "KLRG1")

# Loop through both matrices
for (m in markers_maturation) {
  p1 <- plotEmbedding(proj_filt, colorBy = "GeneScoreMatrix", name = m,   
                      baseSize = 14, size = 2, plotAs = "points") + 
        ggtitle(paste(m, "- Accessibility"))
  p2 <- plotEmbedding(proj_filt, colorBy = "GeneIntegrationMatrix_v3", name = m,
                      baseSize = 14, size = 2, plotAs = "points") + 
        ggtitle(paste(m, "- Expression"))
  print(p1 + p2)
}

# GZMB/PRF1 - cytotoxic
# GZMK/XCL1 - more immunomodulatory / CD56bright

# The issue:
# Both cytotoxic and immunomod NK have accessible PRF1/GZMB loci
# But only cytotoxic subset actively transcribes them

markers_function <- c("PRF1", "GZMB", "GZMK", "XCL1", "AREG")
```

```{r}
p2g <- getPeak2GeneLinks(proj_filt,
  corCutOff = 0.45,
  FDRCutOff = 1e-04,
  varCutOffATAC = 0.25, # increase cutoff?
  varCutOffRNA = 0.25, # increase cutoff?
  resolution = 1,
  returnLoops = TRUE
)

# Plot tracks at key loci grouped by cluster
plotBrowserTrack(ArchRProj = proj_filt,
  groupBy = "Clusters_res1",
  geneSymbol = "TCF7",
  upstream = 50000,
  downstream = 50000,
  features = getPeakSet(proj_filt),
  loops = getCoAccessibility(proj_filt),
  baseSize = 12,
  borderWidth = 0.4,
  tickWidth = 0.4,
  facetbaseSize = 10,
  geneAnnotation = getGeneAnnotation(proj_filt),
  title = "")



# Compare multiple loci
loci_to_check <- c("ZEB2", "SELL", "FCGR3A", "PRF1", "GZMB", "GZMK", "KLRC2")

for (gene in loci_to_check) {
  p <- plotBrowserTrack(proj_filt,
    groupBy = "Clusters_res1",
    geneSymbol = gene,
    upstream = 50000,
    downstream = 50000,
    features = getPeakSet(proj_filt),
    loops = getCoAccessibility(proj_filt),
    baseSize = 12,
    borderWidth = 0.4,
  tickWidth = 0.4,
  facetbaseSize = 10,
  geneAnnotation = getGeneAnnotation(proj_filt),
  title = ""
  )
  p
}
```

# ___Label transfer QC___

```{r}
p1 <- plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "cellColData",
  name = "predicted_NK_group",
  embedding = "UMAP",
  baseSize = 14,
  size = 2,
  plotAs = "points"
)

p2 <- plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "cellColData",
  name = "prediction_score",
  embedding = "UMAP",
  baseSize = 14,
  size = 2,
  plotAs = "points"
)


p2 + p1

pdf(paste0(params$out,"figures/2026-02_snATAC-seq_label_transfer_NK2+NK3_combined.pdf"),
    width = 10,height = 5)
p2 + p1
dev.off()

```

# ___NK minor-subset [hNK_Bm1, hNK_Bm2, hNK_Bm3] barplot per group___

```{r barplot minor NK subtypes}
tab <- table(proj_filt$predicted_NK_subtype,
             proj_filt$group)
tab_prop <- prop.table(tab, margin = 2) 
tab_prop

input <- reshape2::melt(tab_prop)

prop_plot <- ggplot(input, aes(x = Var2, y = value, fill = Var1)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_void() +
  labs(x = "", y = "Percentage of NK cells",
       fill = "NK subsets") +  
  ggtitle("snATAC-seq of BM metastasis samples") +
  theme(axis.text.x = element_text(angle = 45,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",
                                    angle = 45),
        axis.title.y = element_text(size = 16,
                                    color = "black", 
                                    angle = 90),
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                             color = "black")) +
  scale_fill_manual(values = COLOR_CODE_v2)   

prop_plot 
```

```{r}
plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "cellColData",
  name = "predicted_NK_subtype",
  embedding = "UMAP",
  baseSize = 14,
  size = 2,
  plotAs = "points"
)

tab <- table(proj_filt$predicted_NK_subtype,proj_filt$Clusters_res1)

tab_prop <- prop.table(tab, margin = 2) 
tab_prop

input <- reshape2::melt(tab_prop)

prop_plot <- ggplot(input, aes(x = Var2, y = value, fill = Var1)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_void() +
  labs(x = "", y = "Percentage of NK cells",
       fill = "NK subsets") +  
  ggtitle("snATAC-seq of BM metastasis samples") +
  theme(axis.text.x = element_text(angle = 45,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",
                                    angle = 45),
        axis.title.y = element_text(size = 16,
                                    color = "black", 
                                    angle = 90),
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                             color = "black")) +
  scale_fill_manual(values = COLOR_CODE_v2)   

prop_plot 

```


# ___NK major-subset [cyto, immuno] barplot per group___

```{r barplot major NK subtypes}
tab <- table(proj_filt$predicted_NK_group,
             proj_filt$group)
tab_prop <- prop.table(tab, margin = 2) 
tab_prop

input <- reshape2::melt(tab_prop)

COLOR_CODE_NK_TYPES <- c("immuno"   = "#7EA3AC",
                         "cyto"   = "#4f4e92")

prop_plot <- ggplot(input, aes(x = Var2, y = value, fill = Var1)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_void() +
  labs(x = "", y = "Percentage of NK cells",
       fill = "NK subsets") +  
  ggtitle("snATAC-seq of BM metastasis samples") +
  theme(axis.text.x = element_text(angle = 45,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",
                                    angle = 45),
        axis.title.y = element_text(size = 16,
                                    color = "black", 
                                    angle = 90),
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                             color = "black")) +
  scale_fill_manual(values = COLOR_CODE_NK_TYPES)   

prop_plot 
```

```{r}
plotEmbedding(
  ArchRProj = proj_filt,
  colorBy = "cellColData",
  name = "Clusters_res1",
  embedding = "UMAP",
  baseSize = 14,
  size = 2,
  plotAs = "points"
)

tab <- table(proj_filt$predicted_NK_group,proj_filt$Clusters_res1)

tab_prop <- prop.table(tab, margin = 2) 
tab_prop

input <- reshape2::melt(tab_prop)

COLOR_CODE_NK_TYPES <- c("immuno"   = "#7EA3AC",
                         "cyto"   = "#4f4e92")

prop_plot <- ggplot(input, aes(x = Var2, y = value, fill = Var1)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_void() +
  labs(x = "", y = "Percentage of NK cells",
       fill = "NK subsets") +  
  ggtitle("snATAC-seq of BM metastasis samples") +
  theme(axis.text.x = element_text(angle = 45,
                                   color = "black",
                                   size = 16),
        axis.text.y = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 16),
        legend.text = element_text(angle = 0,
                                   hjust = 1,
                                   color = "black",
                                   size = 18),
        axis.title.x = element_text(size = 16,
                                    color = "black",
                                    angle = 45),
        axis.title.y = element_text(size = 16,
                                    color = "black", 
                                    angle = 90),
        legend.title = element_text(size = 16,
                                    color = "black"),
        title = element_text(size = 20,
                             color = "black")) +
  scale_fill_manual(values = COLOR_CODE_NK_TYPES)   

prop_plot 

```



# ___RNA+ATAC network analysis___

```{r}
# Add group coverages
proj_filt <- addGroupCoverages(
  ArchRProj = proj_filt,
  groupBy = "Clusters_res1", 
  minCells = 40,
  force = TRUE
)

# Call peaks using MACS2
proj_filt <- addReproduciblePeakSet(
  ArchRProj = proj_filt,
  groupBy = "Clusters_res1",
  pathToMacs2 = macs2_path
)

proj_filt <- addPeakMatrix(proj_filt)
getAvailableMatrices(proj_filt)
getPeakSet(proj_filt)


# run Peak2GeneLinks
proj_filt <- addPeak2GeneLinks(
  ArchRProj = proj_filt,
  reducedDims = "IterativeLSI_20260203",
  useMatrix = "GeneIntegrationMatrix_v3",
  maxDist = 50000,        
  corCutOff = 0.45     # increase correlation?
)

# retrieve Peak2GeneLinks above thresholds
p2g <- getPeak2GeneLinks(
  ArchRProj = proj_filt,
  corCutOff = 0.45,
  FDRCutOff = 0.01
)

```

```{r}
gene_of_interest <- "CD6"

# Visualize at a genomic locus
track <- plotBrowserTrack(
  ArchRProj = proj_filt,
  groupBy = "Clusters_res1",
  geneSymbol = gene_of_interest,
  upstream = 100000,
  downstream = 100000,
  loops = p2g
)

grid::grid.newpage()
grid::grid.draw(track$CD6) # change depending on the gene

# Plot peak-to-gene heatmap
clustering_data <- plotPeak2GeneHeatmap(proj_filt, 
                          groupBy = "Clusters_res1",
                          seed = 2805,
                          returnMatrices = TRUE)

p <- plotPeak2GeneHeatmap(proj_filt, 
                          groupBy = "Clusters_res1",
                          seed = 2805)

p

pdf(paste0(params$out,"figures/snATAC_p2g_heatmap.pdf"),
    height = 8, width = 10)
p
dev.off()

# per Group
# c("group", "Clusters_res1", "predicted_NK_subtype", "predicted_NK_group")
perGroup_p2g <- plotPeak2GeneHeatmap(proj_filt, 
                          groupBy = "group",
                          seed = 2805)

perGroup_p2g

pdf(paste0(params$out,"figures/snATAC_p2g_perGroup_heatmap.pdf"),
    height = 10, width = 10)
perGroup_p2g
dev.off()
```

```{r}
# Explore the structure
str(clustering_data)
names(clustering_data)

# Extract k-means cluster assignments
kmeans_clusters <- clustering_data$KmeansAssignment

# This gives you which module (1-25) each peak-to-gene link belongs to
head(kmeans_clusters)

# Extract k-means cluster assignments (module IDs)
kmeans_clusters <- clustering_data$ATAC$kmeansId

# Get the peak-to-gene link information
p2g_df <- as.data.frame(clustering_data$Peak2GeneLinks)

# Add module assignments
p2g_df$module <- kmeans_clusters

# View the data
head(p2g_df)

# Check number of links per module
table(p2g_df$module)

# Get unique genes per module
genes_by_module <- split(p2g_df$gene, p2g_df$module)
module_gene_list <- lapply(genes_by_module, unique)

# Count genes per module
sapply(module_gene_list, length)

# View top genes in each module
lapply(module_gene_list, head, 10)

# Create summary table
module_summary <- do.call(rbind, lapply(sort(unique(kmeans_clusters)), function(i) {
  sub_df <- p2g_df[p2g_df$module == i, ]
  genes <- unique(sub_df$gene)
  data.frame(
    module = i,
    n_links = nrow(sub_df),
    n_genes = length(genes),
    mean_correlation = mean(sub_df$Correlation),
    top_genes = paste(head(genes, 5), collapse = ", ")
  )
}))

print(module_summary)

# Save genes by module for downstream analysis (GO enrichment, etc.)
for (mod in sort(unique(kmeans_clusters))) {
  genes <- unique(p2g_df$gene[p2g_df$module == mod])
  write.csv(
    data.frame(gene = genes),
    paste0(params$out,"tables/module_", mod, "_genes.csv"),
    row.names = FALSE
  )
}

# Calculate Module Activity Per Cluster
# Extract matrices
atac_mat <- clustering_data$ATAC$matrix
rna_mat <- clustering_data$RNA$matrix
col_data <- as.data.frame(clustering_data$ATAC$colData)

# Calculate mean z-score per module per cluster
clusters <- unique(col_data$groupBy)

module_activity <- sapply(sort(unique(kmeans_clusters)), function(mod) {
  # Get rows belonging to this module
  mod_idx <- which(kmeans_clusters == mod)
  
  # Calculate mean per cluster
  sapply(clusters, function(cl) {
    cell_idx <- which(col_data$groupBy == cl)
    mean(rna_mat[mod_idx, cell_idx])
  })
})

colnames(module_activity) <- paste0("M", sort(unique(kmeans_clusters)))
rownames(module_activity) <- clusters

# Heatmap of module activity by cluster
library(pheatmap)

moduleScores_ByCluster <- pheatmap(
  t(module_activity),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("#592e83", "white", "#80aea0"))(100),
  main = "Module Activity by Cluster (RNA Z-scores)"
)

moduleScores_ByCluster

pdf(paste0(params$out,"figures/snATAC_moduleScores_ByCluster.pdf"),
    height = 8, width = 10)
moduleScores_ByCluster
dev.off()

# Identify cluster-enriched modules
# Find modules where C6 (hNK_Bm1) has highest activity
C6_enriched <- which(apply(module_activity, 2, which.max) == which(clusters == "C6"))
print(paste("Modules enriched in C6 (hNK_Bm1):", paste(names(C6_enriched), collapse = ", ")))

# Get genes from C6-enriched modules
hNK_Bm1_genes <- unique(p2g_df$gene[p2g_df$module %in% as.numeric(gsub("M", "", names(C6_enriched)))])
print(paste("Number of hNK_Bm1-enriched genes:", length(hNK_Bm1_genes)))
head(hNK_Bm1_genes, 20)
```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# Run GO enrichment for each module
module_GO_results <- lapply(sort(unique(kmeans_clusters)), function(mod) {
  genes <- unique(p2g_df$gene[p2g_df$module == mod])
  
  if (length(genes) < 10) return(NULL)
  
  tryCatch({
    enrichGO(
      gene = genes,
      universe = 
      OrgDb = org.Hs.eg.db,
      keyType = "SYMBOL",
      ont = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.1
    )
  }, error = function(e) NULL)
})

names(module_GO_results) <- paste0("Module_", sort(unique(kmeans_clusters)))

# View results for a specific module
module_GO_results$Module_7

# Plot top terms
dotplot(module_GO_results$Module_7, 
        showCategory = 10, 
        title = "Module 7 GO Enrichment")

# Combine top terms from all modules into summary
GO_summary <- do.call(rbind, lapply(names(module_GO_results), function(mod) {
  res <- module_GO_results[[mod]]
  if (!is.null(res) && nrow(as.data.frame(res)) > 0) {
    df <- head(as.data.frame(res), 3)
    df$module <- mod
    df[, c("module", "Description", "pvalue", "p.adjust", "Count")]
  }
}))

print(GO_summary)
```


```{r}

```


```{r}
# 3. Look at specific genes in modules of interest
# For example, genes high in C6 (hNK_Bm1-specific):

markerGenes <- getMarkerFeatures(
  ArchRProj = proj_filt,
  useMatrix = "GeneIntegrationMatrix_v3",
  groupBy = "Clusters_res1",
  useGroups = "C6",
  bgdGroups = c("C1", "C2", "C3", "C4", "C5", "C7")
)

# 4. Check NK-relevant TF activity differences between C6 vs others
plotEmbedding(proj_filt, 
              colorBy = "MotifMatrix", 
              name = "z:TBX21_789", 
              embedding = "UMAP")
```

# ___Browser output___

```{r}
ArchRBrowser(ArchRProj = proj_filt)
```

```{r}
plotFragmentSizes(ArchRProj = proj_filt)
plotTSSEnrichment(ArchRProj = proj_filt)
```


# ___Save & exit___

```{r}
# Save archR project
saveArchRProject(ArchRProj = proj_filt, 
                 outputDirectory = paste0(params$out,"/ArchR_v20260203/"), 
                 load = FALSE)

saveRDS(proj_filt, file = paste0(params$out,"/ArchR_v20260203/ATAC_analysis_v20260203.Rds"))
```

```{r}
save.image(paste0(params$out,"/Rds/archR_analysis_v20260204.RData"))
```

# ___The End___

```{r}
sessionInfo()
```