---
title: "NK project: Paper revision figures"
author: "Dr. Sara Wernig-Zorc"
date: "2026-04-01"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  out: "/home/rstudio/mnt_out/"
  anno: "/home/rstudio/mnt_resources/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```

# ___Project specific setup___

```{r load project specific R packages and color code, echo = FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
library("ArchR")
set.seed(2805)
library("BSgenome.Hsapiens.UCSC.hg38")
```

# ___Load RNA data___

```{r}
fetahu2023 <- readRDS("~/mnt_out/Rds/fetahu2023_withNewMetadata.Rds")

fetahu2023 <- fetahu2023 %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)

unique(fetahu2023$orig.ident)

fetahu2023$celltype  <-  fetahu2023$orig.ident


ALL_NK <- readRDS("~/mnt_out/Rds/ALL_NK_cluster_withNewMetadata.Rds")

ALL_NK <- ALL_NK %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)


unique(ALL_NK$predicted_NK_type_de.n350)

ALL_NB <- readRDS("~/mnt_out/Rds/ALL_NB_cluster_withNewMetadata.Rds")

ALL_NB <- ALL_NB %>% 
  NormalizeData(normalization.method = "LogNormalize",
                scale.factor = 10000)

```

# ___CD6 expression per cell type___

```{r}
do_FeaturePlot(fetahu2023,
            features = "CD6",
            reduction = "UMAP",
            pt.size = 0.05,
            order = TRUE)

VlnPlot(fetahu2023,
         group.by = "celltype",
         split.by = "group",
         features = "CD6",
         log = TRUE,
         layer = "data")

VlnPlot(fetahu2023,
         group.by = "celltype",
         features = "CD6",
         log = TRUE,
         layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_UMAP_v2)

VlnPlot(fetahu2023,
        group.by = "celltype",
        split.by = "group",
        features = ALCAM_ligands,
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP_v1)
```

# ___ALCAM expression per cell type___

```{r fig.height=3, fig.width=10}
do_FeaturePlot(fetahu2023,
            features = "ALCAM",
            reduction = "UMAP",
            pt.size = 0.05,
            order = TRUE)

VlnPlot(fetahu2023,
         group.by = "cell_type",
         split.by = "group",
         features = "ALCAM",
         log = TRUE,
         layer = "data",
        pt.size = 0.01)

VlnPlot(fetahu2023,
         group.by = "cell_type",
         features = "ALCAM",
         log = TRUE,
         layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_UMAP_v2)
```

# ___Other potential L-R pairs___

```{r}
new_candidates <- c("DLL3",  "NOTCH1", 
                    "PRSS3", "F2R",
                    "EPHA4", "FGFR1",
                    "CD6",	"DAG1")

new_candidates <- c("PRSS3", # on NB cells
                    "F2R")   # on NK cells

candidate_plot <- VlnPlot(fetahu2023,
         group.by = "cell_type",
        split.by = "group",
         features = new_candidates,
         log = TRUE,
         layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP_v1)

candidate_plot
```

Explanation/interpretation: Tumor-derived DLL3 activates NOTCH1 on NK cells, and the downstream canonical effector (RBPJ) sits in the same module as your entire chromatin repression complex. The signaling cascade is directly wired into the epigenetic block: DLL3 -> NOTCH1 -> RBPJ (M4) -> HES/HEY targets -> ID2 upregulation -> maturation arrest. Notably, DLL1 is also in M5, raising the possibility of an autocrine/paracrine NK–NK NOTCH loop amplifying the tumor-initiated signal. This is the extrinsic trigger that could initiate or maintain the ATRXdel chromatin silencing axis.

ATRXdel NK cells retain aberrant TGFB1 accessibility (M21/M22 not fully silenced, consistent with maturation failure) and likely produce TGF-β1 via NF-κB-driven transcription (M3/M4 axis) rather than via exhaustion. The PDCD1/TOX2/LAG3 co-accessibility in the same modules is chromatin priming rather than active exhaustion transcription — distinguishing ATRXdel TGFB1 production mechanistically from what would occur in MYCNamp cells. Immature NK cells failing to silence TGFB1 is a direct consequence of the maturation block, and the TGF-β they produce acts back on the tumor and potentially on neighboring NK cells in a way that perpetuates the immunosuppressive microenvironment.

# ___Other ALCAM ligands___

Addressing reviewer question regarding other ALCAM ligands [LILRB4, CD9, L1CAM, 
SOSTDC1, S100B, LGALS1, LGALS3, LGALS8, SDCBP, EZR, ADAM17, MMP14]

```{r fig.height=10, fig.width=10}
ALCAM_ligands <- c("CD6","LILRB4", "CD9", "L1CAM", 
                   "S100B", "LGALS1", "LGALS3", "LGALS8",
                    "SDCBP", "EZR", "ADAM17", "MMP14")

VlnPlot(ALL_NK,
        group.by = "predicted_NK_type_de.n350",
        split.by = "group",
        features = ALCAM_ligands,
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP_v1)
```

# ___Other CD6 receptors___

Addressing reviewer question regarding other CD6 receptors [CDCP1, CD44, GADS, LCP2]

## ___On NK Cells___

```{r  fig.height=3, fig.width=6}
CD6_receptors <- c("CDCP1", "CD44", "LCP2")

VlnPlot(ALL_NK,
        group.by = "predicted_NK_type_de.n350",
        split.by = "group",
        features = CD6_receptors,
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP_v1)
```

## ___On TU cells___

```{r fig.height=6, fig.width=6}
CD6_receptors <- c("ALCAM","CDCP1", "CD44", "LCP2")

VlnPlot(ALL_NB,
        group.by = "group",
        features = CD6_receptors,
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP_v1)
```


# ___CD39 expression___

```{r  fig.height=3, fig.width=6}
# CD39
gene_of_interest <- "ENTPD1"

VlnPlot(ALL_NK,
        group.by = "predicted_NK_type_de.n350",
        split.by = "group",
        features = gene_of_interest,
        log = TRUE,
        layer = "data",
        pt.size = 0,
        cols = COLOR_CODE_GROUP_v1)
```



# ___NK cell gene modules___

I defined 25 gene modules based on integrated ATAC and RNA data. Here I want to
address the function of each modules. 

```{r}
proj_filt <- readRDS(file = paste0(params$out,"/ArchR_v20260203/ATAC_analysis_v20260203.Rds"))

library(ArchR)
library(SummarizedExperiment)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)

set.seed(42)

# Resolve Bioconductor/plyr conflicts with dplyr
filter    <- dplyr::filter
select    <- dplyr::select
mutate    <- dplyr::mutate
summarise <- dplyr::summarise
rename    <- dplyr::rename
count     <- dplyr::count

module_dir   <- "/home/rstudio/mnt_out/tables/NK_module_genes"    # dir containing module_N_genes.csv
cluster_col  <- "Clusters_res1"          # cellColData column for C1–C7 
# THIS IS THE FINAL CLUSTERING I AM USING
group_col    <- "group"                  # cellColData column for ATRXdel etc.
matrix_name  <- "GeneScoreMatrix"       # or "GeneIntegrationMatrix"
n_ctrl_bins  <- 25L                      # expression bins for ctrl sampling
n_ctrl_genes <- 100L                     # ctrl genes per module 
z_cap        <- 2.0                      # cap z-scores for colour scale

group_order   <- c("ATRXdel", "ATRXwtMYCNwt", "control", "MYCNamp")
cluster_order <- paste0("C", 1:7)

module_files <- list.files(module_dir,
                           pattern  = "module_\\d+_genes\\.csv",
                           full.names = TRUE)
module_nums  <- as.integer(gsub(".*module_(\\d+)_genes\\.csv", "\\1",
                                basename(module_files)))
module_files <- module_files[order(module_nums)]

module_genes <- lapply(module_files, function(f) read.csv(f)$gene)
names(module_genes) <- paste0("M", sort(module_nums))
message("Loaded ", length(module_genes), " modules")

all_module_genes <- unique(unlist(module_genes))
se <- getMatrixFromProject(proj_filt)

# Confirm assay name
avail_assays <- assayNames(se)
avail_assays

# Gene names 
gene_names <- rowData(se)$name
if (is.null(gene_names)) gene_names <- rownames(se)
rownames(se) <- gene_names

expr_mat   <- as.matrix(assay(se))

message(sprintf("Matrix: %d genes × %d cells", nrow(expr_mat), ncol(expr_mat)))

# For each module:
#   1. Pool all genes present in matrix
#   2. Bin ALL matrix genes by mean expression rank (25 bins)
#   3. For each expression bin represented by module genes,
#      sample n_ctrl_genes/n_bins control genes from the same bin
#      (excluding module genes themselves)
#   4. score = colMeans(module_genes) − colMeans(ctrl_genes)
#
# This mirrors Seurat::AddModuleScore 

# Bin all genes in full matrix by mean expression rank
full_gene_means <- rowMeans(as.matrix(assay(se)), na.rm = TRUE)
gene_bins <- cut(rank(full_gene_means, ties.method = "average"),
                 breaks = n_ctrl_bins,
                 labels = FALSE)
names(gene_bins) <- names(full_gene_means)

full_expr <- as.matrix(assay(se))   # full matrix for ctrl sampling

score_mat <- matrix(NA_real_,
                    nrow     = ncol(expr_mat),
                    ncol     = length(module_genes),
                    dimnames = list(colnames(expr_mat), names(module_genes)))

for (mod in names(module_genes)) {
  genes_use <- intersect(module_genes[[mod]], rownames(expr_mat))
  n_found   <- length(genes_use)
  if (n_found == 0) {
    warning(mod, ": no genes found in matrix — skipping")
    next
  }

  # Bins occupied by module genes
  mod_bins    <- gene_bins[genes_use]
  bin_table   <- table(mod_bins)

  # Sample ctrl genes proportional to bin representation
  ctrl_genes <- c()
  for (b in names(bin_table)) {
    bin_pool    <- names(gene_bins)[gene_bins == as.integer(b)]
    bin_pool    <- setdiff(bin_pool, genes_use)            # exclude module genes
    n_from_bin  <- as.integer(bin_table[b]) *
                   (n_ctrl_genes / n_found)
    n_sample    <- min(ceiling(n_from_bin), length(bin_pool))
    if (n_sample > 0)
      ctrl_genes <- c(ctrl_genes, sample(bin_pool, n_sample))
  }

  mod_score  <- colMeans(full_expr[genes_use,  , drop = FALSE], na.rm = TRUE)
  ctrl_score <- if (length(ctrl_genes) > 0)
    colMeans(full_expr[ctrl_genes, , drop = FALSE], na.rm = TRUE)
  else
    rep(0, ncol(full_expr))

  score_mat[, mod] <- mod_score - ctrl_score
  message(sprintf("  %s: %d / %d genes scored", mod, n_found,
                  length(module_genes[[mod]])))
}

cell_meta <- as.data.frame(
  getCellColData(proj_filt, select = c(cluster_col, group_col))
)
colnames(cell_meta) <- c("cluster", "group")

# ArchR cell names include "#" (SampleName#Barcode)
score_df <- as.data.frame(score_mat)
score_df$cell    <- rownames(score_df)
cell_meta$cell   <- rownames(cell_meta)
score_df <- left_join(score_df, cell_meta, by = "cell")

stopifnot(sum(!is.na(score_df$cluster)) > 0)

# AGGREGATE: MEAN PER CLUSTER × GROUP
agg <- score_df %>%
  pivot_longer(cols = all_of(names(module_genes)),
               names_to  = "Module",
               values_to = "Score") %>%
  filter(!is.na(cluster), !is.na(group), !is.na(Score)) %>%
  group_by(cluster, group, Module) %>%
  summarise(MeanScore = mean(Score, na.rm = TRUE),
            nCells    = dplyr::n(),
            .groups   = "drop")

# warn if any combination has very few cells
sparse <- filter(agg, nCells < 10)
if (nrow(sparse) > 0) {
  message("Warning: ", nrow(sparse),
          " cluster × group × module combos have <10 cells — interpret with caution!!")
}

# Z-SCORE PER MODULE ACROSS ALL CLUSTER×GROUP COMBOS
agg <- agg %>%
  group_by(Module) %>%
  mutate(ZScore = as.numeric(scale(MeanScore))) %>%
  ungroup() %>%
  mutate(ZScore = pmax(pmin(ZScore, z_cap), -z_cap))   # cap extremes

# FACTOR ORDERING
agg$Module  <- factor(agg$Module,  levels = paste0("M", 1:25))
agg$cluster <- factor(agg$cluster, levels = cluster_order)
agg$group   <- factor(agg$group,   levels = group_order)

# x-axis labels: "C1|ATRXdel", sorted cluster-first, group-second
agg <- agg %>%
  mutate(clust_group = paste0(cluster, "|", group))

x_levels <- agg %>%
  distinct(clust_group, cluster, group) %>%
  arrange(cluster, group) %>%
  pull(clust_group)

agg$clust_group <- factor(agg$clust_group, levels = x_levels)

mat <- agg %>%
  select(Module, clust_group, ZScore) %>%
  pivot_wider(names_from  = clust_group,
              values_from = ZScore) %>%
  column_to_rownames("Module") %>%
  as.matrix()

mat <- mat[paste0("M", 1:25), x_levels, drop = FALSE]

# column anno
col_meta <- agg %>%
  distinct(clust_group, cluster, group) %>%
  arrange(factor(clust_group, levels = x_levels))

ha_top <- HeatmapAnnotation(
  Cluster = col_meta$cluster,
  Group   = col_meta$group,
  col     = list(
    Cluster = COLOR_CODE_perCLUST[levels(col_meta$cluster)],
    Group   = COLOR_CODE_perGROUP[levels(col_meta$group)]
  ),
  annotation_name_side = "left",
  annotation_name_gp   = gpar(fontsize = 8),
  simple_anno_size     = unit(3, "mm"),
  gap                  = unit(1, "mm")
)

# color scale
col_fun <- colorRamp2(
  c(-z_cap, 0, z_cap),
  c("#9B59B6","#FFFFFF","#1E8449")
)

# plot heatmap
# column_split by cluster creates visual separation between the 7 cluster blocks

ht <- Heatmap(
  mat,
  name                  = "RNA Z-score",
  col                   = col_fun,
  top_annotation        = ha_top,

  # Row order: fixed M1→M25 (change to TRUE to cluster modules)
  cluster_rows          = FALSE,
  row_names_side        = "left",
  row_names_gp          = gpar(fontsize = 9),
  row_title             = NULL,

  # Column order: fixed, split by cluster
  cluster_columns       = FALSE,
  column_split          = col_meta$cluster,    # gap between clusters
  column_gap            = unit(2, "mm"),
  column_title_gp       = gpar(fontsize = 9, fontface = "bold"),
  column_names_gp       = gpar(fontsize = 7),
  column_names_rot      = 60,

  # Visuals
  border                = TRUE,
  rect_gp               = gpar(col = "white", lwd = 0.3),

  # Legend
  heatmap_legend_param  = list(
    title          = "Z-score",
    at             = c(-2, -1, 0, 1, 2),
    legend_height  = unit(3, "cm"),
    title_gp       = gpar(fontsize = 9),
    labels_gp      = gpar(fontsize = 8)
  ),

  width  = unit(22, "cm"),
  height = unit(14, "cm")
)

# save
pdf(paste0(params$out,"figures/module_activity_cluster_group.pdf"), 
    width = 16, height = 9)
draw(ht,
     heatmap_legend_side   = "right",
     annotation_legend_side = "right",
     padding = unit(c(5, 5, 5, 5), "mm"))
dev.off()

# export the scored matrix as CSV
write.csv(agg, paste0(params$out,"tables/module_scores_cluster_group_long.csv"), 
                      row.names = FALSE)
```

# ___NK cell cluster annotation based on functional NK cell gene modules___

Barplot with annotated ATAC clusters

```{r}
# rename clusters
cluster_names <- c("C1" = "CD56bright 1",     # Immunomodulatory NK cells
                   "C2" = "CD56bright 2",     # Immunomodulatory NK cells
                   "C3" = "CD56bright 3",     # Immunomodulatory NK cells
                   "C4" = "Intermediate 1",   # Transitional NK cells
                   "C5" = "Intermediate 2",   # Transitional NK cells
                   "C6" = "CD56dim 1",        # Cytotoxic NK cells
                   "C7" = "CD56dim 2")        # Cytotoxic NK cells

proj_filt$Cluster_names <- cluster_names[proj_filt$Clusters_res1]

# metaclusters

# rename clusters
metacluster_names <- c("C1" = "CD56bright",     # Immunomodulatory NK cells
                   "C2" = "CD56bright",     # Immunomodulatory NK cells
                   "C3" = "CD56bright",     # Immunomodulatory NK cells
                   "C4" = "Intermediate",   # Transitional NK cells
                   "C5" = "Intermediate",   # Transitional NK cells
                   "C6" = "CD56dim",        # Cytotoxic NK cells
                   "C7" = "CD56dim")        # Cytotoxic NK cells

proj_filt$Metacluster_names <- metacluster_names[proj_filt$Clusters_res1]


# Rename sample ATRXdel to ATRXdel
proj_filt$group <- ifelse(proj_filt$group == "ATRXdel", "ATRXdel", proj_filt$group)

unique(proj_filt$Cluster_names)
unique(proj_filt$group)
```


## ___Bar plot per cluster___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "Cluster_names")))

cluster_order <- c(
  "CD56bright 1", "CD56bright 2", "CD56bright 3",
  "Intermediate 1", "Intermediate 2",
  "CD56dim 1", "CD56dim 2")

meta$Cluster_names <- factor(meta$Cluster_names, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, Cluster_names, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = Cluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perCLUST_v2, breaks = cluster_order) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1),
    legend.position = "right",
    plot.title     = element_text(face = "bold")
  )

# per sample
df_sample <- meta %>%
  count(sample, Cluster_names, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = Cluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perCLUST_v2, breaks = cluster_order) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1),
    legend.position = "right",
    plot.title     = element_text(face = "bold")
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_cluster_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_cluster_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```

## ___Bar plot per metacluster___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "Metacluster_names")))

cluster_order <- c(
  "CD56bright",
  "Intermediate",
  "CD56dim")

meta$Metacluster_names <- factor(meta$Metacluster_names, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, Metacluster_names, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = Metacluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perMETACLUST, breaks = cluster_order) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1),
    legend.position = "right",
    plot.title     = element_text(face = "bold")
  )

# per sample
df_sample <- meta %>%
  count(sample, Metacluster_names, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = Metacluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perMETACLUST, breaks = cluster_order) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1),
    legend.position = "right",
    plot.title     = element_text(face = "bold")
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_metacluster_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_metacluster_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```


### ___with the label transfer from RNA___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "majority_vote_label")))

cluster_order <- c(
  "hNK_Bm1",
  "hNK_Bm2",
  "hNK_Bm3",
  "hNK_Bm4")

meta$majority_vote_label <- factor(meta$majority_vote_label, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, majority_vote_label, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = majority_vote_label)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_v2, breaks = cluster_order) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1),
    legend.position = "right",
    plot.title     = element_text(face = "bold")
  )

# per sample
df_sample <- meta %>%
  count(sample, majority_vote_label, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = majority_vote_label)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_v2, breaks = cluster_order) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1),
    legend.position = "right",
    plot.title     = element_text(face = "bold")
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_metacluster_labelTransfer_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_metacluster_labelTransfer_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```

```{r}
saveRDS(proj_filt, file = paste0(params$out,"/Rds/snATAC_NK_v20260401.Rds"))
```

# ___Adaptive NK cell signature: ModuleScore___

The question we want to answer here is if we have any adaptive NK cells in the 
ATAC data. When doing the label transfer from RNA data of matched patients,
we saw a clear signature for adaptive NK cells in a small proportion of NK cells,
however in the ATAC data, the label transfer did not identify any adaptive NK cells.

```{r}

```

