---
title: "NK project: Manuscript Main figure 4"
subtitle: "CD6 - ALCAM, other R - L pairs"
author: "Dr. Sara Wernig-Zorc"
date: "2026-07-02"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  out: "/home/rstudio/mnt_out/NK_project/"
  anno: "/home/rstudio/mnt_resources/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, error = FALSE,
                      messages = FALSE, warning = FALSE)
library(parallel)
future::plan("sequential")
options(future.globals.maxSize = 80 * 10^9)
options(Seurat.object.assay.version = "v5")
print(params)
```

# Setup

```{r env, echo=FALSE, include=FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
library(ggplot2)
library(ggpubr)
library(patchwork)
library(dplyr)
set.seed(2805)

fig_out <- paste0(params$out, "figures/revision/")
dir.create(fig_out, showWarnings = FALSE, recursive = TRUE)

NK_SUBSET_ORDER <- c("hNK_Bm1", "hNK_Bm2", "hNK_Bm3", "hNK_Bm4")
ALCAM_NK_GENES  <- c("CD6", "LGALS1", "LGALS8", "S100B","LILRB4", "L1CAM", "ALCAM")
CD6_NB_GENES    <- c("ALCAM",  "CD44", "CDCP1")
CTRL_ATRX       <- list(c("control", "ATRXdel"))

GRADIENT_COLORS <- c(low = "#5B283D", 
                     mid = "white", 
                     high = "#375344")

GRADIENT_COLORS_v2 <- c("#375344", 
                        "#ADCB9C",
                        "white",
                        "#B99FBD",
                        "#5B283D")
```

```{r dotplot function, echo=FALSE, include=FALSE}
fc_dotplot <- function(seurat_obj, genes, group_col = "group", x_col = NULL,
                       ctrl = "control", test = "ATRXdel",
                       fc_limits = c(-2, 2), base_size = 9) {
  vars <- c(genes, group_col, if (!is.null(x_col)) x_col)
  df   <- Seurat::FetchData(seurat_obj, vars = vars, layer = "data")

  grp_vars  <- c(group_col, if (!is.null(x_col)) x_col)
  join_vars <- if (!is.null(x_col)) x_col else character(0)

  summ <- df %>%
    tidyr::pivot_longer(cols = dplyr::all_of(genes),
                        names_to = "gene", values_to = "expr") %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(c("gene", grp_vars)))) %>%
    dplyr::summarise(pct  = mean(expr > 0) * 100,
                     avg  = mean(expr),
                     .groups = "drop")

  ctrl_s <- dplyr::filter(summ, .data[[group_col]] == ctrl)
  test_s <- dplyr::filter(summ, .data[[group_col]] == test)
  by_cols <- c("gene", join_vars)

  fc_df <- dplyr::inner_join(
    dplyr::select(ctrl_s, dplyr::all_of(c(by_cols, "avg"))),
    dplyr::select(test_s, dplyr::all_of(c(by_cols, "avg", "pct"))),
    by = by_cols, suffix = c("_ctrl", "_test")
  ) %>%
    dplyr::mutate(
      log2FC  = log2((avg_test + 0.01) / (avg_ctrl + 0.01)),
      log2FC  = pmax(pmin(log2FC, fc_limits[2]), fc_limits[1])
    )

  # BH-adjusted Wilcoxon per gene × x_col group
  .empty_stat <- data.frame(p = NA_real_, p.adj = NA_real_,
                             p.adj.signif = NA_character_)
  stat_df <- df %>%
    tidyr::pivot_longer(cols = dplyr::all_of(genes),
                        names_to = "gene", values_to = "expr") %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(c("gene", join_vars)))) %>%
    dplyr::group_modify(~ {
      dd <- .x
      if (length(unique(dd[[group_col]])) < 2) return(.empty_stat)
      tryCatch(
        rstatix::wilcox_test(dd, expr ~ !!rlang::sym(group_col),
                             exact = FALSE) %>%
          rstatix::adjust_pvalue(method = "BH") %>%
          rstatix::add_significance("p.adj",
            cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
            symbols   = c("****", "***", "**", "*", "ns")) %>%
          dplyr::select(p, p.adj, p.adj.signif),
        error = function(e) .empty_stat
      )
    }) %>%
    dplyr::ungroup()

  plot_df <- dplyr::left_join(fc_df, stat_df, by = by_cols) %>%
    dplyr::mutate(gene = factor(gene, levels = rev(genes)))

  # guard: ensure column exists even if all stat tests failed
  if (!"p.adj.signif" %in% names(plot_df)) plot_df$p.adj.signif <- NA_character_

  # create .xvar column to avoid rlang quoting issues with special column names
  plot_df$.xvar <- if (!is.null(x_col)) as.character(plot_df[[x_col]]) else "ATRXdel"
  if (!is.null(x_col) && is.factor(plot_df[[x_col]])) {
    plot_df$.xvar <- factor(plot_df$.xvar, levels = levels(plot_df[[x_col]]))
  }

  message("fc_dotplot: fc_df rows = ", nrow(fc_df), ", plot_df rows = ", nrow(plot_df))

  ggplot2::ggplot(
    plot_df,
    ggplot2::aes(x = .xvar, y = gene, size = pct, color = log2FC)
  ) +
    ggplot2::geom_point() +
    ggplot2::geom_text(ggplot2::aes(label = ifelse(is.na(p.adj.signif), "", p.adj.signif)),
                       size = 2.5, vjust = -1.5, color = "black") +
    ggplot2::scale_size_continuous(name = "% expressing",
                                   range = c(1, 8), limits = c(0, 100)) +
    ggplot2::scale_color_gradient2(
      low = GRADIENT_COLORS["low"], mid = GRADIENT_COLORS["mid"],
      high = GRADIENT_COLORS["high"], midpoint = 0,
      limits = fc_limits, oob = scales::squish,
      name = "log2FC (ATRXdel vs. ctrl)"
    ) +
    ggplot2::theme_classic(base_size = base_size) +
    ggplot2::theme(
      axis.title    = ggplot2::element_blank(),
      axis.text.x   = ggplot2::element_text(angle = 45, hjust = 1,
                                             size = base_size - 1),
      axis.text.y   = ggplot2::element_text(face = "italic",
                                             size = base_size - 1),
      axis.line     = ggplot2::element_line(linewidth = 0.4),
      axis.ticks    = ggplot2::element_line(linewidth = 0.4),
      legend.key.size   = ggplot2::unit(3, "mm"),
      legend.text       = ggplot2::element_text(size = base_size - 2),
      legend.background = ggplot2::element_blank(),
      strip.background  = ggplot2::element_blank(),
      strip.text        = ggplot2::element_text(size = base_size)
    )
}
```

# Load data

```{r load-data}
fetahu2023 <- readRDS("~/mnt_out/NK_project/Rds/fetahu2023_withNewMetadata.Rds")
fetahu2023 <- NormalizeData(fetahu2023, 
                            normalization.method = "LogNormalize",
                            scale.factor = 10000) %>% 
  ScaleData()
fetahu2023$celltype <- fetahu2023$orig.ident

ALL_NK <- readRDS("~/mnt_out/NK_project/Rds/ALL_NK_cluster_withNewMetadata.Rds")
ALL_NK <- NormalizeData(ALL_NK, 
                        normalization.method = "LogNormalize",
                        scale.factor = 10000) %>% 
  ScaleData()

ALL_NB <- readRDS("~/mnt_out/NK_project/Rds/ALL_NB_cluster_withNewMetadata.Rds")
ALL_NB <- NormalizeData(ALL_NB, 
                        normalization.method = "LogNormalize",
                        scale.factor = 10000) %>% 
  ScaleData()

# NK:  all groups, only the 4 annotated subsets (for multi-group violin)
ALL_NK_all <- subset(ALL_NK,
                     subset = predicted_NK_type_de.n350 %in% NK_SUBSET_ORDER)
ALL_NK_all$group <- factor(ALL_NK_all$group,
                            levels = names(COLOR_CODE_GROUP))
ALL_NK_all$predicted_NK_type_de.n350 <- factor(
  ALL_NK_all$predicted_NK_type_de.n350, levels = NK_SUBSET_ORDER)

# NK:  control + ATRXdel, only the 4 annotated subsets
ALL_NK_2 <- subset(ALL_NK,
                   subset = group %in% c("control", "ATRXdel") &
                            predicted_NK_type_de.n350 %in% NK_SUBSET_ORDER)
ALL_NK_2$group <- factor(ALL_NK_2$group, levels = c("control", "ATRXdel"))
ALL_NK_2$predicted_NK_type_de.n350 <- factor(ALL_NK_2$predicted_NK_type_de.n350,
                                              levels = NK_SUBSET_ORDER)

# combined group × NK subset label for dotplots showing both groups (4A, 4D)
ALL_NK_2$group_nksubset <- paste0(as.character(ALL_NK_2$group), "_",
                                   as.character(ALL_NK_2$predicted_NK_type_de.n350))
nk_gc_levels <- as.vector(rbind(paste0("ATRXdel_", NK_SUBSET_ORDER),
                                 paste0("control_",  NK_SUBSET_ORDER)))
ALL_NK_2$group_nksubset <- factor(ALL_NK_2$group_nksubset, levels = nk_gc_levels)

# NB:  tumor-only object → no control; use ATRXdel only
# droplevels() removes the unused MYCNamp / ATRXwtMYCNwt factor levels that
# Seurat::subset() retains — SCpubr checks all levels, not just present ones.
ALL_NB_A <- subset(ALL_NB, subset = group == "ATRXdel")
ALL_NB_A$group <- droplevels(ALL_NB_A$group)

# Full atlas: all groups
fetahu_all <- fetahu2023
fetahu_all$group    <- factor(fetahu_all$group,    levels = names(COLOR_CODE_GROUP))
fetahu_all$celltype <- factor(fetahu_all$celltype, levels = names(COLOR_CODE_FETAHU2023_v2))

# Full atlas: control + ATRXdel
fetahu_2 <- subset(fetahu2023, subset = group %in% c("control", "ATRXdel"))
fetahu_2$group    <- factor(fetahu_2$group,    levels = c("control", "ATRXdel"))
fetahu_2$celltype <- factor(fetahu_2$celltype, levels = names(COLOR_CODE_FETAHU2023_v2))

# Atlas + NB tumor cells merged, with group_celltype combined annotation for 4E
ALL_NB_fetahu <- subset(ALL_NB, subset = group %in% c("control", "ATRXdel"))
ALL_NB_fetahu$group    <- factor(ALL_NB_fetahu$group, levels = c("control", "ATRXdel"))
ALL_NB_fetahu$celltype <- "NB"
fetahu_2_nb <- merge(fetahu_2, ALL_NB_fetahu)
fetahu_2_nb <- NormalizeData(fetahu_2_nb, normalization.method = "LogNormalize",
                              scale.factor = 10000)
fetahu_2_nb$group    <- factor(fetahu_2_nb$group, levels = c("control", "ATRXdel"))
ct_levels_nb         <- c(levels(fetahu_2$celltype), "NB")
fetahu_2_nb$celltype <- factor(as.character(fetahu_2_nb$celltype), levels = ct_levels_nb)

# combined label: control_X then ATRXdel_X within each cell type
fetahu_2_nb$group_celltype <- paste0(as.character(fetahu_2_nb$group), "_",
                                      as.character(fetahu_2_nb$celltype))
gc_levels <- as.vector(rbind(paste0("control_",  ct_levels_nb),
                              paste0("ATRXdel_", ct_levels_nb)))
fetahu_2_nb$group_celltype <- factor(fetahu_2_nb$group_celltype, levels = gc_levels)

cat("NK_2 cells:", ncol(ALL_NK_2), "\n")
cat("NB_A cells:", ncol(ALL_NB_A), "\n")
cat("fetahu_2 cells:", ncol(fetahu_2), "\n")
cat("fetahu_2_nb cells:", ncol(fetahu_2_nb), "\n")
```

# Figure 4A — CD6 expression per NK cell state

**Layout: one panel per NK subset (hNK_Bm1–4); x = group (ctrl vs ATRXdel)**

```{r cd6-nk-violin, fig.height=3, fig.width=3}
p_cd6_vio <- do_ViolinPlot(
  ALL_NK_2,
  features   = "CD6",
  group.by   = "group",
  colors.use = COLOR_CODE_GROUP[c("control", "ATRXdel")],
  legend.title = "",
  font.size  = 12,
  plot_boxplot = FALSE,
  plot.grid = FALSE,
  grid.type = "blank") +
  ggpubr::stat_compare_means(
    comparisons = CTRL_ATRX, method = "wilcox.test",
    label = "p.signif", size = 3, tip.length = 0.01
  )
p_cd6_vio

ggsave(paste0(fig_out, "CD6_NKsubset_violin.pdf"),
       p_cd6_vio, width = 3, height = 3, device = cairo_pdf)
```

```{r cd6-nk-dot, fig.height=5, fig.width=3}
# BH-adjusted Wilcoxon + log2FC per NK subset
.cd6_stats <- Seurat::FetchData(ALL_NK_2,
                                 vars = c("CD6", "group", "predicted_NK_type_de.n350"),
                                 layer = "data") %>%
  dplyr::group_by(predicted_NK_type_de.n350) %>%
  dplyr::group_modify(~ {
    w  <- rstatix::wilcox_test(.x, CD6 ~ group, exact = FALSE) %>%
      rstatix::adjust_pvalue(method = "BH") %>%
      rstatix::add_significance("p.adj",
        cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
        symbols   = c("****", "***", "**", "*", "ns"))
    fc <- log2((mean(.x$CD6[.x$group == "ATRXdel"]) + 0.01) /
                (mean(.x$CD6[.x$group == "control"])  + 0.01))
    data.frame(signif = w$p.adj.signif, log2FC = round(fc, 2))
  }) %>%
  dplyr::ungroup()

.lvls <- levels(ALL_NK_2$NK_subtype_per_group)
.cd6_stats <- .cd6_stats %>%
  dplyr::mutate(
    y1    = sapply(predicted_NK_type_de.n350, function(s)
                     which(grepl("ATRXdel", .lvls) & grepl(s, .lvls, fixed = TRUE))),
    y2    = sapply(predicted_NK_type_de.n350, function(s)
                     which(grepl("control",  .lvls) & grepl(s, .lvls, fixed = TRUE))),
    y_mid = (y1 + y2) / 2,
    label = paste0(signif, "  log2FC=", log2FC)
  )

p_cd6_dot <- do_DotPlot(
  ALL_NK_2,
  features   = "CD6",
  group.by   = "NK_subtype_per_group",
  dot_border = FALSE,
  font.size  = 10,
  dot.scale  = 12,
  use_viridis  = TRUE,
  viridis.palette = "A",
  viridis.direction = -1,
  colors.use   = GRADIENT_COLORS_v2
)

.x_range <- ggplot_build(p_cd6_dot)$layout$panel_params[[1]]$x.range
.x_brk   <- .x_range[2] + 0.15 * diff(.x_range)

p_cd6_dot <- p_cd6_dot +
  ggplot2::geom_segment(
    data = .cd6_stats,
    ggplot2::aes(y = y1, yend = y2),
    x = .x_brk, xend = .x_brk,
    inherit.aes = FALSE, color = "black", linewidth = 0.4
  ) +
  ggplot2::geom_text(
    data = .cd6_stats,
    ggplot2::aes(y = y_mid, label = label),
    x = .x_brk + 0.05 * diff(.x_range),
    inherit.aes = FALSE, hjust = 0, size = 2.5
  ) +
  ggplot2::coord_cartesian(clip = "off") +
  ggplot2::theme(plot.margin = ggplot2::margin(t = 5, r = 45, b = 5, l = 5))

p_cd6_dot

ggsave(paste0(fig_out, "CD6_NKsubset_dotplot.pdf"),
       p_cd6_dot, width = 3, height = 5, device = cairo_pdf)
```

# Figure 4B — CD6 binding partners on NB tumor cells (ATRXdel)

**Genes: ALCAM, CDCP1, CD44 | NB = tumor-only, no control**

```{r cd6lig-nb-violin, fig.height=3, fig.width=5}
p_cd6lig_vio <- do_ViolinPlot(
  ALL_NB_A,
  features     = CD6_NB_GENES,
  colors.use   = COLOR_CODE_GROUP["ATRXdel"],
  legend.title = "",
  font.size    = 9,
  plot_boxplot = FALSE,
  y_cut = c(3,3,3),
)
p_cd6lig_vio

ggsave(paste0(fig_out, "CD6ligands_NB_violin.pdf"),
       p_cd6lig_vio, width = 5, height = 3, device = cairo_pdf)
```

```{r cd6lig-nb-dot, fig.height=4, fig.width=6}
# BH-adjusted Wilcoxon + log2FC per gene, for two comparisons vs ATRXdel
.nb_stats <- Seurat::FetchData(ALL_NB,
                                vars = c(CD6_NB_GENES, "group"),
                                layer = "data") %>%
  tidyr::pivot_longer(cols = dplyr::all_of(CD6_NB_GENES),
                      names_to = "gene", values_to = "expr") %>%
  dplyr::group_by(gene) %>%
  dplyr::group_modify(~ {
    dd <- .x
    purrr::map_dfr(list(c("ATRXdel","ATRXwtMYCNwt"), c("ATRXdel","MYCNamp")), function(grps) {
      sub <- dplyr::filter(dd, group %in% grps)
      sub$group <- factor(sub$group, levels = grps)
      w <- tryCatch(
        rstatix::wilcox_test(sub, expr ~ group, exact = FALSE) %>%
          rstatix::adjust_pvalue(method = "BH") %>%
          rstatix::add_significance("p.adj",
            cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
            symbols   = c("****", "***", "**", "*", "ns")),
        error = function(e) data.frame(p.adj.signif = "ns")
      )
      fc <- log2((mean(sub$expr[sub$group == "ATRXdel"]) + 0.01) /
                  (mean(sub$expr[sub$group == grps[2]])   + 0.01))
      data.frame(comparison = paste(grps, collapse = " vs "),
                 grp2 = grps[2],
                 signif = w$p.adj.signif[1],
                 log2FC = round(fc, 2))
    })
  }) %>%
  dplyr::ungroup()

p_cd6lig_dot <- do_DotPlot(
  ALL_NB,
  features     = CD6_NB_GENES,
  group.by     = "group",
  scale        = TRUE,
  dot.scale    = 15,
  font.size    = 12,
  dot_border   = FALSE,
  use_viridis  = TRUE,
  viridis.palette = "A",
  viridis.direction = -1,
  colors.use   = GRADIENT_COLORS_v2
)

# detect y-axis group order from the built plot
.pb       <- ggplot_build(p_cd6lig_dot)
.y_groups <- .pb$layout$panel_params[[1]]$y$limits   # groups bottom → top
.x_range  <- .pb$layout$panel_params[[1]]$x.range

.y_pos <- function(grp) which(.y_groups == grp)
.x_brk1 <- .x_range[2] + 0.20 * diff(.x_range)   # ATRXdel vs ATRXwtMYCNwt
.x_brk2 <- .x_range[2] + 0.55 * diff(.x_range)   # ATRXdel vs MYCNamp
.x_tick  <- 0.06 * diff(.x_range)

.make_cmp <- function(grp2, x_brk) {
  sub <- dplyr::filter(.nb_stats, grp2 == !!grp2) %>%
    dplyr::arrange(factor(gene, levels = CD6_NB_GENES))
  lbl <- paste(paste0(sub$gene, ": ", sub$signif, " (FC=", sub$log2FC, ")"),
               collapse = "\n")
  y1  <- .y_pos("ATRXdel")
  y2  <- .y_pos(grp2)
  list(
    ggplot2::geom_segment(data = data.frame(x_brk = x_brk, y1 = y1, y2 = y2),
      ggplot2::aes(y = y1, yend = y2), x = x_brk, xend = x_brk,
      inherit.aes = FALSE, color = "black", linewidth = 0.4),
    ggplot2::geom_text(data = data.frame(x_brk = x_brk + .x_tick,
                                          y_mid = (y1 + y2) / 2, label = lbl),
      ggplot2::aes(y = y_mid, label = label), x = x_brk + .x_tick,
      inherit.aes = FALSE, hjust = 0, size = 2.2, lineheight = 0.9)
  )
}

p_cd6lig_dot <- p_cd6lig_dot +
  .make_cmp("ATRXwtMYCNwt", .x_brk1) +
  .make_cmp("MYCNamp",       .x_brk2) +
  ggplot2::coord_cartesian(clip = "off") +
  ggplot2::theme(plot.margin = ggplot2::margin(t = 5, r = 120, b = 5, l = 5))

p_cd6lig_dot

ggsave(paste0(fig_out, "CD6ligands_NB_dotplot.pdf"),
       p_cd6lig_dot, width = 6, height = 4, device = cairo_pdf)
```

# Figure 4C — ALCAM expression in NB tumor cells (ATRXdel)

```{r alcam-nb-violin, fig.height=3, fig.width=2.5}
p_alcam_vio <- do_ViolinPlot(
  ALL_NB,
  features     = "ALCAM",
  group.by     = "group",
  colors.use   = COLOR_CODE_GROUP,
  legend.title = "",
  font.size    = 9,
  plot_boxplot = FALSE
)
p_alcam_vio

ggsave(paste0(fig_out, "ALCAM_NB_violin.pdf"),
       p_alcam_vio, width = 2.5, height = 3, device = cairo_pdf)
```

# Figure 4D — ALCAM-binding partners on NK subsets (ctrl vs ATRXdel)

**Genes: CD6, LGALS1, LGALS8, LILRB4, L1CAM, S100B**
**Layout: one panel per gene; x = group (ctrl vs ATRXdel), facet by NK subset within each panel**

```{r alcamlig-nk-violin, fig.height=10, fig.width=8}
p_alcamlig_vio <- VlnPlot(
  ALL_NK_2,
  features = ALCAM_NK_GENES,
  split.by = "group",
  group.by = "predicted_NK_type_de.n350",
  cols     = COLOR_CODE_GROUP,
  log      = TRUE,
  pt.size  = 0
) & ggplot2::coord_cartesian(ylim = c(NA, 5))
p_alcamlig_vio

ggsave(paste0(fig_out, "ALCAMligands_NK_violin.pdf"),
       p_alcamlig_vio, width = 8, height = 10, device = cairo_pdf)
```

```{r alcamlig-nk-allgroups-violin, fig.height=8, fig.width=8}
# Same layout as alcamlig-nk-violin but including all groups (not only ctrl + ATRXdel)
p_alcamlig_vio_all <- VlnPlot(
  ALL_NK_all,
  features = ALCAM_NK_GENES,
  split.by = "group",
  group.by = "predicted_NK_type_de.n350",
  cols     = COLOR_CODE_GROUP,
  log      = TRUE,
  pt.size  = 0
) & ggplot2::coord_cartesian(ylim = c(NA, 5))
p_alcamlig_vio_all

ggsave(paste0(fig_out, "ALCAMligands_NK_allgroups_violin.pdf"),
       p_alcamlig_vio_all, width = 8, height = 8, device = cairo_pdf)
```

```{r alcamlig-nk-allgroups-violin-2, fig.height=8, fig.width=8}
# Same layout as alcamlig-nk-violin but including all groups (not only ctrl + ATRXdel)
p_alcamlig_vio_all <- VlnPlot(
  ALL_NK_all,
  features     = ALCAM_NK_GENES,
  group.by     = "group",
  cols         = COLOR_CODE_GROUP,
  log          = TRUE,
  pt.size      = 0
) & ggplot2::coord_cartesian(ylim = c(NA, 5))
p_alcamlig_vio_all

ggsave(paste0(fig_out, "ALCAMligands_NK_allgroups_violin_perGroup.pdf"),
       p_alcamlig_vio_all, width = 8, height = 8, device = cairo_pdf)
```

```{r alcamlig-nk-group-stats}
# BH-adjusted Wilcoxon (control vs ATRXdel) per gene, across all NK cells (group level)
# p-values BH-adjusted globally across all genes
.alcamlig_nk_group_stats <- Seurat::FetchData(
  ALL_NK_2,
  vars  = c(ALCAM_NK_GENES, "group"),
  layer = "data"
) %>%
  tidyr::pivot_longer(cols = dplyr::all_of(ALCAM_NK_GENES),
                      names_to = "gene", values_to = "expr") %>%
  dplyr::group_by(gene) %>%
  dplyr::group_modify(~ {
    dd <- .x
    if (length(unique(dd$group)) < 2)
      return(data.frame(log2FC = NA_real_, p = NA_real_))
    w <- tryCatch(
      rstatix::wilcox_test(dd, expr ~ group, exact = FALSE),
      error = function(e) NULL
    )
    fc <- log2((mean(dd$expr[dd$group == "ATRXdel"]) + 0.01) /
                (mean(dd$expr[dd$group == "control"])  + 0.01))
    if (is.null(w)) return(data.frame(log2FC = round(fc, 3), p = NA_real_))
    data.frame(log2FC = round(fc, 3), p = w$p[1])
  }) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    p.adj        = p.adjust(p, method = "BH"),
    p.adj.signif = dplyr::case_when(
      is.na(p.adj)   ~ NA_character_,
      p.adj < 0.0001 ~ "****",
      p.adj < 0.001  ~ "***",
      p.adj < 0.01   ~ "**",
      p.adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    )
  ) %>%
  dplyr::arrange(gene)

knitr::kable(.alcamlig_nk_group_stats, digits = 4,
             caption = "BH-adjusted Wilcoxon: control vs ATRXdel — ALCAM-binding NK genes (all NK cells pooled)")
```

```{r alcamlig-nk-stats}
# BH-adjusted Wilcoxon (control vs ATRXdel) per gene × NK subset
# p-values BH-adjusted globally across all gene×subset combinations
.alcamlig_nk_stats <- Seurat::FetchData(
  ALL_NK_2,
  vars  = c(ALCAM_NK_GENES, "group", "predicted_NK_type_de.n350"),
  layer = "data"
) %>%
  tidyr::pivot_longer(cols = dplyr::all_of(ALCAM_NK_GENES),
                      names_to = "gene", values_to = "expr") %>%
  dplyr::group_by(gene, predicted_NK_type_de.n350) %>%
  dplyr::group_modify(~ {
    dd <- .x
    if (length(unique(dd$group)) < 2)
      return(data.frame(log2FC = NA_real_, p = NA_real_))
    w <- tryCatch(
      rstatix::wilcox_test(dd, expr ~ group, exact = FALSE),
      error = function(e) NULL
    )
    fc <- log2((mean(dd$expr[dd$group == "ATRXdel"]) + 0.01) /
                (mean(dd$expr[dd$group == "control"])  + 0.01))
    if (is.null(w)) return(data.frame(log2FC = round(fc, 3), p = NA_real_))
    data.frame(log2FC = round(fc, 3), p = w$p[1])
  }) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    p.adj        = p.adjust(p, method = "BH"),
    p.adj.signif = dplyr::case_when(
      is.na(p.adj)   ~ NA_character_,
      p.adj < 0.0001 ~ "****",
      p.adj < 0.001  ~ "***",
      p.adj < 0.01   ~ "**",
      p.adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    )
  ) %>%
  dplyr::rename(NK_subset = predicted_NK_type_de.n350) %>%
  dplyr::arrange(gene, NK_subset)

knitr::kable(.alcamlig_nk_stats, digits = 4,
             caption = "BH-adjusted Wilcoxon: control vs ATRXdel — ALCAM-binding NK genes per NK subset")
```

```{r alcamlig-nk-dot, fig.height=3, fig.width=3}
# x = group (control vs ATRXdel), y = gene, color = avg expression, size = % expressing
p_alcamlig_dot <- do_DotPlot(
  ALL_NK_2,
  features     = ALCAM_NK_GENES,
  group.by     = "group",
  dot_border   = FALSE,
  font.size    = 9,
  dot.scale    = 10,
  use_viridis  = TRUE,
  viridis.palette = "A",
  viridis.direction = -1,
  colors.use   = GRADIENT_COLORS_v2
)
p_alcamlig_dot

ggsave(paste0(fig_out, "ALCAMligands_NK_dotplot.pdf"),
       p_alcamlig_dot, width = 3, height = 3, device = cairo_pdf)
```


# Figure 4E — ALCAM + CD6 on all cell types (ctrl vs ATRXdel)

**Stacked: ALCAM on top, CD6 on bottom**

```{r alcam-cd6-allt-violin, fig.height=7, fig.width=18}
.vio_gene <- function(gene) {
  Seurat::VlnPlot(
    fetahu_2,
    features = gene,
    group.by = "celltype",
    split.by = "group",
    cols     = unname(COLOR_CODE_GROUP[c("control", "ATRXdel")]),
    pt.size  = 0,
    log = TRUE
  ) + ggplot2::ggtitle(gene) +
    ggplot2::theme(plot.title = ggplot2::element_text(face = "italic"))
}

p_stacked_vio <- patchwork::wrap_plots(
  .vio_gene("ALCAM"),
  .vio_gene("CD6"),
  ncol = 1
)
p_stacked_vio

ggsave(paste0(fig_out, "ALCAM_CD6_allCelltypes_violin.pdf"),
       p_stacked_vio, width = 18, height = 7, device = cairo_pdf)
```

```{r alcam-cd6-allt-allgroups-violin, fig.height=7, fig.width=18}
.vio_gene_all <- function(gene) {
  Seurat::VlnPlot(
    fetahu_all,
    features = gene,
    group.by = "celltype",
    split.by = "group",
    cols     = unname(COLOR_CODE_GROUP),
    pt.size  = 0,
    log      = TRUE
  ) + ggplot2::ggtitle(gene) +
    ggplot2::theme(plot.title = ggplot2::element_text(face = "italic")
}

p_stacked_vio_all <- patchwork::wrap_plots(
  .vio_gene_all("ALCAM"),
  .vio_gene_all("CD6"),
  ncol = 1
)
p_stacked_vio_all

ggsave(paste0(fig_out, "ALCAM_CD6_allCelltypes_allgroups_violin.pdf"),
       p_stacked_vio_all, width = 18, height = 7, device = cairo_pdf)
```

```{r alcam-cd6-allt-stats}
# BH-adjusted Wilcoxon (control vs ATRXdel) per gene × cell type
# p-values BH-adjusted globally across all gene×celltype combinations
.allt_stats <- Seurat::FetchData(
  fetahu_2,
  vars  = c("ALCAM", "CD6", "group", "celltype"),
  layer = "data"
) %>%
  tidyr::pivot_longer(cols = c("ALCAM", "CD6"),
                      names_to = "gene", values_to = "expr") %>%
  dplyr::group_by(gene, celltype) %>%
  dplyr::group_modify(~ {
    dd <- .x
    if (length(unique(dd$group)) < 2)
      return(data.frame(log2FC = NA_real_, p = NA_real_))
    w <- tryCatch(
      rstatix::wilcox_test(dd, expr ~ group, exact = FALSE),
      error = function(e) NULL
    )
    fc <- log2((mean(dd$expr[dd$group == "ATRXdel"]) + 0.01) /
                (mean(dd$expr[dd$group == "control"])  + 0.01))
    if (is.null(w)) return(data.frame(log2FC = round(fc, 3), p = NA_real_))
    data.frame(log2FC = round(fc, 3), p = w$p[1])
  }) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    p.adj        = p.adjust(p, method = "BH"),
    p.adj.signif = dplyr::case_when(
      is.na(p.adj)   ~ NA_character_,
      p.adj < 0.0001 ~ "****",
      p.adj < 0.001  ~ "***",
      p.adj < 0.01   ~ "**",
      p.adj < 0.05   ~ "*",
      TRUE           ~ "ns"
    )
  ) %>%
  dplyr::arrange(gene, celltype)

knitr::kable(.allt_stats, digits = 4,
             caption = "BH-adjusted Wilcoxon: control vs ATRXdel — ALCAM & CD6 per cell type (atlas)")
```

```{r alcam-cd6-allt-dot, fig.height=5, fig.width=10}
p_alcam_dot <- do_DotPlot(
  fetahu_2_nb,
  features   = "ALCAM",
  group.by   = "group_celltype",
  dot_border = FALSE,
  font.size  = 8,
  flip       = TRUE,
  dot.scale  = 6,
  use_viridis = TRUE,
  viridis.palette = "A",
  viridis.direction = -1
)

p_cd6_dot <- do_DotPlot(
  fetahu_2_nb,
  features   = "CD6",
  group.by   = "group_celltype",
  dot_border = FALSE,
  font.size  = 8,
  flip       = TRUE,
  dot.scale  = 6,
  use_viridis = TRUE,
  viridis.palette = "A",
  viridis.direction = -1
)

p_stacked_dot <- patchwork::wrap_plots(p_alcam_dot, p_cd6_dot, nrow = 2)
p_stacked_dot

ggsave(paste0(fig_out, "ALCAM_CD6_allCelltypes_dotplot.pdf"),
       p_stacked_dot, width = 10, height = 5, device = cairo_pdf)
```

# SessionInfo

```{r}
sessionInfo()
```
