---
title: "NK project: Paper revision figures, part II"
subtitle: "Bulk RNA/ATAC/ChIP-seq"
author: "Dr. Sara Wernig-Zorc"
date: "2026-06-03"
output: 
  html_document: 
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  out: "/home/rstudio/mnt_out/"
  anno: "/home/rstudio/mnt_resources/"
  markers: "/home/rstudio/mnt_resources/MES_ADRN/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE, 
                      error = FALSE, 
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 80)
options(future.globals.maxSize = 80 * 10^9)
print(params)
```

# ___Project specific setup___

```{r load project specific R packages and color code, echo = FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
library(Gviz)
set.seed(2805)
```


*Data from: van Gerven et al., 2023, Lorenzi et al., 2025 and Qadeer et al., 2019*

# ___Bulk RNA-seq: ALCAM expression in isogenic NB cell lines___

## ___van Gerven et al., 2023___

*Isogenic NB cell lines: GI-M-EN, NB139, SK-N-AS*

```{r load-isogenic-counts}
raw_counts_isogenic <- read.csv(
  "/home/rstudio/mnt_data/nb_tumors/vanGerven2023.bulkRNA/processed_data/GSE226770_RNA_raw_counts_isogenic_cell_lines.csv",
  row.names = 1,
  check.names = FALSE
)

CLONE_TYPE_LEVELS <- c("WT", "KO", "Del2-10", "Del2-13", "Del10-12")

COLOR_CODE_CLONE <- c(
  "WT"       = "#FAA21B",
  "KO"       = "#4f4e92",
  "Del2-10"  = "#7b5ea7",
  "Del2-13"  = "#9b7ec8",
  "Del10-12" = "#bba0d8"
)

# Build metadata from column names
sample_meta_isogenic <- data.frame(
  sample     = colnames(raw_counts_isogenic),
  cell_line  = sub("\\.(WT|KO|Del).*", "", colnames(raw_counts_isogenic)),
  clone_type = sub("^[^.]+\\.([^.]+)\\..*", "\\1", colnames(raw_counts_isogenic)),
  stringsAsFactors = FALSE
)
rownames(sample_meta_isogenic) <- sample_meta_isogenic$sample

sample_meta_isogenic$clone_type <- factor(sample_meta_isogenic$clone_type,
                                           levels = CLONE_TYPE_LEVELS)  # WT = reference
sample_meta_isogenic$cell_line  <- factor(sample_meta_isogenic$cell_line,
                                           levels = c("GIMEN", "NB139", "SKNAS"))
```

```{r deseq2-isogenic}
# Pre-filter: require >= 10 counts in at least 4 samples (= smallest single replicate group)
dds_isogenic <- DESeqDataSetFromMatrix(
  countData = raw_counts_isogenic,
  colData   = sample_meta_isogenic,
  # cell_line is the paired factor (isogenic design).
  # clone_type captures each deletion separately; WT is the reference level.
  design    = ~ cell_line + clone_type
)
keep         <- rowSums(counts(dds_isogenic) >= 10) >= 4
dds_isogenic <- dds_isogenic[keep, ]
cat("Genes retained after filtering:", nrow(dds_isogenic), "\n")

# Full DESeq2 pipeline
dds_isogenic <- DESeq(dds_isogenic)

# VST for visualization (blind=FALSE: uses design dispersion estimates)
vst_isogenic     <- vst(dds_isogenic, blind = FALSE)
vst_mat_isogenic <- assay(vst_isogenic)
```

```{r deseq2-isogenic-results}
library(ashr)  # required for lfcShrink(type = "ashr")
# Each clone type vs WT, controlling for cell line
clone_types_tested <- c("KO", "Del2-10", "Del2-13", "Del10-12")

res_isogenic_list <- lapply(clone_types_tested, function(ct) {
  res <- results(dds_isogenic,
                 contrast = c("clone_type", ct, "WT"),
                 alpha    = 0.05)
  res <- lfcShrink(dds_isogenic,
                   contrast = c("clone_type", ct, "WT"),
                   res      = res,
                   type     = "ashr")
  df <- as.data.frame(res) %>%
    rownames_to_column("ensembl_id") %>%
    arrange(padj)
  df[["clone_type"]] <- ct  # plain character assignment avoids list-column from mutate()
  df
})
res_isogenic_df <- bind_rows(res_isogenic_list)
res_isogenic_df$clone_type <- as.character(res_isogenic_df$clone_type)

cat("Significant DEGs per clone type (padj < 0.05, |LFC| > 1):\n")
sig_degs <- res_isogenic_df[
  !is.na(res_isogenic_df$padj) &
  abs(res_isogenic_df$log2FoldChange) > 1 &
  res_isogenic_df$padj < 0.05, ]
table(as.character(sig_degs$clone_type))

# ALCAM result per clone type
ALCAM_ENSG <- "ENSG00000170017"
alcam_res  <- res_isogenic_df[res_isogenic_df$ensembl_id == ALCAM_ENSG,
                               c("clone_type", "log2FoldChange", "lfcSE", "pvalue", "padj")]
alcam_res$clone_type <- as.character(alcam_res$clone_type)
print(alcam_res)
```

```{r alcam-size-factor-normalized}
# log2(size-factor normalized + 1) — re-uses size factors from dds_isogenic
norm_counts_isogenic <- counts(dds_isogenic, normalized = TRUE)
log_norm_isogenic    <- log2(norm_counts_isogenic + 1)

alcam_lognorm <- data.frame(
  sample     = colnames(log_norm_isogenic),
  expression = as.numeric(log_norm_isogenic[ALCAM_ENSG, ]),
  cell_line  = sample_meta_isogenic$cell_line,
  clone_type = sample_meta_isogenic$clone_type
)
```

```{r alcam-raw-counts-diagnostic}
# Diagnostic: raw ALCAM counts per sample
alcam_raw <- data.frame(
  sample     = colnames(raw_counts_isogenic),
  raw_count  = as.numeric(raw_counts_isogenic[ALCAM_ENSG, ]),
  cell_line  = sample_meta_isogenic$cell_line,
  clone_type = sample_meta_isogenic$clone_type
)
alcam_raw %>% arrange(cell_line, clone_type)
```

```{r alcam-lognorm-plot, fig.height=5, fig.width=10}
# Build one panel per cell line with only the comparisons that exist in that panel,
# then combine with patchwork — this avoids ggpubr silently dropping all brackets
# when a clone type is absent from a facet.

make_alcam_panel <- function(cl, comparisons, show_legend = FALSE) {
  df <- droplevels(alcam_lognorm[alcam_lognorm$cell_line == cl, ])
  ggboxplot(
    df,
    x             = "clone_type",
    y             = "expression",
    fill          = "clone_type",
    add           = "jitter",
    add.params    = list(size = 2.5, alpha = 0.8, color = "black"),
    outlier.shape = NA
  ) +
    scale_fill_manual(values = COLOR_CODE_CLONE, name = "Clone type",
                      drop = TRUE) +
    stat_compare_means(
      comparisons = comparisons,
      method      = "wilcox.test",
      label       = "p.format"
    ) +
    labs(title = cl,
         y     = expression(log[2] ~ "(normalized counts + 1)"),
         x     = "") +
    theme_pubr(base_size = 12) +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1),
      legend.position = if (show_legend) "right" else "none",
      plot.title      = element_text(face = "bold", size = 12)
    )
}

p_gimen <- make_alcam_panel("GIMEN",
  list(c("WT", "KO"), c("WT", "Del2-10"), c("WT", "Del2-13")))
p_nb139 <- make_alcam_panel("NB139",
  list(c("WT", "Del10-12"), c("WT", "Del2-10")))
p_sknas <- make_alcam_panel("SKNAS",
  list(c("WT", "KO")),
  show_legend = TRUE)

alcam_lognorm_plot <- (p_gimen | p_nb139 | p_sknas) +
  plot_annotation(
    title = "ALCAM expression in isogenic NB cell lines",
    subtitle = "van Gerven et al., 2023",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )

alcam_lognorm_plot

pdf(paste0(params$out, "NK_project/figures/isogenic_ALCAM_lognorm_per_clonetype.pdf"),
    height = 5, width = 10)
alcam_lognorm_plot
dev.off()
```

```{r alcam-lognorm-stats}
alcam_lognorm %>%
  group_by(cell_line, clone_type) %>%
  summarise(n      = n(),
            mean   = round(mean(expression), 3),
            median = round(median(expression), 3),
            sd     = round(sd(expression), 3),
            .groups = "drop")
```

## ___Qadeer et al., 2019___

*Isogenic NB cell lines: SK-N-FI*

```{r load-sknfi-fpkm}
# grep the ALCAM expression
alcam_line_sknfi <- grep(
  "^ALCAM\\s",
  readLines("/home/rstudio/mnt_data/nb_tumors/qadeer2019.bulkRNA.bulkChIP/processed_data/GSE100148/GSE100148_SKNFI_MS2_IFF_FPKM.txt"),
  value = TRUE
)
alcam_fpkm_sknfi <- as.numeric(strsplit(trimws(alcam_line_sknfi), "\\s+")[[1]][-1])
names(alcam_fpkm_sknfi) <- c("FI.MS2.R1", "FI.MS2.R2", "FI.MS2.IFF.R1", "FI.MS2.IFF.R2")
alcam_fpkm_sknfi

# Metadata: MS2 = ATRXwt (parental); IFF = ATRXdel (ATRX-deleted)
sample_meta_sknfi <- data.frame(
  sample      = names(alcam_fpkm_sknfi),
  ATRX_status = ifelse(grepl("IFF", names(alcam_fpkm_sknfi)), "ATRXdel", "ATRXwt"),
  stringsAsFactors = FALSE
)
sample_meta_sknfi$ATRX_status <- factor(sample_meta_sknfi$ATRX_status,
                                         levels = c("ATRXwt", "ATRXdel"))

COLOR_CODE_ATRX <- c("ATRXwt" = "#FAA21B", "ATRXdel" = "#4f4e92")
```

```{r alcam-sknfi-df}
alcam_sknfi <- data.frame(
  sample      = names(alcam_fpkm_sknfi),
  expression  = log2(alcam_fpkm_sknfi + 1),
  ATRX_status = sample_meta_sknfi$ATRX_status
)
alcam_sknfi
```

```{r alcam-sknfi-plot, fig.height=5, fig.width=3}
alcam_sknfi_plot <- ggboxplot(
  alcam_sknfi,
  x             = "ATRX_status",
  y             = "expression",
  fill          = "ATRX_status",
  add           = "jitter",
  add.params    = list(size = 3, alpha = 0.9, color = "black"),
  outlier.shape = NA
) +
  scale_fill_manual(values = COLOR_CODE_ATRX, name = "ATRX status") +
  stat_compare_means(
    comparisons = list(c("ATRXwt", "ATRXdel")),
    method      = "wilcox.test",
    label       = "p.format"
  ) +
  labs(
    title   = "SK-N-FI",
    subtitle = "Qadeer et al., 2019",
    y       = expression(log[2] ~ "(FPKM + 1)"),
    x       = "",
    caption = ""
  ) +
  theme_pubr(base_size = 12) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title      = element_text(face = "bold", size = 13),
    plot.caption    = element_text(size = 9, color = "gray40")
  )

alcam_sknfi_plot

pdf(paste0(params$out, "NK_project/figures/SKNFI_ALCAM_ATRXwt_vs_ATRXdel.pdf"),
    height = 5, width = 3)
alcam_sknfi_plot
dev.off()
```

# ___DEG pathway enrichment analysis: ATRXdel vs ATRXwt___

## ___van Gerven et al., 2023___

*Isogenic NB cell lines: GI-M-EN, NB139, SK-N-AS*

*One representative deletion per cell line selected: GIMEN KO, NB139 Del2-10, SKNAS KO.*

```{r deseq2-selected-clones}
# Select the purest deletion type per cell line (4 WT + 4 del each, balanced)
selected_samples <- c(
  grep("^GIMEN\\.WT\\.",     colnames(raw_counts_isogenic), value = TRUE),
  grep("^GIMEN\\.KO\\.",     colnames(raw_counts_isogenic), value = TRUE),
  grep("^NB139\\.WT\\.",     colnames(raw_counts_isogenic), value = TRUE),
  grep("^NB139\\.Del2-10\\.",colnames(raw_counts_isogenic), value = TRUE),
  grep("^SKNAS\\.WT\\.",     colnames(raw_counts_isogenic), value = TRUE),
  grep("^SKNAS\\.KO\\.",     colnames(raw_counts_isogenic), value = TRUE)
)

meta_selected <- sample_meta_isogenic[selected_samples, , drop = FALSE]
meta_selected$ATRX_status <- factor(
  ifelse(meta_selected$clone_type == "WT", "ATRXwt", "ATRXdel"),
  levels = c("ATRXwt", "ATRXdel")
)

# Re-use pre-filtered gene set from dds_isogenic
dds_selected <- DESeqDataSetFromMatrix(
  countData = counts(dds_isogenic)[, selected_samples],
  colData   = meta_selected,
  design    = ~ cell_line + ATRX_status
)
dds_selected <- DESeq(dds_selected)
cat("Samples:", ncol(dds_selected), "| Genes:", nrow(dds_selected), "\n")
table(meta_selected$cell_line, meta_selected$ATRX_status)
```

```{r deseq2-selected-results}
res_selected <- results(dds_selected,
                        contrast = c("ATRX_status", "ATRXdel", "ATRXwt"),
                        alpha    = 0.05)
res_selected <- lfcShrink(dds_selected,
                           contrast = c("ATRX_status", "ATRXdel", "ATRXwt"),
                           res      = res_selected,
                           type     = "ashr")

res_selected_df <- as.data.frame(res_selected) %>%
  rownames_to_column("ensembl_id") %>%
  arrange(padj)

n_sig  <- sum(res_selected_df$padj < 0.05 & abs(res_selected_df$log2FoldChange) > 1, na.rm = TRUE)
n_up   <- sum(res_selected_df$padj < 0.05 & res_selected_df$log2FoldChange > 1,  na.rm = TRUE)
n_down <- sum(res_selected_df$padj < 0.05 & res_selected_df$log2FoldChange < -1, na.rm = TRUE)
cat("Significant DEGs (padj<0.05, |LFC|>1):", n_sig, "— Up:", n_up, "| Down:", n_down, "\n")

# ALCAM result
res_selected_df[res_selected_df$ensembl_id == ALCAM_ENSG,
                c("ensembl_id", "log2FoldChange", "lfcSE", "pvalue", "padj")]
```

```{r deg-id-conversion}
# Convert Ensembl → Entrez + Symbol for downstream enrichment
gene_map <- bitr(res_selected_df$ensembl_id,
                 fromType = "ENSEMBL",
                 toType   = c("ENTREZID", "SYMBOL"),
                 OrgDb    = org.Hs.eg.db)

res_selected_annot <- res_selected_df %>%
  left_join(gene_map, by = c("ensembl_id" = "ENSEMBL"))

# Gene universe: expressed, tested genes with a valid Entrez ID
geneUniverse_entrez <- unique(gene_map$ENTREZID)
cat("Gene universe:", length(geneUniverse_entrez), "genes\n")

# Significant subsets (upregulated = primary interest given expected ALCAM↑ with ATRXdel)
sig_up   <- res_selected_annot %>% filter(padj < 0.05, log2FoldChange > 1,  !is.na(ENTREZID))
sig_down <- res_selected_annot %>% filter(padj < 0.05, log2FoldChange < -1, !is.na(ENTREZID))
cat("Up:", nrow(sig_up), "| Down:", nrow(sig_down), "genes with Entrez IDs\n")
```

```{r volcano-selected, fig.height=8, fig.width=8}
top_labels <- res_selected_annot %>%
  filter(!is.na(padj), !is.na(SYMBOL)) %>%
  slice_min(padj, n = 15) %>%
  pull(SYMBOL)

volcano_plot <- EnhancedVolcano(
  res_selected_annot,
  lab             = res_selected_annot$SYMBOL,
  x               = "log2FoldChange",
  y               = "padj",
  selectLab       = top_labels,
  pCutoff         = 0.05,
  FCcutoff        = 1,
  title           = "ATRXdel vs ATRXwt",
  subtitle        = "Isogenic NB cell lines — GIMEN KO, NB139 Del2-10, SKNAS KO",
  caption         = paste0("Up: ", n_up, " | Down: ", n_down, " (padj<0.05, |LFC|>1)"),
  legendPosition  = "right",
  drawConnectors  = TRUE,
  widthConnectors = 0.4
)
volcano_plot

pdf(paste0(params$out, "NK_project/figures/DEG_volcano_ATRXdel_vs_ATRXwt.pdf"),
    height = 8, width = 8)
volcano_plot
dev.off()
```

```{r ora-go, fig.height=6, fig.width=8}
ego_up <- enrichGO(gene          = unique(sig_up$ENTREZID),
                   universe      = geneUniverse_entrez,
                   OrgDb         = org.Hs.eg.db,
                   ont           = "BP",
                   pAdjustMethod = "fdr",
                   pvalueCutoff  = 0.05,
                   qvalueCutoff  = 0.1,
                   minGSSize     = 10,
                   maxGSSize     = 500,
                   readable      = TRUE)

ego_down <- enrichGO(gene          = unique(sig_down$ENTREZID),
                     universe      = geneUniverse_entrez,
                     OrgDb         = org.Hs.eg.db,
                     ont           = "BP",
                     pAdjustMethod = "fdr",
                     pvalueCutoff  = 0.05,
                     qvalueCutoff  = 0.2,
                     minGSSize     = 2,
                     maxGSSize     = 5000,
                     readable      = TRUE)

if (!is.null(ego_up) && nrow(ego_up) > 0) {
  p_go_up <- dotplot(ego_up, showCategory = 20, title = "GO:BP — upregulated in ATRXdel")
  print(p_go_up)
  pdf(paste0(params$out, "NK_project/figures/ORA_GO_up_ATRXdel.pdf"), height = 6, width = 8)
  print(p_go_up); dev.off()
  if (nrow(ego_up) >= 10) {
    p_tree_up <- treeplot(pairwise_termsim(ego_up), nCluster = 4, label_format = 30)
    print(p_tree_up)
    pdf(paste0(params$out, "NK_project/figures/ORA_GO_up_ATRXdel_tree.pdf"), height = 8, width = 12)
    print(p_tree_up); dev.off()
  }
} else { cat("No significant GO:BP terms for upregulated genes.\n") }

if (!is.null(ego_down) && nrow(ego_down) > 0) {
  p_go_down <- dotplot(ego_down, showCategory = 20, title = "GO:BP — downregulated in ATRXdel")
  print(p_go_down)
  pdf(paste0(params$out, "NK_project/figures/ORA_GO_down_ATRXdel.pdf"), height = 6, width = 8)
  print(p_go_down); dev.off()
  if (nrow(ego_down) >= 10) {
    p_tree_down <- treeplot(pairwise_termsim(ego_down), nCluster = 4, label_format = 30)
    print(p_tree_down)
    pdf(paste0(params$out, "NK_project/figures/ORA_GO_down_ATRXdel_tree.pdf"), height = 8, width = 12)
    print(p_tree_down); dev.off()
  }
} else { cat("No significant GO:BP terms for downregulated genes.\n") }
```

```{r ora-kegg, fig.height=6, fig.width=8}
ekegg_up <- enrichKEGG(gene          = unique(sig_up$ENTREZID),
                       universe      = geneUniverse_entrez,
                       organism      = "hsa",
                       keyType       = "kegg",
                       pAdjustMethod = "fdr",
                       pvalueCutoff  = 0.05,
                       qvalueCutoff  = 0.1,
                       minGSSize     = 10,
                       maxGSSize     = 500)

ekegg_down <- enrichKEGG(gene          = unique(sig_down$ENTREZID),
                         universe      = geneUniverse_entrez,
                         organism      = "hsa",
                         keyType       = "kegg",
                         pAdjustMethod = "fdr",
                         pvalueCutoff  = 0.05,
                         qvalueCutoff  = 0.2,
                         minGSSize     = 2,
                         maxGSSize     = 5000)

if (!is.null(ekegg_up) && nrow(ekegg_up) > 0) {
  p_kegg_up <- dotplot(ekegg_up, showCategory = 20, title = "KEGG — upregulated in ATRXdel")
  print(p_kegg_up)
  pdf(paste0(params$out, "NK_project/figures/ORA_KEGG_up_ATRXdel.pdf"), height = 6, width = 8)
  print(p_kegg_up); dev.off()
} else { cat("No significant KEGG terms for upregulated genes.\n") }

if (!is.null(ekegg_down) && nrow(ekegg_down) > 0) {
  p_kegg_down <- dotplot(ekegg_down, showCategory = 20, title = "KEGG — downregulated in ATRXdel")
  print(p_kegg_down)
  pdf(paste0(params$out, "NK_project/figures/ORA_KEGG_down_ATRXdel.pdf"), height = 6, width = 8)
  print(p_kegg_down); dev.off()
} else { cat("No significant KEGG terms for downregulated genes.\n") }
```


## ___Qadeer et al., 2019___

*Isogenic NB cell lines: SK-N-FI*

*Note: For Qadeer et al., 2019 only FPKM data is available, but raw counts required for DESeq2.*


# ___GSEAnalysis: MES and inflammation signatures___

## ___van Gerven et al., 2023___

```{r gsea-setup}
# Ranked gene list: shrunken LFC descending, named by Entrez ID
ranked_df <- res_selected_annot %>%
  filter(!is.na(ENTREZID), !is.na(log2FoldChange)) %>%
  arrange(desc(log2FoldChange)) %>%
  distinct(ENTREZID, .keep_all = TRUE)

ranked_genes <- setNames(ranked_df$log2FoldChange, ranked_df$ENTREZID)
cat("Genes in ranked list:", length(ranked_genes), "\n")

# Hallmark gene sets
hallmark_t2g <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, entrez_gene) %>%
  mutate(entrez_gene = as.character(entrez_gene))

# NB MES/ADRN gene sets from mounted resource path
mes_adrn_files <- list.files(params$markers, full.names = TRUE)
mes_adrn_t2g <- lapply(mes_adrn_files, function(f) {
  gs_name   <- tools::file_path_sans_ext(basename(f))
  gene_syms <- readLines(f)
  gene_syms <- gene_syms[nzchar(trimws(gene_syms))]
  entrez    <- bitr(gene_syms, fromType = "SYMBOL", toType = "ENTREZID",
                    OrgDb = org.Hs.eg.db)$ENTREZID
  data.frame(gs_name = gs_name, entrez_gene = as.character(entrez))
}) %>% bind_rows()

cat("MES/ADRN gene sets:", length(unique(mes_adrn_t2g$gs_name)), "\n")
cat("Hallmark gene sets:", length(unique(hallmark_t2g$gs_name)), "\n")
```

```{r gsea-mesenchymal, fig.height=5, fig.width=8}
mes_hallmark <- hallmark_t2g %>%
  filter(grepl("EPITHELIAL_MESENCHYMAL", gs_name))
mes_t2g <- bind_rows(mes_hallmark, mes_adrn_t2g)

# pvalueCutoff=0.25: MSigDB/Broad convention for GSEA (permutation-based, 0.05 is too strict)
# pvalueCutoff=1 to retrieve all results; filter by padj<0.25 for significance calls
gsea_mes <- GSEA(ranked_genes,
                 TERM2GENE     = mes_t2g,
                 pAdjustMethod = "fdr",
                 pvalueCutoff  = 1,
                 verbose       = FALSE)

sig_sets_mes <- gsea_mes@result$ID[gsea_mes@result$p.adjust < 0.25]

if (length(sig_sets_mes) > 0) {
  p_dot_mes <- dotplot(gsea_mes, showCategory = sig_sets_mes,
                       title = "GSEA — mesenchymal signatures (ATRXdel vs ATRXwt)")
  print(p_dot_mes)
  pdf(paste0(params$out, "NK_project/figures/GSEA_mesenchymal_ATRXdel_dotplot.pdf"),
      height = 5, width = 8)
  print(p_dot_mes); dev.off()
} else { cat("No mesenchymal gene sets significant at FDR<0.25.\n") }

# Running enrichment plots for ALL tested sets (shows NES + p even if not significant)
for (gs in gsea_mes@result$ID) {
  gs_padj <- gsea_mes@result$p.adjust[gsea_mes@result$ID == gs]
  gs_label <- ifelse(gs_padj < 0.25, gs, paste0(gs, " (ns, padj=", round(gs_padj, 3), ")"))
  p_es <- gseaplot2(gsea_mes, geneSetID = gs, title = gs_label, pvalue_table = TRUE)
  print(p_es)
  pdf(paste0(params$out, "NK_project/figures/GSEA_mes_", gs, ".pdf"), height = 6, width = 8)
  print(p_es); dev.off()
}
```

```{r gsea-inflammation, fig.height=5, fig.width=8}
inflam_t2g <- hallmark_t2g %>%
  filter(grepl("INFLAMMATORY|TNF_ALPHA|IL6_JAK|INTERFERON|IL2_STAT5", gs_name))

gsea_inflam <- GSEA(ranked_genes,
                    TERM2GENE     = inflam_t2g,
                    pAdjustMethod = "fdr",
                    pvalueCutoff  = 1,
                    verbose       = FALSE)

sig_sets_inflam <- gsea_inflam@result$ID[gsea_inflam@result$p.adjust < 0.25]

if (length(sig_sets_inflam) > 0) {
  p_dot_inflam <- dotplot(gsea_inflam, showCategory = sig_sets_inflam,
                          title = "GSEA — inflammatory signatures (ATRXdel vs ATRXwt)")
  print(p_dot_inflam)
  pdf(paste0(params$out, "NK_project/figures/GSEA_inflammation_ATRXdel_dotplot.pdf"),
      height = 5, width = 8)
  print(p_dot_inflam); dev.off()
} else { cat("No inflammatory gene sets significant at FDR<0.25.\n") }

# Running enrichment plots for ALL tested sets
for (gs in gsea_inflam@result$ID) {
  gs_padj <- gsea_inflam@result$p.adjust[gsea_inflam@result$ID == gs]
  gs_label <- ifelse(gs_padj < 0.25, gs, paste0(gs, " (ns, padj=", round(gs_padj, 3), ")"))
  p_es <- gseaplot2(gsea_inflam, geneSetID = gs, title = gs_label, pvalue_table = TRUE)
  print(p_es)
  pdf(paste0(params$out, "NK_project/figures/GSEA_inflam_", gs, ".pdf"), height = 6, width = 8)
  print(p_es); dev.off()
}
```

## ___Qadeer et al., 2019___

```{r sknfi-ranked-genes}
# Parse the malformed SKNFI FPKM file.
# Strategy: for each line, extract the gene name (first token) and then extract
# all decimal numbers via regex — this handles merged values with trailing dots
# (e.g. "1.23.4.56" → matches "1.23" then "4.56") without any whitespace assumptions.
sknfi_lines <- readLines(
  "/home/rstudio/mnt_data/nb_tumors/qadeer2019.bulkRNA.bulkChIP/processed_data/GSE100148/GSE100148_SKNFI_MS2_IFF_FPKM.txt"
)
# Strip non-printable characters (e.g. \f form feeds) and drop blank lines
sknfi_lines <- gsub("[^[:print:]]", "", sknfi_lines)
sknfi_lines <- sknfi_lines[nzchar(trimws(sknfi_lines))]

parse_sknfi_line <- function(l) {
  l <- trimws(l)
  gene <- strsplit(l, "\\s+")[[1]][1]
  gene <- gsub("[^A-Za-z0-9_\\-\\.]", "", gene)   # remove any residual non-word chars
  rest <- sub("^\\S+\\s+", "", l)
  # Extract all decimal or integer numbers; stray dots between merged values are skipped
  nums <- as.numeric(regmatches(rest, gregexpr("\\d+\\.\\d+|\\d+", rest))[[1]])
  if (nzchar(gene) && length(nums) >= 4) c(gene, nums[1:4]) else NULL
}

sknfi_parsed <- Filter(Negate(is.null), lapply(sknfi_lines[-1], parse_sknfi_line))

sknfi_df <- data.frame(
  gene   = sapply(sknfi_parsed, `[`, 1),
  MS2_R1 = as.numeric(sapply(sknfi_parsed, `[`, 2)),
  MS2_R2 = as.numeric(sapply(sknfi_parsed, `[`, 3)),
  IFF_R1 = as.numeric(sapply(sknfi_parsed, `[`, 4)),
  IFF_R2 = as.numeric(sapply(sknfi_parsed, `[`, 5)),
  stringsAsFactors = FALSE
)
sknfi_df <- sknfi_df[complete.cases(sknfi_df), ]
cat("Genes parsed from SKNFI file:", nrow(sknfi_df), "\n")

# log2 fold change (ATRXdel/IFF vs ATRXwt/MS2); pseudocount avoids log(0)
sknfi_df$mean_wt  <- (sknfi_df$MS2_R1 + sknfi_df$MS2_R2) / 2
sknfi_df$mean_del <- (sknfi_df$IFF_R1 + sknfi_df$IFF_R2) / 2
sknfi_df$log2FC   <- log2(sknfi_df$mean_del + 0.1) - log2(sknfi_df$mean_wt + 0.1)

# Convert gene symbols → Entrez IDs for GSEA
gene_map_sknfi   <- bitr(sknfi_df$gene, fromType = "SYMBOL", toType = "ENTREZID",
                          OrgDb = org.Hs.eg.db)
sknfi_df_mapped  <- merge(sknfi_df, gene_map_sknfi, by.x = "gene", by.y = "SYMBOL")
sknfi_df_mapped  <- sknfi_df_mapped[!duplicated(sknfi_df_mapped$ENTREZID), ]

ranked_sknfi <- sort(setNames(sknfi_df_mapped$log2FC, sknfi_df_mapped$ENTREZID),
                     decreasing = TRUE)
cat("SKNFI ranked gene list:", length(ranked_sknfi), "genes\n")

# ALCAM check
alcam_entrez <- gene_map_sknfi$ENTREZID[gene_map_sknfi$SYMBOL == "ALCAM"]
cat("ALCAM log2FC in SKNFI:", ranked_sknfi[alcam_entrez], "\n")
```

```{r gsea-sknfi-mesenchymal, fig.height=5, fig.width=8}
gsea_mes_sknfi <- GSEA(ranked_sknfi,
                       TERM2GENE     = mes_t2g,
                       pAdjustMethod = "fdr",
                       pvalueCutoff  = 1,
                       verbose       = FALSE)

sig_sets_mes_sknfi <- gsea_mes_sknfi@result$ID[gsea_mes_sknfi@result$p.adjust < 0.25]
if (length(sig_sets_mes_sknfi) > 0) {
  print(dotplot(gsea_mes_sknfi, showCategory = sig_sets_mes_sknfi,
                title = "GSEA mesenchymal — SKNFI (ATRXdel vs ATRXwt)"))
} else { cat("No mesenchymal gene sets significant at FDR<0.25 in SKNFI.\n") }

for (gs in gsea_mes_sknfi@result$ID) {
  gs_padj <- gsea_mes_sknfi@result$p.adjust[gsea_mes_sknfi@result$ID == gs]
  gs_label <- ifelse(gs_padj < 0.25, gs, paste0(gs, " (ns, padj=", round(gs_padj, 3), ")"))
  p_es <- gseaplot2(gsea_mes_sknfi, geneSetID = gs, title = gs_label, pvalue_table = TRUE)
  print(p_es)
  pdf(paste0(params$out, "NK_project/figures/GSEA_SKNFI_mes_", gs, ".pdf"), height = 6, width = 8)
  print(p_es); dev.off()
}
```

```{r gsea-sknfi-inflammation, fig.height=5, fig.width=8}
gsea_inflam_sknfi <- GSEA(ranked_sknfi,
                          TERM2GENE     = inflam_t2g,
                          pAdjustMethod = "fdr",
                          pvalueCutoff  = 1,
                          verbose       = FALSE)

sig_sets_inflam_sknfi <- gsea_inflam_sknfi@result$ID[gsea_inflam_sknfi@result$p.adjust < 0.25]
if (length(sig_sets_inflam_sknfi) > 0) {
  print(dotplot(gsea_inflam_sknfi, showCategory = sig_sets_inflam_sknfi,
                title = "GSEA inflammation — SKNFI (ATRXdel vs ATRXwt)"))
} else { cat("No inflammatory gene sets significant at FDR<0.25 in SKNFI.\n") }

for (gs in gsea_inflam_sknfi@result$ID) {
  gs_padj <- gsea_inflam_sknfi@result$p.adjust[gsea_inflam_sknfi@result$ID == gs]
  gs_label <- ifelse(gs_padj < 0.25, gs, paste0(gs, " (ns, padj=", round(gs_padj, 3), ")"))
  p_es <- gseaplot2(gsea_inflam_sknfi, geneSetID = gs, title = gs_label, pvalue_table = TRUE)
  print(p_es)
  pdf(paste0(params$out, "NK_project/figures/GSEA_SKNFI_inflam_", gs, ".pdf"),
      height = 6, width = 8)
  print(p_es); dev.off()
}
```

# ___Bulk ATAC-seq & ChIP-seq: ATRX binding, histone marks and chromatin accessibility at ALCAM gene locus___


```{r gviz-setup}
library(Gviz)
library(rtracklayer)
library(GenomicFeatures)
library(txdbmaker)
library(GenomeInfoDbData)

# ALCAM locus (based on genome version hg38 and gencode v43 gene annotations): chr3:105,366,909-105,576,900
# Plot window with +/- 80 kb flanking
ALCAM_CHR   <- "chr3"
ALCAM_START <- 105280000
ALCAM_END   <- 105640000

alcam_region <- GRanges(seqnames = ALCAM_CHR,
                         ranges   = IRanges(ALCAM_START, ALCAM_END))

# Helper: import a bigWig over the ALCAM region
import_bw <- function(path, region = alcam_region) {
  import(path, which = region, as = "GRanges")
}

# Helper: average multiple bigWig replicates → single GRanges with mean score
avg_bw <- function(paths, region = alcam_region) {
  rle_list <- lapply(paths, function(f) {
    bw <- import(f, which = region, as = "RleList")
    bw[[ALCAM_CHR]][ALCAM_START:ALCAM_END]
  })
  avg_rle <- Reduce("+", rle_list) / length(rle_list)
  GRanges(seqnames = ALCAM_CHR,
          ranges   = IRanges(start = ALCAM_START:ALCAM_END, width = 1),
          score    = as.numeric(avg_rle))
}

# Gene model from GTF
txdb_alcam <- suppressMessages(
  txdbmaker::makeTxDbFromGFF(
    "/home/rstudio/mnt_resources/GTF/gencode.v43.annotation.gtf",
    format    = "gtf",
    organism  = "Homo sapiens",
    chrominfo = Seqinfo(genome = "hg38")
  )
)

# ENCODE cCREs
# read.table + manual GRanges construction avoids rtracklayer parse error
ccre_df <- read.table(
  "/home/rstudio/mnt_resources/regulatory_regions/GRCh38-cCREs.bed",
  header    = FALSE, sep = "\t",
  col.names = c("chr", "start", "end", "accession", "accession2", "type"),
  colClasses = c("character", "integer", "integer",
                 "character", "character", "character")
)
# BED coordinates are 0-based half-open → convert to 1-based closed for GRanges
ccre_gr <- GRanges(
  seqnames = ccre_df$chr,
  ranges   = IRanges(start = ccre_df$start + 1L, end = ccre_df$end),
  accession = ccre_df$accession,
  type      = ccre_df$type
)
ccre_alcam <- subsetByOverlaps(ccre_gr, alcam_region)
cat("cCREs overlapping ALCAM window:", length(ccre_alcam), "\n")
print(as.data.frame(ccre_alcam)[, c("accession", "type")])

# EPD (Eukaryotic Promoter Database) — experimentally validated ALCAM promoters
epd_df    <- read.table(
  "/home/rstudio/mnt_resources/regulatory_regions/homo_sapiens/human_EPDnew_promoters_hg38.bed",
  header       = FALSE, sep = "\t",
  comment.char = "#",   # skip the #chrom header line
  col.names    = c("chr", "start", "end", "name", "score", "strand", "thickStart", "thickEnd"),
  colClasses   = c("character", "integer", "integer", "character",
                   "numeric",   "character", "integer",   "integer")
)
epd_alcam_df <- epd_df[grep("^ALCAM", epd_df$name), ]
epd_alcam    <- GRanges(
  seqnames = epd_alcam_df$chr,
  ranges   = IRanges(start = epd_alcam_df$start + 1L, end = epd_alcam_df$end),
  strand   = epd_alcam_df$strand,
  name     = epd_alcam_df$name
)
cat("EPD promoters for ALCAM:\n")
print(as.data.frame(epd_alcam)[, c("seqnames", "start", "end", "name")])
```

```{r gviz-annotation-tracks}
axis_track <- GenomeAxisTrack()
ideo_track <- IdeogramTrack(genome = "hg38", chromosome = ALCAM_CHR)

gene_track <- GeneRegionTrack(
  txdb_alcam,
  chromosome  = ALCAM_CHR,
  start       = ALCAM_START,
  end         = ALCAM_END,
  name        = "ALCAM",
  showId      = TRUE,
  geneSymbols = TRUE,
  fill        = "#333333",
  col         = NA,
  fontsize    = 10
)

# Color cCREs by type: PLS/promoter = red, ELS/enhancer = orange, CTCF-only = gray
ccre_cols <- ifelse(grepl("PLS|DNase-H3K4me3", ccre_alcam$type), "#e63946",
             ifelse(grepl("ELS|pELS|dELS",      ccre_alcam$type), "#E07B39",
                    "#aaaaaa"))
ccre_track <- AnnotationTrack(
  ccre_alcam,
  name     = "ENCODE cCREs",
  fill     = ccre_cols,
  col      = NA,
  fontsize = 8,
  stacking = "dense"
)

# EPD promoter track — marks exact TSS positions validated by EPD
epd_track <- AnnotationTrack(
  epd_alcam,
  name     = "EPD promoters",
  id       = epd_alcam$name,
  showFeatureId = TRUE,
  fill     = "#c1121f",
  col      = NA,
  fontsize = 8,
  stacking = "dense"
)
```

```{r gviz-atrx-tracks}
CHIP_DIR <- "/home/rstudio/mnt_data/nb_tumors/qadeer2019.bulkRNA.bulkChIP/processed_data/GSE100148/GSE100148_RAW/"

chla90_atrx_gr <- import_bw(file.path(CHIP_DIR, "GSM3389350_CHLA90_ATRX.bigWig"))
sknfi_atrx_gr  <- import_bw(file.path(CHIP_DIR, "GSM3389345_SKNFI_ATRX.bigWig"))

track_chla90_atrx <- DataTrack(chla90_atrx_gr, name = "CHLA90 ATRX",
                                type = "histogram", fill.histogram = "#4f4e92",
                                col.histogram = NA, fontsize = 10)
track_sknfi_atrx  <- DataTrack(sknfi_atrx_gr,  name = "SKNFI ATRX",
                                type = "histogram", fill.histogram = "#7b5ea7",
                                col.histogram = NA, fontsize = 10)
```

```{r gviz-atac-tracks}
ATAC_DIR <- "/home/rstudio/mnt_data/nb_tumors/lorenzi2026.bulkATAC/GSE312501/"

p53_avg <- avg_bw(file.path(ATAC_DIR, c(
  "GSM9347420_p53_DMSO_REP1.mLb.clN.bigWig",
  "GSM9347424_p53_DMSO_REP2.mLb.clN.bigWig",
  "GSM9347428_p53_DMSO_REP3.mLb.clN.bigWig"
)))
e6_avg <- avg_bw(file.path(ATAC_DIR, c(
  "GSM9347422_E6_DMSO_REP1.mLb.clN.bigWig",
  "GSM9347426_E6_DMSO_REP2.mLb.clN.bigWig",
  "GSM9347430_E6_DMSO_REP3.mLb.clN.bigWig"
)))

atac_ylim <- c(0, max(c(score(p53_avg), score(e6_avg)), na.rm = TRUE) * 1.05)

track_atac_p53 <- DataTrack(p53_avg, name = "ATAC p53\n(ATRXwt)",
                             type = "histogram", fill.histogram = "#7EA3AC",
                             col.histogram = NA, ylim = atac_ylim, fontsize = 10)
track_atac_e6  <- DataTrack(e6_avg,  name = "ATAC E6\n(ATRXdel)",
                             type = "histogram", fill.histogram = "#4f4e92",
                             col.histogram = NA, ylim = atac_ylim, fontsize = 10)
```

```{r gviz-histone-tracks}
chla90_h3k9ac_gr   <- import_bw(file.path(CHIP_DIR, "GSM2672020_CHLA90_H3K9ac.bigWig"))
chla90_h3k27me3_gr <- import_bw(file.path(CHIP_DIR, "GSM2672018_CHLA90_H3K27me3.bigWig"))
chla90_h3k9me3_gr  <- import_bw(file.path(CHIP_DIR, "GSM2672019_CHLA90_H3K9me3.bigWig"))

track_h3k9ac   <- DataTrack(chla90_h3k9ac_gr,   name = "CHLA90\nH3K9ac",
                             type = "histogram", fill.histogram = "#2a9d8f",
                             col.histogram = NA, fontsize = 10)
track_h3k27me3 <- DataTrack(chla90_h3k27me3_gr, name = "CHLA90\nH3K27me3",
                             type = "histogram", fill.histogram = "#e63946",
                             col.histogram = NA, fontsize = 10)
track_h3k9me3  <- DataTrack(chla90_h3k9me3_gr,  name = "CHLA90\nH3K9me3",
                             type = "histogram", fill.histogram = "#8d8d8d",
                             col.histogram = NA, fontsize = 10)
```

```{r gviz-ratio-track}
# Log2(ATRXdel / ATRXwt) ratio track for ATAC-seq (E6 / p53)
# Pseudocount avoids log(0); positive = more open in ATRXdel, negative = more open in ATRXwt
p53_scores <- score(p53_avg)
e6_scores  <- score(e6_avg)
ratio_scores <- log2((e6_scores + 0.01) / (p53_scores + 0.01))

ratio_gr <- GRanges(seqnames = ALCAM_CHR,
                     ranges   = IRanges(start = ALCAM_START:ALCAM_END, width = 1),
                     score    = ratio_scores)

track_atac_ratio <- DataTrack(
  ratio_gr,
  name           = "ATAC\nlog2(del/WT)",
  type           = "histogram",
  fill.histogram = "#E07B39",
  col.histogram  = NA,
  fontsize       = 10,
  # Symmetric y-axis around 0
  ylim           = c(-max(abs(ratio_scores), na.rm = TRUE),
                      max(abs(ratio_scores), na.rm = TRUE))
)
```

## ___Lorenzi et al., 2025___

*Isogenic NB cell lines: SK-N-SH*

```{r atac-profile-r, fig.height=5, fig.width=7}
# R-based equivalent of deepTools computeMatrix + plotProfile
# TSS-centered profile around ALCAM_1 EPD promoter (TSS at 105,366,908)
ALCAM_TSS  <- 105366908L
PROF_WIN   <- 5000L   # ±5kb
BIN_SIZE   <- 50L

get_binned_profile <- function(bw_paths, condition, chr = ALCAM_CHR,
                                tss = ALCAM_TSS, window = PROF_WIN,
                                bin_size = BIN_SIZE) {
  region     <- GRanges(chr, IRanges(tss - window, tss + window))
  bin_starts <- seq(tss - window, tss + window - bin_size + 1L, by = bin_size)
  bins       <- GRanges(chr, IRanges(start = bin_starts, width = bin_size))
  seqlevels(bins) <- chr  # ensure only chr3 is in seqlevels

  rep_mat <- sapply(bw_paths, function(f) {
    cov     <- import(f, which = region, as = "RleList")
    cov_chr <- cov[chr]  # subset RleList to chr3 only → names(cov_chr) == seqlevels(bins)
    ba      <- binnedAverage(bins, cov_chr, "avg")
    ba$avg
  })

  data.frame(
    position  = bin_starts - tss,
    signal    = rowMeans(rep_mat, na.rm = TRUE),
    condition = condition
  )
}

p53_profile <- get_binned_profile(
  file.path(ATAC_DIR, c("GSM9347420_p53_DMSO_REP1.mLb.clN.bigWig",
                          "GSM9347424_p53_DMSO_REP2.mLb.clN.bigWig",
                          "GSM9347428_p53_DMSO_REP3.mLb.clN.bigWig")),
  condition = "p53 (ATRXwt)"
)
e6_profile <- get_binned_profile(
  file.path(ATAC_DIR, c("GSM9347422_E6_DMSO_REP1.mLb.clN.bigWig",
                          "GSM9347426_E6_DMSO_REP2.mLb.clN.bigWig",
                          "GSM9347430_E6_DMSO_REP3.mLb.clN.bigWig")),
  condition = "E6 (ATRXdel)"
)

profile_df <- rbind(p53_profile, e6_profile)
profile_df$condition <- factor(profile_df$condition,
                                levels = c("p53 (ATRXwt)", "E6 (ATRXdel)"))

ATAC_PROFILE_COLORS <- c("p53 (ATRXwt)" = "#FAA21B", "E6 (ATRXdel)" = "#4f4e92")

atac_profile_plot <- ggplot(profile_df, aes(x = position / 1000, y = signal,
                                             color = condition)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  scale_color_manual(values = ATAC_PROFILE_COLORS, name = "") +
  scale_x_continuous(labels = function(x) paste0(x, " kb")) +
  labs(
    title    = "ATAC-seq signal around ALCAM TSS",
    subtitle = "EPD promoter ALCAM_1 (chr3:105,366,908)",
    x        = "Distance from TSS",
    y        = "Mean ATAC-seq signal (bigWig)"
  ) +
  theme_pubr(base_size = 12) +
  theme(legend.position = "top",
        plot.title      = element_text(face = "bold", size = 13))

atac_profile_plot

pdf(paste0(params$out, "NK_project/figures/ALCAM_ATAC_TSS_profile.pdf"),
    height = 5, width = 7)
atac_profile_plot
dev.off()
```

```{r atrx-profile-r, fig.height=5, fig.width=7}
# ATRX ChIP-seq profile: CHLA90 vs SKNFI around ALCAM TSS
chla90_atrx_profile <- get_binned_profile(
  file.path(CHIP_DIR, "GSM3389350_CHLA90_ATRX.bigWig"),
  condition = "CHLA90"
)
sknfi_atrx_profile <- get_binned_profile(
  file.path(CHIP_DIR, "GSM3389345_SKNFI_ATRX.bigWig"),
  condition = "SKNFI"
)

atrx_profile_df <- rbind(chla90_atrx_profile, sknfi_atrx_profile)
atrx_profile_df$condition <- factor(atrx_profile_df$condition,
                                     levels = c("CHLA90", "SKNFI"))

atrx_profile_plot <- ggplot(atrx_profile_df,
                              aes(x = position / 1000, y = signal, color = condition)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  scale_color_manual(values = c("CHLA90" = "#4f4e92", "SKNFI" = "#FAA21B"), name = "") +
  scale_x_continuous(labels = function(x) paste0(x, " kb")) +
  labs(
    title    = "ATRX ChIP-seq signal around ALCAM TSS",
    subtitle = "Qadeer et al., 2019",
    x        = "Distance from TSS",
    y        = "ATRX ChIP-seq signal (bigWig)"
  ) +
  theme_pubr(base_size = 12) +
  theme(legend.position = "top",
        plot.title      = element_text(face = "bold", size = 13))

atrx_profile_plot

pdf(paste0(params$out, "NK_project/figures/ALCAM_ATRX_ChIP_TSS_profile.pdf"),
    height = 5, width = 7)
atrx_profile_plot
dev.off()
```

## ___Qadeer et al., 2019___

```{r histone-profile-r, fig.height=6, fig.width=7}
# Histone mark profiles: CHLA90 vs SKNFI for H3K9ac, H3K27me3, H3K9me3
histone_marks <- list(
  list(file = "GSM2672020_CHLA90_H3K9ac.bigWig",   cell = "CHLA90", mark = "H3K9ac"),
  list(file = "GSM2672010_SKNFI_H3K9ac.bigWig",    cell = "SKNFI",  mark = "H3K9ac"),
  list(file = "GSM2672018_CHLA90_H3K27me3.bigWig",  cell = "CHLA90", mark = "H3K27me3"),
  list(file = "GSM2672008_SKNFI_H3K27me3.bigWig",   cell = "SKNFI",  mark = "H3K27me3"),
  list(file = "GSM2672019_CHLA90_H3K9me3.bigWig",   cell = "CHLA90", mark = "H3K9me3"),
  list(file = "GSM2672009_SKNFI_H3K9me3.bigWig",    cell = "SKNFI",  mark = "H3K9me3")
)

histone_profile_df <- do.call(rbind, lapply(histone_marks, function(m) {
  df       <- get_binned_profile(file.path(CHIP_DIR, m$file), condition = m$cell)
  df$mark  <- m$mark
  df
}))

histone_profile_df$mark      <- factor(histone_profile_df$mark,
                                        levels = c("H3K9ac", "H3K27me3", "H3K9me3"))
histone_profile_df$condition <- factor(histone_profile_df$condition,
                                        levels = c("CHLA90", "SKNFI"))

MARK_COLORS <- c("H3K9ac" = "#2a9d8f", "H3K27me3" = "#e63946", "H3K9me3" = "#8d8d8d")
CELL_LTYPE  <- c("CHLA90" = "solid",   "SKNFI"    = "dashed")

histone_profile_plot <- ggplot(histone_profile_df,
                                aes(x = position / 1000, y = signal,
                                    color = mark, linetype = condition)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.4) +
  scale_color_manual(values = MARK_COLORS, name = "Histone mark") +
  scale_linetype_manual(values = CELL_LTYPE, name = "Cell line") +
  scale_x_continuous(labels = function(x) paste0(x, " kb")) +
  facet_wrap(~ mark, ncol = 1, scales = "free_y") +
  labs(
    title    = "Histone mark ChIP-seq signal around ALCAM TSS",
    subtitle = "Qadeer et al., 2019 — CHLA90 (solid) vs SKNFI (dashed)",
    x        = "Distance from TSS",
    y        = "ChIP-seq signal (bigWig)"
  ) +
  theme_pubr(base_size = 12) +
  theme(legend.position = "right",
        strip.text      = element_text(face = "bold"),
        plot.title      = element_text(face = "bold", size = 13))

histone_profile_plot

pdf(paste0(params$out, "NK_project/figures/ALCAM_histone_TSS_profile.pdf"),
    height = 6, width = 7)
histone_profile_plot
dev.off()
```

```{r gviz-plot, fig.height=16, fig.width=10}
all_tracks <- list(
  ideo_track, axis_track, gene_track, epd_track, ccre_track,
  track_chla90_atrx, track_sknfi_atrx,
  track_atac_p53, track_atac_e6, track_atac_ratio,
  track_h3k9ac, track_h3k27me3, track_h3k9me3
)
track_sizes <- c(0.5, 0.5, 1.5, 0.3, 0.5, 1.5, 1.5, 1.5, 1.5, 1.0, 1.5, 1.5, 1.5)

plotTracks(all_tracks, from = ALCAM_START, to = ALCAM_END, sizes = track_sizes,
           background.panel = "#FFFFFF", background.title = "#F0F0F0",
           col.title = "#333333", fontsize = 10, title.width = 1.5)

pdf(paste0(params$out, "NK_project/figures/ALCAM_genome_browser.pdf"),
    height = 14, width = 10)
plotTracks(all_tracks, from = ALCAM_START, to = ALCAM_END, sizes = track_sizes,
           background.panel = "#FFFFFF", background.title = "#F0F0F0",
           col.title = "#333333", fontsize = 10, title.width = 1.5)
dev.off()
```

```{r gviz-promoter-zoom, fig.height=16, fig.width=8}
# Zoom into ALCAM promoter region (±30 kb around EPD TSS at chr3:105,366,908)
plotTracks(all_tracks, from = 105336000, to = 105400000, sizes = track_sizes,
           background.panel = "#FFFFFF", background.title = "#F0F0F0",
           col.title = "#333333", fontsize = 10, title.width = 1.5)

pdf(paste0(params$out, "NK_project/figures/ALCAM_genome_browser_promoter_zoom.pdf"),
    height = 16, width = 8)
plotTracks(all_tracks, from = 105336000, to = 105400000, sizes = track_sizes,
           background.panel = "#FFFFFF", background.title = "#F0F0F0",
           col.title = "#333333", fontsize = 10, title.width = 1.5)
dev.off()
```

# ___(alternative method) deepTools bigWig comparison in BASH___

```{bash deeptools-install, eval=FALSE}
# Install deepTools into the conda environment (run once)
conda install -c bioconda deeptools -y
```

```{bash deeptools-compare, eval=FALSE}
ATAC_DIR="/home/rstudio/mnt_data/nb_tumors/lorenzi2026.bulkATAC/GSE312501"
CHIP_DIR="/home/rstudio/mnt_data/nb_tumors/qadeer2019.bulkRNA.bulkChIP/processed_data/GSE100148/GSE100148_RAW"
OUT_DIR="/home/rstudio/mnt_out/NK_project/figures/deeptools"
mkdir -p "$OUT_DIR"

ALCAM_REGION="chr3:105280000:105640000"
ALCAM_TSS="chr3 105366908"  # for reference-point matrix

#  ATAC-seq: log2 ratio bigWig (p53 ATRXwt / E6 ATRXdel, merged replicates) 
bigwigMerge \
  "$ATAC_DIR/GSM9347420_p53_DMSO_REP1.mLb.clN.bigWig" \
  "$ATAC_DIR/GSM9347424_p53_DMSO_REP2.mLb.clN.bigWig" \
  "$ATAC_DIR/GSM9347428_p53_DMSO_REP3.mLb.clN.bigWig" \
  -o "$OUT_DIR/p53_merged.bedGraph" 2>/dev/null &&
  bedGraphToBigWig "$OUT_DIR/p53_merged.bedGraph" <(fetchChromSizes hg38) \
    "$OUT_DIR/p53_merged.bw" || true

bigwigCompare \
  --bigwig1 "$ATAC_DIR/GSM9347420_p53_DMSO_REP1.mLb.clN.bigWig" \
  --bigwig2 "$ATAC_DIR/GSM9347422_E6_DMSO_REP1.mLb.clN.bigWig" \
  --operation log2 --pseudocount 1 \
  --binSize 50 \
  -o "$OUT_DIR/ATAC_log2_p53_vs_E6.bw"

#  Profile plot: ATAC signal ±5kb around ALCAM TSS 
computeMatrix reference-point \
  --referencePoint TSS \
  --regionsFileName <(echo -e "chr3\t105361908\t105371908\tALCAM\t0\t+") \
  --scoreFileName \
    "$ATAC_DIR/GSM9347420_p53_DMSO_REP1.mLb.clN.bigWig" \
    "$ATAC_DIR/GSM9347424_p53_DMSO_REP2.mLb.clN.bigWig" \
    "$ATAC_DIR/GSM9347428_p53_DMSO_REP3.mLb.clN.bigWig" \
    "$ATAC_DIR/GSM9347422_E6_DMSO_REP1.mLb.clN.bigWig" \
    "$ATAC_DIR/GSM9347426_E6_DMSO_REP2.mLb.clN.bigWig" \
    "$ATAC_DIR/GSM9347430_E6_DMSO_REP3.mLb.clN.bigWig" \
  --beforeRegionStartLength 5000 \
  --afterRegionStartLength 5000 \
  --binSize 50 \
  --samplesLabel p53_R1 p53_R2 p53_R3 E6_R1 E6_R2 E6_R3 \
  -o "$OUT_DIR/ATAC_ALCAM_matrix.gz"

plotProfile \
  --matrixFile "$OUT_DIR/ATAC_ALCAM_matrix.gz" \
  --outFileName "$OUT_DIR/ATAC_ALCAM_profile.pdf" \
  --samplesLabel "p53 (ATRXwt)" "p53 (ATRXwt)" "p53 (ATRXwt)" \
                 "E6 (ATRXdel)" "E6 (ATRXdel)" "E6 (ATRXdel)" \
  --colors "#7EA3AC" "#7EA3AC" "#7EA3AC" "#4f4e92" "#4f4e92" "#4f4e92" \
  --plotTitle "ATAC-seq at ALCAM TSS" \
  --averageType mean \
  --perGroup \
  --plotHeight 5 --plotWidth 7

#  ChIP-seq profile: ATRX + histone marks 
computeMatrix reference-point \
  --referencePoint TSS \
  --regionsFileName <(echo -e "chr3\t105361908\t105371908\tALCAM\t0\t+") \
  --scoreFileName \
    "$CHIP_DIR/GSM3389350_CHLA90_ATRX.bigWig" \
    "$CHIP_DIR/GSM3389345_SKNFI_ATRX.bigWig" \
    "$CHIP_DIR/GSM2672020_CHLA90_H3K9ac.bigWig" \
    "$CHIP_DIR/GSM2672018_CHLA90_H3K27me3.bigWig" \
    "$CHIP_DIR/GSM2672019_CHLA90_H3K9me3.bigWig" \
  --beforeRegionStartLength 5000 \
  --afterRegionStartLength 5000 \
  --binSize 50 \
  -o "$OUT_DIR/ChIP_ALCAM_matrix.gz"

plotProfile \
  --matrixFile "$OUT_DIR/ChIP_ALCAM_matrix.gz" \
  --outFileName "$OUT_DIR/ChIP_ALCAM_profile.pdf" \
  --samplesLabel "CHLA90 ATRX" "SKNFI ATRX" "CHLA90 H3K9ac" "CHLA90 H3K27me3" "CHLA90 H3K9me3" \
  --colors "#4f4e92" "#7b5ea7" "#2a9d8f" "#e63946" "#8d8d8d" \
  --plotTitle "ChIP-seq at ALCAM TSS" \
  --plotHeight 5 --plotWidth 7
```





