---
title: "NK project: DEG analysis - hNK_Bm2 vs. hNK_Bm3 per genetic group"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_NK: "/home/rstudio/mnt_out/NK_project/Rds/ALL_NK_cluster_withNewMetadata.Rds"
  out: "/home/rstudio/mnt_out/NK_project/"
  GO_db: "/home/rstudio/mnt_resources/GSEA_GeneSetEnrichments/"
---

PROJECT DESCRIPTION

Do hNK_Bm2 and hNK_Bm3 differ from each other in terms of gene expression, and does the
nature or extent of their difference change across genetic contexts (control, ATRXdel,
MYCNamp, ATRXwtMYCNwt)?

Strategy: for each genetic group separately, run pseudobulk DEG between hNK_Bm2 and hNK_Bm3
using DElegate, then overlay significant genes across groups with Venn diagrams to identify
context-shared vs. context-specific differences.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE,
                      error = FALSE,
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r include=FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
```

# ___Load data___

```{r include=FALSE}
suppressPackageStartupMessages(library('pheatmap'))
suppressPackageStartupMessages(library("DOSE"))
suppressPackageStartupMessages(library("enrichplot"))
suppressPackageStartupMessages(library("ggupset"))
suppressPackageStartupMessages(library("EnhancedVolcano"))
suppressPackageStartupMessages(library("ggVennDiagram"))
suppressPackageStartupMessages(library("ggbeeswarm"))
```

```{r include=FALSE}
ALL_NK <- readRDS(file = params$input_NK)
```

```{r}
counts_matrix <- GetAssayData(ALL_NK, slot = "counts")

hist(Matrix::rowSums(counts_matrix), breaks = 100000000, xlim = (c(0, 100)))
abline(v = 2)
hist(log(Matrix::rowSums(counts_matrix)), breaks = 1000)
abline(v = 3)

geneUniverse <- rownames(counts_matrix)[Matrix::rowSums(counts_matrix) > 1]
length(geneUniverse)

geneUniverse_entrezIDs <- map_gene_to_entrez(geneUniverse)

hallmark_genes_H <- msigdbr(species = "human", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)
```

Publication theme and shared helpers.

```{r}
# Publication-quality ggplot2 theme
pub_theme <- theme_classic(base_size = 10) +
  theme(
    axis.line        = element_line(linewidth = 0.4, color = "black"),
    axis.ticks       = element_line(linewidth = 0.4, color = "black"),
    axis.text        = element_text(size = 8,  color = "black"),
    axis.title       = element_text(size = 9,  color = "black"),
    plot.title       = element_text(size = 10, face = "bold"),
    plot.subtitle    = element_text(size = 8),
    legend.key.size  = unit(3,  "mm"),
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 9),
    strip.background = element_blank(),
    strip.text       = element_text(size = 9)
  )
theme_set(pub_theme)

# Cluster and group colors (from 00_NK_paper_setup.R)
COL_BM2   <- COLOR_CODE_v2["hNK_Bm2"]   # "#679F8F"
COL_BM3   <- COLOR_CODE_v2["hNK_Bm3"]   # "#BF4E4E"
COL_CLUST <- COLOR_CODE_v2[c("hNK_Bm2", "hNK_Bm3")]

# Directional volcano colors: up in Bm3 = COL_BM3, up in Bm2 = COL_BM2
volcano_colors <- function(df, padj_cut = 0.05, fc_cut = 0.5) {
  col_vec <- rep("gray70", nrow(df))
  col_vec[!is.na(df$padj) & df$padj < padj_cut & df$log_fc >  fc_cut] <- COL_BM3
  col_vec[!is.na(df$padj) & df$padj < padj_cut & df$log_fc < -fc_cut] <- COL_BM2
  names(col_vec) <- rownames(df)
  col_vec
}

# Enrichment helpers
run_enrichment <- function(gene_symbols,
                           type = c("BP", "MF", "CC", "KEGG", "HALLMARK"),
                           term2gene = NULL,
                           label = "") {
  type <- match.arg(type)
  if (length(gene_symbols) == 0) {
    message(label, ": no input genes - skipping"); return(invisible(NULL))
  }
  entrez_ids <- na.omit(map_gene_to_entrez(gene_symbols))
  if (length(entrez_ids) == 0) {
    message(label, ": no valid Entrez IDs - skipping"); return(invisible(NULL))
  }
  tryCatch(
    {
      if (type %in% c("BP", "MF", "CC")) {
        enrichGO(gene = entrez_ids, ont = type, OrgDb = "org.Hs.eg.db",
                 readable = TRUE, pAdjustMethod = "bonferroni",
                 pvalueCutoff = 0.05)
      } else if (type == "KEGG") {
        enrichKEGG(gene = entrez_ids, organism = "hsa", keyType = "kegg",
                   pAdjustMethod = "bonferroni", pvalueCutoff = 0.05)
      } else if (type == "HALLMARK") {
        enricher(gene = entrez_ids, TERM2GENE = term2gene,
                 pAdjustMethod = "bonferroni", pvalueCutoff = 0.05)
      }
    },
    error = function(e) {
      message(label, " (", type, ") error: ", conditionMessage(e)); NULL
    }
  )
}

# Show enrichment dotplot inline; save PDF when out_path provided
show_enrichment <- function(result, label = "", out_path = NULL,
                            out_w = 5, out_h = 4.5) {
  if (is.null(result) || nrow(result@result) == 0) {
    print(paste0("0 enriched terms - ", label)); return(invisible(NULL))
  }
  signif_rows <- result@result[result@result$p.adjust < 0.05, ]
  if (nrow(signif_rows) == 0) {
    print(paste0("0 terms with bonferroni p.adjust < 0.05 - ", label))
    return(invisible(NULL))
  }
  p <- dotplot(result, showCategory = 20) +
    ggtitle(label) + pub_theme +
    theme(axis.text.y = element_text(size = 7))
  print(p)
  if (!is.null(out_path)) {
    pdf(out_path, height = out_h, width = out_w, useDingbats = FALSE)
    print(p); dev.off()
  }
}

run_treeplot <- function(result, label = "", n_cluster = 5, n_words = 1,
                         show_cat = 15, out_path = NULL, out_w = 9, out_h = 4) {
  if (is.null(result) || nrow(result@result) == 0) {
    message("treeplot skipped (no enriched terms): ", label); return(invisible(NULL))
  }
  signif_rows <- result@result[result@result$p.adjust < 0.05, ]
  if (nrow(signif_rows) < 2) {
    message("treeplot skipped (< 2 significant terms): ", label); return(invisible(NULL))
  }
  p <- tryCatch(
    {
      enrichres2 <- pairwise_termsim(result)
      treeplot(enrichres2, showCategory = show_cat, color = "p.adjust",
               offset = 10, offset_tiplab = 0.5,
               nCluster = min(n_cluster, nrow(signif_rows)),
               nWords = n_words, split = "subcategory")
    },
    error = function(e) { message("treeplot error: ", conditionMessage(e)); NULL }
  )
  if (!is.null(p)) {
    print(p)
    if (!is.null(out_path)) {
      pdf(out_path, height = out_h, width = out_w, useDingbats = FALSE)
      print(p); dev.off()
    }
  }
}

# Top-DEG violin + boxplot + jitter
# top_n genes per direction (up in Bm3, up in Bm2) ranked by |log_fc|.
# Violins coloured by cluster; boxplot shows median/IQR; points are individual cells.
plot_top_deg_violins <- function(seurat_obj, deg_df,
                                 group_col = "predicted_NK_type_de.n350",
                                 top_n = 12, out_path = NULL,
                                 out_w = 9, out_h = 5) {
  if (nrow(deg_df) == 0) {
    message("no DEGs - skipping violin plot"); return(invisible(NULL))
  }
  top_genes <- deg_df %>%
    mutate(abs_lfc   = abs(log_fc),
           direction = ifelse(log_fc > 0, "up in hNK_Bm3", "up in hNK_Bm2")) %>%
    group_by(direction) %>%
    slice_max(abs_lfc, n = top_n) %>%
    ungroup() %>%
    arrange(desc(log_fc)) %>%
    mutate(feature = factor(feature, levels = unique(feature)))

  genes <- levels(top_genes$feature)
  genes <- intersect(genes, rownames(seurat_obj))
  if (length(genes) == 0) {
    message("no genes found in object - skipping violin plot"); return(invisible(NULL))
  }

  expr <- as.data.frame(
    t(as.matrix(GetAssayData(seurat_obj, slot = "data")[genes, , drop = FALSE]))
  )
  expr$cluster <- seurat_obj@meta.data[[group_col]]
  expr_long <- expr %>%
    pivot_longer(-cluster, names_to = "gene", values_to = "expr") %>%
    mutate(gene = factor(gene, levels = genes)) %>%
    left_join(top_genes %>% dplyr::select(feature, direction, log_fc),
              by = c("gene" = "feature"))

  p <- ggplot(expr_long, aes(x = cluster, y = expr,
                              fill = cluster, color = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.3, alpha = 0.6) +
    geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white",
                 color = "black", linewidth = 0.35) +
    geom_quasirandom(size = 0.25, alpha = 0.35, color = "black",
                     groupOnX = TRUE, bandwidth = 0.5) +
    facet_wrap(~ gene, nrow = 2, scales = "free_y") +
    scale_fill_manual(values  = COL_CLUST, guide = "none") +
    scale_color_manual(values = COL_CLUST, guide = "none") +
    labs(x = NULL, y = "log-normalised expression",
         subtitle = paste0("top ", top_n, " DEGs per direction (|log_fc| ranked)")) +
    theme(
      axis.text.x   = element_text(size = 7, angle = 30, hjust = 1),
      axis.text.y   = element_text(size = 7),
      strip.text    = element_text(size = 7, face = "italic"),
      panel.spacing = unit(2, "mm")
    )

  print(p)
  if (!is.null(out_path)) {
    pdf(out_path, height = out_h, width = out_w, useDingbats = FALSE)
    print(p); dev.off()
  }
}
```

# ___DEG: hNK_Bm2 vs. hNK_Bm3 per genetic group___

*Replicates:* patient ID (`rados_et_al_2025`).
*Comparison:* hNK_Bm2 (reference) vs. hNK_Bm3 - **positive log_fc = higher in hNK_Bm3**.
*Cutoffs:* padj < 0.05 & |log_fc| > 0.5.
*Volcano colors:* hNK_Bm3-high = `r COL_BM3`, hNK_Bm2-high = `r COL_BM2`.

```{r}
Idents(ALL_NK) <- ALL_NK@meta.data$predicted_NK_type_de.n350

hNK_Bm2_Bm3 <- subset(ALL_NK, idents = c("hNK_Bm2", "hNK_Bm3"), invert = FALSE)
hNK_Bm2_Bm3
table(hNK_Bm2_Bm3$predicted_NK_type_de.n350, hNK_Bm2_Bm3$group)
```

## ___1. control___

```{r}
Idents(hNK_Bm2_Bm3) <- hNK_Bm2_Bm3$group
hNK_Bm2_Bm3_ctrl <- subset(hNK_Bm2_Bm3, idents = "control")
table(hNK_Bm2_Bm3_ctrl$predicted_NK_type_de.n350)
```

```{r}
DEG_ctrl <- DElegate::findDE(hNK_Bm2_Bm3_ctrl,
                             group_column = "predicted_NK_type_de.n350",
                             order_results = TRUE, compare_is_ref = TRUE,
                             replicate_column = "rados_et_al_2025", compare = "hNK_Bm2")

DEG_ctrl_signif <- subset(DEG_ctrl, padj < 0.1)
DEG_ctrl_up     <- subset(DEG_ctrl_signif, log_fc >  0.5 & padj < 0.05)
DEG_ctrl_down   <- subset(DEG_ctrl_signif, log_fc < -0.5 & padj < 0.05)

cat("Total (padj<0.1):", nrow(DEG_ctrl_signif),
    "| up Bm3:", nrow(DEG_ctrl_up), "| up Bm2:", nrow(DEG_ctrl_down), "\n")
DEG_ctrl_signif

write.table(DEG_ctrl, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm2-vs-hNKBm3_control_2026-06-25.csv"))
```

### ___Volcano - control___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_ctrl) <- DEG_ctrl$feature

p_volcano_ctrl <- EnhancedVolcano(
  DEG_ctrl, lab = rownames(DEG_ctrl), x = "log_fc", y = "padj",
  title = "control", subtitle = "hNK_Bm2 (ref) vs. hNK_Bm3",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(DEG_ctrl$padj[DEG_ctrl$padj > 0], na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3, colCustom = volcano_colors(DEG_ctrl), colAlpha = 0.7,
  legendLabSize = 8, legendIconSize = 3, titleLabSize = 10, subtitleLabSize = 8,
  axisLabSize = 9, boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_ctrl

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_control_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_ctrl; dev.off()
```

### ___Top DEG expression - control___

```{r fig.height=5, fig.width=9}
plot_top_deg_violins(hNK_Bm2_Bm3_ctrl, deg_df = bind_rows(DEG_ctrl_up, DEG_ctrl_down),
  top_n = 12, out_path = paste0(params$out,
    "figures/DEG_hNKBm2-vs-hNKBm3_control_topDEG_violins.pdf"))
```

### ___Pathway enrichment - control (up in hNK_Bm3)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_ctrl_up$feature
show_enrichment(run_enrichment(genes, "BP",       label = "ctrl Bm3 up: GO-BP"), "ctrl Bm3 up: GO-BP")
show_enrichment(run_enrichment(genes, "MF",       label = "ctrl Bm3 up: GO-MF"), "ctrl Bm3 up: GO-MF")
show_enrichment(run_enrichment(genes, "CC",       label = "ctrl Bm3 up: GO-CC"), "ctrl Bm3 up: GO-CC")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ctrl Bm3 up: KEGG"),  "ctrl Bm3 up: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ctrl Bm3 up: HALLMARK",
                               term2gene = hallmark_genes_H), "ctrl Bm3 up: HALLMARK")
```

### ___Pathway enrichment - control (up in hNK_Bm2)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_ctrl_down$feature
show_enrichment(run_enrichment(genes, "BP",       label = "ctrl Bm2 up: GO-BP"), "ctrl Bm2 up: GO-BP")
show_enrichment(run_enrichment(genes, "MF",       label = "ctrl Bm2 up: GO-MF"), "ctrl Bm2 up: GO-MF")
show_enrichment(run_enrichment(genes, "CC",       label = "ctrl Bm2 up: GO-CC"), "ctrl Bm2 up: GO-CC")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ctrl Bm2 up: KEGG"),  "ctrl Bm2 up: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ctrl Bm2 up: HALLMARK",
                               term2gene = hallmark_genes_H), "ctrl Bm2 up: HALLMARK")
```


## ___2. ATRXdel___

```{r}
Idents(hNK_Bm2_Bm3) <- hNK_Bm2_Bm3$group
hNK_Bm2_Bm3_ATRXdel <- subset(hNK_Bm2_Bm3, idents = "ATRXdel")
table(hNK_Bm2_Bm3_ATRXdel$predicted_NK_type_de.n350)
```

```{r}
DEG_ATRXdel <- DElegate::findDE(hNK_Bm2_Bm3_ATRXdel,
                                group_column = "predicted_NK_type_de.n350",
                                order_results = TRUE, compare_is_ref = TRUE,
                                replicate_column = "rados_et_al_2025", compare = "hNK_Bm2")

DEG_ATRXdel_signif <- subset(DEG_ATRXdel, padj < 0.1)
DEG_ATRXdel_up     <- subset(DEG_ATRXdel_signif, log_fc >  0.5 & padj < 0.05)
DEG_ATRXdel_down   <- subset(DEG_ATRXdel_signif, log_fc < -0.5 & padj < 0.05)

cat("Total (padj<0.1):", nrow(DEG_ATRXdel_signif),
    "| up Bm3:", nrow(DEG_ATRXdel_up), "| up Bm2:", nrow(DEG_ATRXdel_down), "\n")
DEG_ATRXdel_signif

write.table(DEG_ATRXdel, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm2-vs-hNKBm3_ATRXdel_2026-06-25.csv"))
```

### ___Volcano - ATRXdel___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_ATRXdel) <- DEG_ATRXdel$feature

p_volcano_ATRXdel <- EnhancedVolcano(
  DEG_ATRXdel, lab = rownames(DEG_ATRXdel), x = "log_fc", y = "padj",
  title = "ATRXdel", subtitle = "hNK_Bm2 (ref) vs. hNK_Bm3",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(DEG_ATRXdel$padj[DEG_ATRXdel$padj > 0], na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3, colCustom = volcano_colors(DEG_ATRXdel), colAlpha = 0.7,
  legendLabSize = 8, legendIconSize = 3, titleLabSize = 10, subtitleLabSize = 8,
  axisLabSize = 9, boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_ATRXdel

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_ATRXdel_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_ATRXdel; dev.off()
```

### ___Top DEG expression - ATRXdel___

```{r fig.height=5, fig.width=9}
plot_top_deg_violins(hNK_Bm2_Bm3_ATRXdel, deg_df = bind_rows(DEG_ATRXdel_up, DEG_ATRXdel_down),
  top_n = 12, out_path = paste0(params$out,
    "figures/DEG_hNKBm2-vs-hNKBm3_ATRXdel_topDEG_violins.pdf"))
```

### ___Pathway enrichment - ATRXdel (up in hNK_Bm3)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_ATRXdel_up$feature
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXdel Bm3 up: GO-BP"), "ATRXdel Bm3 up: GO-BP")
show_enrichment(run_enrichment(genes, "MF",       label = "ATRXdel Bm3 up: GO-MF"), "ATRXdel Bm3 up: GO-MF")
show_enrichment(run_enrichment(genes, "CC",       label = "ATRXdel Bm3 up: GO-CC"), "ATRXdel Bm3 up: GO-CC")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXdel Bm3 up: KEGG"),  "ATRXdel Bm3 up: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXdel Bm3 up: HALLMARK",
                               term2gene = hallmark_genes_H), "ATRXdel Bm3 up: HALLMARK")
```

### ___Pathway enrichment - ATRXdel (up in hNK_Bm2)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_ATRXdel_down$feature
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXdel Bm2 up: GO-BP"), "ATRXdel Bm2 up: GO-BP")
show_enrichment(run_enrichment(genes, "MF",       label = "ATRXdel Bm2 up: GO-MF"), "ATRXdel Bm2 up: GO-MF")
show_enrichment(run_enrichment(genes, "CC",       label = "ATRXdel Bm2 up: GO-CC"), "ATRXdel Bm2 up: GO-CC")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXdel Bm2 up: KEGG"),  "ATRXdel Bm2 up: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXdel Bm2 up: HALLMARK",
                               term2gene = hallmark_genes_H), "ATRXdel Bm2 up: HALLMARK")
```


## ___3. MYCNamp___

```{r}
Idents(hNK_Bm2_Bm3) <- hNK_Bm2_Bm3$group
hNK_Bm2_Bm3_MYCNamp <- subset(hNK_Bm2_Bm3, idents = "MYCNamp")
table(hNK_Bm2_Bm3_MYCNamp$predicted_NK_type_de.n350)
```

```{r}
DEG_MYCNamp <- DElegate::findDE(hNK_Bm2_Bm3_MYCNamp,
                                group_column = "predicted_NK_type_de.n350",
                                order_results = TRUE, compare_is_ref = TRUE,
                                replicate_column = "rados_et_al_2025", compare = "hNK_Bm2")

DEG_MYCNamp_signif <- subset(DEG_MYCNamp, padj < 0.1)
DEG_MYCNamp_up     <- subset(DEG_MYCNamp_signif, log_fc >  0.5 & padj < 0.05)
DEG_MYCNamp_down   <- subset(DEG_MYCNamp_signif, log_fc < -0.5 & padj < 0.05)

cat("Total (padj<0.1):", nrow(DEG_MYCNamp_signif),
    "| up Bm3:", nrow(DEG_MYCNamp_up), "| up Bm2:", nrow(DEG_MYCNamp_down), "\n")
DEG_MYCNamp_signif

write.table(DEG_MYCNamp, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm2-vs-hNKBm3_MYCNamp_2026-06-25.csv"))
```

### ___Volcano - MYCNamp___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_MYCNamp) <- DEG_MYCNamp$feature

p_volcano_MYCNamp <- EnhancedVolcano(
  DEG_MYCNamp, lab = rownames(DEG_MYCNamp), x = "log_fc", y = "padj",
  title = "MYCNamp", subtitle = "hNK_Bm2 (ref) vs. hNK_Bm3",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(DEG_MYCNamp$padj[DEG_MYCNamp$padj > 0], na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3, colCustom = volcano_colors(DEG_MYCNamp), colAlpha = 0.7,
  legendLabSize = 8, legendIconSize = 3, titleLabSize = 10, subtitleLabSize = 8,
  axisLabSize = 9, boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_MYCNamp

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_MYCNamp_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_MYCNamp; dev.off()
```

### ___Top DEG expression - MYCNamp___

```{r fig.height=5, fig.width=9}
plot_top_deg_violins(hNK_Bm2_Bm3_MYCNamp, deg_df = bind_rows(DEG_MYCNamp_up, DEG_MYCNamp_down),
  top_n = 12, out_path = paste0(params$out,
    "figures/DEG_hNKBm2-vs-hNKBm3_MYCNamp_topDEG_violins.pdf"))
```

### ___Pathway enrichment - MYCNamp (up in hNK_Bm3)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_MYCNamp_up$feature
show_enrichment(run_enrichment(genes, "BP",       label = "MYCNamp Bm3 up: GO-BP"), "MYCNamp Bm3 up: GO-BP")
show_enrichment(run_enrichment(genes, "MF",       label = "MYCNamp Bm3 up: GO-MF"), "MYCNamp Bm3 up: GO-MF")
show_enrichment(run_enrichment(genes, "CC",       label = "MYCNamp Bm3 up: GO-CC"), "MYCNamp Bm3 up: GO-CC")
show_enrichment(run_enrichment(genes, "KEGG",     label = "MYCNamp Bm3 up: KEGG"),  "MYCNamp Bm3 up: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "MYCNamp Bm3 up: HALLMARK",
                               term2gene = hallmark_genes_H), "MYCNamp Bm3 up: HALLMARK")
```

### ___Pathway enrichment - MYCNamp (up in hNK_Bm2)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_MYCNamp_down$feature
show_enrichment(run_enrichment(genes, "BP",       label = "MYCNamp Bm2 up: GO-BP"), "MYCNamp Bm2 up: GO-BP")
show_enrichment(run_enrichment(genes, "MF",       label = "MYCNamp Bm2 up: GO-MF"), "MYCNamp Bm2 up: GO-MF")
show_enrichment(run_enrichment(genes, "CC",       label = "MYCNamp Bm2 up: GO-CC"), "MYCNamp Bm2 up: GO-CC")
show_enrichment(run_enrichment(genes, "KEGG",     label = "MYCNamp Bm2 up: KEGG"),  "MYCNamp Bm2 up: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "MYCNamp Bm2 up: HALLMARK",
                               term2gene = hallmark_genes_H), "MYCNamp Bm2 up: HALLMARK")
```


## ___4. ATRXwtMYCNwt___

```{r}
Idents(hNK_Bm2_Bm3) <- hNK_Bm2_Bm3$group
hNK_Bm2_Bm3_ATRXwtMYCNwt <- subset(hNK_Bm2_Bm3, idents = "ATRXwtMYCNwt")
table(hNK_Bm2_Bm3_ATRXwtMYCNwt$predicted_NK_type_de.n350)
```

```{r}
DEG_ATRXwtMYCNwt <- DElegate::findDE(hNK_Bm2_Bm3_ATRXwtMYCNwt,
                                     group_column = "predicted_NK_type_de.n350",
                                     order_results = TRUE, compare_is_ref = TRUE,
                                     replicate_column = "rados_et_al_2025", compare = "hNK_Bm2")

DEG_ATRXwtMYCNwt_signif <- subset(DEG_ATRXwtMYCNwt, padj < 0.1)
DEG_ATRXwtMYCNwt_up     <- subset(DEG_ATRXwtMYCNwt_signif, log_fc >  0.5 & padj < 0.05)
DEG_ATRXwtMYCNwt_down   <- subset(DEG_ATRXwtMYCNwt_signif, log_fc < -0.5 & padj < 0.05)

cat("Total (padj<0.1):", nrow(DEG_ATRXwtMYCNwt_signif),
    "| up Bm3:", nrow(DEG_ATRXwtMYCNwt_up), "| up Bm2:", nrow(DEG_ATRXwtMYCNwt_down), "\n")
DEG_ATRXwtMYCNwt_signif

write.table(DEG_ATRXwtMYCNwt, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm2-vs-hNKBm3_ATRXwtMYCNwt_2026-06-25.csv"))
```

### ___Volcano - ATRXwtMYCNwt___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_ATRXwtMYCNwt) <- DEG_ATRXwtMYCNwt$feature

p_volcano_ATRXwtMYCNwt <- EnhancedVolcano(
  DEG_ATRXwtMYCNwt, lab = rownames(DEG_ATRXwtMYCNwt), x = "log_fc", y = "padj",
  title = "ATRXwtMYCNwt", subtitle = "hNK_Bm2 (ref) vs. hNK_Bm3",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(DEG_ATRXwtMYCNwt$padj[DEG_ATRXwtMYCNwt$padj > 0], na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3, colCustom = volcano_colors(DEG_ATRXwtMYCNwt), colAlpha = 0.7,
  legendLabSize = 8, legendIconSize = 3, titleLabSize = 10, subtitleLabSize = 8,
  axisLabSize = 9, boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_ATRXwtMYCNwt

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_ATRXwtMYCNwt_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_ATRXwtMYCNwt; dev.off()
```

### ___Top DEG expression - ATRXwtMYCNwt___

```{r fig.height=5, fig.width=9}
plot_top_deg_violins(hNK_Bm2_Bm3_ATRXwtMYCNwt,
  deg_df = bind_rows(DEG_ATRXwtMYCNwt_up, DEG_ATRXwtMYCNwt_down),
  top_n = 12, out_path = paste0(params$out,
    "figures/DEG_hNKBm2-vs-hNKBm3_ATRXwtMYCNwt_topDEG_violins.pdf"))
```

### ___Pathway enrichment - ATRXwtMYCNwt (up in hNK_Bm3)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_ATRXwtMYCNwt_up$feature
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXwtMYCNwt Bm3 up: GO-BP"), "ATRXwtMYCNwt Bm3 up: GO-BP")
show_enrichment(run_enrichment(genes, "MF",       label = "ATRXwtMYCNwt Bm3 up: GO-MF"), "ATRXwtMYCNwt Bm3 up: GO-MF")
show_enrichment(run_enrichment(genes, "CC",       label = "ATRXwtMYCNwt Bm3 up: GO-CC"), "ATRXwtMYCNwt Bm3 up: GO-CC")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXwtMYCNwt Bm3 up: KEGG"),  "ATRXwtMYCNwt Bm3 up: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXwtMYCNwt Bm3 up: HALLMARK",
                               term2gene = hallmark_genes_H), "ATRXwtMYCNwt Bm3 up: HALLMARK")
```

### ___Pathway enrichment - ATRXwtMYCNwt (up in hNK_Bm2)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_ATRXwtMYCNwt_down$feature
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXwtMYCNwt Bm2 up: GO-BP"), "ATRXwtMYCNwt Bm2 up: GO-BP")
show_enrichment(run_enrichment(genes, "MF",       label = "ATRXwtMYCNwt Bm2 up: GO-MF"), "ATRXwtMYCNwt Bm2 up: GO-MF")
show_enrichment(run_enrichment(genes, "CC",       label = "ATRXwtMYCNwt Bm2 up: GO-CC"), "ATRXwtMYCNwt Bm2 up: GO-CC")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXwtMYCNwt Bm2 up: KEGG"),  "ATRXwtMYCNwt Bm2 up: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXwtMYCNwt Bm2 up: HALLMARK",
                               term2gene = hallmark_genes_H), "ATRXwtMYCNwt Bm2 up: HALLMARK")
```


## ___Combined 4-panel volcano (publication figure)___

```{r fig.height=10, fig.width=7}
p_volcano_4panel <- (p_volcano_ctrl | p_volcano_ATRXdel) /
                    (p_volcano_MYCNamp | p_volcano_ATRXwtMYCNwt) +
  plot_annotation(title = "hNK_Bm2 vs. hNK_Bm3 - DEG by genetic group",
                  theme = theme(plot.title = element_text(size = 11, face = "bold")))

p_volcano_4panel

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_volcano_4panel.pdf"),
    height = 10, width = 7, useDingbats = FALSE)
p_volcano_4panel; dev.off()
```


# ___Summary: number of significant DEGs per group___

```{r}
summary_DEGs <- data.frame(
  group = c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt"),
  n_up_in_Bm3 = c(nrow(DEG_ctrl_up),   nrow(DEG_ATRXdel_up),
                  nrow(DEG_MYCNamp_up), nrow(DEG_ATRXwtMYCNwt_up)),
  n_up_in_Bm2 = c(nrow(DEG_ctrl_down),   nrow(DEG_ATRXdel_down),
                  nrow(DEG_MYCNamp_down), nrow(DEG_ATRXwtMYCNwt_down))
)
summary_DEGs
```

```{r fig.height=3.5, fig.width=4.5}
summary_DEGs_long <- summary_DEGs %>%
  pivot_longer(cols = c(n_up_in_Bm3, n_up_in_Bm2),
               names_to = "direction", values_to = "n_genes") %>%
  mutate(group = factor(group, levels = c("control","ATRXdel","MYCNamp","ATRXwtMYCNwt")),
         direction = factor(direction, levels = c("n_up_in_Bm3","n_up_in_Bm2"),
                            labels = c("up in hNK_Bm3", "up in hNK_Bm2")))

p_summary <- ggplot(summary_DEGs_long, aes(x = group, y = n_genes, fill = direction)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("up in hNK_Bm3" = COL_BM3, "up in hNK_Bm2" = COL_BM2)) +
  scale_x_discrete(labels = function(x) sub("ATRXwtMYCNwt", "ATRXwt\nMYCNwt", x)) +
  labs(title = "Significant DEGs per genetic group", subtitle = "|log_fc| > 0.5, padj < 0.05",
       x = NULL, y = "Number of genes", fill = NULL) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8), legend.position = "top")

p_summary

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_summary_barplot.pdf"),
    height = 3.5, width = 4.5, useDingbats = FALSE)
p_summary; dev.off()
```


# ___Overlap of DEGs across genetic groups___

## ___Genes higher in hNK_Bm3___

```{r fig.height=4.5, fig.width=4.5}
up_Bm3_lists <- list(control = DEG_ctrl_up$feature, ATRXdel = DEG_ATRXdel_up$feature,
                     MYCNamp = DEG_MYCNamp_up$feature, ATRXwtMYCNwt = DEG_ATRXwtMYCNwt_up$feature)

p_venn_Bm3 <- ggVennDiagram(up_Bm3_lists, set_size = 3.5, label_alpha = 0, label_size = 3.5,
  category.names = c("control","ATRXdel","MYCNamp","ATRXwtMYCNwt"),
  edge_size = 0.5, label_percent_digit = 1, label_color = "white") +
  scale_fill_gradient(low = "#D9C97A", high = COL_BM3, name = "n genes") +
  labs(title = "up in hNK_Bm3 - overlap across groups") +
  theme(plot.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 8), legend.title = element_text(size = 9)) +
  scale_x_continuous(expand = expansion(mult = .2))

p_venn_Bm3

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_Venn_upInBm3.pdf"),
    height = 4.5, width = 4.5, useDingbats = FALSE)
p_venn_Bm3; dev.off()

common_up_Bm3_all4 <- Reduce(intersect, up_Bm3_lists)
cat("Genes higher in hNK_Bm3 in ALL 4 groups:\n"); print(common_up_Bm3_all4)
common_up_Bm3_3of4 <- names(which(table(unlist(up_Bm3_lists)) >= 3))
cat("\nGenes higher in hNK_Bm3 in >= 3 of 4 groups:\n"); print(common_up_Bm3_3of4)
```

## ___Genes higher in hNK_Bm2___

```{r fig.height=4.5, fig.width=4.5}
up_Bm2_lists <- list(control = DEG_ctrl_down$feature, ATRXdel = DEG_ATRXdel_down$feature,
                     MYCNamp = DEG_MYCNamp_down$feature, ATRXwtMYCNwt = DEG_ATRXwtMYCNwt_down$feature)

p_venn_Bm2 <- ggVennDiagram(up_Bm2_lists, set_size = 3.5, label_alpha = 0, label_size = 3.5,
  category.names = c("control","ATRXdel","MYCNamp","ATRXwtMYCNwt"),
  edge_size = 0.5, label_percent_digit = 1, label_color = "white") +
  scale_fill_gradient(low = "#C8DDD9", high = COL_BM2, name = "n genes") +
  labs(title = "up in hNK_Bm2 - overlap across groups") +
  theme(plot.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 8), legend.title = element_text(size = 9)) +
  scale_x_continuous(expand = expansion(mult = .2))

p_venn_Bm2

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_Venn_upInBm2.pdf"),
    height = 4.5, width = 4.5, useDingbats = FALSE)
p_venn_Bm2; dev.off()

common_up_Bm2_all4 <- Reduce(intersect, up_Bm2_lists)
cat("Genes higher in hNK_Bm2 in ALL 4 groups:\n"); print(common_up_Bm2_all4)
common_up_Bm2_3of4 <- names(which(table(unlist(up_Bm2_lists)) >= 3))
cat("\nGenes higher in hNK_Bm2 in >= 3 of 4 groups:\n"); print(common_up_Bm2_3of4)
```

## ___Enrichment on consistently different genes___

### ___Consistently higher in hNK_Bm3 (>= 3 of 4 groups)___

```{r fig.height=4.5, fig.width=5}
genes <- common_up_Bm3_3of4; print(genes)
show_enrichment(run_enrichment(genes, "BP", label = "consistent Bm3 up: GO-BP"),
  "consistent Bm3 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_consistent_Bm3up_GO-BP_dotplot.pdf"))
show_enrichment(run_enrichment(genes, "KEGG", label = "consistent Bm3 up: KEGG"),
  "consistent Bm3 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_consistent_Bm3up_KEGG_dotplot.pdf"))
show_enrichment(run_enrichment(genes, "HALLMARK", label = "consistent Bm3 up: HALLMARK",
                               term2gene = hallmark_genes_H),
  "consistent Bm3 up: HALLMARK",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_consistent_Bm3up_HALLMARK_dotplot.pdf"))
```

### ___Consistently higher in hNK_Bm2 (>= 3 of 4 groups)___

```{r fig.height=4.5, fig.width=5}
genes <- common_up_Bm2_3of4; print(genes)
show_enrichment(run_enrichment(genes, "BP", label = "consistent Bm2 up: GO-BP"),
  "consistent Bm2 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_consistent_Bm2up_GO-BP_dotplot.pdf"))
show_enrichment(run_enrichment(genes, "KEGG", label = "consistent Bm2 up: KEGG"),
  "consistent Bm2 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_consistent_Bm2up_KEGG_dotplot.pdf"))
show_enrichment(run_enrichment(genes, "HALLMARK", label = "consistent Bm2 up: HALLMARK",
                               term2gene = hallmark_genes_H),
  "consistent Bm2 up: HALLMARK",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_consistent_Bm2up_HALLMARK_dotplot.pdf"))
```


# ___Pairwise group logFC correlations___

```{r fig.height=3, fig.width=9}
all_DEG_merged <- DEG_ctrl %>%
  dplyr::select(feature, log_fc, padj) %>% dplyr::rename(log_fc_ctrl = log_fc, padj_ctrl = padj) %>%
  full_join(DEG_ATRXdel %>% dplyr::select(feature, log_fc, padj) %>%
              dplyr::rename(log_fc_ATRXdel = log_fc, padj_ATRXdel = padj), by = "feature") %>%
  full_join(DEG_MYCNamp %>% dplyr::select(feature, log_fc, padj) %>%
              dplyr::rename(log_fc_MYCNamp = log_fc, padj_MYCNamp = padj), by = "feature") %>%
  full_join(DEG_ATRXwtMYCNwt %>% dplyr::select(feature, log_fc, padj) %>%
              dplyr::rename(log_fc_ATRXwtMYCNwt = log_fc, padj_ATRXwtMYCNwt = padj), by = "feature")

sc_opts <- list(geom_point(size = 0.5, alpha = 0.3, color = "gray50"),
                geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.6, fill = "gray80"),
                geom_hline(yintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.3),
                geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.3))

p_sc1 <- ggplot(all_DEG_merged, aes(x = log_fc_ctrl, y = log_fc_ATRXdel)) + sc_opts +
  labs(title = "control vs. ATRXdel", x = "log_fc (control)", y = "log_fc (ATRXdel)")
p_sc2 <- ggplot(all_DEG_merged, aes(x = log_fc_ctrl, y = log_fc_MYCNamp)) + sc_opts +
  labs(title = "control vs. MYCNamp", x = "log_fc (control)", y = "log_fc (MYCNamp)")
p_sc3 <- ggplot(all_DEG_merged, aes(x = log_fc_ctrl, y = log_fc_ATRXwtMYCNwt)) + sc_opts +
  labs(title = "control vs. ATRXwtMYCNwt", x = "log_fc (control)", y = "log_fc (ATRXwtMYCNwt)")

p_sc1 + p_sc2 + p_sc3

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_logFC_correlation_across_groups.pdf"),
    height = 3, width = 9, useDingbats = FALSE)
p_sc1 + p_sc2 + p_sc3; dev.off()
```

```{r}
cat("Spearman r of log_fc between groups:\n")
cat("ctrl vs ATRXdel:       r =", round(cor(all_DEG_merged$log_fc_ctrl, all_DEG_merged$log_fc_ATRXdel, use="complete.obs", method="spearman"), 3), "\n")
cat("ctrl vs MYCNamp:       r =", round(cor(all_DEG_merged$log_fc_ctrl, all_DEG_merged$log_fc_MYCNamp, use="complete.obs", method="spearman"), 3), "\n")
cat("ctrl vs ATRXwtMYCNwt:  r =", round(cor(all_DEG_merged$log_fc_ctrl, all_DEG_merged$log_fc_ATRXwtMYCNwt, use="complete.obs", method="spearman"), 3), "\n")
cat("ATRXdel vs MYCNamp:    r =", round(cor(all_DEG_merged$log_fc_ATRXdel, all_DEG_merged$log_fc_MYCNamp, use="complete.obs", method="spearman"), 3), "\n")
cat("ATRXdel vs ATRXwtMYCNwt: r =", round(cor(all_DEG_merged$log_fc_ATRXdel, all_DEG_merged$log_fc_ATRXwtMYCNwt, use="complete.obs", method="spearman"), 3), "\n")
cat("MYCNamp vs ATRXwtMYCNwt: r =", round(cor(all_DEG_merged$log_fc_MYCNamp, all_DEG_merged$log_fc_ATRXwtMYCNwt, use="complete.obs", method="spearman"), 3), "\n")
```


# ___DEG: hNK_Bm2 vs. hNK_Bm3 - all groups pooled, paired by patient___

All 16 patients pooled. The pseudobulk DESeq2 model includes patient as a blocking factor
(replicate_column creates design ~ patient + cluster), so the Bm2-Bm3 comparison is made
within each patient before averaging across patients.

```{r}
table(hNK_Bm2_Bm3$predicted_NK_type_de.n350, hNK_Bm2_Bm3$rados_et_al_2025)
```

```{r}
DEG_paired_all <- DElegate::findDE(hNK_Bm2_Bm3,
                                   group_column = "predicted_NK_type_de.n350",
                                   order_results = TRUE, compare_is_ref = TRUE,
                                   replicate_column = "rados_et_al_2025", compare = "hNK_Bm2")

DEG_paired_all_signif <- subset(DEG_paired_all, padj < 0.1)
DEG_paired_all_up     <- subset(DEG_paired_all_signif, log_fc >  0.5 & padj < 0.05)
DEG_paired_all_down   <- subset(DEG_paired_all_signif, log_fc < -0.5 & padj < 0.05)

cat("Total (padj<0.1):", nrow(DEG_paired_all_signif),
    "| up Bm3 (|logFC|>0.5, p<0.05):", nrow(DEG_paired_all_up),
    "| up Bm2:", nrow(DEG_paired_all_down), "\n")
DEG_paired_all_signif

write.table(DEG_paired_all, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_2026-06-25.csv"))
```

## ___Volcano - all groups, paired___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_paired_all) <- DEG_paired_all$feature

p_volcano_paired <- EnhancedVolcano(
  DEG_paired_all, lab = rownames(DEG_paired_all), x = "log_fc", y = "padj",
  title = "all groups - paired by patient", subtitle = "hNK_Bm2 (ref) vs. hNK_Bm3",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(DEG_paired_all$padj[DEG_paired_all$padj > 0], na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3, colCustom = volcano_colors(DEG_paired_all), colAlpha = 0.7,
  legendLabSize = 8, legendIconSize = 3, titleLabSize = 10, subtitleLabSize = 8,
  axisLabSize = 9, boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_paired

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_paired; dev.off()
```

## ___Top DEG expression - all groups, paired___

Violin body = distribution; boxplot = median/IQR; points = individual cells (quasi-random jitter).

```{r fig.height=5, fig.width=9}
plot_top_deg_violins(hNK_Bm2_Bm3, deg_df = bind_rows(DEG_paired_all_up, DEG_paired_all_down),
  top_n = 12, out_path = paste0(params$out,
    "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_topDEG_violins.pdf"))
```

## ___Pathway enrichment - all groups, paired (up in hNK_Bm3)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_paired_all_up$feature

go_BP_p    <- run_enrichment(genes, "BP",       label = "paired Bm3 up: GO-BP")
go_MF_p    <- run_enrichment(genes, "MF",       label = "paired Bm3 up: GO-MF")
go_CC_p    <- run_enrichment(genes, "CC",       label = "paired Bm3 up: GO-CC")
kegg_p     <- run_enrichment(genes, "KEGG",     label = "paired Bm3 up: KEGG")
hallmark_p <- run_enrichment(genes, "HALLMARK", label = "paired Bm3 up: HALLMARK",
                             term2gene = hallmark_genes_H)

show_enrichment(go_BP_p,    "paired Bm3 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm3up_GO-BP_dotplot.pdf"))
show_enrichment(go_MF_p,    "paired Bm3 up: GO-MF",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm3up_GO-MF_dotplot.pdf"))
show_enrichment(go_CC_p,    "paired Bm3 up: GO-CC",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm3up_GO-CC_dotplot.pdf"))
show_enrichment(kegg_p,     "paired Bm3 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm3up_KEGG_dotplot.pdf"))
show_enrichment(hallmark_p, "paired Bm3 up: HALLMARK",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm3up_HALLMARK_dotplot.pdf"))
```

```{r}
run_treeplot(kegg_p,  label = "paired Bm3 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm3up_KEGG_treeplot.pdf"))
run_treeplot(go_BP_p, label = "paired Bm3 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm3up_GO-BP_treeplot.pdf"))
```

## ___Pathway enrichment - all groups, paired (up in hNK_Bm2)___

```{r fig.height=4.5, fig.width=5}
genes <- DEG_paired_all_down$feature

go_BP_p2    <- run_enrichment(genes, "BP",       label = "paired Bm2 up: GO-BP")
go_MF_p2    <- run_enrichment(genes, "MF",       label = "paired Bm2 up: GO-MF")
go_CC_p2    <- run_enrichment(genes, "CC",       label = "paired Bm2 up: GO-CC")
kegg_p2     <- run_enrichment(genes, "KEGG",     label = "paired Bm2 up: KEGG")
hallmark_p2 <- run_enrichment(genes, "HALLMARK", label = "paired Bm2 up: HALLMARK",
                              term2gene = hallmark_genes_H)

show_enrichment(go_BP_p2,    "paired Bm2 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm2up_GO-BP_dotplot.pdf"))
show_enrichment(go_MF_p2,    "paired Bm2 up: GO-MF",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm2up_GO-MF_dotplot.pdf"))
show_enrichment(go_CC_p2,    "paired Bm2 up: GO-CC",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm2up_GO-CC_dotplot.pdf"))
show_enrichment(kegg_p2,     "paired Bm2 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm2up_KEGG_dotplot.pdf"))
show_enrichment(hallmark_p2, "paired Bm2 up: HALLMARK",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm2up_HALLMARK_dotplot.pdf"))
```

```{r}
run_treeplot(kegg_p2,  label = "paired Bm2 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm2up_KEGG_treeplot.pdf"))
run_treeplot(go_BP_p2, label = "paired Bm2 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_Bm2up_GO-BP_treeplot.pdf"))
```

## ___Overlap: paired all-groups vs. per-group___

```{r fig.height=4.5, fig.width=4.5}
paired_up_Bm3_list <- list(paired_all = DEG_paired_all_up$feature,
  control = DEG_ctrl_up$feature, ATRXdel = DEG_ATRXdel_up$feature,
  MYCNamp = DEG_MYCNamp_up$feature)

p_venn_paired_Bm3 <- ggVennDiagram(paired_up_Bm3_list, set_size = 3.5, label_alpha = 0,
  label_size = 3.5, category.names = c("paired_all","control","ATRXdel","MYCNamp"),
  edge_size = 0.5, label_percent_digit = 1, label_color = "white") +
  scale_fill_gradient(low = "#D9C97A", high = COL_BM3, name = "n genes") +
  labs(title = "up in hNK_Bm3: paired vs. per-group") +
  theme(plot.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 8), legend.title = element_text(size = 9)) +
  scale_x_continuous(expand = expansion(mult = .2))

p_venn_paired_Bm3

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_Venn_paired-vs-perGroup_upInBm3.pdf"),
    height = 4.5, width = 4.5, useDingbats = FALSE)
p_venn_paired_Bm3; dev.off()

only_in_paired_up <- setdiff(DEG_paired_all_up$feature,
  unique(c(DEG_ctrl_up$feature, DEG_ATRXdel_up$feature,
           DEG_MYCNamp_up$feature, DEG_ATRXwtMYCNwt_up$feature)))
cat("Genes up Bm3 only in paired (power gain):", length(only_in_paired_up), "\n")
print(only_in_paired_up)
```

```{r fig.height=4.5, fig.width=4.5}
paired_up_Bm2_list <- list(paired_all = DEG_paired_all_down$feature,
  control = DEG_ctrl_down$feature, ATRXdel = DEG_ATRXdel_down$feature,
  MYCNamp = DEG_MYCNamp_down$feature)

p_venn_paired_Bm2 <- ggVennDiagram(paired_up_Bm2_list, set_size = 3.5, label_alpha = 0,
  label_size = 3.5, category.names = c("paired_all","control","ATRXdel","MYCNamp"),
  edge_size = 0.5, label_percent_digit = 1, label_color = "white") +
  scale_fill_gradient(low = "#C8DDD9", high = COL_BM2, name = "n genes") +
  labs(title = "up in hNK_Bm2: paired vs. per-group") +
  theme(plot.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 8), legend.title = element_text(size = 9)) +
  scale_x_continuous(expand = expansion(mult = .2))

p_venn_paired_Bm2

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_Venn_paired-vs-perGroup_upInBm2.pdf"),
    height = 4.5, width = 4.5, useDingbats = FALSE)
p_venn_paired_Bm2; dev.off()

only_in_paired_down <- setdiff(DEG_paired_all_down$feature,
  unique(c(DEG_ctrl_down$feature, DEG_ATRXdel_down$feature,
           DEG_MYCNamp_down$feature, DEG_ATRXwtMYCNwt_down$feature)))
cat("Genes up Bm2 only in paired (power gain):", length(only_in_paired_down), "\n")
print(only_in_paired_down)
```


# ___Single-cell heatmap + PCA: pooled DEGs___

Significant genes from the pooled paired analysis (|log_fc| > 0.5, padj < 0.05),
sorted by log_fc descending (hNK_Bm3-high at top, hNK_Bm2-high at bottom).

## ___Prepare gene list___

```{r}
DEG_heatmap_genes <- bind_rows(DEG_paired_all_up, DEG_paired_all_down) %>%
  arrange(desc(log_fc))
cat("Total significant DEGs:", nrow(DEG_heatmap_genes), "\n")

if (nrow(DEG_heatmap_genes) > 80) {
  DEG_heatmap_genes <- DEG_heatmap_genes %>%
    mutate(abs_logfc = abs(log_fc)) %>% slice_max(abs_logfc, n = 80) %>%
    arrange(desc(log_fc)) %>% dplyr::select(-abs_logfc)
  message("Capping to top 80 genes by |log_fc| for heatmap readability")
}

genes_for_heatmap <- intersect(DEG_heatmap_genes$feature, rownames(hNK_Bm2_Bm3))
cat("Genes present in object:", length(genes_for_heatmap), "\n")
```

## ___Single-cell heatmap___

```{r fig.height=12, fig.width=8}
Idents(hNK_Bm2_Bm3) <- hNK_Bm2_Bm3$predicted_NK_type_de.n350
hNK_Bm2_Bm3 <- ScaleData(hNK_Bm2_Bm3, features = genes_for_heatmap, verbose = FALSE)

p_heatmap <- DoHeatmap(
  hNK_Bm2_Bm3, features = genes_for_heatmap,
  group.by = "predicted_NK_type_de.n350", group.colors = unname(COL_CLUST),
  label = TRUE, size = 3.5, draw.lines = TRUE
) +
  scale_fill_gradientn(colors = rev(RColorBrewer::brewer.pal(11, "RdBu")),
                       na.value = "white", name = "scaled\nexpr.") +
  labs(title    = "hNK_Bm2 vs. hNK_Bm3 - pooled paired analysis DEGs",
       subtitle = "sorted by log_fc  up-Bm3-high top  |  down-Bm2-high bottom") +
  theme(plot.title = element_text(size = 10, face = "bold"), plot.subtitle = element_text(size = 8),
        axis.text.y = element_text(size = 7), legend.text = element_text(size = 8),
        legend.title = element_text(size = 9))

p_heatmap

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_singleCell_heatmap.pdf"),
    height = 12, width = 8, useDingbats = FALSE)
p_heatmap; dev.off()
```

## ___PCA: cell separation driven by DEGs___

```{r}
all_deg_genes <- intersect(union(DEG_paired_all_up$feature, DEG_paired_all_down$feature),
                           rownames(hNK_Bm2_Bm3))

expr_mat  <- as.matrix(GetAssayData(hNK_Bm2_Bm3, slot = "data")[all_deg_genes, ])
gene_vars <- apply(expr_mat, 1, var)
expr_mat  <- expr_mat[gene_vars > 0, ]
cat("Genes used for PCA (non-zero variance):", nrow(expr_mat), "\n")

pca_res       <- prcomp(t(expr_mat), scale. = TRUE, center = TRUE)
var_explained <- summary(pca_res)$importance[2, ] * 100

pca_df <- data.frame(PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2],
                     cluster = hNK_Bm2_Bm3$predicted_NK_type_de.n350,
                     patient = hNK_Bm2_Bm3$rados_et_al_2025)
cat("Variance: PC1 =", round(var_explained[1], 1), "%, PC2 =",
    round(var_explained[2], 1), "%\n")
```

```{r fig.height=3.5, fig.width=3.5}
p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.35, size = 0.8) +
  stat_ellipse(type = "norm", linetype = 2, linewidth = 0.7) +
  scale_color_manual(values = COL_CLUST) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(title    = "PCA: hNK_Bm2 vs. hNK_Bm3",
       subtitle = "features = significant DEGs (pooled analysis)",
       x = paste0("PC1 (", round(var_explained[1], 1), "% variance)"),
       y = paste0("PC2 (", round(var_explained[2], 1), "% variance)"),
       color = NULL) +
  theme(legend.position = "top")

p_pca
```

## ___Top genes driving PC1___

```{r fig.height=5, fig.width=3.5}
n_top <- 30

loadings_pc1 <- data.frame(gene = rownames(pca_res$rotation),
                            loading = pca_res$rotation[, 1]) %>%
  left_join(DEG_paired_all %>% dplyr::select(feature, log_fc), by = c("gene" = "feature")) %>%
  mutate(abs_loading = abs(loading)) %>%
  slice_max(abs_loading, n = n_top) %>%
  arrange(loading) %>%
  mutate(gene = factor(gene, levels = gene),
         higher_in = case_when(log_fc > 0 ~ "hNK_Bm3", log_fc < 0 ~ "hNK_Bm2", TRUE ~ "n.s."))

p_loadings <- ggplot(loadings_pc1, aes(x = gene, y = loading, fill = higher_in)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  coord_flip() +
  scale_fill_manual(values = c("hNK_Bm3" = COL_BM3, "hNK_Bm2" = COL_BM2, "n.s." = "gray70"),
                    na.value = "gray70") +
  labs(title    = paste0("Top ", n_top, " genes: PC1 loadings"),
       subtitle = "fill = cluster with higher expression",
       x = NULL, y = "PC1 loading", fill = NULL) +
  theme(legend.position = "top", axis.text.y = element_text(size = 7))

p_loadings

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_PCA.pdf"),
    height = 5, width = 7, useDingbats = FALSE)
p_pca + p_loadings + plot_layout(widths = c(1, 1.2)); dev.off()

p_pca + p_loadings + plot_layout(widths = c(1, 1.2))
```

## ___Scree plot___

```{r fig.height=3, fig.width=4}
scree_df <- data.frame(PC = seq_along(var_explained), var = var_explained,
                       cum = cumsum(var_explained))

p_scree <- ggplot(scree_df[1:min(20, nrow(scree_df)), ], aes(x = PC, y = var)) +
  geom_col(fill = COL_BM2, color = NA, width = 0.7) +
  geom_line(aes(y = cum), color = "black", linewidth = 0.5) +
  geom_point(aes(y = cum), color = "black", size = 1.5) +
  scale_x_continuous(breaks = seq(2, 20, 2)) +
  labs(title = "Scree plot (top 20 PCs)",
       subtitle = "bars = % variance per PC; line = cumulative",
       x = "Principal component", y = "% variance explained")

p_scree

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_allGroups_paired_PCA_scree.pdf"),
    height = 3, width = 4, useDingbats = FALSE)
p_scree; dev.off()
```


# ___Expression of CD160 and CD52 in hNK_Bm2 and hNK_Bm3___

CD160 and CD52 expression across hNK_Bm2 and hNK_Bm3 cells, split by genetic group.

```{r fig.height=4, fig.width=8}
GENES_OF_INTEREST <- c("CD160", "CD52")

# Violin plots per cluster, split by genetic group
p_vln_cd <- VlnPlot(
  hNK_Bm2_Bm3,
  features    = GENES_OF_INTEREST,
  group.by    = "predicted_NK_type_de.n350",
  split.by    = "group",
  split.plot  = TRUE,
  cols        = unname(COLOR_CODE_GROUP[c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt")]),
  pt.size     = 0,
  ncol        = 2
) & pub_theme &
  theme(axis.title.x = element_blank(), legend.position = "right")

p_vln_cd

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_CD160-CD52_violin.pdf"),
    height = 4, width = 8, useDingbats = FALSE)
p_vln_cd; dev.off()
```

```{r fig.height=4, fig.width=7}
# Dot plot: cluster × genetic group for CD160 and CD52
hNK_Bm2_Bm3$clust_group <- paste0(hNK_Bm2_Bm3$predicted_NK_type_de.n350, " | ",
                                    hNK_Bm2_Bm3$group)
Idents(hNK_Bm2_Bm3) <- factor(
  hNK_Bm2_Bm3$clust_group,
  levels = as.vector(outer(c("hNK_Bm2", "hNK_Bm3"),
                           c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt"),
                           paste, sep = " | "))
)

p_dot_cd <- DotPlot(
  hNK_Bm2_Bm3,
  features = GENES_OF_INTEREST,
  cols     = c("lightgrey", "#B2182B")
) +
  coord_flip() +
  labs(title = "CD160 & CD52 expression", x = NULL, y = NULL) +
  pub_theme +
  theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 7))

p_dot_cd

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_CD160-CD52_dotplot.pdf"),
    height = 4, width = 7, useDingbats = FALSE)
p_dot_cd; dev.off()
```

```{r fig.height=3.5, fig.width=6}
# Box plot of per-cell normalized expression, Bm2 vs Bm3 per genetic group
Idents(hNK_Bm2_Bm3) <- hNK_Bm2_Bm3$predicted_NK_type_de.n350

expr_df <- bind_rows(lapply(GENES_OF_INTEREST, function(gene) {
  vals <- FetchData(hNK_Bm2_Bm3, vars = c(gene, "predicted_NK_type_de.n350", "group"))
  colnames(vals)[1] <- "expression"
  vals$gene <- gene
  vals
})) %>%
  mutate(
    cluster = factor(predicted_NK_type_de.n350, levels = c("hNK_Bm2", "hNK_Bm3")),
    group   = factor(group, levels = c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt"))
  )

p_box_cd <- ggplot(expr_df, aes(x = cluster, y = expression, fill = cluster)) +
  geom_violin(scale = "width", linewidth = 0.3, alpha = 0.85) +
  geom_boxplot(width = 0.12, fill = "white", outlier.size = 0.3, linewidth = 0.3) +
  facet_grid(gene ~ group, scales = "free_y") +
  scale_fill_manual(values = c("hNK_Bm2" = unname(COL_BM2),
                               "hNK_Bm3" = unname(COL_BM3))) +
  labs(
    title    = "CD160 & CD52 expression — hNK_Bm2 vs hNK_Bm3",
    subtitle = "per genetic group",
    x = NULL, y = "normalized expression"
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1), legend.position = "none")

p_box_cd

pdf(paste0(params$out, "figures/DEG_hNKBm2-vs-hNKBm3_CD160-CD52_boxplot.pdf"),
    height = 3.5, width = 6, useDingbats = FALSE)
p_box_cd; dev.off()
```

```{r}
# Wilcoxon test: Bm2 vs Bm3 per gene per genetic group
Idents(hNK_Bm2_Bm3) <- hNK_Bm2_Bm3$predicted_NK_type_de.n350

stat_cd <- bind_rows(lapply(GENES_OF_INTEREST, function(gene) {
  bind_rows(lapply(c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt"), function(grp) {
    sub_obj <- subset(hNK_Bm2_Bm3, subset = group == grp)
    bm2_vals <- FetchData(subset(sub_obj, idents = "hNK_Bm2"), vars = gene)[[1]]
    bm3_vals <- FetchData(subset(sub_obj, idents = "hNK_Bm3"), vars = gene)[[1]]
    if (length(bm2_vals) < 3 || length(bm3_vals) < 3) return(NULL)
    wt <- wilcox.test(bm3_vals, bm2_vals)
    data.frame(
      gene       = gene,
      group      = grp,
      n_Bm2      = length(bm2_vals),
      n_Bm3      = length(bm3_vals),
      median_Bm2 = median(bm2_vals),
      median_Bm3 = median(bm3_vals),
      log2FC     = log2((mean(bm3_vals) + 0.01) / (mean(bm2_vals) + 0.01)),
      p_value    = wt$p.value,
      stringsAsFactors = FALSE
    )
  }))
}))

stat_cd$p_adj <- p.adjust(stat_cd$p_value, method = "BH")
stat_cd$sig   <- stat_cd$p_adj < 0.05
stat_cd <- stat_cd %>%
  mutate(across(c(p_value, p_adj, log2FC, median_Bm2, median_Bm3), ~signif(.x, 3)))

cat("CD160 & CD52: Bm3 vs Bm2 Wilcoxon tests (BH-adjusted):\n")
print(stat_cd)

write.csv(stat_cd,
          paste0(params$out, "tables/CD160-CD52_Bm2-vs-Bm3_wilcoxon_2026-06-25.csv"),
          row.names = FALSE)
```

# ___The End___

```{r}
sessionInfo()
```

```{r render, eval=FALSE, include=FALSE}
rmarkdown::render(
  input         = "/home/rstudio/NK-NB_study/code/36_20260625_DEG_hNKBm2_vs_hNKBm3_analysis.Rmd",
  output_format = "html_document",
  output_file   = "36_20260625_DEG_hNKBm2_vs_hNKBm3_analysis.html",
  output_dir    = "/home/rstudio/mnt_out/NK_project/"
)
```
