---
title: "NK project: DEG analysis - all NK clusters, one-vs-rest paired by patient"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  input_NK: "/home/rstudio/mnt_out/NK_project/Rds/ALL_NK_cluster_withNewMetadata.Rds"
  out: "/home/rstudio/mnt_out/NK_project/"
---

PROJECT DESCRIPTION

Characterise each NK sub-population (hNK_Bm1–4) by identifying genes that
distinguish it from all other clusters combined (one-vs-rest pseudobulk DEG).
Patient ID is used as a blocking factor so that the comparison is made within
each donor before averaging across donors, accounting for the matched sample
design.

Positive log_fc = higher in the cluster of interest relative to all other NK cells.
Cutoffs: padj < 0.05 & |log_fc| > 0.5 (Bonferroni correction).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE,
                      error = FALSE,
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = 'future::multicore', workers = 20)
options(future.globals.maxSize = 8 * 10^9)

print(params)
```

```{r include=FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
```

# ___Load data___

```{r include=FALSE}
suppressPackageStartupMessages(library("DOSE"))
suppressPackageStartupMessages(library("enrichplot"))
suppressPackageStartupMessages(library("ggupset"))
suppressPackageStartupMessages(library("EnhancedVolcano"))
suppressPackageStartupMessages(library("ggbeeswarm"))
```

```{r include=FALSE}
ALL_NK <- readRDS(file = params$input_NK)
```

```{r}
counts_matrix <- GetAssayData(ALL_NK, slot = "counts")

hist(Matrix::rowSums(counts_matrix), breaks = 100000000, xlim = c(0, 100))
abline(v = 2)
hist(log(Matrix::rowSums(counts_matrix)), breaks = 1000)
abline(v = 3)

geneUniverse <- rownames(counts_matrix)[Matrix::rowSums(counts_matrix) > 1]
length(geneUniverse)

geneUniverse_entrezIDs <- map_gene_to_entrez(geneUniverse)

hallmark_genes_H <- msigdbr(species = "human", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)

cat("Clusters present:\n")
print(table(ALL_NK$predicted_NK_type_de.n350))
```

Publication theme and shared color helpers.

```{r}
pub_theme <- theme_classic(base_size = 9) +
  theme(
    axis.line        = element_line(linewidth = 0.4, color = "black"),
    axis.ticks       = element_line(linewidth = 0.4, color = "black"),
    axis.text        = element_text(size = 8,  color = "black"),
    axis.title       = element_text(size = 9,  color = "black"),
    plot.title       = element_text(size = 10, face = "bold"),
    plot.subtitle    = element_text(size = 8),
    legend.key.size  = unit(3,  "mm"),
    legend.text      = element_text(size = 8),
    legend.title     = element_text(size = 9),
    strip.background = element_blank(),
    strip.text       = element_text(size = 9)
  )
theme_set(pub_theme)

# All 4 cluster colors from COLOR_CODE_v2
COL_CLUST_ALL <- COLOR_CODE_v2[c("hNK_Bm1", "hNK_Bm2", "hNK_Bm3", "hNK_Bm4")]

# Volcano color helper: cluster-of-interest color for up, gray for down/NS
volcano_colors_ovr <- function(df, clust_col, padj_cut = 0.05, fc_cut = 0.5) {
  col_vec <- rep("gray70", nrow(df))
  col_vec[!is.na(df$padj) & df$padj < padj_cut & df$log_fc >  fc_cut] <- clust_col
  col_vec[!is.na(df$padj) & df$padj < padj_cut & df$log_fc < -fc_cut] <- "gray40"
  names(col_vec) <- rownames(df)
  col_vec
}

# Enrichment helpers
run_enrichment <- function(gene_symbols,
                           type = c("BP", "MF", "CC", "KEGG", "HALLMARK"),
                           term2gene = NULL,
                           label = "") {
  type <- match.arg(type)
  if (length(gene_symbols) == 0) {
    message(label, ": no input genes - skipping"); return(invisible(NULL))
  }
  entrez_ids <- na.omit(map_gene_to_entrez(gene_symbols))
  if (length(entrez_ids) == 0) {
    message(label, ": no valid Entrez IDs - skipping"); return(invisible(NULL))
  }
  tryCatch(
    {
      if (type %in% c("BP", "MF", "CC")) {
        enrichGO(gene = entrez_ids, ont = type, OrgDb = "org.Hs.eg.db",
                 readable = TRUE, pAdjustMethod = "bonferroni",
                 pvalueCutoff = 0.05)
      } else if (type == "KEGG") {
        enrichKEGG(gene = entrez_ids, organism = "hsa", keyType = "kegg",
                   pAdjustMethod = "bonferroni", pvalueCutoff = 0.05)
      } else if (type == "HALLMARK") {
        enricher(gene = entrez_ids, TERM2GENE = term2gene,
                 pAdjustMethod = "bonferroni", pvalueCutoff = 0.05)
      }
    },
    error = function(e) {
      message(label, " (", type, ") error: ", conditionMessage(e)); NULL
    }
  )
}

show_enrichment <- function(result, label = "", out_path = NULL,
                            out_w = 5, out_h = 4.5) {
  if (is.null(result) || nrow(result@result) == 0) {
    print(paste0("0 enriched terms - ", label)); return(invisible(NULL))
  }
  signif_rows <- result@result[result@result$p.adjust < 0.05, ]
  if (nrow(signif_rows) == 0) {
    print(paste0("0 terms with bonferroni p.adjust < 0.05 - ", label))
    return(invisible(NULL))
  }
  p <- dotplot(result, showCategory = 20) +
    ggtitle(label) + pub_theme +
    theme(axis.text.y = element_text(size = 7))
  print(p)
  if (!is.null(out_path)) {
    pdf(out_path, height = out_h, width = out_w, useDingbats = FALSE)
    print(p); dev.off()
  }
}

run_treeplot <- function(result, label = "", n_cluster = 5, n_words = 1,
                         show_cat = 15, out_path = NULL, out_w = 9, out_h = 4) {
  if (is.null(result) || nrow(result@result) == 0) {
    message("treeplot skipped (no enriched terms): ", label); return(invisible(NULL))
  }
  signif_rows <- result@result[result@result$p.adjust < 0.05, ]
  if (nrow(signif_rows) < 2) {
    message("treeplot skipped (< 2 significant terms): ", label); return(invisible(NULL))
  }
  p <- tryCatch(
    {
      enrichres2 <- pairwise_termsim(result)
      treeplot(enrichres2, showCategory = show_cat, color = "p.adjust",
               offset = 10, offset_tiplab = 0.5,
               nCluster = min(n_cluster, nrow(signif_rows)),
               nWords = n_words, split = "subcategory")
    },
    error = function(e) { message("treeplot error: ", conditionMessage(e)); NULL }
  )
  if (!is.null(p)) {
    tryCatch(print(p),
             error = function(e) message("treeplot render error: ", conditionMessage(e)))
    if (!is.null(out_path)) {
      tryCatch({
        pdf(out_path, height = out_h, width = out_w, useDingbats = FALSE)
        print(p); dev.off()
      }, error = function(e) {
        try(dev.off(), silent = TRUE)
        message("treeplot PDF error: ", conditionMessage(e))
      })
    }
  }
}

# Violin + boxplot + jitter for top marker genes of a given cluster
plot_top_markers_violins <- function(seurat_obj, deg_df, cluster_name,
                                     group_col = "predicted_NK_type_de.n350",
                                     top_n = 10, out_path = NULL,
                                     out_w = 8, out_h = 4) {
  if (nrow(deg_df) == 0) {
    message("no DEGs - skipping violin plot"); return(invisible(NULL))
  }
  top_genes <- deg_df %>%
    filter(log_fc > 0.5, padj < 0.05) %>%
    slice_max(log_fc, n = top_n) %>%
    arrange(desc(log_fc)) %>%
    mutate(feature = factor(feature, levels = unique(feature)))

  genes <- levels(top_genes$feature)
  genes <- intersect(genes, rownames(seurat_obj))
  if (length(genes) == 0) {
    message("no genes in object - skipping violin plot"); return(invisible(NULL))
  }

  expr <- as.data.frame(
    t(as.matrix(GetAssayData(seurat_obj, slot = "data")[genes, , drop = FALSE]))
  )
  expr$cluster <- seurat_obj@meta.data[[group_col]]
  expr_long <- expr %>%
    pivot_longer(-cluster, names_to = "gene", values_to = "expr") %>%
    mutate(gene      = factor(gene, levels = genes),
           is_target = cluster == cluster_name)

  p <- ggplot(expr_long, aes(x = cluster, y = expr,
                              fill = cluster, color = cluster)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0.3, alpha = 0.6) +
    geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white",
                 color = "black", linewidth = 0.35) +
    geom_quasirandom(size = 0.2, alpha = 0.3, color = "black",
                     groupOnX = TRUE, bandwidth = 0.5) +
    facet_wrap(~ gene, nrow = 2, scales = "free_y") +
    scale_fill_manual(values  = COL_CLUST_ALL, guide = "none") +
    scale_color_manual(values = COL_CLUST_ALL, guide = "none") +
    labs(x = NULL, y = "log-normalised expression",
         title    = paste0(cluster_name, " - top ", top_n, " marker genes"),
         subtitle = "log_fc > 0.5, padj < 0.05 (one-vs-rest)") +
    theme(
      axis.text.x   = element_text(size = 7, angle = 30, hjust = 1),
      axis.text.y   = element_text(size = 7),
      strip.text    = element_text(size = 7, face = "italic"),
      panel.spacing = unit(2, "mm")
    )

  print(p)
  if (!is.null(out_path)) {
    pdf(out_path, height = out_h, width = out_w, useDingbats = FALSE)
    print(p); dev.off()
  }
}
```

Helper to run one-vs-rest DEG for a single cluster.
`compare = "other"` with `compare_is_ref = TRUE` → positive log_fc = higher in cluster of interest.

```{r}
run_ovr_DEG <- function(seurat_obj, cluster_name,
                        group_col   = "predicted_NK_type_de.n350",
                        patient_col = "rados_et_al_2025") {
  seurat_obj$ovr_group <- ifelse(
    seurat_obj@meta.data[[group_col]] == cluster_name,
    cluster_name, "other"
  )
  DElegate::findDE(
    seurat_obj,
    group_column     = "ovr_group",
    order_results    = TRUE,
    compare_is_ref   = TRUE,
    replicate_column = patient_col,
    compare          = "other"
  )
}
```

# ___DEG one-vs-rest per cluster___

## ___1. hNK_Bm1___

```{r}
DEG_Bm1 <- run_ovr_DEG(ALL_NK, "hNK_Bm1")

DEG_Bm1_signif <- subset(DEG_Bm1, padj < 0.1)
DEG_Bm1_up     <- subset(DEG_Bm1, padj < 0.05 & log_fc >  0.5)
DEG_Bm1_down   <- subset(DEG_Bm1, padj < 0.05 & log_fc < -0.5)

cat("Total (padj<0.1):", nrow(DEG_Bm1_signif),
    "| up Bm1:", nrow(DEG_Bm1_up), "| down:", nrow(DEG_Bm1_down), "\n")
DEG_Bm1_signif

write.table(DEG_Bm1, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm1_vs_rest_2026-06-25.csv"))
```

### ___Volcano - hNK_Bm1___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_Bm1) <- DEG_Bm1$feature

p_volcano_Bm1 <- EnhancedVolcano(
  DEG_Bm1, lab = rownames(DEG_Bm1), x = "log_fc", y = "padj",
  title = "hNK_Bm1", subtitle = "vs. all other NK clusters",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(pmax(DEG_Bm1$padj, 1e-300), na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3,
  colCustom = volcano_colors_ovr(DEG_Bm1, COL_CLUST_ALL["hNK_Bm1"]),
  colAlpha = 0.7, legendLabSize = 8, legendIconSize = 3,
  titleLabSize = 10, subtitleLabSize = 8, axisLabSize = 9,
  boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_Bm1

pdf(paste0(params$out, "figures/DEG_hNKBm1_vs_rest_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_Bm1; dev.off()
```

### ___Top marker expression - hNK_Bm1___

```{r fig.height=4, fig.width=8}
plot_top_markers_violins(ALL_NK, deg_df = DEG_Bm1, cluster_name = "hNK_Bm1",
  top_n = 10, out_path = paste0(params$out,
    "figures/DEG_hNKBm1_vs_rest_topMarkers_violins.pdf"))
```

### ___Pathway enrichment - hNK_Bm1 marker genes___

```{r fig.height=6, fig.width=9}
genes <- DEG_Bm1_up$feature

go_BP_Bm1    <- run_enrichment(genes, "BP",       label = "Bm1 up: GO-BP")
go_MF_Bm1    <- run_enrichment(genes, "MF",       label = "Bm1 up: GO-MF")
go_CC_Bm1    <- run_enrichment(genes, "CC",       label = "Bm1 up: GO-CC")
kegg_Bm1     <- run_enrichment(genes, "KEGG",     label = "Bm1 up: KEGG")
hallmark_Bm1 <- run_enrichment(genes, "HALLMARK", label = "Bm1 up: HALLMARK",
                               term2gene = hallmark_genes_H)

show_enrichment(go_BP_Bm1, "Bm1 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm1_vs_rest_up_GO-BP_dotplot.pdf"))
show_enrichment(go_MF_Bm1, "Bm1 up: GO-MF",
  out_path = paste0(params$out, "figures/DEG_hNKBm1_vs_rest_up_GO-MF_dotplot.pdf"))
show_enrichment(go_CC_Bm1, "Bm1 up: GO-CC",
  out_path = paste0(params$out, "figures/DEG_hNKBm1_vs_rest_up_GO-CC_dotplot.pdf"))
show_enrichment(kegg_Bm1, "Bm1 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm1_vs_rest_up_KEGG_dotplot.pdf"))
show_enrichment(hallmark_Bm1, "Bm1 up: HALLMARK",
  out_path = paste0(params$out, "figures/DEG_hNKBm1_vs_rest_up_HALLMARK_dotplot.pdf"))
```

```{r}
run_treeplot(kegg_Bm1,  label = "Bm1 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm1_vs_rest_up_KEGG_treeplot.pdf"))
run_treeplot(go_BP_Bm1, label = "Bm1 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm1_vs_rest_up_GO-BP_treeplot.pdf"))
```


## ___2. hNK_Bm2___

```{r}
DEG_Bm2 <- run_ovr_DEG(ALL_NK, "hNK_Bm2")

DEG_Bm2_signif <- subset(DEG_Bm2, padj < 0.1)
DEG_Bm2_up     <- subset(DEG_Bm2, padj < 0.05 & log_fc >  0.5)
DEG_Bm2_down   <- subset(DEG_Bm2, padj < 0.05 & log_fc < -0.5)

cat("Total (padj<0.1):", nrow(DEG_Bm2_signif),
    "| up Bm2:", nrow(DEG_Bm2_up), "| down:", nrow(DEG_Bm2_down), "\n")
DEG_Bm2_signif

write.table(DEG_Bm2, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm2_vs_rest_2026-06-25.csv"))
```

### ___Volcano - hNK_Bm2___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_Bm2) <- DEG_Bm2$feature

p_volcano_Bm2 <- EnhancedVolcano(
  DEG_Bm2, lab = rownames(DEG_Bm2), x = "log_fc", y = "padj",
  title = "hNK_Bm2", subtitle = "vs. all other NK clusters",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(pmax(DEG_Bm2$padj, 1e-300), na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3,
  colCustom = volcano_colors_ovr(DEG_Bm2, COL_CLUST_ALL["hNK_Bm2"]),
  colAlpha = 0.7, legendLabSize = 8, legendIconSize = 3,
  titleLabSize = 10, subtitleLabSize = 8, axisLabSize = 9,
  boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_Bm2

pdf(paste0(params$out, "figures/DEG_hNKBm2_vs_rest_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_Bm2; dev.off()
```

### ___Top marker expression - hNK_Bm2___

```{r fig.height=4, fig.width=8}
plot_top_markers_violins(ALL_NK, deg_df = DEG_Bm2, cluster_name = "hNK_Bm2",
  top_n = 10, out_path = paste0(params$out,
    "figures/DEG_hNKBm2_vs_rest_topMarkers_violins.pdf"))
```

### ___Pathway enrichment - hNK_Bm2 marker genes___

```{r fig.height=6, fig.width=9}
genes <- DEG_Bm2_up$feature

go_BP_Bm2    <- run_enrichment(genes, "BP",       label = "Bm2 up: GO-BP")
go_MF_Bm2    <- run_enrichment(genes, "MF",       label = "Bm2 up: GO-MF")
go_CC_Bm2    <- run_enrichment(genes, "CC",       label = "Bm2 up: GO-CC")
kegg_Bm2     <- run_enrichment(genes, "KEGG",     label = "Bm2 up: KEGG")
hallmark_Bm2 <- run_enrichment(genes, "HALLMARK", label = "Bm2 up: HALLMARK",
                               term2gene = hallmark_genes_H)

show_enrichment(go_BP_Bm2, "Bm2 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm2_vs_rest_up_GO-BP_dotplot.pdf"))
show_enrichment(go_MF_Bm2, "Bm2 up: GO-MF",
  out_path = paste0(params$out, "figures/DEG_hNKBm2_vs_rest_up_GO-MF_dotplot.pdf"))
show_enrichment(go_CC_Bm2, "Bm2 up: GO-CC",
  out_path = paste0(params$out, "figures/DEG_hNKBm2_vs_rest_up_GO-CC_dotplot.pdf"))
show_enrichment(kegg_Bm2, "Bm2 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm2_vs_rest_up_KEGG_dotplot.pdf"))
show_enrichment(hallmark_Bm2, "Bm2 up: HALLMARK",
  out_path = paste0(params$out, "figures/DEG_hNKBm2_vs_rest_up_HALLMARK_dotplot.pdf"))
```

```{r}
run_treeplot(kegg_Bm2,  label = "Bm2 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm2_vs_rest_up_KEGG_treeplot.pdf"))
run_treeplot(go_BP_Bm2, label = "Bm2 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm2_vs_rest_up_GO-BP_treeplot.pdf"))
```


## ___3. hNK_Bm3___

```{r}
DEG_Bm3 <- run_ovr_DEG(ALL_NK, "hNK_Bm3")

DEG_Bm3_signif <- subset(DEG_Bm3, padj < 0.1)
DEG_Bm3_up     <- subset(DEG_Bm3, padj < 0.05 & log_fc >  0.5)
DEG_Bm3_down   <- subset(DEG_Bm3, padj < 0.05 & log_fc < -0.5)

cat("Total (padj<0.1):", nrow(DEG_Bm3_signif),
    "| up Bm3:", nrow(DEG_Bm3_up), "| down:", nrow(DEG_Bm3_down), "\n")
DEG_Bm3_signif

write.table(DEG_Bm3, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm3_vs_rest_2026-06-25.csv"))
```

### ___Volcano - hNK_Bm3___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_Bm3) <- DEG_Bm3$feature

p_volcano_Bm3 <- EnhancedVolcano(
  DEG_Bm3, lab = rownames(DEG_Bm3), x = "log_fc", y = "padj",
  title = "hNK_Bm3", subtitle = "vs. all other NK clusters",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(pmax(DEG_Bm3$padj, 1e-300), na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3,
  colCustom = volcano_colors_ovr(DEG_Bm3, COL_CLUST_ALL["hNK_Bm3"]),
  colAlpha = 0.7, legendLabSize = 8, legendIconSize = 3,
  titleLabSize = 10, subtitleLabSize = 8, axisLabSize = 9,
  boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_Bm3

pdf(paste0(params$out, "figures/DEG_hNKBm3_vs_rest_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_Bm3; dev.off()
```

### ___Top marker expression - hNK_Bm3___

```{r fig.height=4, fig.width=8}
plot_top_markers_violins(ALL_NK, deg_df = DEG_Bm3, cluster_name = "hNK_Bm3",
  top_n = 10, out_path = paste0(params$out,
    "figures/DEG_hNKBm3_vs_rest_topMarkers_violins.pdf"))
```

### ___Pathway enrichment - hNK_Bm3 marker genes___

```{r fig.height=6, fig.width=9}
genes <- DEG_Bm3_up$feature

go_BP_Bm3    <- run_enrichment(genes, "BP",       label = "Bm3 up: GO-BP")
go_MF_Bm3    <- run_enrichment(genes, "MF",       label = "Bm3 up: GO-MF")
go_CC_Bm3    <- run_enrichment(genes, "CC",       label = "Bm3 up: GO-CC")
kegg_Bm3     <- run_enrichment(genes, "KEGG",     label = "Bm3 up: KEGG")
hallmark_Bm3 <- run_enrichment(genes, "HALLMARK", label = "Bm3 up: HALLMARK",
                               term2gene = hallmark_genes_H)

show_enrichment(go_BP_Bm3, "Bm3 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm3_vs_rest_up_GO-BP_dotplot.pdf"))
show_enrichment(go_MF_Bm3, "Bm3 up: GO-MF",
  out_path = paste0(params$out, "figures/DEG_hNKBm3_vs_rest_up_GO-MF_dotplot.pdf"))
show_enrichment(go_CC_Bm3, "Bm3 up: GO-CC",
  out_path = paste0(params$out, "figures/DEG_hNKBm3_vs_rest_up_GO-CC_dotplot.pdf"))
show_enrichment(kegg_Bm3, "Bm3 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm3_vs_rest_up_KEGG_dotplot.pdf"))
show_enrichment(hallmark_Bm3, "Bm3 up: HALLMARK",
  out_path = paste0(params$out, "figures/DEG_hNKBm3_vs_rest_up_HALLMARK_dotplot.pdf"))
```

```{r}
run_treeplot(kegg_Bm3,  label = "Bm3 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm3_vs_rest_up_KEGG_treeplot.pdf"))
run_treeplot(go_BP_Bm3, label = "Bm3 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm3_vs_rest_up_GO-BP_treeplot.pdf"))
```


## ___4. hNK_Bm4___

```{r}
DEG_Bm4 <- run_ovr_DEG(ALL_NK, "hNK_Bm4")

DEG_Bm4_signif <- subset(DEG_Bm4, padj < 0.1)
DEG_Bm4_up     <- subset(DEG_Bm4, padj < 0.05 & log_fc >  0.5)
DEG_Bm4_down   <- subset(DEG_Bm4, padj < 0.05 & log_fc < -0.5)

cat("Total (padj<0.1):", nrow(DEG_Bm4_signif),
    "| up Bm4:", nrow(DEG_Bm4_up), "| down:", nrow(DEG_Bm4_down), "\n")
DEG_Bm4_signif

write.table(DEG_Bm4, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out, "tables/DEG_hNKBm4_vs_rest_2026-06-25.csv"))
```

### ___Volcano - hNK_Bm4___

```{r fig.height=5, fig.width=3.5}
rownames(DEG_Bm4) <- DEG_Bm4$feature

p_volcano_Bm4 <- EnhancedVolcano(
  DEG_Bm4, lab = rownames(DEG_Bm4), x = "log_fc", y = "padj",
  title = "hNK_Bm4", subtitle = "vs. all other NK clusters",
  pCutoff = 0.05, FCcutoff = 0.5,
  ylim = c(0, ceiling(-log10(min(pmax(DEG_Bm4$padj, 1e-300), na.rm = TRUE))) + 0.5),
  pointSize = 1.5, labSize = 3,
  colCustom = volcano_colors_ovr(DEG_Bm4, COL_CLUST_ALL["hNK_Bm4"]),
  colAlpha = 0.7, legendLabSize = 8, legendIconSize = 3,
  titleLabSize = 10, subtitleLabSize = 8, axisLabSize = 9,
  boxedLabels = TRUE, drawConnectors = TRUE,
  widthConnectors = 0.5, colConnectors = "black", legendPosition = "right"
) + pub_theme

p_volcano_Bm4

pdf(paste0(params$out, "figures/DEG_hNKBm4_vs_rest_volcano.pdf"),
    height = 5, width = 3.5, useDingbats = FALSE)
p_volcano_Bm4; dev.off()
```

### ___Top marker expression - hNK_Bm4___

```{r fig.height=4, fig.width=8}
plot_top_markers_violins(ALL_NK, deg_df = DEG_Bm4, cluster_name = "hNK_Bm4",
  top_n = 10, out_path = paste0(params$out,
    "figures/DEG_hNKBm4_vs_rest_topMarkers_violins.pdf"))
```

### ___Pathway enrichment - hNK_Bm4 marker genes___

```{r fig.height=6, fig.width=9}
genes <- DEG_Bm4_up$feature

go_BP_Bm4    <- run_enrichment(genes, "BP",       label = "Bm4 up: GO-BP")
go_MF_Bm4    <- run_enrichment(genes, "MF",       label = "Bm4 up: GO-MF")
go_CC_Bm4    <- run_enrichment(genes, "CC",       label = "Bm4 up: GO-CC")
kegg_Bm4     <- run_enrichment(genes, "KEGG",     label = "Bm4 up: KEGG")
hallmark_Bm4 <- run_enrichment(genes, "HALLMARK", label = "Bm4 up: HALLMARK",
                               term2gene = hallmark_genes_H)

show_enrichment(go_BP_Bm4, "Bm4 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm4_vs_rest_up_GO-BP_dotplot.pdf"))
show_enrichment(go_MF_Bm4, "Bm4 up: GO-MF",
  out_path = paste0(params$out, "figures/DEG_hNKBm4_vs_rest_up_GO-MF_dotplot.pdf"))
show_enrichment(go_CC_Bm4, "Bm4 up: GO-CC",
  out_path = paste0(params$out, "figures/DEG_hNKBm4_vs_rest_up_GO-CC_dotplot.pdf"))
show_enrichment(kegg_Bm4, "Bm4 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm4_vs_rest_up_KEGG_dotplot.pdf"))
show_enrichment(hallmark_Bm4, "Bm4 up: HALLMARK",
  out_path = paste0(params$out, "figures/DEG_hNKBm4_vs_rest_up_HALLMARK_dotplot.pdf"))
```

```{r}
run_treeplot(kegg_Bm4,  label = "Bm4 up: KEGG",
  out_path = paste0(params$out, "figures/DEG_hNKBm4_vs_rest_up_KEGG_treeplot.pdf"))
run_treeplot(go_BP_Bm4, label = "Bm4 up: GO-BP",
  out_path = paste0(params$out, "figures/DEG_hNKBm4_vs_rest_up_GO-BP_treeplot.pdf"))
```


## ___Combined 4-panel volcano (publication figure)___

```{r fig.height=10, fig.width=7}
p_volcano_4panel <- (p_volcano_Bm1 | p_volcano_Bm2) /
                    (p_volcano_Bm3 | p_volcano_Bm4) +
  plot_annotation(title = "NK clusters - one-vs-rest marker genes",
                  theme = theme(plot.title = element_text(size = 11, face = "bold")))

p_volcano_4panel

pdf(paste0(params$out, "figures/DEG_allNKclusters_vs_rest_volcano_4panel.pdf"),
    height = 10, width = 7, useDingbats = FALSE)
p_volcano_4panel; dev.off()
```


# ___Summary: marker gene counts per cluster___

```{r fig.height=3.5, fig.width=4}
summary_markers <- data.frame(
  cluster  = c("hNK_Bm1", "hNK_Bm2", "hNK_Bm3", "hNK_Bm4"),
  n_up     = c(nrow(DEG_Bm1_up), nrow(DEG_Bm2_up), nrow(DEG_Bm3_up), nrow(DEG_Bm4_up)),
  n_down   = c(nrow(DEG_Bm1_down), nrow(DEG_Bm2_down), nrow(DEG_Bm3_down), nrow(DEG_Bm4_down))
)
summary_markers

summary_long <- summary_markers %>%
  pivot_longer(c(n_up, n_down), names_to = "direction", values_to = "n_genes") %>%
  mutate(cluster   = factor(cluster, levels = c("hNK_Bm1","hNK_Bm2","hNK_Bm3","hNK_Bm4")),
         direction = factor(direction, levels = c("n_up","n_down"),
                            labels = c("higher in cluster", "lower in cluster")))

p_summary <- ggplot(summary_long, aes(x = cluster, y = n_genes, fill = cluster)) +
  geom_col(width = 0.7) +
  facet_wrap(~ direction, scales = "free_y") +
  scale_fill_manual(values = COL_CLUST_ALL, guide = "none") +
  labs(title = "Significant DEGs per cluster (one-vs-rest)",
       subtitle = "|log_fc| > 0.5, padj < 0.05",
       x = NULL, y = "Number of genes") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))

p_summary

pdf(paste0(params$out, "figures/DEG_allNKclusters_vs_rest_summary_barplot.pdf"),
    height = 3.5, width = 4, useDingbats = FALSE)
p_summary; dev.off()
```


# ___Single-cell heatmap: all clusters × all significant marker genes___

One row per gene, one column per cell. Genes are ordered in four blocks — one per cluster —
sorted by their one-vs-rest log_fc within each block (cluster-defining genes at the top of
each block). Only upregulated markers (log_fc > 0.5, padj < 0.05) are included.
Genes that appear as markers of more than one cluster are assigned to the cluster
with the highest log_fc.

```{r}
# Collect all upregulated markers and resolve ambiguity by highest log_fc
all_up_markers <- bind_rows(
  DEG_Bm1_up %>% mutate(primary_cluster = "hNK_Bm1"),
  DEG_Bm2_up %>% mutate(primary_cluster = "hNK_Bm2"),
  DEG_Bm3_up %>% mutate(primary_cluster = "hNK_Bm3"),
  DEG_Bm4_up %>% mutate(primary_cluster = "hNK_Bm4")
)

# Assign each gene to cluster with highest log_fc
gene_assignments <- all_up_markers %>%
  group_by(feature) %>%
  slice_max(log_fc, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(feature, primary_cluster, log_fc) %>%
  arrange(primary_cluster,
          factor(primary_cluster, levels = c("hNK_Bm1","hNK_Bm2","hNK_Bm3","hNK_Bm4")),
          desc(log_fc))

# Final ordered gene list for heatmap (sorted within each cluster block by log_fc desc)
gene_order <- gene_assignments %>%
  arrange(factor(primary_cluster, levels = c("hNK_Bm1","hNK_Bm2","hNK_Bm3","hNK_Bm4")),
          desc(log_fc)) %>%
  pull(feature)

# Keep only genes present in the object
gene_order <- intersect(gene_order, rownames(ALL_NK))

cat("Total unique marker genes for heatmap:", length(gene_order), "\n")
cat("Per cluster:\n")
print(table(gene_assignments$primary_cluster))
```

```{r}
Idents(ALL_NK) <- ALL_NK$predicted_NK_type_de.n350
ALL_NK <- ScaleData(ALL_NK, features = gene_order, verbose = FALSE)
```

```{r fig.height=14, fig.width=9}
p_heatmap_all <- DoHeatmap(
  ALL_NK, features = gene_order,
  group.by     = "predicted_NK_type_de.n350",
  group.colors = unname(COL_CLUST_ALL),
  label        = TRUE, size = 3.5, draw.lines = TRUE
) +
  scale_fill_gradientn(
    colors   = rev(RColorBrewer::brewer.pal(11, "RdBu")),
    na.value = "white", name = "scaled\nexpr."
  ) +
  labs(title    = "NK cluster marker genes (one-vs-rest)",
       subtitle = "log_fc > 0.5, padj < 0.05 | genes sorted by cluster block then log_fc") +
  theme(
    plot.title   = element_text(size = 10, face = "bold"),
    plot.subtitle = element_text(size = 8),
    axis.text.y  = element_text(size = 6),
    legend.text  = element_text(size = 8),
    legend.title = element_text(size = 9)
  )

p_heatmap_all

pdf(paste0(params$out, "figures/DEG_allNKclusters_vs_rest_singleCell_heatmap.pdf"),
    height = 14, width = 9, useDingbats = FALSE)
p_heatmap_all; dev.off()
```

## ___Cluster-block gene summary table___

```{r}
gene_assignments %>%
  group_by(primary_cluster) %>%
  arrange(desc(log_fc)) %>%
  slice_head(n = 20) %>%
  ungroup() %>%
  dplyr::select(primary_cluster, feature, log_fc) %>%
  dplyr::rename(cluster = primary_cluster, gene = feature, log_fc_ovr = log_fc)
```


# ___The End___

```{r}
sessionInfo()
```

```{r render, eval=FALSE, include=FALSE}
rmarkdown::render(
  input         = "/home/rstudio/NK-NB_study/code/37_20260625_DEG_allNKclusters_oneVsRest_analysis.Rmd",
  output_format = "html_document",
  output_file   = "37_20260625_DEG_allNKclusters_oneVsRest_analysis.html",
  output_dir    = "/home/rstudio/mnt_out/NK_project/"
)
```
