---
title: "NK project: bulk ATAC-seq - NB isogenic cell lines E6 (ATRXdel) vs p53 (ATRXwt)"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  data_dir: "/home/rstudio/mnt_data/nb_tumors/lorenzi2026.bulkATAC/GSE312501"
  nk_tables: "/home/rstudio/mnt_out/NK_project/tables/"
  out:       "/home/rstudio/mnt_out/NK_project/"
  tss_window: 2000
---

**PROJECT DESCRIPTION**

Validation of NK cell finding direction in an isogenic neuroblastoma cell line system.
E6 = SK-N-SH ATRXdel (ATRX-deleted); p53 = SK-N-SH ATRXwt (ATRX wild-type).
Genome: hg38 (UCSC chromosome naming).

Question: do TFs and genes found to be differentially expressed or active in NK cells
from ATRXdel patients (vs. ATRXwt/control) also show concordant chromatin accessibility
changes at their loci in the isogenic ATRXdel NB cell line?

Samples (Lorenzi 2026, GSE312501):
  E6:    ATRXdel clone, DMSO (n=3) + RA (n=3)
  p53:   ATRXwt clone,  DMSO (n=3) + RA (n=3)
  CHLA90, SKNMM: included in QC for context.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = TRUE,
                      error = FALSE,
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
options(future.globals.maxSize = 8 * 10^9)
print(params)
```

```{r include=FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
```

# ___Load packages___

```{r include=FALSE}
suppressPackageStartupMessages({
  library(rtracklayer)
  library(GenomicRanges)
  library(IRanges)
  library(limma)
  library(TxDb.Hsapiens.UCSC.hg38.knownGene)
  library(org.Hs.eg.db)
  library(AnnotationDbi)
  library(ggrepel)
  library(pheatmap)
  library(RColorBrewer)
  library(EnhancedVolcano)
  library(tidyverse)
  library(purrr)
})
```

```{r}
theme_set(pub_theme)

# Cell-line colors (map onto project palette where logical)
COL_ATRX <- c(
  "ATRXdel" = unname(COLOR_CODE_GROUP["ATRXdel"]),  # "#4f4e92"
  "ATRXwt"  = unname(COLOR_CODE_GROUP["control"])    # "#7EA3AC"
)
COL_LINE <- c(
  "p53"    = unname(COLOR_CODE_GROUP["control"]),
  "E6"     = unname(COLOR_CODE_GROUP["ATRXdel"]),
  "CHLA90" = unname(COLOR_CODE_GROUP["MYCNamp"]),
  "SKNMM"  = unname(COLOR_CODE_GROUP["ATRXwtMYCNwt"])
)
COL_TREAT <- c("DMSO" = "gray30", "RA" = "#C56A30")
```

# ___Sample metadata___

```{r}
bw_files   <- list.files(params$data_dir, pattern = "\\.bigWig$",        full.names = TRUE)
peak_files <- list.files(params$data_dir, pattern = "\\.broadPeak\\.gz$", full.names = TRUE)

parse_fname <- function(fnames) {
  base <- gsub("^GSM[0-9]+_", "", basename(fnames))
  base <- gsub("\\.mLb.*$", "", base)
  parts <- strsplit(base, "_")
  data.frame(
    file      = fnames,
    sample_id = base,
    cell_line = sapply(parts, `[[`, 1),
    treatment = sapply(parts, `[[`, 2),
    replicate = sapply(parts, `[[`, 3),
    stringsAsFactors = FALSE
  )
}

meta_bw   <- parse_fname(bw_files)
meta_peak <- parse_fname(peak_files)

meta_bw$ATRX <- ifelse(meta_bw$cell_line == "E6", "ATRXdel",
                 ifelse(meta_bw$cell_line == "p53", "ATRXwt", NA))
meta_bw$ATRX <- factor(meta_bw$ATRX, levels = c("ATRXwt", "ATRXdel"))

cat("BigWig files found:", nrow(meta_bw), "\n")
print(table(meta_bw$cell_line, meta_bw$treatment))
```

# ___Consensus peak set (E6 + p53 DMSO)___

Build one merged peak set from all DMSO E6 and p53 replicates so the differential
analysis uses exactly the same genomic windows for both conditions.

```{r}
import_broadpeak <- function(peak_file) {
  df <- read.table(peak_file, header = FALSE,
                   col.names = c("chr","start","end","name","score","strand",
                                 "signalValue","pValue","qValue"))
  std_chroms <- paste0("chr", c(1:22, "X", "Y", "M"))
  df <- df[df$chr %in% std_chroms, ]
  GRanges(seqnames = df$chr,
          ranges   = IRanges(start = df$start + 1L, end = df$end),
          score    = df$signalValue)
}

focal_peaks_meta <- meta_peak[meta_peak$cell_line %in% c("E6","p53") &
                               meta_peak$treatment == "DMSO", ]

peak_list <- lapply(focal_peaks_meta$file, import_broadpeak)
names(peak_list) <- focal_peaks_meta$sample_id

consensus_peaks <- GenomicRanges::reduce(unlist(GRangesList(peak_list)))

std_chroms <- paste0("chr", c(1:22, "X", "Y", "M"))
consensus_peaks <- consensus_peaks[as.character(seqnames(consensus_peaks)) %in% std_chroms]
seqlevels(consensus_peaks) <- seqlevelsInUse(consensus_peaks)

consensus_peaks$peak_id <- paste0("peak_", seq_len(length(consensus_peaks)))

cat("Consensus peaks (E6 + p53 DMSO):", length(consensus_peaks), "\n")
hist(width(consensus_peaks), breaks = 200, main = "Peak width distribution",
     xlab = "Peak width (bp)", col = "gray70", border = NA)
abline(v = median(width(consensus_peaks)), col = "#4f4e92", lwd = 2, lty = 2)
```

# ___Score peaks from bigWig___

For each sample, compute mean bigWig coverage within each consensus peak.
Only DMSO E6 and p53 samples are used for the primary differential analysis;
all 24 samples are scored for the QC PCA.

```{r}
score_bw_at_peaks <- function(bw_path, peaks) {
  # Use bare GRanges (no seqinfo) for the query to avoid seqinfo validation
  query <- GRanges(seqnames = as.character(seqnames(peaks)), ranges = ranges(peaks))
  bw_gr <- tryCatch(
    rtracklayer::import.bw(bw_path, which = query),
    error = function(e) {
      message("ERROR importing ", basename(bw_path), ": ", conditionMessage(e))
      NULL
    }
  )
  if (is.null(bw_gr) || length(bw_gr) == 0L) {
    message("No signal returned from ", basename(bw_path))
    return(rep(0, length(peaks)))
  }
  # Weighted mean: sum(score * overlap_width) / peak_width
  hits   <- findOverlaps(query, bw_gr)
  result <- rep(0.0, length(peaks))
  if (length(hits) > 0L) {
    ov_w <- width(pintersect(query[queryHits(hits)], bw_gr[subjectHits(hits)],
                              ignore.strand = TRUE))
    agg  <- tapply(bw_gr$score[subjectHits(hits)] * ov_w, queryHits(hits), sum)
    qi   <- as.integer(names(agg))
    result[qi] <- as.numeric(agg) / width(peaks)[qi]
  }
  result
}

# Verify chromosome names match between peaks and bigWig files
bw_chroms <- seqlevels(rtracklayer::BigWigFile(bw_files[1]))
cat("BigWig chromosomes (first 10):", paste(head(bw_chroms, 10), collapse = ", "), "\n")
cat("Peak chromosomes:", paste(seqlevels(consensus_peaks), collapse = ", "), "\n")
cat("Matching chromosomes:",
    length(intersect(bw_chroms, seqlevels(consensus_peaks))), "\n")

cat("Scoring", nrow(meta_bw), "bigWig files at", length(consensus_peaks),
    "peaks — this may take several minutes...\n")

signal_mat <- vapply(setNames(meta_bw$file, meta_bw$sample_id),
                     score_bw_at_peaks, peaks = consensus_peaks,
                     FUN.VALUE = numeric(length(consensus_peaks)))

# Replace any NAs with 0 (peaks with no coverage in a sample)
signal_mat[is.na(signal_mat)] <- 0

cat("Signal matrix:", nrow(signal_mat), "peaks ×", ncol(signal_mat), "samples\n")
cat("Fraction of peaks with zero signal across all samples:",
    round(mean(rowSums(signal_mat) == 0), 3), "\n")

# Remove peaks with zero signal in all samples; re-index to 1..N
keep <- rowSums(signal_mat) > 0
signal_mat      <- signal_mat[keep, ]
consensus_peaks <- consensus_peaks[keep]
rownames(signal_mat) <- seq_len(nrow(signal_mat))   # 1, 2, ..., N — matches queryHits() output
cat("Peaks after removing zero-signal peaks:", length(consensus_peaks), "\n")
```

# ___QC: all 24 samples___

```{r fig.height=4, fig.width=5}
lib_sizes <- colSums(signal_mat)
lib_df <- data.frame(
  sample_id = names(lib_sizes),
  total_signal = lib_sizes,
  cell_line = meta_bw$cell_line[match(names(lib_sizes), meta_bw$sample_id)],
  treatment = meta_bw$treatment[match(names(lib_sizes), meta_bw$sample_id)]
)

p_libsize <- ggplot(lib_df, aes(x = sample_id, y = total_signal / 1e6,
                                 fill = cell_line, alpha = treatment)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = COL_LINE) +
  scale_alpha_manual(values = c("DMSO" = 1, "RA" = 0.55)) +
  labs(title = "Total ATAC signal per sample",
       x = NULL, y = "Total signal (× 10⁶)", fill = "cell line", alpha = "treatment") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

p_libsize

pdf(paste0(params$out, "figures/bulkATAC_NBcelllines_library_sizes.pdf"),
    height = 4, width = 5, useDingbats = FALSE)
p_libsize; dev.off()
```

```{r fig.height=4.5, fig.width=5}
# PCA of log2-normalised signal across all 24 samples
log2_signal <- log2(signal_mat + 0.01)

pca_all  <- prcomp(t(log2_signal), scale. = TRUE, center = TRUE)
var_expl <- summary(pca_all)$importance[2, ] * 100

pca_df <- data.frame(
  PC1       = pca_all$x[, 1], PC2 = pca_all$x[, 2],
  sample_id = colnames(log2_signal)
) %>%
  left_join(meta_bw %>% dplyr::select(sample_id, cell_line, treatment), by = "sample_id")

p_pca_all <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cell_line, shape = treatment,
                                  label = sample_id)) +
  geom_point(size = 3, alpha = 0.9) +
  ggrepel::geom_text_repel(size = 2.5, max.overlaps = 20) +
  scale_color_manual(values = COL_LINE) +
  scale_shape_manual(values = c("DMSO" = 16, "RA" = 17)) +
  labs(title    = "PCA: all bulk ATAC-seq samples",
       subtitle = paste0("PC1 ", round(var_expl[1], 1), "%, PC2 ", round(var_expl[2], 1), "%"),
       x = paste0("PC1 (", round(var_expl[1], 1), "% var)"),
       y = paste0("PC2 (", round(var_expl[2], 1), "% var)"),
       color = "cell line", shape = "treatment")

p_pca_all

pdf(paste0(params$out, "figures/bulkATAC_NBcelllines_PCA_allSamples.pdf"),
    height = 4.5, width = 5, useDingbats = FALSE)
p_pca_all; dev.off()
```

# ___Differential accessibility: E6 (ATRXdel) vs p53 (ATRXwt) — DMSO___

```{r}
dmso_meta <- meta_bw[meta_bw$cell_line %in% c("E6", "p53") &
                      meta_bw$treatment == "DMSO", ]
dmso_meta$ATRX <- factor(ifelse(dmso_meta$cell_line == "E6", "ATRXdel", "ATRXwt"),
                          levels = c("ATRXwt", "ATRXdel"))

signal_dmso <- log2_signal[, dmso_meta$sample_id]

design <- model.matrix(~ ATRX, data = dmso_meta)
fit    <- lmFit(signal_dmso, design)
fit    <- eBayes(fit)

dap_dmso <- topTable(fit, coef = "ATRXATRXdel", number = Inf, sort.by = "P") %>%
  tibble::rownames_to_column("peak_id") %>%
  mutate(
    peak_id   = as.integer(peak_id),
    chr       = as.character(seqnames(consensus_peaks))[peak_id],
    start     = start(consensus_peaks)[peak_id],
    end       = end(consensus_peaks)[peak_id],
    sig       = adj.P.Val < 0.05 & abs(logFC) > 1,
    direction = case_when(
      sig & logFC > 0 ~ "more open ATRXdel",
      sig & logFC < 0 ~ "more open ATRXwt",
      TRUE            ~ "NS"
    )
  )

cat("DAP summary (adj.P < 0.05, |log2FC| > 1):\n")
print(table(dap_dmso$direction))

write.table(dap_dmso, sep = "\t", row.names = FALSE, quote = FALSE,
            file = paste0(params$out,
                          "tables/bulkATAC_NBcelllines_E6-vs-p53_DMSO_DAP_2026-06-25.csv"))
```

## ___Volcano: E6 vs p53 DMSO___

```{r fig.height=5, fig.width=4}
p_volcano_dmso <- ggplot(dap_dmso, aes(x = logFC, y = -log10(adj.P.Val),
                                        color = direction)) +
  geom_point(size = 0.8, alpha = 0.5) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray60", linewidth = 0.35) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray60", linewidth = 0.35) +
  scale_color_manual(values = c(
    "more open ATRXdel" = COL_ATRX["ATRXdel"],
    "more open ATRXwt"  = COL_ATRX["ATRXwt"],
    "NS"                = "gray80"
  )) +
  labs(title    = "Differential chromatin accessibility",
       subtitle = "E6 (ATRXdel) vs p53 (ATRXwt) — DMSO",
       x        = "log2 fold change (E6 / p53)",
       y        = "-log10 (adj. p-value)",
       color    = NULL) +
  theme(legend.position = "top")

p_volcano_dmso

pdf(paste0(params$out, "figures/bulkATAC_NBcelllines_E6-vs-p53_DMSO_volcano.pdf"),
    height = 5, width = 4, useDingbats = FALSE)
p_volcano_dmso; dev.off()
```

## ___MA plot___

```{r fig.height=4, fig.width=4}
dap_dmso$AveExpr_raw <- 2^dap_dmso$AveExpr

p_ma <- ggplot(dap_dmso, aes(x = AveExpr, y = logFC, color = direction)) +
  geom_point(size = 0.6, alpha = 0.4) +
  geom_hline(yintercept = c(-1, 0, 1), linetype = c("dashed","solid","dashed"),
             color = c("gray60","black","gray60"), linewidth = c(0.35,0.5,0.35)) +
  scale_color_manual(values = c(
    "more open ATRXdel" = COL_ATRX["ATRXdel"],
    "more open ATRXwt"  = COL_ATRX["ATRXwt"],
    "NS"                = "gray80"
  )) +
  labs(title = "MA plot: E6 vs p53 DMSO",
       x = "Average log2 expression", y = "log2 FC (E6 / p53)", color = NULL) +
  theme(legend.position = "top")

p_ma

pdf(paste0(params$out, "figures/bulkATAC_NBcelllines_E6-vs-p53_DMSO_MAplot.pdf"),
    height = 4, width = 4, useDingbats = FALSE)
p_ma; dev.off()
```

# ___Annotate peaks to nearest gene___

Map each consensus peak to the nearest TSS (TxDb hg38, ±50 kb cutoff).
Peaks within the TSS window (±`r params$tss_window` bp) are labelled "promoter".

```{r}
txdb  <- TxDb.Hsapiens.UCSC.hg38.knownGene
genes_hg38 <- genes(txdb)

gene_symbols <- mapIds(org.Hs.eg.db, keys = genes_hg38$gene_id,
                       keytype = "ENTREZID", column = "SYMBOL",
                       multiVals = "first")
genes_hg38$symbol <- gene_symbols

# Strand-aware TSS
tss_pos <- ifelse(as.character(strand(genes_hg38)) == "+",
                  start(genes_hg38), end(genes_hg38))
tss_gr <- GRanges(seqnames = seqnames(genes_hg38),
                  ranges   = IRanges(start = tss_pos, width = 1),
                  gene_id  = genes_hg38$gene_id,
                  symbol   = genes_hg38$symbol)

# For each peak, find nearest TSS
dist_to_tss <- distanceToNearest(consensus_peaks, tss_gr, ignore.strand = TRUE)

peak_anno <- data.frame(
  peak_id     = queryHits(dist_to_tss),          # integer index into consensus_peaks
  gene_id     = tss_gr$gene_id[subjectHits(dist_to_tss)],
  gene_symbol = tss_gr$symbol[subjectHits(dist_to_tss)],
  dist_to_tss = mcols(dist_to_tss)$distance
) %>%
  filter(dist_to_tss <= 50000) %>%
  mutate(region = case_when(
    dist_to_tss <= params$tss_window ~ "promoter",
    dist_to_tss <= 10000             ~ "gene-proximal",
    TRUE                             ~ "distal"
  ))

# Merge annotation into DAP table
dap_dmso <- dap_dmso %>%
  left_join(peak_anno %>%
              group_by(peak_id) %>%
              slice_min(dist_to_tss, n = 1, with_ties = FALSE),
            by = "peak_id")

cat("Peak region breakdown:\n")
print(table(peak_anno$region))
cat("\nTop DAPs with gene annotation:\n")
dap_dmso %>% filter(sig) %>% arrange(adj.P.Val) %>%
  dplyr::select(gene_symbol, logFC, adj.P.Val, region, chr, start, end) %>%
  head(20)
```

# ___Pathway analysis: chromatin accessibility signatures (DMSO)___

Gene-set over-representation (ORA) for the six accessibility gene sets (direction × region),
plus GSEA on the full ranked gene list.

```{r include=FALSE}
suppressPackageStartupMessages({
  library(clusterProfiler)
  library(DOSE)
  library(enrichplot)
  library(msigdbr)
})

hallmark_genes_H <- msigdbr(species = "human", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)

geneUniverse_atac <- unique(na.omit(peak_anno$gene_symbol))
geneUniverse_entrez <- na.omit(map_gene_to_entrez(geneUniverse_atac))
```

## ___Gene sets: direction × region___

```{r}
# All sig DAPs annotated (joined in dap_dmso)
sig_dap <- dap_dmso %>%
  filter(sig, !is.na(gene_symbol), !is.na(region))

# Define 6 gene sets
gene_sets_atac <- list(
  ATRXdel_promoter  = unique(sig_dap$gene_symbol[sig_dap$direction == "more open ATRXdel" & sig_dap$region == "promoter"]),
  ATRXdel_proximal  = unique(sig_dap$gene_symbol[sig_dap$direction == "more open ATRXdel" & sig_dap$region == "gene-proximal"]),
  ATRXdel_distal    = unique(sig_dap$gene_symbol[sig_dap$direction == "more open ATRXdel" & sig_dap$region == "distal"]),
  ATRXwt_promoter   = unique(sig_dap$gene_symbol[sig_dap$direction == "more open ATRXwt"  & sig_dap$region == "promoter"]),
  ATRXwt_proximal   = unique(sig_dap$gene_symbol[sig_dap$direction == "more open ATRXwt"  & sig_dap$region == "gene-proximal"]),
  ATRXwt_distal     = unique(sig_dap$gene_symbol[sig_dap$direction == "more open ATRXwt"  & sig_dap$region == "distal"])
)

cat("Gene set sizes:\n")
sapply(gene_sets_atac, length)
```

## ___Enrichment helpers___

```{r}
run_enrichment <- function(gene_symbols,
                           type = c("BP", "MF", "CC", "KEGG", "HALLMARK"),
                           term2gene = NULL,
                           label = "") {
  type <- match.arg(type)
  if (length(gene_symbols) == 0) {
    message(label, ": no input genes — skipping"); return(invisible(NULL))
  }
  entrez_ids <- na.omit(map_gene_to_entrez(gene_symbols))
  if (length(entrez_ids) == 0) {
    message(label, ": no valid Entrez IDs — skipping"); return(invisible(NULL))
  }
  tryCatch({
    if (type %in% c("BP", "MF", "CC")) {
      enrichGO(gene = entrez_ids, universe = geneUniverse_entrez,
               ont = type, OrgDb = "org.Hs.eg.db",
               readable = TRUE, pAdjustMethod = "bonferroni",
               pvalueCutoff = 0.05)
    } else if (type == "KEGG") {
      enrichKEGG(gene = entrez_ids, universe = geneUniverse_entrez,
                 organism = "hsa", keyType = "kegg",
                 pAdjustMethod = "bonferroni", pvalueCutoff = 0.05)
    } else if (type == "HALLMARK") {
      enricher(gene = entrez_ids, universe = as.character(geneUniverse_entrez),
               TERM2GENE = term2gene,
               pAdjustMethod = "bonferroni", pvalueCutoff = 0.05)
    }
  }, error = function(e) {
    message(label, " (", type, ") error: ", conditionMessage(e)); NULL
  })
}

show_enrichment <- function(result, label = "", out_path = NULL,
                            out_w = 5, out_h = 4.5) {
  if (is.null(result) || nrow(result@result) == 0) {
    message("0 enriched terms — ", label); return(invisible(NULL))
  }
  signif_rows <- result@result[result@result$p.adjust < 0.05, ]
  if (nrow(signif_rows) == 0) {
    message("0 terms with bonferroni p.adjust < 0.05 — ", label)
    return(invisible(NULL))
  }
  p <- dotplot(result, showCategory = 20) +
    ggtitle(label) + pub_theme +
    theme(axis.text.y = element_text(size = 7))
  print(p)
  if (!is.null(out_path)) {
    pdf(out_path, height = out_h, width = out_w, useDingbats = FALSE)
    print(p); dev.off()
  }
  invisible(result)
}
```

## ___ORA: ATRXdel open — promoter___

```{r fig.height=3, fig.width=3}
genes <- gene_sets_atac$ATRXdel_promoter
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXdel open — promoter: GO-BP"),
                "ATRXdel open — promoter: GO-BP",
                out_path = paste0(params$out, "figures/bulkATAC_ATRXdel-promoter_GO-BP.pdf"))
show_enrichment(run_enrichment(genes, "MF",       label = "ATRXdel open — promoter: GO-MF"),
                "ATRXdel open — promoter: GO-MF")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXdel open — promoter: KEGG"),
                "ATRXdel open — promoter: KEGG",
                out_path = paste0(params$out, "figures/bulkATAC_ATRXdel-promoter_KEGG.pdf"))
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXdel open — promoter: HALLMARK",
                               term2gene = hallmark_genes_H),
                "ATRXdel open — promoter: HALLMARK",
                out_path = paste0(params$out, "figures/bulkATAC_ATRXdel-promoter_HALLMARK.pdf"))
```

## ___ORA: ATRXdel open — gene-proximal___

```{r fig.height=3, fig.width=3}
genes <- gene_sets_atac$ATRXdel_proximal
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXdel open — proximal: GO-BP"),
                "ATRXdel open — proximal: GO-BP")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXdel open — proximal: KEGG"),
                "ATRXdel open — proximal: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXdel open — proximal: HALLMARK",
                               term2gene = hallmark_genes_H),
                "ATRXdel open — proximal: HALLMARK")
```

## ___ORA: ATRXdel open — distal___

```{r fig.height=3, fig.width=3}
genes <- gene_sets_atac$ATRXdel_distal
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXdel open — distal: GO-BP"),
                "ATRXdel open — distal: GO-BP")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXdel open — distal: KEGG"),
                "ATRXdel open — distal: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXdel open — distal: HALLMARK",
                               term2gene = hallmark_genes_H),
                "ATRXdel open — distal: HALLMARK")
```

## ___ORA: ATRXwt open — promoter___

```{r fig.height=3, fig.width=3}
genes <- gene_sets_atac$ATRXwt_promoter
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXwt open — promoter: GO-BP"),
                "ATRXwt open — promoter: GO-BP",
                out_path = paste0(params$out, "figures/bulkATAC_ATRXwt-promoter_GO-BP.pdf"))
show_enrichment(run_enrichment(genes, "MF",       label = "ATRXwt open — promoter: GO-MF"),
                "ATRXwt open — promoter: GO-MF")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXwt open — promoter: KEGG"),
                "ATRXwt open — promoter: KEGG",
                out_path = paste0(params$out, "figures/bulkATAC_ATRXwt-promoter_KEGG.pdf"))
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXwt open — promoter: HALLMARK",
                               term2gene = hallmark_genes_H),
                "ATRXwt open — promoter: HALLMARK",
                out_path = paste0(params$out, "figures/bulkATAC_ATRXwt-promoter_HALLMARK.pdf"))
```

## ___ORA: ATRXwt open — gene-proximal___

```{r fig.height=3, fig.width=3}
genes <- gene_sets_atac$ATRXwt_proximal
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXwt open — proximal: GO-BP"),
                "ATRXwt open — proximal: GO-BP")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXwt open — proximal: KEGG"),
                "ATRXwt open — proximal: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXwt open — proximal: HALLMARK",
                               term2gene = hallmark_genes_H),
                "ATRXwt open — proximal: HALLMARK")
```

## ___ORA: ATRXwt open — distal___

```{r fig.height=3, fig.width=3}
genes <- gene_sets_atac$ATRXwt_distal
show_enrichment(run_enrichment(genes, "BP",       label = "ATRXwt open — distal: GO-BP"),
                "ATRXwt open — distal: GO-BP")
show_enrichment(run_enrichment(genes, "KEGG",     label = "ATRXwt open — distal: KEGG"),
                "ATRXwt open — distal: KEGG")
show_enrichment(run_enrichment(genes, "HALLMARK", label = "ATRXwt open — distal: HALLMARK",
                               term2gene = hallmark_genes_H),
                "ATRXwt open — distal: HALLMARK")
```

## ___GSEA: ranked by logFC at best peak per gene___

Rank all annotated genes by the logFC of their best (most significant) associated DAP.
Positive rank = more open in ATRXdel; negative = more open in ATRXwt.

```{r fig.height=3, fig.width=4}
# For each gene, take the peak with the lowest adj.P.Val
gene_ranked <- dap_dmso %>%
  filter(!is.na(gene_symbol)) %>%
  group_by(gene_symbol) %>%
  slice_min(adj.P.Val, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(logFC))

ranked_vec <- setNames(gene_ranked$logFC, gene_ranked$gene_symbol)
ranked_entrez <- na.omit(map_gene_to_entrez(names(ranked_vec)))
ranked_vec_entrez <- ranked_vec[match(names(ranked_entrez), names(ranked_vec))]
names(ranked_vec_entrez) <- as.character(ranked_entrez)
ranked_vec_entrez <- sort(ranked_vec_entrez, decreasing = TRUE)
ranked_vec_entrez <- ranked_vec_entrez[!duplicated(names(ranked_vec_entrez))]

cat("Genes in ranked list:", length(ranked_vec_entrez), "\n")

# GSEA — Hallmark
hallmark_gsea_list <- msigdbr(species = "human", category = "H") %>%
  dplyr::select(gs_name, entrez_gene) %>%
  mutate(entrez_gene = as.character(entrez_gene))

gsea_hallmark <- tryCatch(
  GSEA(geneList  = ranked_vec_entrez,
       TERM2GENE = hallmark_gsea_list,
       pvalueCutoff = 0.05,
       pAdjustMethod = "bonferroni",
       verbose = FALSE),
  error = function(e) { message("GSEA Hallmark error: ", conditionMessage(e)); NULL }
)

if (!is.null(gsea_hallmark) && nrow(gsea_hallmark@result) > 0) {
  p_gsea_h <- dotplot(gsea_hallmark, showCategory = 20, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA — Hallmark (ranked by ATAC logFC, ATRXdel vs ATRXwt)") +
    pub_theme + theme(axis.text.y = element_text(size = 7))
  print(p_gsea_h)
  pdf(paste0(params$out, "figures/bulkATAC_GSEA_Hallmark_ATRXdel-vs-ATRXwt.pdf"),
      height = 5, width = 10, useDingbats = FALSE)
  print(p_gsea_h); dev.off()
} else {
  message("No significant Hallmark GSEA terms (adj.p < 0.05)")
}

# GSEA — GO biological process
gsea_go_bp <- tryCatch(
  gseGO(geneList     = ranked_vec_entrez,
        OrgDb        = org.Hs.eg.db,
        ont          = "BP",
        pvalueCutoff = 0.05,
        pAdjustMethod = "bonferroni",
        verbose = FALSE),
  error = function(e) { message("gseGO BP error: ", conditionMessage(e)); NULL }
)

if (!is.null(gsea_go_bp) && nrow(gsea_go_bp@result) > 0) {
  p_gsea_go <- dotplot(gsea_go_bp, showCategory = 20, split = ".sign") +
    facet_grid(. ~ .sign) +
    ggtitle("GSEA — GO-BP (ranked by ATAC logFC)") +
    pub_theme + theme(axis.text.y = element_text(size = 7))
  print(p_gsea_go)
  pdf(paste0(params$out, "figures/bulkATAC_GSEA_GO-BP_ATRXdel-vs-ATRXwt.pdf"),
      height = 5, width = 10, useDingbats = FALSE)
  print(p_gsea_go); dev.off()
} else {
  message("No significant GO-BP GSEA terms (adj.p < 0.05)")
}
```

# ___TF footprinting: DeepTools (ATRXdel vs ATRXwt — DMSO)___

Aggregate footprint profiles are computed with:

1. **motifmatchr** (R): scan consensus peaks for JASPAR2020 motifs → per-TF site BED files
2. **DeepTools `computeMatrix reference-point`**: extract bigWig signal ±250 bp around each motif centre
3. **DeepTools `plotProfile --outFileNameData`**: export per-sample average signal profiles
4. **ggplot2** (R): overlay ATRXdel vs ATRXwt mean ± SEM with pub_theme

Conditions: E6 (ATRXdel) vs p53 (ATRXwt) — DMSO replicates only.
TFs: `TFs` vector from `00_NK_paper_setup.R`.

## ___Setup___

```{r fp-setup}
suppressPackageStartupMessages({
  library(motifmatchr)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(TFBSTools)
  library(JASPAR2020)
  library(jsonlite)
})

ft_dir      <- file.path(params$out, "footprinting")
ft_bed_dir  <- file.path(ft_dir, "motif_beds")
ft_mat_dir  <- file.path(ft_dir, "matrices")
ft_prof_dir <- file.path(ft_dir, "profiles")
ft_fig_dir  <- file.path(params$out, "figures")
for (d in c(ft_bed_dir, ft_mat_dir, ft_prof_dir)) {
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
}

dt_ok <- Sys.which("computeMatrix") != ""
if (!dt_ok) message("computeMatrix (deepTools) not found — install via: pip install deeptools")

TFs_fp <- unique(TFs)
cat("TFs of interest:", paste(TFs_fp, collapse = ", "), "\n")

# DMSO bigWig files: ATRXdel (E6) first, then ATRXwt (p53)
dmso_meta   <- meta_bw[meta_bw$cell_line %in% c("E6", "p53") &
                        meta_bw$treatment == "DMSO", ]
dmso_meta   <- dmso_meta[order(dmso_meta$ATRX), ]   # ATRXdel < ATRXwt alphabetically
bws_del     <- dmso_meta$file[dmso_meta$cell_line == "E6"]
bws_wt      <- dmso_meta$file[dmso_meta$cell_line == "p53"]
bws_all     <- c(bws_del, bws_wt)
labels_all  <- c(dmso_meta$sample_id[dmso_meta$cell_line == "E6"],
                 dmso_meta$sample_id[dmso_meta$cell_line == "p53"])
n_del <- length(bws_del)
n_wt  <- length(bws_wt)
cat(sprintf("BigWig files: ATRXdel n=%d, ATRXwt n=%d\n", n_del, n_wt))

FP_WINDOW <- 250L   # bp upstream/downstream of motif centre
FP_BINSIZE <- 10L   # bp per bin in computeMatrix
```

## ___Motif scanning (motifmatchr)___

```{r fp-motifs}
# Fetch JASPAR2020 CORE vertebrate PWMs for TFs of interest
opts    <- list(collection = "CORE", tax_group = "vertebrates", all_versions = FALSE)
pfm_all <- TFBSTools::getMatrixSet(JASPAR2020, opts = opts)

tf_upper <- toupper(TFs_fp)
keep_idx <- sapply(pfm_all, function(m) toupper(TFBSTools::name(m)) %in% tf_upper)
pfm_tfs  <- pfm_all[keep_idx]

cat("Motifs retrieved for", length(pfm_tfs), "of", length(TFs_fp), "TFs\n")
if (length(pfm_tfs) > 0)
  cat("Matched:", paste(unique(sapply(pfm_tfs, TFBSTools::name)), collapse = ", "), "\n")

# Scan consensus peaks for motif occurrences
motif_hits <- motifmatchr::matchMotifs(
  pwms    = pfm_tfs,
  subject = consensus_peaks,
  genome  = BSgenome.Hsapiens.UCSC.hg38,
  out     = "positions",
  p.cutoff = 5e-5
)

# Write per-TF BED files of motif match centres (0-based, required by DeepTools)
tf_beds <- list()
for (i in seq_along(motif_hits)) {
  tf_name <- TFBSTools::name(pfm_tfs[[i]])
  hits_gr <- motif_hits[[i]]
  if (length(hits_gr) == 0) {
    message("No motif sites found for: ", tf_name); next
  }
  centres <- data.frame(
    chr    = as.character(seqnames(hits_gr)),
    start  = as.integer(start(hits_gr) + floor(width(hits_gr) / 2)) - 1L,   # 0-based
    end    = as.integer(start(hits_gr) + floor(width(hits_gr) / 2)),
    name   = paste0(tf_name, "_", seq_along(hits_gr)),
    score  = 0,
    strand = as.character(strand(hits_gr))
  )
  bed_path <- file.path(ft_bed_dir, paste0(tf_name, "_centres.bed"))
  write.table(centres, bed_path, sep = "\t", row.names = FALSE,
              col.names = FALSE, quote = FALSE)
  tf_beds[[tf_name]] <- bed_path
  cat(tf_name, ":", nrow(centres), "motif sites →", bed_path, "\n")
}
```

## ___computeMatrix: signal around motif centres___

```{r fp-computematrix}
if (dt_ok) {
  for (tf in names(tf_beds)) {
    mat_out <- file.path(ft_mat_dir, paste0(tf, "_matrix.gz"))
    if (file.exists(mat_out)) {
      cat(tf, ": matrix exists — skipping\n"); next
    }
    cmd <- paste(
      "computeMatrix reference-point",
      "--referencePoint center",
      sprintf("-b %d -a %d --binSize %d", FP_WINDOW, FP_WINDOW, FP_BINSIZE),
      "--missingDataAsZero --skipZeros",
      "-S", paste(bws_all, collapse = " "),
      "-R", tf_beds[[tf]],
      "-o", mat_out,
      "--samplesLabel", paste(labels_all, collapse = " "),
      "-p 4"
    )
    cat("Running computeMatrix:", tf, "\n")
    ret <- system(cmd)
    if (ret != 0) message("computeMatrix failed for: ", tf)
  }
}
```

## ___plotProfile: aggregate footprint PDFs___

```{r fp-plotprofile}
# Colour string: n_del replicates ATRXdel colour + n_wt replicates ATRXwt colour
col_del <- unname(COL_ATRX["ATRXdel"])
col_wt  <- unname(COL_ATRX["ATRXwt"])
colors_str <- paste(c(rep(col_del, n_del), rep(col_wt, n_wt)), collapse = " ")

if (dt_ok) {
  for (tf in names(tf_beds)) {
    mat_out  <- file.path(ft_mat_dir,  paste0(tf, "_matrix.gz"))
    prof_pdf <- file.path(ft_prof_dir, paste0(tf, "_profile.pdf"))
    prof_tab <- file.path(ft_prof_dir, paste0(tf, "_profile_data.tab"))
    if (!file.exists(mat_out)) next

    cmd <- paste(
      "plotProfile",
      "-m", mat_out,
      "--perGroup",
      "--colors", colors_str,
      sprintf("--plotTitle '%s footprint: ATRXdel vs ATRXwt'", tf),
      "--yAxisLabel 'Mean ATAC signal'",
      sprintf("--xAxisLabel 'Distance from motif centre (bp)'"),
      "--outFileName", prof_pdf,
      "--outFileNameData", prof_tab,
      "--dpi 150"
    )
    ret <- system(cmd)
    if (ret != 0) message("plotProfile failed for: ", tf)
  }
  cat("Footprint profile PDFs:", ft_prof_dir, "\n")
}
```

## ___ggplot2: footprint profiles per TF___

Parse the computeMatrix output (gzipped; JSON header + tab-delimited data) and plot
mean ± SEM per condition in ggplot2.

```{r fp-parse-matrix}
parse_fp_matrix <- function(matrix_gz) {
  con   <- gzfile(matrix_gz, "r")
  lines <- readLines(con)
  close(con)

  hdr <- jsonlite::fromJSON(sub("^@", "", lines[1]))
  n_samples  <- length(hdr$sample_labels)
  upstream   <- hdr[["upstream"]]
  downstream <- hdr[["downstream"]]
  bin_sz     <- hdr[["bin size"]]
  n_bins     <- (upstream + downstream) / bin_sz
  # Bin centres relative to motif (negative = upstream)
  bin_pos    <- seq(-upstream + bin_sz / 2, downstream - bin_sz / 2, by = bin_sz)

  data_lines <- lines[-1]
  data_lines <- data_lines[nzchar(data_lines)]
  mat <- do.call(rbind, lapply(data_lines, function(l) {
    parts <- strsplit(l, "\t")[[1]]
    suppressWarnings(as.numeric(parts[7:length(parts)]))  # cols 1-6: chr/start/end/name/score/strand
  }))

  if (ncol(mat) != n_bins * n_samples) {
    message("Unexpected matrix dimensions: ", nrow(mat), " x ", ncol(mat)); return(NULL)
  }

  # Per-sample mean signal per bin → named list
  sample_means <- lapply(seq_len(n_samples), function(i) {
    cols <- ((i - 1L) * n_bins + 1L):(i * n_bins)
    colMeans(mat[, cols, drop = FALSE], na.rm = TRUE)
  })
  names(sample_means) <- hdr$sample_labels

  list(bin_pos = bin_pos, sample_means = sample_means,
       sample_labels = hdr$sample_labels, n_bins = n_bins)
}
```

```{r fp-ggplot, fig.height=3, fig.width=4}
fp_plots <- list()

for (tf in names(tf_beds)) {
  mat_gz <- file.path(ft_mat_dir, paste0(tf, "_matrix.gz"))
  if (!file.exists(mat_gz)) next

  pf <- parse_fp_matrix(mat_gz)
  if (is.null(pf)) next

  del_samps <- pf$sample_labels[seq_len(n_del)]
  wt_samps  <- pf$sample_labels[seq(n_del + 1L, length(pf$sample_labels))]

  del_mat <- do.call(rbind, pf$sample_means[del_samps])
  wt_mat  <- do.call(rbind, pf$sample_means[wt_samps])

  long_df <- dplyr::bind_rows(
    data.frame(
      position  = pf$bin_pos,
      mean_sig  = colMeans(del_mat, na.rm = TRUE),
      sem       = apply(del_mat, 2, sd, na.rm = TRUE) / sqrt(nrow(del_mat)),
      condition = "ATRXdel"
    ),
    data.frame(
      position  = pf$bin_pos,
      mean_sig  = colMeans(wt_mat, na.rm = TRUE),
      sem       = apply(wt_mat, 2, sd, na.rm = TRUE) / sqrt(nrow(wt_mat)),
      condition = "ATRXwt"
    )
  ) %>%
    dplyr::mutate(condition = factor(condition, levels = c("ATRXdel", "ATRXwt")))

  p <- ggplot(long_df, aes(x = position, y = mean_sig,
                            color = condition, fill = condition)) +
    geom_ribbon(aes(ymin = mean_sig - sem, ymax = mean_sig + sem),
                alpha = 0.15, color = NA) +
    geom_line(linewidth = 0.7) +
    geom_vline(xintercept = 0, linetype = "dashed",
               color = "gray50", linewidth = 0.35) +
    scale_color_manual(values = COL_ATRX, name = NULL) +
    scale_fill_manual( values = COL_ATRX, name = NULL) +
    scale_x_continuous(
      breaks = c(-FP_WINDOW, -FP_WINDOW %/% 2L, 0L,
                  FP_WINDOW %/% 2L, FP_WINDOW),
      labels = c(paste0("-", FP_WINDOW), paste0("-", FP_WINDOW %/% 2L),
                 "0", as.character(FP_WINDOW %/% 2L), as.character(FP_WINDOW))
    ) +
    labs(
      title    = paste(tf, "— ATAC footprint"),
      subtitle = "ATRXdel (E6) vs ATRXwt (p53), DMSO; ribbon = ±SEM",
      x        = "Distance from motif centre (bp)",
      y        = "Mean ATAC-seq signal"
    ) +
    pub_theme +
    theme(legend.position = "top")

  fp_plots[[tf]] <- p
  print(p)

  ggsave(file.path(ft_fig_dir,
                   paste0("bulkATAC_footprint_", tf, "_ATRXdel-vs-ATRXwt.pdf")),
         p, width = 4, height = 3, device = cairo_pdf)
}
```

## ___Combined footprint panel (all TFs)___

```{r fp-combined, fig.height=6, fig.width=9}
if (length(fp_plots) > 0) {
  n_col      <- min(3L, length(fp_plots))
  n_row      <- ceiling(length(fp_plots) / n_col)
  p_combined <- patchwork::wrap_plots(fp_plots, ncol = n_col) +
    patchwork::plot_layout(guides = "collect") &
    theme(legend.position = "bottom")

  print(p_combined)
  ggsave(
    file.path(ft_fig_dir, "bulkATAC_footprints_allTFs_ATRXdel-vs-ATRXwt.pdf"),
    p_combined,
    width  = n_col * 4,
    height = n_row * 3,
    device = cairo_pdf
  )
} else {
  message("No footprint matrices found — run computeMatrix chunks first.")
}
```

# ___The End___

```{r}
sessionInfo()
```

```{r render, eval=FALSE, include=FALSE}
rmarkdown::render(
  input         = "/home/rstudio/NK-NB_study/code/38_20260625_bulkATAC_NBcelllines_ATRXdel_vs_ATRXwt.Rmd",
  output_format = "html_document",
  output_file   = "38_20260625_bulkATAC_NBcelllines_ATRXdel_vs_ATRXwt.html",
  output_dir    = "/home/rstudio/mnt_out/NK_project/"
)
```
