---
title: "NK project: snATAC-seq — Data prep (module scoring)"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  out:  "/home/rstudio/mnt_out/NK_project/"
  anno: "/home/rstudio/mnt_resources/"
---

Preprocessing script — run **before** 39a_, 39b_, and 39c_.

Loads the ArchR project, adds per-cell module scores for M1–M25, and saves
`proj_filt` + `score_df` as Rds files that all three downstream scripts consume.

Outputs:
- `Rds/snATAC_NK_v20260625_modules.Rds`  — ArchR project with module scores
- `Rds/snATAC_NK_score_df_v20260625.Rds` — score data frame (cells × modules)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = FALSE,
                      error = FALSE,
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = "future::multicore", workers = 20)
options(future.globals.maxSize = 40 * 10^9)
print(params)
```

```{r include=FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
```

# ___Load packages___

```{r include=FALSE}
suppressPackageStartupMessages({
  library(ArchR)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(JASPAR2024)
  library(TFBSTools)
  library(motifmatchr)
  library(RSQLite)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrepel)
  library(patchwork)
  library(tidyverse)
})

addArchRThreads(threads = 20)
addArchRGenome("hg38")

filter    <- dplyr::filter
select    <- dplyr::select
mutate    <- dplyr::mutate
summarise <- dplyr::summarise
rename    <- dplyr::rename
count     <- dplyr::count
group_by  <- dplyr::group_by
ungroup   <- dplyr::ungroup
arrange   <- dplyr::arrange
slice_min <- dplyr::slice_min
sub       <- base::sub
gsub      <- base::gsub

# Remap group label strings to canonical names used throughout the script.
# Logic: anything with "atrx" but no "wt" → ATRXdel (catches ATRXdel/ATRXmut/ATRXm/ATRX_del etc.)
#        anything with "atrx" and "wt"    → ATRXwtMYCNwt
remap_group <- function(x) {
  x <- as.character(x)
  x[grepl("atrx", x, ignore.case = TRUE) & !grepl("wt", x, ignore.case = TRUE)] <- "ATRXdel"
  x[grepl("atrx", x, ignore.case = TRUE) &  grepl("wt", x, ignore.case = TRUE)] <- "ATRXwtMYCNwt"
  x[grepl("mycn", x, ignore.case = TRUE) &  grepl("amp", x, ignore.case = TRUE)] <- "MYCNamp"
  x[grepl("^control$",            x, ignore.case = TRUE)] <- "control"
  x
}

# Extract canonical group name from an ArchR clust_group column name.
# ArchR appends "._.Rep1" to coverage column names, e.g. "C2_control._.Rep1".
clust_col_to_group <- function(col_names) {
  # strip cluster prefix (everything up to and including the first "_")
  suffix <- sub("^[^_]+_", "", col_names)
  # strip ArchR's rep suffix (e.g. "._.Rep1")
  suffix <- sub("\\._\\.Rep.*$", "", suffix)
  remap_group(suffix)
}
```

```{r}
theme_set(pub_theme)

COL_ATAC_CLUSTER <- c(
  "C1" = "#1B4F8A",
  "C2" = "#2E7BB5",
  "C3" = "#6AADD5",
  "C4" = "#4A9B6F",
  "C5" = "#8FC494",
  "C6" = "#C0392B",
  "C7" = "#E8837A"
)

COL_GROUP <- c(
  "control"      = unname(COLOR_CODE_GROUP["control"]),
  "ATRXdel"      = unname(COLOR_CODE_GROUP["ATRXdel"]),
  "MYCNamp"      = unname(COLOR_CODE_GROUP["MYCNamp"]),
  "ATRXwtMYCNwt" = unname(COLOR_CODE_GROUP["ATRXwtMYCNwt"])
)

GROUP_ORDER   <- c( "control","ATRXdel", "MYCNamp", "ATRXwtMYCNwt")
CLUSTER_ORDER <- paste0("C", 1:7)

CLUSTER_NAMES <- c(
  "C1" = "CD56bright 1",
  "C2" = "CD56bright 2",
  "C3" = "CD56bright 3",
  "C4" = "Intermediate 1",
  "C5" = "Intermediate 2",
  "C6" = "CD56dim 1",
  "C7" = "CD56dim 2"
)

METACLUSTER <- c(
  "C1" = "CD56bright", "C2" = "CD56bright", "C3" = "CD56bright",
  "C4" = "Intermediate", "C5" = "Intermediate",
  "C6" = "CD56dim", "C7" = "CD56dim"
)
```

# ___Load ArchR project___

```{r}
proj_filt <- readRDS(file = paste0(params$out, "ArchR_v20260203/ATAC_analysis_v20260203.Rds"))

# Update output directory and Arrow file paths to current mount point
proj_filt@projectMetadata$outputDirectory <- paste0(params$out, "ArchR_v20260203")
proj_filt@sampleColData$ArrowFiles <- paste0(
  params$out, "ArchR_v20260203/ArrowFiles/",
  basename(as.character(proj_filt@sampleColData$ArrowFiles))
)

cat("Arrow file path set to:\n",
    as.character(proj_filt@sampleColData$ArrowFiles), "\n")
cat("Arrow file exists:", file.exists(as.character(proj_filt@sampleColData$ArrowFiles)), "\n")

# Inspect raw values before any conversion — catches spelling mismatches
raw_groups   <- sort(unique(as.character(proj_filt$group)))
raw_clusters <- sort(unique(as.character(proj_filt$Clusters_res1)))
cat("Raw group values in project:  ", paste(raw_groups,   collapse = ", "), "\n")
cat("Raw cluster values in project:", paste(raw_clusters, collapse = ", "), "\n")

proj_filt$group <- factor(remap_group(proj_filt$group), levels = GROUP_ORDER)

na_count <- sum(is.na(proj_filt$group))
if (na_count > 0) {
  warning(na_count, " cells have NA group after remapping — check raw_groups above")
} else {
  cat("Group remapping OK. Canonical values:", paste(levels(proj_filt$group), collapse = ", "), "\n")
  print(table(proj_filt$group, useNA = "always"))
}

# Add metadata
proj_filt$Cluster_name  <- CLUSTER_NAMES[as.character(proj_filt$Clusters_res1)]
proj_filt$Metacluster   <- METACLUSTER[as.character(proj_filt$Clusters_res1)]
proj_filt$Clusters_res1 <- factor(proj_filt$Clusters_res1, levels = CLUSTER_ORDER)

cat("\nCells:", nrow(getCellColData(proj_filt)), "\n")
cat("Clusters (Clusters_res1):\n")
print(table(proj_filt$Clusters_res1))
cat("\nGenetic groups (after remap):\n")
print(table(proj_filt$group, useNA = "always"))
cat("\nCluster x group:\n")
print(table(proj_filt$Clusters_res1, proj_filt$group))
```

# ___NK cell gene module scoring (M1–M25)___

Adds per-cell gene-activity module scores to `proj_filt` using
`addModuleScore()` (GeneScoreMatrix). Results are stored both inside the
ArchR project (as `proj_filt$M1.score`, ...) and in `score_df`.

```{r}
library(ArchR)
library(SummarizedExperiment)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)

set.seed(42)

# Resolve Bioconductor/plyr conflicts with dplyr
filter    <- dplyr::filter
select    <- dplyr::select
mutate    <- dplyr::mutate
summarise <- dplyr::summarise
rename    <- dplyr::rename
count     <- dplyr::count

module_dir   <- "/home/rstudio/mnt_out/NK_project/tables/NK_module_genes"
cluster_col  <- "Clusters_res1"
group_col    <- "group"
z_cap        <- 2.0

group_order   <- c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt")
cluster_order <- paste0("C", 1:7)

# Load gene lists for M1-M25
module_files <- list.files(module_dir, pattern = "module_\\d+_genes\\.csv",
                           full.names = TRUE)
module_nums  <- as.integer(gsub(".*module_(\\d+)_genes\\.csv", "\\1",
                                basename(module_files)))
module_files <- module_files[order(module_nums)]
module_genes <- lapply(module_files, function(f) read.csv(f)$gene)
names(module_genes) <- paste0("M", sort(module_nums))
message("Loaded ", length(module_genes), " modules: ",
        paste(names(module_genes), collapse = ", "))

# ---- ArchR native addModuleScore (GeneScoreMatrix) — one module at a time ----
# Passing all modules as a named list fails in current ArchR (assigns 25 values
# to all cells instead of per-cell scores). Add each module individually instead,
# matching the pattern used for MatureNK / AdaptiveNK below.
gsm_genes <- getFeatures(proj_filt, useMatrix = "GeneScoreMatrix")
for (mod_name in names(module_genes)) {
  genes_valid <- intersect(module_genes[[mod_name]], gsm_genes)
  if (length(genes_valid) == 0) {
    message("No valid genes for ", mod_name, " in GeneScoreMatrix — skipping")
    next
  }
  message(mod_name, ": ", length(genes_valid), "/", length(module_genes[[mod_name]]),
          " genes found in GeneScoreMatrix")
  proj_filt <- addModuleScore(
    ArchRProj = proj_filt,
    useMatrix = "GeneScoreMatrix",
    name      = mod_name,
    features  = list(score = genes_valid)
  )
}

# Retrieve scores — ArchR names columns "<name>.score", e.g. "M1.score"
module_cols <- paste0(names(module_genes), ".score")
score_raw   <- as.data.frame(
  getCellColData(proj_filt, select = c(cluster_col, group_col, module_cols))
)
colnames(score_raw)[1:2] <- c("cluster", "group")
score_raw$group <- remap_group(as.character(score_raw$group))

# Rename "M1.score" → "M1" etc. for downstream compatibility
score_df        <- score_raw
colnames(score_df)[3:(2 + length(module_genes))] <- names(module_genes)
score_df$cell   <- rownames(score_df)

stopifnot(sum(!is.na(score_df$cluster)) > 0)
message("score_df: ", nrow(score_df), " cells, ",
        sum(!is.na(score_df$M1)), " with M1 scores")

# ___Save___

```{r}
dir.create(paste0(params$out, "Rds"), showWarnings = FALSE, recursive = TRUE)

saveRDS(proj_filt,
        file = paste0(params$out, "Rds/snATAC_NK_v20260625_modules.Rds"))
saveRDS(score_df,
        file = paste0(params$out, "Rds/snATAC_NK_score_df_v20260625.Rds"))

message("Saved proj_filt (",
        nCells(proj_filt), " cells) → snATAC_NK_v20260625_modules.Rds")
message("Saved score_df (", nrow(score_df), " rows) → snATAC_NK_score_df_v20260625.Rds")
```

# ___The End___

```{r render, eval=FALSE, include=FALSE}
rmarkdown::render(
  "/home/rstudio/code/NK-NB_study/code/39_00_20260625_snATAC_NK_prep.Rmd",
  output_file = paste0(params$out, "html/39_00_20260625_snATAC_NK_prep.html")
)
```
