---
title: "NK project: snATAC-seq — Module scoring and GRN analysis"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  out:  "/home/rstudio/mnt_out/NK_project/"
  anno: "/home/rstudio/mnt_resources/"
---

PROJECT DESCRIPTION

NK cell snATAC-seq re-analysed with unsupervised LSI + clustering at resolution 1,
yielding 7 distinct chromatin-accessibility clusters (C1-C7).
Clusters are annotated as CD56bright (C1-C3), Intermediate (C4-C5), and CD56dim (C6-C7).

Questions:
1. Do TFs relevant to NK cell development and function show differential footprinting
   between genetic patient groups (ATRXdel, control, MYCNamp, ATRXwtMYCNwt) within
   each of the 7 ATAC clusters?
2. Is the GSK3B gene locus and promoter differentially accessible between genetic groups
   within each cluster?
   
Requires: `39_` completed first (saves `Rds/snATAC_NK_v20260625_modules.Rds`).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = FALSE,
                      error = FALSE,
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = "future::multicore", workers = 20)
options(future.globals.maxSize = 40 * 10^9)
print(params)
```

```{r include=FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
```

# ___Load packages___

```{r include=FALSE}
suppressPackageStartupMessages({
  library(ArchR)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(JASPAR2024)
  library(TFBSTools)
  library(motifmatchr)
  library(RSQLite)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrepel)
  library(patchwork)
  library(tidyverse)
})

addArchRThreads(threads = 20)
addArchRGenome("hg38")

filter    <- dplyr::filter
select    <- dplyr::select
mutate    <- dplyr::mutate
summarise <- dplyr::summarise
rename    <- dplyr::rename
count     <- dplyr::count
group_by  <- dplyr::group_by
ungroup   <- dplyr::ungroup
arrange   <- dplyr::arrange
slice_min <- dplyr::slice_min
sub       <- base::sub
gsub      <- base::gsub

# Remap group label strings to canonical names used throughout the script.
# Logic: anything with "atrx" but no "wt" → ATRXdel (catches ATRXdel/ATRXmut/ATRXm/ATRX_del etc.)
#        anything with "atrx" and "wt"    → ATRXwtMYCNwt
remap_group <- function(x) {
  x <- as.character(x)
  x[grepl("atrx", x, ignore.case = TRUE) & !grepl("wt", x, ignore.case = TRUE)] <- "ATRXdel"
  x[grepl("atrx", x, ignore.case = TRUE) &  grepl("wt", x, ignore.case = TRUE)] <- "ATRXwtMYCNwt"
  x[grepl("mycn", x, ignore.case = TRUE) &  grepl("amp", x, ignore.case = TRUE)] <- "MYCNamp"
  x[grepl("^control$",            x, ignore.case = TRUE)] <- "control"
  x
}

# Extract canonical group name from an ArchR clust_group column name.
# ArchR appends "._.Rep1" to coverage column names, e.g. "C2_control._.Rep1".
clust_col_to_group <- function(col_names) {
  # strip cluster prefix (everything up to and including the first "_")
  suffix <- sub("^[^_]+_", "", col_names)
  # strip ArchR's rep suffix (e.g. "._.Rep1")
  suffix <- sub("\\._\\.Rep.*$", "", suffix)
  remap_group(suffix)
}
```

```{r}
theme_set(pub_theme)

COL_ATAC_CLUSTER <- c(
  "C1" = "#1B4F8A",
  "C2" = "#2E7BB5",
  "C3" = "#6AADD5",
  "C4" = "#4A9B6F",
  "C5" = "#8FC494",
  "C6" = "#C0392B",
  "C7" = "#E8837A"
)

COL_GROUP <- c(
  "control"      = unname(COLOR_CODE_GROUP["control"]),
  "ATRXdel"      = unname(COLOR_CODE_GROUP["ATRXdel"]),
  "MYCNamp"      = unname(COLOR_CODE_GROUP["MYCNamp"]),
  "ATRXwtMYCNwt" = unname(COLOR_CODE_GROUP["ATRXwtMYCNwt"])
)

GROUP_ORDER   <- c( "control","ATRXdel", "MYCNamp", "ATRXwtMYCNwt")
CLUSTER_ORDER <- paste0("C", 1:7)

CLUSTER_NAMES <- c(
  "C1" = "CD56bright 1",
  "C2" = "CD56bright 2",
  "C3" = "CD56bright 3",
  "C4" = "Intermediate 1",
  "C5" = "Intermediate 2",
  "C6" = "CD56dim 1",
  "C7" = "CD56dim 2"
)

METACLUSTER <- c(
  "C1" = "CD56bright", "C2" = "CD56bright", "C3" = "CD56bright",
  "C4" = "Intermediate", "C5" = "Intermediate",
  "C6" = "CD56dim", "C7" = "CD56dim"
)
```

# ___Load from prep script (39_00_)___

Requires `39_00_20260625_snATAC_NK_prep.Rmd` to have been rendered first.

```{r}
proj_filt <- readRDS(paste0(params$out, "Rds/snATAC_NK_v20260625_modules.Rds"))
score_df  <- readRDS(paste0(params$out, "Rds/snATAC_NK_score_df_v20260625.Rds"))

# Restore Arrow file paths to current mount point
proj_filt@projectMetadata$outputDirectory <- paste0(params$out, "ArchR_v20260203")
proj_filt@sampleColData$ArrowFiles <- paste0(
  params$out, "ArchR_v20260203/ArrowFiles/",
  basename(as.character(proj_filt@sampleColData$ArrowFiles))
)

cat("Loaded proj_filt:", nCells(proj_filt), "cells\n")
cat("Loaded score_df:", nrow(score_df), "rows,",
    sum(grepl("^M\\d+$", colnames(score_df))), "module columns\n")
print(table(proj_filt$Clusters_res1, proj_filt$group))
```

# ___PART I___

# ___UMAP: cluster overview___

```{r fig.height=4.5, fig.width=9}
p_umap_cluster <- plotEmbedding(
  proj_filt,
  colorBy    = "cellColData",
  name       = "Clusters_res1",
  embedding  = "UMAP",
  pal        = COL_ATAC_CLUSTER,
  labelMeans = TRUE,
  rastr      = FALSE
) + ggtitle("snATAC-seq NK cells — 7 clusters") + pub_theme

p_umap_group <- plotEmbedding(
  proj_filt,
  colorBy    = "cellColData",
  name       = "group",
  embedding  = "UMAP",
  pal        = COL_GROUP,
  rastr      = FALSE
) + ggtitle("Genetic group") + pub_theme

p_umap_cluster | p_umap_group

pdf(paste0(params$out, "figures/snATAC_NK_UMAP_clusters-and-groups.pdf"),
    height = 4.5, width = 9, useDingbats = FALSE)
p_umap_cluster | p_umap_group; dev.off()
```


# ___Cell composition: genetic groups per cluster___

```{r fig.height=4, fig.width=5}
cell_meta <- data.frame(
  Clusters_res1 = factor(as.character(proj_filt$Clusters_res1), levels = CLUSTER_ORDER),
  group         = factor(as.character(proj_filt$group), levels = GROUP_ORDER),
  row.names     = getCellNames(proj_filt),
  stringsAsFactors = FALSE
)

comp_df <- cell_meta %>%
  count(Clusters_res1, group) %>%
  group_by(Clusters_res1) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_comp <- ggplot(comp_df, aes(x = Clusters_res1, y = pct, fill = group)) +
  geom_col(width = 0.75) +
  scale_fill_manual(values = COL_GROUP) +
  labs(title    = "Cell composition per ATAC cluster",
       subtitle = "Percentage of cells per genetic group",
       x = "ATAC cluster", y = "% cells", fill = "genetic group") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p_comp

pdf(paste0(params$out, "figures/snATAC_NK_composition_per_cluster.pdf"),
    height = 4, width = 5, useDingbats = FALSE)
p_comp; dev.off()
```

# ___Cell composition: NK functional subsets (label transfer)___

Label-transfer from matched scRNA-seq data assigns each snATAC-seq cell to one of the
four NK maturation states (hNK_Bm1–4).
- **Immunomodulatory**: hNK_Bm2 (CD56bright immature), hNK_Bm3 (transitional)
- **Cytotoxic**: hNK_Bm1 (CD56dim mature), hNK_Bm4 (CD56dim highly cytotoxic)

```{r}
NK_SUBSET_ORDER <- c("hNK_Bm1", "hNK_Bm2", "hNK_Bm3", "hNK_Bm4")
COL_NK_SUBSET   <- COLOR_CODE_v2[NK_SUBSET_ORDER]

FUNCTIONAL_MAP <- c(
  "hNK_Bm1" = "cytotoxic",
  "hNK_Bm2" = "immunomodulatory",
  "hNK_Bm3" = "immunomodulatory",
  "hNK_Bm4" = "cytotoxic"
)
COL_FUNCTIONAL   <- c("immunomodulatory" = "#4A9B6F", "cytotoxic" = "#C0392B")
FUNCTIONAL_ORDER <- c("immunomodulatory", "cytotoxic")

# Build metadata — majority_vote_label = label transfer column
lt_meta <- data.frame(
  Clusters_res1  = factor(as.character(proj_filt$Clusters_res1), levels = CLUSTER_ORDER),
  group          = factor(as.character(proj_filt$group), levels = GROUP_ORDER),
  nk_subset      = factor(as.character(proj_filt$majority_vote_label), levels = NK_SUBSET_ORDER),
  row.names      = getCellNames(proj_filt),
  stringsAsFactors = FALSE
) %>%
  mutate(functional = factor(FUNCTIONAL_MAP[as.character(nk_subset)],
                             levels = FUNCTIONAL_ORDER))

cat("Label transfer subset distribution:\n")
print(table(lt_meta$nk_subset, useNA = "always"))
cat("\nFunctional category distribution:\n")
print(table(lt_meta$functional, useNA = "always"))
```

## ___NK subset per ATAC cluster___

```{r fig.height=4, fig.width=9}
df_cl_sub <- lt_meta %>%
  filter(!is.na(nk_subset)) %>%
  count(Clusters_res1, nk_subset) %>%
  group_by(Clusters_res1) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_cl_sub <- ggplot(df_cl_sub, aes(x = Clusters_res1, y = pct, fill = nk_subset)) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = COL_NK_SUBSET) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "NK maturation subset per ATAC cluster",
       x = "ATAC cluster", y = "% cells", fill = "NK subset") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

df_cl_func <- lt_meta %>%
  filter(!is.na(functional)) %>%
  count(Clusters_res1, functional) %>%
  group_by(Clusters_res1) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_cl_func <- ggplot(df_cl_func, aes(x = Clusters_res1, y = pct, fill = functional)) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = COL_FUNCTIONAL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Cytotoxic vs immunomodulatory per ATAC cluster",
       x = "ATAC cluster", y = "% cells", fill = NULL) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p_cl_sub | p_cl_func

pdf(paste0(params$out, "figures/snATAC_NK_NKsubset_per_cluster.pdf"),
    height = 4, width = 9, useDingbats = FALSE)
p_cl_sub | p_cl_func; dev.off()
```

## ___NK subset per genetic group___

```{r fig.height=4, fig.width=9}
df_grp_sub <- lt_meta %>%
  filter(!is.na(nk_subset), !is.na(group)) %>%
  count(group, nk_subset) %>%
  group_by(group) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_grp_sub <- ggplot(df_grp_sub, aes(x = group, y = pct, fill = nk_subset)) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = COL_NK_SUBSET) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "NK maturation subset per genetic group",
       x = NULL, y = "% cells", fill = "NK subset") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

df_grp_func <- lt_meta %>%
  filter(!is.na(functional), !is.na(group)) %>%
  count(group, functional) %>%
  group_by(group) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_grp_func <- ggplot(df_grp_func, aes(x = group, y = pct, fill = functional)) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = COL_FUNCTIONAL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Cytotoxic vs immunomodulatory per genetic group",
       x = NULL, y = "% cells", fill = NULL) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p_grp_sub | p_grp_func

pdf(paste0(params$out, "figures/snATAC_NK_NKsubset_per_group.pdf"),
    height = 4, width = 9, useDingbats = FALSE)
p_grp_sub | p_grp_func; dev.off()
```


# ___NK cell gene modules___

I defined 25 gene modules based on integrated ATAC and RNA data. Here I want to
address the function of each modules. 

```{r}
library(ArchR)
library(SummarizedExperiment)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)

set.seed(42)

# Resolve Bioconductor/plyr conflicts with dplyr
filter    <- dplyr::filter
select    <- dplyr::select
mutate    <- dplyr::mutate
summarise <- dplyr::summarise
rename    <- dplyr::rename
count     <- dplyr::count

module_dir   <- "/home/rstudio/mnt_out/NK_project/tables/NK_module_genes"
cluster_col  <- "Clusters_res1"
group_col    <- "group"
z_cap        <- 2.0

group_order   <- c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt")
cluster_order <- paste0("C", 1:7)

# Load gene lists for M1-M25
module_files <- list.files(module_dir, pattern = "module_\\d+_genes\\.csv",
                           full.names = TRUE)
module_nums  <- as.integer(gsub(".*module_(\\d+)_genes\\.csv", "\\1",
                                basename(module_files)))
module_files <- module_files[order(module_nums)]
module_genes <- lapply(module_files, function(f) read.csv(f)$gene)
names(module_genes) <- paste0("M", sort(module_nums))
message("Loaded ", length(module_genes), " modules: ",
        paste(names(module_genes), collapse = ", "))

# M1-M25 module scores already computed by 39_00_ and loaded via score_df above.
message("score_df: ", nrow(score_df), " cells, ",
        sum(!is.na(score_df$M1)), " with M1 scores")

# AGGREGATE: MEAN PER CLUSTER × GROUP
agg <- score_df %>%
  pivot_longer(cols = all_of(names(module_genes)),
               names_to  = "Module",
               values_to = "Score") %>%
  filter(!is.na(cluster), !is.na(group), !is.na(Score)) %>%
  group_by(cluster, group, Module) %>%
  summarise(MeanScore = mean(Score, na.rm = TRUE),
            nCells    = dplyr::n(),
            .groups   = "drop")

# warn if any combination has very few cells
sparse <- filter(agg, nCells < 10)
if (nrow(sparse) > 0) {
  message("Warning: ", nrow(sparse),
          " cluster × group × module combos have <10 cells — interpret with caution!!")
}

# Z-SCORE PER MODULE ACROSS ALL CLUSTER×GROUP COMBOS
agg <- agg %>%
  group_by(Module) %>%
  mutate(ZScore = as.numeric(scale(MeanScore))) %>%
  ungroup() %>%
  mutate(ZScore = pmax(pmin(ZScore, z_cap), -z_cap))   # cap extremes

# FACTOR ORDERING
agg$Module  <- factor(agg$Module,  levels = paste0("M", 1:25))
agg$cluster <- factor(agg$cluster, levels = cluster_order)
agg$group   <- factor(agg$group,   levels = group_order)

# x-axis labels: "C1|ATRXdel", sorted cluster-first, group-second
agg <- agg %>%
  mutate(clust_group = paste0(cluster, "|", group))

x_levels <- agg %>%
  distinct(clust_group, cluster, group) %>%
  arrange(cluster, group) %>%
  pull(clust_group)

agg$clust_group <- factor(agg$clust_group, levels = x_levels)

mat <- agg %>%
  select(Module, clust_group, ZScore) %>%
  pivot_wider(names_from  = clust_group,
              values_from = ZScore) %>%
  column_to_rownames("Module") %>%
  as.matrix()

mat <- mat[intersect(paste0("M", 1:25), rownames(mat)),
           intersect(x_levels,          colnames(mat)),
           drop = FALSE]

# column anno
col_meta <- agg %>%
  distinct(clust_group, cluster, group) %>%
  arrange(factor(clust_group, levels = x_levels))

# Extract unique present values as plain character — avoids NA-in-levels from ArchR factors
clust_levels <- intersect(cluster_order,
                          unique(na.omit(as.character(col_meta$cluster))))
group_levels <- intersect(group_order,
                          unique(na.omit(as.character(col_meta$group))))

ha_top <- HeatmapAnnotation(
  Cluster = factor(as.character(col_meta$cluster), levels = clust_levels),
  Group   = factor(as.character(col_meta$group),   levels = group_levels),
  col     = list(
    "Cluster" = COLOR_CODE_perCLUST[clust_levels],
    "Group"   = COLOR_CODE_GROUP[group_levels]
  ),
  annotation_name_side = "left",
  annotation_name_gp   = gpar(fontsize = 8),
  simple_anno_size     = unit(3, "mm"),
  gap                  = unit(1, "mm")
)

# color scale
col_fun <- colorRamp2(
  c(-z_cap, 0, z_cap),
  c("#9B59B6","#FFFFFF","#1E8449")
)

# plot heatmap
# column_split by cluster creates visual separation between the 7 cluster blocks

ht <- Heatmap(
  mat,
  name                  = "RNA Z-score",
  col                   = col_fun,
  top_annotation        = ha_top,

  # Row order: fixed M1→M25 (change to TRUE to cluster modules)
  cluster_rows          = FALSE,
  row_names_side        = "left",
  row_names_gp          = gpar(fontsize = 9),
  row_title             = NULL,

  # Column order: fixed, split by cluster
  cluster_columns       = FALSE,
  column_split          = col_meta$cluster,    # gap between clusters
  column_gap            = unit(2, "mm"),
  column_title_gp       = gpar(fontsize = 9, fontface = "bold"),
  column_names_gp       = gpar(fontsize = 7),
  column_names_rot      = 60,

  # Visuals
  border                = TRUE,
  rect_gp               = gpar(col = "white", lwd = 0.3),

  # Legend
  heatmap_legend_param  = list(
    title          = "Z-score",
    at             = c(-2, -1, 0, 1, 2),
    legend_height  = unit(3, "cm"),
    title_gp       = gpar(fontsize = 9),
    labels_gp      = gpar(fontsize = 8)
  ),

  width  = unit(22, "cm"),
  height = unit(14, "cm")
)

# save
pdf(paste0(params$out,"figures/module_activity_cluster_group.pdf"), 
    width = 16, height = 9)
draw(ht,
     heatmap_legend_side   = "right",
     annotation_legend_side = "right",
     padding = unit(c(5, 5, 5, 5), "mm"))
dev.off()

# export the scored matrix as CSV
write.csv(agg, paste0(params$out,"tables/module_scores_cluster_group_long.csv"), 
                      row.names = FALSE)
```

# ___NK cell cluster annotation based on functional NK cell gene modules___

Barplot with annotated ATAC clusters

```{r}
proj_filt$Cluster_names     <- CLUSTER_NAMES[proj_filt$Clusters_res1]
proj_filt$Metacluster_names <- METACLUSTER[proj_filt$Clusters_res1]

unique(proj_filt$Cluster_names)
unique(proj_filt$group)
```


## ___Barplot per cluster___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "Cluster_names")))

cluster_order <- c(
  "CD56bright 1", "CD56bright 2", "CD56bright 3",
  "Intermediate 1", "Intermediate 2",
  "CD56dim 1", "CD56dim 2")

meta$Cluster_names <- factor(meta$Cluster_names, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, Cluster_names, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = Cluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perCLUST_v2, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# per sample
df_sample <- meta %>%
  count(sample, Cluster_names, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = Cluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perCLUST_v2, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_cluster_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_cluster_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```

## ___Bar plot per metacluster___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "Metacluster_names")))

cluster_order <- c(
  "CD56bright",
  "Intermediate",
  "CD56dim")

meta$Metacluster_names <- factor(meta$Metacluster_names, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, Metacluster_names, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = Metacluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perMETACLUST, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# per sample
df_sample <- meta %>%
  count(sample, Metacluster_names, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = Metacluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perMETACLUST, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_metacluster_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_metacluster_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```


### ___with the label transfer from RNA___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "majority_vote_label")))

cluster_order <- c(
  "hNK_Bm1",
  "hNK_Bm2",
  "hNK_Bm3",
  "hNK_Bm4")

meta$majority_vote_label <- factor(meta$majority_vote_label, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, majority_vote_label, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = majority_vote_label)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_v2, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# per sample
df_sample <- meta %>%
  count(sample, majority_vote_label, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = majority_vote_label)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_v2, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_metacluster_labelTransfer_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_metacluster_labelTransfer_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```

```{r}
saveRDS(proj_filt, file = paste0(params$out,"/Rds/snATAC_NK_v20260401.Rds"))
```


# ___Publication-ready module score violin plots___

Publication-quality violin plots matching the figure style: pure violins colored per
genetic group, Wilcoxon significance vs control, expression-formatted x-axis labels,
colored y-axis title identifying the NK subset or module.

```{r}
library(ggsignif)
library(ggtext)

GROUP_ORDER_PUB <- c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt")

COL_GROUP_PUB <- c(
  "control"      = unname(COLOR_CODE_GROUP["control"]),
  "ATRXdel"      = unname(COLOR_CODE_GROUP["ATRXdel"]),
  "MYCNamp"      = unname(COLOR_CODE_GROUP["MYCNamp"]),
  "ATRXwtMYCNwt" = unname(COLOR_CODE_GROUP["ATRXwtMYCNwt"])
)

# Group label normalizer — handles ATRXm/ATRXmut/ATRXdel spelling variants
remap_group_32 <- remap_group

# x-axis label formatter — returns a parsed expression for ggplot2 label rendering
pub_x_labels <- function(breaks) {
  label_map <- list(
    "control"      = "CTRL",
    "ATRXdel"      = expression(ATRX^del),
    "MYCNamp"      = expression(MYCN^amp),
    "ATRXwtMYCNwt" = expression(ATRX^wt~MYCN^wt)
  )
  lapply(breaks, function(b) {
    if (b %in% names(label_map)) label_map[[b]] else b
  })
}

# Helper: publication violin for one module/score column, comparing all groups vs control
pub_violin_module <- function(data, score_col, y_label, title_label,
                               title_color = "black",
                               ref_group   = "control",
                               base_size   = 9) {
  df <- data %>%
    dplyr::select(group, score = dplyr::all_of(score_col)) %>%
    dplyr::filter(!is.na(group), !is.na(score)) %>%
    dplyr::mutate(group = factor(group, levels = GROUP_ORDER_PUB))

  non_ref <- setdiff(levels(df$group), ref_group)

  # Wilcoxon tests vs reference group
  stat_res <- lapply(non_ref, function(g) {
    x1 <- df$score[df$group == ref_group]
    x2 <- df$score[df$group == g]
    if (length(x1) < 3 || length(x2) < 3) return(NULL)
    p <- wilcox.test(x1, x2)$p.value
    symb <- dplyr::case_when(
      p < 1e-4 ~ "****", p < 0.001 ~ "***",
      p < 0.01 ~ "**",   p < 0.05  ~ "*",
      TRUE     ~ "ns"
    )
    list(pair = c(ref_group, g), label = symb)
  })
  stat_res <- Filter(Negate(is.null), stat_res)

  comparisons  <- lapply(stat_res, `[[`, "pair")
  annotations  <- sapply(stat_res, `[[`, "label")

  ggplot(df, aes(x = group, y = score, fill = group)) +
    geom_violin(scale = "width", trim = FALSE, alpha = 0.95, linewidth = 0) +
    {if (length(comparisons) > 0)
       ggsignif::geom_signif(
         comparisons      = comparisons,
         annotations      = annotations,
         map_signif_level = FALSE,
         step_increase    = 0.10,
         tip_length       = 0.01,
         textsize         = 3,
         color            = "black",
         vjust            = 0.3
       )} +
    scale_fill_manual(values = COL_GROUP_PUB) +
    scale_x_discrete(labels = pub_x_labels) +
    labs(
      title = title_label,
      x     = NULL,
      y     = y_label
    ) +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x     = element_text(angle = 45, hjust = 1, size = base_size - 1),
      axis.text.y     = element_text(size = base_size - 1),
      axis.title.y    = element_text(size = base_size, face = "plain"),
      axis.line       = element_line(linewidth = 0.4),
      axis.ticks      = element_line(linewidth = 0.4),
      plot.title      = element_text(size = base_size, face = "bold",
                                     color = title_color, hjust = 0),
      plot.margin     = margin(t = 25, r = 8, b = 5, l = 8)
    )
}
```

## ___Module scores within immunomodulatory NK subsets (Bm2 and Bm3)___

For cells labeled as hNK_Bm2 or hNK_Bm3 by label transfer, show a defined
"Mature NK" gene module score across genetic groups — recapitulating the figure style.

```{r}
# ArchR native addModuleScore for mature NK signature (GeneScoreMatrix)
mature_NK_genes_valid <- intersect(mature_NK_genes,
                                   getFeatures(proj_filt, useMatrix = "GeneScoreMatrix"))
cat("Mature NK genes in GeneScoreMatrix:", length(mature_NK_genes_valid),
    "/", length(mature_NK_genes), "\n")
proj_filt <- addModuleScore(
  ArchRProj = proj_filt,
  useMatrix = "GeneScoreMatrix",
  name      = "MatureNK",
  features  = list(score = mature_NK_genes_valid)
)

# Retrieve score — column is "MatureNK.score"
mature_cd <- as.data.frame(
  getCellColData(proj_filt, select = c("MatureNK.score", "group", "majority_vote_label"))
)

df_mature_atac <- data.frame(
  cell     = rownames(mature_cd),
  score    = mature_cd[["MatureNK.score"]],
  group    = remap_group(as.character(mature_cd[["group"]])),
  nk_label = as.character(mature_cd[["majority_vote_label"]]),
  stringsAsFactors = FALSE
)
cat("NK label transfer counts:\n")
print(table(df_mature_atac$nk_label, useNA = "always"))

NK_SUBSET_COLORS <- COLOR_CODE_v2[c("hNK_Bm2", "hNK_Bm3")]
```

```{r fig.height=7, fig.width=3.5}
nk_score_plots <- lapply(c("hNK_Bm2", "hNK_Bm3"), function(subset) {
  df <- df_mature_atac %>%
    dplyr::filter(nk_label == subset, !is.na(group)) %>%
    dplyr::select(group, score)

  pub_violin_module(
    data        = df,
    score_col   = "score",
    y_label     = "Mature NK gene score (ATAC)",
    title_label = paste0("Immunomodulatory ", subset),
    title_color = unname(NK_SUBSET_COLORS[subset])
  )
})

combined_nk <- patchwork::wrap_plots(nk_score_plots, ncol = 1)
combined_nk

ggsave(paste0(params$out, "figures/MatureNK_ATACscore_Bm2-Bm3_pub.pdf"),
       combined_nk, width = 3.5, height = 7, device = cairo_pdf)
```

## ___All NK cells: module scores per genetic group (all 25 modules)___

```{r fig.height=18, fig.width=14}
# score_df columns: M1..M25, cell, cluster, group (from module scoring section above)

score_df_pub <- score_df %>%
  dplyr::filter(!is.na(group)) %>%
  dplyr::mutate(group = factor(remap_group(group), levels = GROUP_ORDER_PUB))

plots_by_group <- lapply(paste0("M", 1:25), function(mod) {
  pub_violin_module(
    data        = score_df_pub,
    score_col   = mod,
    y_label     = "Module score",
    title_label = mod,
    title_color = "black"
  ) + theme(plot.margin = margin(t = 25, r = 4, b = 4, l = 4))
})

patchwork::wrap_plots(plots_by_group, ncol = 5)

ggsave(paste0(params$out, "figures/NK_module_scores_M1-M25_per_group_pub.pdf"),
       patchwork::wrap_plots(plots_by_group, ncol = 5),
       width = 14, height = 18, device = cairo_pdf)
```

## ___All NK cells: module scores per ATAC cluster (all 25 modules)___

```{r}
pub_violin_cluster <- function(data, score_col, y_label, title_label, base_size = 9) {
  df <- data %>%
    dplyr::select(cluster, score = dplyr::all_of(score_col)) %>%
    dplyr::filter(!is.na(cluster), !is.na(score)) %>%
    dplyr::mutate(cluster = factor(cluster, levels = CLUSTER_ORDER))

  ggplot(df, aes(x = cluster, y = score, fill = cluster)) +
    geom_violin(scale = "width", trim = FALSE, alpha = 0.95, linewidth = 0) +
    scale_fill_manual(values = COLOR_CODE_perCLUST[CLUSTER_ORDER]) +
    labs(title = title_label, x = NULL, y = y_label) +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x     = element_text(angle = 45, hjust = 1, size = base_size - 1),
      axis.text.y     = element_text(size = base_size - 1),
      axis.title.y    = element_text(size = base_size),
      axis.line       = element_line(linewidth = 0.4),
      axis.ticks      = element_line(linewidth = 0.4),
      plot.title      = element_text(size = base_size, face = "bold", hjust = 0),
      plot.margin     = margin(t = 10, r = 4, b = 4, l = 4)
    )
}
```

```{r fig.height=18, fig.width=14}
score_df_cl <- score_df %>%
  dplyr::filter(!is.na(cluster)) %>%
  dplyr::mutate(cluster = factor(cluster, levels = CLUSTER_ORDER))

plots_by_cluster <- lapply(paste0("M", 1:25), function(mod) {
  pub_violin_cluster(
    data        = score_df_cl,
    score_col   = mod,
    y_label     = "Module score",
    title_label = mod
  )
})

patchwork::wrap_plots(plots_by_cluster, ncol = 5)

ggsave(paste0(params$out, "figures/NK_module_scores_M1-M25_per_cluster_pub.pdf"),
       patchwork::wrap_plots(plots_by_cluster, ncol = 5),
       width = 14, height = 18, device = cairo_pdf)
```

## ___Module–cluster activity heatmap (cluster annotation sanity check)___

Mean module score per cluster collapsed across genetic groups, Z-scored per module
across clusters. Rows cluster automatically — modules active in the same NK states
should group together.

```{r fig.height=8, fig.width=5}
# Collapse score_df to mean per cluster × module (ignore group here)
agg_cl <- score_df %>%
  pivot_longer(cols      = all_of(names(module_genes)),
               names_to  = "Module",
               values_to = "Score") %>%
  dplyr::filter(!is.na(cluster), !is.na(Score)) %>%
  dplyr::group_by(cluster, Module) %>%
  dplyr::summarise(MeanScore = mean(Score, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(Module  = factor(Module,  levels = paste0("M", 1:25)),
                cluster = factor(cluster, levels = CLUSTER_ORDER))

mat_cl <- agg_cl %>%
  pivot_wider(names_from = cluster, values_from = MeanScore) %>%
  column_to_rownames("Module") %>%
  as.matrix()
cl_cols_avail <- intersect(CLUSTER_ORDER, colnames(mat_cl))
mat_cl <- mat_cl[intersect(paste0("M", 1:25), rownames(mat_cl)),
                 cl_cols_avail,
                 drop = FALSE]
storage.mode(mat_cl) <- "numeric"

# Z-score each module across clusters (highlights relative enrichment per cluster)
mat_cl_z <- t(scale(t(mat_cl)))
mat_cl_z[is.nan(mat_cl_z)] <- 0

ha_cl <- HeatmapAnnotation(
  Cluster = cl_cols_avail,
  col     = list(Cluster = COLOR_CODE_perCLUST[cl_cols_avail]),
  simple_anno_size     = unit(3, "mm"),
  annotation_name_side = "left",
  annotation_name_gp   = gpar(fontsize = 8)
)

ht_cl <- Heatmap(
  mat_cl_z,
  name              = "Z-score",
  col               = colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027")),
  top_annotation    = ha_cl,
  cluster_rows      = TRUE,
  cluster_columns   = FALSE,
  row_names_side    = "left",
  row_names_gp      = gpar(fontsize = 9),
  column_names_side = "top",
  column_names_gp   = gpar(fontsize = 9),
  border            = TRUE,
  rect_gp           = gpar(col = "white", lwd = 0.5),
  heatmap_legend_param = list(title = "Z-score", at = c(-2, 0, 2),
                               title_gp = gpar(fontsize = 9),
                               labels_gp = gpar(fontsize = 8)),
  width  = unit(5,  "cm"),
  height = unit(12, "cm")
)

pdf(paste0(params$out, "figures/NK_module_cluster_heatmap_sanity.pdf"),
    width = 5, height = 8)
draw(ht_cl, padding = unit(c(5, 5, 5, 5), "mm"))
dev.off()

draw(ht_cl)
```

## ___Pathway enrichment analysis for M1-M25 (GO:BP + KEGG)___

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# Run GO:BP enrichment for each module
enrich_list <- lapply(names(module_genes), function(mod) {
  genes   <- unique(module_genes[[mod]])
  id_map  <- tryCatch(
    bitr(genes, fromType = "SYMBOL", toType = "ENTREZID",
         OrgDb = org.Hs.eg.db, drop = TRUE),
    error = function(e) NULL
  )
  if (is.null(id_map) || nrow(id_map) < 5) return(NULL)

  res <- tryCatch(
    enrichGO(gene          = id_map$ENTREZID,
             OrgDb         = org.Hs.eg.db,
             ont           = "BP",
             pAdjustMethod = "BH",
             pvalueCutoff  = 0.05,
             qvalueCutoff  = 0.2,
             readable      = TRUE),
    error = function(e) NULL
  )
  if (is.null(res) || nrow(as.data.frame(res)) == 0) return(NULL)
  as.data.frame(res) %>% dplyr::mutate(Module = mod)
})
names(enrich_list) <- names(module_genes)

enrich_df <- dplyr::bind_rows(Filter(Negate(is.null), enrich_list))
cat("Modules with GO:BP hits:", sum(sapply(enrich_list, Negate(is.null))),
    "/", length(module_genes), "\n")

write.csv(enrich_df,
          paste0(params$out, "tables/NK_module_GO_enrichment.csv"),
          row.names = FALSE)
```

```{r fig.height=8, fig.width=6}
# Top 3 terms per module → dot plot
top_terms <- enrich_df %>%
  dplyr::mutate(Module = factor(Module, levels = paste0("M", 1:25))) %>%
  dplyr::group_by(Module) %>%
  dplyr::slice_min(p.adjust, n = 3, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    Description = stringr::str_wrap(Description, width = 45),
    log10p      = -log10(p.adjust)
  )

ggplot(top_terms, aes(x = Module, y = Description, size = Count,
                      colour = log10p)) +
  geom_point() +
  scale_colour_gradient(low = "#fee090", high = "#d73027",
                        name = expression(-log[10](p[adj]))) +
  scale_size_continuous(range = c(2, 6), name = "Gene count") +
  labs(x = NULL, y = NULL,
       title = "Top GO:BP terms per module") +
  theme_classic(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    axis.text.y  = element_text(size = 7),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3),
    plot.title   = element_text(face = "bold", size = 10)
  )

ggsave(paste0(params$out, "figures/NK_module_GO_dotplot.pdf"),
       width = 6, height = 8, device = cairo_pdf)
```

## ___Combined GRN module scores___

Modules grouped into seven biologically defined GRN programmes (based on module gene content
and NK cell biology):

| GRN | NK state | Modules |
|-----|----------|---------|
| Immunomodulatory GRN | CD56bright / Immature | M2, M4 |
| Cytotoxic GRN | CD56dim Cytotoxic | M8, M10, M15, M19, M20 |
| Adaptive GRN | Adaptive / Memory-like | M12, M13, M16 |
| IFN Response GRN | IFN Response | M7, M11 |
| Antigen licensing GRN | MHC-I / Antigen Processing | M9, M14, M18 |
| Tissue Resident GRN | Tissue Resident | M17 |
| Transitional GRN | Developmental / Transitional | M1, M3, M5, M6 |

```{r}
GRN_MODULES <- list(
  "Immunomodulatory GRN"  = paste0("M", c(2, 4)),
  "Cytotoxic GRN"         = paste0("M", c(8, 10, 15, 19, 20)),
  "Adaptive GRN"          = paste0("M", c(12, 13, 16)),
  "IFN Response GRN"      = paste0("M", c(7, 11)),
  "Antigen licensing GRN" = paste0("M", c(9, 14, 18)),
  "Tissue Resident GRN"   = paste0("M", c(17)),
  "Transitional GRN"      = paste0("M", c(1, 3, 5, 6))
)

GRN_COLORS <- c(
  "Immunomodulatory GRN"  = unname(COL_FUNCTIONAL["immunomodulatory"]),  # #4A9B6F — matches COL_FUNCTIONAL
  "Cytotoxic GRN"         = unname(COL_FUNCTIONAL["cytotoxic"]),          # #C0392B — matches COL_FUNCTIONAL
  "Adaptive GRN"          = "#7B241C",                                     # dark maroon — deeper shade of cytotoxic red, adaptive = terminally mature
  "IFN Response GRN"      = unname(COLOR_CODE_v2["hNK_Bm4"]),             # #CDA646 — gold
  "Antigen licensing GRN" = unname(COLOR_CODE_v2["hNK_Bm1"]),             # #6376AE — slate blue
  "Tissue Resident GRN"   = unname(COLOR_CODE_GROUP["MYCNamp"]),           # #A7563C — warm terracotta
  "Transitional GRN"      = unname(COLOR_CODE_GROUP["control"])            # #7EA3AC — muted steel blue, neutral
)

GRN_ORDER <- c(
  "Immunomodulatory GRN", "Cytotoxic GRN", "Adaptive GRN",
  "IFN Response GRN", "Antigen licensing GRN",
  "Tissue Resident GRN", "Transitional GRN"
)

# Compute per-cell GRN score = mean of constituent module scores
mod_cols <- names(module_genes)
for (grn in names(GRN_MODULES)) {
  mods_use <- intersect(GRN_MODULES[[grn]], mod_cols)
  score_df[[grn]] <- rowMeans(score_df[, mods_use, drop = FALSE], na.rm = TRUE)
}

grn_df <- score_df %>%
  dplyr::select(cell, cluster, group, all_of(GRN_ORDER)) %>%
  dplyr::mutate(
    group   = factor(remap_group(group),                   levels = GROUP_ORDER_PUB),
    cluster = factor(as.character(cluster), levels = CLUSTER_ORDER)
  )

cat("GRN score summary:\n")
print(summary(grn_df[, GRN_ORDER]))
```

```{r fig.height=4, fig.width=10}
# Violin: GRN score per genetic group
grn_group_plots <- lapply(GRN_ORDER, function(grn) {
  df <- grn_df %>%
    dplyr::filter(!is.na(group)) %>%
    dplyr::select(group, score = dplyr::all_of(grn))

  pub_violin_module(
    data        = df,
    score_col   = "score",
    y_label     = "GRN score",
    title_label = grn,
    title_color = GRN_COLORS[grn]
  )
})

p_grn_group <- patchwork::wrap_plots(grn_group_plots, ncol = 4)
p_grn_group

ggsave(paste0(params$out, "figures/NK_GRN_scores_per_group_pub.pdf"),
       p_grn_group, width = 14, height = 8, device = cairo_pdf)
```

```{r fig.height=4, fig.width=10}
# Violin: GRN score per ATAC cluster

pub_violin_grn_cluster <- function(data, score_col, grn_name, base_size = 9) {
  df <- data %>%
    dplyr::filter(!is.na(cluster)) %>%
    dplyr::select(cluster, score = dplyr::all_of(score_col))

  ggplot(df, aes(x = cluster, y = score, fill = cluster)) +
    geom_violin(scale = "width", trim = FALSE, alpha = 0.95, linewidth = 0) +
    scale_fill_manual(values = COLOR_CODE_perCLUST[CLUSTER_ORDER]) +
    labs(title = grn_name, x = NULL, y = "GRN score") +
    pub_theme +
    theme(
      legend.position = "none",
      axis.text.x  = element_text(angle = 45, hjust = 1, size = base_size - 1),
      axis.text.y  = element_text(size = base_size - 1),
      axis.line    = element_line(linewidth = 0.4),
      axis.ticks   = element_line(linewidth = 0.4),
      plot.title   = element_text(size = base_size, face = "bold",
                                  color = GRN_COLORS[grn_name], hjust = 0),
      plot.margin  = margin(t = 10, r = 4, b = 4, l = 4)
    )
}

grn_cluster_plots <- lapply(GRN_ORDER, function(grn) {
  pub_violin_grn_cluster(grn_df, grn, grn)
})

p_grn_cluster <- patchwork::wrap_plots(grn_cluster_plots, ncol = 4)
p_grn_cluster

ggsave(paste0(params$out, "figures/NK_GRN_scores_per_cluster_pub.pdf"),
       p_grn_cluster, width = 14, height = 8, device = cairo_pdf)
```

```{r fig.height=5, fig.width=6}
# Heatmap: mean GRN score per cluster × GRN (compact overview)
grn_cl_mat <- sapply(GRN_ORDER, function(grn) {
  grn_df %>%
    dplyr::filter(!is.na(cluster)) %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(m = mean(.data[[grn]], na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(factor(cluster, levels = CLUSTER_ORDER)) %>%
    dplyr::pull(m)
})
cl_avail <- intersect(CLUSTER_ORDER, unique(na.omit(as.character(grn_df$cluster))))
rownames(grn_cl_mat) <- cl_avail

grn_cl_z <- scale(grn_cl_mat)   # Z-score across GRNs per cluster

# t(grn_cl_z): rows = GRNs, cols = clusters — row annotation marks GRN category by colour
grn_row_order <- intersect(GRN_ORDER, rownames(t(grn_cl_z)))

ra_grn <- rowAnnotation(
  GRN  = grn_row_order,
  col  = list(GRN = GRN_COLORS),
  simple_anno_size     = unit(3, "mm"),
  annotation_name_gp   = gpar(fontsize = 8),
  show_annotation_name = FALSE
)

ht_grn <- Heatmap(
  t(grn_cl_z)[grn_row_order, , drop = FALSE],
  name              = "Z-score",
  col               = colorRamp2(c(-1.5, 0, 1.5), c("#4575b4", "white", "#d73027")),
  left_annotation   = ra_grn,
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  row_names_side    = "right",
  row_names_gp      = gpar(fontsize = 8),
  column_names_side = "top",
  column_names_gp   = gpar(fontsize = 9),
  border            = TRUE,
  rect_gp           = gpar(col = "white", lwd = 1),
  heatmap_legend_param = list(title = "Z-score",
                               title_gp = gpar(fontsize = 9),
                               labels_gp = gpar(fontsize = 8)),
  width  = unit(5, "cm"),
  height = unit(7, "cm")
)

pdf(paste0(params$out, "figures/NK_GRN_cluster_heatmap.pdf"), width = 6, height = 7)
draw(ht_grn, padding = unit(c(5, 5, 5, 5), "mm"))
dev.off()

draw(ht_grn)
```

## ___Dotplot: GRN scores per NK subset × genetic group___

```{r fig.height=5.5, fig.width=10}
# Join NK label-transfer subset (majority_vote_label) into GRN data via cell names
grn_dot_df <- grn_df %>%
  dplyr::mutate(
    nk_subset = factor(
      as.character(lt_meta[cell, "nk_subset"]),
      levels = NK_SUBSET_ORDER
    ),
    group = factor(as.character(group), levels = GROUP_ORDER_PUB)
  ) %>%
  dplyr::filter(!is.na(nk_subset), !is.na(group))

# Aggregate mean score + % positive per NK subset × group × GRN,
# then z-score each GRN across all NK subset × group combinations
dot_agg <- grn_dot_df %>%
  tidyr::pivot_longer(cols = dplyr::all_of(GRN_ORDER),
                      names_to = "GRN", values_to = "score") %>%
  dplyr::filter(!is.na(score)) %>%
  dplyr::group_by(nk_subset, group, GRN) %>%
  dplyr::summarise(
    mean_score = mean(score, na.rm = TRUE),
    pct_pos    = mean(score > 0, na.rm = TRUE) * 100,
    n_cells    = dplyr::n(),
    .groups    = "drop"
  ) %>%
  dplyr::filter(n_cells >= 5) %>%
  dplyr::group_by(GRN) %>%
  dplyr::mutate(z_score = as.numeric(scale(mean_score))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    z_score = pmax(pmin(z_score, 2), -2),   # cap at ±2
    GRN     = factor(GRN, levels = rev(GRN_ORDER))
  )

p_grn_dot <- ggplot(dot_agg,
                    aes(x = nk_subset, y = GRN,
                        size = pct_pos, colour = z_score)) +
  geom_point() +
  facet_grid(. ~ group, scales = "free_x", space = "free_x") +
  scale_size_continuous(range = c(0.5, 5),
                        name  = "% cells\n(score > 0)",
                        limits = c(0, 100)) +
  scale_colour_gradient2(
    low      = "#4575b4",
    mid      = "white",
    high     = "#d73027",
    midpoint = 0,
    limits   = c(-2, 2),
    name     = "GRN score\n(z-score)"
  ) +
  labs(
    title = "GRN scores per NK subset and genetic group",
    x = NULL, y = NULL
  ) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 7,
                                  face = "italic"),
    axis.text.y   = element_text(size = 8),
    strip.text    = element_text(size = 8, face = "bold"),
    panel.spacing = unit(1, "mm"),
    panel.border  = element_rect(colour = "grey80", fill = NA, linewidth = 0.4)
  )

p_grn_dot

ggsave(paste0(params$out, "figures/NK_GRN_scores_NKsubset-group_dotplot.pdf"),
       p_grn_dot, width = 10, height = 5.5, device = cairo_pdf)
```

## ___Dotplot: all module scores (M1–M25) per NK subset × genetic group___

```{r fig.height=12, fig.width=10}
# Reuse grn_dot_df (already has nk_subset + group); add all module scores
mod_dot_df <- score_df %>%
  dplyr::mutate(
    nk_subset = factor(
      as.character(lt_meta[cell, "nk_subset"]),
      levels = NK_SUBSET_ORDER
    ),
    group = factor(remap_group(as.character(group)), levels = GROUP_ORDER_PUB)
  ) %>%
  dplyr::filter(!is.na(nk_subset), !is.na(group))

mod_dot_agg <- mod_dot_df %>%
  tidyr::pivot_longer(cols = dplyr::all_of(names(module_genes)),
                      names_to = "Module", values_to = "score") %>%
  dplyr::filter(!is.na(score)) %>%
  dplyr::group_by(nk_subset, group, Module) %>%
  dplyr::summarise(
    mean_score = mean(score, na.rm = TRUE),
    pct_pos    = mean(score > 0, na.rm = TRUE) * 100,
    n_cells    = dplyr::n(),
    .groups    = "drop"
  ) %>%
  dplyr::filter(n_cells >= 5) %>%
  dplyr::group_by(Module) %>%
  dplyr::mutate(z_score = as.numeric(scale(mean_score))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    z_score = pmax(pmin(z_score, 2), -2),
    Module  = factor(Module, levels = rev(paste0("M", 1:25)))
  )

p_mod_dot <- ggplot(mod_dot_agg,
                    aes(x = nk_subset, y = Module,
                        size = pct_pos, colour = z_score)) +
  geom_point() +
  facet_grid(. ~ group, scales = "free_x", space = "free_x") +
  scale_size_continuous(range = c(0.3, 4),
                        name  = "% cells\n(score > 0)",
                        limits = c(0, 100)) +
  scale_colour_gradient2(
    low      = "#4575b4",
    mid      = "white",
    high     = "#d73027",
    midpoint = 0,
    limits   = c(-2, 2),
    name     = "Module score\n(z-score)"
  ) +
  labs(
    title = "Module scores per NK subset and genetic group",
    x = NULL, y = NULL
  ) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 7, face = "italic"),
    axis.text.y   = element_text(size = 7),
    strip.text    = element_text(size = 8, face = "bold"),
    panel.spacing = unit(1, "mm"),
    panel.border  = element_rect(colour = "grey80", fill = NA, linewidth = 0.4)
  )

p_mod_dot

ggsave(paste0(params$out, "figures/NK_module_scores_NKsubset-group_dotplot.pdf"),
       p_mod_dot, width = 10, height = 12, device = cairo_pdf)
```

# ___Adaptive NK cell signature: ModuleScore___

The question we want to answer here is if we have any adaptive NK cells in the
ATAC data. When doing the label transfer from RNA data of matched patients,
we saw a clear signature for adaptive NK cells in a small proportion of NK cells,
however in the ATAC data, the label transfer did not identify any adaptive NK cells.

```{r}
# ArchR native addModuleScore for adaptive NK signature (GeneScoreMatrix)
adaptive_NK_genes_valid <- intersect(adaptive_NK_genes,
                                     getFeatures(proj_filt, useMatrix = "GeneScoreMatrix"))
cat("Adaptive NK genes in GeneScoreMatrix:", length(adaptive_NK_genes_valid),
    "/", length(adaptive_NK_genes), "\n")
proj_filt <- addModuleScore(
  ArchRProj = proj_filt,
  useMatrix = "GeneScoreMatrix",
  name      = "AdaptiveNK",
  features  = list(score = adaptive_NK_genes_valid)
)

# Retrieve score — column is "AdaptiveNK.score"
adapt_cd <- as.data.frame(
  getCellColData(proj_filt,
                 select = c("AdaptiveNK.score", "group", "Clusters_res1",
                            "majority_vote_label"))
)

df_adapt_atac <- data.frame(
  cell     = rownames(adapt_cd),
  score    = adapt_cd[["AdaptiveNK.score"]],
  group    = factor(remap_group(as.character(adapt_cd[["group"]])),
                    levels = GROUP_ORDER_PUB),
  cluster  = factor(as.character(adapt_cd[["Clusters_res1"]]), levels = CLUSTER_ORDER),
  nk_label = as.character(adapt_cd[["majority_vote_label"]]),
  stringsAsFactors = FALSE
)
```

```{r fig.height=3.5, fig.width=3.5}
# Adaptive NK score per genetic group
p_adapt_group <- pub_violin_module(
  data        = df_adapt_atac %>% dplyr::filter(!is.na(group)),
  score_col   = "score",
  y_label     = "Adaptive NK gene score (ATAC)",
  title_label = "Adaptive NK signature — per group",
  title_color = "#2C7BB6"
)
p_adapt_group

ggsave(paste0(params$out, "figures/AdaptiveNK_ATACscore_per_group_pub.pdf"),
       p_adapt_group, width = 3.5, height = 3.5, device = cairo_pdf)
```

```{r fig.height=3.5, fig.width=4}
# Adaptive NK score per ATAC cluster C1-C7
p_adapt_cluster <- pub_violin_cluster(
  data        = df_adapt_atac %>% dplyr::filter(!is.na(cluster)),
  score_col   = "score",
  y_label     = "Adaptive NK gene score (ATAC)",
  title_label = "Adaptive NK signature — per ATAC cluster"
)
p_adapt_cluster

ggsave(paste0(params$out, "figures/AdaptiveNK_ATACscore_per_ATACcluster_pub.pdf"),
       p_adapt_cluster, width = 4, height = 3.5, device = cairo_pdf)
```

```{r fig.height=3.5, fig.width=5}
# Adaptive NK score per NK label-transfer subset, split by genetic group
p_adapt_nk <- df_adapt_atac %>%
  dplyr::filter(!is.na(nk_label), !is.na(group)) %>%
  dplyr::mutate(nk_label = factor(nk_label,
                                  levels = c("hNK_Bm1","hNK_Bm2","hNK_Bm3","hNK_Bm4"))) %>%
  ggplot(aes(x = nk_label, y = score, fill = group)) +
  geom_violin(scale = "width", position = position_dodge(0.85),
              trim = FALSE, alpha = 0.85, linewidth = 0) +
  scale_fill_manual(values = COL_GROUP_PUB) +
  labs(title = "Adaptive NK signature — per NK subset (ATAC)",
       x = NULL, y = "Adaptive NK gene score (ATAC)", fill = "genetic group") +
  theme_classic(base_size = 9) +
  theme(axis.text.x    = element_text(angle = 35, hjust = 1),
        axis.line      = element_line(linewidth = 0.4),
        axis.ticks     = element_line(linewidth = 0.4),
        plot.title     = element_text(face = "bold", size = 9),
        legend.key.size = unit(3, "mm"),
        legend.text    = element_text(size = 8))
p_adapt_nk

ggsave(paste0(params$out, "figures/AdaptiveNK_ATACscore_per_NKsubset_pub.pdf"),
       p_adapt_nk, width = 5, height = 3.5, device = cairo_pdf)
```

```{r}
# Extract GeneScoreMatrix for per-gene accessibility plots (KLRC2, FCER1G)
se_markers  <- getMatrixFromProject(proj_filt, useMatrix = "GeneScoreMatrix")
gene_names  <- rowData(se_markers)$name
if (is.null(gene_names)) gene_names <- rownames(se_markers)
rownames(se_markers) <- gene_names
full_expr   <- as.matrix(assay(se_markers))
```

```{r fig.height=3.5, fig.width=7}
# Gene accessibility for key adaptive NK markers: KLRC2 (gained) and FCER1G (lost)
marker_genes <- intersect(c("KLRC2", "FCER1G"), gene_names)

dplyr::bind_rows(lapply(marker_genes, function(g) {
  data.frame(
    gene     = g,
    score    = as.numeric(full_expr[g, ]),
    nk_label = as.character(proj_filt$majority_vote_label),
    group    = factor(remap_group(as.character(proj_filt$group)), levels = GROUP_ORDER_PUB),
    stringsAsFactors = FALSE
  )
})) %>%
  dplyr::filter(!is.na(nk_label), !is.na(group)) %>%
  dplyr::mutate(nk_label = factor(nk_label,
                                  levels = c("hNK_Bm1","hNK_Bm2","hNK_Bm3","hNK_Bm4"))) %>%
  ggplot(aes(x = nk_label, y = score, fill = group)) +
  geom_violin(scale = "width", position = position_dodge(0.85),
              trim = FALSE, alpha = 0.85, linewidth = 0) +
  facet_wrap(~ gene, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = COL_GROUP_PUB) +
  labs(x = NULL, y = "Gene accessibility score (ATAC)", fill = "genetic group") +
  theme_classic(base_size = 9) +
  theme(axis.text.x    = element_text(angle = 35, hjust = 1),
        strip.text     = element_text(face = "bold", size = 9),
        legend.key.size = unit(3, "mm"),
        legend.text    = element_text(size = 8)) ->
  p_markers
p_markers

ggsave(paste0(params$out, "figures/AdaptiveNK_KLRC2_FCER1G_ATACscore_pub.pdf"),
       p_markers, width = 7, height = 3.5, device = cairo_pdf)
```



# ___Save intermediate ArchR project and score data___

```{r}
saveRDS(proj_filt,
        file = paste0(params$out, "Rds/snATAC_NK_v20260625_modules.Rds"))
saveRDS(score_df,
        file = paste0(params$out, "Rds/snATAC_NK_score_df_v20260625.Rds"))
cat("Saved proj_filt and score_df.\n")
```

# ___The End___

```{r render, eval=FALSE, include=FALSE}
rmarkdown::render(
  "/home/rstudio/code/NK-NB_study/code/39a_20260625_snATAC_NK_module_scoring.Rmd",
  output_file = paste0(params$out, "html/39a_20260625_snATAC_NK_module_scoring.html")
)
```
