---
title: "NK project: snATAC-seq — TF footprinting and GSK3B chromatin accessibility"
author: "Dr. Sara Wernig-Zorc"
date: "`r format(Sys.time(), '%B %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
    collapsed: false
    code_folding: hide
    highlight: pygments
    theme: spacelab
    df_print: paged
params:
  out:  "/home/rstudio/mnt_out/NK_project/"
  anno: "/home/rstudio/mnt_resources/"
---

PROJECT DESCRIPTION

NK cell snATAC-seq re-analysed with unsupervised LSI + clustering at resolution 1,
yielding 7 distinct chromatin-accessibility clusters (C1-C7).
Clusters are annotated as CD56bright (C1-C3), Intermediate (C4-C5), and CD56dim (C6-C7).

Questions:
1. Do TFs relevant to NK cell development and function show differential footprinting
   between genetic patient groups (ATRXdel, control, MYCNamp, ATRXwtMYCNwt) within
   each of the 7 ATAC clusters?
2. Is the GSK3B gene locus and promoter differentially accessible between genetic groups
   within each cluster?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = FALSE,
                      error = FALSE,
                      messages = FALSE,
                      warning = FALSE)

library(parallel)
future::plan(strategy = "future::multicore", workers = 20)
options(future.globals.maxSize = 40 * 10^9)
print(params)
```

```{r include=FALSE}
source("/home/rstudio/00_NK_paper_setup.R")
source("/home/rstudio/00_NK_paper_utils.R")
```

# ___Load packages___

```{r include=FALSE}
suppressPackageStartupMessages({
  library(ArchR)
  library(BSgenome.Hsapiens.UCSC.hg38)
  library(JASPAR2024)
  library(TFBSTools)
  library(motifmatchr)
  library(RSQLite)
  library(ComplexHeatmap)
  library(circlize)
  library(ggrepel)
  library(patchwork)
  library(tidyverse)
})

addArchRThreads(threads = 20)
addArchRGenome("hg38")

filter    <- dplyr::filter
select    <- dplyr::select
mutate    <- dplyr::mutate
summarise <- dplyr::summarise
rename    <- dplyr::rename
count     <- dplyr::count
group_by  <- dplyr::group_by
ungroup   <- dplyr::ungroup
arrange   <- dplyr::arrange
slice_min <- dplyr::slice_min
sub       <- base::sub
gsub      <- base::gsub

# Remap group label strings to canonical names used throughout the script.
# Logic: anything with "atrx" but no "wt" → ATRXdel (catches ATRXdel/ATRXmut/ATRXm/ATRX_del etc.)
#        anything with "atrx" and "wt"    → ATRXwtMYCNwt
remap_group <- function(x) {
  x <- as.character(x)
  x[grepl("atrx", x, ignore.case = TRUE) & !grepl("wt", x, ignore.case = TRUE)] <- "ATRXdel"
  x[grepl("atrx", x, ignore.case = TRUE) &  grepl("wt", x, ignore.case = TRUE)] <- "ATRXwtMYCNwt"
  x[grepl("mycn", x, ignore.case = TRUE) &  grepl("amp", x, ignore.case = TRUE)] <- "MYCNamp"
  x[grepl("^control$",            x, ignore.case = TRUE)] <- "control"
  x
}

# Extract canonical group name from an ArchR clust_group column name.
# ArchR appends "._.Rep1" to coverage column names, e.g. "C2_control._.Rep1".
clust_col_to_group <- function(col_names) {
  # strip cluster prefix (everything up to and including the first "_")
  suffix <- sub("^[^_]+_", "", col_names)
  # strip ArchR's rep suffix (e.g. "._.Rep1")
  suffix <- sub("\\._\\.Rep.*$", "", suffix)
  remap_group(suffix)
}
```

```{r}
theme_set(pub_theme)

COL_ATAC_CLUSTER <- c(
  "C1" = "#1B4F8A",
  "C2" = "#2E7BB5",
  "C3" = "#6AADD5",
  "C4" = "#4A9B6F",
  "C5" = "#8FC494",
  "C6" = "#C0392B",
  "C7" = "#E8837A"
)

COL_GROUP <- c(
  "control"      = unname(COLOR_CODE_GROUP["control"]),
  "ATRXdel"      = unname(COLOR_CODE_GROUP["ATRXdel"]),
  "MYCNamp"      = unname(COLOR_CODE_GROUP["MYCNamp"]),
  "ATRXwtMYCNwt" = unname(COLOR_CODE_GROUP["ATRXwtMYCNwt"])
)

GROUP_ORDER   <- c( "control","ATRXdel", "MYCNamp", "ATRXwtMYCNwt")
CLUSTER_ORDER <- paste0("C", 1:7)

CLUSTER_NAMES <- c(
  "C1" = "CD56bright 1",
  "C2" = "CD56bright 2",
  "C3" = "CD56bright 3",
  "C4" = "Intermediate 1",
  "C5" = "Intermediate 2",
  "C6" = "CD56dim 1",
  "C7" = "CD56dim 2"
)

METACLUSTER <- c(
  "C1" = "CD56bright", "C2" = "CD56bright", "C3" = "CD56bright",
  "C4" = "Intermediate", "C5" = "Intermediate",
  "C6" = "CD56dim", "C7" = "CD56dim"
)
```

# ___Load ArchR project___

```{r}
proj_filt <- readRDS(file = paste0(params$out, "ArchR_v20260203/ATAC_analysis_v20260203.Rds"))

# Update output directory and Arrow file paths to current mount point
proj_filt@projectMetadata$outputDirectory <- paste0(params$out, "ArchR_v20260203")
proj_filt@sampleColData$ArrowFiles <- paste0(
  params$out, "ArchR_v20260203/ArrowFiles/",
  basename(as.character(proj_filt@sampleColData$ArrowFiles))
)

cat("Arrow file path set to:\n",
    as.character(proj_filt@sampleColData$ArrowFiles), "\n")
cat("Arrow file exists:", file.exists(as.character(proj_filt@sampleColData$ArrowFiles)), "\n")

# Inspect raw values before any conversion — catches spelling mismatches
raw_groups   <- sort(unique(as.character(proj_filt$group)))
raw_clusters <- sort(unique(as.character(proj_filt$Clusters_res1)))
cat("Raw group values in project:  ", paste(raw_groups,   collapse = ", "), "\n")
cat("Raw cluster values in project:", paste(raw_clusters, collapse = ", "), "\n")

proj_filt$group <- factor(remap_group(proj_filt$group), levels = GROUP_ORDER)

na_count <- sum(is.na(proj_filt$group))
if (na_count > 0) {
  warning(na_count, " cells have NA group after remapping — check raw_groups above")
} else {
  cat("Group remapping OK. Canonical values:", paste(levels(proj_filt$group), collapse = ", "), "\n")
  print(table(proj_filt$group, useNA = "always"))
}

# Add metadata
proj_filt$Cluster_name  <- CLUSTER_NAMES[as.character(proj_filt$Clusters_res1)]
proj_filt$Metacluster   <- METACLUSTER[as.character(proj_filt$Clusters_res1)]
proj_filt$Clusters_res1 <- factor(proj_filt$Clusters_res1, levels = CLUSTER_ORDER)

cat("\nCells:", nrow(getCellColData(proj_filt)), "\n")
cat("Clusters (Clusters_res1):\n")
print(table(proj_filt$Clusters_res1))
cat("\nGenetic groups (after remap):\n")
print(table(proj_filt$group, useNA = "always"))
cat("\nCluster x group:\n")
print(table(proj_filt$Clusters_res1, proj_filt$group))
```

# ___PART I___

# ___UMAP: cluster overview___

```{r fig.height=4.5, fig.width=9}
p_umap_cluster <- plotEmbedding(
  proj_filt,
  colorBy    = "cellColData",
  name       = "Clusters_res1",
  embedding  = "UMAP",
  pal        = COL_ATAC_CLUSTER,
  labelMeans = TRUE,
  rastr      = FALSE
) + ggtitle("snATAC-seq NK cells — 7 clusters") + pub_theme

p_umap_group <- plotEmbedding(
  proj_filt,
  colorBy    = "cellColData",
  name       = "group",
  embedding  = "UMAP",
  pal        = COL_GROUP,
  rastr      = FALSE
) + ggtitle("Genetic group") + pub_theme

p_umap_cluster | p_umap_group

pdf(paste0(params$out, "figures/snATAC_NK_UMAP_clusters-and-groups.pdf"),
    height = 4.5, width = 9, useDingbats = FALSE)
p_umap_cluster | p_umap_group; dev.off()
```


# ___Cell composition: genetic groups per cluster___

```{r fig.height=4, fig.width=5}
cell_meta <- data.frame(
  Clusters_res1 = factor(as.character(proj_filt$Clusters_res1), levels = CLUSTER_ORDER),
  group         = factor(as.character(proj_filt$group), levels = GROUP_ORDER),
  row.names     = getCellNames(proj_filt),
  stringsAsFactors = FALSE
)

comp_df <- cell_meta %>%
  count(Clusters_res1, group) %>%
  group_by(Clusters_res1) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_comp <- ggplot(comp_df, aes(x = Clusters_res1, y = pct, fill = group)) +
  geom_col(width = 0.75) +
  scale_fill_manual(values = COL_GROUP) +
  labs(title    = "Cell composition per ATAC cluster",
       subtitle = "Percentage of cells per genetic group",
       x = "ATAC cluster", y = "% cells", fill = "genetic group") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p_comp

pdf(paste0(params$out, "figures/snATAC_NK_composition_per_cluster.pdf"),
    height = 4, width = 5, useDingbats = FALSE)
p_comp; dev.off()
```

# ___Cell composition: NK functional subsets (label transfer)___

Label-transfer from matched scRNA-seq data assigns each snATAC-seq cell to one of the
four NK maturation states (hNK_Bm1–4).
- **Immunomodulatory**: hNK_Bm2 (CD56bright immature), hNK_Bm3 (transitional)
- **Cytotoxic**: hNK_Bm1 (CD56dim mature), hNK_Bm4 (CD56dim highly cytotoxic)

```{r}
NK_SUBSET_ORDER <- c("hNK_Bm1", "hNK_Bm2", "hNK_Bm3", "hNK_Bm4")
COL_NK_SUBSET   <- COLOR_CODE_v2[NK_SUBSET_ORDER]

FUNCTIONAL_MAP <- c(
  "hNK_Bm1" = "cytotoxic",
  "hNK_Bm2" = "immunomodulatory",
  "hNK_Bm3" = "immunomodulatory",
  "hNK_Bm4" = "cytotoxic"
)
COL_FUNCTIONAL   <- c("immunomodulatory" = "#4A9B6F", "cytotoxic" = "#C0392B")
FUNCTIONAL_ORDER <- c("immunomodulatory", "cytotoxic")

# Build metadata — majority_vote_label = label transfer column
lt_meta <- data.frame(
  Clusters_res1  = factor(as.character(proj_filt$Clusters_res1), levels = CLUSTER_ORDER),
  group          = factor(as.character(proj_filt$group), levels = GROUP_ORDER),
  nk_subset      = factor(as.character(proj_filt$majority_vote_label), levels = NK_SUBSET_ORDER),
  row.names      = getCellNames(proj_filt),
  stringsAsFactors = FALSE
) %>%
  mutate(functional = factor(FUNCTIONAL_MAP[as.character(nk_subset)],
                             levels = FUNCTIONAL_ORDER))

cat("Label transfer subset distribution:\n")
print(table(lt_meta$nk_subset, useNA = "always"))
cat("\nFunctional category distribution:\n")
print(table(lt_meta$functional, useNA = "always"))
```

## ___NK subset per ATAC cluster___

```{r fig.height=4, fig.width=9}
df_cl_sub <- lt_meta %>%
  filter(!is.na(nk_subset)) %>%
  count(Clusters_res1, nk_subset) %>%
  group_by(Clusters_res1) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_cl_sub <- ggplot(df_cl_sub, aes(x = Clusters_res1, y = pct, fill = nk_subset)) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = COL_NK_SUBSET) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "NK maturation subset per ATAC cluster",
       x = "ATAC cluster", y = "% cells", fill = "NK subset") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

df_cl_func <- lt_meta %>%
  filter(!is.na(functional)) %>%
  count(Clusters_res1, functional) %>%
  group_by(Clusters_res1) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_cl_func <- ggplot(df_cl_func, aes(x = Clusters_res1, y = pct, fill = functional)) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = COL_FUNCTIONAL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Cytotoxic vs immunomodulatory per ATAC cluster",
       x = "ATAC cluster", y = "% cells", fill = NULL) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p_cl_sub | p_cl_func

pdf(paste0(params$out, "figures/snATAC_NK_NKsubset_per_cluster.pdf"),
    height = 4, width = 9, useDingbats = FALSE)
p_cl_sub | p_cl_func; dev.off()
```

## ___NK subset per genetic group___

```{r fig.height=4, fig.width=9}
df_grp_sub <- lt_meta %>%
  filter(!is.na(nk_subset), !is.na(group)) %>%
  count(group, nk_subset) %>%
  group_by(group) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_grp_sub <- ggplot(df_grp_sub, aes(x = group, y = pct, fill = nk_subset)) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = COL_NK_SUBSET) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "NK maturation subset per genetic group",
       x = NULL, y = "% cells", fill = "NK subset") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

df_grp_func <- lt_meta %>%
  filter(!is.na(functional), !is.na(group)) %>%
  count(group, functional) %>%
  group_by(group) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

p_grp_func <- ggplot(df_grp_func, aes(x = group, y = pct, fill = functional)) +
  geom_col(width = 0.75, colour = "white", linewidth = 0.2) +
  scale_fill_manual(values = COL_FUNCTIONAL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Cytotoxic vs immunomodulatory per genetic group",
       x = NULL, y = "% cells", fill = NULL) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

p_grp_sub | p_grp_func

pdf(paste0(params$out, "figures/snATAC_NK_NKsubset_per_group.pdf"),
    height = 4, width = 9, useDingbats = FALSE)
p_grp_sub | p_grp_func; dev.off()
```


# ___NK cell gene modules___

I defined 25 gene modules based on integrated ATAC and RNA data. Here I want to
address the function of each modules. 

```{r}
library(ArchR)
library(SummarizedExperiment)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)

set.seed(42)

# Resolve Bioconductor/plyr conflicts with dplyr
filter    <- dplyr::filter
select    <- dplyr::select
mutate    <- dplyr::mutate
summarise <- dplyr::summarise
rename    <- dplyr::rename
count     <- dplyr::count

module_dir   <- "/home/rstudio/mnt_out/NK_project/tables/NK_module_genes"
cluster_col  <- "Clusters_res1"
group_col    <- "group"
z_cap        <- 2.0

group_order   <- c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt")
cluster_order <- paste0("C", 1:7)

# Load gene lists for M1-M25
module_files <- list.files(module_dir, pattern = "module_\\d+_genes\\.csv",
                           full.names = TRUE)
module_nums  <- as.integer(gsub(".*module_(\\d+)_genes\\.csv", "\\1",
                                basename(module_files)))
module_files <- module_files[order(module_nums)]
module_genes <- lapply(module_files, function(f) read.csv(f)$gene)
names(module_genes) <- paste0("M", sort(module_nums))
message("Loaded ", length(module_genes), " modules: ",
        paste(names(module_genes), collapse = ", "))

# ---- ArchR native addModuleScore (GeneScoreMatrix) — one module at a time ----
# Passing all modules as a named list fails in current ArchR (assigns 25 values
# to all cells instead of per-cell scores). Add each module individually instead,
# matching the pattern used for MatureNK / AdaptiveNK below.
gsm_genes <- getFeatures(proj_filt, useMatrix = "GeneScoreMatrix")
for (mod_name in names(module_genes)) {
  genes_valid <- intersect(module_genes[[mod_name]], gsm_genes)
  if (length(genes_valid) == 0) {
    message("No valid genes for ", mod_name, " in GeneScoreMatrix — skipping")
    next
  }
  message(mod_name, ": ", length(genes_valid), "/", length(module_genes[[mod_name]]),
          " genes found in GeneScoreMatrix")
  proj_filt <- addModuleScore(
    ArchRProj = proj_filt,
    useMatrix = "GeneScoreMatrix",
    name      = mod_name,
    features  = list(score = genes_valid)
  )
}

# Retrieve scores — ArchR names columns "<name>.score", e.g. "M1.score"
module_cols <- paste0(names(module_genes), ".score")
score_raw   <- as.data.frame(
  getCellColData(proj_filt, select = c(cluster_col, group_col, module_cols))
)
colnames(score_raw)[1:2] <- c("cluster", "group")
score_raw$group <- remap_group(as.character(score_raw$group))

# Rename "M1.score" → "M1" etc. for downstream compatibility
score_df        <- score_raw
colnames(score_df)[3:(2 + length(module_genes))] <- names(module_genes)
score_df$cell   <- rownames(score_df)

stopifnot(sum(!is.na(score_df$cluster)) > 0)
message("score_df: ", nrow(score_df), " cells, ",
        sum(!is.na(score_df$M1)), " with M1 scores")

# AGGREGATE: MEAN PER CLUSTER × GROUP
agg <- score_df %>%
  pivot_longer(cols = all_of(names(module_genes)),
               names_to  = "Module",
               values_to = "Score") %>%
  filter(!is.na(cluster), !is.na(group), !is.na(Score)) %>%
  group_by(cluster, group, Module) %>%
  summarise(MeanScore = mean(Score, na.rm = TRUE),
            nCells    = dplyr::n(),
            .groups   = "drop")

# warn if any combination has very few cells
sparse <- filter(agg, nCells < 10)
if (nrow(sparse) > 0) {
  message("Warning: ", nrow(sparse),
          " cluster × group × module combos have <10 cells — interpret with caution!!")
}

# Z-SCORE PER MODULE ACROSS ALL CLUSTER×GROUP COMBOS
agg <- agg %>%
  group_by(Module) %>%
  mutate(ZScore = as.numeric(scale(MeanScore))) %>%
  ungroup() %>%
  mutate(ZScore = pmax(pmin(ZScore, z_cap), -z_cap))   # cap extremes

# FACTOR ORDERING
agg$Module  <- factor(agg$Module,  levels = paste0("M", 1:25))
agg$cluster <- factor(agg$cluster, levels = cluster_order)
agg$group   <- factor(agg$group,   levels = group_order)

# x-axis labels: "C1|ATRXdel", sorted cluster-first, group-second
agg <- agg %>%
  mutate(clust_group = paste0(cluster, "|", group))

x_levels <- agg %>%
  distinct(clust_group, cluster, group) %>%
  arrange(cluster, group) %>%
  pull(clust_group)

agg$clust_group <- factor(agg$clust_group, levels = x_levels)

mat <- agg %>%
  select(Module, clust_group, ZScore) %>%
  pivot_wider(names_from  = clust_group,
              values_from = ZScore) %>%
  column_to_rownames("Module") %>%
  as.matrix()

mat <- mat[intersect(paste0("M", 1:25), rownames(mat)),
           intersect(x_levels,          colnames(mat)),
           drop = FALSE]

# column anno
col_meta <- agg %>%
  distinct(clust_group, cluster, group) %>%
  arrange(factor(clust_group, levels = x_levels))

# Extract unique present values as plain character — avoids NA-in-levels from ArchR factors
clust_levels <- intersect(cluster_order,
                          unique(na.omit(as.character(col_meta$cluster))))
group_levels <- intersect(group_order,
                          unique(na.omit(as.character(col_meta$group))))

ha_top <- HeatmapAnnotation(
  Cluster = factor(as.character(col_meta$cluster), levels = clust_levels),
  Group   = factor(as.character(col_meta$group),   levels = group_levels),
  col     = list(
    "Cluster" = COLOR_CODE_perCLUST[clust_levels],
    "Group"   = COLOR_CODE_GROUP[group_levels]
  ),
  annotation_name_side = "left",
  annotation_name_gp   = gpar(fontsize = 8),
  simple_anno_size     = unit(3, "mm"),
  gap                  = unit(1, "mm")
)

# color scale
col_fun <- colorRamp2(
  c(-z_cap, 0, z_cap),
  c("#9B59B6","#FFFFFF","#1E8449")
)

# plot heatmap
# column_split by cluster creates visual separation between the 7 cluster blocks

ht <- Heatmap(
  mat,
  name                  = "RNA Z-score",
  col                   = col_fun,
  top_annotation        = ha_top,

  # Row order: fixed M1→M25 (change to TRUE to cluster modules)
  cluster_rows          = FALSE,
  row_names_side        = "left",
  row_names_gp          = gpar(fontsize = 9),
  row_title             = NULL,

  # Column order: fixed, split by cluster
  cluster_columns       = FALSE,
  column_split          = col_meta$cluster,    # gap between clusters
  column_gap            = unit(2, "mm"),
  column_title_gp       = gpar(fontsize = 9, fontface = "bold"),
  column_names_gp       = gpar(fontsize = 7),
  column_names_rot      = 60,

  # Visuals
  border                = TRUE,
  rect_gp               = gpar(col = "white", lwd = 0.3),

  # Legend
  heatmap_legend_param  = list(
    title          = "Z-score",
    at             = c(-2, -1, 0, 1, 2),
    legend_height  = unit(3, "cm"),
    title_gp       = gpar(fontsize = 9),
    labels_gp      = gpar(fontsize = 8)
  ),

  width  = unit(22, "cm"),
  height = unit(14, "cm")
)

# save
pdf(paste0(params$out,"figures/module_activity_cluster_group.pdf"), 
    width = 16, height = 9)
draw(ht,
     heatmap_legend_side   = "right",
     annotation_legend_side = "right",
     padding = unit(c(5, 5, 5, 5), "mm"))
dev.off()

# export the scored matrix as CSV
write.csv(agg, paste0(params$out,"tables/module_scores_cluster_group_long.csv"), 
                      row.names = FALSE)
```

# ___NK cell cluster annotation based on functional NK cell gene modules___

Barplot with annotated ATAC clusters

```{r}
proj_filt$Cluster_names     <- CLUSTER_NAMES[proj_filt$Clusters_res1]
proj_filt$Metacluster_names <- METACLUSTER[proj_filt$Clusters_res1]

unique(proj_filt$Cluster_names)
unique(proj_filt$group)
```


## ___Barplot per cluster___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "Cluster_names")))

cluster_order <- c(
  "CD56bright 1", "CD56bright 2", "CD56bright 3",
  "Intermediate 1", "Intermediate 2",
  "CD56dim 1", "CD56dim 2")

meta$Cluster_names <- factor(meta$Cluster_names, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, Cluster_names, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = Cluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perCLUST_v2, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# per sample
df_sample <- meta %>%
  count(sample, Cluster_names, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = Cluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perCLUST_v2, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_cluster_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_cluster_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```

## ___Bar plot per metacluster___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "Metacluster_names")))

cluster_order <- c(
  "CD56bright",
  "Intermediate",
  "CD56dim")

meta$Metacluster_names <- factor(meta$Metacluster_names, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, Metacluster_names, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = Metacluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perMETACLUST, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# per sample
df_sample <- meta %>%
  count(sample, Metacluster_names, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = Metacluster_names)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_perMETACLUST, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_metacluster_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_metacluster_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```


### ___with the label transfer from RNA___

```{r}
meta <- as.data.frame(getCellColData(proj_filt, 
                                     select = c("sample", "group", "majority_vote_label")))

cluster_order <- c(
  "hNK_Bm1",
  "hNK_Bm2",
  "hNK_Bm3",
  "hNK_Bm4")

meta$majority_vote_label <- factor(meta$majority_vote_label, levels = cluster_order)
meta$sample <- factor(meta$sample, levels = PATIENT_ORDER)
meta$group <- factor(meta$group, levels = GROUP_ORDER)

# per group
df_group <- meta %>%
  count(group, majority_vote_label, name = "n_cells") %>%
  group_by(group) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_group <- ggplot(df_group, aes(x = group, y = pct, fill = majority_vote_label)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_v2, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per group",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# per sample
df_sample <- meta %>%
  count(sample, majority_vote_label, name = "n_cells") %>%
  group_by(sample) %>%
  mutate(pct = n_cells / sum(n_cells) * 100) %>%
  ungroup()

p_sample <- ggplot(df_sample, aes(x = sample, y = pct, fill = majority_vote_label)) +
  geom_bar(stat = "identity", width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = COLOR_CODE_v2, breaks = cluster_order) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NK-cell cluster composition per sample",
    x     = NULL,
    y     = "% of cells",
    fill  = "Cluster"
  ) +
  pub_theme +
  theme(
    axis.text.x    = element_text(angle = 35, hjust = 1, size = 8),
    legend.position = "right"
  )

# plots
library(patchwork)
combined <- p_group | p_sample

ggsave(paste0(params$out,"figures/NK_metacluster_labelTransfer_barplots.pdf"), combined,
       width = 10, height = 4, device = cairo_pdf)
ggsave(paste0(params$out,"figures/NK_metacluster_labelTransfer_barplots.png"), combined,
       width = 10, height = 4, dpi = 300)
```

```{r}
saveRDS(proj_filt, file = paste0(params$out,"/Rds/snATAC_NK_v20260401.Rds"))
```


# ___Publication-ready module score violin plots___

Publication-quality violin plots matching the figure style: pure violins colored per
genetic group, Wilcoxon significance vs control, expression-formatted x-axis labels,
colored y-axis title identifying the NK subset or module.

```{r}
library(ggsignif)
library(ggtext)

GROUP_ORDER_PUB <- c("control", "ATRXdel", "MYCNamp", "ATRXwtMYCNwt")

COL_GROUP_PUB <- c(
  "control"      = unname(COLOR_CODE_GROUP["control"]),
  "ATRXdel"      = unname(COLOR_CODE_GROUP["ATRXdel"]),
  "MYCNamp"      = unname(COLOR_CODE_GROUP["MYCNamp"]),
  "ATRXwtMYCNwt" = unname(COLOR_CODE_GROUP["ATRXwtMYCNwt"])
)

# Group label normalizer — handles ATRXm/ATRXmut/ATRXdel spelling variants
remap_group_32 <- remap_group

# x-axis label formatter — returns a parsed expression for ggplot2 label rendering
pub_x_labels <- function(breaks) {
  label_map <- list(
    "control"      = "CTRL",
    "ATRXdel"      = expression(ATRX^del),
    "MYCNamp"      = expression(MYCN^amp),
    "ATRXwtMYCNwt" = expression(ATRX^wt~MYCN^wt)
  )
  lapply(breaks, function(b) {
    if (b %in% names(label_map)) label_map[[b]] else b
  })
}

# Helper: publication violin for one module/score column, comparing all groups vs control
pub_violin_module <- function(data, score_col, y_label, title_label,
                               title_color = "black",
                               ref_group   = "control",
                               base_size   = 9) {
  df <- data %>%
    dplyr::select(group, score = dplyr::all_of(score_col)) %>%
    dplyr::filter(!is.na(group), !is.na(score)) %>%
    dplyr::mutate(group = factor(group, levels = GROUP_ORDER_PUB))

  non_ref <- setdiff(levels(df$group), ref_group)

  # Wilcoxon tests vs reference group
  stat_res <- lapply(non_ref, function(g) {
    x1 <- df$score[df$group == ref_group]
    x2 <- df$score[df$group == g]
    if (length(x1) < 3 || length(x2) < 3) return(NULL)
    p <- wilcox.test(x1, x2)$p.value
    symb <- dplyr::case_when(
      p < 1e-4 ~ "****", p < 0.001 ~ "***",
      p < 0.01 ~ "**",   p < 0.05  ~ "*",
      TRUE     ~ "ns"
    )
    list(pair = c(ref_group, g), label = symb)
  })
  stat_res <- Filter(Negate(is.null), stat_res)

  comparisons  <- lapply(stat_res, `[[`, "pair")
  annotations  <- sapply(stat_res, `[[`, "label")

  ggplot(df, aes(x = group, y = score, fill = group)) +
    geom_violin(scale = "width", trim = FALSE, alpha = 0.95, linewidth = 0) +
    {if (length(comparisons) > 0)
       ggsignif::geom_signif(
         comparisons      = comparisons,
         annotations      = annotations,
         map_signif_level = FALSE,
         step_increase    = 0.10,
         tip_length       = 0.01,
         textsize         = 3,
         color            = "black",
         vjust            = 0.3
       )} +
    scale_fill_manual(values = COL_GROUP_PUB) +
    scale_x_discrete(labels = pub_x_labels) +
    labs(
      title = title_label,
      x     = NULL,
      y     = y_label
    ) +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x     = element_text(angle = 45, hjust = 1, size = base_size - 1),
      axis.text.y     = element_text(size = base_size - 1),
      axis.title.y    = element_text(size = base_size, face = "plain"),
      axis.line       = element_line(linewidth = 0.4),
      axis.ticks      = element_line(linewidth = 0.4),
      plot.title      = element_text(size = base_size, face = "bold",
                                     color = title_color, hjust = 0),
      plot.margin     = margin(t = 25, r = 8, b = 5, l = 8)
    )
}
```

## ___Module scores within immunomodulatory NK subsets (Bm2 and Bm3)___

For cells labeled as hNK_Bm2 or hNK_Bm3 by label transfer, show a defined
"Mature NK" gene module score across genetic groups — recapitulating the figure style.

```{r}
# ArchR native addModuleScore for mature NK signature (GeneScoreMatrix)
mature_NK_genes_valid <- intersect(mature_NK_genes,
                                   getFeatures(proj_filt, useMatrix = "GeneScoreMatrix"))
cat("Mature NK genes in GeneScoreMatrix:", length(mature_NK_genes_valid),
    "/", length(mature_NK_genes), "\n")
proj_filt <- addModuleScore(
  ArchRProj = proj_filt,
  useMatrix = "GeneScoreMatrix",
  name      = "MatureNK",
  features  = list(score = mature_NK_genes_valid)
)

# Retrieve score — column is "MatureNK.score"
mature_cd <- as.data.frame(
  getCellColData(proj_filt, select = c("MatureNK.score", "group", "majority_vote_label"))
)

df_mature_atac <- data.frame(
  cell     = rownames(mature_cd),
  score    = mature_cd[["MatureNK.score"]],
  group    = remap_group(as.character(mature_cd[["group"]])),
  nk_label = as.character(mature_cd[["majority_vote_label"]]),
  stringsAsFactors = FALSE
)
cat("NK label transfer counts:\n")
print(table(df_mature_atac$nk_label, useNA = "always"))

NK_SUBSET_COLORS <- COLOR_CODE_v2[c("hNK_Bm2", "hNK_Bm3")]
```

```{r fig.height=7, fig.width=3.5}
nk_score_plots <- lapply(c("hNK_Bm2", "hNK_Bm3"), function(subset) {
  df <- df_mature_atac %>%
    dplyr::filter(nk_label == subset, !is.na(group)) %>%
    dplyr::select(group, score)

  pub_violin_module(
    data        = df,
    score_col   = "score",
    y_label     = "Mature NK gene score (ATAC)",
    title_label = paste0("Immunomodulatory ", subset),
    title_color = unname(NK_SUBSET_COLORS[subset])
  )
})

combined_nk <- patchwork::wrap_plots(nk_score_plots, ncol = 1)
combined_nk

ggsave(paste0(params$out, "figures/MatureNK_ATACscore_Bm2-Bm3_pub.pdf"),
       combined_nk, width = 3.5, height = 7, device = cairo_pdf)
```

## ___All NK cells: module scores per genetic group (all 25 modules)___

```{r fig.height=18, fig.width=14}
# score_df columns: M1..M25, cell, cluster, group (from module scoring section above)

score_df_pub <- score_df %>%
  dplyr::filter(!is.na(group)) %>%
  dplyr::mutate(group = factor(remap_group(group), levels = GROUP_ORDER_PUB))

plots_by_group <- lapply(paste0("M", 1:25), function(mod) {
  pub_violin_module(
    data        = score_df_pub,
    score_col   = mod,
    y_label     = "Module score",
    title_label = mod,
    title_color = "black"
  ) + theme(plot.margin = margin(t = 25, r = 4, b = 4, l = 4))
})

patchwork::wrap_plots(plots_by_group, ncol = 5)

ggsave(paste0(params$out, "figures/NK_module_scores_M1-M25_per_group_pub.pdf"),
       patchwork::wrap_plots(plots_by_group, ncol = 5),
       width = 14, height = 18, device = cairo_pdf)
```

## ___All NK cells: module scores per ATAC cluster (all 25 modules)___

```{r}
pub_violin_cluster <- function(data, score_col, y_label, title_label, base_size = 9) {
  df <- data %>%
    dplyr::select(cluster, score = dplyr::all_of(score_col)) %>%
    dplyr::filter(!is.na(cluster), !is.na(score)) %>%
    dplyr::mutate(cluster = factor(cluster, levels = CLUSTER_ORDER))

  ggplot(df, aes(x = cluster, y = score, fill = cluster)) +
    geom_violin(scale = "width", trim = FALSE, alpha = 0.95, linewidth = 0) +
    scale_fill_manual(values = COLOR_CODE_perCLUST[CLUSTER_ORDER]) +
    labs(title = title_label, x = NULL, y = y_label) +
    theme_classic(base_size = base_size) +
    theme(
      legend.position = "none",
      axis.text.x     = element_text(angle = 45, hjust = 1, size = base_size - 1),
      axis.text.y     = element_text(size = base_size - 1),
      axis.title.y    = element_text(size = base_size),
      axis.line       = element_line(linewidth = 0.4),
      axis.ticks      = element_line(linewidth = 0.4),
      plot.title      = element_text(size = base_size, face = "bold", hjust = 0),
      plot.margin     = margin(t = 10, r = 4, b = 4, l = 4)
    )
}
```

```{r fig.height=18, fig.width=14}
score_df_cl <- score_df %>%
  dplyr::filter(!is.na(cluster)) %>%
  dplyr::mutate(cluster = factor(cluster, levels = CLUSTER_ORDER))

plots_by_cluster <- lapply(paste0("M", 1:25), function(mod) {
  pub_violin_cluster(
    data        = score_df_cl,
    score_col   = mod,
    y_label     = "Module score",
    title_label = mod
  )
})

patchwork::wrap_plots(plots_by_cluster, ncol = 5)

ggsave(paste0(params$out, "figures/NK_module_scores_M1-M25_per_cluster_pub.pdf"),
       patchwork::wrap_plots(plots_by_cluster, ncol = 5),
       width = 14, height = 18, device = cairo_pdf)
```

## ___Module–cluster activity heatmap (cluster annotation sanity check)___

Mean module score per cluster collapsed across genetic groups, Z-scored per module
across clusters. Rows cluster automatically — modules active in the same NK states
should group together.

```{r fig.height=8, fig.width=5}
# Collapse score_df to mean per cluster × module (ignore group here)
agg_cl <- score_df %>%
  pivot_longer(cols      = all_of(names(module_genes)),
               names_to  = "Module",
               values_to = "Score") %>%
  dplyr::filter(!is.na(cluster), !is.na(Score)) %>%
  dplyr::group_by(cluster, Module) %>%
  dplyr::summarise(MeanScore = mean(Score, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(Module  = factor(Module,  levels = paste0("M", 1:25)),
                cluster = factor(cluster, levels = CLUSTER_ORDER))

mat_cl <- agg_cl %>%
  pivot_wider(names_from = cluster, values_from = MeanScore) %>%
  column_to_rownames("Module") %>%
  as.matrix()
cl_cols_avail <- intersect(CLUSTER_ORDER, colnames(mat_cl))
mat_cl <- mat_cl[intersect(paste0("M", 1:25), rownames(mat_cl)),
                 cl_cols_avail,
                 drop = FALSE]
storage.mode(mat_cl) <- "numeric"

# Z-score each module across clusters (highlights relative enrichment per cluster)
mat_cl_z <- t(scale(t(mat_cl)))
mat_cl_z[is.nan(mat_cl_z)] <- 0

ha_cl <- HeatmapAnnotation(
  Cluster = cl_cols_avail,
  col     = list(Cluster = COLOR_CODE_perCLUST[cl_cols_avail]),
  simple_anno_size     = unit(3, "mm"),
  annotation_name_side = "left",
  annotation_name_gp   = gpar(fontsize = 8)
)

ht_cl <- Heatmap(
  mat_cl_z,
  name              = "Z-score",
  col               = colorRamp2(c(-2, 0, 2), c("#4575b4", "white", "#d73027")),
  top_annotation    = ha_cl,
  cluster_rows      = TRUE,
  cluster_columns   = FALSE,
  row_names_side    = "left",
  row_names_gp      = gpar(fontsize = 9),
  column_names_side = "top",
  column_names_gp   = gpar(fontsize = 9),
  border            = TRUE,
  rect_gp           = gpar(col = "white", lwd = 0.5),
  heatmap_legend_param = list(title = "Z-score", at = c(-2, 0, 2),
                               title_gp = gpar(fontsize = 9),
                               labels_gp = gpar(fontsize = 8)),
  width  = unit(5,  "cm"),
  height = unit(12, "cm")
)

pdf(paste0(params$out, "figures/NK_module_cluster_heatmap_sanity.pdf"),
    width = 5, height = 8)
draw(ht_cl, padding = unit(c(5, 5, 5, 5), "mm"))
dev.off()

draw(ht_cl)
```

## ___Pathway enrichment analysis for M1-M25 (GO:BP + KEGG)___

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# Run GO:BP enrichment for each module
enrich_list <- lapply(names(module_genes), function(mod) {
  genes   <- unique(module_genes[[mod]])
  id_map  <- tryCatch(
    bitr(genes, fromType = "SYMBOL", toType = "ENTREZID",
         OrgDb = org.Hs.eg.db, drop = TRUE),
    error = function(e) NULL
  )
  if (is.null(id_map) || nrow(id_map) < 5) return(NULL)

  res <- tryCatch(
    enrichGO(gene          = id_map$ENTREZID,
             OrgDb         = org.Hs.eg.db,
             ont           = "BP",
             pAdjustMethod = "BH",
             pvalueCutoff  = 0.05,
             qvalueCutoff  = 0.2,
             readable      = TRUE),
    error = function(e) NULL
  )
  if (is.null(res) || nrow(as.data.frame(res)) == 0) return(NULL)
  as.data.frame(res) %>% dplyr::mutate(Module = mod)
})
names(enrich_list) <- names(module_genes)

enrich_df <- dplyr::bind_rows(Filter(Negate(is.null), enrich_list))
cat("Modules with GO:BP hits:", sum(sapply(enrich_list, Negate(is.null))),
    "/", length(module_genes), "\n")

write.csv(enrich_df,
          paste0(params$out, "tables/NK_module_GO_enrichment.csv"),
          row.names = FALSE)
```

```{r fig.height=8, fig.width=6}
# Top 3 terms per module → dot plot
top_terms <- enrich_df %>%
  dplyr::mutate(Module = factor(Module, levels = paste0("M", 1:25))) %>%
  dplyr::group_by(Module) %>%
  dplyr::slice_min(p.adjust, n = 3, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    Description = stringr::str_wrap(Description, width = 45),
    log10p      = -log10(p.adjust)
  )

ggplot(top_terms, aes(x = Module, y = Description, size = Count,
                      colour = log10p)) +
  geom_point() +
  scale_colour_gradient(low = "#fee090", high = "#d73027",
                        name = expression(-log[10](p[adj]))) +
  scale_size_continuous(range = c(2, 6), name = "Gene count") +
  labs(x = NULL, y = NULL,
       title = "Top GO:BP terms per module") +
  theme_classic(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    axis.text.y  = element_text(size = 7),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3),
    plot.title   = element_text(face = "bold", size = 10)
  )

ggsave(paste0(params$out, "figures/NK_module_GO_dotplot.pdf"),
       width = 6, height = 8, device = cairo_pdf)
```

## ___Combined GRN module scores___

Modules grouped into seven biologically defined GRN programmes (based on module gene content
and NK cell biology):

| GRN | NK state | Modules |
|-----|----------|---------|
| Immunomodulatory GRN | CD56bright / Immature | M2, M4 |
| Cytotoxic GRN | CD56dim Cytotoxic | M8, M10, M15, M19, M20 |
| Adaptive GRN | Adaptive / Memory-like | M12, M13, M16 |
| IFN Response GRN | IFN Response | M7, M11 |
| Antigen licensing GRN | MHC-I / Antigen Processing | M9, M14, M18 |
| Tissue Resident GRN | Tissue Resident | M17 |
| Transitional GRN | Developmental / Transitional | M1, M3, M5, M6 |

```{r}
GRN_MODULES <- list(
  "Immunomodulatory GRN"  = paste0("M", c(2, 4)),
  "Cytotoxic GRN"         = paste0("M", c(8, 10, 15, 19, 20)),
  "Adaptive GRN"          = paste0("M", c(12, 13, 16)),
  "IFN Response GRN"      = paste0("M", c(7, 11)),
  "Antigen licensing GRN" = paste0("M", c(9, 14, 18)),
  "Tissue Resident GRN"   = paste0("M", c(17)),
  "Transitional GRN"      = paste0("M", c(1, 3, 5, 6))
)

GRN_COLORS <- c(
  "Immunomodulatory GRN"  = unname(COL_FUNCTIONAL["immunomodulatory"]),  # #4A9B6F — matches COL_FUNCTIONAL
  "Cytotoxic GRN"         = unname(COL_FUNCTIONAL["cytotoxic"]),          # #C0392B — matches COL_FUNCTIONAL
  "Adaptive GRN"          = "#7B241C",                                     # dark maroon — deeper shade of cytotoxic red, adaptive = terminally mature
  "IFN Response GRN"      = unname(COLOR_CODE_v2["hNK_Bm4"]),             # #CDA646 — gold
  "Antigen licensing GRN" = unname(COLOR_CODE_v2["hNK_Bm1"]),             # #6376AE — slate blue
  "Tissue Resident GRN"   = unname(COLOR_CODE_GROUP["MYCNamp"]),           # #A7563C — warm terracotta
  "Transitional GRN"      = unname(COLOR_CODE_GROUP["control"])            # #7EA3AC — muted steel blue, neutral
)

GRN_ORDER <- c(
  "Immunomodulatory GRN", "Cytotoxic GRN", "Adaptive GRN",
  "IFN Response GRN", "Antigen licensing GRN",
  "Tissue Resident GRN", "Transitional GRN"
)

# Compute per-cell GRN score = mean of constituent module scores
mod_cols <- names(module_genes)
for (grn in names(GRN_MODULES)) {
  mods_use <- intersect(GRN_MODULES[[grn]], mod_cols)
  score_df[[grn]] <- rowMeans(score_df[, mods_use, drop = FALSE], na.rm = TRUE)
}

grn_df <- score_df %>%
  dplyr::select(cell, cluster, group, all_of(GRN_ORDER)) %>%
  dplyr::mutate(
    group   = factor(remap_group(group),                   levels = GROUP_ORDER_PUB),
    cluster = factor(as.character(cluster), levels = CLUSTER_ORDER)
  )

cat("GRN score summary:\n")
print(summary(grn_df[, GRN_ORDER]))
```

```{r fig.height=4, fig.width=10}
# Violin: GRN score per genetic group
grn_group_plots <- lapply(GRN_ORDER, function(grn) {
  df <- grn_df %>%
    dplyr::filter(!is.na(group)) %>%
    dplyr::select(group, score = dplyr::all_of(grn))

  pub_violin_module(
    data        = df,
    score_col   = "score",
    y_label     = "GRN score",
    title_label = grn,
    title_color = GRN_COLORS[grn]
  )
})

p_grn_group <- patchwork::wrap_plots(grn_group_plots, ncol = 4)
p_grn_group

ggsave(paste0(params$out, "figures/NK_GRN_scores_per_group_pub.pdf"),
       p_grn_group, width = 14, height = 8, device = cairo_pdf)
```

```{r fig.height=4, fig.width=10}
# Violin: GRN score per ATAC cluster

pub_violin_grn_cluster <- function(data, score_col, grn_name, base_size = 9) {
  df <- data %>%
    dplyr::filter(!is.na(cluster)) %>%
    dplyr::select(cluster, score = dplyr::all_of(score_col))

  ggplot(df, aes(x = cluster, y = score, fill = cluster)) +
    geom_violin(scale = "width", trim = FALSE, alpha = 0.95, linewidth = 0) +
    scale_fill_manual(values = COLOR_CODE_perCLUST[CLUSTER_ORDER]) +
    labs(title = grn_name, x = NULL, y = "GRN score") +
    pub_theme +
    theme(
      legend.position = "none",
      axis.text.x  = element_text(angle = 45, hjust = 1, size = base_size - 1),
      axis.text.y  = element_text(size = base_size - 1),
      axis.line    = element_line(linewidth = 0.4),
      axis.ticks   = element_line(linewidth = 0.4),
      plot.title   = element_text(size = base_size, face = "bold",
                                  color = GRN_COLORS[grn_name], hjust = 0),
      plot.margin  = margin(t = 10, r = 4, b = 4, l = 4)
    )
}

grn_cluster_plots <- lapply(GRN_ORDER, function(grn) {
  pub_violin_grn_cluster(grn_df, grn, grn)
})

p_grn_cluster <- patchwork::wrap_plots(grn_cluster_plots, ncol = 4)
p_grn_cluster

ggsave(paste0(params$out, "figures/NK_GRN_scores_per_cluster_pub.pdf"),
       p_grn_cluster, width = 14, height = 8, device = cairo_pdf)
```

```{r fig.height=5, fig.width=6}
# Heatmap: mean GRN score per cluster × GRN (compact overview)
grn_cl_mat <- sapply(GRN_ORDER, function(grn) {
  grn_df %>%
    dplyr::filter(!is.na(cluster)) %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(m = mean(.data[[grn]], na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(factor(cluster, levels = CLUSTER_ORDER)) %>%
    dplyr::pull(m)
})
cl_avail <- intersect(CLUSTER_ORDER, unique(na.omit(as.character(grn_df$cluster))))
rownames(grn_cl_mat) <- cl_avail

grn_cl_z <- scale(grn_cl_mat)   # Z-score across GRNs per cluster

# t(grn_cl_z): rows = GRNs, cols = clusters — row annotation marks GRN category by colour
grn_row_order <- intersect(GRN_ORDER, rownames(t(grn_cl_z)))

ra_grn <- rowAnnotation(
  GRN  = grn_row_order,
  col  = list(GRN = GRN_COLORS),
  simple_anno_size     = unit(3, "mm"),
  annotation_name_gp   = gpar(fontsize = 8),
  show_annotation_name = FALSE
)

ht_grn <- Heatmap(
  t(grn_cl_z)[grn_row_order, , drop = FALSE],
  name              = "Z-score",
  col               = colorRamp2(c(-1.5, 0, 1.5), c("#4575b4", "white", "#d73027")),
  left_annotation   = ra_grn,
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  row_names_side    = "right",
  row_names_gp      = gpar(fontsize = 8),
  column_names_side = "top",
  column_names_gp   = gpar(fontsize = 9),
  border            = TRUE,
  rect_gp           = gpar(col = "white", lwd = 1),
  heatmap_legend_param = list(title = "Z-score",
                               title_gp = gpar(fontsize = 9),
                               labels_gp = gpar(fontsize = 8)),
  width  = unit(5, "cm"),
  height = unit(7, "cm")
)

pdf(paste0(params$out, "figures/NK_GRN_cluster_heatmap.pdf"), width = 6, height = 7)
draw(ht_grn, padding = unit(c(5, 5, 5, 5), "mm"))
dev.off()

draw(ht_grn)
```

## ___Dotplot: GRN scores per NK subset × genetic group___

```{r fig.height=5.5, fig.width=10}
# Join NK label-transfer subset (majority_vote_label) into GRN data via cell names
grn_dot_df <- grn_df %>%
  dplyr::mutate(
    nk_subset = factor(
      as.character(lt_meta[cell, "nk_subset"]),
      levels = NK_SUBSET_ORDER
    ),
    group = factor(as.character(group), levels = GROUP_ORDER_PUB)
  ) %>%
  dplyr::filter(!is.na(nk_subset), !is.na(group))

# Aggregate mean score + % positive per NK subset × group × GRN,
# then z-score each GRN across all NK subset × group combinations
dot_agg <- grn_dot_df %>%
  tidyr::pivot_longer(cols = dplyr::all_of(GRN_ORDER),
                      names_to = "GRN", values_to = "score") %>%
  dplyr::filter(!is.na(score)) %>%
  dplyr::group_by(nk_subset, group, GRN) %>%
  dplyr::summarise(
    mean_score = mean(score, na.rm = TRUE),
    pct_pos    = mean(score > 0, na.rm = TRUE) * 100,
    n_cells    = dplyr::n(),
    .groups    = "drop"
  ) %>%
  dplyr::filter(n_cells >= 5) %>%
  dplyr::group_by(GRN) %>%
  dplyr::mutate(z_score = as.numeric(scale(mean_score))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    z_score = pmax(pmin(z_score, 2), -2),   # cap at ±2
    GRN     = factor(GRN, levels = rev(GRN_ORDER))
  )

p_grn_dot <- ggplot(dot_agg,
                    aes(x = nk_subset, y = GRN,
                        size = pct_pos, colour = z_score)) +
  geom_point() +
  facet_grid(. ~ group, scales = "free_x", space = "free_x") +
  scale_size_continuous(range = c(0.5, 5),
                        name  = "% cells\n(score > 0)",
                        limits = c(0, 100)) +
  scale_colour_gradient2(
    low      = "#4575b4",
    mid      = "white",
    high     = "#d73027",
    midpoint = 0,
    limits   = c(-2, 2),
    name     = "GRN score\n(z-score)"
  ) +
  labs(
    title = "GRN scores per NK subset and genetic group",
    x = NULL, y = NULL
  ) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 7,
                                  face = "italic"),
    axis.text.y   = element_text(size = 8),
    strip.text    = element_text(size = 8, face = "bold"),
    panel.spacing = unit(1, "mm"),
    panel.border  = element_rect(colour = "grey80", fill = NA, linewidth = 0.4)
  )

p_grn_dot

ggsave(paste0(params$out, "figures/NK_GRN_scores_NKsubset-group_dotplot.pdf"),
       p_grn_dot, width = 10, height = 5.5, device = cairo_pdf)
```

## ___Dotplot: all module scores (M1–M25) per NK subset × genetic group___

```{r fig.height=12, fig.width=10}
# Reuse grn_dot_df (already has nk_subset + group); add all module scores
mod_dot_df <- score_df %>%
  dplyr::mutate(
    nk_subset = factor(
      as.character(lt_meta[cell, "nk_subset"]),
      levels = NK_SUBSET_ORDER
    ),
    group = factor(remap_group(as.character(group)), levels = GROUP_ORDER_PUB)
  ) %>%
  dplyr::filter(!is.na(nk_subset), !is.na(group))

mod_dot_agg <- mod_dot_df %>%
  tidyr::pivot_longer(cols = dplyr::all_of(names(module_genes)),
                      names_to = "Module", values_to = "score") %>%
  dplyr::filter(!is.na(score)) %>%
  dplyr::group_by(nk_subset, group, Module) %>%
  dplyr::summarise(
    mean_score = mean(score, na.rm = TRUE),
    pct_pos    = mean(score > 0, na.rm = TRUE) * 100,
    n_cells    = dplyr::n(),
    .groups    = "drop"
  ) %>%
  dplyr::filter(n_cells >= 5) %>%
  dplyr::group_by(Module) %>%
  dplyr::mutate(z_score = as.numeric(scale(mean_score))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    z_score = pmax(pmin(z_score, 2), -2),
    Module  = factor(Module, levels = rev(paste0("M", 1:25)))
  )

p_mod_dot <- ggplot(mod_dot_agg,
                    aes(x = nk_subset, y = Module,
                        size = pct_pos, colour = z_score)) +
  geom_point() +
  facet_grid(. ~ group, scales = "free_x", space = "free_x") +
  scale_size_continuous(range = c(0.3, 4),
                        name  = "% cells\n(score > 0)",
                        limits = c(0, 100)) +
  scale_colour_gradient2(
    low      = "#4575b4",
    mid      = "white",
    high     = "#d73027",
    midpoint = 0,
    limits   = c(-2, 2),
    name     = "Module score\n(z-score)"
  ) +
  labs(
    title = "Module scores per NK subset and genetic group",
    x = NULL, y = NULL
  ) +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 7, face = "italic"),
    axis.text.y   = element_text(size = 7),
    strip.text    = element_text(size = 8, face = "bold"),
    panel.spacing = unit(1, "mm"),
    panel.border  = element_rect(colour = "grey80", fill = NA, linewidth = 0.4)
  )

p_mod_dot

ggsave(paste0(params$out, "figures/NK_module_scores_NKsubset-group_dotplot.pdf"),
       p_mod_dot, width = 10, height = 12, device = cairo_pdf)
```

# ___Adaptive NK cell signature: ModuleScore___

The question we want to answer here is if we have any adaptive NK cells in the
ATAC data. When doing the label transfer from RNA data of matched patients,
we saw a clear signature for adaptive NK cells in a small proportion of NK cells,
however in the ATAC data, the label transfer did not identify any adaptive NK cells.

```{r}
# ArchR native addModuleScore for adaptive NK signature (GeneScoreMatrix)
adaptive_NK_genes_valid <- intersect(adaptive_NK_genes,
                                     getFeatures(proj_filt, useMatrix = "GeneScoreMatrix"))
cat("Adaptive NK genes in GeneScoreMatrix:", length(adaptive_NK_genes_valid),
    "/", length(adaptive_NK_genes), "\n")
proj_filt <- addModuleScore(
  ArchRProj = proj_filt,
  useMatrix = "GeneScoreMatrix",
  name      = "AdaptiveNK",
  features  = list(score = adaptive_NK_genes_valid)
)

# Retrieve score — column is "AdaptiveNK.score"
adapt_cd <- as.data.frame(
  getCellColData(proj_filt,
                 select = c("AdaptiveNK.score", "group", "Clusters_res1",
                            "majority_vote_label"))
)

df_adapt_atac <- data.frame(
  cell     = rownames(adapt_cd),
  score    = adapt_cd[["AdaptiveNK.score"]],
  group    = factor(remap_group(as.character(adapt_cd[["group"]])),
                    levels = GROUP_ORDER_PUB),
  cluster  = factor(as.character(adapt_cd[["Clusters_res1"]]), levels = CLUSTER_ORDER),
  nk_label = as.character(adapt_cd[["majority_vote_label"]]),
  stringsAsFactors = FALSE
)
```

```{r fig.height=3.5, fig.width=3.5}
# Adaptive NK score per genetic group
p_adapt_group <- pub_violin_module(
  data        = df_adapt_atac %>% dplyr::filter(!is.na(group)),
  score_col   = "score",
  y_label     = "Adaptive NK gene score (ATAC)",
  title_label = "Adaptive NK signature — per group",
  title_color = "#2C7BB6"
)
p_adapt_group

ggsave(paste0(params$out, "figures/AdaptiveNK_ATACscore_per_group_pub.pdf"),
       p_adapt_group, width = 3.5, height = 3.5, device = cairo_pdf)
```

```{r fig.height=3.5, fig.width=4}
# Adaptive NK score per ATAC cluster C1-C7
p_adapt_cluster <- pub_violin_cluster(
  data        = df_adapt_atac %>% dplyr::filter(!is.na(cluster)),
  score_col   = "score",
  y_label     = "Adaptive NK gene score (ATAC)",
  title_label = "Adaptive NK signature — per ATAC cluster"
)
p_adapt_cluster

ggsave(paste0(params$out, "figures/AdaptiveNK_ATACscore_per_ATACcluster_pub.pdf"),
       p_adapt_cluster, width = 4, height = 3.5, device = cairo_pdf)
```

```{r fig.height=3.5, fig.width=5}
# Adaptive NK score per NK label-transfer subset, split by genetic group
p_adapt_nk <- df_adapt_atac %>%
  dplyr::filter(!is.na(nk_label), !is.na(group)) %>%
  dplyr::mutate(nk_label = factor(nk_label,
                                  levels = c("hNK_Bm1","hNK_Bm2","hNK_Bm3","hNK_Bm4"))) %>%
  ggplot(aes(x = nk_label, y = score, fill = group)) +
  geom_violin(scale = "width", position = position_dodge(0.85),
              trim = FALSE, alpha = 0.85, linewidth = 0) +
  scale_fill_manual(values = COL_GROUP_PUB) +
  labs(title = "Adaptive NK signature — per NK subset (ATAC)",
       x = NULL, y = "Adaptive NK gene score (ATAC)", fill = "genetic group") +
  theme_classic(base_size = 9) +
  theme(axis.text.x    = element_text(angle = 35, hjust = 1),
        axis.line      = element_line(linewidth = 0.4),
        axis.ticks     = element_line(linewidth = 0.4),
        plot.title     = element_text(face = "bold", size = 9),
        legend.key.size = unit(3, "mm"),
        legend.text    = element_text(size = 8))
p_adapt_nk

ggsave(paste0(params$out, "figures/AdaptiveNK_ATACscore_per_NKsubset_pub.pdf"),
       p_adapt_nk, width = 5, height = 3.5, device = cairo_pdf)
```

```{r}
# Extract GeneScoreMatrix for per-gene accessibility plots (KLRC2, FCER1G)
se_markers  <- getMatrixFromProject(proj_filt, useMatrix = "GeneScoreMatrix")
gene_names  <- rowData(se_markers)$name
if (is.null(gene_names)) gene_names <- rownames(se_markers)
rownames(se_markers) <- gene_names
full_expr   <- as.matrix(assay(se_markers))
```

```{r fig.height=3.5, fig.width=7}
# Gene accessibility for key adaptive NK markers: KLRC2 (gained) and FCER1G (lost)
marker_genes <- intersect(c("KLRC2", "FCER1G"), gene_names)

dplyr::bind_rows(lapply(marker_genes, function(g) {
  data.frame(
    gene     = g,
    score    = as.numeric(full_expr[g, ]),
    nk_label = as.character(proj_filt$majority_vote_label),
    group    = factor(remap_group(as.character(proj_filt$group)), levels = GROUP_ORDER_PUB),
    stringsAsFactors = FALSE
  )
})) %>%
  dplyr::filter(!is.na(nk_label), !is.na(group)) %>%
  dplyr::mutate(nk_label = factor(nk_label,
                                  levels = c("hNK_Bm1","hNK_Bm2","hNK_Bm3","hNK_Bm4"))) %>%
  ggplot(aes(x = nk_label, y = score, fill = group)) +
  geom_violin(scale = "width", position = position_dodge(0.85),
              trim = FALSE, alpha = 0.85, linewidth = 0) +
  facet_wrap(~ gene, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = COL_GROUP_PUB) +
  labs(x = NULL, y = "Gene accessibility score (ATAC)", fill = "genetic group") +
  theme_classic(base_size = 9) +
  theme(axis.text.x    = element_text(angle = 35, hjust = 1),
        strip.text     = element_text(face = "bold", size = 9),
        legend.key.size = unit(3, "mm"),
        legend.text    = element_text(size = 8)) ->
  p_markers
p_markers

ggsave(paste0(params$out, "figures/AdaptiveNK_KLRC2_FCER1G_ATACscore_pub.pdf"),
       p_markers, width = 7, height = 3.5, device = cairo_pdf)
```


# ___TF footprinting: setup___


```{r}
TF_LIST <- unique(TFs)
cat("TFs to footprint:", paste(TF_LIST, collapse = ", "), "\n")

proj_filt$clust_group <- factor(paste0(
  as.character(proj_filt$Clusters_res1), "_",
  as.character(proj_filt$group)
))
cat("\nclust_group combinations:\n")
print(sort(table(proj_filt$clust_group), decreasing = TRUE))
```

## ___Add group coverages (cluster x group)___

Creates pseudo-bulk coverage tracks for each cluster x genetic group combination.
`force = FALSE` skips recomputation if coverages already exist.

```{r}
proj_filt <- addGroupCoverages(
  ArchRProj    = proj_filt,
  groupBy      = "clust_group",
  minCells     = 5,       # lowered from 10 — some clusters have few ATRXdel cells
  maxCells     = 500,
  maxFragments = 25 * 10^6,
  force        = TRUE,    # recompute now that group names are correctly remapped
  verbose      = TRUE
)
```

## ___Match TF names to JASPAR motifs___

```{r}
# Load motif positions directly — avoids stale internal path stored in the Rds
motif_pos_path <- paste0(params$out, "ArchR_v20260203/Annotations/Motif-Positions-In-Peaks.rds")
cat("Motif positions path:", motif_pos_path, "\n")
cat("File exists:", file.exists(motif_pos_path), "\n")

if (file.exists(motif_pos_path)) {
  motif_positions <- readRDS(motif_pos_path)
} else {
  # Fallback: let ArchR try (may fail if annotation path is also stale)
  motif_positions <- tryCatch(
    getPositions(proj_filt),
    error = function(e) {
      stop("Cannot load motif positions from disk or via getPositions().\n",
           "Expected file: ", motif_pos_path, "\n",
           "Original error: ", conditionMessage(e))
    }
  )
}

motif_names_all <- names(motif_positions)
cat("Total motifs loaded:", length(motif_names_all), "\n")

match_tf_to_motif <- function(tf, motif_names) {
  exact <- motif_names[grepl(paste0("_", tf, "$"), motif_names, ignore.case = TRUE)]
  if (length(exact) > 0) return(exact[1])
  partial <- motif_names[grepl(tf, motif_names, ignore.case = TRUE)]
  if (length(partial) > 0) return(partial[1])
  NA_character_
}

tf_motif_map <- data.frame(
  TF         = TF_LIST,
  motif_name = sapply(TF_LIST, match_tf_to_motif, motif_names = motif_names_all),
  stringsAsFactors = FALSE
)
tf_motif_map_valid <- tf_motif_map[!is.na(tf_motif_map$motif_name), ]
cat("\nMatched TF -> motif:\n")
print(tf_motif_map_valid)

tf_motif_names <- tf_motif_map_valid$motif_name
```

## ___Compute TF footprints___

```{r}
footprints <- getFootprints(
  ArchRProj = proj_filt,
  positions = motif_positions[tf_motif_names],
  groupBy   = "clust_group",
  flank     = 250
)
cat("Footprint SE dimensions:", dim(footprints), "\n")
cat("Groups:", paste(colnames(footprints), collapse = ", "), "\n")
```

# ___Plot footprints by cluster (C1–C7, all groups pooled)___

Compare TF footprints across the 7 ATAC clusters, pooling all genetic groups.

```{r}
proj_filt <- addGroupCoverages(
  ArchRProj    = proj_filt,
  groupBy      = "Clusters_res1",
  minCells     = 5,
  maxCells     = 500,
  maxFragments = 25 * 10^6,
  force        = TRUE,
  verbose      = TRUE
)
```

```{r}
footprints_cluster <- getFootprints(
  ArchRProj = proj_filt,
  positions = motif_positions[tf_motif_names],
  groupBy   = "Clusters_res1",
  flank     = 250
)
cat("Cluster footprint columns:", paste(colnames(footprints_cluster), collapse = ", "), "\n")
```

```{r fig.height=4, fig.width=7}
cl_col_to_canonical <- function(col_names) {
  sub("\\._\\..*$", "", col_names)   # "C1._.Rep1" → "C1"
}

# pal names must match what plotFootprints uses (strips ._.* suffix internally)
cl_pal <- setNames(
  COL_ATAC_CLUSTER[cl_col_to_canonical(colnames(footprints_cluster))],
  cl_col_to_canonical(colnames(footprints_cluster))
)
cl_pal[is.na(cl_pal)] <- "gray50"

tryCatch({
  plotFootprints(
    seFoot       = footprints_cluster,
    ArchRProj    = proj_filt,
    normMethod   = "Subtract",
    plotName     = "Footprints_by-cluster",
    addDOC       = FALSE,
    smoothWindow = 10,
    pal          = cl_pal
  )
  message("Footprint PDF saved: by-cluster")
}, error = function(e) message("Footprint error (by-cluster): ", conditionMessage(e)))
```

# ___Custom ggplot2 footprint plots___

```{r fig.height=5, fig.width=10}
extract_footprint_df <- function(seFoot, tf_motif, groups) {
  if (!(tf_motif %in% rownames(seFoot))) return(NULL)
  mat <- assay(seFoot, "footprint")
  row_i <- which(rownames(mat) == tf_motif)
  if (length(row_i) == 0) return(NULL)
  df <- as.data.frame(t(mat[row_i, , drop = FALSE]))
  df$position <- seq(-250, 250, length.out = nrow(df))
  colnames(df)[1] <- "footprint_val"
  # mat is positions x groups in ArchR output
  df_long <- as.data.frame(mat[row_i, , drop = TRUE])
  df_long <- data.frame(
    clust_group = colnames(mat),
    footprint   = as.numeric(mat[row_i, ]),
    position    = rep(seq(-250, 250, length.out = ncol(mat)), each = 1)
  )
  # Reshape: mat is [motifs x positions], groups are in colnames
  # ArchR footprint SE: rows=motifs, cols=groups; values are vectors per position
  # Use the correct structure
  pos_seq <- seq(-250, 250, length.out = ncol(mat))
  do.call(rbind, lapply(groups[groups %in% colnames(mat)], function(g) {
    data.frame(
      clust_group = g,
      group       = clust_col_to_group(g),   # strips cluster prefix + ._.Rep suffix, remaps
      position    = pos_seq,
      footprint   = as.numeric(mat[row_i, g, drop = FALSE]),
      stringsAsFactors = FALSE
    )
  }))
}

plot_cluster_footprints <- function(seFoot, cluster_id, tf_motif_map, group_colors,
                                    cluster_name) {
  # Match by prefix — handles ATRXm / ATRXdel / ATRXmut naming variants
  all_groups <- colnames(seFoot)[grepl(paste0("^", cluster_id, "_"), colnames(seFoot))]
  df_list <- lapply(seq_len(nrow(tf_motif_map)), function(i) {
    df <- extract_footprint_df(seFoot, tf_motif_map$motif_name[i], all_groups)
    if (!is.null(df)) df$TF <- tf_motif_map$TF[i]
    df
  })
  df_all <- bind_rows(df_list)
  if (is.null(df_all) || nrow(df_all) == 0) {
    message("No footprint data for cluster ", cluster_id); return(invisible(NULL))
  }
  df_all$group <- factor(df_all$group, levels = names(group_colors))
  ggplot(df_all, aes(x = position, y = footprint, color = group)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.3) +
    geom_line(linewidth = 0.6, alpha = 0.85) +
    facet_wrap(~ TF, scales = "free_y", ncol = 4) +
    scale_color_manual(values = group_colors) +
    labs(
      title    = paste0(cluster_id, ": ", cluster_name),
      subtitle = "TF footprint — genetic groups (bias-subtracted)",
      x        = "Distance from motif center (bp)",
      y        = "Normalized insertion frequency",
      color    = "genetic group"
    ) +
    theme(
      legend.position = "top",
      strip.text      = element_text(size = 8, face = "italic"),
      panel.spacing   = unit(2, "mm")
    )
}

for (cl in CLUSTER_ORDER) {
  p <- tryCatch(
    plot_cluster_footprints(footprints, cl, tf_motif_map_valid, COL_GROUP, CLUSTER_NAMES[cl]),
    error = function(e) { message("ggplot footprint error for ", cl, ": ", conditionMessage(e)); NULL }
  )
  if (!is.null(p)) {
    tryCatch(print(p), error = function(e) message("print error: ", conditionMessage(e)))
    tryCatch({
      pdf(paste0(params$out, "figures/snATAC_NK_footprints_", cl, "_genetic-groups.pdf"),
          height = 5, width = 10, useDingbats = FALSE)
      print(p); dev.off()
    }, error = function(e) { try(dev.off(), silent = TRUE); message(conditionMessage(e)) })
  }
}
```

# ___ChromVAR TF deviation scores___

```{r fig.height=6, fig.width=10}
tryCatch({
  chrom_se <- getMatrixFromProject(proj_filt, useMatrix = "MotifMatrix")

  row_idx <- which(rowData(chrom_se)$name %in% tf_motif_map_valid$motif_name)
  if (length(row_idx) == 0) {
    row_idx <- which(grepl(paste(TF_LIST, collapse = "|"),
                           rowData(chrom_se)$name, ignore.case = TRUE))
  }
  if (length(row_idx) == 0) stop("No matching TFs in MotifMatrix")

  chrom_sub <- chrom_se[row_idx, ]
  cell_df <- data.frame(
    Clusters_res1 = as.character(proj_filt$Clusters_res1),
    group         = as.character(proj_filt$group),
    clust_group   = as.character(proj_filt$clust_group),
    cell          = getCellNames(proj_filt),
    stringsAsFactors = FALSE
  )
  dev_mat <- assay(chrom_sub, 2)
  rownames(dev_mat) <- rowData(chrom_sub)$name

  clust_group_levels <- levels(proj_filt$clust_group)
  pb_mat <- vapply(clust_group_levels, function(cg) {
    cells <- cell_df$cell[cell_df$clust_group == cg]
    cells <- cells[cells %in% colnames(dev_mat)]
    if (length(cells) < 5) return(rep(NA_real_, nrow(dev_mat)))
    rowMeans(dev_mat[, cells, drop = FALSE], na.rm = TRUE)
  }, numeric(nrow(dev_mat)))

  motif_to_tf    <- setNames(tf_motif_map_valid$TF, tf_motif_map_valid$motif_name)
  rownames(pb_mat) <- ifelse(rownames(pb_mat) %in% names(motif_to_tf),
                             motif_to_tf[rownames(pb_mat)], rownames(pb_mat))
  pb_mat <- pb_mat[, colSums(is.na(pb_mat)) < nrow(pb_mat) / 2, drop = FALSE]

  col_ann_df <- data.frame(
    cluster   = sub("_.*", "", colnames(pb_mat)),
    group     = sub("^[^_]+_", "", colnames(pb_mat)),
    row.names = colnames(pb_mat)
  )
  col_ha <- HeatmapAnnotation(
    cluster = col_ann_df$cluster,
    group   = col_ann_df$group,
    col     = list(
      cluster = COL_ATAC_CLUSTER[unique(col_ann_df$cluster)],
      group   = COL_GROUP[unique(col_ann_df$group)]
    ),
    annotation_name_gp = gpar(fontsize = 8)
  )

  pb_z <- t(scale(t(pb_mat)))
  pb_z[pb_z >  2] <-  2
  pb_z[pb_z < -2] <- -2

  ht <- Heatmap(
    pb_z,
    name              = "deviation\nz-score",
    top_annotation    = col_ha,
    row_names_gp      = gpar(fontsize = 8, fontface = "italic"),
    column_names_gp   = gpar(fontsize = 7),
    col               = colorRamp2(c(-2, 0, 2),
                                   c(COL_GROUP["control"], "white", COL_GROUP["ATRXdel"])),
    cluster_rows      = TRUE,
    cluster_columns   = FALSE,
    column_order      = colnames(pb_mat),
    show_column_names = TRUE,
    border            = TRUE,
    column_title      = "ChromVAR TF deviation: NK ATAC clusters x genetic groups",
    column_title_gp   = gpar(fontsize = 10, fontface = "bold")
  )
  draw(ht)
  pdf(paste0(params$out, "figures/snATAC_NK_chromVAR_TF_deviation_heatmap.pdf"),
      height = 6, width = 10, useDingbats = FALSE)
  draw(ht); dev.off()
  cat("ChromVAR heatmap saved. TFs:", paste(rownames(pb_z), collapse = ", "), "\n")

}, error = function(e) {
  message("ChromVAR MotifMatrix not available or error: ", conditionMessage(e))
  message("Run addDeviationsMatrix() on the ArchR project if needed.")
})
```

# ___Footprint depth summary heatmap___

```{r fig.height=6, fig.width=9}
compute_footprint_score <- function(seFoot, tf_motif, groups, center_bp = 25) {
  if (!(tf_motif %in% rownames(seFoot))) return(NULL)
  mat   <- assay(seFoot, "footprint")
  row_i <- which(rownames(mat) == tf_motif)
  if (length(row_i) == 0) return(NULL)
  pos        <- seq(-250, 250, length.out = ncol(mat))
  center_idx <- abs(pos) <= center_bp
  flank_idx  <- abs(pos) >= 100 & abs(pos) <= 200
  sapply(groups, function(g) {
    if (!(g %in% colnames(mat))) return(NA_real_)
    mean(mat[row_i, flank_idx], na.rm = TRUE) -
    mean(mat[row_i, center_idx], na.rm = TRUE)
  })
}

all_cg <- colnames(footprints)
score_df <- bind_rows(lapply(seq_len(nrow(tf_motif_map_valid)), function(i) {
  sc <- compute_footprint_score(footprints, tf_motif_map_valid$motif_name[i], all_cg)
  if (is.null(sc)) return(NULL)
  data.frame(TF = tf_motif_map_valid$TF[i], clust_group = names(sc),
             score = as.numeric(sc), stringsAsFactors = FALSE)
}))

score_df <- score_df %>%
  filter(!is.na(score)) %>%
  mutate(
    cluster = factor(base::sub("_.*", "", clust_group), levels = CLUSTER_ORDER),
    group   = factor(clust_col_to_group(clust_group), levels = GROUP_ORDER)
  )

score_wide <- score_df %>%
  mutate(col_id = paste0(cluster, "|", group)) %>%
  dplyr::select(TF, col_id, score) %>%
  pivot_wider(names_from = col_id, values_from = score) %>%
  tibble::column_to_rownames("TF")

x_levels <- as.vector(outer(CLUSTER_ORDER, GROUP_ORDER, paste, sep = "|"))
x_levels <- x_levels[x_levels %in% colnames(score_wide)]
score_wide <- score_wide[, x_levels, drop = FALSE]

score_z <- t(scale(t(as.matrix(score_wide))))
score_z[score_z >  2] <-  2
score_z[score_z < -2] <- -2

col_ann_sc <- data.frame(
  cluster   = base::sub("\\|.*", "", colnames(score_wide)),
  group     = base::sub(".*\\|", "", colnames(score_wide)),
  row.names = colnames(score_wide)
)
ha_sc <- HeatmapAnnotation(
  cluster = col_ann_sc$cluster,
  group   = col_ann_sc$group,
  col     = list(
    cluster = COL_ATAC_CLUSTER[unique(col_ann_sc$cluster)],
    group   = COL_GROUP[unique(col_ann_sc$group)]
  ),
  annotation_name_gp = gpar(fontsize = 8)
)

ht_score <- Heatmap(
  score_z,
  name             = "footprint\ndepth (z)",
  top_annotation   = ha_sc,
  row_names_gp     = gpar(fontsize = 8, fontface = "italic"),
  column_names_gp  = gpar(fontsize = 7),
  col              = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
  cluster_rows     = TRUE,
  cluster_columns  = FALSE,
  column_split     = col_ann_sc$cluster,
  border           = TRUE,
  column_title     = "TF footprint depth: 7 ATAC clusters x genetic groups",
  column_title_gp  = gpar(fontsize = 10, fontface = "bold")
)
draw(ht_score)
pdf(paste0(params$out, "figures/snATAC_NK_TF_footprintDepth_heatmap.pdf"),
    height = 6, width = 9, useDingbats = FALSE)
draw(ht_score); dev.off()
```

# ___TF footprinting: by genetic group (all clusters pooled)___

Compare footprints between ATRXdel / control / MYCNamp / ATRXwtMYCNwt across all NK cells, ignoring cluster identity.

## ___Group coverages — by group only___

```{r}
proj_filt <- addGroupCoverages(
  ArchRProj    = proj_filt,
  groupBy      = "group",
  minCells     = 5,
  maxCells     = 500,
  maxFragments = 25 * 10^6,
  force        = TRUE,
  verbose      = TRUE
)
```

## ___Footprints — by group___

```{r}
footprints_group <- getFootprints(
  ArchRProj = proj_filt,
  positions = motif_positions[tf_motif_names],
  groupBy   = "group",
  flank     = 250
)
cat("Group footprint columns:", paste(colnames(footprints_group), collapse = ", "), "\n")
```

## ___Plot: by group (ArchR native)___

```{r fig.height=4, fig.width=7}
# Map "._.Rep1"-suffixed column names to canonical group names for colors
grp_col_to_canonical <- function(col_names) {
  x <- sub("\\._\\.Rep.*$", "", col_names)
  remap_group(x)
}

grp_pal <- setNames(
  COL_GROUP[grp_col_to_canonical(colnames(footprints_group))],
  colnames(footprints_group)
)
grp_pal[is.na(grp_pal)] <- "gray50"

tryCatch({
  plotFootprints(
    seFoot       = footprints_group,
    ArchRProj    = proj_filt,
    normMethod   = "Subtract",
    plotName     = "Footprints_all-clusters_by-group",
    addDOC       = FALSE,
    smoothWindow = 10,
    pal          = grp_pal,
    plotDir      = paste0(params$out, "figures/footprints/")
  )
  message("Footprint PDF saved: by-group")
}, error = function(e) message("Footprint error (by-group): ", conditionMessage(e)))
```

## ___Plot: by group (ggplot2)___

```{r fig.height=5, fig.width=10}
mat_grp <- assay(footprints_group, "footprint")
pos_seq <- seq(-250, 250, length.out = ncol(mat_grp))

df_grp <- bind_rows(lapply(seq_len(nrow(tf_motif_map_valid)), function(i) {
  motif <- tf_motif_map_valid$motif_name[i]
  tf    <- tf_motif_map_valid$TF[i]
  if (!(motif %in% rownames(mat_grp))) return(NULL)
  row_i <- which(rownames(mat_grp) == motif)
  bind_rows(lapply(colnames(mat_grp), function(g) {
    data.frame(
      TF        = tf,
      group     = grp_col_to_canonical(g),
      position  = pos_seq,
      footprint = as.numeric(mat_grp[row_i, g]),
      stringsAsFactors = FALSE
    )
  }))
})) %>%
  filter(!is.na(group)) %>%
  mutate(group = factor(group, levels = GROUP_ORDER))

p_grp <- ggplot(df_grp, aes(x = position, y = footprint, color = group)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.3) +
  geom_line(linewidth = 0.6, alpha = 0.85) +
  facet_wrap(~ TF, scales = "free_y", ncol = 4) +
  scale_color_manual(values = COL_GROUP) +
  labs(
    title    = "TF footprints — all NK cells, by genetic group",
    subtitle = "bias-subtracted, all clusters pooled",
    x        = "Distance from motif center (bp)",
    y        = "Normalized insertion frequency",
    color    = "genetic group"
  ) +
  theme(legend.position = "top", strip.text = element_text(size = 8, face = "italic"),
        panel.spacing = unit(2, "mm"))

tryCatch(print(p_grp), error = function(e) message(conditionMessage(e)))
tryCatch({
  pdf(paste0(params$out, "figures/snATAC_NK_footprints_by-group.pdf"),
      height = 5, width = 10, useDingbats = FALSE)
  print(p_grp); dev.off()
}, error = function(e) { try(dev.off(), silent = TRUE); message(conditionMessage(e)) })
```

# ___TF footprinting: by cluster (ggplot2)___

```{r fig.height=5, fig.width=10}
mat_cl  <- assay(footprints_cluster, "footprint")
pos_seq <- seq(-250, 250, length.out = ncol(mat_cl))

df_cl <- bind_rows(lapply(seq_len(nrow(tf_motif_map_valid)), function(i) {
  motif <- tf_motif_map_valid$motif_name[i]
  tf    <- tf_motif_map_valid$TF[i]
  if (!(motif %in% rownames(mat_cl))) return(NULL)
  row_i <- which(rownames(mat_cl) == motif)
  bind_rows(lapply(colnames(mat_cl), function(g) {
    data.frame(
      TF        = tf,
      cluster   = cl_col_to_canonical(g),
      position  = pos_seq,
      footprint = as.numeric(mat_cl[row_i, g]),
      stringsAsFactors = FALSE
    )
  }))
})) %>%
  filter(!is.na(cluster)) %>%
  mutate(cluster = factor(cluster, levels = CLUSTER_ORDER))

p_cl <- ggplot(df_cl, aes(x = position, y = footprint, color = cluster)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.3) +
  geom_line(linewidth = 0.6, alpha = 0.85) +
  facet_wrap(~ TF, scales = "free_y", ncol = 4) +
  scale_color_manual(values = COL_ATAC_CLUSTER) +
  labs(
    title    = "TF footprints — all genetic groups, by ATAC cluster",
    subtitle = "bias-subtracted, all groups pooled",
    x        = "Distance from motif center (bp)",
    y        = "Normalized insertion frequency",
    color    = "cluster"
  ) +
  theme(legend.position = "top", strip.text = element_text(size = 8, face = "italic"),
        panel.spacing = unit(2, "mm"))

tryCatch(print(p_cl), error = function(e) message(conditionMessage(e)))
tryCatch({
  pdf(paste0(params$out, "figures/snATAC_NK_footprints_by-cluster.pdf"),
      height = 5, width = 10, useDingbats = FALSE)
  print(p_cl); dev.off()
}, error = function(e) { try(dev.off(), silent = TRUE); message(conditionMessage(e)) })
```

# ___TF footprinting: control samples only, by cluster (C1–C7)___

Healthy donor NK cells only — isolates cluster-intrinsic TF binding differences
from any confounding effect of genetic group (ATRXdel / MYCNamp / ATRXwtMYCNwt).

```{r}
# ctrl_cluster = Clusters_res1 for control cells, NA for all others
# ArchR skips NA cells during addGroupCoverages
proj_filt$ctrl_cluster <- ifelse(
  as.character(proj_filt$group) == "control",
  as.character(proj_filt$Clusters_res1),
  NA_character_
)
cat("Control cells per cluster:\n")
print(table(proj_filt$ctrl_cluster, useNA = "always"))
```

```{r}
proj_filt <- addGroupCoverages(
  ArchRProj    = proj_filt,
  groupBy      = "ctrl_cluster",
  minCells     = 5,
  maxCells     = 500,
  maxFragments = 25 * 10^6,
  force        = TRUE,
  verbose      = TRUE
)
```

```{r}
footprints_ctrl <- getFootprints(
  ArchRProj = proj_filt,
  positions = motif_positions[tf_motif_names],
  groupBy   = "ctrl_cluster",
  flank     = 250
)
cat("Control footprint columns:", paste(colnames(footprints_ctrl), collapse = ", "), "\n")
```

```{r fig.height=4, fig.width=7}
# ArchR native plot — pal names must match stripped colnames (no ._.Rep* suffix)
ctrl_cl_canonical <- sub("\\._\\..*$", "", colnames(footprints_ctrl))
ctrl_pal <- setNames(
  COL_ATAC_CLUSTER[ctrl_cl_canonical],
  ctrl_cl_canonical
)
ctrl_pal[is.na(ctrl_pal)] <- "gray50"

tryCatch({
  plotFootprints(
    seFoot       = footprints_ctrl,
    ArchRProj    = proj_filt,
    normMethod   = "Subtract",
    plotName     = "Footprints_control-only_by-cluster",
    addDOC       = FALSE,
    smoothWindow = 10,
    pal          = ctrl_pal
  )
  message("Footprint PDF saved: control-only by-cluster")
}, error = function(e) message("Footprint error (ctrl by-cluster): ", conditionMessage(e)))
```

```{r fig.height=5, fig.width=10}
mat_ctrl <- assay(footprints_ctrl, "footprint")
pos_seq  <- seq(-250, 250, length.out = ncol(mat_ctrl))

df_ctrl <- bind_rows(lapply(seq_len(nrow(tf_motif_map_valid)), function(i) {
  motif <- tf_motif_map_valid$motif_name[i]
  tf    <- tf_motif_map_valid$TF[i]
  if (!(motif %in% rownames(mat_ctrl))) return(NULL)
  row_i <- which(rownames(mat_ctrl) == motif)
  bind_rows(lapply(colnames(mat_ctrl), function(g) {
    data.frame(
      TF        = tf,
      cluster   = sub("\\._\\..*$", "", g),
      position  = pos_seq,
      footprint = as.numeric(mat_ctrl[row_i, g]),
      stringsAsFactors = FALSE
    )
  }))
})) %>%
  dplyr::filter(!is.na(cluster)) %>%
  dplyr::mutate(cluster = factor(cluster, levels = CLUSTER_ORDER))

p_ctrl <- ggplot(df_ctrl, aes(x = position, y = footprint, color = cluster)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.3) +
  geom_line(linewidth = 0.6, alpha = 0.85) +
  facet_wrap(~ TF, scales = "free_y", ncol = 4) +
  scale_color_manual(values = COL_ATAC_CLUSTER) +
  labs(
    title    = "TF footprints — control samples only, by ATAC cluster",
    subtitle = "bias-subtracted; healthy donor NK cells",
    x        = "Distance from motif center (bp)",
    y        = "Normalized insertion frequency",
    color    = "cluster"
  ) +
  theme(legend.position = "top", strip.text = element_text(size = 8, face = "italic"),
        panel.spacing = unit(2, "mm"))

tryCatch(print(p_ctrl), error = function(e) message(conditionMessage(e)))
tryCatch({
  pdf(paste0(params$out, "figures/snATAC_NK_footprints_control-only_by-cluster.pdf"),
      height = 5, width = 10, useDingbats = FALSE)
  print(p_ctrl); dev.off()
}, error = function(e) { try(dev.off(), silent = TRUE); message(conditionMessage(e)) })
```

# ___TF motif activity: quantitative analysis (MotifMatrix z-scores)___

Per-cell chromVAR deviation z-scores provide statistical power to test group and
cluster differences for each TF motif.

## ___Extract MotifMatrix z-scores___

```{r}
mm   <- getMatrixFromProject(proj_filt, useMatrix = "MotifMatrix")
zMat <- assays(mm)[["z"]]
rownames(zMat) <- sub("^z:", "", rownames(zMat))
cat("MotifMatrix z dimensions:", dim(zMat), "\n")

# Subset to our TF list
tf_names_z <- intersect(tf_motif_map_valid$motif_name, rownames(zMat))
cat("TFs in our list found in MotifMatrix:", length(tf_names_z), "/",
    nrow(tf_motif_map_valid), "\n")
zMat_sub <- zMat[tf_names_z, , drop = FALSE]

# Map motif name → TF display name
motif_to_tf <- setNames(tf_motif_map_valid$TF,
                        tf_motif_map_valid$motif_name)[tf_names_z]

# Per-cell metadata aligned to zMat columns
cell_meta_z <- data.frame(
  cell    = colnames(zMat_sub),
  group   = factor(remap_group(as.character(proj_filt$group)),   levels = GROUP_ORDER),
  cluster = factor(as.character(proj_filt$Clusters_res1),        levels = CLUSTER_ORDER),
  stringsAsFactors = FALSE
)
cat("Cells with group annotation:", sum(!is.na(cell_meta_z$group)), "\n")
```

## ___Statistical tests: by genetic group___

```{r}
res_group <- bind_rows(lapply(tf_names_z, function(motif) {
  df <- data.frame(
    z     = as.numeric(zMat_sub[motif, ]),
    group = cell_meta_z$group
  ) %>% dplyr::filter(!is.na(group))

  kw  <- tryCatch(kruskal.test(z ~ group, data = df), error = function(e) NULL)
  pw  <- tryCatch(
    pairwise.wilcox.test(df$z, df$group, p.adjust.method = "BH"),
    error = function(e) NULL
  )

  kw_p  <- if (!is.null(kw))  kw$p.value  else NA_real_
  kw_adj <- p.adjust(kw_p, method = "BH", n = length(tf_names_z))

  pw_df <- if (!is.null(pw)) {
    p_mat <- pw$p.value
    pairs <- which(!is.na(p_mat), arr.ind = TRUE)
    data.frame(
      motif      = motif,
      TF         = motif_to_tf[motif],
      comparison = "group",
      kw_p       = kw_p,
      kw_p_adj   = kw_adj,
      pair       = paste0(rownames(p_mat)[pairs[, 1]], "_vs_",
                          colnames(p_mat)[pairs[, 2]]),
      pw_p       = p_mat[pairs],
      stringsAsFactors = FALSE
    )
  } else {
    data.frame(motif = motif, TF = motif_to_tf[motif], comparison = "group",
               kw_p = kw_p, kw_p_adj = kw_adj,
               pair = NA_character_, pw_p = NA_real_, stringsAsFactors = FALSE)
  }
  pw_df
}))

res_group$pw_p_adj <- p.adjust(res_group$pw_p, method = "BH")
cat("Significant TFs (Kruskal-Wallis BH < 0.05) by genetic group:",
    sum(res_group$kw_p_adj < 0.05, na.rm = TRUE), "\n")
print(res_group %>% dplyr::filter(kw_p_adj < 0.05) %>%
      dplyr::distinct(TF, kw_p, kw_p_adj) %>% dplyr::arrange(kw_p_adj))
```

## ___Statistical tests: by cluster (all groups)___

```{r}
res_cluster <- bind_rows(lapply(tf_names_z, function(motif) {
  df <- data.frame(
    z       = as.numeric(zMat_sub[motif, ]),
    cluster = cell_meta_z$cluster
  ) %>% dplyr::filter(!is.na(cluster))

  kw  <- tryCatch(kruskal.test(z ~ cluster, data = df), error = function(e) NULL)
  pw  <- tryCatch(
    pairwise.wilcox.test(df$z, df$cluster, p.adjust.method = "BH"),
    error = function(e) NULL
  )

  kw_p  <- if (!is.null(kw)) kw$p.value else NA_real_
  kw_adj <- p.adjust(kw_p, method = "BH", n = length(tf_names_z))

  pw_df <- if (!is.null(pw)) {
    p_mat <- pw$p.value
    pairs <- which(!is.na(p_mat), arr.ind = TRUE)
    data.frame(
      motif      = motif,
      TF         = motif_to_tf[motif],
      comparison = "cluster",
      kw_p       = kw_p,
      kw_p_adj   = kw_adj,
      pair       = paste0(rownames(p_mat)[pairs[, 1]], "_vs_",
                          colnames(p_mat)[pairs[, 2]]),
      pw_p       = p_mat[pairs],
      stringsAsFactors = FALSE
    )
  } else {
    data.frame(motif = motif, TF = motif_to_tf[motif], comparison = "cluster",
               kw_p = kw_p, kw_p_adj = kw_adj,
               pair = NA_character_, pw_p = NA_real_, stringsAsFactors = FALSE)
  }
  pw_df
}))

res_cluster$pw_p_adj <- p.adjust(res_cluster$pw_p, method = "BH")
cat("Significant TFs (Kruskal-Wallis BH < 0.05) by cluster:",
    sum(res_cluster$kw_p_adj < 0.05, na.rm = TRUE), "\n")
print(res_cluster %>% dplyr::filter(kw_p_adj < 0.05) %>%
      dplyr::distinct(TF, kw_p, kw_p_adj) %>% dplyr::arrange(kw_p_adj))
```

## ___Statistical tests: by cluster, control cells only___

```{r}
ctrl_mask <- !is.na(cell_meta_z$group) & cell_meta_z$group == "control"

res_ctrl_cluster <- bind_rows(lapply(tf_names_z, function(motif) {
  df <- data.frame(
    z       = as.numeric(zMat_sub[motif, ctrl_mask]),
    cluster = cell_meta_z$cluster[ctrl_mask]
  ) %>% dplyr::filter(!is.na(cluster))

  if (length(unique(df$cluster)) < 2) {
    return(data.frame(motif = motif, TF = motif_to_tf[motif],
                      comparison = "cluster_ctrl",
                      kw_p = NA_real_, kw_p_adj = NA_real_,
                      pair = NA_character_, pw_p = NA_real_,
                      stringsAsFactors = FALSE))
  }

  kw  <- tryCatch(kruskal.test(z ~ cluster, data = df), error = function(e) NULL)
  pw  <- tryCatch(
    pairwise.wilcox.test(df$z, df$cluster, p.adjust.method = "BH"),
    error = function(e) NULL
  )

  kw_p  <- if (!is.null(kw)) kw$p.value else NA_real_
  kw_adj <- p.adjust(kw_p, method = "BH", n = length(tf_names_z))

  pw_df <- if (!is.null(pw)) {
    p_mat <- pw$p.value
    pairs <- which(!is.na(p_mat), arr.ind = TRUE)
    data.frame(
      motif      = motif,
      TF         = motif_to_tf[motif],
      comparison = "cluster_ctrl",
      kw_p       = kw_p,
      kw_p_adj   = kw_adj,
      pair       = paste0(rownames(p_mat)[pairs[, 1]], "_vs_",
                          colnames(p_mat)[pairs[, 2]]),
      pw_p       = p_mat[pairs],
      stringsAsFactors = FALSE
    )
  } else {
    data.frame(motif = motif, TF = motif_to_tf[motif], comparison = "cluster_ctrl",
               kw_p = kw_p, kw_p_adj = kw_adj,
               pair = NA_character_, pw_p = NA_real_, stringsAsFactors = FALSE)
  }
  pw_df
}))

res_ctrl_cluster$pw_p_adj <- p.adjust(res_ctrl_cluster$pw_p, method = "BH")
cat("Significant TFs (Kruskal-Wallis BH < 0.05) by cluster in controls:",
    sum(res_ctrl_cluster$kw_p_adj < 0.05, na.rm = TRUE), "\n")
print(res_ctrl_cluster %>% dplyr::filter(kw_p_adj < 0.05) %>%
      dplyr::distinct(TF, kw_p, kw_p_adj) %>% dplyr::arrange(kw_p_adj))
```

## ___Save results___

```{r}
out_dir_stats <- paste0(params$out, "tables/TF_motif_stats/")
dir.create(out_dir_stats, showWarnings = FALSE, recursive = TRUE)

write.csv(res_group,        paste0(out_dir_stats, "TF_motif_z_by_group.csv"),        row.names = FALSE)
write.csv(res_cluster,      paste0(out_dir_stats, "TF_motif_z_by_cluster.csv"),      row.names = FALSE)
write.csv(res_ctrl_cluster, paste0(out_dir_stats, "TF_motif_z_by_cluster_ctrl.csv"), row.names = FALSE)
message("Saved stats tables to ", out_dir_stats)
```

## ___Heatmap: median z-score per TF × group___

```{r fig.height=6, fig.width=5}
z_by_group <- t(apply(zMat_sub, 1, function(z) {
  tapply(z, cell_meta_z$group, median, na.rm = TRUE)
}))[, GROUP_ORDER, drop = FALSE]
rownames(z_by_group) <- motif_to_tf[rownames(z_by_group)]

# Keep only TFs significant by group (kw_p_adj < 0.05), or all if none pass
sig_tfs_group <- res_group %>%
  dplyr::filter(kw_p_adj < 0.05) %>%
  dplyr::pull(TF) %>% unique()
z_plot_group  <- if (length(sig_tfs_group) > 1) z_by_group[sig_tfs_group, , drop = FALSE] else z_by_group

zsc_group <- t(scale(t(z_plot_group)))
zsc_group <- pmin(pmax(zsc_group, -z_cap), z_cap)

col_fun_g <- colorRamp2(c(-z_cap, 0, z_cap),
                        c("#2166AC", "white", "#D6604D"))

ht_group <- Heatmap(
  zsc_group,
  name            = "z-score\n(row-scaled)",
  col             = col_fun_g,
  top_annotation  = HeatmapAnnotation(
    Group = GROUP_ORDER,
    col   = list(Group = COL_GROUP),
    show_annotation_name = FALSE
  ),
  cluster_columns = FALSE,
  cluster_rows    = TRUE,
  show_row_names  = TRUE,
  row_names_gp    = gpar(fontsize = 7),
  column_names_gp = gpar(fontsize = 8),
  column_title    = "TF motif activity by genetic group",
  heatmap_legend_param = list(title_gp = gpar(fontsize = 8),
                              labels_gp = gpar(fontsize = 7))
)
draw(ht_group)
pdf(paste0(params$out, "figures/TF_motif_z_heatmap_by_group.pdf"),
    width = 5, height = 6, useDingbats = FALSE)
draw(ht_group); dev.off()
```

## ___Heatmap: median z-score per TF × cluster___

```{r fig.height=6, fig.width=5}
z_by_cluster <- t(apply(zMat_sub, 1, function(z) {
  tapply(z, cell_meta_z$cluster, median, na.rm = TRUE)
}))[, CLUSTER_ORDER, drop = FALSE]
rownames(z_by_cluster) <- motif_to_tf[rownames(z_by_cluster)]

sig_tfs_cluster <- res_cluster %>%
  dplyr::filter(kw_p_adj < 0.05) %>%
  dplyr::pull(TF) %>% unique()
z_plot_cluster  <- if (length(sig_tfs_cluster) > 1) z_by_cluster[sig_tfs_cluster, , drop = FALSE] else z_by_cluster

zsc_cluster <- t(scale(t(z_plot_cluster)))
zsc_cluster <- pmin(pmax(zsc_cluster, -z_cap), z_cap)

col_fun_cl <- colorRamp2(c(-z_cap, 0, z_cap),
                         c("#2166AC", "white", "#D6604D"))

ht_cluster <- Heatmap(
  zsc_cluster,
  name            = "z-score\n(row-scaled)",
  col             = col_fun_cl,
  top_annotation  = HeatmapAnnotation(
    Cluster = CLUSTER_ORDER,
    col     = list(Cluster = COL_ATAC_CLUSTER),
    show_annotation_name = FALSE
  ),
  cluster_columns = FALSE,
  cluster_rows    = TRUE,
  show_row_names  = TRUE,
  row_names_gp    = gpar(fontsize = 7),
  column_names_gp = gpar(fontsize = 8),
  column_title    = "TF motif activity by ATAC cluster",
  heatmap_legend_param = list(title_gp = gpar(fontsize = 8),
                              labels_gp = gpar(fontsize = 7))
)
draw(ht_cluster)
pdf(paste0(params$out, "figures/TF_motif_z_heatmap_by_cluster.pdf"),
    width = 5, height = 6, useDingbats = FALSE)
draw(ht_cluster); dev.off()
```

## ___Heatmap: median z-score per TF × cluster, control cells only___

```{r fig.height=6, fig.width=5}
z_by_ctrl_cl <- t(apply(zMat_sub[, ctrl_mask, drop = FALSE], 1, function(z) {
  tapply(z, cell_meta_z$cluster[ctrl_mask], median, na.rm = TRUE)
}))
# Keep only CLUSTER_ORDER columns that exist
ctrl_cl_cols <- intersect(CLUSTER_ORDER, colnames(z_by_ctrl_cl))
z_by_ctrl_cl <- z_by_ctrl_cl[, ctrl_cl_cols, drop = FALSE]
rownames(z_by_ctrl_cl) <- motif_to_tf[rownames(z_by_ctrl_cl)]

sig_tfs_ctrl_cl <- res_ctrl_cluster %>%
  dplyr::filter(kw_p_adj < 0.05) %>%
  dplyr::pull(TF) %>% unique()
z_plot_ctrl_cl  <- if (length(sig_tfs_ctrl_cl) > 1) z_by_ctrl_cl[sig_tfs_ctrl_cl, , drop = FALSE] else z_by_ctrl_cl

zsc_ctrl_cl <- t(scale(t(z_plot_ctrl_cl)))
zsc_ctrl_cl <- pmin(pmax(zsc_ctrl_cl, -z_cap), z_cap)

ht_ctrl_cl <- Heatmap(
  zsc_ctrl_cl,
  name            = "z-score\n(row-scaled)",
  col             = colorRamp2(c(-z_cap, 0, z_cap), c("#2166AC", "white", "#D6604D")),
  top_annotation  = HeatmapAnnotation(
    Cluster = ctrl_cl_cols,
    col     = list(Cluster = COL_ATAC_CLUSTER[ctrl_cl_cols]),
    show_annotation_name = FALSE
  ),
  cluster_columns = FALSE,
  cluster_rows    = TRUE,
  show_row_names  = TRUE,
  row_names_gp    = gpar(fontsize = 7),
  column_names_gp = gpar(fontsize = 8),
  column_title    = "TF motif activity by cluster — control cells only",
  heatmap_legend_param = list(title_gp = gpar(fontsize = 8),
                              labels_gp = gpar(fontsize = 7))
)
draw(ht_ctrl_cl)
pdf(paste0(params$out, "figures/TF_motif_z_heatmap_by_cluster_ctrl.pdf"),
    width = 5, height = 6, useDingbats = FALSE)
draw(ht_ctrl_cl); dev.off()
```


# ___PART II___

# ___GSK3B differential chromatin accessibility___

**Reviewer #2, question #2**

Assess chromatin accessibility at the GSK3B gene body and promoter region (TSS +/- 2 kb)
across all 7 ATAC clusters comparing genetic groups.

## ___Define GSK3B locus___

```{r}
# Get GSK3B coordinates from the project's gene annotation
gene_anno  <- getGeneAnnotation(proj_filt)
gsk3b_gene <- gene_anno$genes[which(gene_anno$genes$symbol == "GSK3B")]

if (length(gsk3b_gene) == 0) {
  # Fallback: hg38 coordinates (chr3, minus strand)
  gsk3b_gene <- GRanges("chr3", IRanges(119508042, 119742202), strand = "-",
                         symbol = "GSK3B")
  cat("GSK3B gene annotation not found in project; using hg38 hardcoded coordinates.\n")
} else {
  cat("GSK3B gene:", as.character(seqnames(gsk3b_gene)),
      start(gsk3b_gene), "-", end(gsk3b_gene),
      "strand:", as.character(strand(gsk3b_gene)), "\n")
}

# Define promoter: TSS +/- 2 kb
tss_w <- 2000L
if (as.character(strand(gsk3b_gene)) == "-") {
  gsk3b_tss <- end(gsk3b_gene)
} else {
  gsk3b_tss <- start(gsk3b_gene)
}

gsk3b_promoter <- GRanges(
  seqnames = seqnames(gsk3b_gene),
  ranges   = IRanges(gsk3b_tss - tss_w, gsk3b_tss + tss_w),
  strand   = strand(gsk3b_gene)
)

# Extended locus for browser track: gene body + 5 kb flanks
gsk3b_locus <- GRanges(
  seqnames = seqnames(gsk3b_gene),
  ranges   = IRanges(
    pmax(1L, start(gsk3b_gene) - 5000L),
    end(gsk3b_gene) + 5000L
  )
)

cat("GSK3B promoter (TSS +/-2kb):",
    as.character(seqnames(gsk3b_promoter)),
    start(gsk3b_promoter), "-", end(gsk3b_promoter), "\n")
cat("GSK3B locus (+/-5kb):",
    as.character(seqnames(gsk3b_locus)),
    start(gsk3b_locus), "-", end(gsk3b_locus), "\n")
```

## ___Browser track at GSK3B locus___

```{r fig.height=6, fig.width=8}
# Plot for each cluster: one track per genetic group within the cluster
for (cl in CLUSTER_ORDER) {
  cl_groups <- paste0(cl, "_", GROUP_ORDER)
  cl_groups <- cl_groups[cl_groups %in%
                           names(getGroupCoverages(proj_filt, groupBy = "clust_group"))]
  if (length(cl_groups) == 0) {
    # try without filtering — plotBrowserTrack handles group selection
    cl_groups <- paste0(cl, "_", GROUP_ORDER)
  }

  tryCatch({
    p_track <- plotBrowserTrack(
      ArchRProj   = proj_filt,
      groupBy     = "clust_group",
      useGroups   = cl_groups,
      geneSymbol  = "GSK3B",
      region      = gsk3b_locus,
      pal         = setNames(COL_GROUP[sub(paste0("^", cl, "_"), "", cl_groups)], cl_groups),
      plotSummary = c("bulkTrack", "featureTrack", "geneTrack"),
      sizes       = c(10, 1.5, 2),
      tileSize    = 100,
      upstream    = 5000,
      downstream  = 5000
    )
    grid::grid.newpage()
    grid::grid.draw(p_track)
    pdf(paste0(params$out, "figures/snATAC_NK_GSK3B_browserTrack_", cl, ".pdf"),
        height = 6, width = 8, useDingbats = FALSE)
    grid::grid.draw(p_track); dev.off()
    message("Browser track saved for cluster ", cl)
  }, error = function(e) {
    message("Browser track error for ", cl, ": ", conditionMessage(e))
  })
}
```

## ___Peak accessibility at GSK3B gene body + promoter___

Extract peaks overlapping the GSK3B locus, aggregate accessibility per cell,
and compare genetic groups within each cluster.

```{r}
peak_se <- getMatrixFromProject(proj_filt, useMatrix = "PeakMatrix")

# Peaks overlapping GSK3B gene body OR promoter
gsk3b_query <- c(gsk3b_gene, gsk3b_promoter)
peak_gr     <- getPeakSet(proj_filt)
ov_hits     <- findOverlaps(peak_gr, gsk3b_query, ignore.strand = TRUE)
gsk3b_peak_idx <- unique(queryHits(ov_hits))

cat("Peaks overlapping GSK3B locus (gene + promoter):", length(gsk3b_peak_idx), "\n")
if (length(gsk3b_peak_idx) == 0) {
  cat("WARNING: No peaks found at GSK3B locus. Check chromosome naming.\n")
}
```

```{r}
if (length(gsk3b_peak_idx) > 0) {
  acc_mat <- assay(peak_se, "PeakMatrix")[gsk3b_peak_idx, , drop = FALSE]

  # Determine which peaks are in promoter vs gene body
  is_promoter <- gsk3b_peak_idx %in% queryHits(findOverlaps(
    peak_gr, gsk3b_promoter, ignore.strand = TRUE
  ))
  is_genebody <- gsk3b_peak_idx %in% queryHits(findOverlaps(
    peak_gr, gsk3b_gene, ignore.strand = TRUE
  ))

  cat("Promoter peaks:", sum(is_promoter), "\n")
  cat("Gene body peaks:", sum(is_genebody), "\n")

  # Per-cell sum accessibility over GSK3B region (binarised peak matrix → count)
  cell_acc <- data.frame(
    cell       = colnames(acc_mat),
    acc_all    = colSums(acc_mat),
    acc_promo  = if (any(is_promoter)) colSums(acc_mat[is_promoter, , drop = FALSE]) else 0,
    acc_body   = if (any(is_genebody)) colSums(acc_mat[is_genebody, , drop = FALSE]) else 0
  )

  cell_df_gsk <- data.frame(
    Clusters_res1 = as.character(proj_filt$Clusters_res1),
    group         = as.character(proj_filt$group),
    cell          = getCellNames(proj_filt),
    stringsAsFactors = FALSE
  )
  cell_df_gsk <- left_join(cell_df_gsk, cell_acc, by = "cell") %>%
    mutate(
      Clusters_res1 = factor(Clusters_res1, levels = CLUSTER_ORDER),
      group         = factor(group, levels = GROUP_ORDER)
    )

  head(cell_df_gsk)
}
```

## ___Pseudobulk accessibility at GSK3B per cluster x group___

```{r fig.height=6, fig.width=10}
if (exists("cell_df_gsk") && nrow(cell_df_gsk) > 0 && !all(cell_df_gsk$acc_all == 0)) {

  pb_gsk <- cell_df_gsk %>%
    group_by(Clusters_res1, group) %>%
    summarise(
      n_cells    = n(),
      mean_all   = mean(acc_all,   na.rm = TRUE),
      mean_promo = mean(acc_promo, na.rm = TRUE),
      mean_body  = mean(acc_body,  na.rm = TRUE),
      .groups = "drop"
    ) %>%
    filter(n_cells >= 5)

  # Long format for plotting
  pb_long <- pb_gsk %>%
    pivot_longer(cols = c(mean_promo, mean_body),
                 names_to = "region", values_to = "accessibility") %>%
    mutate(region = recode(region,
                           mean_promo = "GSK3B promoter (TSS +/-2kb)",
                           mean_body  = "GSK3B gene body"))

  p_pb <- ggplot(pb_long, aes(x = Clusters_res1, y = accessibility, fill = group)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    facet_wrap(~ region, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = COL_GROUP) +
    labs(
      title    = "GSK3B chromatin accessibility per ATAC cluster",
      subtitle = "Pseudobulk mean (cells >=5 per clust_group)",
      x        = "ATAC cluster", y = "mean accessibility", fill = "genetic group"
    ) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          legend.position = "top")

  p_pb
  pdf(paste0(params$out, "figures/snATAC_NK_GSK3B_pseudobulk_accessibility.pdf"),
      height = 6, width = 10, useDingbats = FALSE)
  p_pb; dev.off()
}
```

## ___Per-cell violin plots at GSK3B per cluster___

```{r fig.height=5, fig.width=12}
if (exists("cell_df_gsk")) {
  regions_to_plot <- list(
    "GSK3B promoter (TSS +/-2kb)" = "acc_promo",
    "GSK3B gene body"             = "acc_body"
  )

  for (region_label in names(regions_to_plot)) {
    col_nm <- regions_to_plot[[region_label]]
    p_vln <- ggplot(cell_df_gsk,
                    aes(x = group, y = .data[[col_nm]], fill = group)) +
      geom_violin(scale = "width", linewidth = 0.3, alpha = 0.8) +
      geom_boxplot(width = 0.12, fill = "white", outlier.size = 0.3,
                   linewidth = 0.3) +
      facet_wrap(~ Clusters_res1, nrow = 2, labeller = as_labeller(CLUSTER_NAMES)) +
      scale_fill_manual(values = COL_GROUP) +
      labs(
        title    = region_label,
        subtitle = "Per-cell chromatin accessibility — genetic groups within each ATAC cluster",
        x = NULL, y = "peak accessibility", fill = "genetic group"
      ) +
      theme(
        axis.text.x    = element_text(angle = 40, hjust = 1, size = 7),
        legend.position = "none"
      )
    print(p_vln)
    pdf(paste0(params$out, "figures/snATAC_NK_GSK3B_",
               gsub("[^A-Za-z0-9]+", "_", region_label), "_violin.pdf"),
        height = 5, width = 12, useDingbats = FALSE)
    print(p_vln); dev.off()
  }
}
```

## ___Wilcoxon tests: ATRXdel vs control per cluster___

```{r}
if (exists("cell_df_gsk")) {
  regions_stat <- c(
    "GSK3B promoter" = "acc_promo",
    "GSK3B gene body" = "acc_body"
  )

  stat_results <- bind_rows(lapply(names(regions_stat), function(rl) {
    col_nm <- regions_stat[[rl]]
    bind_rows(lapply(CLUSTER_ORDER, function(cl) {
      sub_df <- cell_df_gsk %>% filter(Clusters_res1 == cl)
      bind_rows(lapply(
        list(c("ATRXdel", "control"),
             c("ATRXdel", "MYCNamp"),
             c("ATRXdel", "ATRXwtMYCNwt"),
             c("MYCNamp", "control")),
        function(grp_pair) {
          g1 <- sub_df %>% filter(group == grp_pair[1]) %>% pull(col_nm)
          g2 <- sub_df %>% filter(group == grp_pair[2]) %>% pull(col_nm)
          if (length(g1) < 5 || length(g2) < 5) return(NULL)
          wt <- wilcox.test(g1, g2)
          data.frame(
            region    = rl,
            cluster   = cl,
            group1    = grp_pair[1],
            group2    = grp_pair[2],
            n1        = length(g1),
            n2        = length(g2),
            median1   = median(g1),
            median2   = median(g2),
            p_value   = wt$p.value,
            stringsAsFactors = FALSE
          )
        }
      ))
    }))
  }))

  stat_results$p_adj <- p.adjust(stat_results$p_value, method = "BH")
  stat_results$sig   <- stat_results$p_adj < 0.05
  stat_results <- stat_results %>%
    arrange(region, cluster, p_adj) %>%
    mutate(across(c(p_value, p_adj), ~signif(.x, 3)))

  cat("GSK3B accessibility — Wilcoxon tests (BH-adjusted):\n")
  print(stat_results)

  write.csv(stat_results,
            paste0(params$out, "tables/snATAC_NK_GSK3B_accessibility_wilcoxon.csv"),
            row.names = FALSE)
}
```

## ___Peak-level accessibility heatmap at GSK3B___

```{r fig.height=5, fig.width=8}
if (exists("cell_df_gsk") && length(gsk3b_peak_idx) > 0) {
  # Pseudobulk mean per peak per clust_group
  cg_levels <- levels(proj_filt$clust_group)
  cell_cg        <- as.character(proj_filt$clust_group)
  names(cell_cg) <- getCellNames(proj_filt)

  pb_peak_mat <- vapply(cg_levels, function(cg) {
    cells <- names(cell_cg)[cell_cg == cg]
    cells <- cells[cells %in% colnames(acc_mat)]
    if (length(cells) < 5) return(rep(NA_real_, nrow(acc_mat)))
    rowMeans(acc_mat[, cells, drop = FALSE], na.rm = TRUE)
  }, numeric(nrow(acc_mat)))

  # Row annotation: promoter vs gene body
  peak_region <- ifelse(is_promoter, "promoter", ifelse(is_genebody, "gene body", "other"))
  row_ha <- rowAnnotation(
    region = peak_region,
    col    = list(region = c("promoter" = "#D62728", "gene body" = "#1F77B4",
                             "other" = "gray80")),
    annotation_name_gp = gpar(fontsize = 8)
  )

  col_ann_pk <- data.frame(
    cluster   = sub("_.*", "", colnames(pb_peak_mat)),
    group     = sub("^[^_]+_", "", colnames(pb_peak_mat)),
    row.names = colnames(pb_peak_mat)
  )
  ha_pk <- HeatmapAnnotation(
    cluster = col_ann_pk$cluster,
    group   = col_ann_pk$group,
    col     = list(
      cluster = COL_ATAC_CLUSTER[unique(col_ann_pk$cluster)],
      group   = COL_GROUP[unique(col_ann_pk$group)]
    ),
    annotation_name_gp = gpar(fontsize = 8)
  )

  pb_peak_z <- t(scale(t(pb_peak_mat)))
  pb_peak_z[is.nan(pb_peak_z)] <- 0
  pb_peak_z[pb_peak_z >  2]   <-  2
  pb_peak_z[pb_peak_z < -2]   <- -2

  ht_peak <- Heatmap(
    pb_peak_z,
    name              = "accessibility\n(z-score)",
    top_annotation    = ha_pk,
    left_annotation   = row_ha,
    row_names_gp      = gpar(fontsize = 7),
    column_names_gp   = gpar(fontsize = 7),
    col               = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
    cluster_rows      = TRUE,
    cluster_columns   = FALSE,
    column_split      = col_ann_pk$cluster,
    border            = TRUE,
    column_title      = "GSK3B peak accessibility: 7 clusters x genetic groups",
    column_title_gp   = gpar(fontsize = 10, fontface = "bold"),
    show_row_names    = nrow(pb_peak_z) <= 50
  )
  draw(ht_peak)
  pdf(paste0(params$out, "figures/snATAC_NK_GSK3B_peak_accessibility_heatmap.pdf"),
      height = 5, width = 8, useDingbats = FALSE)
  draw(ht_peak); dev.off()
}
```

# ___TF footprinting: cytotoxic vs immunomodulatory NK (label transfer)___

**Reviewer #2, question #2**

Minimal functional label transfer: hNK_Bm1 + hNK_Bm4 → cytotoxic; hNK_Bm2 + hNK_Bm3 → immunomodulatory.
Compares TF footprints between the two functional categories, and then within each category
across genetic groups.

## ___Add functional label to ArchR project___

```{r}
# Add functional category to cell metadata from majority_vote_label
proj_filt$functional <- factor(
  FUNCTIONAL_MAP[as.character(proj_filt$majority_vote_label)],
  levels = FUNCTIONAL_ORDER
)

# Combined functional × group label for within-category group comparisons
proj_filt$functional_group <- factor(
  ifelse(
    !is.na(proj_filt$functional) & !is.na(proj_filt$group),
    paste0(as.character(proj_filt$functional), "_", as.character(proj_filt$group)),
    NA_character_
  )
)

cat("Functional label distribution:\n")
print(table(proj_filt$functional, useNA = "always"))
cat("\nFunctional × group distribution:\n")
print(table(proj_filt$functional_group, useNA = "always"))
```

## ___Footprints: cytotoxic vs immunomodulatory (all groups pooled)___

```{r}
proj_filt <- addGroupCoverages(
  ArchRProj    = proj_filt,
  groupBy      = "functional",
  minCells     = 10,
  maxCells     = 1000,
  maxFragments = 25 * 10^6,
  force        = TRUE,
  verbose      = TRUE
)

footprints_func <- getFootprints(
  ArchRProj = proj_filt,
  positions = motif_positions[tf_motif_names],
  groupBy   = "functional",
  flank     = 250
)
cat("Functional footprint columns:", paste(colnames(footprints_func), collapse = ", "), "\n")
```

```{r fig.height=5, fig.width=10}
func_col_canonical <- function(col_names)
  sub("\\._\\.Rep.*$", "", col_names)

mat_func <- assay(footprints_func, "footprint")
pos_seq   <- seq(-250, 250, length.out = ncol(mat_func))

df_func <- dplyr::bind_rows(lapply(seq_len(nrow(tf_motif_map_valid)), function(i) {
  motif <- tf_motif_map_valid$motif_name[i]
  tf    <- tf_motif_map_valid$TF[i]
  if (!(motif %in% rownames(mat_func))) return(NULL)
  row_i <- which(rownames(mat_func) == motif)
  dplyr::bind_rows(lapply(colnames(mat_func), function(g) {
    data.frame(
      TF         = tf,
      functional = func_col_canonical(g),
      position   = pos_seq,
      footprint  = as.numeric(mat_func[row_i, g]),
      stringsAsFactors = FALSE
    )
  }))
})) %>%
  dplyr::filter(!is.na(functional)) %>%
  dplyr::mutate(functional = factor(functional, levels = FUNCTIONAL_ORDER))

p_func <- ggplot(df_func, aes(x = position, y = footprint, color = functional)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.3) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  facet_wrap(~ TF, scales = "free_y", ncol = 4) +
  scale_color_manual(values = COL_FUNCTIONAL, labels = c("Immunomodulatory", "Cytotoxic")) +
  labs(
    title    = "TF footprints — cytotoxic vs immunomodulatory NK cells",
    subtitle = "bias-subtracted; hNK_Bm1+Bm4 = cytotoxic, hNK_Bm2+Bm3 = immunomodulatory",
    x        = "Distance from motif center (bp)",
    y        = "Normalized insertion frequency",
    color    = NULL
  ) +
  theme(legend.position = "top", strip.text = element_text(size = 8, face = "italic"),
        panel.spacing = unit(2, "mm"))

tryCatch(print(p_func), error = function(e) message(conditionMessage(e)))
tryCatch({
  pdf(paste0(params$out, "figures/snATAC_NK_footprints_functional-label.pdf"),
      height = 5, width = 10, useDingbats = FALSE)
  print(p_func); dev.off()
}, error = function(e) { try(dev.off(), silent = TRUE); message(conditionMessage(e)) })
```

## ___Footprints: functional × genetic group___

```{r}
proj_filt <- addGroupCoverages(
  ArchRProj    = proj_filt,
  groupBy      = "functional_group",
  minCells     = 5,
  maxCells     = 500,
  maxFragments = 25 * 10^6,
  force        = TRUE,
  verbose      = TRUE
)

footprints_func_grp <- getFootprints(
  ArchRProj = proj_filt,
  positions = motif_positions[tf_motif_names],
  groupBy   = "functional_group",
  flank     = 250
)
cat("Functional × group footprint columns:",
    paste(colnames(footprints_func_grp), collapse = ", "), "\n")
```

```{r fig.height=5, fig.width=12}
# Parse column names: "cytotoxic_ATRXdel._.Rep1" → functional + group
fg_col_canonical <- function(col_names) {
  x <- sub("\\._\\.Rep.*$", "", col_names)    # strip ._.Rep* suffix
  data.frame(
    functional = sub("_([^_]+)$", "", x),      # everything before last _group
    group      = remap_group(sub(".*_", "", x)), # last segment = group
    stringsAsFactors = FALSE
  )
}

mat_fg  <- assay(footprints_func_grp, "footprint")
pos_seq <- seq(-250, 250, length.out = ncol(mat_fg))
fg_meta <- fg_col_canonical(colnames(mat_fg))

FUNC_GROUP_ORDER <- as.vector(outer(FUNCTIONAL_ORDER, GROUP_ORDER,
                                    paste, sep = "_"))

df_fg <- dplyr::bind_rows(lapply(seq_len(nrow(tf_motif_map_valid)), function(i) {
  motif <- tf_motif_map_valid$motif_name[i]
  tf    <- tf_motif_map_valid$TF[i]
  if (!(motif %in% rownames(mat_fg))) return(NULL)
  row_i <- which(rownames(mat_fg) == motif)
  do.call(rbind, lapply(seq_along(colnames(mat_fg)), function(j) {
    data.frame(
      TF         = tf,
      functional = fg_meta$functional[j],
      group      = fg_meta$group[j],
      position   = pos_seq,
      footprint  = as.numeric(mat_fg[row_i, j]),
      stringsAsFactors = FALSE
    )
  }))
})) %>%
  dplyr::filter(!is.na(functional), !is.na(group)) %>%
  dplyr::mutate(
    functional = factor(functional, levels = FUNCTIONAL_ORDER),
    group      = factor(group,      levels = GROUP_ORDER)
  )

# One panel per functional category, lines colored by group
for (func_cat in FUNCTIONAL_ORDER) {
  df_sub <- dplyr::filter(df_fg, functional == func_cat)
  if (nrow(df_sub) == 0) next

  p_fg <- ggplot(df_sub, aes(x = position, y = footprint, color = group)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.3) +
    geom_line(linewidth = 0.6, alpha = 0.85) +
    facet_wrap(~ TF, scales = "free_y", ncol = 4) +
    scale_color_manual(values = COL_GROUP) +
    labs(
      title    = paste0("TF footprints — ", func_cat, " NK cells, by genetic group"),
      subtitle = "bias-subtracted",
      x        = "Distance from motif center (bp)",
      y        = "Normalized insertion frequency",
      color    = "genetic group"
    ) +
    theme(legend.position = "top", strip.text = element_text(size = 8, face = "italic"),
          panel.spacing = unit(2, "mm"))

  tryCatch(print(p_fg), error = function(e) message(conditionMessage(e)))
  tryCatch({
    pdf(paste0(params$out, "figures/snATAC_NK_footprints_", func_cat, "_by-group.pdf"),
        height = 5, width = 10, useDingbats = FALSE)
    print(p_fg); dev.off()
  }, error = function(e) { try(dev.off(), silent = TRUE); message(conditionMessage(e)) })
}
```

# ___Motif accessibility enrichment: chromVAR by functional label___

**Reviewer #2, question #2**

ChromVAR deviation scores quantify per-cell TF motif accessibility enrichment relative
to background peaks. Here we compare the TFs of interest between cytotoxic and
immunomodulatory NK cells, and across genetic groups within each functional category.

## ___Extract chromVAR deviations for TFs of interest___

```{r}
chrom_func <- tryCatch({
  chrom_se_full <- getMatrixFromProject(proj_filt, useMatrix = "MotifMatrix")
  # Match TFs of interest to MotifMatrix row names
  row_idx <- which(rowData(chrom_se_full)$name %in% tf_motif_map_valid$motif_name)
  if (length(row_idx) == 0)
    row_idx <- which(grepl(paste(TF_LIST, collapse = "|"),
                           rowData(chrom_se_full)$name, ignore.case = TRUE))
  if (length(row_idx) == 0) stop("No TF motifs found in MotifMatrix")

  chrom_se_tf <- chrom_se_full[row_idx, ]
  dev_mat     <- assay(chrom_se_tf, 2)          # assay 2 = z-score deviations
  rownames(dev_mat) <- rowData(chrom_se_tf)$name

  # Remap motif names to TF gene symbols
  motif_to_tf <- setNames(tf_motif_map_valid$TF, tf_motif_map_valid$motif_name)
  rownames(dev_mat) <- ifelse(rownames(dev_mat) %in% names(motif_to_tf),
                              motif_to_tf[rownames(dev_mat)], rownames(dev_mat))

  cell_meta_cv <- data.frame(
    functional = as.character(proj_filt$functional),
    group      = as.character(proj_filt$group),
    func_grp   = paste0(as.character(proj_filt$functional), "_",
                        as.character(proj_filt$group)),
    row.names  = getCellNames(proj_filt),
    stringsAsFactors = FALSE
  )
  list(dev_mat = dev_mat, cell_meta = cell_meta_cv)

}, error = function(e) {
  message("MotifMatrix not available: ", conditionMessage(e))
  NULL
})
```

## ___Heatmap: TF deviation by functional label (cytotoxic vs immunomodulatory)___

```{r fig.height=5, fig.width=4}
if (!is.null(chrom_func)) {
  dev_mat   <- chrom_func$dev_mat
  cell_meta <- chrom_func$cell_meta

  # Pseudobulk mean per functional category
  pb_func <- vapply(FUNCTIONAL_ORDER, function(f) {
    cells <- rownames(cell_meta)[cell_meta$functional == f & !is.na(cell_meta$functional)]
    cells <- cells[cells %in% colnames(dev_mat)]
    if (length(cells) < 10) return(rep(NA_real_, nrow(dev_mat)))
    rowMeans(dev_mat[, cells, drop = FALSE], na.rm = TRUE)
  }, numeric(nrow(dev_mat)))

  pb_func <- pb_func[, colSums(is.na(pb_func)) == 0, drop = FALSE]

  ha_func <- HeatmapAnnotation(
    functional = FUNCTIONAL_ORDER[seq_len(ncol(pb_func))],
    col        = list(functional = COL_FUNCTIONAL),
    simple_anno_size     = unit(3, "mm"),
    annotation_name_side = "left",
    annotation_name_gp   = gpar(fontsize = 8)
  )

  pb_func_z <- t(scale(t(pb_func)))
  pb_func_z[is.nan(pb_func_z)] <- 0
  pb_func_z[pb_func_z >  2] <-  2
  pb_func_z[pb_func_z < -2] <- -2

  ht_func <- Heatmap(
    pb_func_z,
    name              = "deviation\nz-score",
    top_annotation    = ha_func,
    row_names_gp      = gpar(fontsize = 8, fontface = "italic"),
    column_names_gp   = gpar(fontsize = 9),
    col               = colorRamp2(c(-2, 0, 2),
                                   c(COL_FUNCTIONAL["cytotoxic"], "white",
                                     COL_FUNCTIONAL["immunomodulatory"])),
    cluster_rows      = TRUE,
    cluster_columns   = FALSE,
    border            = TRUE,
    column_title      = "ChromVAR: cytotoxic vs immunomodulatory NK",
    column_title_gp   = gpar(fontsize = 9, fontface = "bold")
  )
  draw(ht_func)
  pdf(paste0(params$out, "figures/snATAC_NK_chromVAR_TF_functional-label_heatmap.pdf"),
      height = 5, width = 4, useDingbats = FALSE)
  draw(ht_func); dev.off()
}
```

## ___Heatmap: TF deviation by functional × genetic group___

```{r fig.height=5, fig.width=7}
if (!is.null(chrom_func)) {
  dev_mat   <- chrom_func$dev_mat
  cell_meta <- chrom_func$cell_meta

  fg_levels <- as.vector(outer(FUNCTIONAL_ORDER, GROUP_ORDER, paste, sep = "_"))

  pb_fg <- vapply(fg_levels, function(fg) {
    cells <- rownames(cell_meta)[cell_meta$func_grp == fg & !is.na(cell_meta$functional)]
    cells <- cells[cells %in% colnames(dev_mat)]
    if (length(cells) < 5) return(rep(NA_real_, nrow(dev_mat)))
    rowMeans(dev_mat[, cells, drop = FALSE], na.rm = TRUE)
  }, numeric(nrow(dev_mat)))

  # Drop columns with insufficient cells
  keep_cols <- colSums(is.na(pb_fg)) == 0
  pb_fg     <- pb_fg[, keep_cols, drop = FALSE]
  fg_keep   <- fg_levels[keep_cols]

  fg_ann <- data.frame(
    functional = sub("_.*", "", fg_keep),
    group      = sub("^[^_]+_", "", fg_keep),
    row.names  = fg_keep
  )

  ha_fg <- HeatmapAnnotation(
    functional = fg_ann$functional,
    group      = fg_ann$group,
    col        = list(
      functional = COL_FUNCTIONAL[unique(fg_ann$functional)],
      group      = COL_GROUP[unique(fg_ann$group)]
    ),
    simple_anno_size     = unit(3, "mm"),
    annotation_name_side = "left",
    annotation_name_gp   = gpar(fontsize = 8)
  )

  pb_fg_z <- t(scale(t(pb_fg)))
  pb_fg_z[is.nan(pb_fg_z)] <- 0
  pb_fg_z[pb_fg_z >  2] <-  2
  pb_fg_z[pb_fg_z < -2] <- -2

  ht_fg <- Heatmap(
    pb_fg_z,
    name              = "deviation\nz-score",
    top_annotation    = ha_fg,
    row_names_gp      = gpar(fontsize = 8, fontface = "italic"),
    column_names_gp   = gpar(fontsize = 7),
    col               = colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
    cluster_rows      = TRUE,
    cluster_columns   = FALSE,
    column_split      = fg_ann$functional,
    column_gap        = unit(3, "mm"),
    border            = TRUE,
    column_title_gp   = gpar(fontsize = 9, fontface = "bold"),
    column_title      = "ChromVAR TF deviation: functional label × genetic group"
  )
  draw(ht_fg)
  pdf(paste0(params$out, "figures/snATAC_NK_chromVAR_TF_functional-x-group_heatmap.pdf"),
      height = 5, width = 7, useDingbats = FALSE)
  draw(ht_fg); dev.off()
}
```

## ___Violin plots: TF deviation per functional label (per TF)___

```{r fig.height=5, fig.width=10}
if (!is.null(chrom_func)) {
  dev_mat   <- chrom_func$dev_mat
  cell_meta <- chrom_func$cell_meta

  df_dev <- as.data.frame(t(dev_mat)) %>%
    tibble::rownames_to_column("cell") %>%
    tidyr::pivot_longer(-cell, names_to = "TF", values_to = "deviation") %>%
    dplyr::left_join(tibble::rownames_to_column(cell_meta, "cell"), by = "cell") %>%
    dplyr::filter(!is.na(functional)) %>%
    dplyr::mutate(
      functional = factor(functional, levels = FUNCTIONAL_ORDER),
      group      = factor(group, levels = GROUP_ORDER)
    )

  p_dev_func <- ggplot(df_dev, aes(x = functional, y = deviation, fill = functional)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0, alpha = 0.9) +
    ggpubr::stat_compare_means(
      comparisons = list(c("immunomodulatory", "cytotoxic")),
      method      = "wilcox.test",
      label       = "p.signif",
      size        = 3,
      symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
                         symbols   = c("****", "***", "**", "*", "ns"))
    ) +
    facet_wrap(~ TF, scales = "free_y", ncol = 4) +
    scale_fill_manual(values = COL_FUNCTIONAL) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
    labs(
      title = "ChromVAR TF deviation: cytotoxic vs immunomodulatory NK",
      x = NULL, y = "chromVAR deviation z-score", fill = NULL
    ) +
    theme(legend.position = "top",
          strip.text      = element_text(size = 8, face = "italic"),
          axis.text.x     = element_blank(),
          axis.ticks.x    = element_blank())

  tryCatch(print(p_dev_func), error = function(e) message(conditionMessage(e)))
  tryCatch({
    pdf(paste0(params$out, "figures/snATAC_NK_chromVAR_TF_functional-label_violins.pdf"),
        height = 5, width = 10, useDingbats = FALSE)
    print(p_dev_func); dev.off()
  }, error = function(e) { try(dev.off(), silent = TRUE); message(conditionMessage(e)) })
}
```

## ___Violin plots: TF deviation per group, split by functional label___

```{r fig.height=8, fig.width=10}
if (!is.null(chrom_func)) {
  # One row per TF, facet column = functional category, fill = group
  p_dev_grp <- ggplot(df_dev, aes(x = group, y = deviation, fill = group)) +
    geom_violin(scale = "width", trim = TRUE, linewidth = 0, alpha = 0.9) +
    facet_grid(TF ~ functional, scales = "free_y") +
    scale_fill_manual(values = COL_GROUP) +
    scale_x_discrete(labels = function(x) {
      lbl <- c(control = "CTRL", ATRXdel = "ATRXdel",
                MYCNamp = "MYCNamp", ATRXwtMYCNwt = "ATRXwt")
      ifelse(x %in% names(lbl), lbl[x], x)
    }) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    labs(
      title = "ChromVAR TF deviation by genetic group and functional label",
      x = NULL, y = "chromVAR deviation z-score", fill = "genetic group"
    ) +
    theme(legend.position  = "top",
          strip.text.x     = element_text(size = 9, face = "bold"),
          strip.text.y     = element_text(size = 7, face = "italic"),
          axis.text.x      = element_text(angle = 45, hjust = 1, size = 7),
          panel.spacing    = unit(1.5, "mm"))

  tryCatch(print(p_dev_grp), error = function(e) message(conditionMessage(e)))
  tryCatch({
    n_tfs <- length(unique(df_dev$TF))
    pdf(paste0(params$out,
               "figures/snATAC_NK_chromVAR_TF_group-x-functional_violins.pdf"),
        height = max(6, n_tfs * 1.5), width = 7, useDingbats = FALSE)
    print(p_dev_grp); dev.off()
  }, error = function(e) { try(dev.off(), silent = TRUE); message(conditionMessage(e)) })
}
```

# ___Save updated ArchR project___

```{r eval=FALSE}
saveRDS(proj_filt,
        file = paste0(params$out,
                      "ArchR_v20260203/ATAC_analysis_v20260625_withClustGroup.Rds"))
```

```{r render, eval=FALSE, include=FALSE}
rmarkdown::render(
  "/home/rstudio/code/NK-NB_study/code/39_20260625_snATAC_NK_TF_footprinting.Rmd",
  output_file = paste0(params$out, "html/39_20260625_snATAC_NK_TF_footprinting.html")
)
```

# ___The End___


